### 2018-11-20 ms/spaceness.tex
added ms folder

\documentclass[12pt]{amsart}
\usepackage{geometry} % see geometry.pdf on how to lay out the page. There's lots.
\geometry{a4paper} % or letter or a5paper or ... etc
\usepackage{graphicx}
% \geometry{landscape} % rotated page geometry

% See the ``Article customise'' template for come common customisations

\title{ Evolution on the Final Frontier: How Dispersal and Competition Shape Genetic Variation in Continuous Space }
\author{C. J. Battey, Peter Ralph, Andrew Kern}
\date{November 2018} % delete this line to display the current date

%%% BEGIN DOCUMENT
\begin{document}

\maketitle
\tableofcontents

\section{Abstract}
Random mating is a core assumption of many population genetic models, but likely occurs in few real populations. Instead most organisms mate with nearby individuals and disperse across only a small part of the full range of a species. This process creates the pattern known as isolation by distance (IBD; Wright 1942) -- a positive correlation between genetic and geographic distances among individuals. Though this pattern is both expected from first principles and widely observed in natural populations (Aguillon et al. 2017, Pinto et al. 2013, Battey et al. 2018), existing population genetic models either exclude it entirely (by assuming random mating) or take the heuristic approach of binning the landscape into discrete blocks of randomly mating populations that exchange migrants each generation.

Efforts to model genomes evolving in continuous space have a long history but have been limited by computational inefficiency and a lack of analytic tractability. Both Wright (1942) and Malecót (196X) developed analytic expectations for genetic variation of populations in continuous space, but these models fail to capture the spatial dynamics of real populations (Felsenstein XXXX). Specifically, the externally fixed population sizes used in Wright's and Malecót's models creates a form of global population regulation in which the fitness of each individual depends on the fitness of every other individual in the population despite a lack of interaction among distant individuals. Over time this causes populations to become increasingly clumped until all individuals occur in a single contiguous area occupying a fraction of the available range -- what Felsenstein 197X described as "the pain in the torus".

In reality global population size is an emergent property of individual fitness, which in turn is regulated by competition among interacting individuals at the local scale.

\section{Introduction}

\section{Methods}
\subsection{ A Forward-Time Model of Evolution in Continuous Space }

\subsubsection{ Mating and Dispersal }

\subsubsection{ Competition }

\subsubsection{ Boundary Conditions (Refrain from the Torus?) }

\subsubsection{ Tree Sequence Recording }

\subsection{ Summary Statistics }

\subsection{ Demographic Modeling }

\subsection{ Association Tests }

\subsection{ Machine Learning }

\section{Results}

\subsection{ Summary Statistics }

\subsection{ Impacts on Demographic Inference and GWAS }

\subsection{ Estimating Model Parameters with Machine Learning }

\section{ Discussion }

\section{Figures}

\begin{figure}[h]
	\centerline{
	\includegraphics{../figures/10kouts_coalescence_by_generation}
	}
\end{figure}

\begin{figure}[h]
	\centerline{
	\includegraphics{../figures/sumstat_by_neighbors_spat_v_rm.pdf}
	}
\end{figure}

\begin{figure}[h]
	\centerline{
	\includegraphics{../figures/sfs_spatial_v_rm.pdf}
	}
\end{figure}

\end{document}

### 2018-11-20 scripts/ibs_viz.R

@@ -1,0 +1,63 @@
+setwd("~/spaceness/")
+library(plyr);library(ggplot2);library(magrittr);library(data.table);library(wesanderson)
+
+theme_set(theme_classic()+theme(axis.text=element_text(size=7),
+                                axis.title=element_text(size=7),
+                                strip.text = element_text(size=7),
+                                title = element_text(size=8,hjust=0.5),
+                                legend.title = element_text(size=7),
+                                legend.text = element_text(size=7),
+                                strip.background = element_blank()))
+
+ss <- fread("sumstats/ss_spatial_W16.txt",data.table=F)
+rm <- fread("sumstats/ss_randmates_W16.txt",data.table=F)
+ss$model <- "spatial"
+rm$model <- "random mating"
+ss <- rbind(ss,rm)
+sfs <- ss[,12:112]
+ibsbins <- ss[,113:(ncol(ss)-1)]
+ss <- ss[,c(1:11,ncol(ss))]
+colnames(ss) <- c("sigma","segsites","pi","thetaW","tajD","het_o","fis",
+                  "gen_sp_corr","ibs_95p","ibs_n_blocks_per_pair","ibs_over_1e6","model")
+#ss <- subset(ss,sigma != 0)
+popsize <- fread("sumstats/popsize_spatial_W16.txt")
+rmpopsize <- fread("sumstats/popsize_randmates_W16.txt")
+popsize <- rbind(popsize,rmpopsize)
+names(popsize) <- c("sigma","census_n")
+ss <- merge(ss,popsize,by="sigma")
+
+ibsbins$sigma <- ss$sigma
+tmp <- melt(ibsbins,id.vars=c("sigma"))
+tmp$variable <- unlist(lapply(tmp$variable,function(e) as.numeric(gsub("V","",e))-111))
+tmp <- subset(tmp,sigma != 0)
+
+######### IBS block size plots #############
+ggplot(data=subset(tmp,sigma==min(tmp$sigma)| sigma == max(tmp$sigma)),
+       aes(x=variable,y=log(value)/1e8,col=factor(sigma)))+
+  xlim(0,1000)+
+  geom_line()
+
+ggplot(data=tmp,aes(x=variable,y=log(value),col=sigma,z=sigma))+
+  scale_fill_viridis_c()+
+  xlim(0,10000)+
+  xlab("IBS Block Size")+ylab("log(count)")+
[…]

### 2018-11-20 scripts/pairwise_ibs.py

@@ -1,0 +1,27 @@
+import numpy as np
+def getPairwiseIbsTractLengths(x,y,positions,maxlen):
+    '''
+    input:
+    x: haplotype as 1D array
+    y: haplotype as 1D array
+    positions: a 1D array of SNP positions (integer, 1-based)
+
+    Returns:
+    1d array listing distances between adjacent SNPs in a pair of sequences.
+
+    Example:
+    haps,positions,labels,locs=getHapsPosLabelsLocs(direc="mutated_treeSeq_directory")
+    ibs_lengths_1_2=getPairwiseIbsTractLengths(haps[0][:,0],haps[0][:,1],positions[0])
+    '''
+    snps=np.equal(x,y)
+    snp_positions=positions[snps]
+    l=len(snp_positions)
+    ibs_tracts=[]
+    if(l==0):
+        ibs_tracts=[maxlen]
+    else:
+        if(l>1):
+            ibs_tracts=snp_positions[np.arange(1,l,2)]-snp_positions[np.arange(0,l-1,2)] #middle blocks
+        np.append(ibs_tracts,snp_positions[0])          #first block
+        np.append(ibs_tracts,maxlen-snp_positions[l-1]) #last block
+    return ibs_tracts

### 2018-11-20 scripts/plot_sumstats.R

@@ -23,8 +23,10 @@ sfs <- pd[,8:(ncol(pd)-1)]
 sfs$dispersal <- pd$dispersal
 sfs$segsites <- pd$segsites
 sfs$type <- pd$type
-pd <- melt(pd[,c("segsites","pi","tajD","het_o","Fis","dispersal","type","Census N")],id.vars=c("dispersal","type"))
-pd$variable[pd$variable=="pi"] <- expression(pi)
+pd$mean_neighbors <- ((pi*pd$dispersal^2)/16^2)*1000
+pd$max_neighbors <- (3*((pi*pd$dispersal^2)/16^2))*1000
+pd <- melt(pd[,c("segsites","pi","tajD","het_o","Fis","dispersal","type","Census N","mean_neighbors","max_neighbors")],
+           id.vars=c("dispersal","type","mean_neighbors","max_neighbors"))
 
 #summary stat plot
 pd$variable <- factor(pd$variable,levels=c("Census N","segsites","pi","tajD","het_o","Fis"))
@@ -41,7 +43,7 @@ ss_by_dispersal <- ggplot(data=pd,aes(x=dispersal/16*100,y=value,fill=type,col=t
   #scale_color_brewer(palette = "Dark2",name="Model")+
   scale_fill_manual(values=discrete_palette,name="Model")+
   scale_color_manual(values=discrete_palette,guide=F)+
-  xlab("Dispersal (% Total Range)")+
+  xlab("Interaction Radius (% range width)")+
   facet_wrap(~variable,scales="free")+
   geom_point(size=1.4,shape=21,stroke=0.02,col="black",alpha=0.7)+
   geom_smooth(lwd=0.5,fill=NA)
@@ -50,6 +52,30 @@ pdf("~/spaceness/figures/sumstat_by_dispersal_spat_v_rm.pdf",width=7.5,height=3.
 ss_by_dispersal <- ss_by_dispersal + guides(fill = guide_legend(override.aes = list(size=4)))
 print(ss_by_dispersal)
 dev.off()
+
+ss_by_neighbors <- ggplot(data=pd,aes(x=mean_neighbors,y=value,fill=type,col=type))+
+  theme_classic()+
+  theme(axis.title.y=element_blank(),
+        strip.background = element_blank(),
+        legend.box.background = element_blank(),
+        legend.title = element_text(size=9),
[…]

### 2018-11-20 scripts/process_slimout.py

@@ -1,0 +1,36 @@
+from slimtools import *
+import argparse
+#os.chdir("/Users/cj/spaceness")
+
+parser = argparse.ArgumentParser(
+            description='Calculate summary statistics from a set of \
+                         tree sequences produced by msprime or SLiM')
+parser.add_argument('--indir', dest='indir',
+                    help='full path to directory with tree sequences')
+parser.add_argument('--outpath', dest='outdir',
+                   help='full path for output file.')
+parser.add_argument('--subset', dest='subset',type=bool,
+                   help='Boolean. If False, process all files in directory.\
+                         If True, start and stop give the 0-based \
+                         index bounds of files to process as returned \
+                         by os.listdir(indir)')
+parser.add_argument('--start', dest='start',type=int,
+                   help='First file to process.')
+parser.add_argument('--stop', dest='stop',type=int,
+                   help='Last file to process.')
+args=parser.parse_args()
+
+print("loading haplotype matrices")
+haps,pos,labels,locs=get_ms_outs("sims/mutated/spatial/W16/",
+                                 subset=args.subset,
+                                 start=args.start,
+                                 stop=args.stop)
+pos=discretize_snp_positions(pos)
+
+print("calculating summary statistics")
+ss=getHaplotypeSumStats(haps,
+                        pos,
+                        labels,
+                        locs,
+                        args.outdir,
+                        verbose=False)

### 2018-11-20 scripts/run_slim.py

@@ -28,12 +28,11 @@ def run_one_slim_sim(sampled_param,
     distribution. Output will be named with the name and value of
     the sampled parameter.
     '''
-
     #get sampled param names and values
     val=numpy.random.uniform(min,max)
 
-    #get output file path
     filename = sampled_param+"_"+str(val)+"_.trees"
+
     filepath = os.path.join(outdir,filename)
 
     #set strings for defining SLiM variables
@@ -47,7 +46,7 @@ def run_one_slim_sim(sampled_param,
              slim_recipe]
 
     #run it
-    print("starting slim simulation with "+args.sampled_param+"="+str(val))
+    print("writing tree sequence output to "+filepath)
     subprocess.check_output(command)
 
     return None

### 2018-11-20 scripts/sample_and_mutate_treeseqs.py

@@ -1,0 +1,27 @@
+from slimtools import *
+
+#check treeseq directory for coalescence
+treeseqs,out=check_treeseq_coalescence("/Users/cj/spaceness/sims/slimout/spatial/W16")
+np.mean(out[0]) #mean proportion uncoalesced gene trees
+np.mean(out[1]) #median roots per gene tree
+
+sample_treeseq_directory(indir="/Users/cj/spaceness/sims/slimout/random_mating/W16",
+                         outdir="/Users/cj/spaceness/sims/sampled/random_mating/W16",
+                         nSamples=50,
+                         recapitate=False,
+                         recombination_rate=1e-9)
+
+mutate_treeseqs("/Users/cj/spaceness/sims/sampled/random_mating/W16/",
+                "/Users/cj/spaceness/sims/mutated/random_mating/W16/",
+                1e-8)
+
+haps,positions,labels,locs=get_ms_outs("/Users/cj/spaceness/sims/mutated/random_mating/W16",False)
+
+positions=discretize_snp_positions(positions)
+
+popsizes=get_pop_sizes(indir="/Users/cj/spaceness/sims/slimout/random_mating/W16/",
+              outpath="/Users/cj/spaceness/sumstats/popsize_randmates_W16.txt")
+
+ss=getHaplotypeSumStats(haps,positions,labels,locs,
+                        outfile="/Users/cj/spaceness/sumstats/ss_spatial_W16_sample2.txt",
+                        verbose=False)

### 2018-11-20 scripts/slimtools.py

@@ -5,7 +5,11 @@ import shlex, scipy, time, re
 from shutil import copyfile
 import allel
 import time
+from scipy import spatial
+from scipy import stats
 from tqdm import tqdm
+import itertools
+import matplotlib.pyplot as plt
 
 def run_one_slim_sim(sampled_param,
                      min,
@@ -69,17 +73,28 @@ def check_treeseq_coalescence(indir):
         j=j+1
     return treeseqs,out
 
-def sample_treeseq_directory(indir,outdir,nSamples,recapitate,recombination_rate,output_pop_sizes,pop_size_outpath):
+def get_pop_sizes(indir,outpath):
+    trees = trees=[f for f in os.listdir(indir) if not f.startswith(".")]
+    labels=[float(re.sub("sigma_|_.trees*|.trees","",x)) for x in trees]
+    out=np.empty(shape=(len(trees),2))
+    for i in tqdm(range(len(trees))):
+        ts=pyslim.load(os.path.join(indir,trees[i]))
+        popsize=ts.num_samples/2
+        out[i]=[labels[i],popsize]
+    np.savetxt(outpath,out)
+    return out
+
+def sample_treeseq_directory(indir,outdir,nSamples,recapitate,recombination_rate):
     '''
     Sample individuals from a tree sequence then simplify and (optionally)
     recapitate the tree with msprime.
     '''
     trees=[f for f in os.listdir(indir) if not f.startswith(".")]
-    pop_sizes=[]
+    #pop_sizes=[]
     for i in tqdm(range(len(trees))):
         ts=pyslim.load(os.path.join(indir,trees[i]))
-        if(output_pop_sizes):
-            pop_sizes.append(ts.num_samples/2)
+        #if(output_pop_sizes):
+            #pop_sizes.append(ts.num_samples/2)
         sample_inds=np.unique([ts.node(j).individual for j in ts.samples()]) #final-generation individuals
         subsample=np.random.choice(sample_inds,nSamples,replace=False) #get nSamples random sample inds
         subsample_nodes=[ts.individual(x).nodes for x in subsample] #node numbers for sampled individuals
@@ -89,8 +104,8 @@ def sample_treeseq_directory(indir,outdir,nSamples,recapitate,recombination_rate
         if(recapitate):
             ts=ts.recapitate(recombination_rate=recombination_rate)
         o.dump(os.path.join(outdir,trees[i]))
-    if(output_pop_sizes):
[…]

### 2018-11-20 scripts/treeseq_to_vcf.py

@@ -1,0 +1,5 @@
+#smcpp runs for simulated nwf model output
+import pyslim, msprime, numpy as np, os
+os.chdir("/Users/cj/spaceness/demography/smcpp")
+
+ts=pyslim.load("data/sigma_0.3095957995572229_.trees1500000.trees")

### 2018-11-20 ms/spaceness.tex
updated title
misc

@@
 
-\title{ Evolution on the Final Frontier: How Dispersal and Competition Shape Genetic Variation in Continuous Space }
+\title{ Space is the Place: How Dispersal and Competition Shape Genetic Variation in Continuous Space }
 \author{C. J. Battey, Peter Ralph, Andrew Kern}
@@
 \section{Abstract}
+
+\section{Introduction}
 Random mating is a core assumption of many population genetic models, but likely occurs in few real populations.
@@
 In reality global population size is an emergent property of individual fitness, which in turn is regulated by competition among interacting individuals at the local scale.
-
-\section{Introduction}
 

### 2018-12-06 batch_scripts/1_run_slim_simulations.sh
added gwas, start numbered scripts, parallelized sumstats

@@ -1,6 +1,7 @@
 #!/bin/bash
-for i in {1..200} 
+for i in {1..100} 
 do
+	sbatch /projects/kernlab/cbattey2/spaceness/batch_scripts/runslim_spatial.srun
 	sbatch /projects/kernlab/cbattey2/spaceness/batch_scripts/runslim_random_mating.srun
 done
 

### 2018-12-06 batch_scripts/2_get_sumstats.sh

@@ -1,0 +1,50 @@
+#!/bin/bash
+# get summary statistics for all tree sequences in a directory
+cd ~/projects/spaceness
+
+#spatial mate choice sims
+files=~/projects/spaceness/sims/slimout/spatial/W35/*
+for f in $files
+do
+ 
+ command="python scripts/sumstats_from_treeseq.py \
+ --infile $f \
+ --outfile ~/projects/spaceness/sumstats/ss_spatial_W35.txt \
+ --sampling random \
+ --mu 0.25e-8 \
+ --sampling_location 25,25 \
+ --seed 12345"
+
+cp ~/projects/spaceness/batch_scripts/slurm_sumstats_header.srun ~/projects/spaceness/batch_scripts/tmp.srun
+
+echo $command >> ~/projects/spaceness/batch_scripts/tmp.srun
+
+sbatch ~/projects/spaceness/batch_scripts/tmp.srun
+	
+sleep 0.05
+
+done
+
+#random mating sims
+files=~/projects/spaceness/sims/slimout/random_mating/W35/*
+for f in $files
+do
+ 
+ command="python scripts/sumstats_from_treeseq.py \
+ --infile $f \
+ --outfile ~/projects/spaceness/sumstats/ss_randmates_W35.txt \
+ --sampling random \
+ --mu 0.25e-8 \
+ --sampling_location 25,25 \
+ --seed 12345"
+
+cp ~/projects/spaceness/batch_scripts/slurm_sumstats_header.srun ~/projects/spaceness/batch_scripts/tmp.srun
+
+echo $command >> ~/projects/spaceness/batch_scripts/tmp.srun
+
+sbatch ~/projects/spaceness/batch_scripts/tmp.srun
+	
+sleep 0.05
+
+done
+

### 2018-12-06 batch_scripts/3_run_gwas.sh

@@ -1,0 +1,31 @@
+#!/bin/bash
+# get summary statistics for all tree sequences in a directory
+cd ~/projects/spaceness
+
+#spatial mate choice sims
+files=~/projects/spaceness/sims/slimout/spatial/W35/*
+for f in $files
+do
+ 
+ command="python scripts/run_gwas.py \
+ --treeseq $f \
+ --outdir ~/projects/spaceness/gwas/out \
+ --plink_path ~/projects/plink-1.07-x86_64/plink \
+ --vcftools_path ~/projects/vcftools_0.1.13/bin/vcftools \
+ --nSamples 1000 \
+ --sampling random \
+ --mu 0.25e-8 \
+ --phenotype transform_coord \
+ --phenotype_mean 175 \
+ --phenotype_sd 8 \
+ --seed 12345"
+ 
+cp ~/projects/spaceness/batch_scripts/slurm_gwas_header.srun ~/projects/spaceness/batch_scripts/tmp.srun
+
+echo $command >> ~/projects/spaceness/batch_scripts/tmp.srun
+
+sbatch ~/projects/spaceness/batch_scripts/tmp.srun
+	
+sleep 0.01
+
+done

### 2018-12-06 scripts/gwas_viz.R

@@ -1,0 +1,60 @@
+#visualizing GWAS on SLiM simulations
+library(ggplot2);library(magrittr);library(plyr);library(data.table)
+setwd("~/spaceness/gwas/")
+
+#individual manhattan plot
+gwas <- fread("out/sigma_0.3707859307192668_.trees1500000.trees.assoc.linear")
+
+gwas <- subset(gwas,TEST=="ADD")
+nrow(gwas)
+
+gwas$p_adj <- p.adjust(gwas$P,method="fdr")
+gwas$p_minus_log_10 <- -log(gwas$P,10)
+  
+  
+png("gwas_sim_spatial_phen_PCA_corr_sigma0905.png",width=6.5,height=1,units="in",res=1200)
+ggplot(data=gwas,aes(x=BP,y=p_minus_log_10))+
+  theme_classic()+theme(axis.text=element_text(size=7),
+                        axis.title = element_text(size=7),
+                        axis.text.x=element_blank(),
+                        axis.ticks.x=element_blank(),
+                        axis.title.x=element_blank())+
+  geom_point(size=0.2,shape=21,stroke=0.4)+
+  ylab(expression(italic(-log[10](p))))+
+  geom_hline(yintercept=-log(.05/nrow(gwas),10),col="grey",linetype=2)+
+  annotate(geom="text",label=expression(italic(p[corr]==0.05)),
+           x=4e6,y=-log(.05/nrow(gwas),10)-0.07*max(gwas$p_minus_log_10),size=1.5,col="white")
+dev.off()
+
+
+#how does the number of genome-wide significant SNPs change with dispersal? 
+files <- list.files("out",full.names = T) %>% grep("assoc.linear",.,value = T)
+pd <- data.frame(sigma=numeric(),n_sig=integer(),n_snps=integer(),prop_sig=numeric())
+for(i in files){
+  gwas <- fread(i)
+  gwas$p_adj <- p.adjust(gwas$P,method="fdr")
+  n_sig <- nrow(subset(gwas,p_adj<=0.05))
+  n_snps <- nrow(gwas)
+  prop_sig <- n_sig/n_snps
+  sigma <- strsplit(i,"_") %>% unlist() %>% .[2] %>% as.numeric()
[…]

### 2018-12-06 scripts/ibs_viz.R

@@ -9,19 +9,19 @@ theme_set(theme_classic()+theme(axis.text=element_text(size=7),
                                 legend.text = element_text(size=7),
                                 strip.background = element_blank()))
 
-ss <- fread("sumstats/ss_spatial_W16.txt",data.table=F)
-rm <- fread("sumstats/ss_randmates_W16.txt",data.table=F)
+ss <- fread("sumstats/ss_spatial_W35.txt",data.table=F)
+rm <- fread("sumstats/ss_randmates_W35.txt",data.table=F)
 ss$model <- "spatial"
 rm$model <- "random mating"
 ss <- rbind(ss,rm)
-sfs <- ss[,12:112]
-ibsbins <- ss[,113:(ncol(ss)-1)]
-ss <- ss[,c(1:11,ncol(ss))]
+sfs <- ss[,15:(ncol(ss)-1)]
+#ibsbins <- ss[,113:(ncol(ss)-1)]
+ss <- ss[,c(1:14,ncol(ss))]
 colnames(ss) <- c("sigma","segsites","pi","thetaW","tajD","het_o","fis",
-                  "gen_sp_corr","ibs_95p","ibs_n_blocks_per_pair","ibs_over_1e6","model")
+                  "gen_sp_corr","ibs_mean","ibs_var","ibs_skew","ibs_95p","ibs_num_blocks_per_pair","ibs_mean_spat_corr","model")
 #ss <- subset(ss,sigma != 0)
-popsize <- fread("sumstats/popsize_spatial_W16.txt")
-rmpopsize <- fread("sumstats/popsize_randmates_W16.txt")
+popsize <- fread("sumstats/popsize_spatial_W35.txt")
+rmpopsize <- fread("sumstats/popsize_randmates_W35.txt")
 popsize <- rbind(popsize,rmpopsize)
 names(popsize) <- c("sigma","census_n")
 ss <- merge(ss,popsize,by="sigma")
@@ -44,20 +44,39 @@ ggplot(data=tmp,aes(x=variable,y=log(value),col=sigma,z=sigma))+
   stat_summary_hex(fun="mean",bins = 100)
 
 ############ mixed summary stat blocks ###########
-mss <- melt(ss,id.vars=c("sigma","model","census_n"))
-mss <- subset(mss,sigma<=3.5)
[…]

### 2018-12-06 scripts/run_gwas.py

@@ -1,0 +1,150 @@
+from slimtools import *
+import msprime as msp,pyslim,numpy as np,os,subprocess as sp
+import argparse
+import allel
+
+parser = argparse.ArgumentParser(description='Run a GWAS in plink on individuals \
+                                              sampled from a SLiM simulation.\
+                                              Requires plink, vcftools, msprime, and pyslim.')
+parser.add_argument('--treeseq', dest='treeseq',
+                    help='path to simulation tree sequence.')
+parser.add_argument('--outdir', dest='outdir',
+                   help='Output file directory')
+parser.add_argument('--plink_path', dest='plink_path',
+                   help='call for plink')
+parser.add_argument('--vcftools_path', dest='vcftools_path',
+                   help='call for vcftools')
+parser.add_argument('--nSamples', dest='nSamples', type=int,
+                   help='Number of individuals to sample.')
+parser.add_argument('--sampling',dest='sampling',
+                   help='Type of sampling to use. Options: \
+                   random (nSamples random individuals). \
+                   point (nSamples individuals closest to sampling_location). \
+                   pair (nSamples individuals closest to sampling_location \
+                   and its reflection over the x axis)')
+parser.add_argument('--mu', dest='mu',type=float,
+                   help='mutation rate per base per unit time (msprime generations or SLiM time steps)')
+parser.add_argument('--phenotype', dest='phenotype',
+                   help='Type of phenotype to generate. \
+                         options: "gaussian" (phenotypes drawn from a normal distribution); \
+                                  "transform_coord" (phenotype is a transformation of the x coordinate); \
+                                  "corner_bimodal" (phenotype drawn from different normal distributions \
+                                  in/outside a corner of the landscape).')
+parser.add_argument('--phenotype_mean', dest='phenotype_mean',
+                   help='mean for gaussian phenotypes')
+parser.add_argument('--phenotype_sd', dest='phenotype_sd',
+                   help='standard deviation for gaussian phenotypes')
+parser.add_argument('--seed', dest='seed',type=int,
+                   help='Random seed for sampling individuals and adding mutations to tree sequences.')
+
+args=parser.parse_args()
+
[…]

### 2018-12-06 scripts/sample_and_mutate_treeseqs.py

@@ -1,27 +1,28 @@
 from slimtools import *
 
 #check treeseq directory for coalescence
-treeseqs,out=check_treeseq_coalescence("/Users/cj/spaceness/sims/slimout/spatial/W16")
-np.mean(out[0]) #mean proportion uncoalesced gene trees
-np.mean(out[1]) #median roots per gene tree
+treeseqs,out=check_treeseq_coalescence("/Users/cj/spaceness/sims/slimout/wf/")
+np.mean(out[0]) #mean proportion uncoalesced trees
+np.mean(out[1]) #mean roots per tree
 
-sample_treeseq_directory(indir="/Users/cj/spaceness/sims/slimout/random_mating/W16",
-                         outdir="/Users/cj/spaceness/sims/sampled/random_mating/W16",
+sample_treeseq_directory(indir="/Users/cj/spaceness/sims/slimout/wf/",
+                         outdir="/Users/cj/spaceness/sims/sampled/wf/",
                          nSamples=50,
                          recapitate=False,
-                         recombination_rate=1e-9)
+                         recombination_rate=1e-8)
 
-mutate_treeseqs("/Users/cj/spaceness/sims/sampled/random_mating/W16/",
-                "/Users/cj/spaceness/sims/mutated/random_mating/W16/",
-                1e-8)
+mutate_treeseqs("/Users/cj/spaceness/sims/sampled/wf",
+                "/Users/cj/spaceness/sims/mutated/wf",
+                4e-8)
 
-haps,positions,labels,locs=get_ms_outs("/Users/cj/spaceness/sims/mutated/random_mating/W16",False)
+haps,positions,labels,locs=get_ms_outs("/Users/cj/spaceness/sims/mutated/wf",False)
 
 positions=discretize_snp_positions(positions)
 
-popsizes=get_pop_sizes(indir="/Users/cj/spaceness/sims/slimout/random_mating/W16/",
-              outpath="/Users/cj/spaceness/sumstats/popsize_randmates_W16.txt")
+popsizes=get_pop_sizes(indir="/Users/cj/spaceness/sims/slimout/wf/",
+              outpath="/Users/cj/spaceness/sumstats/popsize_wfN1000.txt")
 
 ss=getHaplotypeSumStats(haps,positions,labels,locs,
-                        outfile="/Users/cj/spaceness/sumstats/ss_spatial_W16_sample2.txt",
-                        verbose=False)
[…]

### 2018-12-06 scripts/slimtools.py

@@ -86,8 +86,8 @@ def get_pop_sizes(indir,outpath):
 
 def sample_treeseq_directory(indir,outdir,nSamples,recapitate,recombination_rate):
     '''
-    Sample individuals from a tree sequence then simplify and (optionally)
-    recapitate the tree with msprime.
+    Sample individuals from all tree sequences a directory then simplify and (optionally)
+    recapitate the treeseq_to_vcf.py with msprime.
     '''
     trees=[f for f in os.listdir(indir) if not f.startswith(".")]
     #pop_sizes=[]
@@ -108,19 +108,53 @@ def sample_treeseq_directory(indir,outdir,nSamples,recapitate,recombination_rate
     #    np.savetxt(pop_size_outpath,pop_sizes)
     return None
 
-def mutate_treeseqs(indir,outdir,mu):
+def sample_treeseq(infile,outfile,nSamples,recapitate,recombination_rate,write_to_file,seed=12345):
+    '''
+    Sample individuals from a tree sequence then simplify and (optionally)
+    recapitate the tree with msprime.
+    '''
+    if type(infile)==str:
+        ts=pyslim.load(infile)
+    else:
+        ts=infile
+    if(not seed==0):
+        np.random.seed(seed)
+    sample_inds=np.unique([ts.node(j).individual for j in ts.samples()]) #final-generation individuals
+    subsample=np.random.choice(sample_inds,nSamples,replace=False) #get nSamples random sample inds
+    subsample_nodes=[ts.individual(x).nodes for x in subsample] #node numbers for sampled individuals
+    subsample_nodes=[a for b in subsample_nodes for a in b] #flatten the list
+    subsample_nodes=np.sort(np.array(subsample_nodes))
+    o=ts.simplify(subsample_nodes)
+    if(recapitate):
+        ts=ts.recapitate(recombination_rate=recombination_rate)
+    if(write_to_file):
+        o.dump(outfile)
+
+    return o
+
+def mutate_treeseqs(indir,outdir,mu,seed):
     '''
     Add mutations at constant rate "mu" to all tree sequences in directory "indir"
     '''
     trees=[f for f in os.listdir(indir) if not f.startswith(".")]
     for i in tqdm(trees):
         ts=pyslim.load(os.path.join(indir,i))
-        ts=msp.mutate(ts,mu)
+        ts=msp.mutate(ts,mu,random_seed=seed)
[…]

### 2018-12-06 scripts/sumstats_from_treeseq.py

@@ -1,0 +1,56 @@
+from slimtools import *
+import argparse
+#os.chdir("/Users/cj/spaceness")
+
+parser = argparse.ArgumentParser(
+            description='Calculate summary statistics from a  \
+                         tree sequence produced by msprime or SLiM')
+parser.add_argument('--infile', dest='infile',
+                    help='full path to tree sequence')
+parser.add_argument('--outfile', dest='outfile',
+                   help='full path for output summary statistics.')
+parser.add_argument('--seed', dest='seed',type=int,
+                   help='Random seed for sampling individuals and adding mutations to tree sequences.')
+parser.add_argument('--sampling',dest='sampling',
+                   help='Type of sampling to use. Options: \
+                   random (nSamples random individuals). \
+                   point (nSamples individuals closest to sampling_location). \
+                   pair (nSamples individuals closest to sampling_location \
+                   and its reflection over the x axis)')
+parser.add_argument('--mu', dest='mu',
+                   help='mutation rate per base per unit time in SLiM simulations.')
+parser.add_argument('--sampling_location', dest='sampling_location',
+                   help='Sampling site location given as "x,y"')
+args=parser.parse_args()
+
+#params for debugging
+args=argparse.Namespace(infile='/Users/cj/spaceness/sims/slimout/spatial/W35/sigma_0.2130878654199541_.trees',
+                        outfile='/Users/cj/Desktop/test_sumstats.txt',
+                        sampling='random',
+                        mu=0.25e-8,
+                        seed=12345)
+
+print("sampling individuals and adding mutations")
+ts=sample_treeseq(infile=args.infile,
+                  outfile="",
+                  nSamples=100,
+                  recapitate=False,
+                  recombination_rate=1e-8,
+                  write_to_file=False,
+                  seed=args.seed)
+ts=msp.mutate(ts,args.mu,random_seed=args.seed)
+
+print("loading haplotype matrices")
+haps,pos,locs=get_ms_outs(ts)
+label=float(re.sub("sigma_|_.trees*|.trees","",os.path.basename(args.infile)))
+pos=discretize_snp_positions(pos)
+
+print("calculating summary statistics")
+ss=getSLiMSumStats(haps,
+                   pos,
+                   label,
+                   locs,
+                   args.outfile,
+                   maxlen=1e8,
+                   ibs_tracts=True,
+                   verbose=True)

### 2018-12-06 batch_scripts/run_slim_batch_small_long.sh
put old scripts in old folders

@@ -1,6 +1,0 @@
-#!/bin/bash
-for i in {1..100} 
-do
-	sbatch /projects/kernlab/cbattey2/spaceness/scripts/runslim_small_long.srun
-done
-

### 2018-12-06 batch_scripts/run_slim_batch_spatial.sh

@@ -1,7 +1,0 @@
-#!/bin/bash
-for i in {1..100} 
-do
-	sbatch /projects/kernlab/cbattey2/spaceness/batch_scripts/runslim_spatial.srun
-
-done
-

### 2018-12-06 batch_scripts/run_slim_batch_wf.sh

@@ -1,6 +1,0 @@
-#!/bin/bash
-for i in {1..200} 
-do
-	sbatch /projects/kernlab/cbattey2/spaceness/scripts/runslim_wf.srun
-done
-

### 2018-12-06 scripts/pairwise_ibs.py

@@ -1,27 +1,0 @@
-import numpy as np
-def getPairwiseIbsTractLengths(x,y,positions,maxlen):
-    '''
-    input:
-    x: haplotype as 1D array
-    y: haplotype as 1D array
-    positions: a 1D array of SNP positions (integer, 1-based)
-
-    Returns:
-    1d array listing distances between adjacent SNPs in a pair of sequences.
-
-    Example:
-    haps,positions,labels,locs=getHapsPosLabelsLocs(direc="mutated_treeSeq_directory")
-    ibs_lengths_1_2=getPairwiseIbsTractLengths(haps[0][:,0],haps[0][:,1],positions[0])
-    '''
-    snps=np.equal(x,y)
-    snp_positions=positions[snps]
-    l=len(snp_positions)
-    ibs_tracts=[]
-    if(l==0):
-        ibs_tracts=[maxlen]
-    else:
-        if(l>1):
-            ibs_tracts=snp_positions[np.arange(1,l,2)]-snp_positions[np.arange(0,l-1,2)] #middle blocks
-        np.append(ibs_tracts,snp_positions[0])          #first block
-        np.append(ibs_tracts,maxlen-snp_positions[l-1]) #last block
-    return ibs_tracts

### 2018-12-06 scripts/sample_treeseqs.py

@@ -1,40 +1,0 @@
-def sample_treeseq_directory(indir,
-                             type="random",
-                             outdir,
-                             nSamples,
-                             recapitate,
-                             recombination_rate,
-                             n_localities,
-                             width):
-    '''
-    loop sample_treeseq over a directory.
-    '''
-    trees=os.listdir(indir)
-    for i in range(len(trees)):
-        ts=pyslim.load(os.path.join(indir,trees[i]))
-        if(type=="random"):
-            sample_inds=np.unique([ts.node(j).individual for j in ts.samples()]) #final-generation individuals
-            subsample=np.random.choice(sample_inds,nSamples,replace=False) #get nSamples random sample inds
-            subsample_nodes=[ts.individual(x).nodes for x in subsample] #node numbers for sampled individuals
-            subsample_nodes=[a for b in subsample_nodes for a in b] #flatten the list
-            subsample_nodes=np.sort(np.array(subsample_nodes))
-            o=ts.simplify(subsample_nodes)
-        else if(type=="grid"):
-            sampling_locs=width/n_localities
-
-        if(recapitate):
-            ts=ts.recapitate(recombination_rate=recombination_rate)
-        o.dump(os.path.join(outdir,trees[i]))
-
-    return None
-
-
-indir="/Users/cj/spaceness/sims/slimout/k5/"
-type="grid"
-nSamples=50
-recapitate=True
-recombination_rate=1e-9
-n_localities=4
-width=35
-
-sample_treeseq_directory()

### 2018-12-06 scripts/treeseq_to_vcf.py

@@ -1,5 +1,0 @@
-#smcpp runs for simulated nwf model output
-import pyslim, msprime, numpy as np, os
-os.chdir("/Users/cj/spaceness/demography/smcpp")
-
-ts=pyslim.load("data/sigma_0.3095957995572229_.trees1500000.trees")

### 2018-12-06 scripts/gwas_viz.R
fixed argparse bugs and -assoc call

@@ -3,7 +3,7 @@ library(ggplot2);library(magrittr);library(plyr);library(data.table)
 setwd("~/spaceness/gwas/")
 
 #individual manhattan plot
-gwas <- fread("out/sigma_0.3707859307192668_.trees1500000.trees.assoc.linear")
+gwas <- fread("out_phen_corner/sigma_0.20156213958750416_.trees1500000.trees.assoc.linear")
 
 gwas <- subset(gwas,TEST=="ADD")
 nrow(gwas)
@@ -12,7 +12,7 @@ gwas$p_adj <- p.adjust(gwas$P,method="fdr")
 gwas$p_minus_log_10 <- -log(gwas$P,10)
   
   
-png("gwas_sim_spatial_phen_PCA_corr_sigma0905.png",width=6.5,height=1,units="in",res=1200)
+png("gwas_sim_spatial_phen_corner_PCA_corr_sigma020.png",width=6.5,height=1,units="in",res=1200)
 ggplot(data=gwas,aes(x=BP,y=p_minus_log_10))+
   theme_classic()+theme(axis.text=element_text(size=7),
                         axis.title = element_text(size=7),
@@ -28,7 +28,7 @@ dev.off()
 
 
 #how does the number of genome-wide significant SNPs change with dispersal? 
-files <- list.files("out",full.names = T) %>% grep("assoc.linear",.,value = T)
+files <- list.files("out_phen_corner",full.names = T) %>% grep("assoc.linear",.,value = T)
 pd <- data.frame(sigma=numeric(),n_sig=integer(),n_snps=integer(),prop_sig=numeric())
 for(i in files){
   gwas <- fread(i)
@@ -41,7 +41,7 @@ for(i in files){
 }
 pd$pc_corr <- T
 
-files <- list.files("out",full.names = T) %>% grep("qassoc",.,value = T)
+files <- list.files("out_phen_corner",full.names = T) %>% grep("qassoc",.,value = T)
 for(i in files){
   gwas <- fread(i)
   gwas$p_adj <- p.adjust(gwas$P,method="fdr")
@@ -52,9 +52,9 @@ for(i in files){
   pd <- rbind(pd,data.frame(sigma=sigma,n_sig=n_sig,n_snps=n_snps,prop_sig=prop_sig,pc_corr=F))
 }
 
-ggplot(data=pd,aes(x=sigma,y=n_sig,col=pc_corr))+
[…]

### 2018-12-06 scripts/run_gwas.py

@@ -6,7 +6,7 @@ import allel
 parser = argparse.ArgumentParser(description='Run a GWAS in plink on individuals \
                                               sampled from a SLiM simulation.\
                                               Requires plink, vcftools, msprime, and pyslim.')
-parser.add_argument('--treeseq', dest='treeseq',
+parser.add_argument('--infile', dest='infile',
                     help='path to simulation tree sequence.')
 parser.add_argument('--outdir', dest='outdir',
                    help='Output file directory')
@@ -40,20 +40,22 @@ parser.add_argument('--seed', dest='seed',type=int,
 args=parser.parse_args()
 
 ### debug
-# args = argparse.Namespace(treeseq='/Users/cj/spaceness/sims/slimout/spatial/W50/coalesced/sigma_0.20156213958750416_.trees1500000.trees',
-#                           outdir="/Users/cj/spaceness/gwas/out/normal_phen/",
+# args = argparse.Namespace(infile='/Users/cj/spaceness/sims/slimout/spatial/W50/coalesced/sigma_0.20156213958750416_.trees1500000.trees',
+#                           outdir="/Users/cj/Desktop/",
 #                           plink_path="plink",
 #                           vcftools_path="vcftools",
 #                           nSamples=1000,
 #                           mu=0.25e-8,
 #                           phenotype="transform_coord",
 #                           phenotype_mean=175,
-#                           phenotype_sd=8)
+#                           phenotype_sd=8,
+#                           seed=123)
 
 #sample individuals and add mutations
+simname=os.path.basename(args.infile)
 ts=sample_treeseq(infile=args.infile,
                   outfile="",
-                  nSamples=100,
+                  nSamples=args.nSamples,
                   recapitate=False,
                   recombination_rate=1e-8,
                   write_to_file=False,
@@ -116,31 +118,30 @@ if args.phenotype=="corner_bimodal":
 
 
 #run plink association analysis with PC coords as covariates
-if args.pca==True:
-    sp.check_output([args.plink_path,
-                     "--noweb",
-                     "--file",
-                     os.path.join(args.outdir,simname),
-                     "--pheno",
-                     os.path.join(args.outdir,simname)+".phenotypes",
-                     "--allow-no-sex",
-                     "--out",
-                     os.path.join(args.outdir,simname),
-                     "--linear",
-                     "--covar",
-                     os.path.join(args.outdir,simname)+".pca",
-                     "--hide-covar"])
-else:
-    sp.check_output([args.plink_path,
[…]

### 2018-12-06 scripts/run_gwas.py
added phenotype mean args

@@ -100,19 +100,21 @@ if args.phenotype=="gaussian":
     for i in range(args.nSamples):
         phenfile.write("msp_"+str(i)+" "+"msp_"+str(i)+" "+str(phenotypes[i])+"\n")
     phenfile.close()
+
 if args.phenotype=="transform_coord":
     phenfile=open(os.path.join(args.outdir,simname)+".phenotypes","w")
     for i in range(args.nSamples):
-        ind_phenotype=np.random.normal(locs[i][0]/2.5+160,7,1)[0] #if W=50, means will range from 160 - 180 to mimic a cline in cm height
+        ind_phenotype=np.random.normal(locs[i][0]/2.5+args.phenotype_mean,args.phenotype_sd,1)[0]
         phenfile.write("msp_"+str(i)+" "+"msp_"+str(i)+" "+str(ind_phenotype)+"\n")
     phenfile.close()
+
 if args.phenotype=="corner_bimodal":
     phenfile=open(os.path.join(args.outdir,simname)+".phenotypes","w")
     for i in range(args.nSamples):
-        if locs[i][0]<25 and locs[i][1]<25:
-            ind_phenotype=np.random.normal(140,10,1)[0]
+        if locs[i][0]<20 and locs[i][1]<20:
+            ind_phenotype=np.random.normal(args.phenotype_mean,args.phenotype_sd,1)[0]
         else:
-            ind_phenotype=np.random.normal(110,10,1)[0]
+            ind_phenotype=np.random.normal(args.phenotype_mean+3*args.phenotype_sd,args.phenotype_sd,1)[0]
         phenfile.write("msp_"+str(i)+" "+"msp_"+str(i)+" "+str(ind_phenotype)+"\n")
     phenfile.close()
 
@@ -132,6 +134,7 @@ sp.check_output([args.plink_path,
                  os.path.join(args.outdir,simname)+".pca",
                  "--hide-covar"])
 
+#run plink without covariates
 sp.check_output([args.plink_path,
                  "--noweb",
                  "--file",

### 2018-12-06 batch_scripts/3_run_gwas.sh
added patchy phenotypes, full parallel for gwas batch script

@@ -3,21 +3,21 @@
 cd ~/projects/spaceness
 
 #spatial mate choice sims
-files=~/projects/spaceness/sims/slimout/spatial/W35/*
+files=~/projects/spaceness/sims/slimout/spatial/W50/coalesced/*
 for f in $files
 do
  
  command="python scripts/run_gwas.py \
- --treeseq $f \
- --outdir ~/projects/spaceness/gwas/out \
+ --infile $f \
+ --outdir ~/projects/spaceness/gwas/out_normal \
  --plink_path ~/projects/plink-1.07-x86_64/plink \
  --vcftools_path ~/projects/vcftools_0.1.13/bin/vcftools \
  --nSamples 1000 \
  --sampling random \
  --mu 0.25e-8 \
- --phenotype transform_coord \
- --phenotype_mean 175 \
- --phenotype_sd 8 \
+ --phenotype gaussian \
+ --phenotype_mean 100 \
+ --phenotype_sd 10 \
  --seed 12345"
  
 cp ~/projects/spaceness/batch_scripts/slurm_gwas_header.srun ~/projects/spaceness/batch_scripts/tmp.srun
@@ -25,7 +25,66 @@ cp ~/projects/spaceness/batch_scripts/slurm_gwas_header.srun ~/projects/spacenes
 echo $command >> ~/projects/spaceness/batch_scripts/tmp.srun
 
 sbatch ~/projects/spaceness/batch_scripts/tmp.srun
-	
-sleep 0.01
+#sleep 0.1 
+
+command="python scripts/run_gwas.py \
+ --infile $f \
+ --outdir ~/projects/spaceness/gwas/out_clinal \
+ --plink_path ~/projects/plink-1.07-x86_64/plink \
+ --vcftools_path ~/projects/vcftools_0.1.13/bin/vcftools \
+ --nSamples 1000 \
+ --sampling random \
+ --mu 0.25e-8 \
+ --phenotype gaussian \
+ --phenotype_mean 100 \
+ --phenotype_sd 10 \
+ --seed 12345"
+ 
+cp ~/projects/spaceness/batch_scripts/slurm_gwas_header.srun ~/projects/spaceness/batch_scripts/tmp.srun
+
+echo $command >> ~/projects/spaceness/batch_scripts/tmp.srun
+
+sbatch ~/projects/spaceness/batch_scripts/tmp.srun
+#sleep 0.1
+
[…]

### 2018-12-06 scripts/gwas_viz.R

@@ -1,60 +1,75 @@
 #visualizing GWAS on SLiM simulations
-library(ggplot2);library(magrittr);library(plyr);library(data.table)
+library(ggplot2);library(magrittr);library(plyr);library(data.table);library(fdrtool)
 setwd("~/spaceness/gwas/")
 
 #individual manhattan plot
-gwas <- fread("out_phen_corner/sigma_0.20156213958750416_.trees1500000.trees.assoc.linear")
-
+gwas <- fread("~/spaceness/gwas/out_phen_corner/sigma_0.9050194305800039_.trees1500000.trees.assoc.linear")
 gwas <- subset(gwas,TEST=="ADD")
-nrow(gwas)
-
-gwas$p_adj <- p.adjust(gwas$P,method="fdr")
+#gwas$p_adj <- p.adjust(gwas$P,method="fdr")
+gwas$p_adj <- qvalue(gwas$P)$qvalues
 gwas$p_minus_log_10 <- -log(gwas$P,10)
-  
-  
-png("gwas_sim_spatial_phen_corner_PCA_corr_sigma020.png",width=6.5,height=1,units="in",res=1200)
+cutoff <- (max(subset(gwas,p_adj<=0.05)$P)+min(subset(gwas,p_adj>0.05)$P))/2
+if(is.infinite(cutoff)) cutoff <- 0.05/nrow(gwas) #... can you calculate an FDR qval if it's outside the range of the data? 
+
+png("gwas_sim_spatial_phen_corner_PCA_corr.png",width=6.5,height=1,units="in",res=1200)
 ggplot(data=gwas,aes(x=BP,y=p_minus_log_10))+
   theme_classic()+theme(axis.text=element_text(size=7),
                         axis.title = element_text(size=7),
                         axis.text.x=element_blank(),
                         axis.ticks.x=element_blank(),
                         axis.title.x=element_blank())+
-  geom_point(size=0.2,shape=21,stroke=0.4)+
-  ylab(expression(italic(-log[10](p))))+
-  geom_hline(yintercept=-log(.05/nrow(gwas),10),col="grey",linetype=2)+
-  annotate(geom="text",label=expression(italic(p[corr]==0.05)),
[…]

### 2018-12-06 scripts/run_gwas.py

@@ -2,6 +2,7 @@ from slimtools import *
 import msprime as msp,pyslim,numpy as np,os,subprocess as sp
 import argparse
 import allel
+import scipy
 
 parser = argparse.ArgumentParser(description='Run a GWAS in plink on individuals \
                                               sampled from a SLiM simulation.\
@@ -30,9 +31,9 @@ parser.add_argument('--phenotype', dest='phenotype',
                                   "transform_coord" (phenotype is a transformation of the x coordinate); \
                                   "corner_bimodal" (phenotype drawn from different normal distributions \
                                   in/outside a corner of the landscape).')
-parser.add_argument('--phenotype_mean', dest='phenotype_mean',
+parser.add_argument('--phenotype_mean', dest='phenotype_mean',type=float,
                    help='mean for gaussian phenotypes')
-parser.add_argument('--phenotype_sd', dest='phenotype_sd',
+parser.add_argument('--phenotype_sd', dest='phenotype_sd',type=float,
                    help='standard deviation for gaussian phenotypes')
 parser.add_argument('--seed', dest='seed',type=int,
                    help='Random seed for sampling individuals and adding mutations to tree sequences.')
@@ -52,6 +53,7 @@ args=parser.parse_args()
 #                           seed=123)
 
 #sample individuals and add mutations
+np.random.seed(args.seed)
 simname=os.path.basename(args.infile)
 ts=sample_treeseq(infile=args.infile,
                   outfile="",
@@ -66,6 +68,7 @@ ts=msp.mutate(ts,args.mu,random_seed=args.seed)
 haps=ts.genotype_matrix()
 sample_inds=np.unique([ts.node(j).individual for j in ts.samples()])
 locs=[[ts.individual(x).location[0],ts.individual(x).location[1]] for x in sample_inds]
+np.savetxt(os.path.join(args.outdir,simname)+"_locs.txt",locs)
 
 #run a PCA
 genotype_counts=allel.HaplotypeArray(haps).to_genotypes(ploidy=2).to_allele_counts()
@@ -100,7 +103,7 @@ if args.phenotype=="gaussian":
     for i in range(args.nSamples):
         phenfile.write("msp_"+str(i)+" "+"msp_"+str(i)+" "+str(phenotypes[i])+"\n")
     phenfile.close()
-
[…]

### 2018-12-07 ms/spaceness.tex
misc

@@
 
-\section{Introduction}
-Random mating is a core assumption of many population genetic models, but likely occurs in few real populations.
-Instead most organisms mate with nearby individuals and disperse across only a small part of the full range of a species.
-This process creates the pattern known as isolation by distance (IBD; Wright 1942) -- a positive correlation between genetic and geographic distances among individuals.
-Though this pattern is both expected from first principles and widely observed in natural populations (Aguillon et al.
-2017, Pinto et al.
-2013, Battey et al.
-2018), existing population genetic models either exclude it entirely (by assuming random mating) or take the heuristic approach of binning the landscape into discrete blocks of randomly mating populations that exchange migrants each generation.
-
-Efforts to model genomes evolving in continuous space have a long history but have been limited by computational inefficiency and a lack of analytic tractability.
-Both Wright (1942) and Malecót (196X) developed analytic expectations for genetic variation of populations in continuous space, but these models fail to capture the spatial dynamics of real populations (Felsenstein XXXX).
-Specifically, the externally fixed population sizes used in Wright's and Malecót's models creates a form of global population regulation in which the fitness of each individual depends on the fitness of every other individual in the population despite a lack of interaction among distant individuals.
-Over time this causes populations to become increasingly clumped until all individuals occur in a single contiguous area occupying a fraction of the available range -- what Felsenstein 197X described as "the pain in the torus".
-
-In reality global population size is an emergent property of individual fitness, which in turn is regulated by competition among interacting individuals at the local scale.
+\section{Introduction} %Edit away!
 
@@
 
-\subsubsection{ Boundary Conditions (Refrain from the Torus?) }
+\subsubsection{ Boundary Conditions }
 
@@
 \subsection{ Association Tests }
-
-\subsection{ Machine Learning }
 

### 2019-03-06 ms/spaceness.tex
ms updates from overlead

@@
-\documentclass[12pt]{amsart}
-\usepackage{geometry} % see geometry.pdf on how to lay out the page. There's lots.
-\geometry{a4paper} % or letter or a5paper or ... etc
-\usepackage{graphicx}
-% \geometry{landscape} % rotated page geometry
+%\documentclass{article}
 
-% See the ``Article customise'' template for come common customisations
+\documentclass[9pt,twocolumn,twoside]{gsajnl}
+\usepackage{amsmath}
+% Use the documentclass option 'lineno' to view line numbers
 
-\title{ Space is the Place: How Dispersal and Competition Shape Genetic Variation in Continuous Space }
-\author{C. J. Battey, Peter Ralph, Andrew Kern}
-\date{November 2018} % delete this line to display the current date
+\articletype{inv} % article type
+% {inv} Investigation 
+% {gs} Genomic Selection
+% {goi} Genetics of Immunity 
+% {gos} Genetics of Sex 
+% {mp} Multiparental Populations
 
-%%% BEGIN DOCUMENT
+%Simulating Isolation by Distance in Continuous Space
+%Space is the Place: How Dispersal and Competition Shape Genetic Variation in Continuous Space
+%Impacts of Continuous Spatial Structure on Analyses of Population Genetic Data
+\title{Space is the Place: How Dispersal Shapes Genetic Variation in Continuous Space}
+
+\author[$\ast$,1]{C.J. Battey}
+\author[$\dagger$,1]{Peter Ralph}
+\author[$\ddagger$,2]{Andrew Kern}
+
+\affil[$\ast$]{University of Oregon Dept. Biology, Institute for Ecology Evolution}
+\keywords{Space; Population Structure; Demography; Haplotype block sharing}
+
+\runningtitle{SPACENESS} % For use in the footer
+
+%% For the footnote.
+%% Give the last name of the first author if only one author;
+% \runningauthor{FirstAuthorLastname}
+%% last names of both authors if there are two authors;
+% \runningauthor{FirstAuthorLastname and SecondAuthorLastname}
+%% last name of the first author followed by et al, if more than two authors.
+\runningauthor{Battey \textit{et al.}}
+
+\begin{abstract}
+Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations exchanging migrants.
+As the availability of population-level, whole-genome data allows inference of increasingly fine-scale patterns of ancestry in many species, models incorporating realistic demographic and spatial processes are needed to accurately describe spatial structure and control for its impacts on analyses of selection and demography.
+Here we implement a forward-time continuous-space model of molecular evolution and use it to study the impacts of limited dispersal in continuous landscapes on population genetic summary statistics, demographic inference, and association studies.
+Low dispersal slows the geographic spread of ancestry and inflates branch lengths for mid-frequency alleles because of slow coalescence among distant individuals.
+This inflation can bias demographic inference from the site frequency spectrum towards inferring a recent spike in population sizes.
+In a GWAS setting spatially correlated phenotypes cause spurious signals of genetic association when principal components analyses fail to capture the relevant spatial scale of phenotypic variation, which can occur under either very limited or very high dispersal.
+Last we use a machine learning regression approach to evaluate summary statistics as estimators of dispersal distance and find that correlation coefficients between spatial distance and moments of the distribution of pairwise haplotype block lengths retain strong signal even at high dispersal distances.
+\end{abstract}
+
 \begin{document}
@@
 \maketitle
-\tableofcontents
+\thispagestyle{firststyle}
+\marginmark
+\firstpagefootnote
 
-\section{Abstract}
+\correspondingauthoraffiliation{3}{University of Oregon Dept. Biology, Institute for Ecology and Evolution.}
+\vspace{-35pt}% Only used for adjusting extra space in the left column of the first page
 
-\section{Introduction} %Edit away!
+\section{Introduction}
+The inescapable reality that biological organisms live, move, and reproduce in continuous spatial landscapes has been all but ignored in population genetic models.
 
-\section{Methods}
-\subsection{ A Forward-Time Model of Evolution in Continuous Space }
+In most sexual organisms individuals mate with other nearby individuals, leading to a positive correlation between genetic and geographic distances.
+This pattern of "isolation by distance" \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, but is only heuristically approximated by existing forward-time models of molecular evolution based on discrete demes exchanging migrants.
+As the availability of population-level whole-genome data allows inference of increasingly fine-scale patterns of ancestry in many species, models incorporating biologically realistic demographic and spatial processes are needed to accurately describe continuous spatial structure and control for its impacts on analyses of selection and demography.
+This situation is particularly clear in human population genetics, where individual ancestries are often interpreted as mixtures of discrete populations in part because of modeling assumptions underlying discrete clustering algorithms and association analyses are often strongly dependent on statistical corrections for population stratification \citep{Berg2018}.
 
-\subsubsection{ Mating and Dispersal }
+The best-studied approaches to population genetics in continuous space were developed by \citep{Wright1943} and \citep{Malecot1948}, who derived expressions for pairwise genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals.
+However, \citep{Felsenstein1975} showed that these assumptions are incompatible -- over time, a population meeting them will clump into a small number of geographic clusters occupying only a part of the available range.
+(NOTE: where to put Rousset et al.
+lit?)
 
-\subsubsection{ Competition }
+NOTE: probably cut this paragraph???
 
-\subsubsection{ Boundary Conditions }
+This process is analogous to genetic drift, with spatial locations and population densities taking the place of sites and allele frequencies.
+Because individuals disperse from their natal location, new offspring are likely to be close to their parents.
+If by chance an individual has no offspring the local population density is then lowered, which decreases the chance of new individuals dispersing into that area.
+This process compounds over time until only one or two patches of habitat are occupied at any given time.
+The resulting spatial clustering clearly does not describe spatial dynamics in most real populations, because declining growth rates in regions of high population density are common in both lab ( ) and wild populations ( ).
 
-\subsubsection{ Tree Sequence Recording }
+One method for modeling populations in continuous space is then to assume the existence of a grid of discrete randomly-mating populations connected by migration, which prevents clustering by forcing all regions to be occupied in every generation.
+Though good approximations of continuous structure given high dispersal (Barton(?)), these models are not truly continuous and limit investigation of spatial structure below the level of the deme.
+An alternative approach is to model the geographic spread of ancestry backwards in time through a diffusion approximation -- an approach that has recently made significant progress in modeling both dispersal and demographic parameters \citep{Barton2010,Kelleher2014,Ringbauer2017,Ringbauer2018}.
+However, these models currently lack a well-described forward-time analog that produces biologically realistic spatial patterns.
 
-\subsection{ Summary Statistics }
+A direct approach to the clustering problem of classical forward-time models is to incorporate density dependence by scaling individual fitness to local population density.
+This shifts reproductive output towards regions of low-density, preventing populations from clustering in space over time.
+This approach was taken previously by \citep{Doebeli2003} who used an individual based model with continuous space and density dependent fitness to study the probability of speciation along continuous environmental gradients.
+(**Sentence referencing other continuous IBM's with density dependence**).
+However to our knowledge all previous implementations of continuous space models have focused on individuals or at most a few genotypes as the unit of analysis, which limits our ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms.
 
-\subsection{ Demographic Modeling }
+Here we describe an implementation of an individual-based model in continuous space that incorporates overlapping generations, a Gaussian dispersal kernel, and density-dependent fitness, and scales to chromosome-scale alignments across tens of thousands of individuals.
+We use this model to describe the impacts of variation in dispersal distance on a variety of summary statistics typically analyzed under discrete population models, and apply a machine-learning regression approach to evaluate a new class of summary statistics based on correlations between geographic distance and moments of the distribution of pairwise haplotype block lengths as estimators of dispersal distance.
+We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum; finding that inflation of branch lengths for mid-frequency alleles associated with slow coalescence of distant individuals **TBD**.
 
-\subsection{ Association Tests }
+Last, we use our model to study the impacts of isolation by distance in continuous space on genome-wide association studies (GWAS).
+Variation in ancestry proportions between case and control cohorts will cause inflation of test statistics at SNPs with different allele frequencies among populations (CITE).
+In addition, because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within and among populations can lead to inflated effect size estimates in GWAS of quantitative traits conducted on samples from large geographic regions \citep{Mathieson2012}.
+For example, imagine a population of flowering plants occurring over a large latitudinal range with populations subject to isolation by distance.
+Flowering time is strongly dependent on photoperiod and temperature, both of which typically vary with latitude.
+If we were to conduct a GWAS for flowering time without correcting for population structure we would find that all SNPs with allele frequencies varying clinally with latitude were significantly associated with flowering time, even if flowering time was determined purely by the environment.
+
+A common method of correcting for population structure in GWAS is to run a principal components analysis on the genotype matrix and include PC coordinates as covariates in a linear regression for each SNP \citep{Price2006}.
+Mixed model GWAS employs a different statistical framework but takes a similar approach by incorporating a kinship matrix describing pairwise genetic relatedness among samples as a random effect in a linear mixed model \citep{Kang2010}.
+In both approaches the key question is whether or not the ancestry components of the model adequately describe "background" levels of genetic variation in the sample relative to any confounding effects of ancestry and phenotype.
+Recent studies in European and British populations \citep{Kerminen2018,Berg2018,Sohail2018} have found evidence of residual population stratification in effect size estimates from large-scale GWAS employing both linear and mixed model approaches, suggesting that existing structure corrections may be inadequate.
+For traits controlled by a few large effect loci this confounding effect is likely to be relatively small, but for analyses of complex traits with thousands of putatively associated SNPs the potential for confounding is high.
+Here we simulate a range of spatially correlated phenotypes for simulated individuals and seek to identify conditions under which existing GWAS structure corrections fail.
+\section{Materials and Methods}
+\label{sec:materials:methods}
+
+\subsection{A Forward-Time Model of Evolution in Continuous Space}
+
+We implemented our model using the non-Wright-Fisher module in the program SLiM v3.0 \citep{Haller2019}.
+Each time step consists of three stages: reproduction, dispersal, and competition.
+To reduce the parameter space we use the same parameter, denoted $\sigma$, to modulate the spatial scale of interactions at all three stages by adjusting the standard deviation of the corresponding Gaussian functions.
+As in previous work \citep{Wright1943,Ringbauer2017}, $\sigma$ as applied in our dispersal step is approximately equal to the mean parent-offspring distance.
+
+At the beginning of the simulation individuals are distributed randomly on a continuous landscape.
+Mates are selected proportional to distances among individuals weighted by a Gaussian function with mean $1/(2\pi\sigma^2)$ and standard deviation $\sigma$.
+Wright's \cite{Wright1943} "Neighborhood Size", defined as $4\pi\sigma^2 K$ where $K$ is the population density, is then the approximate number of individuals available for mating in our simulation.
+The number of offspring is drawn from a Poisson distribution with mean $1/L$, where $L$ is the average lifespan.
+If new offspring are produced, we simulate dispersal by taking two draws from a normal distribution with mean 0 and standard deviation $\sigma$ and adding these to the x and y coordinates of the first parent.
+
+Density-dependent competition is simulated by adjusting the probability of survival of each individual proportional to the local density of individuals up to a distance $3\sigma$ away, scaled according to the same Gaussian function used for mate selection.
+Given a per-unit carrying capacity $K$ and average fecundity $F$, the probability of survival $d$ for individual $i$ after adjustment for density is:
+
+\begin{equation}
+    d_{i}=min(0.95,\frac{1}{1+\rho*\sum_{i}{G(x)}})
+\end{equation}
+where 
+\begin{equation}
+    \rho = F/((1+F)*K)
+\end{equation} 
+and $G(x)$ is the Gaussian function used to weight distances to other individuals.
+
+**NOTE: better ways to write this down? Should we write out the normal function too?
+
+A major challenge in all spatial models is dealing with range edges.
+When local population density is used to model competition, edge or corner populations can be assigned artificially high fitness values because they lack neighbors within their interaction radius but outside the bounds of the simulation.
+We approximate a decline in habitat suitability near edges by decreasing the probability of survival proportional to the square root of distance to edges in units of $\sigma$.
+The final probability of survival for individuals within one $\sigma$ of an edge is
+
+\begin{multline}
+    s_{i}=d_{i} min(1,\sqrt{x_{i}/\sigma})
+    min(1,\sqrt{y_{i}/\sigma})\\
+    min(1,\sqrt{(x_{max}-x_{i})/\sigma})
+    min(1,\sqrt{(y_{max}-y_{i})/\sigma)}
+\end{multline}
+
+where $x$ and $y$ are spatial coordinates.
+
+The full history of the genomes of final-generation individuals are stored after each timestep by SLiM and recorded as tree sequences \citep{Kelleher2018}.
+To speed up our simulations we set the mutation rate to 0 in SLiM simulations and later added mutations to the resulting tree sequences with msprime \citep{Kelleher2016}.
+Note that because time in SLiM is measured in timesteps rather than generation time, the desired per-generation rate must be scaled by generation time in this procedure (see "Genealogical Parameters" below for further information).
+
+We compared our model output to a two-dimensional stepping-stone model and a one-population coalescent model implemented in msprime \citep{Kelleher2016}.
+To isolate spatial effects from other components of the model such as overlapping generations and density-dependent probability of survival, we also ran a second version of the SLiM model with random mate selection and dispersal.
+
+We ran 400 simulations for the spatial and random-mating SLiM models on a square landscape of size 50 with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), average lifetime $L=4$, genome size $1e8$, and recombination rates $1e-9$.
+A mutation rate of $1e-8 mutations/site/generation$ was then applied to the coalesced tree sequence with msprime.
+$\sigma$ values were drawn from a uniform distribution bounded by 0.2 and 4.
+Simulations were run until all extant individuals shared a common ancestor within the simulation (i.e.
+the tree sequence had coalesced).
+Coalescent stepping-stone simulations were run on a 5x5 grid with a total $N_{e}$ of 10,000 ($N_{e}$ per population = 400) and 10 haploid samples per population; allowing migration between all adjacent populations.
+We ran 500 simulations drawing the number of migrants exchanged by adjacent demes from a uniform distribution between 0 and 40.
+
+\subsection{Genealogical Parameters}
+To study the geographic spread of genealogical ancestry in our simulations we output the full pedigree of all individuals over 50 generations with $\sigma$ set to 0.2, 0.35, 0.5, 1, 2, and 3.5 (corresponding to Wright's Neighborhood Sizes of 2.5, 7.7, 15.7, 62.8, 251.3, and 769.7, respectively).
+We then selected the individual closest to the center of the landscape in the final generation of each simulation, pruned the pedigree to include only its genealogical ancestors (that is, regardless of the amount of genetic material actually inherited from a given ancestor) and plotted the age of its most recent ancestor in each region of the full landscape using the 2-dimensional summary feature of ggplot2 in R \citep{Wickham2016}.
+We calculated the rate of spread of genealogical ancestry by estimating the mean and maximum distances between the focal individual and its ancestors in each generation and then fitting a linear model for distance as a function of generations using ordinary least squares regression.
+
+Because generation time in our model is not fixed and may vary both across simulations we also calculated "effective" generation times for all simulations.
+To do this we output the full tree sequence of all individuals in every time step for 200 time steps.
+Starting at the final time step we then calculated the average age of parents in each previous generation, sampling past individuals in proportion to their genetic contribution to the final generation, which can be extracted from the tree sequence.
+When adding mutations to the tree sequences produced by SLiM, we scaled the mutation rate by the generation time because msprime's "mutate()" function assumes that time is measured in units of generations.
+So to achieve a $1e-8 mutations/site/generation$ mutation rate for a simulation in which the generation time was 5 time steps, we entered $1e-8/5$ for the mutation rate in msprime.mutate().
+We then verified that this procedure produced the correct number of mutations by comparing a subset simulations with SLiM-generated mutations (which are applied only at mating events) with mutations added by msprime, finding that they were nearly identical in all cases.
+\subsection{Summary Statistics}
+We calculated a set of 20 summary statistics (Supplementary Table 1) from 60 diploid individuals sampled randomly from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}.
+Statistics included common single-population summaries such as mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{is}$), and Tajima's D, as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distances \citep{Rousset1997}, which we summarized as the correlation coefficient between $log10(spatial distance)$ and the proportion of identical base pairs among individuals.
+
+Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) haplotype block sharing among samples.
+Intuitively, haplotype block length distributions should contain more information about the genealogy of a set of samples than allele frequency measures alone because they reflect the influence of two Poisson processes (mutation and recombination) acting along its branches, rather than just one.
+
+The full distribution of lengths of IBS tracts for each pair of individuals was first calculated with a custom python function.
+We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $1e6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.
+To assess the degree to which the IBS tract length distribution shifts with spatial distance, we then estimated correlation coefficients between spatial distance and each moment of the pairwise IBS tract distribution.
+Because more closely related individuals on average share longer haplotype blocks we expect that spatial distance will be negatively correlated with mean haplotype block length, and that this correlation will be strongest (i.e.
+most negative) when dispersal is low.
+The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks and is thus most informative of recent demographic events \citep{Ringbauer2017} (cite Coop \& Ralph here \& above).
+
+\subsection{Demographic Modeling}
+We fit single-population demographic models to the site frequency spectra of 60 randomly sampled individuals from each simulation with the program Stairwayplot \citep{Liu2015}.
+Site frequency spectra used for input data were calculated in scikit-allel.
+100 bootstrap replicates were generated for each simulation by resampling genotype matrices with replacement over sites and recalculating the site frequency spectra.
+Models were then fit across all bootstrap replicates using default settings in Stairwayplot and the median estimate of Ne per time unit across bootstrap replicates was used to represent the output of each simulation.
+The resulting inferred population size histories were plotted with line colors scaled to $\sigma$ to visualize how dispersal scaling impacts estimates of demographic history.
+
+\subsection{Association Studies}
+To assess the degree to which spatial structure confounds GWAS we simulated four types of phenotypic variation for 1000 randomly sampled final-generation individuals in each SLiM simulation and conducted a linear-regression GWAS with PC covariates in PLINK \citep{PURCELL2007}.
+SNPs with a minor allele frequency less than 1\% were excluded from the analysis.
+Phenotype values were meant to roughly reflect the distribution of height across Europe, which has recently been found to be confounded with population structure in large scale GWAS \citep{Berg2018,Sohail2018}.
+Conceptually our approach is similar to that taken in \citep{Mathieson2012}, though here we model continuous spatial variation and focus on genome-wide effects rather than low-frequency alleles specifically.
+
+In the first simulation phenotypes for all individuals were drawn from a normal distribution with mean 110 and standard deviation 10.
+Next we simulated clinal environmental influences on phenotype by drawing the phenotypes from independent normal distributions in which the mean was scaled by an individual's x position such that it varied by 2 standard deviations across the map.
+Third, we approximate concentrated environmental effects by drawing phenotypes for individuals with x and y coordinates below 20 from a normal distribution with mean 2 standard deviations above the rest of the map.
+Last, we simulated a "patchy" environmental influence on phenotypes by selecting 10 random points on the map and drawing phenotypes for all individuals within two map units of any selected point from a normal distribution with mean 2 standard deviations above the rest of the map.
+
+Principal components analysis (PCA) was conducted in scikit allel on the matrix of derived allele counts by individual for each simulation.
+SNPs were first filtered to remove strongly linked sites by calculating LD between all pairs of SNPs in a 200-SNP moving window and dropping one of each pair of sites with an $R^2$ over 0.1.
+The LD-pruned allele count matrix was then centered and all sites scaled to unit variance when conducting the PCA, following recommendations in \citep{Patterson2006}.
+
+We ran linear-model GWAS both with and without the first 10 principal components as covariates in PLINK and summarized results across simulations by counting the number of significant SNPs with an expected false positive rate of less than 5\% after adjusting p values with the R function p.adjust(...,method="fdr").
+We also examined $p$ values for systemic inflation by estimating the expected values from a uniform distribution (because no SNPs were used when generating phenotypes), plotting observed against expected values for all simulations, and using the two-dimensional binning feature of ggplot2 \citep{Wickham2016} to find the mean $\sigma$ value in each region of quantile-quantile space.
 
 \section{Results}
+\subsection{Runtime and Memory Requirements}
+\subsection{Summary Statistics}
+\subsection{Demographic Modeling}
+\subsection{GWAS}
 
-\subsection{ Summary Statistics }
+\section{Discussion}
 
-\subsection{ Impacts on Demographic Inference and GWAS }
+\section{Data Availability}
 
-\subsection{ Estimating Model Parameters with Machine Learning }
+\section{Acknowledgements}
 
-\section{ Discussion }
+\begin{figure*}[h!]
+\centering
+\includegraphics[width=\textwidth]{sfs_spatial_v_rm.pdf}
+\caption{Scaled allele frequency distributions for random mating and spatial SLiM models under varying $\sigma$.
+The rightmost panel is the difference between the spatial and random-mating distributions, showing an inflation of mid-frequency sites for neighborhood sizes of less than 50.
+Lines are a local regression (loess) model at span 2.}
+\label{fig:spectrum}
+\end{figure*}
 
-\section{Figures}
+\begin{figure*}[h!]
+\centering
+\includegraphics[width=\textwidth]{min_ancestor_age_maps.pdf}
+\caption{Geographic spread of genealogical ancestors from a random individual over 50 generations, by Neighborhood size and $\sigma$. Colors are scaled by the age of the most recent ancestor.}
+\label{fig:spectrum}
+\end{figure*}
 
-\begin{figure}[h]
-	\centerline{
-	\includegraphics{../figures/10kouts_coalescence_by_generation}
-	}
+\begin{figure*}[h!]
+\centering
+\includegraphics[width=\textwidth]{gwas_summary_sigma225.png}
+\caption{Impacts of spatially correlated phenotypes and isolation by distance on GWAS.
+Colors represent different phenotype distributions as indicated in the figure legend.
+A: numbers of significant SNPs after FDR correction for linear-model GWAS conducted with (top) or without (bottom) PC covariates.
+B: Example Manhattan plots for a single simulation with varying phenotype distributions.
+Points in black are significant at an estimated FDR cutoff of 0.05, while the dotted line shows the Bonferroni-corrrected \textit{p} value of 0.05.
+In practice most human GWAS studies use a significance cutoff between these points.
+C and D show the corresponding spatial and principal component positions, colored by phenotype.}
+\label{fig:spectrum}
+\end{figure*}
+
+\newpage
+
+\bibliography{spaceness}
+
+\end{document}
+
+\section{Genetics Submission Guidelines}
+Manuscripts submitted to \textit{GENETICS} should contain a clear description of the experimental design in sufficient detail so that the experimental analysis could be repeated by another scientist.
+If the level of detail necessary to explain the protocol goes beyond two paragraphs, give a short description in the main body of the paper and prepare a detailed description for supporting information.
+For example, details would include indicating how many individuals were used, and if applicable how individuals or groups were combined for analysis.
+If working with mutants indicate how many independent mutants were isolated.
+If working with populations indicate how samples were collected and whether they were random with respect to the target population.
+
+\subsection{Statistical Analysis}
+
+It is important to indicate what statistical analysis has been performed; not just the name of the software and options selected, but the method and model applied.
+In the case of many genes being examined simultaneously, or many phenotypes, a multiple comparison correction should be used to control the type I error rate, or a rationale for not applying a correction must be provided.
+The type of correction applied should be clearly stated.
+It should also be clear whether the p-values reported are raw, or after correction.
+Corrected p-values are often appropriate, but raw p-values should be available in the supporting materials so that others may perform their own corrections.
+In large scale data exploration studies (e.g.
+genome wide expression studies) a clear and complete description of the replication structure must be provided.
+
+\subsection{Data Availability}
+
+At the end of the Materials and Methods section, include a statement on reagent and data availability.
+Please read the Data and Reagent Policy before writing the statement.
+Make sure to list the accession numbers or DOIs of any data you have placed in public repositories.
+List the file names and descriptions of any data you will upload as supplemental information.
+The statement should also include any applicable IRB numbers.
+You may include specifications for how to properly acknowledge or cite the data.
+
+For example: Strains are available upon request.
+File S1 contains detailed descriptions of all supplemental files.
+File S2 contains SNP ID numbers and locations.
+File S3 contains genotypes for each individual.
+Sequence data are available at GenBank and the accession numbers are listed in File S3.
+Gene expression data are available at GEO with the accession number: GDS1234.
+Code used to generate the simulated data is provided in file S4.
+
+\section{Results and Discussion}
+
+The results and discussion should not be repetitive.
+The results section should give a factual presentation of the data and all tables and figures should be referenced; the discussion should not summarize the results but provide an interpretation of the results, and should clearly delineate between the findings of the particular study and the possible impact of those findings in a larger context.
+Authors are encouraged to cite recent work relevant to their interpretations.
+Present and discuss results only once, not in both the Results and Discussion sections.
+It is sometimes acceptable to combine results and discussion.
+The text should be as succinct as possible.
+Heed Strunk and White's dictum: "Omit needless words!"
+
+\section{Additional guidelines}
+
+\subsection{Numbers} In the text, write out numbers nine or less except as part of a date, a fraction or decimal, a percentage, or a unit of measurement.
+Use Arabic numbers for those larger than nine, except as the first word of a sentence; however, try to avoid starting a sentence with such a number.
+
+\subsection{Units} Use abbreviations of the customary units of measurement only when they are preceded by a number: "3 min" but "several minutes".
+Write "percent" as one word, except when used with a number: "several percent" but "75\%." To indicate temperature in centigrade, use ° (for example, 37°); include a letter after the degree symbol only when some other scale is intended (for example, 45°K).
+
+\subsection{Nomenclature and Italicization} Italicize names of organisms even when  when the species is not indicated.
+Italicize the first three letters of the names of restriction enzyme cleavage sites, as in HindIII.
+Write the names of strains in roman except when incorporating specific genotypic designations.
+Italicize genotype names and symbols, including all components of alleles, but not when the name of a gene is the same as the name of an enzyme.
+Do not use "+" to indicate wild type.
+Carefully distinguish between genotype (italicized) and phenotype (not italicized) in both the writing and the symbolism.
+
+\subsection{Cross References}
+Use the \verb|\nameref| command with the \verb|\label| command to insert cross-references to section headings.
+For example, a \verb|\label| has been defined in the section \nameref{sec:materials:methods}.
+
+\section{In-text Citations}
+
+Add citations using the \verb|\citep{}| command, for example \citep{neher2013genealogies} or for multiple citations, \citep{neher2013genealogies, rodelsperger2014characterization,Falush16}
+
+\section{Examples of Article Components}
+\label{sec:examples}
+
+The sections below show examples of different header levels, which you can use in the primary sections of the manuscript (Results, Discussion, etc.) to organize your content.
+
+\section{First level section header}
+
+Use this level to group two or more closely related headings in a long article.
+
+\subsection{Second level section header}
+
+Second level section text.
+
+\subsubsection{Third level section header:}
+
+Third level section text. These headings may be numbered, but only when the numbers must be cited in the text.
+
+\section{Figures and Tables}
+
+Figures and Tables should be labelled and referenced in the standard way using the \verb|\label{}| and \verb|\ref{}| commands.
+
+\subsection{Sample Figure}
+
+Figure \ref{fig:spectrum} shows an example figure.
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=\linewidth]{example-figure}
+\caption{Example figure from \url{10.1534/genetics.114.173807}.
+Please include your figures in the manuscript for the review process.
+You can upload figures to Overleaf via the Project menu.
+Upon acceptance, we'll ask for your figure files to be uploaded in any of the following formats: TIFF (.tiff), JPEG (.jpg), Microsoft PowerPoint (.ppt), EPS (.eps), or Adobe Illustrator (.ai).
+Images should be a minimum of 300 dpi in resolution and 500 dpi minimum if line art images.
+RGB, CMYK, and Grayscale are all acceptable.
+Halftones should be high contrast with sharp detail, because some loss of detail and contrast is inevitable in the production process.
+Figures should be 10-20 cm in width and 1-25 cm in height.
+Graph axes must be exactly perpendicular and all lines of equal density.
+Label multiple figure parts with A, B, etc.
+in bolded type, and use Arrows and numbers to draw attention to areas you want to highlight.
+Legends should start with a brief title and should be a self-contained description of the content of the figure that provides enough detail to fully understand the data presented.
+All conventional symbols used to indicate figure data points are available for typesetting; unconventional symbols should not be used.
+Italicize all mathematical variables (both in the figure legend and figure) , genotypes, and additional symbols that are normally italicized.
+}%
+\label{fig:spectrum}
 \end{figure}
 
-\begin{figure}[h]
-	\centerline{
-	\includegraphics{../figures/sumstat_by_neighbors_spat_v_rm.pdf}
-	}
+\subsection{Sample Video}
+
+Figure \ref{video:spectrum} shows how to include a video in your manuscript.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics[width=\linewidth]{example-figure}
+\caption{Example movie (the figure file above is used as a placeholder for this example).
+\textit{GENETICS} supports video and movie files that can be linked from any portion of the article - including the abstract.
+Acceptable formats include .asf, avi, .wav, and all types of Windows Media files.
+}%
+\label{video:spectrum}
 \end{figure}
 
-\begin{figure}[h]
-	\centerline{
-	\includegraphics{../figures/sfs_spatial_v_rm.pdf}
-	}
-\end{figure}
+\subsection{Sample Table}
 
-\end{document}
+Table \ref{tab:shape-functions} shows an example table.
+Avoid shading, color type, line drawings, graphics, or other illustrations within tables.
+Use tables for data only; present drawings, graphics, and illustrations as separate figures.
+Histograms should not be used to present data that can be captured easily in text or small tables, as they take up much more space.
+
+Tables numbers are given in Arabic numerals. Tables should not be numbered 1A, 1B, etc., but if necessary, interior parts of the table can be labeled A, B, etc. for easy reference in the text.
+
+\begin{table*}[htbp]
+\centering
+\caption{\bf Students and their grades}
+\begin{tableminipage}{\textwidth}
+\begin{tabularx}{\textwidth}{XXXX}
+\hline
+Student & Grade\footnote{This is an example of a footnote in a table.
+Lowercase, superscript italic letters (a, b, c, etc.) are used by default.
+You can also use *, **, and *** to indicate conventional levels of statistical significance, explained below the table.} & Rank & Notes \\
+\hline
+Alice & 82\% & 1 & Performed very well.\\
+Bob & 65\% & 3 & Not up to his usual standard.\\
+Charlie & 73\% & 2 & A good attempt.\\
+\hline
+\end{tabularx}
+  \label{tab:shape-functions}
+\end{tableminipage}
+\end{table*}
+
+\section{Sample Equation}
+
+Let $X_1, X_2, \ldots, X_n$ be a sequence of independent and identically distributed random variables with $\text{E}[X_i] = \mu$ and $\text{Var}[X_i] = \sigma^2 < \infty$, and let
+\begin{equation}
+S_n = \frac{X_1 + X_2 + \cdots + X_n}{n}
+      = \frac{1}{n}\sum_{i}^{n} X_i
+\label{eq:refname1}
+\end{equation}
+denote their mean. Then as $n$ approaches infinity, the random variables $\sqrt{n}(S_n - \mu)$ converge in distribution to a normal $\mathcal{N}(0, \sigma^2)$.

### 2019-03-13 ms/spaceness.tex
ms updates

@@
 
-\documentclass[9pt,twocolumn,twoside]{gsajnl}
+\documentclass[9pt,twocolumn,twoside,lineno]{gsajnl}
 \usepackage{amsmath}
@@
 
-\articletype{inv} % article type
+\articletype{preprint} % article type
 % {inv} Investigation 
@@
 \author[$\ast$,1]{C.J. Battey}
-\author[$\dagger$,1]{Peter Ralph}
-\author[$\ddagger$,2]{Andrew Kern}
+\author[$\ast$]{Peter Ralph}
+\author[$\ast$]{Andrew Kern}
 
@@
 As the availability of population-level, whole-genome data allows inference of increasingly fine-scale patterns of ancestry in many species, models incorporating realistic demographic and spatial processes are needed to accurately describe spatial structure and control for its impacts on analyses of selection and demography.
-Here we implement a forward-time continuous-space model of molecular evolution and use it to study the impacts of limited dispersal in continuous landscapes on population genetic summary statistics, demographic inference, and association studies.
-Low dispersal slows the geographic spread of ancestry and inflates branch lengths for mid-frequency alleles because of slow coalescence among distant individuals.
-This inflation can bias demographic inference from the site frequency spectrum towards inferring a recent spike in population sizes.
-In a GWAS setting spatially correlated phenotypes cause spurious signals of genetic association when principal components analyses fail to capture the relevant spatial scale of phenotypic variation, which can occur under either very limited or very high dispersal.
+Here we implement a forward-time simulation of molecular evolution in continuous space and use it to study the impacts of dispersal and density-dependent competition on population genetic summary statistics, demographic inference, and association studies.
+Low dispersal slows the geographic spread of ancestry and increases branch lengths corresponding to mid-frequency alleles because of slow coalescence among distant groups of individuals.
+As a result demographic inference from the site frequency spectrum can be biased towards inferring a spike in past population sizes.
+Spatially correlated phenotypes cause spurious signals of genetic association when principal components analyses fail to capture the relevant spatial scale of phenotypic variation, which can occur under either very limited or very high dispersal.
 Last we use a machine learning regression approach to evaluate summary statistics as estimators of dispersal distance and find that correlation coefficients between spatial distance and moments of the distribution of pairwise haplotype block lengths retain strong signal even at high dispersal distances.
@@
 
-\correspondingauthoraffiliation{3}{University of Oregon Dept. Biology, Institute for Ecology and Evolution.}
+\correspondingauthoraffiliation{1}{301 Pacific Hall, University of Oregon Dept. Biology, Institute for Ecology and Evolution. cbattey2@uoregon.edu.}
 \vspace{-35pt}% Only used for adjusting extra space in the left column of the first page
@@
 lit?)
-
-NOTE: probably cut this paragraph???
-
-This process is analogous to genetic drift, with spatial locations and population densities taking the place of sites and allele frequencies.
-Because individuals disperse from their natal location, new offspring are likely to be close to their parents.
-If by chance an individual has no offspring the local population density is then lowered, which decreases the chance of new individuals dispersing into that area.
-This process compounds over time until only one or two patches of habitat are occupied at any given time.
-The resulting spatial clustering clearly does not describe spatial dynamics in most real populations, because declining growth rates in regions of high population density are common in both lab ( ) and wild populations ( ).
 
@@
 Last, we use our model to study the impacts of isolation by distance in continuous space on genome-wide association studies (GWAS).
-Variation in ancestry proportions between case and control cohorts will cause inflation of test statistics at SNPs with different allele frequencies among populations (CITE).
-In addition, because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within and among populations can lead to inflated effect size estimates in GWAS of quantitative traits conducted on samples from large geographic regions \citep{Mathieson2012}.
+Variation in ancestry proportions between case and control cohorts is expected to cause inflation of test statistics at SNPs with different allele frequencies among populations (CITE).
+In addition, because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within and among populations can lead to inflated effect size estimates in GWAS of quantitative traits conducted on samples from large geographic regions, and this effect is particularly prominent in low-frequency alleles \citep{Mathieson2012}.
 For example, imagine a population of flowering plants occurring over a large latitudinal range with populations subject to isolation by distance.
@@
 A common method of correcting for population structure in GWAS is to run a principal components analysis on the genotype matrix and include PC coordinates as covariates in a linear regression for each SNP \citep{Price2006}.
-Mixed model GWAS employs a different statistical framework but takes a similar approach by incorporating a kinship matrix describing pairwise genetic relatedness among samples as a random effect in a linear mixed model \citep{Kang2010}.
+Mixed model GWAS employs a different statistical framework but takes a similar approach by incorporating a kinship matrix describing pairwise genetic relatedness among samples as a random effect in a linear mixed model \citep{Yu2006,Kang2008}.
 In both approaches the key question is whether or not the ancestry components of the model adequately describe "background" levels of genetic variation in the sample relative to any confounding effects of ancestry and phenotype.
@@
 
-We ran 400 simulations for the spatial and random-mating SLiM models on a square landscape of size 50 with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), average lifetime $L=4$, genome size $1e8$, and recombination rates $1e-9$.
-A mutation rate of $1e-8 mutations/site/generation$ was then applied to the coalesced tree sequence with msprime.
-$\sigma$ values were drawn from a uniform distribution bounded by 0.2 and 4.
+We ran 400 simulations for the spatial and random-mating SLiM models on a square landscape of size 50 with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), average lifetime $L=4$, genome size $1\times10^{8}$, recombination rate $1\times10^{-9}$, and drawing $/sigma$ values from a uniform distribution bounded by 0.2 and 4.
+To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM and later applied mutations to the coalesced tree sequence with msprime's 'mutate( )' function.
+Because msprime assumes that timesteps are in units of generations, we divided the per-generation rate of $1\times10^{-8} mut/site/gen$ by the generation time estimated for each value of $\sigma$ (see 'Genealogical Parameters' below) to convert the rate to units of mutations/site/timestep.
+We then verified that this procedure produced the correct number of mutations by comparing a subset of simulations with SLiM-generated mutations (which are applied only at mating events) with mutations added by msprime, finding that they were nearly identical in all cases.
 Simulations were run until all extant individuals shared a common ancestor within the simulation (i.e.
 the tree sequence had coalesced).
-Coalescent stepping-stone simulations were run on a 5x5 grid with a total $N_{e}$ of 10,000 ($N_{e}$ per population = 400) and 10 haploid samples per population; allowing migration between all adjacent populations.
-We ran 500 simulations drawing the number of migrants exchanged by adjacent demes from a uniform distribution between 0 and 40.
+
+Coalescent stepping-stone simulations were on a 5x5 grid with 10 haploid samples per deme and migration between adjacent populations.
+To facilitate comparison across simulation types we ran one coalescent simulation for every SLiM parameter set.
+Because $\sigma$ is approximately equal to the mean distance between parent and offspring, we approximate the relationship between $\sigma$ and the migration rate in a grid simulation as $\sigma=4md$, where $m$ is the migration rate and $d$ is the ratio of landscape widths in continuous space vs coalescent simulations (here, 10).
+That is, an individual in an interior population has a $4m$ probability of being a migrant, and if it is a migrant will move $d$ units on the map.
+We set the total $N_{e}$ equal to the census size of each SLiM simulation divided by the variance in reproductive output (see below).
 
@@
 
-Because generation time in our model is not fixed and may vary both across simulations we also calculated "effective" generation times for all simulations.
-To do this we output the full tree sequence of all individuals in every time step for 200 time steps.
-Starting at the final time step we then calculated the average age of parents in each previous generation, sampling past individuals in proportion to their genetic contribution to the final generation, which can be extracted from the tree sequence.
-When adding mutations to the tree sequences produced by SLiM, we scaled the mutation rate by the generation time because msprime's "mutate()" function assumes that time is measured in units of generations.
-So to achieve a $1e-8 mutations/site/generation$ mutation rate for a simulation in which the generation time was 5 time steps, we entered $1e-8/5$ for the mutation rate in msprime.mutate().
-We then verified that this procedure produced the correct number of mutations by comparing a subset simulations with SLiM-generated mutations (which are applied only at mating events) with mutations added by msprime, finding that they were nearly identical in all cases.
+Next we estimated the variance in reproductive output and the generation time of individuals in our simulations.
+To do this we output the full tree sequence of all individuals for 200 time steps.
+Starting at the final time step we then calculated the average age of parents in each previous generation as an estimate of generation time, sampling past individuals in proportion to their genetic contribution to the final generation.
+To estimate variance in reproductive output we calculated the variance in number of offspring for all individuals born in timesteps 50-150 (to ensure that the simulation captured the full lifespan of the sampled individuals).
+
 \subsection{Summary Statistics}
@@
 Site frequency spectra used for input data were calculated in scikit-allel.
-100 bootstrap replicates were generated for each simulation by resampling genotype matrices with replacement over sites and recalculating the site frequency spectra.
-Models were then fit across all bootstrap replicates using default settings in Stairwayplot and the median estimate of Ne per time unit across bootstrap replicates was used to represent the output of each simulation.
+100 bootstrap replicates were generated for each simulation by resampling over sites.
+Models were then fit across all bootstrap replicates using default settings in Stairwayplot and the median estimate of $Ne$ per generation across bootstrap replicates was used to represent the output of each simulation.
 The resulting inferred population size histories were plotted with line colors scaled to $\sigma$ to visualize how dispersal scaling impacts estimates of demographic history.
@@
 To assess the degree to which spatial structure confounds GWAS we simulated four types of phenotypic variation for 1000 randomly sampled final-generation individuals in each SLiM simulation and conducted a linear-regression GWAS with PC covariates in PLINK \citep{PURCELL2007}.
-SNPs with a minor allele frequency less than 1\% were excluded from the analysis.
+SNPs with a minor allele frequency less than 1\% were excluded from this analysis.
 Phenotype values were meant to roughly reflect the distribution of height across Europe, which has recently been found to be confounded with population structure in large scale GWAS \citep{Berg2018,Sohail2018}.
@@
 \section{Results}
+
 \subsection{Runtime and Memory Requirements}
+
 \subsection{Summary Statistics}
+
 \subsection{Demographic Modeling}
+
 \subsection{GWAS}
+Spatially correlated phenotypes in populations with low dispersal distances cause uncorrected association analyses to report significant associations at most genome-wide SNPs (Figure 1).
+No inflation was observed for the "normal" phenotype distribution which is random with respect to space, demonstrating that this effect is caused by the interaction of isolation by distance and spatially correlated phenotypes.
+The deflation in $p$ values for spatial phenotypes is particularly pronounced for the corner and clinal distributions, likely because in these cases our sample captures more phenotypic variation than in the "patchy" distribution.
+When PCA positions are included as covariates the vast majority of SNPs no longer pass a 5\% FDR significance threshold, but up to 1000 genome-wide significant hits (out of an average of 150,000 SNPs) were observed even in relatively high dispersal for the corner and clinal distributions.
+
+Notably, many SNPs inferred as significant are in close proximity along the genome and likely reflect haplotypes that became common in one region of the landscape (Figure 1, supplementary figures X) [note: add example manhattan plots for s=0.3 and s=3 to fig, supp figs for manhattan plots across a larger range of sigma].
+Thus the assumption that patterns of linkage disequilibrium are uncorrelated with allele frequency differentiation, which is the basis of LD score regression \citep{Bulik-Sullivan2015}, may not hold for these simulations.
+###TODO revisit this and
+
+In the case of clinal phenotypes, PCA covariates perform very well in correcting for population stratification at low dispersal distances but the correction breaks down above neighborhood sizes of 250.
+This pattern appears to reflect a loss of descriptive power in the first ten axes of the PCA, which describe less of the total variation in the data as neighborhood size increases (Figure 1 PANEL TO BE ADDED).
+Essentially, PCA loses descriptive power faster than the breakdown in correlations between allele frequencies and spatial position that our GWAS is picking up on, so eventually the statistical correction breaks down.
+
+Examining quantile-quantile plots shows that $p$ value deflation is most pronounced for the corner phenotype distribution and is particularly strong for simulations with neighborhood sizes of less than 20.
+These results are similar to the observation of high inflation for spatially discrete phenotypes in \citep{Mathieson2012}.
+Notably here we find that the deflation in $p$ values is not limited to low-frequency alleles, and is not entirely controlled by including PCA covariates.
+Quantile-quantile plots of clinal and nonspatial phenotypes also show evidence of $p$ value $inflation$ (i.e.
+points below the 1:1 line), which is observed only when PCA covariates are included in the analysis (see Supplementary Figure X for q-q plots of uncorrected GWAS results).
+This likely reflects the loss of power in GWAS when both phenotype and ancestry are correlated with space, which creates a partially confounded study design.
 
@@
 \centering
-\includegraphics[width=\textwidth]{gwas_summary_sigma225.png}
+\includegraphics[width=\textwidth]{gwas_summary_nsig_qqplots.pdf}
 \caption{Impacts of spatially correlated phenotypes and isolation by distance on GWAS.
@@
 A: numbers of significant SNPs after FDR correction for linear-model GWAS conducted with (top) or without (bottom) PC covariates.
-B: Example Manhattan plots for a single simulation with varying phenotype distributions.
-Points in black are significant at an estimated FDR cutoff of 0.05, while the dotted line shows the Bonferroni-corrrected \textit{p} value of 0.05.
-In practice most human GWAS studies use a significance cutoff between these points.
-C and D show the corresponding spatial and principal component positions, colored by phenotype.}
+B: Quantile-quantile plots of observed versus $p$ values relative to those expected from a uniform distribution assuming no effect on phenotypes.
+The dotted lines show the 95\% confidence region assuming binomial sampling and an alpha level of 0.05.
+Points above the 1:1 line reflect inflated -log10(p).}
 \label{fig:spectrum}
@@
 \end{document}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2019-03-15 ms/spaceness.tex
ms updates

@@
 %Impacts of Continuous Spatial Structure on Analyses of Population Genetic Data
-\title{Space is the Place: How Dispersal Shapes Genetic Variation in Continuous Space}
+\title{Space is the Place: Impacts of Continuous Spatial Structure on Demographic Modeling and Association Studies}
 
@@
 As the availability of population-level, whole-genome data allows inference of increasingly fine-scale patterns of ancestry in many species, models incorporating realistic demographic and spatial processes are needed to accurately describe spatial structure and control for its impacts on analyses of selection and demography.
-Here we implement a forward-time simulation of molecular evolution in continuous space and use it to study the impacts of dispersal and density-dependent competition on population genetic summary statistics, demographic inference, and association studies.
+Here we implement a forward-time simulation of evolution in continuous space and use it to study the impacts of dispersal and density-dependent competition on population genetic summary statistics, demographic inference, and association studies.
 Low dispersal slows the geographic spread of ancestry and increases branch lengths corresponding to mid-frequency alleles because of slow coalescence among distant groups of individuals.
-As a result demographic inference from the site frequency spectrum can be biased towards inferring a spike in past population sizes.
-Spatially correlated phenotypes cause spurious signals of genetic association when principal components analyses fail to capture the relevant spatial scale of phenotypic variation, which can occur under either very limited or very high dispersal.
-Last we use a machine learning regression approach to evaluate summary statistics as estimators of dispersal distance and find that correlation coefficients between spatial distance and moments of the distribution of pairwise haplotype block lengths retain strong signal even at high dispersal distances.
+As a result demographic inference from the site frequency spectrum is biased towards inferring a turbulent past demographic history and a recent decline in populations when neighborhood sizes are below 20.
+In association analyses spatially correlated environments cause genome-wide signals of genetic association with purely environmental phenotypes, and this bias is only partially corrected by regressing out principal components positions during the analysis.
+We demonstrate that stratification bias in continuous populations is worst when environmental effects are spatially discrete, but that corrections for clinal phenotypes can also fail under high dispersal because PCA describes less total variation in the data when isolation by distance is weak.
+
 \end{abstract}
@@
 
-**NOTE: better ways to write this down? Should we write out the normal function too?
-
 A major challenge in all spatial models is dealing with range edges.
@@
 
-The full history of the genomes of final-generation individuals are stored after each timestep by SLiM and recorded as tree sequences \citep{Kelleher2018}.
-To speed up our simulations we set the mutation rate to 0 in SLiM simulations and later added mutations to the resulting tree sequences with msprime \citep{Kelleher2016}.
-Note that because time in SLiM is measured in timesteps rather than generation time, the desired per-generation rate must be scaled by generation time in this procedure (see "Genealogical Parameters" below for further information).
+To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, or density-dependent fitness, we  also implemented a second version of our simulation in which mates are selected at random and offspring disperse to a random location on the landscape.
+For all simulations the full history of the genomes of final-generation individuals were stored as tree sequences \citep{Kelleher2018}.
 
-We compared our model output to a two-dimensional stepping-stone model and a one-population coalescent model implemented in msprime \citep{Kelleher2016}.
+%We compared our model output to a two-dimensional stepping-stone model and a one-population coalescent model implemented in msprime \citep{Kelleher2016}.
 To isolate spatial effects from other components of the model such as overlapping generations and density-dependent probability of survival, we also ran a second version of the SLiM model with random mate selection and dispersal.
@@
 Because msprime assumes that timesteps are in units of generations, we divided the per-generation rate of $1\times10^{-8} mut/site/gen$ by the generation time estimated for each value of $\sigma$ (see 'Genealogical Parameters' below) to convert the rate to units of mutations/site/timestep.
-We then verified that this procedure produced the correct number of mutations by comparing a subset of simulations with SLiM-generated mutations (which are applied only at mating events) with mutations added by msprime, finding that they were nearly identical in all cases.
+We then verified that this procedure produced the correct number of mutations by comparing a subset of simulations with SLiM-generated mutations (which are applied only at mating events) with mutations added by msprime (Supplementary Figure 1).
 Simulations were run until all extant individuals shared a common ancestor within the simulation (i.e.
@@
 
-Coalescent stepping-stone simulations were on a 5x5 grid with 10 haploid samples per deme and migration between adjacent populations.
+%Coalescent stepping-stone simulations were on a 5x5 grid with 10 haploid samples per deme and migration between adjacent populations.
 To facilitate comparison across simulation types we ran one coalescent simulation for every SLiM parameter set.
@@
 
-Next we estimated the variance in reproductive output and the generation time of individuals in our simulations.
-To do this we output the full tree sequence of all individuals for 200 time steps.
-Starting at the final time step we then calculated the average age of parents in each previous generation as an estimate of generation time, sampling past individuals in proportion to their genetic contribution to the final generation.
-To estimate variance in reproductive output we calculated the variance in number of offspring for all individuals born in timesteps 50-150 (to ensure that the simulation captured the full lifespan of the sampled individuals).
+We also estimated the variance in offspring and generation time for all simulations.
+To estimate generation times we stored the age of all parents for 200 timesteps and took the mean.
+To estimate variance in offspring we output the complete pedigree of all individuals in SLiM, counted the number of offspring for all individuals born in timesteps 50-150 (to ensure that the simulation captured the full lifespan of the sampled individuals), and calculated the variance.
 
@@
 
-Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) haplotype block sharing among samples.
+Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among samples.
 Intuitively, haplotype block length distributions should contain more information about the genealogy of a set of samples than allele frequency measures alone because they reflect the influence of two Poisson processes (mutation and recombination) acting along its branches, rather than just one.
@@
 \subsection{Demographic Modeling}
-We fit single-population demographic models to the site frequency spectra of 60 randomly sampled individuals from each simulation with the program Stairwayplot \citep{Liu2015}.
-Site frequency spectra used for input data were calculated in scikit-allel.
+We fit single-population demographic models to the site frequency spectra of 20 randomly sampled individuals from each spatial SLiM simulation with the program Stairwayplot \citep{Liu2015}.
+Site frequency spectra used for input data were calculated in scikit-allel \citep{Miles2017}.
 100 bootstrap replicates were generated for each simulation by resampling over sites.
-Models were then fit across all bootstrap replicates using default settings in Stairwayplot and the median estimate of $Ne$ per generation across bootstrap replicates was used to represent the output of each simulation.
+Models were then fit across all bootstrap replicates using default settings in Stairwayplot and the median estimate of $Ne$ per generation was used to represent the output of each simulation.
 The resulting inferred population size histories were plotted with line colors scaled to $\sigma$ to visualize how dispersal scaling impacts estimates of demographic history.
@@
 \subsection{Association Studies}
-To assess the degree to which spatial structure confounds GWAS we simulated four types of phenotypic variation for 1000 randomly sampled final-generation individuals in each SLiM simulation and conducted a linear-regression GWAS with PC covariates in PLINK \citep{PURCELL2007}.
-SNPs with a minor allele frequency less than 1\% were excluded from this analysis.
+To assess the degree to which spatial structure confounds GWAS we simulated four types of nongenetic phenotype variation for 1000 randomly sampled individuals in each spatial SLiM simulation and conducted a linear-regression GWAS with PC covariates in PLINK \citep{PURCELL2007}.
+SNPs with a minor allele frequency less than 0.5\% were excluded from this analysis.
 Phenotype values were meant to roughly reflect the distribution of height across Europe, which has recently been found to be confounded with population structure in large scale GWAS \citep{Berg2018,Sohail2018}.
-Conceptually our approach is similar to that taken in \citep{Mathieson2012}, though here we model continuous spatial variation and focus on genome-wide effects rather than low-frequency alleles specifically.
+Conceptually our approach is similar to that taken in \citep{Mathieson2012}, though here we model fully continuous spatial variation and focus on genome-wide effects rather than low-frequency alleles specifically.
 
@@
 
-\subsection{Runtime and Memory Requirements}
+\subsection{Genealogical Parameters and Run Times}
 
 \subsection{Summary Statistics}
+As expected from theory based on the Wright-Malécot model \citep{Maruyama1972}, limited dispersal increases the coalescence time of a set of randomly-sampled individuals relative to a random-mating model.
+This is reflected in sequence data as a larger number of segregating sites and inflated estimates of expected heterozygosity ($\pi$, Figure 2).
+In practice the difference in segregating sites and expected heterozygosity in our simulations is negligible above neighborhood sizes of 65 ($\sigma$=1).
+Conversely observed heterozygosity (the fraction of heterozygous individuals) is lower than the random mating expectation when dispersal is low, and as a consequence the inbreeding coefficient ($F_{is}$) is positive.
+This is the continuous space version of the Wahlund effect \citep{Wahlund1928} - a deficit of heterozygous individuals when samples are drawn from subdivided populations.
+Similar observations of positive $F_{is}$ have been observed from stepping-stone simulations in \citep{Shirk2014} and \citep{Neel2013}.
+
+Tajima's $D$ is also inflated by low dispersal; reflecting a shift in the underlying genealogy in which nearby individuals coalesce quickly but distant groups of individuals take much longer to coalesce than they would under random mating.
+As a result the number of mid-frequency alleles is inflated, which increases expected heterozygosity relative to the count of segregating sites and leads to increased Tajima's $D$.
+The pattern is similar to that produced by declining population sizes, in which the terminal branches of the genealogy are short relative to those expected in constant populations.
+Here, however, the effect is caused by geographically limited dispersal instead of sampling from a declining population.
+
+Trends in pairwise haplotype block sharing parallel those seen in SNP-based diversity estimates.
+At low dispersal the mean length of an identical-by-state block shared between two randomly sampled individuals is lower than expected under random mating.
+This is because the average coalescence time of any two individuals is higher, which leaves more time for both mutation and recombination to introduce variation along a given stretch of genome.
+
+The strongest signals of limited dispersal are in statistics that directly incorporate spatial information.
+The correlation coefficient between spatial distance and pairwise genetic differentiation ($D_{xy}$) is positive and declines as dispersal increases, as expected under the theory developed by \citep{Rousset1997} and others.
+As in \citep{Ringbauer2017} and \citep{Baharian2016} we observed that the pairwise distribution of haplotype block lengths is more strongly left-skewed under limited dispersal.
+In our summaries this is reflected in negative correlations between spatial distance and the mean, variance, skew, or count of long blocks from the pairwise distribution of identical-by-state block lengths.
+This is because individuals that are closer geographically also have more recent coalescence times when dispersal is limited, which results in fewer mutations and recombination events that can break up long shared haplotypes.
+Note, however, that the mean of the IBS tract length distribution has only weak signal for $\sigma$ because it is a very noisy summary of the data.
+The count of long haplotype blocks and the skew of the pairwise distribution (which reflects the relative length of the right tail of this distribution) appear to contain much stronger signal.
 
@@
 The deflation in $p$ values for spatial phenotypes is particularly pronounced for the corner and clinal distributions, likely because in these cases our sample captures more phenotypic variation than in the "patchy" distribution.
-When PCA positions are included as covariates the vast majority of SNPs no longer pass a 5\% FDR significance threshold, but up to 1000 genome-wide significant hits (out of an average of 150,000 SNPs) were observed even in relatively high dispersal for the corner and clinal distributions.
+When PCA positions are included as covariates the vast majority of SNPs no longer pass a 5\% FDR significance threshold, but up to 1000 genome-wide significant hits (out of an average of 150,000 SNPs) were observed even in relatively high dispersal simulations for the corner and clinal distributions.
 
@@
 Thus the assumption that patterns of linkage disequilibrium are uncorrelated with allele frequency differentiation, which is the basis of LD score regression \citep{Bulik-Sullivan2015}, may not hold for these simulations.
-###TODO revisit this and
+TODO revisit this with more quantitative analysis (need to keep genotypes to do LD scores?)
 
@@
 Essentially, PCA loses descriptive power faster than the breakdown in correlations between allele frequencies and spatial position that our GWAS is picking up on, so eventually the statistical correction breaks down.
-
 Examining quantile-quantile plots shows that $p$ value deflation is most pronounced for the corner phenotype distribution and is particularly strong for simulations with neighborhood sizes of less than 20.
 These results are similar to the observation of high inflation for spatially discrete phenotypes in \citep{Mathieson2012}.
-Notably here we find that the deflation in $p$ values is not limited to low-frequency alleles, and is not entirely controlled by including PCA covariates.
 Quantile-quantile plots of clinal and nonspatial phenotypes also show evidence of $p$ value $inflation$ (i.e.
-points below the 1:1 line), which is observed only when PCA covariates are included in the analysis (see Supplementary Figure X for q-q plots of uncorrected GWAS results).
+points below the 1:1 line), which is observed only when PCA covariates are included in the analysis (see Supplementary Figure X for plots of uncorrected GWAS results).
 This likely reflects the loss of power in GWAS when both phenotype and ancestry are correlated with space, which creates a partially confounded study design.
@@
 \section{Discussion}
+
+\subsection{GWAS}
+
+% As in \citep{Mathieson2012} we observed that the deflation in $p$ values after PCA correction is largest for spatially discrete phenotypes, and is weaker for phenotypes that are clinal over space.
+When dispersal is low and isolation by distance is strong this effect is pronounced enough to create statistically significant associations at nearly every variable site in the genome unless statistical corrections for population structure are implemented.
+For strongly structure populations and phenotypes that vary systematically over space, the estimated effect sizes and biological interpretations of linear-regression GWAS are therefore controlled primarily by the method of stratification control rather than underlying causal associations between phenotype and genotype.
+
+%This is unlikely to be a significant issue for studies seeking to identify a small number of causal alleles as long as there is variation within populations at those sites.
+However analyses of putatively polygenic phenotypes may be badly affected both by deflation of $p$ values from uncontrolled stratification and by inflation of $p$ values caused by the use of PCA covariates.
+Though many studies now employ mixed-model methods \citep{Kang2008,Yu2006,Kang2010} which are likely more robust to stratification (though not immune, see \citep{Berg2018}), using principal components regression as the method of stratification control is still common in analyses of human genetics [quality time w nature genetics].
 
@@
 
-\begin{figure*}[h!]
+\begin{figure*}[p]
 \centering
-\includegraphics[width=\textwidth]{sfs_spatial_v_rm.pdf}
-\caption{Scaled allele frequency distributions for random mating and spatial SLiM models under varying $\sigma$.
-The rightmost panel is the difference between the spatial and random-mating distributions, showing an inflation of mid-frequency sites for neighborhood sizes of less than 50.
-Lines are a local regression (loess) model at span 2.}
+\includegraphics[width=\textwidth]{min_ancestor_age_maps.pdf}
+\caption{[placeholder - working on this] Geographic spread of genealogical ancestors from a random individual over 50 generations, by Neighborhood size and $\sigma$.
+Colors are scaled by the age of the most recent ancestor.}
 \label{fig:spectrum}
@@
 
-\begin{figure*}[h!]
+\begin{figure*}[p]
 \centering
-\includegraphics[width=\textwidth]{min_ancestor_age_maps.pdf}
-\caption{Geographic spread of genealogical ancestors from a random individual over 50 generations, by Neighborhood size and $\sigma$. Colors are scaled by the age of the most recent ancestor.}
+\includegraphics[width=\textwidth]{sumstats_by_sigma.pdf}
+\caption{Summary statistics from spatial and random-mating SLiM simulations. Statistics are calculated from the genotype matrix of 60 randomly sampled individuals in each simulation.}
 \label{fig:spectrum}
@@
 
-\begin{figure*}[h!]
+\afterpage{\clearpage}
+\begin{figure*}[p]
+\centering
+\includegraphics[width=\textwidth]{stairwayplot_by_sigma_facets.pdf}
+\caption{Inferred demographic histories for spatial SLiM simulations from Stairwayplot.
+Lines shown are the median inferred $N_{e}$ across 100 bootstrap replicates for each simulation.
+Colors are scaled to Wright's neighborhood size (NS) and panels are divided by neighborhood size range.
+The black line is the census population size.}
+\label{fig:spectrum}
+\end{figure*}
+
+\afterpage{\clearpage}
+\begin{figure*}[p]
 \centering
 \includegraphics[width=\textwidth]{gwas_summary_nsig_qqplots.pdf}
-\caption{Impacts of spatially correlated phenotypes and isolation by distance on GWAS.
+\caption{Impacts of spatially correlated phenotypes and isolation by distance on linear regression GWAS.
 Colors represent different phenotype distributions as indicated in the figure legend.
-A: numbers of significant SNPs after FDR correction for linear-model GWAS conducted with (top) or without (bottom) PC covariates.
-B: Quantile-quantile plots of observed versus $p$ values relative to those expected from a uniform distribution assuming no effect on phenotypes.
+A: example phenotype distributions and sampling density.
+B: proportion of total variance explained by the first 10 PC axes, by neighborhood size.
+C: numbers of significant SNPs after FDR correction for linear-model GWAS conducted with (top) or without (bottom) PC covariates.
+Note the log scale.
+D: Quantile-quantile plots of observed versus $p$ values relative to those expected from a uniform distribution for neighborhood sizes under 65 ($\sigma$=1).
 The dotted lines show the 95\% confidence region assuming binomial sampling and an alpha level of 0.05.
@@
 
-\newpage
+\afterpage{\clearpage}
 

### 2019-03-15 ms/spaceness.tex
new gwas fig w pca var

@@
 %Impacts of Continuous Spatial Structure on Analyses of Population Genetic Data
-\title{Space is the Place: Impacts of Continuous Spatial Structure on Demographic Modeling and Association Studies}
+\title{Impacts of Isolation by Distance on Demographic Modeling and Association Studies}
 
@@
 In association analyses spatially correlated environments cause genome-wide signals of genetic association with purely environmental phenotypes, and this bias is only partially corrected by regressing out principal components positions during the analysis.
-We demonstrate that stratification bias in continuous populations is worst when environmental effects are spatially discrete, but that corrections for clinal phenotypes can also fail under high dispersal because PCA describes less total variation in the data when isolation by distance is weak.
+We find that association analyses in continuous populations are most biased when environmental effects are spatially discrete, but that corrections for clinal phenotypes can also fail under high dispersal because PCA describes less total variation in the data when isolation by distance is weak.
 
@@
 \centering
-\includegraphics[width=\textwidth]{gwas_summary_nsig_qqplots.pdf}
+\includegraphics[width=\textwidth]{gwas_summary_nsig_qqplots_logX.pdf}
 \caption{Impacts of spatially correlated phenotypes and isolation by distance on linear regression GWAS.
@@
 C: numbers of significant SNPs after FDR correction for linear-model GWAS conducted with (top) or without (bottom) PC covariates.
-Note the log scale.
 D: Quantile-quantile plots of observed versus $p$ values relative to those expected from a uniform distribution for neighborhood sizes under 65 ($\sigma$=1).

### 2019-03-15 ms/spaceness.tex
adding sampling to ms

@@
 Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations exchanging migrants.
-As the availability of population-level, whole-genome data allows inference of increasingly fine-scale patterns of ancestry in many species, models incorporating realistic demographic and spatial processes are needed to accurately describe spatial structure and control for its impacts on analyses of selection and demography.
+As the availability of population-level whole genome data allows inference of increasingly fine-scale patterns of ancestry in many species, models incorporating realistic demographic and spatial processes are needed to accurately describe spatial structure and control for its impacts on analyses of selection and demography.
 Here we implement a forward-time simulation of evolution in continuous space and use it to study the impacts of dispersal and density-dependent competition on population genetic summary statistics, demographic inference, and association studies.
@@
 As a result demographic inference from the site frequency spectrum is biased towards inferring a turbulent past demographic history and a recent decline in populations when neighborhood sizes are below 20.
-In association analyses spatially correlated environments cause genome-wide signals of genetic association with purely environmental phenotypes, and this bias is only partially corrected by regressing out principal components positions during the analysis.
-We find that association analyses in continuous populations are most biased when environmental effects are spatially discrete, but that corrections for clinal phenotypes can also fail under high dispersal because PCA describes less total variation in the data when isolation by distance is weak.
-
+For many common population genetic summary statistics the variation associated with different sampling strategies is larger than that observed across the range of dispersal distances we simulated, pointing to the need to carefully account for the spatial distribution of sampling when interpreting results.
+Last we show that spatially correlated environments cause genome-wide signals of association with purely environmental phenotypes in linear-regression GWAS, and this bias is not fully corrected by regressing out principal components positions during the analysis.
 \end{abstract}
@@
 As the availability of population-level whole-genome data allows inference of increasingly fine-scale patterns of ancestry in many species, models incorporating biologically realistic demographic and spatial processes are needed to accurately describe continuous spatial structure and control for its impacts on analyses of selection and demography.
-This situation is particularly clear in human population genetics, where individual ancestries are often interpreted as mixtures of discrete populations in part because of modeling assumptions underlying discrete clustering algorithms and association analyses are often strongly dependent on statistical corrections for population stratification \citep{Berg2018}.
+This situation is particularly clear in human population genetics, where individual ancestries are often interpreted as mixtures of discrete populations in part because of modeling assumptions underlying discrete clustering algorithms and association analyses are often strongly dependent on statistical corrections for population stratification \citep{Berg2018,Bulik-Sullivan2015,Young2018,Heckerman2016,Kang2010,Sohail2018}.
 
-The best-studied approaches to population genetics in continuous space were developed by \citep{Wright1943} and \citep{Malecot1948}, who derived expressions for pairwise genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals.
-However, \citep{Felsenstein1975} showed that these assumptions are incompatible -- over time, a population meeting them will clump into a small number of geographic clusters occupying only a part of the available range.
-(NOTE: where to put Rousset et al.
-lit?)
+The best-studied approaches to population genetics in continuous space were developed by \citep{Wright1943} and \citep{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals.
+A key finding of Wright's model is that many important aspects of continuous populations can be described in terms of "neighborhood size" -- the number of potential mates for an individual in a given generation, defined as $4\pi\sigma^2d$, where $\sigma$ is the average dispersal distance and $d$ is population density.
+\cite{Maruyama1972} found that the rate of decline in genetic diversity in a 2-dimensional continuous population shifts between distinct regimes when $D\sigma^2$ is greater or less than 1, and proposed that this had the important implication that most population genetic expectations for randmoly mating populations could be applied to continuously distributed populations with relatively little error.
+
+Though some aspects of continuous populations are fairly well described by the Wright-Malécot models, \cite{Felsenstein1975} showed that the assumptions of independent dispersal and Poisson distributed offspring that are the basis of these models are incompatible.
+Over time, a population meeting them will clump into a small number of geographic clusters occupying only a part of the available range.
+Although real populations are often clumped on landscapes due to factors like varying habitat quality and competition among species, the Wright and Malécot models produce much more extreme clumping than is observed in real organisms and fail to account for the density-dependent declines in population growth rate that are widely observed in real populations (CITE).
 
 One method for modeling populations in continuous space is then to assume the existence of a grid of discrete randomly-mating populations connected by migration, which prevents clustering by forcing all regions to be occupied in every generation.
-Though good approximations of continuous structure given high dispersal (Barton(?)), these models are not truly continuous and limit investigation of spatial structure below the level of the deme.
-An alternative approach is to model the geographic spread of ancestry backwards in time through a diffusion approximation -- an approach that has recently made significant progress in modeling both dispersal and demographic parameters \citep{Barton2010,Kelleher2014,Ringbauer2017,Ringbauer2018}.
-However, these models currently lack a well-described forward-time analog that produces biologically realistic spatial patterns.
+Among many other important results drawn from this class of "lattice" or "stepping stone" models, \cite{Rousset1997} showed that the slope of the a linear regression of genetic differentiation ($F_{st}$) against the logarithm of spatial distance is an estimate of neighborhood size.
+Though good approximations of continuous structure given high dispersal (Barton(?)), these models are not truly continuous, force a uniform realized population density across landscapes, and limit investigation of spatial structure below the level of the deme.
+An alternative method is to model the geographic spread of ancestry backwards in time through a diffusion approximation -- an approach that has recently made significant progress in modeling both dispersal and demographic parameters \citep{Barton2010,Kelleher2014,Ringbauer2017,Ringbauer2018}.
+However these models currently lack a well-described forward-time analog that produces biologically realistic spatial patterns and are subject to the same difficulty in incorporating selection and other biological processes as all other reverse time methods (CITE??).
 
-A direct approach to the clustering problem of classical forward-time models is to incorporate density dependence by scaling individual fitness to local population density.
-This shifts reproductive output towards regions of low-density, preventing populations from clustering in space over time.
-This approach was taken previously by \citep{Doebeli2003} who used an individual based model with continuous space and density dependent fitness to study the probability of speciation along continuous environmental gradients.
-(**Sentence referencing other continuous IBM's with density dependence**).
-However to our knowledge all previous implementations of continuous space models have focused on individuals or at most a few genotypes as the unit of analysis, which limits our ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms.
+We took a direct approach to the clustering problem of classical forward-time models by incorporating density dependence into the model.
+By scaling the probability of survival in each timestep to local population density we shift reproductive output towards regions of low-density and prevent populations from clustering.
+A similar approach was taken previously by \citep{Doebeli2003} who used an individual based model with continuous space and density dependent fitness to study the probability of speciation along continuous environmental gradients.
+However to our knowledge all previous implementations of continuous space models have focused on a small number of genetic loci as the unit of analysis, which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms.
 
-Here we describe an implementation of an individual-based model in continuous space that incorporates overlapping generations, a Gaussian dispersal kernel, and density-dependent fitness, and scales to chromosome-scale alignments across tens of thousands of individuals.
-We use this model to describe the impacts of variation in dispersal distance on a variety of summary statistics typically analyzed under discrete population models, and apply a machine-learning regression approach to evaluate a new class of summary statistics based on correlations between geographic distance and moments of the distribution of pairwise haplotype block lengths as estimators of dispersal distance.
-We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum; finding that inflation of branch lengths for mid-frequency alleles associated with slow coalescence of distant individuals **TBD**.
+Here we describe an implementation of an individual-based model in continuous space that incorporates overlapping generations, a Gaussian dispersal kernel, and density-dependent fitness.
+The model scales to chromosome-scale alignments across tens of thousands of individuals, and outputs the full genealogy and recombination history of all final-generation individuals.
+We use this simulation to describe how sampling strategy interacts with isolation by distance to cause systematic variation in population genetic summary statistics typically analyzed under discrete population models.
+We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum; finding that inflation of branch lengths for mid-frequency alleles associated with slow coalescence of distant individuals can bias analyses towards inferring turbulent demographic histories and recent declines in population size.
 
 Last, we use our model to study the impacts of isolation by distance in continuous space on genome-wide association studies (GWAS).
-Variation in ancestry proportions between case and control cohorts is expected to cause inflation of test statistics at SNPs with different allele frequencies among populations (CITE).
-In addition, because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within and among populations can lead to inflated effect size estimates in GWAS of quantitative traits conducted on samples from large geographic regions, and this effect is particularly prominent in low-frequency alleles \citep{Mathieson2012}.
-For example, imagine a population of flowering plants occurring over a large latitudinal range with populations subject to isolation by distance.
-Flowering time is strongly dependent on photoperiod and temperature, both of which typically vary with latitude.
-If we were to conduct a GWAS for flowering time without correcting for population structure we would find that all SNPs with allele frequencies varying clinally with latitude were significantly associated with flowering time, even if flowering time was determined purely by the environment.
+Variation in ancestry proportions between case and control cohorts is expected to cause inflation of test statistics at SNPs with different allele frequencies among populations \citep{Price2006}.
+Because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within continuous populations can also deflate $p$ values in GWAS of quantitative traits \citep{Mathieson2012}.
+These issues are particularly prominent for polygenic traits, because the deflated $p$ values expected from polygenic phenotype correlations are superficially similar to those expected from population structure \citep{Bulik-Sullivan2015,Young2018}.
+However most studies examining the effects of stratification on GWAS have focused on simulating phenotypes for real genotype data \citep{Price2006,Bulik-Sullivan2015} or generating simulated genotypes by sampling from static allele frequency distributions \citep{Bulik-Sullivan2015,Heckerman2016}.
+These approaches address application to human data but are not based on a true generative model and so provide limited insight into the processes generating stratification bias in association analyses.
+Here we simulate both the underlying genealogy and phenotypes of a continuously distributed population and seek to identify regions of parameter space -- i.e.
+the strength of isolation by distance and the spatial distribution of environmental effects on phenotypes -- in which common methods of stratification bias corrections in GWAS are subject to bias.
 
-A common method of correcting for population structure in GWAS is to run a principal components analysis on the genotype matrix and include PC coordinates as covariates in a linear regression for each SNP \citep{Price2006}.
+%[move some of this to methods, rewrite to avoid repitition w above]  A common method of correcting for population structure in GWAS is to run a principal components analysis on the genotype matrix and include PC coordinates as covariates in a linear regression for each SNP \citep{Price2006}.
 Mixed model GWAS employs a different statistical framework but takes a similar approach by incorporating a kinship matrix describing pairwise genetic relatedness among samples as a random effect in a linear mixed model \citep{Yu2006,Kang2008}.
@@
 For traits controlled by a few large effect loci this confounding effect is likely to be relatively small, but for analyses of complex traits with thousands of putatively associated SNPs the potential for confounding is high.
-Here we simulate a range of spatially correlated phenotypes for simulated individuals and seek to identify conditions under which existing GWAS structure corrections fail.
+Here we simulate a range of spatially correlated phenotypes for simulated individuals and seek to identify conditions under which existing GWAS structure corrections are subject to bias caused by continuous population structure.
+
 \section{Materials and Methods}
@@
 
-To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, or density-dependent fitness, we  also implemented a second version of our simulation in which mates are selected at random and offspring disperse to a random location on the landscape.
-For all simulations the full history of the genomes of final-generation individuals were stored as tree sequences \citep{Kelleher2018}.
+To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, and density-dependent fitness, we  also implemented a second version of the SLiM simulation in which mates are selected at random and offspring disperse to a random location on the landscape.
+For all simulations the full genealogy and recombination history of final-generation individuals were stored as tree sequences \citep{Kelleher2018}.
 
@@
 We then verified that this procedure produced the correct number of mutations by comparing a subset of simulations with SLiM-generated mutations (which are applied only at mating events) with mutations added by msprime (Supplementary Figure 1).
-Simulations were run until all extant individuals shared a common ancestor within the simulation (i.e.
+Simulations were run for 1.6 million timesteps (approximately 30N generations), or until all extant individuals shared a common ancestor within the simulation (i.e.
 the tree sequence had coalesced).
@@
 We set the total $N_{e}$ equal to the census size of each SLiM simulation divided by the variance in reproductive output (see below).
+
+\subsection{Sampling}
+Unlike reverse time approaches our model records the genealogy and sequence variation of the complete population, but in real populations genotypes are only observed from a subset of individuals.
+We modeled three sampling strategies meant to approximate common data collection methods in empirical genetic studies (Figure 1).
+"Random" sampling selects individuals at random from across the full landscape.
+"Point" sampling selects individuals proportional to their distance from four equally spaced points on the landscape.
+Last, "midpoint" sampling selects individuals in proportion to their distance from the middle of the landscape.
+In practice random sampling is rarely conducted because of the difficulty of accessing all parts of a species range and the inefficiency of collecting only single individuals at all sites.
+Point and midpoint sampling strategies are both common, particularly for species that require traps or collecting stations to be set up and dismantled at every site.
 
@@
 \subsection{Demographic Modeling}
-We fit single-population demographic models to the site frequency spectra of 20 randomly sampled individuals from each spatial SLiM simulation with the program Stairwayplot \citep{Liu2015}.
+We fit single-population demographic models to the site frequency spectra of 20 individuals from each spatial SLiM simulation with the program Stairwayplot \citep{Liu2015}.
+This analysis was replicated across random, point, and midpoint sampling strategies.
 Site frequency spectra used for input data were calculated in scikit-allel \citep{Miles2017}.
@@
 \subsection{Summary Statistics}
+\subsubsection{Diversity}
 As expected from theory based on the Wright-Malécot model \citep{Maruyama1972}, limited dispersal increases the coalescence time of a set of randomly-sampled individuals relative to a random-mating model.
@@
 Here, however, the effect is caused by geographically limited dispersal instead of sampling from a declining population.
+
+\subsubection{Site Frequency Spectrum}
+
+\subsubsection{Haplotype Block Lengths}
+
+\subsubsection{Spatial Correlations}
 
@@
 \centering
-\includegraphics[width=\textwidth]{sumstats_by_sigma.pdf}
-\caption{Summary statistics from spatial and random-mating SLiM simulations. Statistics are calculated from the genotype matrix of 60 randomly sampled individuals in each simulation.}
+\includegraphics[width=\textwidth]{sumstats_by_neighbors.pdf}
+\caption{[needs some polishing for labels etc] Summary statistics from spatial and random-mating SLiM simulations.
+Statistics are calculated from the genotype matrix of 60 randomly sampled individuals in each simulation.}
 \label{fig:spectrum}
@@
 \centering
-\includegraphics[width=\textwidth]{gwas_summary_nsig_qqplots_logX.pdf}
+\includegraphics[width=\textwidth]{gwas_summary_nsig_qqplots_loglog.pdf}
 \caption{Impacts of spatially correlated phenotypes and isolation by distance on linear regression GWAS.
 Colors represent different phenotype distributions as indicated in the figure legend.
-A: example phenotype distributions and sampling density.
+A: example phenotype distributions and sampling maps, with colors scaled to phenotype values.
 B: proportion of total variance explained by the first 10 PC axes, by neighborhood size.
 C: numbers of significant SNPs after FDR correction for linear-model GWAS conducted with (top) or without (bottom) PC covariates.
-D: Quantile-quantile plots of observed versus $p$ values relative to those expected from a uniform distribution for neighborhood sizes under 65 ($\sigma$=1).
+D: Quantile-quantile plots of observed versus $p$ values relative to those expected from a uniform distribution.
 The dotted lines show the 95\% confidence region assuming binomial sampling and an alpha level of 0.05.

### 2019-03-19 ms/spaceness.tex
added pop params fig

@@
 %\documentclass{article}
+\newcommand{\beginsupplement}{%
+        \setcounter{table}{0}
+        \renewcommand{\thetable}{S\arabic{table}}%
+        \setcounter{figure}{0}
+        \renewcommand{\thefigure}{S\arabic{figure}}%
+     }
+\newcommand{\stopsupplement}{%
+        \setcounter{table}{0}
+        \renewcommand{\thetable}{\arabic{table}}%
+        \setcounter{figure}{0}
+        \renewcommand{\thefigure}{\arabic{figure}}%
+     }
 
@@
 
+% adding commenting commands
+\newif\ifcomments
+\commentstrue
+%\commentsfalse
+\newcommand{\cjb}[1]{\ifcomments{{\color{blue} \it (#1)}}\else{}\fi}
+\newcommand{\plr}[1]{\ifcomments{{\color{purple} \it (#1)}}\else{}\fi}
+\newcommand{\ak}[1]{\ifcomments{{\color{red} \it (#1)}}\else{}\fi}
+
 %Simulating Isolation by Distance in Continuous Space
@@
 %Impacts of Continuous Spatial Structure on Analyses of Population Genetic Data
-\title{Impacts of Isolation by Distance on Demographic Modeling and Association Studies}
+%Impacts of Isolation by Distance on Demographic Modeling and Association Studies
+%Impacts of Dispersal and Sampling on Estimates of Genetic Variation in Continuous Space
+%Space is the Place: How Dispersal and Sampling Shape Estimates of Genetic Variation in Continuous Space
+\title{Space is the Place: How Dispersal and Sampling Shape Analyses of Genetic Variation in Continuous Space}
 
@@
 Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations exchanging migrants.
-As the availability of population-level whole genome data allows inference of increasingly fine-scale patterns of ancestry in many species, models incorporating realistic demographic and spatial processes are needed to accurately describe spatial structure and control for its impacts on analyses of selection and demography.
-Here we implement a forward-time simulation of evolution in continuous space and use it to study the impacts of dispersal and density-dependent competition on population genetic summary statistics, demographic inference, and association studies.
+Here we implement a forward-time simulation of evolution in continuous space and use it to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and association studies.
 Low dispersal slows the geographic spread of ancestry and increases branch lengths corresponding to mid-frequency alleles because of slow coalescence among distant groups of individuals.
@@
 For many common population genetic summary statistics the variation associated with different sampling strategies is larger than that observed across the range of dispersal distances we simulated, pointing to the need to carefully account for the spatial distribution of sampling when interpreting results.
-Last we show that spatially correlated environments cause genome-wide signals of association with purely environmental phenotypes in linear-regression GWAS, and this bias is not fully corrected by regressing out principal components positions during the analysis.
+Last we show that spatially correlated environments cause genome-wide signals of association with purely environmental phenotypes in linear-regression GWAS, and that this bias is not fully controlled by regressing out principal components positions during the analysis.
 \end{abstract}
@@
 The inescapable reality that biological organisms live, move, and reproduce in continuous spatial landscapes has been all but ignored in population genetic models.
-
-In most sexual organisms individuals mate with other nearby individuals, leading to a positive correlation between genetic and geographic distances.
-This pattern of "isolation by distance" \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, but is only heuristically approximated by existing forward-time models of molecular evolution based on discrete demes exchanging migrants.
-As the availability of population-level whole-genome data allows inference of increasingly fine-scale patterns of ancestry in many species, models incorporating biologically realistic demographic and spatial processes are needed to accurately describe continuous spatial structure and control for its impacts on analyses of selection and demography.
+Indeed a near universal rule of reproduction is that individuals mate with other nearby individuals, leading to a positive correlation between genetic and geographic distances.
+This pattern of "isolation by distance" \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, but our mathematical models to describe it are poor approximations of the underlying process.
+As the availability of population-level, whole-genome data allows inference of increasingly fine-scale patterns of ancestry in many species, models incorporating biologically realistic demographic and spatial processes are needed to accurately describe continuous spatial structure and control for its impacts on analyses of selection and demography.
 This situation is particularly clear in human population genetics, where individual ancestries are often interpreted as mixtures of discrete populations in part because of modeling assumptions underlying discrete clustering algorithms and association analyses are often strongly dependent on statistical corrections for population stratification \citep{Berg2018,Bulik-Sullivan2015,Young2018,Heckerman2016,Kang2010,Sohail2018}.
@@
 A key finding of Wright's model is that many important aspects of continuous populations can be described in terms of "neighborhood size" -- the number of potential mates for an individual in a given generation, defined as $4\pi\sigma^2d$, where $\sigma$ is the average dispersal distance and $d$ is population density.
-\cite{Maruyama1972} found that the rate of decline in genetic diversity in a 2-dimensional continuous population shifts between distinct regimes when $D\sigma^2$ is greater or less than 1, and proposed that this had the important implication that most population genetic expectations for randmoly mating populations could be applied to continuously distributed populations with relatively little error.
+\cite{Maruyama1972} found that the rate of decline in genetic diversity in a 2-dimensional continuous population approaches the random mating expectation when $D\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.
 
@@
 Over time, a population meeting them will clump into a small number of geographic clusters occupying only a part of the available range.
-Although real populations are often clumped on landscapes due to factors like varying habitat quality and competition among species, the Wright and Malécot models produce much more extreme clumping than is observed in real organisms and fail to account for the density-dependent declines in population growth rate that are widely observed in real populations (CITE).
+Although real populations are often clumped on landscapes due to factors like varying habitat quality and competition among species, the Wright and Malécot models produce much more extreme clumping than is observed in practice and fail to account for the density-dependent declines in population growth rate that are widely observed in real populations (CITE).
 
-One method for modeling populations in continuous space is then to assume the existence of a grid of discrete randomly-mating populations connected by migration, which prevents clustering by forcing all regions to be occupied in every generation.
+One method for modeling continuous populations is then to assume the existence of a grid of discrete randomly-mating populations connected by migration, which prevents clustering by forcing all regions to be occupied in every generation.
 Among many other important results drawn from this class of "lattice" or "stepping stone" models, \cite{Rousset1997} showed that the slope of the a linear regression of genetic differentiation ($F_{st}$) against the logarithm of spatial distance is an estimate of neighborhood size.
@@
 
+\ak{Andy-- I think we could move some of the text from the following paragraph to methods}
 We took a direct approach to the clustering problem of classical forward-time models by incorporating density dependence into the model.
@@
 
-\subsection{Sampling}
-Unlike reverse time approaches our model records the genealogy and sequence variation of the complete population, but in real populations genotypes are only observed from a subset of individuals.
-We modeled three sampling strategies meant to approximate common data collection methods in empirical genetic studies (Figure 1).
-"Random" sampling selects individuals at random from across the full landscape.
-"Point" sampling selects individuals proportional to their distance from four equally spaced points on the landscape.
-Last, "midpoint" sampling selects individuals in proportion to their distance from the middle of the landscape.
-In practice random sampling is rarely conducted because of the difficulty of accessing all parts of a species range and the inefficiency of collecting only single individuals at all sites.
-Point and midpoint sampling strategies are both common, particularly for species that require traps or collecting stations to be set up and dismantled at every site.
-
 \subsection{Genealogical Parameters}
@@
 
+\subsection{Sampling}
+Our model records the genealogy and sequence variation of the complete population, but in real data genotypes are only observed from a a relatively small number of sampled individuals.
+We modeled three sampling strategies similar to common data collection methods in empirical genetic studies (Figure 1).
+"Random" sampling selects individuals at random from across the full landscape, "point" sampling selects individuals proportional to their distance from four equally spaced points on the landscape, and "midpoint" sampling selects individuals in proportion to their distance from the middle of the landscape.
+Downtream analyses were repeated across all sampling strategies.
+
 \subsection{Summary Statistics}
-We calculated a set of 20 summary statistics (Supplementary Table 1) from 60 diploid individuals sampled randomly from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}.
-Statistics included common single-population summaries such as mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{is}$), and Tajima's D, as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distances \citep{Rousset1997}, which we summarized as the correlation coefficient between $log10(spatial distance)$ and the proportion of identical base pairs among individuals.
+We calculated the site frequency spectrum and a set of 18 summary statistics (Supplementary Table 1) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}.
+Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{is}$), and Tajima's D, as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distances \citep{Rousset1997}, which we summarized as the correlation coefficient between $log10(spatial distance)$ and the proportion of identical base pairs among individuals.
 
 Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among samples.
-Intuitively, haplotype block length distributions should contain more information about the genealogy of a set of samples than allele frequency measures alone because they reflect the influence of two Poisson processes (mutation and recombination) acting along its branches, rather than just one.
-
 The full distribution of lengths of IBS tracts for each pair of individuals was first calculated with a custom python function.
 We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $1e6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.
-To assess the degree to which the IBS tract length distribution shifts with spatial distance, we then estimated correlation coefficients between spatial distance and each moment of the pairwise IBS tract distribution.
+
+We then estimated correlation coefficients between spatial distance and each moment of the pairwise IBS tract distribution.
 Because more closely related individuals on average share longer haplotype blocks we expect that spatial distance will be negatively correlated with mean haplotype block length, and that this correlation will be strongest (i.e.
 most negative) when dispersal is low.
-The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks and is thus most informative of recent demographic events \citep{Ringbauer2017} (cite Coop \& Ralph here \& above).
+The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks so should reflect recent demographic events \citep{Ringbauer2017}.
+
+The effects of sampling on summary statistic estimates were summarized by testing for differences in mean (ANOVA, \citep{Rcore2018}) and variance (Levene's test, \citep{Fox2011}) across sampling strategies for each summary statistic.
 
@@
 This analysis was replicated across random, point, and midpoint sampling strategies.
-Site frequency spectra used for input data were calculated in scikit-allel \citep{Miles2017}.
-100 bootstrap replicates were generated for each simulation by resampling over sites.
-Models were then fit across all bootstrap replicates using default settings in Stairwayplot and the median estimate of $Ne$ per generation was used to represent the output of each simulation.
-The resulting inferred population size histories were plotted with line colors scaled to $\sigma$ to visualize how dispersal scaling impacts estimates of demographic history.
+Site frequency spectra used for input data were calculated in scikit-allel \citep{Miles2017}, and 100 bootstrap replicates were generated for each simulation by resampling over sites.
+Models were fit across all bootstrap replicates using default settings in Stairwayplot and the median estimate of $Ne$ per generation was used to represent the output of each simulation.
 
@@
 We ran linear-model GWAS both with and without the first 10 principal components as covariates in PLINK and summarized results across simulations by counting the number of significant SNPs with an expected false positive rate of less than 5\% after adjusting p values with the R function p.adjust(...,method="fdr").
-We also examined $p$ values for systemic inflation by estimating the expected values from a uniform distribution (because no SNPs were used when generating phenotypes), plotting observed against expected values for all simulations, and using the two-dimensional binning feature of ggplot2 \citep{Wickham2016} to find the mean $\sigma$ value in each region of quantile-quantile space.
+We also examined $p$ values for systemic inflation by estimating the expected values from a uniform distribution (because no SNPs were used when generating phenotypes), plotting observed against expected values for all simulations, and summarizing across simulations by finding the mean $\sigma$ value in each region of quantile-quantile space.
+Results from all analyses were summarized and plotted with the 'ggplot2' \citep{Wickham2016} and "cowplot" \citep{Wilke2019} packages in R \citep{Rcore2018}.
 
@@
 \subsection{Genealogical Parameters and Run Times}
+\subsection{Impacts of Sampling on Summary Statistics}
+For spatial models the mean and variance of all summary statistics other than those involving the mean of the IBS block size distribution was significantly different across sampling strategies (Supplementary Table 1).
+Means were not significantly different across sampling strategies for any statistic under the random mating model.
+Only spatial correlation coefficients had significantly different variance under the random mating model, though in all cases the mean correlation was near zero.
 
-\subsection{Summary Statistics}
-\subsubsection{Diversity}
-As expected from theory based on the Wright-Malécot model \citep{Maruyama1972}, limited dispersal increases the coalescence time of a set of randomly-sampled individuals relative to a random-mating model.
-This is reflected in sequence data as a larger number of segregating sites and inflated estimates of expected heterozygosity ($\pi$, Figure 2).
-In practice the difference in segregating sites and expected heterozygosity in our simulations is negligible above neighborhood sizes of 65 ($\sigma$=1).
-Conversely observed heterozygosity (the fraction of heterozygous individuals) is lower than the random mating expectation when dispersal is low, and as a consequence the inbreeding coefficient ($F_{is}$) is positive.
-This is the continuous space version of the Wahlund effect \citep{Wahlund1928} - a deficit of heterozygous individuals when samples are drawn from subdivided populations.
-Similar observations of positive $F_{is}$ have been observed from stepping-stone simulations in \citep{Shirk2014} and \citep{Neel2013}.
+\subsubsection{Diversity and Site Frequency Spectra}
+Mean within-population genetic distance ($\pi$), the inbreeding coefficient ($F_{is}$) and Tajima's $D$ are inflated at low dispersal distances (Figure 2).
+$\pi$ and $F_{is}$ approach the random mating model at neighborhood sizes of roughly 100.
+Tajima's D varies relatively more across sampling strategies; approaching random mating at $NS\approx65$ with random sampling and $NS>1000$ with midpoint sampling.
 
-Tajima's $D$ is also inflated by low dispersal; reflecting a shift in the underlying genealogy in which nearby individuals coalesce quickly but distant groups of individuals take much longer to coalesce than they would under random mating.
-As a result the number of mid-frequency alleles is inflated, which increases expected heterozygosity relative to the count of segregating sites and leads to increased Tajima's $D$.
-The pattern is similar to that produced by declining population sizes, in which the terminal branches of the genealogy are short relative to those expected in constant populations.
-Here, however, the effect is caused by geographically limited dispersal instead of sampling from a declining population.
+The inflation in $F_{is}$ appears to be caused by a deficit of heterozygous individuals in low-dispersal simulations -- a continuous-space version of the Wahlund effect \citep{Wahlund1928}.
+However under midpoint sampling observed heterozygosity is inflated over the random mating expectation, which we confirmed is caused by a higher proportion of heterozygotes in the middle of the landscape (Supplementary Figure S2 \cjb{CJ - this fig isn't super convincing -- will try denser sampling}).
+\cite{Shirk2014} observed a similar excess of heterozygosity in the middle of the landscape when simulating under a lattice model.
 
-\subsubection{Site Frequency Spectrum}
+The site frequency spectrum showed minimal differences across sampling strategies for random mating models (Supplementary Figure S1).
+Spatial models had fewer low frequency alleles and more mid-frequency alleles than random mating models (Figure 2).
+This effect is relatively weak when sampling is random and strongest for midpoint sampling.
 
 \subsubsection{Haplotype Block Lengths}
+Trends in pairwise haplotype block sharing parallel those in SNP-based diversity estimates.
+At low dispersal the distribution of IBS block lengths in a set of samples is shifted towards smaller values -- resulting in lower means and fewer long IBS blocks.
+The variance and skew of the full IBS tract distribution vary across sampling strategies, but are similar to random-mating models under random sampling.
 
 \subsubsection{Spatial Correlations}
-
-Trends in pairwise haplotype block sharing parallel those seen in SNP-based diversity estimates.
-At low dispersal the mean length of an identical-by-state block shared between two randomly sampled individuals is lower than expected under random mating.
-This is because the average coalescence time of any two individuals is higher, which leaves more time for both mutation and recombination to introduce variation along a given stretch of genome.
-
-The strongest signals of limited dispersal are in statistics that directly incorporate spatial information.
 The correlation coefficient between spatial distance and pairwise genetic differentiation ($D_{xy}$) is positive and declines as dispersal increases, as expected under the theory developed by \citep{Rousset1997} and others.
-As in \citep{Ringbauer2017} and \citep{Baharian2016} we observed that the pairwise distribution of haplotype block lengths is more strongly left-skewed under limited dispersal.
-In our summaries this is reflected in negative correlations between spatial distance and the mean, variance, skew, or count of long blocks from the pairwise distribution of identical-by-state block lengths.
-This is because individuals that are closer geographically also have more recent coalescence times when dispersal is limited, which results in fewer mutations and recombination events that can break up long shared haplotypes.
-Note, however, that the mean of the IBS tract length distribution has only weak signal for $\sigma$ because it is a very noisy summary of the data.
-The count of long haplotype blocks and the skew of the pairwise distribution (which reflects the relative length of the right tail of this distribution) appear to contain much stronger signal.
+As in \citep{Ringbauer2017} and \citep{Baharian2016} we found that the pairwise distribution of haplotype block lengths is more strongly left-skewed under limited dispersal.
+This is reflected in negative correlation coefficients between spatial distance and the mean, variance, skew, and count of long blocks from the pairwise distribution of identical-by-state block lengths.
+Of these summaries the mean of the IBS tract length distribution has very weak sigmal for $\sigma$, likely because it is heavily influenced by the small number of very long IBS tracts.
+Though the count of long IBS blocks and the skew of the full distribution are low and similar to random mating when taken across all samples, the correlation coefficients between spatial distance and the skew or count of long IBS blocks on a pairwise basis varies predictably by neighborhood size.
+All spatial correlations approach the random mating model near neighborhood sizes of 1000 under either point or random sampling.
+Midpoint sampling causes weaker correlations in all sampling schemes.
 
 \subsection{Demographic Modeling}
+Demographic models from Stairwayplot tended to infer patterns of ancient population increases and recent declines when neighborhood sizes were low under all sampling strategies (Figure 3).
+Inflated past population sizes were common in both point and random sampling, demonstrating that the relatively minor shift in the site frequency spectrum observed with random sampling is enough to alter demographic estimates.
+Inference of severe population bottlenecks was also common across neighborhood sizes, particularly under midpoint and point sampling.
+Across all sampling schemes and neighborhood sizes around 20\% of runs inferred a sharp increase in population in the last 100 years.
+
+Estimates stabilize for all sampling schemes as neighborhood sizes approach 750, and trend towards roughly half the census size. 
+\cjb{I still need to figure out what's going on with Ne here...
+pi is $\approx$ expected value for a WF population with $N\approx N_{c}/2$, but variance Ne is more like 1/5 census N.
+I thought it was closer to 1/2 earlier because of a bug in my script where I was dropping individuals who had 0 offspring.
+Working on it...}.
 
 \subsection{GWAS}
-Spatially correlated phenotypes in populations with low dispersal distances cause uncorrected association analyses to report significant associations at most genome-wide SNPs (Figure 1).
-No inflation was observed for the "normal" phenotype distribution which is random with respect to space, demonstrating that this effect is caused by the interaction of isolation by distance and spatially correlated phenotypes.
-The deflation in $p$ values for spatial phenotypes is particularly pronounced for the corner and clinal distributions, likely because in these cases our sample captures more phenotypic variation than in the "patchy" distribution.
-When PCA positions are included as covariates the vast majority of SNPs no longer pass a 5\% FDR significance threshold, but up to 1000 genome-wide significant hits (out of an average of 150,000 SNPs) were observed even in relatively high dispersal simulations for the corner and clinal distributions.
+Spatially correlated phenotypes in populations with low neighborhood size cause uncorrected linear-regression GWAS to report significant associations at most genome-wide SNPs (Figure 1).
+When PCA positions are included as covariates the vast majority of SNPs no longer pass a 5\% FDR significance threshold, but up to 1000 genome-wide significant hits (out of an average of 132,000 SNPs) were observed even in relatively high dispersal simulations for the corner and clinal distributions.
+Corner and patchy phenotype distributions produced the largest number of significant SNPs, and patchy phenotypes in particular showed little difference in the count of significant SNPs after PC correction.
 
-Notably, many SNPs inferred as significant are in close proximity along the genome and likely reflect haplotypes that became common in one region of the landscape (Figure 1, supplementary figures X) [note: add example manhattan plots for s=0.3 and s=3 to fig, supp figs for manhattan plots across a larger range of sigma].
-Thus the assumption that patterns of linkage disequilibrium are uncorrelated with allele frequency differentiation, which is the basis of LD score regression \citep{Bulik-Sullivan2015}, may not hold for these simulations.
-TODO revisit this with more quantitative analysis (need to keep genotypes to do LD scores?)
+For clinal phenotypes PCA covariates perform well in correcting for population stratification at low dispersal distances but the correction breaks down above neighborhood sizes of $\approx250$.
+This pattern appears to reflect a loss of descriptive power in the first ten axes of the PCA, which describe less of the total variation in the data as neighborhood size increases (Figure 4).
 
-In the case of clinal phenotypes, PCA covariates perform very well in correcting for population stratification at low dispersal distances but the correction breaks down above neighborhood sizes of 250.
-This pattern appears to reflect a loss of descriptive power in the first ten axes of the PCA, which describe less of the total variation in the data as neighborhood size increases (Figure 1 PANEL TO BE ADDED).
-Essentially, PCA loses descriptive power faster than the breakdown in correlations between allele frequencies and spatial position that our GWAS is picking up on, so eventually the statistical correction breaks down.
-Examining quantile-quantile plots shows that $p$ value deflation is most pronounced for the corner phenotype distribution and is particularly strong for simulations with neighborhood sizes of less than 20.
-These results are similar to the observation of high inflation for spatially discrete phenotypes in \citep{Mathieson2012}.
+Quantile-quantile plots show that $p$ value deflation is most pronounced for the corner phenotype distribution and is particularly strong for simulations with neighborhood sizes of less than 20.
 Quantile-quantile plots of clinal and nonspatial phenotypes also show evidence of $p$ value $inflation$ (i.e.
 points below the 1:1 line), which is observed only when PCA covariates are included in the analysis (see Supplementary Figure X for plots of uncorrected GWAS results).
-This likely reflects the loss of power in GWAS when both phenotype and ancestry are correlated with space, which creates a partially confounded study design.
 
@@
 \section{Acknowledgements}
+\newpage
 
-\begin{figure*}[p]
+\begin{figure}[htbp]
+\includegraphics{sampling_maps.pdf}
+\caption{Example sampling maps for 60 individuals on a 50x50 landscape under midpoint, point, and random sampling strategies.}
+\label{fig:spectrum}
+\end{figure}
+
+\begin{figure*}[htbp]
 \centering
-\includegraphics[width=\textwidth]{min_ancestor_age_maps.pdf}
-\caption{[placeholder - working on this] Geographic spread of genealogical ancestors from a random individual over 50 generations, by Neighborhood size and $\sigma$.
-Colors are scaled by the age of the most recent ancestor.}
+\includegraphics[width=\textwidth]{pop_params.pdf}
+\caption{Genealogical parameters from spatial and random mating SLiM simulations, by neighborhood size.}
 \label{fig:spectrum}
@@
 \centering
-\includegraphics[width=\textwidth]{sumstats_by_neighbors.pdf}
-\caption{[needs some polishing for labels etc] Summary statistics from spatial and random-mating SLiM simulations.
-Statistics are calculated from the genotype matrix of 60 randomly sampled individuals in each simulation.}
+\includegraphics[width=\textwidth]{sfs_w_sumstats.pdf}
+\caption{A: Site frequency spectra when sampling individuals from the midpoint, from four equally spaced points, and randomly across the landscape.
+Random mating models showed no differences by sampling scheme -- see supplementary figure S1.
+Spatial models have fewer low-frequency alleles and more mid-frequency alleles when neighborhood sizes are small, and this effect is stronger when sampling is more concentrated.
+B: Summary statistics under different sampling strategies, by neighborhood size.}
 \label{fig:spectrum}
@@
 \centering
-\includegraphics[width=\textwidth]{stairwayplot_by_sigma_facets.pdf}
-\caption{Inferred demographic histories for spatial SLiM simulations from Stairwayplot.
+\includegraphics[width=\textwidth]{stairwayplot_by_sigma_facet_sampling_by_bin.pdf}
+\caption{Inferred demographic histories for spatial SLiM simulations from Stairwayplot by sampling scheme and neighborhood size (NS) range.
 Lines shown are the median inferred $N_{e}$ across 100 bootstrap replicates for each simulation.
-Colors are scaled to Wright's neighborhood size (NS) and panels are divided by neighborhood size range.
-The black line is the census population size.}
+Colors are scaled to Wright's neighborhood size.
+The solid black line is the average census population size, and the dotted line is the average variance effective size ($N_{c}/var(n offspring)$).}
 \label{fig:spectrum}
@@
 \caption{Impacts of spatially correlated phenotypes and isolation by distance on linear regression GWAS.
-Colors represent different phenotype distributions as indicated in the figure legend.
 A: example phenotype distributions and sampling maps, with colors scaled to phenotype values.
@@
 C: numbers of significant SNPs after FDR correction for linear-model GWAS conducted with (top) or without (bottom) PC covariates.
-D: Quantile-quantile plots of observed versus $p$ values relative to those expected from a uniform distribution.
+D: Quantile-quantile plots of observed $p$ values relative to those expected from a uniform distribution.
 The dotted lines show the 95\% confidence region assuming binomial sampling and an alpha level of 0.05.
-Points above the 1:1 line reflect inflated -log10(p).}
+Points above the 1:1 line reflect deflated $p$ values.}
 \label{fig:spectrum}
@@
 \bibliography{spaceness}
+
+\section{Supplementary Figures and Tables}
+\beginsupplement
+
+\afterpage{\clearpage}
+\begin{figure*}[p]
+\centering
+\includegraphics[width=\textwidth]{sumstats_by_neighbors_allstats.pdf}
+\caption{Change in summary statistics by neighborhood size and sampling scheme calculated from simulated sequence data of 60 individuals.}
+\label{fig:spectrum} 
+\end{figure*}
+
+\afterpage{\clearpage}
+\begin{figure*}[p]
+\centering
+\includegraphics[width=\textwidth]{fig_S1_sfs_grid_model_by_sampling.pdf}
+\caption{Site frequency spectra for random mating and spatial SLiM models under all sampling schemes.}
+\label{fig:spectrum}
+\end{figure*}
+
+\afterpage{\clearpage}
+\begin{figure*}[p]
+\centering
+\includegraphics[]{het_z_by_ind.pdf}
+\caption{Normalized mean observed heterozygosity by location across 200 randomly-sampled individuals}
+\label{fig:spectrum}
+\end{figure*}
+
+%\afterpage{\clearpage}
+%\begin{figure*}[p]
+%\centering
+%\includegraphics[width=\textwidth]{stairwayplot_by_sigma_facet_sampling_by_bin.pdf}
+%\caption{Inferred demographic histories for spatial SLiM simulations from Stairwayplot by %sampling scheme and neighborhood size (NS) range.
+Lines shown are the median inferred %$N_{e}$ across 100 bootstrap replicates for each simulation.
+Colors are scaled to Wright's %neighborhood size.
+The solid black line is the average census population size, and the %dotted line is the average variance effective size ($N_{e}/var(n offspring)$).}
+%\label{fig:spectrum}
+%\end{figure*}
+
+\begin{table*}[htbp]
+\centering
+\caption{\bf Anova and Levene's test $p$ values for differences by sampling strategy. Bolded values are rejected at \alpha=0.05}
+\begin{tableminipage}{\textwidth}
+\begin{tabularx}{\textwidth}{XXXX}
+  \hline
+ variable & model & p(equal means) & p(equal variance) \\ 
+  \hline
+segsites & random mating & 0.998190 & 0.980730 \\ 
+pi & random mating & 0.997750 & 0.996450 \\ 
+thetaW & random mating & 0.998190 & 0.980730 \\ 
+tajD & random mating & 0.879690 & 0.188770 \\ 
+het\_o & random mating & 0.531540 & 0.433230 \\ 
+fis & random mating & 0.474790 & 0.785730 \\ 
+gen\_dist\_mean & random mating & 0.997770 & 0.996510 \\ 
+gen\_dist\_var & random mating & 0.283630 & 0.647240 \\ 
+gen\_dist\_skew & random mating & 0.958320 & 0.260750 \\ 
+gen\_sp\_corr & random mating & 0.601980 &\textbf{0.000000} \\ 
+ibs\_mean & random mating & 0.997960 & 0.997730 \\ 
+ibs\_var & random mating & 0.486450 & 0.399490 \\ 
+ibs\_skew & random mating & 0.117980 & 0.069770 \\ 
+ibs\_blocks\_per\_pair & random mating & 0.997680 & 0.996570 \\ 
+ibs\_blocks\_over\_1e6\_per\_pair & random mating & 0.834870 & 0.888730 \\ 
+ibs\_mean\_spat\_corr & random mating & 0.073270 & 0.308420 \\ 
+ibs\_1e6blocks\_spat\_corr & random mating & 0.268440 & \textbf{0.002100} \\ 
+ibs\_skew\_spat\_corr & random mating & 0.396920 & \textbf{0.000620} \\ 
+ibs\_blocks\_spat\_corr & random mating & 0.581090 & \textbf{0.000000} \\ 
+segsites & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+pi & spatial & \textbf{0.026510} & \textbf{0.013440} \\ 
+thetaW & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+tajD & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+het\_o & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+fis & spatial & \textbf{0.000000} & \textbf{0.000120} \\ 
+gen\_dist\_mean & spatial & \textbf{0.025390} & \textbf{0.012910} \\ 
+gen\_dist\_var & spatial & \textbf{0.004970} & \textbf{0.006230} \\ 
+gen\_dist\_skew & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+gen\_sp\_corr & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+ibs\_mean & spatial & 0.272400 & 0.114250 \\ 
+ibs\_var & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+ibs\_skew & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+ibs\_blocks\_per\_pair & spatial & \textbf{0.033920} & \textbf{0.016640} \\ 
+ibs\_blocks\_over\_1e6\_per\_pair & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+ibs\_mean\_spat\_corr & spatial & \textbf{0.000000} & 0.590540 \\ 
+ibs\_1e6blocks\_spat\_corr & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+ibs\_skew\_spat\_corr & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+ibs\_blocks\_spat\_corr & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+\hline
+\end{tabularx}
+\end{tableminipage}
+
+\end{table*}
+
+\endsupplement
 

### 2019-04-01 ms/spaceness.tex
ms updates

@@
 
-%Simulating Isolation by Distance in Continuous Space
 %Space is the Place: How Dispersal and Competition Shape Genetic Variation in Continuous Space
@@
 %Impacts of Isolation by Distance on Demographic Modeling and Association Studies
-%Impacts of Dispersal and Sampling on Estimates of Genetic Variation in Continuous Space
-%Space is the Place: How Dispersal and Sampling Shape Estimates of Genetic Variation in Continuous Space
-\title{Space is the Place: How Dispersal and Sampling Shape Analyses of Genetic Variation in Continuous Space}
+%Space is the Place: How Dispersal and Sampling Shape Inference From Genetic Variation in Continuous Space
+%Evolution on the Final Frontier: How Dispersal and Sampling Shape Inference From Genetic Data in Continuous Space
+\title{Effects of Continuous Spatial Structure on Analysis of Population Genetic Data}
 
@@
 \affil[$\ast$]{University of Oregon Dept. Biology, Institute for Ecology Evolution}
-\keywords{Space; Population Structure; Demography; Haplotype block sharing}
+\keywords{Space; Population Structure; Demography; Haplotype block sharing; GWAS}
 
@@
 \begin{abstract}
-Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations exchanging migrants.
+Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations.
+As a result many models used to infer selection, phenotypic associations, or past population sizes assume that samples are a random draw from a panmictic population.
+In reality most populations are structured to some degree across space, but the extent to which this model violation biases common genomic analyses is not well known.
 Here we implement a forward-time simulation of evolution in continuous space and use it to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and association studies.
-Low dispersal slows the geographic spread of ancestry and increases branch lengths corresponding to mid-frequency alleles because of slow coalescence among distant groups of individuals.
-As a result demographic inference from the site frequency spectrum is biased towards inferring a turbulent past demographic history and a recent decline in populations when neighborhood sizes are below 20.
-For many common population genetic summary statistics the variation associated with different sampling strategies is larger than that observed across the range of dispersal distances we simulated, pointing to the need to carefully account for the spatial distribution of sampling when interpreting results.
-Last we show that spatially correlated environments cause genome-wide signals of association with purely environmental phenotypes in linear-regression GWAS, and that this bias is not fully controlled by regressing out principal components positions during the analysis.
+We find that estimates of past population sizes and many common population genetic summary statistics vary significantly with both dispersal and sampling, with most bias occurring when neighborhood sizes are below 100 and sampling is spatially clustered.
+Variation in the site frequency spectrum and Tajima's D is greater across sampling strategies than over the range of demographic parameters we simulated, pointing to the need to carefully account for sampling when interpreting results.
+Last we show that the combination of isolation by distance and spatially correlated environments causes genome-wide signals of association with purely environmental phenotypes in linear-regression GWAS, and that this bias is not fully controlled by regressing out principal components positions during the analysis.
 \end{abstract}
@@
 The inescapable reality that biological organisms live, move, and reproduce in continuous spatial landscapes has been all but ignored in population genetic models.
-Indeed a near universal rule of reproduction is that individuals mate with other nearby individuals, leading to a positive correlation between genetic and geographic distances.
-This pattern of "isolation by distance" \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, but our mathematical models to describe it are poor approximations of the underlying process.
-As the availability of population-level, whole-genome data allows inference of increasingly fine-scale patterns of ancestry in many species, models incorporating biologically realistic demographic and spatial processes are needed to accurately describe continuous spatial structure and control for its impacts on analyses of selection and demography.
-This situation is particularly clear in human population genetics, where individual ancestries are often interpreted as mixtures of discrete populations in part because of modeling assumptions underlying discrete clustering algorithms and association analyses are often strongly dependent on statistical corrections for population stratification \citep{Berg2018,Bulik-Sullivan2015,Young2018,Heckerman2016,Kang2010,Sohail2018}.
+Indeed, a near universal rule of reproduction is that individuals mate with other nearby individuals, leading to a positive correlation between genetic and geographic distances.
+This pattern of "isolation by distance" \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, but classical mathematical models to describing are flawed approximations of the underlying process \citep{Felsenstein1975}.
+As a result many studies focus on describing geographic structure as a mixture of random-mating populations connected by migration \citep{Malecot1948,Wright1931}, and analyze variation within clusters of genetic variation inferred by programs like \textit{structure} \citep{Pritchard2000} as random-mating units.
 
-The best-studied approaches to population genetics in continuous space were developed by \citep{Wright1943} and \citep{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals.
+Though discrete models of between-population differentiation account for much of the genetic variation in species, the assumption that populations are randomly mating at some level has important implications for downstream inference of selection and demography.
+Methods based on the coalescent \citep{Kingman1982} assume that the sampled individuals are a random draw from a random mating population.
+If dispersal or mate selection is limited by space these assumptions are violated -- nearby individuals will be more closely related than an average random pair and drawing multiple samples from the same point on the landscape will not represent a random sample of the genetic variation present in the whole population.
+This is as true at the local scale as it is between the more differentiated units typically analyzed in phylogeographic studies.
+All downstream inferences drawn from the patterns of relationship inferred among sampled individuals will be subject to bias, but the extent depends on the degree of variation in ancestry across the landscape.
+
+This issue is particularly important for analyses of polygenic selection, because the allele frequency variation caused by population structure is similar to the pattern expected when many sites in the genome are associated with a given phenotype \citep{Bulik-Sullivan2015}.
+Indeed two recent studies found that previous evidence of polygenic selection on human height in Europe was badly confounded by population structure \citep{Sohail2018,Berg2018}.
+As the scale of sequence data now available for many species allows inference of increasingly fine-scale patterns of selection and demography, understanding how and when subtle spatial structure is likely to bias results is an important task for population genetics.
+
+Here we describe an implementation of an individual-based model in continuous space that incorporates overlapping generations, a Gaussian dispersal kernel, and density-dependent fitness.
+The model scales to chromosome-scale alignments across tens of thousands of individuals, and outputs the full genealogy and recombination history of all final-generation individuals.
+We use this simulation to test how sampling strategy interacts with isolation by distance to cause systematic variation in population genetic summary statistics typically analyzed under discrete population models.
+We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum.
+
+Last we use our model to study the impacts of isolation by distance in continuous space on genome-wide association studies (GWAS).
+Variation in ancestry proportions between case and control cohorts is expected to cause inflation of test statistics at SNPs with different allele frequencies among populations \citep{Price2006}.
+Because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within continuous populations can also deflate $p$ values in GWAS of traits influenced by the environment \citep{Mathieson2012}.
+Here we simulate both the underlying genealogy and phenotypes of a continuously distributed population and seek to identify regions of parameter space -- i.e.
+the strength of isolation by distance and the spatial distribution of environmental effects on phenotypes -- in which common methods of stratification bias corrections in GWAS are likely to fail.
+
+\subsection{Modeling Evolution in Continuous Space}
+The best-studied approaches to population genetics in continuous space were developed by \cite{Wright1943} and \cite{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals.
 A key finding of Wright's model is that many important aspects of continuous populations can be described in terms of "neighborhood size" -- the number of potential mates for an individual in a given generation, defined as $4\pi\sigma^2d$, where $\sigma$ is the average dispersal distance and $d$ is population density.
-\cite{Maruyama1972} found that the rate of decline in genetic diversity in a 2-dimensional continuous population approaches the random mating expectation when $D\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.
+\cite{Maruyama1972} found that the rate of decline in genetic diversity in a 2-dimensional continuous population approaches the random mating expectation when $d\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.
 
-Though some aspects of continuous populations are fairly well described by the Wright-Malécot models, \cite{Felsenstein1975} showed that the assumptions of independent dispersal and Poisson distributed offspring that are the basis of these models are incompatible.
+Though some aspects of continuous populations are fairly well described by the Wright and Malécot models, \cite{Felsenstein1975} showed that the assumptions of independent dispersal and Poisson distributed offspring that are the basis of these models are incompatible.
 Over time, a population meeting them will clump into a small number of geographic clusters occupying only a part of the available range.
@@
 Among many other important results drawn from this class of "lattice" or "stepping stone" models, \cite{Rousset1997} showed that the slope of the a linear regression of genetic differentiation ($F_{st}$) against the logarithm of spatial distance is an estimate of neighborhood size.
-Though good approximations of continuous structure given high dispersal (Barton(?)), these models are not truly continuous, force a uniform realized population density across landscapes, and limit investigation of spatial structure below the level of the deme.
+Though good approximations of continuous structure given high dispersal, these models are not truly continuous, force a uniform realized population density across landscapes, and limit investigation of spatial structure below the level of the deme.
 An alternative method is to model the geographic spread of ancestry backwards in time through a diffusion approximation -- an approach that has recently made significant progress in modeling both dispersal and demographic parameters \citep{Barton2010,Kelleher2014,Ringbauer2017,Ringbauer2018}.
-However these models currently lack a well-described forward-time analog that produces biologically realistic spatial patterns and are subject to the same difficulty in incorporating selection and other biological processes as all other reverse time methods (CITE??).
 
-\ak{Andy-- I think we could move some of the text from the following paragraph to methods}
-We took a direct approach to the clustering problem of classical forward-time models by incorporating density dependence into the model.
+We took a direct approach to the clustering problem of classical forward-time models by incorporating density dependence.
 By scaling the probability of survival in each timestep to local population density we shift reproductive output towards regions of low-density and prevent populations from clustering.
 A similar approach was taken previously by \citep{Doebeli2003} who used an individual based model with continuous space and density dependent fitness to study the probability of speciation along continuous environmental gradients.
-However to our knowledge all previous implementations of continuous space models have focused on a small number of genetic loci as the unit of analysis, which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms.
-
-Here we describe an implementation of an individual-based model in continuous space that incorporates overlapping generations, a Gaussian dispersal kernel, and density-dependent fitness.
-The model scales to chromosome-scale alignments across tens of thousands of individuals, and outputs the full genealogy and recombination history of all final-generation individuals.
-We use this simulation to describe how sampling strategy interacts with isolation by distance to cause systematic variation in population genetic summary statistics typically analyzed under discrete population models.
-We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum; finding that inflation of branch lengths for mid-frequency alleles associated with slow coalescence of distant individuals can bias analyses towards inferring turbulent demographic histories and recent declines in population size.
-
-Last, we use our model to study the impacts of isolation by distance in continuous space on genome-wide association studies (GWAS).
-Variation in ancestry proportions between case and control cohorts is expected to cause inflation of test statistics at SNPs with different allele frequencies among populations \citep{Price2006}.
-Because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within continuous populations can also deflate $p$ values in GWAS of quantitative traits \citep{Mathieson2012}.
-These issues are particularly prominent for polygenic traits, because the deflated $p$ values expected from polygenic phenotype correlations are superficially similar to those expected from population structure \citep{Bulik-Sullivan2015,Young2018}.
-However most studies examining the effects of stratification on GWAS have focused on simulating phenotypes for real genotype data \citep{Price2006,Bulik-Sullivan2015} or generating simulated genotypes by sampling from static allele frequency distributions \citep{Bulik-Sullivan2015,Heckerman2016}.
-These approaches address application to human data but are not based on a true generative model and so provide limited insight into the processes generating stratification bias in association analyses.
-Here we simulate both the underlying genealogy and phenotypes of a continuously distributed population and seek to identify regions of parameter space -- i.e.
-the strength of isolation by distance and the spatial distribution of environmental effects on phenotypes -- in which common methods of stratification bias corrections in GWAS are subject to bias.
-
-%[move some of this to methods, rewrite to avoid repitition w above]  A common method of correcting for population structure in GWAS is to run a principal components analysis on the genotype matrix and include PC coordinates as covariates in a linear regression for each SNP \citep{Price2006}.
-Mixed model GWAS employs a different statistical framework but takes a similar approach by incorporating a kinship matrix describing pairwise genetic relatedness among samples as a random effect in a linear mixed model \citep{Yu2006,Kang2008}.
-In both approaches the key question is whether or not the ancestry components of the model adequately describe "background" levels of genetic variation in the sample relative to any confounding effects of ancestry and phenotype.
-Recent studies in European and British populations \citep{Kerminen2018,Berg2018,Sohail2018} have found evidence of residual population stratification in effect size estimates from large-scale GWAS employing both linear and mixed model approaches, suggesting that existing structure corrections may be inadequate.
-For traits controlled by a few large effect loci this confounding effect is likely to be relatively small, but for analyses of complex traits with thousands of putatively associated SNPs the potential for confounding is high.
-Here we simulate a range of spatially correlated phenotypes for simulated individuals and seek to identify conditions under which existing GWAS structure corrections are subject to bias caused by continuous population structure.
+However to our knowledge previous implementations of continuous space models have focused on a small number of genetic loci as the unit of analysis, which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms.
+By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.
 
@@
 
-%We compared our model output to a two-dimensional stepping-stone model and a one-population coalescent model implemented in msprime \citep{Kelleher2016}.
-To isolate spatial effects from other components of the model such as overlapping generations and density-dependent probability of survival, we also ran a second version of the SLiM model with random mate selection and dispersal.
-
 We ran 400 simulations for the spatial and random-mating SLiM models on a square landscape of size 50 with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), average lifetime $L=4$, genome size $1\times10^{8}$, recombination rate $1\times10^{-9}$, and drawing $/sigma$ values from a uniform distribution bounded by 0.2 and 4.
@@
 Because msprime assumes that timesteps are in units of generations, we divided the per-generation rate of $1\times10^{-8} mut/site/gen$ by the generation time estimated for each value of $\sigma$ (see 'Genealogical Parameters' below) to convert the rate to units of mutations/site/timestep.
-We then verified that this procedure produced the correct number of mutations by comparing a subset of simulations with SLiM-generated mutations (which are applied only at mating events) with mutations added by msprime (Supplementary Figure 1).
+We then verified that this procedure produced the correct number of mutations by comparing a subset of simulations with SLiM-generated mutations (which are applied only at mating events) with mutations added by msprime.
 Simulations were run for 1.6 million timesteps (approximately 30N generations), or until all extant individuals shared a common ancestor within the simulation (i.e.
@@
 
-%Coalescent stepping-stone simulations were on a 5x5 grid with 10 haploid samples per deme and migration between adjacent populations.
-To facilitate comparison across simulation types we ran one coalescent simulation for every SLiM parameter set.
-Because $\sigma$ is approximately equal to the mean distance between parent and offspring, we approximate the relationship between $\sigma$ and the migration rate in a grid simulation as $\sigma=4md$, where $m$ is the migration rate and $d$ is the ratio of landscape widths in continuous space vs coalescent simulations (here, 10).
-That is, an individual in an interior population has a $4m$ probability of being a migrant, and if it is a migrant will move $d$ units on the map.
-We set the total $N_{e}$ equal to the census size of each SLiM simulation divided by the variance in reproductive output (see below).
+\subsection{Genealogical Parameters}
+We recorded the census population size and estimated the variance in number of offspring and generation time for all simulations.
+To estimate generation times we stored the age of all parents for 200 timesteps after a 100 generation burn-in and took the mean.
+To estimate variance in offspring we tracked the number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period, subset the resulting table to include only the last timestep recorded for each individual, and calculated the variance across all individuals in timesteps 50-100.
 
-\subsection{Genealogical Parameters}
-To study the geographic spread of genealogical ancestry in our simulations we output the full pedigree of all individuals over 50 generations with $\sigma$ set to 0.2, 0.35, 0.5, 1, 2, and 3.5 (corresponding to Wright's Neighborhood Sizes of 2.5, 7.7, 15.7, 62.8, 251.3, and 769.7, respectively).
-We then selected the individual closest to the center of the landscape in the final generation of each simulation, pruned the pedigree to include only its genealogical ancestors (that is, regardless of the amount of genetic material actually inherited from a given ancestor) and plotted the age of its most recent ancestor in each region of the full landscape using the 2-dimensional summary feature of ggplot2 in R \citep{Wickham2016}.
-We calculated the rate of spread of genealogical ancestry by estimating the mean and maximum distances between the focal individual and its ancestors in each generation and then fitting a linear model for distance as a function of generations using ordinary least squares regression.
-
-We also estimated the variance in offspring and generation time for all simulations.
-To estimate generation times we stored the age of all parents for 200 timesteps and took the mean.
-To estimate variance in offspring we output the complete pedigree of all individuals in SLiM, counted the number of offspring for all individuals born in timesteps 50-150 (to ensure that the simulation captured the full lifespan of the sampled individuals), and calculated the variance.
+\cjb{Would it be worthwhile to calculate the rate of spread of ancestry as a function of neighborhood size? IE how long until an individual in the middle of the map has ancestors in every region?}
 
 \subsection{Sampling}
-Our model records the genealogy and sequence variation of the complete population, but in real data genotypes are only observed from a a relatively small number of sampled individuals.
+Our model records the genealogy and sequence variation of the complete population, but in real data genotypes are only observed from a relatively small number of sampled individuals.
 We modeled three sampling strategies similar to common data collection methods in empirical genetic studies (Figure 1).
 "Random" sampling selects individuals at random from across the full landscape, "point" sampling selects individuals proportional to their distance from four equally spaced points on the landscape, and "midpoint" sampling selects individuals in proportion to their distance from the middle of the landscape.
-Downtream analyses were repeated across all sampling strategies.
+Downstream analyses were repeated across all sampling strategies.
 
 \subsection{Summary Statistics}
-We calculated the site frequency spectrum and a set of 18 summary statistics (Supplementary Table 1) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}.
+We calculated the site frequency spectrum and a set of 18 summary statistics (Supplementary Table X \cjb{working on this}) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}.
 Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{is}$), and Tajima's D, as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distances \citep{Rousset1997}, which we summarized as the correlation coefficient between $log10(spatial distance)$ and the proportion of identical base pairs among individuals.
@@
 Models were fit across all bootstrap replicates using default settings in Stairwayplot and the median estimate of $Ne$ per generation was used to represent the output of each simulation.
+
+In preliminary runs we found that inferred population histories were relatively variable even when simulating under a coalescent model, suggesting that some of the differences in demographic estimates for spatial models are caused by the behavior of the optimization algorithm rather than bias in the SFS caused by spatial mate choice and dispersal.
+To separate these effects we ran 100 coalescent simulations with constant population size $6.1\times 10^{-3}$ (the mean $N_{e}$ of random-mating SLiM models estimated from $pi$) and fit stairwayplot models using the same script as for our spatial models.
+We then calculated the standard deviation of inferred $N_{e}$ in each stairwayplot model to summarize the degree of fluctuation around the simulated population size, and asked if standard deviations were higher in spatial relative to coalescent models with a one-tailed t-test.
 
@@
 \subsection{Genealogical Parameters and Run Times}
+Census population size, generation time, and the variance in number of offspring all vary systematically with $\sigma$.
+Census sizes range from $\approx 14,000$ at low $\sigma$ in the random mating model to $\approx 10,000$ for both models when neighborhood sizes approach 1,000.
+Generation times vary between 5.2 and 4.9 timesteps per generation in the spatial model, and variance in offspring varies from 2.2 to 2.6.
+
+%The fitness scaling procedure we use to avoid high fitness at range edges shrinks the suitable area by one $\sigma$ on all sides and likely explains part of the variation in census size, but scaling in offspring variance and the U-shaped distribution of census size in spatial models suggests additional processes at work.
+For spatial models census size initially rises as neighborhood sizes scale from 2 - 10, which seems to reflect an Allee effect \citep{Allee1949} in which some individuals are unable to find mates when the mate selection radius is very small.
+
 \subsection{Impacts of Sampling on Summary Statistics}
@@
 The inflation in $F_{is}$ appears to be caused by a deficit of heterozygous individuals in low-dispersal simulations -- a continuous-space version of the Wahlund effect \citep{Wahlund1928}.
-However under midpoint sampling observed heterozygosity is inflated over the random mating expectation, which we confirmed is caused by a higher proportion of heterozygotes in the middle of the landscape (Supplementary Figure S2 \cjb{CJ - this fig isn't super convincing -- will try denser sampling}).
+However under midpoint sampling observed heterozygosity is inflated over the random mating expectation, which we confirmed is caused by a higher proportion of heterozygotes in the middle of the landscape (Supplementary Figure S2).
 \cite{Shirk2014} observed a similar excess of heterozygosity in the middle of the landscape when simulating under a lattice model.
@@
 \subsubsection{Haplotype Block Lengths}
-Trends in pairwise haplotype block sharing parallel those in SNP-based diversity estimates.
+Trends in pairwise haplotype block sharing parallel those in SNP-based diversity estimates (Supplementary Figure S1).
 At low dispersal the distribution of IBS block lengths in a set of samples is shifted towards smaller values -- resulting in lower means and fewer long IBS blocks.
@@
 \subsection{Demographic Modeling}
-Demographic models from Stairwayplot tended to infer patterns of ancient population increases and recent declines when neighborhood sizes were low under all sampling strategies (Figure 3).
+Demographic models from Stairwayplot tended to infer patterns of ancient population increases and recent declines when neighborhood sizes were below 20 under all sampling strategies (Figure 4).
 Inflated past population sizes were common in both point and random sampling, demonstrating that the relatively minor shift in the site frequency spectrum observed with random sampling is enough to alter demographic estimates.
-Inference of severe population bottlenecks was also common across neighborhood sizes, particularly under midpoint and point sampling.
-Across all sampling schemes and neighborhood sizes around 20\% of runs inferred a sharp increase in population in the last 100 years.
+Inference of severe population bottlenecks was also common at neighborhood sizes under 100 for midpoint and point sampling strategies.
+Above neighborhood sizes of 100 the average inferred demography across all simulations was relatively accurate, with minor fluctuations slightly above the expected variance $Ne$.
+However individual model fits were highly variable and often inferred five-fold or greater population fluctuations even in high-dispersal simulations.
 
-Estimates stabilize for all sampling schemes as neighborhood sizes approach 750, and trend towards roughly half the census size. 
-\cjb{I still need to figure out what's going on with Ne here...
-pi is $\approx$ expected value for a WF population with $N\approx N_{c}/2$, but variance Ne is more like 1/5 census N.
-I thought it was closer to 1/2 earlier because of a bug in my script where I was dropping individuals who had 0 offspring.
-Working on it...}.
+Standard deviations were significantly greater in all sampling strategies for neighborhood sizes under 20 and for midpoint sampling with neighborhood sizes 20-100, but all other sets performed similarly to the coalescent model.
+In summary, spatial mate choice and dispersal causes strong bias in SFS-based demographic estimates for neighborhood sizes below 20 or when sampling is clustered, but otherwise any biases are within the range of variability regularly inferred by Stairwayplot.
+In empirical data, it may be useful to run 10 or more replicate analyses on the same data and take a rolling average, which appears relatively robust for our simulations.
+
+% latex table generated in R 3.5.1 by xtable 1.8-3 package
+% Fri Mar 22 13:50:47 2019
+\begin{table}[ht]
+\centering
+\caption{\bf T-test results comparing standard deviations of inferred $N_{e}$ between spatial and coalescent models, by neighborhood size (NS) and sampling strategy.
+$p$ is the probability that spatial models have higher standard deviations.}
+\begin{tabular}{rllrrrrr}
+  \hline
+sampling & NS range & t & df & $p$ \\ 
+  \hline
+random & 2-20 & 4.2572 & 41.6166 & 0.0001 \\ 
+random & 20-100 & -1.8473 & 171.9905 & 0.9668 \\ 
+random & 100-500 & -2.1297 & 164.3864  & 0.9827 \\ 
+random & 500-1000 & -3.9681 & 147.0497 & 0.9999 \\ 
+point & 2-20 & 7.0802 & 44.3615  & 0.0000 \\ 
+point & 20-100 & -0.2038 & 169.3799 & 0.5806 \\ 
+point & 100-500 & -2.4945 & 152.5000 & 0.9932 \\ 
+point & 500-1000 & -3.8329 & 162.6443& 0.9999 \\ 
+midpoint & 2-20 & 5.9253 & 59.5462 & 0.0000 \\ 
+midpoint & 20-100 & 3.8940 & 171.7005  & 0.0001 \\ 
+midpoint & 100-500 & -2.2764 & 139.5221 & 0.9878 \\ 
+midpoint & 500-1000 & -3.2223 & 165.0792 & 0.9992 \\ 
+   \hline
+\end{tabular}
+\end{table}
 
@@
 
+\subsection{Sampling}
+
+- clustered sampling causes more variation than dispersal scaling for several statistics
+- range edges cause differences in the distribution of genetic variation across a range so sampling the midpoint gives a biased estimate of diversity and heterozygosity.
+Tajima's D is particularly sensitive to this effect.
+- in demographic analyses
+
+\subsection{Demography}
+- low dispersal and clustered sampling bias recent Ne down and inflate past Ne
+- estimates are noisy in all simulations -- need to repeat analyses and average even for coalescent sims (???)
+-
+
 \subsection{GWAS}
+
+- points: PCA correction is pretty good but not perfect. low-effect-size-alleles prob hit worst. 
+- we know that this is also a problem empirically in human data, and was enough of an issue for arabidopsis that PC regression isn't used.
+Approaches like Bayesenv or the Science yellow warbler paper likely contain residual structure, particularly when environmenal clines parallel genetic clines and dispersal is high.
+- caveats: small populations, need to scale, mixed are probably likely better
+- recs: run a gwas for sample x & y coordinates with whatever correction you're using. If there are significant values you're not correcting enough.
+- for clinal traits, include x and y as covariates.
 
@@
 \newpage
+\newpage
 
@@
 
-\begin{figure*}[htbp]
-\centering
-\includegraphics[width=\textwidth]{pop_params.pdf}
+\begin{figure}[htbp]
+\includegraphics{pop_params.pdf}
 \caption{Genealogical parameters from spatial and random mating SLiM simulations, by neighborhood size.}
 \label{fig:spectrum}
-\end{figure*}
+\end{figure}
 
@@
 \includegraphics[width=\textwidth]{sfs_w_sumstats.pdf}
-\caption{A: Site frequency spectra when sampling individuals from the midpoint, from four equally spaced points, and randomly across the landscape.
+\caption{A: Site frequency spectra under midpoint, point, and random sampling schemes.
+Lines are loess curves over the site frequency spectrum from each simulation with colors scaled to neighborhood size.
 Random mating models showed no differences by sampling scheme -- see supplementary figure S1.
 Spatial models have fewer low-frequency alleles and more mid-frequency alleles when neighborhood sizes are small, and this effect is stronger when sampling is more concentrated.
-B: Summary statistics under different sampling strategies, by neighborhood size.}
+B: Summary statistics under each sampling strategy, by neighborhood size.}
 \label{fig:spectrum}
@@
 \centering
-\includegraphics[width=\textwidth]{stairwayplot_by_sigma_facet_sampling_by_bin.pdf}
-\caption{Inferred demographic histories for spatial SLiM simulations from Stairwayplot by sampling scheme and neighborhood size (NS) range.
-Lines shown are the median inferred $N_{e}$ across 100 bootstrap replicates for each simulation.
-Colors are scaled to Wright's neighborhood size.
-The solid black line is the average census population size, and the dotted line is the average variance effective size ($N_{c}/var(n offspring)$).}
+\includegraphics[width=\textwidth]{stairwayplot_facet_rollmean.pdf}
+\caption{Inferred demographic histories for spatial SLiM simulations from Stairwayplot, by sampling scheme and neighborhood size (NS) range.
+The thick line is a rolling mean and thin lines are individual model fits.
+Dashed horizontal lines are the average $N_{e}$ across random-mating SLiM models estimated from $\pi$.}
 \label{fig:spectrum}

### 2019-04-09 ms/spaceness.tex
moved figs to folders, overleaf update 9 Apr

@@
 
-\documentclass[9pt,twocolumn,twoside,lineno]{gsajnl}
+\documentclass[10pt,twoside,lineno]{gsajnl}
 \usepackage{amsmath}
+\linespread{1.5}
 % Use the documentclass option 'lineno' to view line numbers
@@
 \thispagestyle{firststyle}
-\marginmark
+%\marginmark
 \firstpagefootnote
@@
 
-We took a direct approach to the clustering problem of classical forward-time models by incorporating density dependence.
+We took a direct approach to the clustering problem of classical forward-time models by incorporating density dependence into an individual-based model similar to the analytic models developed by Wright and Malécot.
 By scaling the probability of survival in each timestep to local population density we shift reproductive output towards regions of low-density and prevent populations from clustering.
@@
 We approximate a decline in habitat suitability near edges by decreasing the probability of survival proportional to the square root of distance to edges in units of $\sigma$.
-The final probability of survival for individuals within one $\sigma$ of an edge is
+The final probability of survival for individual \textit{i} is then
 
-\begin{multline}
+\begin{equation}
     s_{i}=d_{i} min(1,\sqrt{x_{i}/\sigma})
@@
     min(1,\sqrt{(y_{max}-y_{i})/\sigma)}
-\end{multline}
+\end{equation}
 
-where $x$ and $y$ are spatial coordinates.
+where $x$ and $y$ are spatial coordinates. In practice this creates a one $\sigma$ buffer around the edge of the map in which fitness is depressed.
 
 To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, and density-dependent fitness, we  also implemented a second version of the SLiM simulation in which mates are selected at random and offspring disperse to a random location on the landscape.
+We refer to these two models as the "spatial" and "random mating" models.
 For all simulations the full genealogy and recombination history of final-generation individuals were stored as tree sequences \citep{Kelleher2018}.
@@
 To estimate generation times we stored the age of all parents for 200 timesteps after a 100 generation burn-in and took the mean.
-To estimate variance in offspring we tracked the number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period, subset the resulting table to include only the last timestep recorded for each individual, and calculated the variance across all individuals in timesteps 50-100.
-
-\cjb{Would it be worthwhile to calculate the rate of spread of ancestry as a function of neighborhood size? IE how long until an individual in the middle of the map has ancestors in every region?}
+To estimate variance in offspring we tracked the number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period, subset the resulting table to include only the last timestep recorded for each individual, and calculated the variance in number of offspring across all individuals in timesteps 50-100.
 
@@
 Our model records the genealogy and sequence variation of the complete population, but in real data genotypes are only observed from a relatively small number of sampled individuals.
-We modeled three sampling strategies similar to common data collection methods in empirical genetic studies (Figure 1).
+We modeled three sampling strategies similar to common data collection methods in empirical genetic studies (Figure \ref{fig:samplemap}).
 "Random" sampling selects individuals at random from across the full landscape, "point" sampling selects individuals proportional to their distance from four equally spaced points on the landscape, and "midpoint" sampling selects individuals in proportion to their distance from the middle of the landscape.
@@
 \subsection{Summary Statistics}
-We calculated the site frequency spectrum and a set of 18 summary statistics (Supplementary Table X \cjb{working on this}) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}.
+We calculated the site frequency spectrum and a set of 18 summary statistics (Supplementary Table \cjb{working on this}) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}.
 Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{is}$), and Tajima's D, as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distances \citep{Rousset1997}, which we summarized as the correlation coefficient between $log10(spatial distance)$ and the proportion of identical base pairs among individuals.
@@
 \subsection{Genealogical Parameters and Run Times}
-Census population size, generation time, and the variance in number of offspring all vary systematically with $\sigma$.
+Census population size, generation time, and the variance in number of offspring all vary systematically with $\sigma$ (Figure \ref{fig:genparams}).
 Census sizes range from $\approx 14,000$ at low $\sigma$ in the random mating model to $\approx 10,000$ for both models when neighborhood sizes approach 1,000.
@@
 \subsection{Impacts of Sampling on Summary Statistics}
-For spatial models the mean and variance of all summary statistics other than those involving the mean of the IBS block size distribution was significantly different across sampling strategies (Supplementary Table 1).
+For spatial models the mean and variance of all summary statistics other than those involving the mean of the IBS block size distribution was significantly different across sampling strategies (Table \ref{table:sampling}).
 Means were not significantly different across sampling strategies for any statistic under the random mating model.
@@
 \subsubsection{Diversity and Site Frequency Spectra}
-Mean within-population genetic distance ($\pi$), the inbreeding coefficient ($F_{is}$) and Tajima's $D$ are inflated at low dispersal distances (Figure 2).
+Mean within-population genetic distance ($\pi$), the inbreeding coefficient ($F_{is}$) and Tajima's $D$ are inflated at low dispersal distances (Figure \ref{fig:sumstats}, \ref{fig:allsumstats}).
 $\pi$ and $F_{is}$ approach the random mating model at neighborhood sizes of roughly 100.
@@
 The inflation in $F_{is}$ appears to be caused by a deficit of heterozygous individuals in low-dispersal simulations -- a continuous-space version of the Wahlund effect \citep{Wahlund1928}.
-However under midpoint sampling observed heterozygosity is inflated over the random mating expectation, which we confirmed is caused by a higher proportion of heterozygotes in the middle of the landscape (Supplementary Figure S2).
+However under midpoint sampling observed heterozygosity is inflated over the random mating expectation, which we confirmed is caused by a higher proportion of heterozygotes in the middle of the landscape (Figure S\ref{fig:hetmap}).
 \cite{Shirk2014} observed a similar excess of heterozygosity in the middle of the landscape when simulating under a lattice model.
 
-The site frequency spectrum showed minimal differences across sampling strategies for random mating models (Supplementary Figure S1).
-Spatial models had fewer low frequency alleles and more mid-frequency alleles than random mating models (Figure 2).
+The site frequency spectrum showed minimal differences across sampling strategies for random mating models (Supplementary Figure \ref{fig:allsfs}).
+Spatial models had fewer low frequency alleles and more mid-frequency alleles than random mating models (Figure \ref{fig:sumstats}).
 This effect is relatively weak when sampling is random and strongest for midpoint sampling.
@@
 \subsubsection{Haplotype Block Lengths}
-Trends in pairwise haplotype block sharing parallel those in SNP-based diversity estimates (Supplementary Figure S1).
+Trends in pairwise haplotype block sharing parallel those in SNP-based diversity estimates (Figure \ref{fig:sumstats}, Supplementary Figure \ref{fig:allsumstats}).
 At low dispersal the distribution of IBS block lengths in a set of samples is shifted towards smaller values -- resulting in lower means and fewer long IBS blocks.
@@
 \subsection{Demographic Modeling}
-Demographic models from Stairwayplot tended to infer patterns of ancient population increases and recent declines when neighborhood sizes were below 20 under all sampling strategies (Figure 4).
+Demographic models from Stairwayplot tended to infer patterns of ancient population increases and recent declines when neighborhood sizes were below 20 under all sampling strategies (Figure~\ref{fig:demography}).
 Inflated past population sizes were common in both point and random sampling, demonstrating that the relatively minor shift in the site frequency spectrum observed with random sampling is enough to alter demographic estimates.
@@
 
-Standard deviations were significantly greater in all sampling strategies for neighborhood sizes under 20 and for midpoint sampling with neighborhood sizes 20-100, but all other sets performed similarly to the coalescent model.
+Standard deviations were significantly greater in all sampling strategies for neighborhood sizes under 20 and for midpoint sampling with neighborhood sizes 20-100, but all other sets performed similarly to the coalescent model (Table \ref{table:demography}).
 In summary, spatial mate choice and dispersal causes strong bias in SFS-based demographic estimates for neighborhood sizes below 20 or when sampling is clustered, but otherwise any biases are within the range of variability regularly inferred by Stairwayplot.
@@
 
-% latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Fri Mar 22 13:50:47 2019
-\begin{table}[ht]
-\centering
-\caption{\bf T-test results comparing standard deviations of inferred $N_{e}$ between spatial and coalescent models, by neighborhood size (NS) and sampling strategy.
-$p$ is the probability that spatial models have higher standard deviations.}
-\begin{tabular}{rllrrrrr}
-  \hline
-sampling & NS range & t & df & $p$ \\ 
-  \hline
-random & 2-20 & 4.2572 & 41.6166 & 0.0001 \\ 
-random & 20-100 & -1.8473 & 171.9905 & 0.9668 \\ 
-random & 100-500 & -2.1297 & 164.3864  & 0.9827 \\ 
-random & 500-1000 & -3.9681 & 147.0497 & 0.9999 \\ 
-point & 2-20 & 7.0802 & 44.3615  & 0.0000 \\ 
-point & 20-100 & -0.2038 & 169.3799 & 0.5806 \\ 
-point & 100-500 & -2.4945 & 152.5000 & 0.9932 \\ 
-point & 500-1000 & -3.8329 & 162.6443& 0.9999 \\ 
-midpoint & 2-20 & 5.9253 & 59.5462 & 0.0000 \\ 
-midpoint & 20-100 & 3.8940 & 171.7005  & 0.0001 \\ 
-midpoint & 100-500 & -2.2764 & 139.5221 & 0.9878 \\ 
-midpoint & 500-1000 & -3.2223 & 165.0792 & 0.9992 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
 \subsection{GWAS}
-Spatially correlated phenotypes in populations with low neighborhood size cause uncorrected linear-regression GWAS to report significant associations at most genome-wide SNPs (Figure 1).
+Spatially correlated phenotypes in populations with low neighborhood size cause uncorrected linear-regression GWAS to report significant associations at most genome-wide SNPs \ref{fig:gwas}.
 When PCA positions are included as covariates the vast majority of SNPs no longer pass a 5\% FDR significance threshold, but up to 1000 genome-wide significant hits (out of an average of 132,000 SNPs) were observed even in relatively high dispersal simulations for the corner and clinal distributions.
@@
 For clinal phenotypes PCA covariates perform well in correcting for population stratification at low dispersal distances but the correction breaks down above neighborhood sizes of $\approx250$.
-This pattern appears to reflect a loss of descriptive power in the first ten axes of the PCA, which describe less of the total variation in the data as neighborhood size increases (Figure 4).
+This pattern appears to reflect a loss of descriptive power in the first ten axes of the PCA, which describe less of the total variation in the data as neighborhood size increases (see Figure 4B-C).
 
@@
 
-% As in \citep{Mathieson2012} we observed that the deflation in $p$ values after PCA correction is largest for spatially discrete phenotypes, and is weaker for phenotypes that are clinal over space.
-When dispersal is low and isolation by distance is strong this effect is pronounced enough to create statistically significant associations at nearly every variable site in the genome unless statistical corrections for population structure are implemented.
-For strongly structure populations and phenotypes that vary systematically over space, the estimated effect sizes and biological interpretations of linear-regression GWAS are therefore controlled primarily by the method of stratification control rather than underlying causal associations between phenotype and genotype.
-
-%This is unlikely to be a significant issue for studies seeking to identify a small number of causal alleles as long as there is variation within populations at those sites.
-However analyses of putatively polygenic phenotypes may be badly affected both by deflation of $p$ values from uncontrolled stratification and by inflation of $p$ values caused by the use of PCA covariates.
-Though many studies now employ mixed-model methods \citep{Kang2008,Yu2006,Kang2010} which are likely more robust to stratification (though not immune, see \citep{Berg2018}), using principal components regression as the method of stratification control is still common in analyses of human genetics [quality time w nature genetics].
-
 \section{Data Availability}
@@
 \begin{figure}[htbp]
-\includegraphics{sampling_maps.pdf}
+\includegraphics{figures/sampling_maps.pdf}
 \caption{Example sampling maps for 60 individuals on a 50x50 landscape under midpoint, point, and random sampling strategies.}
-\label{fig:spectrum}
+\label{fig:samplemap}
 \end{figure}
@@
 \begin{figure}[htbp]
-\includegraphics{pop_params.pdf}
+\includegraphics{figures/pop_params.pdf}
 \caption{Genealogical parameters from spatial and random mating SLiM simulations, by neighborhood size.}
-\label{fig:spectrum}
+\label{fig:genparams}
 \end{figure}
@@
 \centering
-\includegraphics[width=\textwidth]{sfs_w_sumstats.pdf}
+\includegraphics[width=\textwidth]{figures/sfs_w_sumstats.pdf}
 \caption{A: Site frequency spectra under midpoint, point, and random sampling schemes.
@@
 B: Summary statistics under each sampling strategy, by neighborhood size.}
-\label{fig:spectrum}
+\label{fig:sumstats}
 \end{figure*}
@@
 \centering
-\includegraphics[width=\textwidth]{stairwayplot_facet_rollmean.pdf}
+\includegraphics[width=\textwidth]{figures/stairwayplot_facet_rollmean.pdf}
 \caption{Inferred demographic histories for spatial SLiM simulations from Stairwayplot, by sampling scheme and neighborhood size (NS) range.
@@
 Dashed horizontal lines are the average $N_{e}$ across random-mating SLiM models estimated from $\pi$.}
-\label{fig:spectrum}
+\label{fig:demography}
 \end{figure*}
@@
 \centering
-\includegraphics[width=\textwidth]{gwas_summary_nsig_qqplots_loglog.pdf}
+\includegraphics[width=\textwidth]{figures/gwas_summary_nsig_qqplots_loglog.pdf}
 \caption{Impacts of spatially correlated phenotypes and isolation by distance on linear regression GWAS.
@@
 Points above the 1:1 line reflect deflated $p$ values.}
-\label{fig:spectrum}
+\label{fig:gwas}
 \end{figure*}
@@
 \centering
-\includegraphics[width=\textwidth]{sumstats_by_neighbors_allstats.pdf}
+\includegraphics[width=\textwidth]{figures/sumstats_by_neighbors_allstats.pdf}
 \caption{Change in summary statistics by neighborhood size and sampling scheme calculated from simulated sequence data of 60 individuals.}
-\label{fig:spectrum} 
+\label{fig:allsumstats} 
 \end{figure*}
@@
 \centering
-\includegraphics[width=\textwidth]{fig_S1_sfs_grid_model_by_sampling.pdf}
+\includegraphics[width=\textwidth]{figures/fig_S1_sfs_grid_model_by_sampling.pdf}
 \caption{Site frequency spectra for random mating and spatial SLiM models under all sampling schemes.}
-\label{fig:spectrum}
+\label{fig:allsfs}
 \end{figure*}
@@
 \centering
-\includegraphics[]{het_z_by_ind.pdf}
+\includegraphics[]{figures/het_z_by_ind.pdf}
 \caption{Normalized mean observed heterozygosity by location across 200 randomly-sampled individuals}
-\label{fig:spectrum}
+\label{fig:hetmap}
 \end{figure*}
@@
 \end{tableminipage}
-
+\label{table:sampling}
 \end{table*}
 
-\endsupplement
+% latex table generated in R 3.5.1 by xtable 1.8-3 package
+% Fri Mar 22 13:50:47 2019
+\begin{table}[ht]
+\centering
+\caption{\bf T-test results comparing standard deviations of inferred $N_{e}$ between spatial and coalescent models, by neighborhood size (NS) and sampling strategy.
+$p$ is the probability that spatial models have higher standard deviations.}
+\begin{tabular}{rllrrrrr}
+  \hline
+sampling & NS range & t & df & $p$ \\ 
+  \hline
+random & 2-20 & 4.2572 & 41.6166 & 0.0001 \\ 
+random & 20-100 & -1.8473 & 171.9905 & 0.9668 \\ 
+random & 100-500 & -2.1297 & 164.3864  & 0.9827 \\ 
+random & 500-1000 & -3.9681 & 147.0497 & 0.9999 \\ 
+point & 2-20 & 7.0802 & 44.3615  & 0.0000 \\ 
+point & 20-100 & -0.2038 & 169.3799 & 0.5806 \\ 
+point & 100-500 & -2.4945 & 152.5000 & 0.9932 \\ 
+point & 500-1000 & -3.8329 & 162.6443& 0.9999 \\ 
+midpoint & 2-20 & 5.9253 & 59.5462 & 0.0000 \\ 
+midpoint & 20-100 & 3.8940 & 171.7005  & 0.0001 \\ 
+midpoint & 100-500 & -2.2764 & 139.5221 & 0.9878 \\ 
+midpoint & 500-1000 & -3.2223 & 165.0792 & 0.9992 \\ 
+   \hline
+\end{tabular}
+\label{table:demography}
+\end{table}
+
+\stopsupplement
 

### 2019-04-11 ms/spaceness.tex
results rewrite, fig 2 update

@@
 \usepackage{amsmath}
-\linespread{1.5}
+\usepackage[nodisplayskipstretch]{setspace}
+\usepackage{makecell} 
+\setstretch{1}
 % Use the documentclass option 'lineno' to view line numbers
@@
 
+\ak{we should start by saying that we don't understand how simple summaries will respond to continuous space model}
 This issue is particularly important for analyses of polygenic selection, because the allele frequency variation caused by population structure is similar to the pattern expected when many sites in the genome are associated with a given phenotype \citep{Bulik-Sullivan2015}.
@@
 the strength of isolation by distance and the spatial distribution of environmental effects on phenotypes -- in which common methods of stratification bias corrections in GWAS are likely to fail.
+
+\section{Materials and Methods}
+\label{sec:materials:methods}
+
+\ak{i moved the following section into M\&M. makes more sense here i think}
 
@@
 By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.
-
-\section{Materials and Methods}
-\label{sec:materials:methods}
-
 \subsection{A Forward-Time Model of Evolution in Continuous Space}
@@
 For all simulations the full genealogy and recombination history of final-generation individuals were stored as tree sequences \citep{Kelleher2018}.
+\ak{add a link to the github repo with code}
 
@@
 the tree sequence had coalesced).
+\ak{maybe worth including a table with some basic runtime results in the supplement?}
 
 \subsection{Genealogical Parameters}
-We recorded the census population size and estimated the variance in number of offspring and generation time for all simulations.
+We recorded the census population size and estimated the variance in number of offspring and generation time \ak{need to explain why this is needed} for all simulations.
 To estimate generation times we stored the age of all parents for 200 timesteps after a 100 generation burn-in and took the mean.
@@
 Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among samples.
-The full distribution of lengths of IBS tracts for each pair of individuals was first calculated with a custom python function.
+The full distribution of lengths of IBS tracts for each pair of individuals was first calculated with a custom python function \ak{point to code on github}.
 We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $1e6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.
@@
 To separate these effects we ran 100 coalescent simulations with constant population size $6.1\times 10^{-3}$ (the mean $N_{e}$ of random-mating SLiM models estimated from $pi$) and fit stairwayplot models using the same script as for our spatial models.
+All coalescent simulations were performed using msprime \citep{Kelleher2016}.
 We then calculated the standard deviation of inferred $N_{e}$ in each stairwayplot model to summarize the degree of fluctuation around the simulated population size, and asked if standard deviations were higher in spatial relative to coalescent models with a one-tailed t-test.
@@
 SNPs with a minor allele frequency less than 0.5\% were excluded from this analysis.
-Phenotype values were meant to roughly reflect the distribution of height across Europe, which has recently been found to be confounded with population structure in large scale GWAS \citep{Berg2018,Sohail2018}.
+Phenotype values were meant to roughly reflect the distribution of height across Europe \ak{how so?
+need more details here.
+Was it matched to mean and variance?}, which has recently been found to be confounded with population structure in large scale GWAS \citep{Berg2018,Sohail2018}.
 Conceptually our approach is similar to that taken in \citep{Mathieson2012}, though here we model fully continuous spatial variation and focus on genome-wide effects rather than low-frequency alleles specifically.
@@
 Last, we simulated a "patchy" environmental influence on phenotypes by selecting 10 random points on the map and drawing phenotypes for all individuals within two map units of any selected point from a normal distribution with mean 2 standard deviations above the rest of the map.
+\ak{in addition to the verbal treatment above I would list the formulae that you are using for the phenotypes here}
 
@@
 \subsection{Genealogical Parameters and Run Times}
-Census population size, generation time, and the variance in number of offspring all vary systematically with $\sigma$ (Figure \ref{fig:genparams}).
+Having implemented our continuous space forward time population genetic simulation, the first thing we were interested in examining was the effect of spatial parameters on population parameters such as the generation time, the census population size, and the variance in offspring number.
+In contrast to intuition based on standard population genetic models, in the context of our non-Wright-Fisher model basic quantitives like the generation time depend on population parameters such as the dispersal distance and population density over the landscape.
+We first examined how the population size, generation time, and  variance in number of offspring vary with neighborhood size which itself is a function of $\sigma$ (Figure \ref{fig:genparams}).
+There are a few things to note.
+First all three quantities are non-linear with respect to neighborhood size.
+Census size largely declines for both the spatial and random mating models, however for spatial models this decline only begins for neighborhood size $\geq 10$.
+By a neighborhood size $\geq 100$ the spatial and random mating models are largely indistinguishable from one another, a sign that our simulations are performing as expected.
 Census sizes range from $\approx 14,000$ at low $\sigma$ in the random mating model to $\approx 10,000$ for both models when neighborhood sizes approach 1,000.
-Generation times vary between 5.2 and 4.9 timesteps per generation in the spatial model, and variance in offspring varies from 2.2 to 2.6.
+
+Generation times similarly show complex behavior with respect to neighborhood sizes, and vary across the parameter range explored between 5.2 and 4.9 timesteps per generation.
+To stress this point further, the generation time here varies because under our non-Wright-Fisher dynamics individuals are free to reproduce more than once over their lifespan and the lifespan of an individual is itself determined by the parameterization of the model.
+Under both the spatial and random mating regimes generation time reaches a nadir at a neighborhood size of $\approx 50$.
+Under the range of neighborhood sizes that we examined here, generation times between the random mating and spatial models are never quite equivalent-- presumably this would cease to be the case at neighborhood sizes higher than we simulated here.
+
+Last we looked at the effect of neighborhood size on the variance in number of offspring in the population-- a key parameter determining the effective population size.
+Surprisingly the spatial and random mating model behave quite differently with respect to the neighborhood size: while the variance in offspring number increases monotonically, or nearly so, under the spatial model, the random mating model actually shows a decline in the variance in offspring number until a neighborhood size $\approx 10$ before it increases and eventually equals what we observe in the spatial case.
 
@@
 
-\subsection{Impacts of Sampling on Summary Statistics}
-For spatial models the mean and variance of all summary statistics other than those involving the mean of the IBS block size distribution was significantly different across sampling strategies (Table \ref{table:sampling}).
-Means were not significantly different across sampling strategies for any statistic under the random mating model.
-Only spatial correlation coefficients had significantly different variance under the random mating model, though in all cases the mean correlation was near zero.
+\subsection{Impacts of Continuous Space on Population Genetic Summary Statistics}
+One of the main goals of this study is to examine the effect of continuous space on canonical summaries of population genetic variation.
+Indeed we have little knowledge to date about how some of our most beloved population genetic summary statistics might be affected by such realistic models.
+Moreover, as we will show, sampling strategies of individuals with respect to space can affect summaries of variation as strongly, if not more so, than the underlying population dynamics.
 
-\subsubsection{Diversity and Site Frequency Spectra}
-Mean within-population genetic distance ($\pi$), the inbreeding coefficient ($F_{is}$) and Tajima's $D$ are inflated at low dispersal distances (Figure \ref{fig:sumstats}, \ref{fig:allsumstats}).
-$\pi$ and $F_{is}$ approach the random mating model at neighborhood sizes of roughly 100.
-Tajima's D varies relatively more across sampling strategies; approaching random mating at $NS\approx65$ with random sampling and $NS>1000$ with midpoint sampling.
+\subsubsection{Site Frequency Spectra and Summaries of Diversity}
+In Figure \ref{fig:sumstats} we examine the effect of varying neighborhood size and sampling strategy on the site frequency spectrum (Figure \ref{fig:sumstats}A) and several canonical population genetic summary statistics (Figure \ref{fig:sumstats}B).
+For populations in continuous space we observe a significant skew in the SFS for smaller neighborhood sizes ($\leq 100$) that is exacerbated by mid point and point sampling of individuals (see Figure \ref{fig:samplemap}).
+As expected these effects are strongest for the smallest neighborhood sizes examined and are consistently in the direction of an enrichment of intermediate frequency variants in comparison to the standard neutral expectation.
+This can be seen succinctly in the response of Tajima's \textit{D} to variation in neighborhood sizes and sampling regime shown in Figure \ref{fig:sumstats}B, whereby we see \textit{D} turning quite positive at small neighborhood sizes.
+This effect is particularly strong for the midpoint and point sampling regimes, but also occurs under random sampling.
+Notably, the point at which Tajima's $D$ approaches 0 differs strongly across sampling strategies -- varying from a neighborhood size of $\approx50$ for random sampling to $\geq1000$ for midpoint sampling.
 
-The inflation in $F_{is}$ appears to be caused by a deficit of heterozygous individuals in low-dispersal simulations -- a continuous-space version of the Wahlund effect \citep{Wahlund1928}.
-However under midpoint sampling observed heterozygosity is inflated over the random mating expectation, which we confirmed is caused by a higher proportion of heterozygotes in the middle of the landscape (Figure S\ref{fig:hetmap}).
-\cite{Shirk2014} observed a similar excess of heterozygosity in the middle of the landscape when simulating under a lattice model.
+One of the most commonly used summaries of variation is Tajima's estimator of the neutral mutation rate $\widehat{\theta}_{\pi}$.
+As can be seen in Figure \ref{fig:sumstats}B $\widehat{\theta}_{\pi}$ can vary as much as nearly three-fold for the smallest neighborhood sizes between the random mating and spatial models depending on the sampling strategies.
+Each sampling strategy approaches the random mating expectation at its own rate, but by neighborhood size of $\approx 100$ all models are equivalent.
+These results are to be expected as at small neighborhood sizes the time to the most recent common ancestor for the population should be pushed back-- i.e.
+it takes longer for lineages to coalesce if dispersal is limited.
+Interestingly the rate of approach to random mating expectations across sampling strategies is reversed relative to that observed in Tajima's D -- midpoint sampling reaches random mating expectations at neighborhood sizes $\approx50$ while random sampling is inflated until neighborhood size $/approx$ 100.
+In effect this means that any estimates of population size based on expected heterozygosity from populations with limited dispersal should be treated with caution.
 
-The site frequency spectrum showed minimal differences across sampling strategies for random mating models (Supplementary Figure \ref{fig:allsfs}).
-Spatial models had fewer low frequency alleles and more mid-frequency alleles than random mating models (Figure \ref{fig:sumstats}).
-This effect is relatively weak when sampling is random and strongest for midpoint sampling.
+Patterns from observed heterozygosity and its derivative $F_{is}$ also depend heavily on neighborhood size under spatial models as well as the sampling scheme.
+$F_{is}$ is inflated above the expectation across most of the parameter space examined and across all sampling strategies.
+This effect is caused by a deficit of heterozygous individuals in low-dispersal simulations -- a continuous-space version of the Wahlund effect \citep{Wahlund1928}.
+Indeed for random sampling under the spatial model $F_{is}$ does not approach the random mating equivalent until neighborhood sizes of nearly 1000.
+The patterns for raw observed heterozygosity are more complex.
+Under midpoint sampling observed heterozygosity is inflated even over the random mating expectation, as a result of the a higher proportion of heterozygotesn occurring in the middle of the landscape (Figure S\ref{fig:hetmap}).
+This squares with a report from \cite{Shirk2014} who observed a similar excess of heterozygosity in the middle of the landscape when simulating under a lattice model.
 
-\subsubsection{Haplotype Block Lengths}
-Trends in pairwise haplotype block sharing parallel those in SNP-based diversity estimates (Figure \ref{fig:sumstats}, Supplementary Figure \ref{fig:allsumstats}).
-At low dispersal the distribution of IBS block lengths in a set of samples is shifted towards smaller values -- resulting in lower means and fewer long IBS blocks.
-The variance and skew of the full IBS tract distribution vary across sampling strategies, but are similar to random-mating models under random sampling.
+Trends in pairwise haplotype block sharing parallel those in allele-frequency-based diversity estimates (Figure \ref{fig:sumstats}, Supplementary Figure \ref{fig:allsumstats}).
+At low dispersal the distribution of IBS block lengths in a set of samples is shifted towards smaller values with respect to the random mating expectation-- resulting in lower means and fewer long IBS blocks.
+The variance and skew of the distribution of haplotype block lengths are only minorly affected by neighborhood size in our simulations when calculated across all pairs of individuals; however, they are strongly dependent on sampling regimes.
+For example, the number of long haplotype blocks declines as neighborhood size increases under midpoint sampling but changes very little across neighborhood sizes under point or random sampling.
+Thus sampling strategies with respect to geography will affect conclusions drawn from haplotype length distributions quite dramatically.
 
-\subsubsection{Spatial Correlations}
-The correlation coefficient between spatial distance and pairwise genetic differentiation ($D_{xy}$) is positive and declines as dispersal increases, as expected under the theory developed by \citep{Rousset1997} and others.
+\subsubsection{Correlations of summary statistics with geographic distance}
+Correlating population genetic summaries such as $F_{st}$ against geographic distance has shown great utility in empirical population genetics \citep{Rousset1997}.
+As we know the exact locations of individuals that are sampled in our simulated populations we have examined the relationship of geographic distance among samples with a number of our summary statistics (Figure \ref{fig:sumstats} and Figure \ref{fig:allsumstats}).
+Starting with a measure of population differentiation,$D_{xy}$, we observe a positive correlation with distance that declines as dispersal increases, as expected under the theory developed by \citep{Rousset1997} and others.
+This relationship varies across sampling strategies of course, with the weakest correlations observed for midpoint sampling.
+Moreover there is clearly strong signal for  neighborhood size in the strength of the correlation between $D_{xy}$ and geographic distance.
+
+We next turn our attention to the effect of geographic distance on haplotype block length sharing.
 As in \citep{Ringbauer2017} and \citep{Baharian2016} we found that the pairwise distribution of haplotype block lengths is more strongly left-skewed under limited dispersal.
-This is reflected in negative correlation coefficients between spatial distance and the mean, variance, skew, and count of long blocks from the pairwise distribution of identical-by-state block lengths.
-Of these summaries the mean of the IBS tract length distribution has very weak sigmal for $\sigma$, likely because it is heavily influenced by the small number of very long IBS tracts.
-Though the count of long IBS blocks and the skew of the full distribution are low and similar to random mating when taken across all samples, the correlation coefficients between spatial distance and the skew or count of long IBS blocks on a pairwise basis varies predictably by neighborhood size.
-All spatial correlations approach the random mating model near neighborhood sizes of 1000 under either point or random sampling.
-Midpoint sampling causes weaker correlations in all sampling schemes.
+This is reflected in negative correlation coefficients between spatial distance and the mean, variance, skew, and count of long blocks from the pairwise distribution of identical-by-state block lengths (Figure \ref{fig:sumstats} and Figure \ref{fig:allsumstats}).
+Of these summaries the mean of the IBS tract length distribution is only weakly affected by neighborhood size, likely because it is heavily influenced by the small number of very long IBS tracts.
+In contrast the count of long IBS blocks and the skew of the pairwise IBS block distribution are strongly dependent on distance among individuals, and the magnitude of this correlation declines predictably with neighborhood size.
+In all spatial correlations random and point sampling are similarly correlated with space across neighborhood sizes, but midpoint sampling causes weaker correlations because it incorporates less genetic and geographic distance than the full sample.
 
-\subsection{Demographic Modeling}
-Demographic models from Stairwayplot tended to infer patterns of ancient population increases and recent declines when neighborhood sizes were below 20 under all sampling strategies (Figure~\ref{fig:demography}).
-Inflated past population sizes were common in both point and random sampling, demonstrating that the relatively minor shift in the site frequency spectrum observed with random sampling is enough to alter demographic estimates.
-Inference of severe population bottlenecks was also common at neighborhood sizes under 100 for midpoint and point sampling strategies.
+\subsubsection{Effects of Sampling on Random Mating and Spatial Models}
+To summarize the effect of sampling strategy on genetic diversity in random mating and spatial models we also asked if summary statistics varied significantly across sampling regimes.
+In table \ref{table:sampling} we show that for spatial models the mean and variance of nearly all summary statistics was significantly different across sampling strategies (Table \ref{table:sampling}) for spatial models, but not for random mating models.
+Essentially, sampling strategy can be safely ignored when the population is randomly mating but will shape estimates of genetic diversity in any population with limited dispersal.
+
+\subsection{Effects of Space on Demographic Inference}
+One of the most important uses for population genetic data is inferring demographic history of populations.
+As demonstrated above, the site frequency spectrum varies across neighborhood sizes and sampling strategies.
+Does this variation lead to different inferences of past population sizes?
+To ask this we inferred population size histories from samples drawn from our simulated populations using a popular software package that uses the SFS as its information, Stairwayplot \citep{Liu2015}.
+
+In Figure \ref{fig:demography} we show inferred population size histories from our simulations as a function of neighborhood size binned in to one of four categories.
+In general we observe that demographic models from Stairwayplot tended to infer patterns of ancient population increases and recent declines when neighborhood sizes were below 20 under all sampling strategies (Figure~\ref{fig:demography}).
+This is consistent with our observations of the SFS from which Stairwayplot is doing its inference.
+Inflated past population sizes were seen in both point and random sampling, demonstrating that the relatively minor shift in the site frequency spectrum observed among sampling regimes is enough to alter demographic estimates.
+More alarmingly, inference of severe population bottlenecks was  common at neighborhood sizes under 100 for midpoint and point sampling strategies.
 Above neighborhood sizes of 100 the average inferred demography across all simulations was relatively accurate, with minor fluctuations slightly above the expected variance $Ne$.
-However individual model fits were highly variable and often inferred five-fold or greater population fluctuations even in high-dispersal simulations.
+While that is so individual model fits were highly variable and often inferred five-fold or greater population fluctuations even in high-dispersal simulations.
 
-Standard deviations were significantly greater in all sampling strategies for neighborhood sizes under 20 and for midpoint sampling with neighborhood sizes 20-100, but all other sets performed similarly to the coalescent model (Table \ref{table:demography}).
+To test whether the variation we observed in inferred demographic histories from Stairwayplot was the result of spatial effects in our simulations rather than the behavior of the optimization routine we compared standard deviations of inferred population sizes in each sampling/neighborhood size bin with those returned by an equilibrium coalescent simulation.
+Were these to be equal we could assume that the variation were purely as to be expected under the noise inherent in the genealogical process.
+Instead we found that standard deviations were significantly greater in all sampling strategies for neighborhood sizes under 20 and for midpoint sampling with neighborhood sizes 20-100, but all other sets performed similarly to the coalescent model (Table \ref{table:demography}).
 In summary, spatial mate choice and dispersal causes strong bias in SFS-based demographic estimates for neighborhood sizes below 20 or when sampling is clustered, but otherwise any biases are within the range of variability regularly inferred by Stairwayplot.
-In empirical data, it may be useful to run 10 or more replicate analyses on the same data and take a rolling average, which appears relatively robust for our simulations.
+This underscores the fact that some \textit{a priori} knowledge about the population dynamics at play will be important to interpreting results of demographic estimation routines.
 
 \subsection{GWAS}
-Spatially correlated phenotypes in populations with low neighborhood size cause uncorrected linear-regression GWAS to report significant associations at most genome-wide SNPs \ref{fig:gwas}.
-When PCA positions are included as covariates the vast majority of SNPs no longer pass a 5\% FDR significance threshold, but up to 1000 genome-wide significant hits (out of an average of 132,000 SNPs) were observed even in relatively high dispersal simulations for the corner and clinal distributions.
-Corner and patchy phenotype distributions produced the largest number of significant SNPs, and patchy phenotypes in particular showed little difference in the count of significant SNPs after PC correction.
+For simulations with neighborhood size under 100 spatial variation in the environment causes uncorrected GWAS to infer significant associations with purely environmental phenotypes at most genome-wide SNPs (Figure \ref{fig:gwas}).
+This effect is particularly strong for clinal and corner environments, for which the lowest dispersal levels cause over 95\% of segregating sites in the sample to return significant associations.
+Patchy environmental distributions, which result in fewer extreme phenotype values in the sample (Figure \ref{fig:gwas}A), cause fewer false-positives overall but still produce up to 1000 spurious associations.
+Notably less than 1\% of simulations with nonspatial environments returned any significant associations (Figure \ref{fig:gwas}C), demonstrating that this effect is caused specifically by the interaction of population structure and spatial variation in the environment rather than by population structure itself.
 
-For clinal phenotypes PCA covariates perform well in correcting for population stratification at low dispersal distances but the correction breaks down above neighborhood sizes of $\approx250$.
-This pattern appears to reflect a loss of descriptive power in the first ten axes of the PCA, which describe less of the total variation in the data as neighborhood size increases (see Figure 4B-C).
+When PCA positions are included as covariates the vast majority of SNPs no longer pass a 5\% FDR significance threshold, but hundreds of genome-wide significant hits (out of an average of 132,000 SNPs) were observed even in relatively high dispersal simulations for the corner and clinal distributions (Figure \ref{fig:gwas}C).
+In most cases the $p$ values for these associations were significant after FDR correction but would not pass the threshold for significance under the more conservative Bonferroni correction, suggesting that the most strongly associated variants from many GWAS of mono- or oligogenic traits are robust to this variety of stratification bias.
+Though patchy environmental distributions produce fewer false positive associations overall, PC correction appears less effective as a correction for in this case -- dropping the count of significant sites by roughly half rather than $>95\%$ as in clinal or corner distributions.
 
-Quantile-quantile plots show that $p$ value deflation is most pronounced for the corner phenotype distribution and is particularly strong for simulations with neighborhood sizes of less than 20.
-Quantile-quantile plots of clinal and nonspatial phenotypes also show evidence of $p$ value $inflation$ (i.e.
-points below the 1:1 line), which is observed only when PCA covariates are included in the analysis (see Supplementary Figure X for plots of uncorrected GWAS results).
+Clinal environments cause an interesting pattern in the count of significant sites after PC correction: at low neighborhood sizes the correction appears to work well and removes nearly all significant associations, but at neighborhood sizes above $\approx250$ the number of significant SNPs increases to between 10 and 400 (Figure \ref{fig:gwas}.
+This appears to reflect a loss of descriptive power in the PCA -- as neighborhood size increases, the total proportion of variance explained by the first 10 PC axes declines from roughly 0.1 to 0.04 (Figure \ref{gwas}B).
+Essentially, PCA seems unable to effectively summarize the weak population structure present in large-neighborhood simulations, but these populations continue to have enough spatial structure to create significant correlations between genotypes and the environment.
+A similar process can also be seen in the corner phenotype distribution, in which the count of significant SNPs initially declines as neighborhood size increases and then increases at approximately the point at which the PCA reaches its nadir.
+
+In addition to examining the count of significant SNPs we also explored the degree of genome-wide inflation of test statistics by plotting quantile-quantile plots for all simulations and phenotype distributions (Figure \ref{fig:gwas}D).
+For clinal environments $-log_{10}(p)$ values are most inflated when neighborhood sizes are large, consistent with the pattern observed in the count of significant associationsa after PC regression.
+In contrast corner and patchy environments cause the greatest inflation in $-log_{10}(p)$ at neighborhood sizes $<100$, which likely reflects the inability of PCA to account for fine-scale structure caused by very limited dispersal.
+We also observed that PC regression appears to cause some degree of overcorrection for all phenotype distributions, visible in Figure \ref{fig:gwas}D as points falling below the 1:1 line.
+This suggests that PC regression is removing some true signal of association in our simulations.
+This result is consistent with a recent empirical study of heritability in human height and body mass index, which found that increasing the number of PC axes used as covariates caused the total proportion of variance explained by SNPs to decline from $\approx0.8$ to $\approx0.75$ \citep{Wainschtein2019}.
+Indeed SNPs with minor alleles frequencies of 0.001 - 0.01, which are expected to reflect fine scale population structure \citep{Mathieson2012,Novembre2009}, are estimated to explain \textit{negative} proportions of the total phenotypic variance in \citep{Wainschtein2019}.
 
@@
 
-\subsection{Sampling}
+Patterns of genetic variation are influenced by the fact that organisms are more likely, on average, to reproduce with others of their species that are geographically proximate, in particular when dispersal distances are short.
+While historically some analytical progress has been made in describing the influence of continuous space on population genetics (e.g.
+\cite{Wright1943,Rousset1997}), the theoretical work is challenging and often untenable for realistic biological models.
+Instead we here use efficient forward time population genetic simulations to describe the myriad influence of space on genetic variation.
+In particular we examine three axes of variation across four different sampling strategies-- 1) population genetic summary statistics, 2) inference of population size history, and 3) the consequences on genome-wide association studies (GWAS).
+We are in particular interested in asking how our empirical inferences from data might be affected by spatial processes.
+As we show below the answers seems to be- often space matters, both because of how populations are sampled with respect to space, and because of the inherent dispersal properties of those populations.
+
+\subsection{Effects of Dispersal}
+Recap the general effects of summary stats-- skew in SFS, inflation of pi for low dispersal. What did we know about this before i.e from literature?
+
+\subsection{Effects of Sampling}
 
@@
 
+\subsection{Where are organisms on this spectrum?}
+Need concrete examples here to relate these simulation results to reality. Where are humans? Mountain goats? Sheep? Does this mean we now need to reinterpret previous results?
+
+\subsection{speculation}
+As we have shown, patterns of population genetic summary statistics contain information about dispersal.
+We imagine that combinations of such summaries when regressed against geographic distance might be sufficient for the construction of supervised machine learning regressors (e.g.
+cite Schrider and Kern 2018) for the accurate estimation of dispersal from genetic data.
+
+\ak{other things to say?}
+
 \section{Data Availability}
+\ak{github repo}
 
 \section{Acknowledgements}
-\newpage
-\newpage
+We thank folks for reading and thinking about this manuscript.
+We thank the Hearth for having such good nearby coffee and members of the 2018-2019 UO fluid dynamics seminar for many helpful discussions.
+CJB and ADK were supported by NIH award R01GM117241.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Figure and Tables %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \afterpage{\clearpage}
+% latex table generated in R 3.5.1 by xtable 1.8-3 package
+% Thu Apr 11 13:11:59 2019
+\begin{table}[ht]
+\centering
+\caption{\bf Neighborhood size estimates from empirical studies.}
+\begin{tabular}{rllrrll}
+  \hline
+ Species & Description & \makecell[l]{Neighborhood\\Size} & Method & Citation \\ 
+  \hline
+  \makecell[l]{\textit{Ipomopsis}\\\textit{aggregata}} & flowering plant & 12.60-37.80 & Genetic & Campbell et al. 1992\\ 
+  \makecell[l]{\textit{Linanthus}\\\textit{parryae}} & flowering plant & 14-27 & Genetic & Wright 1943\\ 
+  \makecell[l]{\textit{Borrichia}\\\textit{frutescens}} & salt marsh plant & 20-30 & Genetic+Survey & Antlfinger 1982 \\ 
+  \makecell[l]{\textit{Oreamnos}\\\textit{americanus}} & mountain goat & 36-100 & Genetic & Shirk and Cushman, 2014 \\ 
+  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{Gainj- and Kalam- \\speaking people,\\Papua New Guinea} & 40-213 & Genetic & Rousset et al. 1997\\ 
+  \makecell[1]{\textit{Formica sp.}} & colonial ants & 50-100 & Genetic & Pamilo 1982 \\ 
+  \makecell[l]{\textit{Astrocaryum}\\\textit{mexicanum}} & palm tree & 102-895 & Genetic+survey & Eguiarte et al. 1993 \\ 
+  \makecell[l]{\textit{Spermophilus}\\\textit{mollis}} & ground squirrel & 204-480 & Genetic+Survey & Antolin et al. 2001\\
+  \makecell[l]{\textit{Sceloperus}\\\textit{olivaceus}} & lizard & 225-270 & Survey & Kerster 1964 \\ 
+  \makecell[l]{\textit{Dieffenbachia}\\\textit{longispatha}} & \makecell[l]{beetle-pollinated\\colonial herb} & 227-611 & Survey & Young 1988 \\ 
+  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{Gainj- and Kalam- \\speaking people,\\Papua New Guinea} & 410 & Survey & Rousset et al. 1997 \\ 
+  \makecell[l]{\textit{Quercus}\\\textit{laevis}} & Oak tree & $\geq440$ & Genetic & Berg and Hamrick, 1994 \\ 
+  \makecell[l]{\textit{Drosophila}\\\textit{pseudoobscura}} & fruit fly & 500-1000 & Genetic & Wright et al. 1941 \\ 
+  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{NE Europe} & 1342-5425 & Genetic & Ringbauer et al. 2018 \\ 
+  \makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 240000-240000 & Survey & Rousset et al. 1997\\ 
+  \makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 360000-360000 & Genetic & Rousset et al. 1997\\ 
+   \hline
+\end{tabular}
+\label{table:NStable}
+\end{table}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Supplement %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Supplementary Figures and Tables}
@@
 \end{figure*}
-
-%\afterpage{\clearpage}
-%\begin{figure*}[p]
-%\centering
-%\includegraphics[width=\textwidth]{stairwayplot_by_sigma_facet_sampling_by_bin.pdf}
-%\caption{Inferred demographic histories for spatial SLiM simulations from Stairwayplot by %sampling scheme and neighborhood size (NS) range.
-Lines shown are the median inferred %$N_{e}$ across 100 bootstrap replicates for each simulation.
-Colors are scaled to Wright's %neighborhood size.
-The solid black line is the average census population size, and the %dotted line is the average variance effective size ($N_{e}/var(n offspring)$).}
-%\label{fig:spectrum}
-%\end{figure*}
 
@@
 \end{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Genetics Submission Guidelines}
-Manuscripts submitted to \textit{GENETICS} should contain a clear description of the experimental design in sufficient detail so that the experimental analysis could be repeated by another scientist.
-If the level of detail necessary to explain the protocol goes beyond two paragraphs, give a short description in the main body of the paper and prepare a detailed description for supporting information.
-For example, details would include indicating how many individuals were used, and if applicable how individuals or groups were combined for analysis.
-If working with mutants indicate how many independent mutants were isolated.
-If working with populations indicate how samples were collected and whether they were random with respect to the target population.
-
-\subsection{Statistical Analysis}
-
-It is important to indicate what statistical analysis has been performed; not just the name of the software and options selected, but the method and model applied.
-In the case of many genes being examined simultaneously, or many phenotypes, a multiple comparison correction should be used to control the type I error rate, or a rationale for not applying a correction must be provided.
-The type of correction applied should be clearly stated.
-It should also be clear whether the p-values reported are raw, or after correction.
-Corrected p-values are often appropriate, but raw p-values should be available in the supporting materials so that others may perform their own corrections.
-In large scale data exploration studies (e.g.
-genome wide expression studies) a clear and complete description of the replication structure must be provided.
-
-\subsection{Data Availability}
-
-At the end of the Materials and Methods section, include a statement on reagent and data availability.
-Please read the Data and Reagent Policy before writing the statement.
-Make sure to list the accession numbers or DOIs of any data you have placed in public repositories.
-List the file names and descriptions of any data you will upload as supplemental information.
-The statement should also include any applicable IRB numbers.
-You may include specifications for how to properly acknowledge or cite the data.
-
-For example: Strains are available upon request.
-File S1 contains detailed descriptions of all supplemental files.
-File S2 contains SNP ID numbers and locations.
-File S3 contains genotypes for each individual.
-Sequence data are available at GenBank and the accession numbers are listed in File S3.
-Gene expression data are available at GEO with the accession number: GDS1234.
-Code used to generate the simulated data is provided in file S4.
-
-\section{Results and Discussion}
-
-The results and discussion should not be repetitive.
-The results section should give a factual presentation of the data and all tables and figures should be referenced; the discussion should not summarize the results but provide an interpretation of the results, and should clearly delineate between the findings of the particular study and the possible impact of those findings in a larger context.
-Authors are encouraged to cite recent work relevant to their interpretations.
-Present and discuss results only once, not in both the Results and Discussion sections.
-It is sometimes acceptable to combine results and discussion.
-The text should be as succinct as possible.
-Heed Strunk and White's dictum: "Omit needless words!"
-
-\section{Additional guidelines}
-
-\subsection{Numbers} In the text, write out numbers nine or less except as part of a date, a fraction or decimal, a percentage, or a unit of measurement.
-Use Arabic numbers for those larger than nine, except as the first word of a sentence; however, try to avoid starting a sentence with such a number.
-
-\subsection{Units} Use abbreviations of the customary units of measurement only when they are preceded by a number: "3 min" but "several minutes".
-Write "percent" as one word, except when used with a number: "several percent" but "75\%." To indicate temperature in centigrade, use ° (for example, 37°); include a letter after the degree symbol only when some other scale is intended (for example, 45°K).
-
-\subsection{Nomenclature and Italicization} Italicize names of organisms even when  when the species is not indicated.
-Italicize the first three letters of the names of restriction enzyme cleavage sites, as in HindIII.
-Write the names of strains in roman except when incorporating specific genotypic designations.
-Italicize genotype names and symbols, including all components of alleles, but not when the name of a gene is the same as the name of an enzyme.
-Do not use "+" to indicate wild type.
-Carefully distinguish between genotype (italicized) and phenotype (not italicized) in both the writing and the symbolism.
-
-\subsection{Cross References}
-Use the \verb|\nameref| command with the \verb|\label| command to insert cross-references to section headings.
-For example, a \verb|\label| has been defined in the section \nameref{sec:materials:methods}.
-
-\section{In-text Citations}
-
-Add citations using the \verb|\citep{}| command, for example \citep{neher2013genealogies} or for multiple citations, \citep{neher2013genealogies, rodelsperger2014characterization,Falush16}
-
-\section{Examples of Article Components}
-\label{sec:examples}
-
-The sections below show examples of different header levels, which you can use in the primary sections of the manuscript (Results, Discussion, etc.) to organize your content.
-
-\section{First level section header}
-
-Use this level to group two or more closely related headings in a long article.
-
-\subsection{Second level section header}
-
-Second level section text.
-
-\subsubsection{Third level section header:}
-
-Third level section text. These headings may be numbered, but only when the numbers must be cited in the text.
-
-\section{Figures and Tables}
-
-Figures and Tables should be labelled and referenced in the standard way using the \verb|\label{}| and \verb|\ref{}| commands.
-
-\subsection{Sample Figure}
-
-Figure \ref{fig:spectrum} shows an example figure.
-
-\begin{figure}[htbp!]
-\centering
-\includegraphics[width=\linewidth]{example-figure}
-\caption{Example figure from \url{10.1534/genetics.114.173807}.
-Please include your figures in the manuscript for the review process.
-You can upload figures to Overleaf via the Project menu.
-Upon acceptance, we'll ask for your figure files to be uploaded in any of the following formats: TIFF (.tiff), JPEG (.jpg), Microsoft PowerPoint (.ppt), EPS (.eps), or Adobe Illustrator (.ai).
-Images should be a minimum of 300 dpi in resolution and 500 dpi minimum if line art images.
-RGB, CMYK, and Grayscale are all acceptable.
-Halftones should be high contrast with sharp detail, because some loss of detail and contrast is inevitable in the production process.
-Figures should be 10-20 cm in width and 1-25 cm in height.
-Graph axes must be exactly perpendicular and all lines of equal density.
-Label multiple figure parts with A, B, etc.
-in bolded type, and use Arrows and numbers to draw attention to areas you want to highlight.
-Legends should start with a brief title and should be a self-contained description of the content of the figure that provides enough detail to fully understand the data presented.
-All conventional symbols used to indicate figure data points are available for typesetting; unconventional symbols should not be used.
-Italicize all mathematical variables (both in the figure legend and figure) , genotypes, and additional symbols that are normally italicized.
-}%
-\label{fig:spectrum}
-\end{figure}
-
-\subsection{Sample Video}
-
-Figure \ref{video:spectrum} shows how to include a video in your manuscript.
-
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=\linewidth]{example-figure}
-\caption{Example movie (the figure file above is used as a placeholder for this example).
-\textit{GENETICS} supports video and movie files that can be linked from any portion of the article - including the abstract.
-Acceptable formats include .asf, avi, .wav, and all types of Windows Media files.
-}%
-\label{video:spectrum}
-\end{figure}
-
-\subsection{Sample Table}
-
-Table \ref{tab:shape-functions} shows an example table.
-Avoid shading, color type, line drawings, graphics, or other illustrations within tables.
-Use tables for data only; present drawings, graphics, and illustrations as separate figures.
-Histograms should not be used to present data that can be captured easily in text or small tables, as they take up much more space.
-
-Tables numbers are given in Arabic numerals. Tables should not be numbered 1A, 1B, etc., but if necessary, interior parts of the table can be labeled A, B, etc. for easy reference in the text.
-
-\begin{table*}[htbp]
-\centering
-\caption{\bf Students and their grades}
-\begin{tableminipage}{\textwidth}
-\begin{tabularx}{\textwidth}{XXXX}
-\hline
-Student & Grade\footnote{This is an example of a footnote in a table.
-Lowercase, superscript italic letters (a, b, c, etc.) are used by default.
-You can also use *, **, and *** to indicate conventional levels of statistical significance, explained below the table.} & Rank & Notes \\
-\hline
-Alice & 82\% & 1 & Performed very well.\\
-Bob & 65\% & 3 & Not up to his usual standard.\\
-Charlie & 73\% & 2 & A good attempt.\\
-\hline
-\end{tabularx}
-  \label{tab:shape-functions}
-\end{tableminipage}
-\end{table*}
-
-\section{Sample Equation}
-
-Let $X_1, X_2, \ldots, X_n$ be a sequence of independent and identically distributed random variables with $\text{E}[X_i] = \mu$ and $\text{Var}[X_i] = \sigma^2 < \infty$, and let
-\begin{equation}
-S_n = \frac{X_1 + X_2 + \cdots + X_n}{n}
-      = \frac{1}{n}\sum_{i}^{n} X_i
-\label{eq:refname1}
-\end{equation}
-denote their mean. Then as $n$ approaches infinity, the random variables $\sqrt{n}(S_n - \mu)$ converge in distribution to a normal $\mathcal{N}(0, \sigma^2)$.

### 2019-04-12 ms/spaceness.tex
gwas text and fig revision

@@
 \subsection{Association Studies}
-To assess the degree to which spatial structure confounds GWAS we simulated four types of nongenetic phenotype variation for 1000 randomly sampled individuals in each spatial SLiM simulation and conducted a linear-regression GWAS with PC covariates in PLINK \citep{PURCELL2007}.
+To assess the degree to which spatial structure confounds GWAS we simulated four types of nongenetic phenotype variation for 1000 randomly sampled individuals in each spatial SLiM simulation and conducted a linear regression GWAS with PC covariates in PLINK \citep{PURCELL2007}.
 SNPs with a minor allele frequency less than 0.5\% were excluded from this analysis.
@@
 \subsection{GWAS}
-For simulations with neighborhood size under 100 spatial variation in the environment causes uncorrected GWAS to infer significant associations with purely environmental phenotypes at most genome-wide SNPs (Figure \ref{fig:gwas}).
-This effect is particularly strong for clinal and corner environments, for which the lowest dispersal levels cause over 95\% of segregating sites in the sample to return significant associations.
-Patchy environmental distributions, which result in fewer extreme phenotype values in the sample (Figure \ref{fig:gwas}A), cause fewer false-positives overall but still produce up to 1000 spurious associations.
-Notably less than 1\% of simulations with nonspatial environments returned any significant associations (Figure \ref{fig:gwas}C), demonstrating that this effect is caused specifically by the interaction of population structure and spatial variation in the environment rather than by population structure itself.
+To ask what confounding effects spatial genetic variation might have on genome-wide association studies we performed GWAS on our simulations using phenotypes that were determined solely by the environment.
+In general we found that for simulations with limited dispersal, i.e.
+neighborhood size $< 100$, spatial variation in the environment causes GWAS to infer significant associations with purely environmental phenotypes at over 25\% of sampled SNPs if no correction for genetic relatedness among samples is performed (Figure \ref{fig:gwas}).
+This effect is particularly strong for clinal and corner environments, for which the lowest dispersal levels cause over 60\% of SNPs in the sample to return significant associations.
+Patchy environmental distributions, which result in fewer extreme phenotype values in the sample (Figure \ref{fig:gwas}A), cause fewer false-positives overall but still produce spurious associations at roughly 10\% of sites at the lowest neighborhood sizes.
+Notably no simulations with nonspatial environments returned more than one significant association (Figure \ref{fig:gwas}C), demonstrating that this effect is caused specifically by the interaction of population structure and spatial variation in the environment rather than by population structure itself.
 
-When PCA positions are included as covariates the vast majority of SNPs no longer pass a 5\% FDR significance threshold, but hundreds of genome-wide significant hits (out of an average of 132,000 SNPs) were observed even in relatively high dispersal simulations for the corner and clinal distributions (Figure \ref{fig:gwas}C).
+When PCA positions are included as covariates for GWAS the vast majority of SNPs no longer pass a 5\% FDR significance threshold, but up to 1.5\% of SNPs are significantly associated at low dispersal distances under corner and patchy environmental distributions (Figure \ref{fig:gwas}C).
+At neighborhood sizes $> 500$ up to 0.31\% of SNPs were significant for corner and clinal environments.
+Given an average of 132,000 SNPs across simulations after MAF filtering, this translates to up to 382 false-positive associations.
 In most cases the $p$ values for these associations were significant after FDR correction but would not pass the threshold for significance under the more conservative Bonferroni correction, suggesting that the most strongly associated variants from many GWAS of mono- or oligogenic traits are robust to this variety of stratification bias.
-Though patchy environmental distributions produce fewer false positive associations overall, PC correction appears less effective as a correction for in this case -- dropping the count of significant sites by roughly half rather than $>95\%$ as in clinal or corner distributions.
 
-Clinal environments cause an interesting pattern in the count of significant sites after PC correction: at low neighborhood sizes the correction appears to work well and removes nearly all significant associations, but at neighborhood sizes above $\approx250$ the number of significant SNPs increases to between 10 and 400 (Figure \ref{fig:gwas}.
-This appears to reflect a loss of descriptive power in the PCA -- as neighborhood size increases, the total proportion of variance explained by the first 10 PC axes declines from roughly 0.1 to 0.04 (Figure \ref{gwas}B).
+Clinal environments cause an interesting pattern in the count of significant sites after PC correction: at low neighborhood sizes the correction removes nearly all significant associations, but at neighborhood sizes above $\approx250$ the proportion of significant SNPs increases to up to 0.4\% (Figure \ref{fig:gwas}.
+This appears to reflect a loss of descriptive power in the PCA -- as neighborhood size increases, the total proportion of variance explained by the first 10 PC axes declines from roughly 0.1 to 0.04 (Figure \ref{fig:gwas}B).
 Essentially, PCA seems unable to effectively summarize the weak population structure present in large-neighborhood simulations, but these populations continue to have enough spatial structure to create significant correlations between genotypes and the environment.
-A similar process can also be seen in the corner phenotype distribution, in which the count of significant SNPs initially declines as neighborhood size increases and then increases at approximately the point at which the PCA reaches its nadir.
+A similar process can also be seen in the corner phenotype distribution, in which the count of significant SNPs initially declines as neighborhood size increases and then increases at approximately the point at which the proportion of variance explained by PCA approaches its minimum.
 
-In addition to examining the count of significant SNPs we also explored the degree of genome-wide inflation of test statistics by plotting quantile-quantile plots for all simulations and phenotype distributions (Figure \ref{fig:gwas}D).
-For clinal environments $-log_{10}(p)$ values are most inflated when neighborhood sizes are large, consistent with the pattern observed in the count of significant associationsa after PC regression.
+In Figure \ref{fig:gwas}D we present quantile-quantile plots that show the degree of genome-wide inflation of test statistics in PC-corrected GWAS across all simulations and environmental distributions.
+For clinal environments $-log_{10}(p)$ values are most inflated when neighborhood sizes are large, consistent with the pattern observed in the count of significant associations after PC regression.
 In contrast corner and patchy environments cause the greatest inflation in $-log_{10}(p)$ at neighborhood sizes $<100$, which likely reflects the inability of PCA to account for fine-scale structure caused by very limited dispersal.
@@
 This suggests that PC regression is removing some true signal of association in our simulations.
-This result is consistent with a recent empirical study of heritability in human height and body mass index, which found that increasing the number of PC axes used as covariates caused the total proportion of variance explained by SNPs to decline from $\approx0.8$ to $\approx0.75$ \citep{Wainschtein2019}.
-Indeed SNPs with minor alleles frequencies of 0.001 - 0.01, which are expected to reflect fine scale population structure \citep{Mathieson2012,Novembre2009}, are estimated to explain \textit{negative} proportions of the total phenotypic variance in \citep{Wainschtein2019}.
 
 \section{Discussion}
-
 Patterns of genetic variation are influenced by the fact that organisms are more likely, on average, to reproduce with others of their species that are geographically proximate, in particular when dispersal distances are short.
 While historically some analytical progress has been made in describing the influence of continuous space on population genetics (e.g.
-\cite{Wright1943,Rousset1997}), the theoretical work is challenging and often untenable for realistic biological models.
+\cite{Wright1943,Rousset1997,Ringbauer2017,Barton2010}), the theoretical work is challenging and often untenable for realistic biological models.
 Instead we here use efficient forward time population genetic simulations to describe the myriad influence of space on genetic variation.
@@
 \subsection{Effects of Dispersal}
+Limited dispersal inflates estimates of population size from genetic variation,
+
 Recap the general effects of summary stats-- skew in SFS, inflation of pi for low dispersal. What did we know about this before i.e from literature?
@@
 - for clinal traits, include x and y as covariates.
+
+This result is consistent with a recent empirical study of heritability in human height and body mass index, which found that increasing the number of PC axes used as covariates caused the total proportion of variance explained by SNPs to decline from $\approx0.8$ to $\approx0.75$ \citep{Wainschtein2019}.
+Indeed SNPs with minor alleles frequencies of 0.001 - 0.01, which are expected to reflect fine scale population structure \citep{Mathieson2012,Novembre2009}, are estimated to explain \textit{negative} proportions of the total phenotypic variance in \citep{Wainschtein2019}.
 
@@
 The thick line is a rolling mean and thin lines are individual model fits.
-Dashed horizontal lines are the average $N_{e}$ across random-mating SLiM models estimated from $\pi$.}
+Dashed horizontal lines are the average $N_{e}$ across random-mating SLiM models estimated from $\theta_{\pi}$.}
 \label{fig:demography}
@@
 \centering
-\includegraphics[width=\textwidth]{figures/gwas_summary_nsig_qqplots_loglog.pdf}
+\includegraphics[width=\textwidth]{figures/gwas_summary.pdf}
 \caption{Impacts of spatially correlated phenotypes and isolation by distance on linear regression GWAS.
-A: example phenotype distributions and sampling maps, with colors scaled to phenotype values.
+A: example environmental distributions and sampling maps, with colors scaled to phenotype values.
 B: proportion of total variance explained by the first 10 PC axes, by neighborhood size.
-C: numbers of significant SNPs after FDR correction for linear-model GWAS conducted with (top) or without (bottom) PC covariates.
+C: Proportion of significant SNPs after FDR correction for linear-model GWAS conducted with (top) or without (bottom) PC covariates, by neighborhood size.
 D: Quantile-quantile plots of observed $p$ values relative to those expected from a uniform distribution.
@@
 \centering
-\caption{\bf Neighborhood size estimates from empirical studies.}
+\caption{\bf Neighborhood size estimates from empirical studies.
+Survey methods indicate field studies to measure dispersal or population density.
+Genetic methods include both sequence-based analyses and enzyme activity assays or crosses used to estimate allele frequencies.}
 \begin{tabular}{rllrrll}
@@
   \hline
-  \makecell[l]{\textit{Ipomopsis}\\\textit{aggregata}} & flowering plant & 12.60-37.80 & Genetic & Campbell et al. 1992\\ 
-  \makecell[l]{\textit{Linanthus}\\\textit{parryae}} & flowering plant & 14-27 & Genetic & Wright 1943\\ 
-  \makecell[l]{\textit{Borrichia}\\\textit{frutescens}} & salt marsh plant & 20-30 & Genetic+Survey & Antlfinger 1982 \\ 
-  \makecell[l]{\textit{Oreamnos}\\\textit{americanus}} & mountain goat & 36-100 & Genetic & Shirk and Cushman, 2014 \\ 
-  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{Gainj- and Kalam- \\speaking people,\\Papua New Guinea} & 40-213 & Genetic & Rousset et al. 1997\\ 
-  \makecell[1]{\textit{Formica sp.}} & colonial ants & 50-100 & Genetic & Pamilo 1982 \\ 
-  \makecell[l]{\textit{Astrocaryum}\\\textit{mexicanum}} & palm tree & 102-895 & Genetic+survey & Eguiarte et al. 1993 \\ 
-  \makecell[l]{\textit{Spermophilus}\\\textit{mollis}} & ground squirrel & 204-480 & Genetic+Survey & Antolin et al. 2001\\
-  \makecell[l]{\textit{Sceloperus}\\\textit{olivaceus}} & lizard & 225-270 & Survey & Kerster 1964 \\ 
-  \makecell[l]{\textit{Dieffenbachia}\\\textit{longispatha}} & \makecell[l]{beetle-pollinated\\colonial herb} & 227-611 & Survey & Young 1988 \\ 
-  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{Gainj- and Kalam- \\speaking people,\\Papua New Guinea} & 410 & Survey & Rousset et al. 1997 \\ 
-  \makecell[l]{\textit{Quercus}\\\textit{laevis}} & Oak tree & $\geq440$ & Genetic & Berg and Hamrick, 1994 \\ 
-  \makecell[l]{\textit{Drosophila}\\\textit{pseudoobscura}} & fruit fly & 500-1000 & Genetic & Wright et al. 1941 \\ 
-  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{NE Europe} & 1342-5425 & Genetic & Ringbauer et al. 2018 \\ 
-  \makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 240000-240000 & Survey & Rousset et al. 1997\\ 
-  \makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 360000-360000 & Genetic & Rousset et al. 1997\\ 
+  \makecell[l]{\textit{Ipomopsis}\\\textit{aggregata}} & flowering plant & 12.60 - 37.80 & Genetic & \citep{Campbell1992}\\
+  \makecell[l]{\textit{Linanthus}\\\textit{parryae}} & flowering plant & 14 - 27 & Genetic & \citep{Wright1943}\\ 
+  \makecell[l]{\textit{Borrichia}\\\textit{frutescens}} & salt marsh plant & 20 - 30 & Genetic+Survey & \citep{Antlfinger1982} \\ 
+  \makecell[l]{\textit{Oreamnos}\\\textit{americanus}} & mountain goat & 36 - 100 & Genetic & \citep{Shirk2014} \\ 
+  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{Gainj- and Kalam- \\speaking people,\\Papua New Guinea} & 40 - 213 & Genetic & \citep{Rousset1997}\\ 
+  \makecell[l]{\textit{Formica sp.}} & colonial ants & 50 - 100 & Genetic & \citep{Pamilo1983} \\ 
+  \makecell[l]{\textit{Astrocaryum}\\\textit{mexicanum}} & palm tree & 102 - 895 & Genetic+survey & \citep{Eguiarte1993} \\ 
+  \makecell[l]{\textit{Spermophilus}\\\textit{mollis}} & ground squirrel & 204 - 480 & Genetic+Survey & \citep{Antolin2001}\\
+  \makecell[l]{\textit{Sceloperus}\\\textit{olivaceus}} & lizard & 225 - 270 & Survey & \citep{Kerster1964} \\ 
+  \makecell[l]{\textit{Dieffenbachia}\\\textit{longispatha}} & \makecell[l]{beetle-pollinated\\colonial herb} & 227 - 611 & Survey & \citep{Young1988} \\ 
+  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{Gainj- and Kalam- \\speaking people,\\Papua New Guinea} & 410 & Survey & \citep{Rousset1997} \\ 
+  \makecell[l]{\textit{Quercus}\\\textit{laevis}} & Oak tree & $>440$ & Genetic & \citep{Berg1995} \\ 
+  \makecell[l]{\textit{Drosophila}\\\textit{pseudoobscura}} & fruit fly & 500 - 1,000 & Survey+Crosses & \citep{Wright1946} \\ 
+  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{POPRES data\\NE Europe} & 1,342 - 5,425 & Genetic & \citep{Ringbauer2017} \\ 
+  \makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 240,000 & Survey & \citep{Rousset1997}\\ 
+  \makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 360,000 & Genetic & \citep{Rousset1997}\\ 
    \hline

### 2019-04-16 ms/spaceness.tex
full discussion draft

@@
 
-\documentclass[10pt,twoside,lineno]{gsajnl}
+\documentclass[11pt,twoside,lineno]{preprint}
 \usepackage{amsmath}
@@
 \usepackage{makecell} 
-\setstretch{1}
+\usepackage{changepage}
+\setstretch{1.5}
+\NeedsTeXFormat{LaTeX2e}
+\newcommand*{\articletypename}{PREPRINT}
 % Use the documentclass option 'lineno' to view line numbers
-
-\articletype{preprint} % article type
-% {inv} Investigation 
-% {gs} Genomic Selection
-% {goi} Genetics of Immunity 
-% {gos} Genetics of Sex 
-% {mp} Multiparental Populations
 
@@
 %\commentsfalse
-\newcommand{\cjb}[1]{\ifcomments{{\color{blue} \it (#1)}}\else{}\fi}
+\newcommand{\cjb}[1]{\ifcomments{{\color{orange} \it (#1)}}\else{}\fi}
 \newcommand{\plr}[1]{\ifcomments{{\color{purple} \it (#1)}}\else{}\fi}
@@
 \author[$\ast$,1]{C.J. Battey}
-\author[$\ast$]{Peter Ralph}
-\author[$\ast$]{Andrew Kern}
+\author[$\ast$]{Peter L. Ralph}
+\author[$\ast$]{Andrew D. Kern}
 
@@
 
-\runningtitle{SPACENESS} % For use in the footer
+\runningtitle{Effects of Continuous Space} % For use in the footer
 
@@
 Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations.
-As a result many models used to infer selection, phenotypic associations, or past population sizes assume that samples are a random draw from a panmictic population.
-In reality most populations are structured to some degree across space, but the extent to which this model violation biases common genomic analyses is not well known.
+As a result many methods of analyzing genetic data assume that samples are a random draw from a random-mating population, but are applied to real populations that are structured over space by limited dispersal.
 Here we implement a forward-time simulation of evolution in continuous space and use it to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and association studies.
-We find that estimates of past population sizes and many common population genetic summary statistics vary significantly with both dispersal and sampling, with most bias occurring when neighborhood sizes are below 100 and sampling is spatially clustered.
-Variation in the site frequency spectrum and Tajima's D is greater across sampling strategies than over the range of demographic parameters we simulated, pointing to the need to carefully account for sampling when interpreting results.
-Last we show that the combination of isolation by distance and spatially correlated environments causes genome-wide signals of association with purely environmental phenotypes in linear-regression GWAS, and that this bias is not fully controlled by regressing out principal components positions during the analysis.
+We find that estimates of population size trajectories and many common summary statistics vary significantly with dispersal, with most bias occurring when Wright's neighborhood size is less than 100 and sampling is spatially clustered.
+In association studies the combination of spatial variation in the environment and limited dispersal creates spurious signals of genetic association with purely environmental phenotypes across much of the genome, and this bias is only partially corrected by regressing out principal components positions in the analysis.
+Last we describe the genealogical process leading to spatial effects we observed on simulated gentoypes, summarize estimates of neighborhood size from empirical studies across a range of organisms, and discuss the relevance of our simulation results for inference from genetic variation in real organisms.
 \end{abstract}
@@
 Indeed, a near universal rule of reproduction is that individuals mate with other nearby individuals, leading to a positive correlation between genetic and geographic distances.
-This pattern of "isolation by distance" \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, but classical mathematical models to describing are flawed approximations of the underlying process \citep{Felsenstein1975}.
-As a result many studies focus on describing geographic structure as a mixture of random-mating populations connected by migration \citep{Malecot1948,Wright1931}, and analyze variation within clusters of genetic variation inferred by programs like \textit{structure} \citep{Pritchard2000} as random-mating units.
+This pattern of "isolation by distance" \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, but classical mathematical models describing it are flawed approximations of the underlying process \citep{Felsenstein1975}.
+As a result many studies focus on describing geographic structure as a set of random-mating populations connected by migration \citep{Malecot1948,Wright1931}, and analyze variation within clusters of genetic variation inferred by programs like \textit{structure} \citep{Pritchard2000} as random-mating units.
 
@@
 Methods based on the coalescent \citep{Kingman1982} assume that the sampled individuals are a random draw from a random mating population.
-If dispersal or mate selection is limited by space these assumptions are violated -- nearby individuals will be more closely related than an average random pair and drawing multiple samples from the same point on the landscape will not represent a random sample of the genetic variation present in the whole population.
-This is as true at the local scale as it is between the more differentiated units typically analyzed in phylogeographic studies.
+If dispersal or mate selection is limited by space these assumptions are violated -- nearby individuals will be more closely related than an average random pair and drawing multiple samples from the same point on the landscape will represent a biased sample of the genetic variation present in the whole population.
 All downstream inferences drawn from the patterns of relationship inferred among sampled individuals will be subject to bias, but the extent depends on the degree of variation in ancestry across the landscape.
 
-\ak{we should start by saying that we don't understand how simple summaries will respond to continuous space model}
-This issue is particularly important for analyses of polygenic selection, because the allele frequency variation caused by population structure is similar to the pattern expected when many sites in the genome are associated with a given phenotype \citep{Bulik-Sullivan2015}.
-Indeed two recent studies found that previous evidence of polygenic selection on human height in Europe was badly confounded by population structure \citep{Sohail2018,Berg2018}.
+Our understanding of how many common summary statistics respond to variation in dispersal distance is limited.
+Tajima's $D$, for example, is often interpreted as reflecting evidence of selection or past population size changes (CITE \cjb{CJ: what's a good citation here?}).
+However this statistic is essentially a summary of the site frequency spectrum, which itself reflects variation in branch lengths of the underlying genealogy of sampled individuals.
+Because the spatial extent of dispersal and mate selection determines the rate of drift in a continuously distributed population \citep{Maruyama1972}, these genealogical summaries may also reflect variation in the scale of spatial interactions both over time and across populations.
+Similarly, the distribution of haplotype block lengths (i.e.
+stretches of identical sequence) among individuals is sometimes interpreted as reflecting historic demography \citep{Harris2013} or evidence of selection \citep{Garud2015}.
+However as this distribution is also an indirect estimate of the branch lengths of the sample genealogy it is also sensitive to variation in dispersal and mate selection \citep{Ringbauer2017,Baharian2016}.
+
+A related issue is the spatial distribution of sampling effort.
+Though individuals in a random-mating population are exchangeable, those in a continuously structured population are not.
+If sampling is clustered at a series of points on the landscape, as is common in empirical studies because of the practical difficulties of sample collection, samples drawn from a spatially structured population may partially reflect spatial biases in the sampling process rather than in the underlying population.
+In addition, range edges have been observed to create complex patterns of heterozygosity in stepping-stone simulations \citep{Neel2013,Shirk2014}, but the effects of this process on many downstream inference procedures is unknown.
+\cjb{must be more out there on sampling -- any recs for papers to read+cite?}
+
+The issue of fine-scale structure for putatively "within-population" analyses is particularly important for studies of polygenic selection.
+This is because the genome-wide allele frequency variation caused by isolation by distance is similar to that expected when many sites in the genome are associated with a given phenotype \citep{Bulik-Sullivan2015}.
+When population structure and the environment covary across space genome-wide association studies (GWAS) may incorrectly infer phenotypic association at many SNPs that have differentiated by drift \citep{Mathieson2012}.
+Indeed two recent studies found that previous evidence of polygenic selection on human height in Europe was confounded by population structure \citep{Sohail2018,Berg2018}, suggesting that existing methods to correct for population structure in GWAS are insufficient.
+However we have little quantitative idea of the population and environmental parameters that can be expected to lead to bias in GWAS.
 As the scale of sequence data now available for many species allows inference of increasingly fine-scale patterns of selection and demography, understanding how and when subtle spatial structure is likely to bias results is an important task for population genetics.
@@
 Here we describe an implementation of an individual-based model in continuous space that incorporates overlapping generations, a Gaussian dispersal kernel, and density-dependent fitness.
-The model scales to chromosome-scale alignments across tens of thousands of individuals, and outputs the full genealogy and recombination history of all final-generation individuals.
+The model scales to chromosome-scale alignments across tens of thousands of individuals and outputs the full genealogy and recombination history of all final-generation individuals.
 We use this simulation to test how sampling strategy interacts with isolation by distance to cause systematic variation in population genetic summary statistics typically analyzed under discrete population models.
@@
 Here we simulate both the underlying genealogy and phenotypes of a continuously distributed population and seek to identify regions of parameter space -- i.e.
-the strength of isolation by distance and the spatial distribution of environmental effects on phenotypes -- in which common methods of stratification bias corrections in GWAS are likely to fail.
+the strength of isolation by distance and the spatial distribution of environmental effects on phenotypes -- in which common methods employed for GWAS are subject to confounding of environmental and genetic effects on phenotypes.
 
@@
 
-\ak{i moved the following section into M\&M. makes more sense here i think}
-
 \subsection{Modeling Evolution in Continuous Space}
-The best-studied approaches to population genetics in continuous space were developed by \cite{Wright1943} and \cite{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals.
+The best-studied approaches to population genetics in continuous space were developed by \cite{Wright1942,Wright1943} and \cite{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals.
 A key finding of Wright's model is that many important aspects of continuous populations can be described in terms of "neighborhood size" -- the number of potential mates for an individual in a given generation, defined as $4\pi\sigma^2d$, where $\sigma$ is the average dispersal distance and $d$ is population density.
@@
 
-Though some aspects of continuous populations are fairly well described by the Wright and Malécot models, \cite{Felsenstein1975} showed that the assumptions of independent dispersal and Poisson distributed offspring that are the basis of these models are incompatible.
+Though some aspects of continuous populations are well described by the Wright and Malécot models, \cite{Felsenstein1975} showed that the assumptions of independent dispersal and Poisson distributed offspring that are the basis of these models are incompatible.
 Over time, a population meeting them will clump into a small number of geographic clusters occupying only a part of the available range.
@@
 \subsection{Genealogical Parameters}
-We recorded the census population size and estimated the variance in number of offspring and generation time \ak{need to explain why this is needed} for all simulations.
+Because basic elements of our simulations like census population size and generation time are not directly parameterized in our model (instead they emerge from the interaction of parameters including average population density, average lifespan, and the spatial scale of interaction), we first monitored a range of genealogical parameters in all simulations.
+We recorded the census population size and estimated the variance in number of offspring and generation time for all simulations.
 To estimate generation times we stored the age of all parents for 200 timesteps after a 100 generation burn-in and took the mean.
@@
 
+\begin{figure}[htbp]
+\centering
+\includegraphics{figures/sampling_maps.pdf}
+\caption{Example sampling maps for 60 individuals on a 50x50 landscape under midpoint, point, and random sampling strategies.}
+\label{fig:samplemap}
+\end{figure}
+
 \subsection{Summary Statistics}
-We calculated the site frequency spectrum and a set of 18 summary statistics (Supplementary Table \cjb{working on this}) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}.
+We calculated the site frequency spectrum and a set of 18 summary statistics (Table \ref{table:sumstats}) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}.
 Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{is}$), and Tajima's D, as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distances \citep{Rousset1997}, which we summarized as the correlation coefficient between $log10(spatial distance)$ and the proportion of identical base pairs among individuals.
@@
 In preliminary runs we found that inferred population histories were relatively variable even when simulating under a coalescent model, suggesting that some of the differences in demographic estimates for spatial models are caused by the behavior of the optimization algorithm rather than bias in the SFS caused by spatial mate choice and dispersal.
-To separate these effects we ran 100 coalescent simulations with constant population size $6.1\times 10^{-3}$ (the mean $N_{e}$ of random-mating SLiM models estimated from $pi$) and fit stairwayplot models using the same script as for our spatial models.
+To separate these effects we ran 100 coalescent simulations with constant population size $6.1\times 10^{-3}$ (the mean $N_{e}$ of random-mating SLiM models estimated from $\Theta_{\pi}$) and fit stairwayplot models using the same script as for our spatial models.
 All coalescent simulations were performed using msprime \citep{Kelleher2016}.
@@
 SNPs with a minor allele frequency less than 0.5\% were excluded from this analysis.
-Phenotype values were meant to roughly reflect the distribution of height across Europe \ak{how so?
-need more details here.
-Was it matched to mean and variance?}, which has recently been found to be confounded with population structure in large scale GWAS \citep{Berg2018,Sohail2018}.
+Phenotype values were set to vary by two standard deviations across the landscape in a rough approximation of the variation seen in height across Europe, which has recently been found to be confounded with population structure in large scale GWAS \citep{Berg2018,Sohail2018}.
 Conceptually our approach is similar to that taken in \citep{Mathieson2012}, though here we model fully continuous spatial variation and focus on genome-wide effects rather than low-frequency alleles specifically.
@@
 In the first simulation phenotypes for all individuals were drawn from a normal distribution with mean 110 and standard deviation 10.
+\begin{equation}
+    P \sim \mathcal{N}(110,10)
+\end{equation}
 Next we simulated clinal environmental influences on phenotype by drawing the phenotypes from independent normal distributions in which the mean was scaled by an individual's x position such that it varied by 2 standard deviations across the map.
+If $x$ is an individual's x coordinate on a 50x50 landscape, the phenotype is then:
+\begin{equation}
+    P \sim \mathcal{N}(110+\frac{2x}{5},10)
+\end{equation}
 Third, we approximate concentrated environmental effects by drawing phenotypes for individuals with x and y coordinates below 20 from a normal distribution with mean 2 standard deviations above the rest of the map.
+\begin{equation}
+    P\sim
+\begin{cases}
+    \mathcal{N}(130,10),& \text{if } x< 20 \text{ and } y<20\\
+    \mathcal{N}(110,10),& \text{otherwise}\\
+\end{cases}
+\end{equation}
+
 Last, we simulated a "patchy" environmental influence on phenotypes by selecting 10 random points on the map and drawing phenotypes for all individuals within two map units of any selected point from a normal distribution with mean 2 standard deviations above the rest of the map.
-\ak{in addition to the verbal treatment above I would list the formulae that you are using for the phenotypes here}
+
+\begin{equation}
+    P\sim
+\begin{cases}
+    \mathcal{N}(130,10),& \text{if } d < 3\\
+    \mathcal{N}(110,10),& \text{otherwise}\\
+\end{cases}
+\end{equation}
+where $d$ is the distance to the nearest patch.
 
@@
 
-\subsection{Genealogical Parameters and Run Times}
+\subsection{Genealogical Parameters}
 Having implemented our continuous space forward time population genetic simulation, the first thing we were interested in examining was the effect of spatial parameters on population parameters such as the generation time, the census population size, and the variance in offspring number.
-In contrast to intuition based on standard population genetic models, in the context of our non-Wright-Fisher model basic quantitives like the generation time depend on population parameters such as the dispersal distance and population density over the landscape.
-We first examined how the population size, generation time, and  variance in number of offspring vary with neighborhood size which itself is a function of $\sigma$ (Figure \ref{fig:genparams}).
+In contrast to intuition based on standard population genetic models, in the context of our non-Wright-Fisher model basic quantities like the generation time depend on life history parameters such as the dispersal distance and carrying capacity of the landscape.
+We first examined how the population size, generation time, and  variance in number of offspring vary with neighborhood size, which itself is a function of $\sigma$ (Figure \ref{fig:genparams}).
 There are a few things to note.
 First all three quantities are non-linear with respect to neighborhood size.
-Census size largely declines for both the spatial and random mating models, however for spatial models this decline only begins for neighborhood size $\geq 10$.
-By a neighborhood size $\geq 100$ the spatial and random mating models are largely indistinguishable from one another, a sign that our simulations are performing as expected.
+Census size largely declines as neighborhood size increases for both the spatial and random mating models.
+However for spatial models this decline only begins for neighborhood size $\geq 10$.
+By a neighborhood size $\geq 100$ the spatial and random mating models are indistinguishable from one another, a sign that our simulations are performing as expected.
 Census sizes range from $\approx 14,000$ at low $\sigma$ in the random mating model to $\approx 10,000$ for both models when neighborhood sizes approach 1,000.
 
+\begin{figure}[htbp]
+\centering
+\includegraphics{figures/pop_params.pdf}
+\caption{Genealogical parameters from spatial and random mating SLiM simulations, by neighborhood size.}
+\label{fig:genparams}
+\end{figure}
+
 Generation times similarly show complex behavior with respect to neighborhood sizes, and vary across the parameter range explored between 5.2 and 4.9 timesteps per generation.
-To stress this point further, the generation time here varies because under our non-Wright-Fisher dynamics individuals are free to reproduce more than once over their lifespan and the lifespan of an individual is itself determined by the parameterization of the model.
+To stress this point further, the generation time here varies because under our non-Wright-Fisher dynamics individuals are free to reproduce more than once over their lifespan and the lifespan of an individual is only indirectly determined by the parameterization of density-dependent competition the model.
 Under both the spatial and random mating regimes generation time reaches a nadir at a neighborhood size of $\approx 50$.
-Under the range of neighborhood sizes that we examined here, generation times between the random mating and spatial models are never quite equivalent-- presumably this would cease to be the case at neighborhood sizes higher than we simulated here.
+Interestingly under the range of neighborhood sizes that we examined, generation times between the random mating and spatial models are never quite equivalent -- presumably this would cease to be the case at neighborhood sizes higher than we simulated here.
 
-Last we looked at the effect of neighborhood size on the variance in number of offspring in the population-- a key parameter determining the effective population size.
+Last we looked at the effect of neighborhood size on the variance in number of offspring in the population -- a key parameter determining the effective population size.
 Surprisingly the spatial and random mating model behave quite differently with respect to the neighborhood size: while the variance in offspring number increases monotonically, or nearly so, under the spatial model, the random mating model actually shows a decline in the variance in offspring number until a neighborhood size $\approx 10$ before it increases and eventually equals what we observe in the spatial case.
@@
 Indeed we have little knowledge to date about how some of our most beloved population genetic summary statistics might be affected by such realistic models.
-Moreover, as we will show, sampling strategies of individuals with respect to space can affect summaries of variation as strongly, if not more so, than the underlying population dynamics.
+Moreover, as we will show, sampling strategies of individuals with respect to space can affect summaries of variation at least as strongly as the underlying population dynamics.
+
+\begin{figure*}[p]
+\centering
+\includegraphics[width=\textwidth]{figures/sfs_w_sumstats.pdf}
+\caption{Site frequency spectrum (A) and summary statistic distributions (B) by sampling strategy and neighborhood size.}
+\label{fig:sumstats}
+\end{figure*}
 
@@
 In table \ref{table:sampling} we show that for spatial models the mean and variance of nearly all summary statistics was significantly different across sampling strategies (Table \ref{table:sampling}) for spatial models, but not for random mating models.
-Essentially, sampling strategy can be safely ignored when the population is randomly mating but will shape estimates of genetic diversity in any population with limited dispersal.
+In summary, sampling strategy can be safely ignored when the population is randomly mating but will shape estimates of genetic diversity in any population with limited dispersal.
 
@@
 While that is so individual model fits were highly variable and often inferred five-fold or greater population fluctuations even in high-dispersal simulations.
-
 To test whether the variation we observed in inferred demographic histories from Stairwayplot was the result of spatial effects in our simulations rather than the behavior of the optimization routine we compared standard deviations of inferred population sizes in each sampling/neighborhood size bin with those returned by an equilibrium coalescent simulation.
@@
 This underscores the fact that some \textit{a priori} knowledge about the population dynamics at play will be important to interpreting results of demographic estimation routines.
+
+\begin{figure}[p]
+\centering
+\includegraphics[width=\textwidth]{figures/stairwayplot_facet_rollmean.pdf}
+\caption{Inferred demographic histories for spatial SLiM simulations from Stairwayplot, by sampling scheme and neighborhood size (NS) range.
+The thick line is a rolling mean and thin lines are individual model fits.
+Dashed horizontal lines are the average $N_{e}$ across random-mating SLiM models estimated from $\theta_{\pi}$.}
+\label{fig:demography}
+\end{figure}
 
@@
 
-When PCA positions are included as covariates for GWAS the vast majority of SNPs no longer pass a 5\% FDR significance threshold, but up to 1.5\% of SNPs are significantly associated at low dispersal distances under corner and patchy environmental distributions (Figure \ref{fig:gwas}C).
+When PCA positions are included as covariates to control for population structure in GWAS the vast majority of SNPs no longer surpass a 5\% FDR significance threshold, but up to 1.5\% of SNPs are significantly associated at low dispersal distances under corner and patchy environmental distributions (Figure \ref{fig:gwas}C).
 At neighborhood sizes $> 500$ up to 0.31\% of SNPs were significant for corner and clinal environments.
@@
 
-Clinal environments cause an interesting pattern in the count of significant sites after PC correction: at low neighborhood sizes the correction removes nearly all significant associations, but at neighborhood sizes above $\approx250$ the proportion of significant SNPs increases to up to 0.4\% (Figure \ref{fig:gwas}.
+Clinal environments cause an interesting pattern in false positives after PC correction: at low neighborhood sizes the correction removes nearly all significant associations, but at neighborhood sizes above $\approx250$ the proportion of significant SNPs increases to up to 0.4\% (Figure \ref{fig:gwas}.
 This appears to reflect a loss of descriptive power in the PCA -- as neighborhood size increases, the total proportion of variance explained by the first 10 PC axes declines from roughly 0.1 to 0.04 (Figure \ref{fig:gwas}B).
@@
 In contrast corner and patchy environments cause the greatest inflation in $-log_{10}(p)$ at neighborhood sizes $<100$, which likely reflects the inability of PCA to account for fine-scale structure caused by very limited dispersal.
-We also observed that PC regression appears to cause some degree of overcorrection for all phenotype distributions, visible in Figure \ref{fig:gwas}D as points falling below the 1:1 line.
-This suggests that PC regression is removing some true signal of association in our simulations.
+Finally, we observed that PC regression appears to cause some degree of overcorrection for all phenotype distributions, visible in Figure \ref{fig:gwas}D as points falling below the 1:1 line.
+
+\begin{figure*}[p]
+\centering
+\includegraphics[width=\textwidth]{figures/gwas_summary.pdf}
+\caption{Impacts of spatially varying environments and isolation by distance on linear regression GWAS.
+Simulated quantitative phenotypes are determined only by an individual's location and the spatial distribution of environmental factors.
+In \textbf{A} we show the phenotypes and locations of sampled individuals under four environmental distributions, with transparency scaled to phenotype.
+As neighborhood size increases a PCA explains less of the total variation in the data (\textbf{B}).
+Spatially correlated environmental factors cause false positives at a large proportion of SNPs, which is partially but not entirely corrected by adding PC positions as covariates (\textbf{C}).
+Quantile-quantile plots in \textbf{D} show inflation of $-log10(p)$ after PC correction across all simulations and environmental distributions, with colors scaled by the median neighborhood size in each region of q-q space.
+%A: example environmental distributions and sampling maps, with colors scaled to phenotype values.
+B: proportion of total variance explained by the first 10 PC axes, by neighborhood size.
+C: Proportion of significant SNPs after FDR correction for GWAS conducted with (top) or without (bottom) PC covariates.
+D: Quantile-quantile plots of observed $p$ values relative to those expected from a uniform distribution.
+The dotted lines show the 95\% confidence region assuming binomial sampling and an alpha level of 0.05.
+Points above the 1:1 line reflect deflated $p$ values.
+}
+\label{fig:gwas}
+\end{figure*}
 
@@
 Instead we here use efficient forward time population genetic simulations to describe the myriad influence of space on genetic variation.
-In particular we examine three axes of variation across four different sampling strategies-- 1) population genetic summary statistics, 2) inference of population size history, and 3) the consequences on genome-wide association studies (GWAS).
+In particular we examine three axes of variation across three different sampling strategies -- 1) population genetic summary statistics, 2) inference of population size history, and 3) the consequences on genome-wide association studies (GWAS).
 We are in particular interested in asking how our empirical inferences from data might be affected by spatial processes.
-As we show below the answers seems to be- often space matters, both because of how populations are sampled with respect to space, and because of the inherent dispersal properties of those populations.
+As we show below the answers seems to be - often space matters, both because of how populations are sampled with respect to space, and because of the inherent dispersal properties of those populations.
 
 \subsection{Effects of Dispersal}
-Limited dispersal inflates estimates of population size from genetic variation,
+Limited dispersal inflates effective population size, creates correlations between genetic and spatial distances, and introduces subtle biases in the site frequency spectrum that are reflected in a positive Tajima's $D$ (Figure \ref{fig:sumstats}).
+At the extreme low end of dispersal distance this can result in an up to three-fold increase in genetic diversity relative to random-mating expectations.
+These effects are strongest when neighborhood sizes are below 100, but in combination with the effects of nonrandom sampling they can persist up to neighborhood sizes of at least 1000 (e.g.
+inflation in Tajima's $D$ and observed heterozygosity under midpoint sampling).
+This general pattern is similar to that predicted by the original analytic model of \cite{Wright1943}, which predicts that populations with neighborhood sizes under 100 will differ substantially from random mating, while those above 10,000 will be nearly indistinguishable from panmixia.
 
-Recap the general effects of summary stats-- skew in SFS, inflation of pi for low dispersal. What did we know about this before i.e from literature?
+The patterns observed in sequence data reflect the effects of space on the underlying genealogy.
+Nearby individuals coalesce rapidly under limited dispersal and so are connected by short branch lengths, while distant individuals take much longer to coalesce than they would under random mating.
+Mutation and recombination events in our simulation both occur at a constant rate along branches of the genealogy, so the genetic distance and number of recombination events separating two individuals is simply a noisy estimate of the branch lengths connecting them.
+These genealogical patterns also relate directly to the site frequency spectrum.
+In our simulations we observed that groups of nearby individuals tend to coalesce rapidly, while coalescence among groups at opposite ends of the landscape takes much longer than under random mating.
+Tip branches (i.e.
+branches subtending only one individual) are then relatively short, and branches in the middle of the genealogy connecting local groups of individuals relatively long.
+These patterns then create the biases we observed in the SFS -- the lowest frequency bins are deflated by the short branch lengths connecting nearby individuals, while mid-frequency bins are inflated by the long branches connecting local groups.
+
+The genealogical patterns introduced by limited dispersal are particularly apparent in the distribution of haplotype block lengths (Figure \ref{fig:sumstats}).
+This is because identical-by-state tract lengths reflect the impacts of two processes acting along the branches of the underlying genealogy -- both mutation and recombination -- rather than just mutation as is the case when looking at the site frequency spectrum or related summaries.
+This means that the pairwise distribution of haplotype block lengths carries with it important information about genealogical variation in the population, and correlation coefficients between moments of the this distribution and geographic location contain signal similar to the correlations between $F_{st}$ or $D_{xy}$ and space \citep{Rousset1997}.
+Indeed this basic logic underlies two recent studies explicitly estimating dispersal from the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}.
+Conversely, because haplotype-based measures of demography are particularly sensitive to variation in the underlying genealogy, inference approaches that assume random mating when analyzing the distribution of shared haplotype block lengths are likely to be strongly affected by spatial processes.
 
 \subsection{Effects of Sampling}
+One of the most important differences between random mating and spatial models is the effect of sampling: in a randomly mating population the spatial distribution of sampling effort has no effect on estimates of genetic variation, but when dispersal is limited sampling strategy can compound spatial patterns in the underlying genealogy and create pervasive impacts on all downstream genetic analyses.
+As expected we found that random sampling provides the most accurate summary of genetic diversity across the landscape.
+However this strategy is often impractical for empirical studies.
+In reality the difficulty of traveling through all parts of a species range and the inefficiency of collecting single individuals at each sampling site means that most studies follow something closer to the "point" sampling strategy we simulated, in which multiple individuals are sampled from nearby points on the landscape.
+For example, in ornithology a sample of 10 individuals per species per locality is a common target when collecting for natural history museums.
+In classical studies of Drosophila variation the situation is considerably worse, in which a single orchard might be sampled with bated traps for instance.
 
-- clustered sampling causes more variation than dispersal scaling for several statistics
-- range edges cause differences in the distribution of genetic variation across a range so sampling the midpoint gives a biased estimate of diversity and heterozygosity.
-Tajima's D is particularly sensitive to this effect.
-- in demographic analyses
+When sampling is clustered at points on a landscape and dispersal is limited, the sampled individuals will be more closely related than a random set of individuals.
+Average coalescence times of individuals collected at a locality will then be more recent and branch lengths shorter than expected by analyses assuming random mating.
+This leads to fewer mutations and recombination events occurring since their last common ancestor, causing a random set of individuals to share longer average IBS tracts and have fewer nucleotide differences.
+For some data summaries, such as Tajima's $D$, Watterson's $\Theta$, or the correlation coefficient between spatial distance and the count of long haplotype blocks, this can result in large differences in estimates between random and point sampling (Figure \ref{fig:sumstats}).
+Inferring underlying demographic parameters from these summary statistics -- for example, estimating dispersal distance as the slope of a regression of $F_{st}$ against the logarithm of spatial distance \citep{Rousset1997} -- may then be subject to bias if sampling is not random across the landscape.
+
+However, the largest sampling effects we observed occurred in our "midpoint" sampling strategy.
+This model is meant to reflect a bias in sampling effort towards the middle of a species' range.
+In empirical studies this sampling strategy could arise if, for example, researchers choose to sample the center of the range and avoid range edges to maximize probability of locating individuals during a short field season.
+Because midpoint sampling provides limited spatial resolution it dramatically reduces the magnitude of observed correlations between spatial and genetic distances.
+More surprisingly, midpoint sampling also leads to strongly positive Tajima's $D$ and an inflation in the proportion of heterozygous individuals in the sample.
+This increase in observed heterozygosity appears to reflect the effects of range edges, which are a fundamental facet of spatial genetic variation that have often been ignored by analytic approaches focusing on infinite toroidal landscapes \citep{Felsenstein1975}.
+If individuals move randomly in a finite two-dimensional landscape then regions in the middle of the landscape receive migrants from all directions while those on the edge receive no migrants from at least one direction.
+The average number of new mutations moving into the middle of the landscape is then higher than the number moving into regions near the range edge, leading to higher heterozygosity and lower inbreeding coefficients ($F_{IS}$) away from range edges.
+Though here we used only a single parameterization of fitness decline at range edges we believe this is a general property of non-infinite landscapes as it has also been observed in previous studies simulating under lattice models \citep{Neel2013,Shirk2014}.
+For empirical studies this suggests that sampling only the middle of the landscape will produce a biased view of many aspects of the genetic variation in a population.
+
+In summary, empirical researchers should collect individuals in as random a manner as practical and should not bias their sample towards the middle of a species' range.
+When sampling is clustered, summary statistics based on segregating sites (e.g.
+Watterson's $\Theta$ and Tajima's $D$), heterozygosity, or the distribution of long haplotype blocks, can be expected to depart significantly from what would be observed under random sampling.
+Comparing the results of analyses conducted on all individuals versus those limited to single individuals per locality may help to reveal any effect of sampling bias if sufficient samples are available.
+Alternatively, sampling strategy could be incorporated directly into a simulation-based inferential framework by, for example, setting the  distribution of samples in a simulation to mimic that used for the empirical target case.
+This could be readily achieved in a supervised machine learning or approximate Bayesian setting (CITES).
 
 \subsection{Demography}
-- low dispersal and clustered sampling bias recent Ne down and inflate past Ne
-- estimates are noisy in all simulations -- need to repeat analyses and average even for coalescent sims (???)
--
+Classical population genetic models collapse many elements of life history variation into a single parameter, $N_{e}$, which is then taken to reflect the degree of variation present in the population when modeling the effects of selection or migration.
+Inferring $N_{e}$ in the past is now a common goal of population genomic analyses and an important step in establishing baseline expectations of genetic variation when searching for signals of selection.
+Here we found that one method of inference of historic $N_{e}$ based on genome-wide estimates of the site frequency spectrum, Stairwayplot \citep{Liu2015}, is relatively robust to variation in dispersal distance when sampling is random and neighborhood size is over 20.
+However, non-random sampling, and particularly midpoint sampling, causes the method to infer inflated estimates of past population sizes and a series of recent bottlenecks (Figure \ref{fig:demography}.
+All sampling strategies lead to inflated ancient and deflated recent $N_{e}$ when neighborhood sizes were less than 20.
+
+These predictions match the biases visible in the raw site frequency spectrum (Figure \ref{fig:sumstats}) -- the deficit of low frequency alleles corresponds to the recent bottlenecks while the inflation of mid-frequency alleles corresponds to the high ancestral $N_{e}$.
+Though we found that Stairwayplot is a noisy estimator of equilibrium demography in general, there was no significant bias in demographic estimates for any sampling strategy for neighborhood sizes over 100.
+Thus many existing analyses are likely robust to biases in inferred $N_{e}$ caused by limited dispersal in continuous landscapes.
+However  barriers to dispersal will likely lead to higher levels of differentiation than we simulated here, and may mimic those seen at the low end of continuous dispersal we simulated.
+
+\cjb{could be another paragraph here discussing the relationship between Ne, the distribution of coalescence times, and dispersal.}
+\ak{wonder if it is worth revisting MSMC or similar now that we know more about how to run the method?}
 
 \subsection{GWAS}
+Over the last twenty years genome-wide association studies (GWAS) have identified tens of thousands of correlations between genetic variation and phentoypes, both in humans and other species.
+This technique is increasingly applied to questions of human health through methods like polygenic risk scores that sum the effect sizes estimated from GWAS to predict an individual's phenotype or disease risk \citep{Khera2018}.
+The most common approach to GWAS, which we followed in our analyses of simulated data here, is to regress the count of derived alleles at a site against individual phenotypes, taking the slope of this regression as an estimate of the effect of the allele on the phenotype.
+As recently reviewed by \cite{Visscher2019}, this is exactly the approach outlined by \cite{Fisher1918} at the dawn of quantitative genetics.
 
-- points: PCA correction is pretty good but not perfect. low-effect-size-alleles prob hit worst. 
-- we know that this is also a problem empirically in human data, and was enough of an issue for arabidopsis that PC regression isn't used.
-Approaches like Bayesenv or the Science yellow warbler paper likely contain residual structure, particularly when environmenal clines parallel genetic clines and dispersal is high.
-- caveats: small populations, need to scale, mixed are probably likely better
-- recs: run a gwas for sample x & y coordinates with whatever correction you're using. If there are significant values you're not correcting enough.
-- for clinal traits, include x and y as covariates.
+Stepping back from the mechanics of GWAS specifically, the approach of quantitative genetics is to decompose the variance in phenotypes into environmental and genetic effects, e.g. 
+\begin{equation}
+P=G+E
+\end{equation}
+\begin{equation}
+var(P)=var(G)+var(E)+cov(G,E)
+\end{equation}
+For GWAS in structured populations, the bias identified in many previous studies \citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015} in which test statistics reflect population structure rather than phenotype association arises because of the presence of a positive covariance between genotypic and environmental variation (the third term above).
+When the environment and genotype covary across space their effects are confounded.
+Note this does not require interaction between genotype and environment (so-called GxE effects), which will introduce an additional source of bias.
+Here we refer simply to the fact that the allele frequency at a site will often covary with the environment when both breeding structure and the environment vary over space.
+To some degree the success of quantitative genetics in fields like agriculture is likely due to the absence of this covariation -- crop and animal breeding operations can be conducted so that the environment is identical (or nearly so) or randomized across populations.
+However in natural populations this situation is extremely unlikely.
+Most populations are structured by a combination of limited dispersal and geographic barriers, and nearly all environments vary over space.
+GWAS in natural populations is forced to confront the confounding effects of population structure and the environment directly.
 
+Incorporating PC positions as covariates in the analysis \citep{Price2006} is designed to address this difficulty by regressing out a baseline level of "average" differentiation.
+However while this approach is quite useful, it does not truly separate the confounding signals of environment and spatially varying genotypes.
+In essence with a PC-corrected GWAS we are asking "what regions of the genome are more associated with this phenotype than the average genome-wide association observed across populations?" In our simulations we observed that this procedure can fail under a variety of circumstances.
+If dispersal is limited and environmental variation is clustered in space (i.e.
+corner or patchy distributions in our simulations), PCA positions fail to capture the fine-scale spatial structure required to remove all signals of association.
+Conversely when dispersal is high we found that PCA loses power to describe population structure before the spatial scale of dispersal breaks down the relationship between genotype and the environment.
+These effects were observed in all spatially varying environmental distributions, but were particularly pronounced for concentrated environmental effects in one region, as was also found in \cite{Mathieson2012}.
+As a result we can expect to see several thousand weak false-positive associations in a PC-corrected GWAS conducted on a human-sized genome in species with neighborhood sizes up to at least 1000.
+
+This does not mean that GWAS is not useful, but does put some limits on the extent of valid interpretation.
+Very few of the associations we identified would be significant at a conservative Bonferroni-adjusted $p$-value cutoff, suggesting that most of the very strong signals of association signals observed in studies of mono- or oligogenic traits are robust to stratification bias.
+Further, the most dramatic effects of stratification inflation we observed occurred at neighborhood sizes below 100 -- smaller than the vast majority of modern human populations (but see below for further discussion of empirical cases).
+However, as recently identified in studies of gentoype associations for human height in Europe \citep{Berg2018,Sohail2018}, PC regression GWAS in modern human populations does leave residual signal of population structure in large-scale GWAS of polygenic traits.
+Indeed, studies in strongly structured species like \textit{Arabadopsis} have long relied on more sophisticated mixed model approaches to correcting for population structure for precisely this reason \citep{Aranzana2005,Sasaki2015}.
+
+A second point that has received less attention in the literature is the issue of overcorrection in GWAS.
+If a truly causal allele segregates at different frequencies in different populations, then correcting for population structure in a regression analysis will result in an underestimate of effect sizes.
+Though our simulations had no causal alleles, we observed some evidence of this effect in the distribution of $p$-values across the genome (Figure \ref{fig:gwas}D): after PC regression many analyses resulted in $p$-values falling below their expected values from a uniform distribution.
 This result is consistent with a recent empirical study of heritability in human height and body mass index, which found that increasing the number of PC axes used as covariates caused the total proportion of variance explained by SNPs to decline from $\approx0.8$ to $\approx0.75$ \citep{Wainschtein2019}.
-Indeed SNPs with minor alleles frequencies of 0.001 - 0.01, which are expected to reflect fine scale population structure \citep{Mathieson2012,Novembre2009}, are estimated to explain \textit{negative} proportions of the total phenotypic variance in \citep{Wainschtein2019}.
+Indeed SNPs with minor alleles frequencies of 0.001 - 0.01, which are expected to reflect fine scale population structure \citep{Mathieson2012,Novembre2009}, are estimated to explain \textit{negative} proportions of the total phenotypic variance in \citep{Wainschtein2019}\ak{i don't understand what the authors mean here-- how can we have a negative variance?}.
+Searching for genetic associations of polygenic traits that vary systematically across but not within populations through existing GWAS approaches is then unlikely to be successful: the signals are fully confounded, and new analytic methods or experiments controlling for variation in the environment will be necessary to rigorously identify causal variants.
 
-\subsection{Where are organisms on this spectrum?}
-Need concrete examples here to relate these simulation results to reality. Where are humans? Mountain goats? Sheep? Does this mean we now need to reinterpret previous results?
+In summary, spatial covariation in population structure and the environment confound the interpretation of GWAS $p$-values, and correction using principal components is insufficient to fully separate these signals for polygenic traits under a variety of environmental and population parameter regimes.
+How more sophisticated mixed-model methods would perform under our simulations is an interesting question that we plan to pursue in a future study, but statistical methods can only take us so far in the absence of controlled experiments.
+One currently popular approach to estimating the degree of bias in GWAS caused by population structure is LD score regression \citep{Bulik-Sullivan2015}.
+Though this approach appears to work well in practice, its interpretation is not always straightforward and it is likely biased by the presence of linked selection \citep{Berg2018}.
+We suggest a straightforward alternative for species in which the primary axes of population differentiation is space (note this is likely not the case for many modern human populations): run a GWAS with spatial coordinates as phenotypes and check for $p$-value inflation or significant associations.
+If significant associations with sample locality are observed after correcting for population structure through PC regression or a kinship matrix, the structure corrections are insufficient.
+This is essentially the approach taken in our "clinal" model (though we also include normally distributed variation in our phenotypes).
+Of course it is possible that genotypes indirectly affect individual locations by adjusting organismal fitness and thus habitat selection across spatially varying environments, but we believe that this hypothesis should be tested against a null of stratification bias inflation rather than accepted as true based on GWAS results.
+\cjb{what do you think of the second half here?}
 
-\subsection{speculation}
-As we have shown, patterns of population genetic summary statistics contain information about dispersal.
-We imagine that combinations of such summaries when regressed against geographic distance might be sufficient for the construction of supervised machine learning regressors (e.g.
-cite Schrider and Kern 2018) for the accurate estimation of dispersal from genetic data.
-
-\ak{other things to say?}
-
-\section{Data Availability}
-\ak{github repo}
-
-\section{Acknowledgements}
-We thank folks for reading and thinking about this manuscript.
-We thank the Hearth for having such good nearby coffee and members of the 2018-2019 UO fluid dynamics seminar for many helpful discussions.
-CJB and ADK were supported by NIH award R01GM117241.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Figure and Tables %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[htbp]
-\includegraphics{figures/sampling_maps.pdf}
-\caption{Example sampling maps for 60 individuals on a 50x50 landscape under midpoint, point, and random sampling strategies.}
-\label{fig:samplemap}
-\end{figure}
-
-\begin{figure}[htbp]
-\includegraphics{figures/pop_params.pdf}
-\caption{Genealogical parameters from spatial and random mating SLiM simulations, by neighborhood size.}
-\label{fig:genparams}
-\end{figure}
-
-\begin{figure*}[p]
-\centering
-\includegraphics[width=\textwidth]{figures/sfs_w_sumstats.pdf}
-\caption{A: Site frequency spectra under midpoint, point, and random sampling schemes.
-Lines are loess curves over the site frequency spectrum from each simulation with colors scaled to neighborhood size.
-Random mating models showed no differences by sampling scheme -- see supplementary figure S1.
-Spatial models have fewer low-frequency alleles and more mid-frequency alleles when neighborhood sizes are small, and this effect is stronger when sampling is more concentrated.
-B: Summary statistics under each sampling strategy, by neighborhood size.}
-\label{fig:sumstats}
-\end{figure*}
-
-\afterpage{\clearpage}
-\begin{figure*}[p]
-\centering
-\includegraphics[width=\textwidth]{figures/stairwayplot_facet_rollmean.pdf}
-\caption{Inferred demographic histories for spatial SLiM simulations from Stairwayplot, by sampling scheme and neighborhood size (NS) range.
-The thick line is a rolling mean and thin lines are individual model fits.
-Dashed horizontal lines are the average $N_{e}$ across random-mating SLiM models estimated from $\theta_{\pi}$.}
-\label{fig:demography}
-\end{figure*}
-
-\afterpage{\clearpage}
-\begin{figure*}[p]
-\centering
-\includegraphics[width=\textwidth]{figures/gwas_summary.pdf}
-\caption{Impacts of spatially correlated phenotypes and isolation by distance on linear regression GWAS.
-A: example environmental distributions and sampling maps, with colors scaled to phenotype values.
-B: proportion of total variance explained by the first 10 PC axes, by neighborhood size.
-C: Proportion of significant SNPs after FDR correction for linear-model GWAS conducted with (top) or without (bottom) PC covariates, by neighborhood size.
-D: Quantile-quantile plots of observed $p$ values relative to those expected from a uniform distribution.
-The dotted lines show the 95\% confidence region assuming binomial sampling and an alpha level of 0.05.
-Points above the 1:1 line reflect deflated $p$ values.}
-\label{fig:gwas}
-\end{figure*}
-
-\afterpage{\clearpage}
 % latex table generated in R 3.5.1 by xtable 1.8-3 package
 % Thu Apr 11 13:11:59 2019
-\begin{table}[ht]
+\begin{table}[hbpt]
+\small
+\begin{adjustwidth}{-.9cm}{}
 \centering
-\caption{\bf Neighborhood size estimates from empirical studies.
-Survey methods indicate field studies to measure dispersal or population density.
-Genetic methods include both sequence-based analyses and enzyme activity assays or crosses used to estimate allele frequencies.}
+\caption{\bf Neighborhood size estimates from empirical studies.}
 \begin{tabular}{rllrrll}
@@
 \label{table:NStable}
+\end{adjustwidth}
 \end{table}
+
+\subsection{Where are natural populations on this spectrum?}
+For how much of the tree of life do spatial patterns circumscribe genomic variation?
+In Table \ref{table:NStable} we gathered estimates of neighborhood size from a range of organisms to get an idea of how likely dispersal is to play an important role in patterns of variation.
+Though this sample is almost certainly biased towards small-neighborhood species (because few studies have quantified neighborhood size in species with very high dispersal or population density), we find that neighborhood sizes in the range we simulated are fairly common across a range of taxa.
+At the extreme low end of empirical neighborhood size estimates we see some flowering plants, large mammals, and colonial insects like ants.
+Species such as this have neighborhood size estimates small enough that spatial processes are likely to strongly influence inference.
+These include some human populations such as the Gainj- and Kalam-speaking people of Papua New Guinea, in which the estimated neighborhood sizes in \citep{Rousset1997} range from 40 to 410 depending on the method of estimation.
+Many more species occur in a middle range of neighborhood sizes between 100 and 1000 -- a range in which spatial processes play a minor role in our analyses under random spatial sampling but are important when sampling of individuals in space is clustered.
+Last, many species likely have neighborhood sizes much larger than we simulated, including modern humans in NE Europe \citep{Ringbauer2017}.
+For these species demographic inference and summary statistics are likely to reflect minimal bias from spatial effects as long as dispersal is truly continuous across the landscape.
+While that is so we caution that association studies in which the effects of population structure are confounded with spatial variation in the environment are still sensitive to dispersal even at these large neighborhood sizes.
+
+\subsection{Future Directions and Limitations}
+As we have shown, a large number of population genetic summary statistics contain information about spatial population processes.
+We imagine that combinations of such summaries might be sufficient for the construction of supervised machine learning regressors (e.g.
+\cite{Schrider2018}) for the accurate estimation of dispersal from genetic data.
+Indeed \cite{Ashander2018} found that inverse interpolation on a vector of summary statistics provided a powerful method of estimating dispersal distances.
+Expanding this approach to include the haplotype-based summary statistics studied here and applying machine learning regressors built for general inference of nonlinear relationships from high-dimensional data may allow precise estimation of spatial parameters under a range of complex models.
+
+Though our continuous space simulation allows incorporation of realistic demographic and spatial processes and is much faster than previous individual-based models, it is inevitably limited by the computational burden of tracking tens of thousands of individuals in every generation.
+In particular the calculations required for our mate selection and competition steps involve summarizing distances across all pairs of individuals and so scale very poorly \ak{how exactly does this scale?
+$O(N^2)$?} as the number of individuals within three-$\sigma$ increases.
+In part the issue of runtime scaling is a function of the spatial process itself -- under very limited dispersal we observed that coalescence requires over 30N generations, so forward-time methods must be run for a very long time to create a complete genealogy as underlies all real genome sequence data.
+The reverse-time model of continuous space evolution described in \cite{Barton2010} and implemented in \cite{Kelleher2014} may allow exploration of parameter regimes with population and landscape sizes more directly comparable to empirical cases like humans.
+However, incorporating selection or other processes into such models will likely be difficult.
+
+A mixed approach may be possible by combining forward- and reverse-time models, as was recently done for a continuous-space Wright-Fisher model in \cite{Lotterhos2019} and for a simulation with linked selection in \cite{Buffalo2019}.
+This would allow us to generate short runs of complex, realistic simulations in forward time in SLiM \citep{Haller2019} and then "finish" the simulations as a coalescent simulation in msprime \citep{Kelleher2016}.
+However a significant difficulty in this "recapitation" approach is scaling the variance in reproductive output and generation time across forward- and reverse-time methods.
+Further development of our understanding of how to merge forward- and reverse-time models is a promising avenue for future research that will be necessary for scaling continuous-space simulations to millions or billions of individuals.
+
+Finally we believe that the difficulties in correcting for population structure in continuous populations using principal components analysis or similar decompositions is a difficult issue, well worth considering on its own-- what does it mean to correct for population structure in a world without discrete demes?
+What are the boundaries between proper corrections between ancestry and underpowering the search for local genetic variation?
+We posit that there is progress yet to be made in deriving either process driven descriptions of ancestry or more generalized unsupervised methods that can better account for shared relatedness, say in the context of GWAS, for populations that exist in space.
+\ak{this paragraph is a bit out there.
+rein me in}
+
+\section{Data Availability}
+\ak{github repo}
+
+\section{Acknowledgements}
+We thank folks for reading and thinking about this manuscript.
+We thank the Hearth for having such good nearby coffee.
+Falling Sky Brewing was key to the success of this research endeavor.
+CJB and ADK were supported by NIH award R01GM117241.
 
@@
 
+\begin{table}
+\label{table:sumstats}
+\small
+\centering
+\caption{\bf Summary statistics calculated on simulated genotypes.}
+\begin{tabular}{rllrrrrr}
+  \hline
+Statistic & Description \\ 
+  \hline
+$\Theta_{/pi}$ & Mean of the distribution of pairwise genetic differences \\
+$\Theta_{/W}$ & Effective population size based on segregating sites \\
+Segregating Sites & Total number of segregating sites in the sample \\
+Tajima's $D$ & Difference in $\Theta_{/pi}$ and $\Theta_{/W}$ over its standard deviation\\
+Observed Heterzygosity & Proportion of heterozygous individuals in the sample \\
+$F_{is}$ & Wright's inbreeding coefficient $1-H_{e}/H_{o}$ \\
+$var(D_{xy})$ & Variance in the distribution of pairwise genetic distances \\
+$skew(D_{xy})$ & Skew of the distribution of pairwise genetic distances \\
+$mean(IBS)$ & \makecell[l]{Mean of the distribution of pairwise identical-by-state (IBS) \\tract lengths taken over all pairs.} \\
+$var(IBS)$ & \makecell[l]{Variance of the distribution of pairwise identical-by-state (IBS) \\tract lengths taken over all pairs.} \\
+$skew(IBS)$ & \makecell[l]{Skew of the distribution of pairwise identical-by-state (IBS) \\tract lengths taken over all pairs.} \\
+$nIBS$ & Number of IBS tracts with length > 2bp across all pairs of individuals. \\
+$mean(IBS>1e6)$ & \makecell[l]{Mean number of IBS tracts over $1\times10^6$bp per pair \\across all pairs in the sample.} \\ 
+$corr(D_{xy},dist)$ & Pearson correlation between genetic distance and $log_{10}(spatial\text{ }distance)$ \\
+$corr(mean(IBS),dist)$ & \makecell[l]{Pearson correlation between the mean of the IBS tract distribution \\for each pair of samples and $log_{10}(spatial\text{ }distance)$} \\
+$corr(nIBS,dist)$ & \makecell[l]{Pearson correlation between the number of IBS tracts \\for each pair of samples and $log_{10}(spatial\text{ }distance)$} \\
+$corr(IBS>1e6,dist)$ & \makecell[l]{Pearson correlation between the number of IBS tracts $> 1\times10^6$bp \\for each pair of samples and $log_{10}(spatial\text{ }distance)$} \\
+$corr(skew(IBS),dist)$ & \makecell[l]{Pearson correlation between the skew of the distribution of pairwise haplotype\\ block lengths for each pair of samples and $log_{10}(spatial\text{ }distance)$} \\
+\end{tabular}
+\end{table}
+
 \begin{table*}[htbp]
+\tiny
 \centering
@@
 \begin{table}[ht]
+\small
 \centering

### 2019-04-17 scripts/gen_time_effects.Rmd
scripts update for ms figs

---
title: "Generation Time Scaling in Continuous Space Models"
author: "CJ Battey"
date: "2/15/2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reticulate);library(data.table);library(ggplot2);library(magrittr)
theme_set(theme_classic())
```

## The issue

In a non-Wright-Fisher SLiM simulation time is measured in timesteps rather than generations, so the branch lengths stored in the edge table are also in units of timesteps. If msprime or tskit are later used to add mutations to a NWF tree sequence, the mutation rate input by the user is also interpreted as mutations per timestep.

This is a non-issue if you add mutations in SLiM (but that slows down already slow simulations) and isn't a big issue if you're comparing simulations with a constant generation time (then you can scale all the mutation rates by the same value). But if generation time varies across simulations then applying a single rate in msprime.mutate( ) will lead to different per-generation rates and bias any genetic measure correlated with theta=4Nmu, which is to say, all of them.

Here I try to figure out if generation time scales with dispersal distance in continuous-space non-Wright-Fisher models (we need a better name for these (cspan??)), see whether we can correct for it by appropriately scaling mu in msprime, and check if it explains some earlier weird results from simulations with high dispersal distance.

## Generation Time Scales with Sigma
I ran some short simulations on small landscapes with mutations in SLiM (16x16, census sizes 1,100 - 400) for 200 generations, remembered all individuals for the second 100 generations, and saved the treeseqs to disk. Peter wrote up this nice code to calculate the average age of parents for individuals born during the all-individuals period. (We also talked about an improved version to estimate "effective generation time" by stepping up sparsetrees of final-generation individuals that would be nice to implement later).
```{python}
import os,numpy as np, msprime, pyslim, re
from tqdm import tqdm
import altair
import pandas
from matplotlib import pyplot as plt
os.chdir("/Users/cj/spaceness")

#read through all-individual ts to get mean parent ages
sims=[x for x in os.listdir("/Users/cj/Desktop/spaceviz/all_ind_trees/w16/") if not x.startswith(".")]
gen_times=np.zeros((len(sims),2))
for i in range(len(sims)):
    ts=pyslim.load(os.path.join("/Users/cj/Desktop/spaceviz/all_ind_trees/w16/",sims[i]))
    sigma=float(re.split("_|s",sims[i])[1])
    node_times = ts.tables.nodes.time
    parent_nodes = ts.tables.edges.parent
    child_nodes = ts.tables.edges.child
    max_T = 100 # number of time steps you recorded everyone for
    # which edges in the EdgeTable correspond to reproduction events we have parent and offspring remembered
    use_these = (node_times[parent_nodes] < max_T)
    parent_ages = node_times[parent_nodes] - node_times[child_nodes]
    gen_times[i,:]= [np.mean(parent_ages[use_these]),sigma]

gen_times=pandas.DataFrame(gen_times,columns=["gen_time","sigma"])
```

```{r,fig.width=4,fig.height=3,fig.align="center"}
ggplot(data=py$gen_times,aes(x=sigma,y=gen_time))+
  geom_point()
```

I also ran a set of small-landscape simulations to coalescence while adding mutations in SLiM -- this slows things down but this size landscape still coalesces in under an hour. I also didn't remember all individuals here, which *really* slows things down. Then I estimated the effective generation time by trying to get msprime.mutate( ) to produce the same number of mutations as SLiM while stochastically optimizing the generation time as a scaling constant for mu.

```{python}
os.chdir("/Users/cj/spaceness")
def guess_gen_time(ts):
    t=pyslim.load(ts)
    slim_muts=t.num_mutations
    g=4
    msp_t=msprime.mutate(t,1e-8/g,keep=False) #slim mutation rate was set to 1e-8
    msp_muts=msp_t.num_mutations
    while abs(msp_muts-slim_muts) > 0.01*slim_muts:
        new_g=g+np.random.uniform(-1,1,1)
        new_msp_t=msprime.mutate(t,1e-8/new_g,keep=False)
        new_msp_muts=new_msp_t.num_mutations
        if abs(new_msp_muts-slim_muts) < abs(msp_muts-slim_muts):
            msp_muts=new_msp_muts
            g=new_g
    return g[0]

sims=os.listdir("sims/slimout/spatial/W16_muts")
gen_times2=np.zeros((len(sims),2))
for i in range(len(sims)):
    g=guess_gen_time(os.path.join("sims/slimout/spatial/W16_muts/",sims[i]))
    sigma=float(re.split("_",sims[i])[1])
    gen_times2[i,:]=[g,sigma]
gen_times2=pandas.DataFrame(gen_times2,columns=["gen_time","sigma"])
```

```{r,fig.width=4.5,fig.height=3,fig.align="center"}
g1 <- py$gen_times
g1$method <- "parent ages"
g2 <- py$gen_times2
g2$method <- "compare n muts"
gen_times <- rbind(g1,g2)
ggplot(data=gen_times,aes(x=sigma,y=gen_time,col=method))+
  scale_color_manual(values=c("forestgreen","cornflowerblue"))+
  geom_point(shape=1)+
  geom_smooth(fill=NA)
```

Mostly consistent, though the estimated generation times are slightly higher when comparing *n* mutations (model artifact or "effective" gen time?). The generation time scaling here is way more severe than in the larger 50x50 simulations I've been working with, but the trend is the same. In the big simulations it looks like generation time still scales by around 10% - enough to mess with pi, theta, IBS distributions, etc. So we can't use the same per-timestep mutation rate across simulations with different sigma.

### msprime v SLiM mutation effects
I ran summary stats using my old method of putting a constant value in msprime.mutate() and compared them to summary stats using SLiM simulations to get an idea of whether or not this matters.

```{r,fig.width=6.5,fig.height=3,fig.align="center"}
setwd("~/spaceness/sumstats")

slim_spat <- fread("ss_spatial_random_W16muts.txt")
slim_spat$model <- "spatial"
slim_spat$muts <- "slim"
msp_spat <- fread("ss_spatial_random_W16_msp_muts.txt")
msp_spat$model <- "spatial"
msp_spat$muts <- "msp"
slim_rand <- fread("ss_random_random_W16muts.txt")
slim_rand$model <- "random"
slim_rand$muts <- "slim"
msp_rand <- fread("ss_random_random_W16_msp_muts.txt")
msp_rand$model <- "random"
msp_rand$muts <- "msp"
ss <- rbind(slim_spat,msp_spat,slim_rand,msp_rand)
ss <- data.frame(ss)[,!grepl("sfs",colnames(ss))]

popsize_spat <- fread("ss_spatial_random_W16muts.txt.popsizes")
popsize_rand <- fread("ss_random_random_W16muts.txt.popsizes")
popsize <- rbind(popsize_spat,popsize_rand)
colnames(popsize) <- c("sigma","census_n")
ss <- merge(ss,popsize,by="sigma")
ss$theta <- 4*(ss$census_n/2)*1e-8
ss$census_n2 <- ss$census_n

mss <- melt(ss,id.vars=c("model","muts","sigma","census_n2"))
pd <- subset(mss,variable %in% c("pi","theta"))
ggplot(data=pd,aes(x=sigma,y=value,col=paste(variable,model)))+
  #ggtitle("Ne estimates for constant mu applied in msprime v SLiM")+
  facet_wrap(~muts)+ylab("")+
  scale_color_brewer(palette = "Paired")+
  geom_point(shape=21,size=0.8,stroke=0.2)+
  geom_smooth(fill=NA,lwd=0.5,span=0.6)
```

Seems like yes.

### Runtime and memory usage for spatial models with mutations
How much longer will it take to simulate spatial models and do the mutations in SLiM? These are runtimes for the "full-size" spatial model with and without mutations. It seems like runtime differences are small past NS=~150 (sigma=~1.5). Memory usage per-generation is around 50% higher to start, but the shape here should change as runtime increases because the whole treeseq is loaded into memory when writing to file. I haven't run long enough simulations yet to compare, but in practice I could get simulations with sigma below 1 to run on talapas with 8G memory without mutations, but those got killed when run with mutations when they tried to write out the treeseq. I have some running with 32G now.

```{r,fig.width=6.5,fig.height=3,fig.align="center"}
runtimes <- fread("~/spaceness/spatial_runtimes.csv")
p1 <- ggplot(data=runtimes,aes(x=NS,y=time/100,col=muts))+geom_point(shape=1)+
  geom_line()+
  ylab("seconds per timestep")

runtimes$time_to_50N_timesteps <- (runtimes$time*5e3)/60^2
p2 <- ggplot(data=runtimes,aes(x=NS,y=time_to_50N_timesteps,col=muts))+geom_point(shape=1)+
  geom_line()

cowplot::plot_grid(p1,p2)

#memory usage
ggplot(data=runtimes,aes(x=NS,y=mem,col=muts))+geom_point(shape=1)+
  geom_line()+
  ylab("max memory in first 100 gen (mb)")

```

<!-- ```{r fig.width=4.5,fig.height=3,fig.align="center",} -->
<!-- model <- loess(gen_time~sigma,py$gen_times,span=0.5) -->
<!-- preds <- predict(model,seq(0.2,5,0.02)) -->
<!-- ggplot(data=data.frame(sigma=seq(0.2,5,0.02),preds),aes(x=sigma,y=preds))+ -->
<!--   geom_line() -->

<!-- sims <- list.files("~/spaceness/sims/slimout/spatial/W16_muts/") -->
<!-- sigmas <- sapply(sims,function(e) strsplit(e,"_") %>% unlist() %>% .[2] %>% as.numeric()) -->
<!-- pred_gen_times <- predict(model,sigmas) -->
<!-- pred_gen_times <- data.frame(sigmas,pred_gen_times) -->

<!-- ggplot(data=pred_gen_times,aes(x=sigma,y=pred_gen_times))+ -->
<!--   geom_line() -->

<!-- ``` -->

### 2019-04-17 scripts/LTTplots.py

@@ -1,0 +1,93 @@
+#spaceness lineage-through-time / coalescence time plots
+from slimtools import *
+from matplotlib import pyplot as plt
+import matplotlib as mpl
+
+#get n nodes by time bin across different sigma values
+os.chdir("/Users/cj/spaceness/sims/slimout/spatial/W50_run3/")
+trees=[x for x in os.listdir() if not x.startswith(".")]
+nodes_per_timebin=[]
+for t in trees:
+    sigma=float(re.sub("sigma_|_.trees*|.trees","",t))
+    ts=sample_treeseq(infile=t,
+                     outfile="",
+                     nSamples=100,
+                     sampling="random",
+                     recapitate=True,
+                     recombination_rate=1e-8,
+                     write_to_file=False,
+                     seed=12345)
+    nodetimes=[x.time for x in ts.nodes()]
+    n_nodes=0
+    nodes_per_timebin.append([0,0,sigma])
+    for m in np.arange(1,1.6e6-1,1e3):
+        n_nodes=n_nodes+len([x for x in nodetimes if  x >= m and x < m+1e3])
+        nodes_per_timebin.append([m+1e3,n_nodes,sigma])
+nodes_per_timebin=np.array(nodes_per_timebin)
+
+#line plot version (but how to get the colorbar??)
+fig = plt.figure(figsize=(2.5,2),dpi=600)
+sigmas=np.unique(nodes_per_timebin[:,2])
+colorscale=np.linspace(0,1,len(sigmas))
+for i in range(len(sigmas)):
+    plt.plot((np.log10(np.array([x for x in nodes_per_timebin if x[2]==sigmas[i]])[:,0]/4)/1e4),
+             np.array([x for x in nodes_per_timebin if x[2]==sigmas[i]])[:,1]/np.max(np.array([x for x in nodes_per_timebin if x[2]==sigmas[i]])[:,1]),
+             c=plt.cm.RdYlBu(colorscale[i]),clip_on=False,linewidth=.75)
+#plt.colorbar(shrink=0.9).set_label("sigma", labelpad=-13, y=1.08, rotation=0)
+plt.xlabel('Generations/N',fontsize=7)
+plt.ylabel('Proportion Total Coalescences',fontsize=7)
+fig.savefig("/Users/cj/Desktop/test2.pdf",bbox_inches='tight')
+
+#scatter plot version
[…]

### 2019-04-17 scripts/distance_by_MAF.py

@@ -1,0 +1,49 @@
+from slimtools import *
+import sklearn.metrics
+from tqdm import tqdm
+infiles=["/Users/cj/spaceness/sims/slimout/spatial/W50_run3/sigma_0.21467883716606018_.trees_3812242",
+         "/Users/cj/spaceness/sims/slimout/spatial/W50_run3/sigma_0.400670106952987_.trees_1541565",
+         "/Users/cj/spaceness/sims/slimout/spatial/W50_run3/sigma_0.8863583600140017_.trees_3937856",
+         "/Users/cj/spaceness/sims/slimout/spatial/W50_run3/sigma_1.257878563742616_.trees_1186380",
+         "/Users/cj/spaceness/sims/slimout/spatial/W50_run3/sigma_2.0108173412933894_.trees_7501871",
+         "/Users/cj/spaceness/sims/slimout/spatial/W50_run3/sigma_4.038506335646773_.trees_9162471"]
+for j in infiles:
+    simname=os.path.basename(j)
+    label=float(re.split("_",simname)[1])
+    ts=sample_treeseq(infile=j,
+                      outfile="",
+                      nSamples=1000,
+                      sampling="random",
+                      recapitate=False,
+                      recombination_rate=1e-9,
+                      write_to_file=False,
+                      sampling_locs=[[12.5,12.5],[12.5,37.5],[37.5,37.5],[37.5,12.5]],
+                      plot=False,
+                      seed=12345)
+
+    #get generation times estimated from short all-individual simulations
+    gentimes=np.loadtxt("/Users/cj/spaceness/W50sp_gentimes.txt")
+    gentime=[x[0] for x in gentimes if np.round(x[1],5)==np.round(label,5)]
+    ts=msp.mutate(ts,1e-8/gentime[0],random_seed=12345)
+
+    #get haplotypes and locations
+    haps=ts.genotype_matrix()
+    genotypes=allel.HaplotypeArray(haps).to_genotypes(ploidy=2)
+    sample_inds=np.unique([ts.node(j).individual for j in ts.samples()]) #add check that nodes corresponding to individuals are sequential for future versions?
+    locs=[[ts.individual(x).location[0],ts.individual(x).location[1]] for x in sample_inds]
+
+    allele_counts=genotypes.count_alleles()
+    dafs=np.array(allele_counts[:,1]/2000)
+    #average distance among allele copies, by MAF bin
[…]

### 2019-04-17 scripts/estimate_gen_times.py

@@ -1,0 +1,76 @@
+#estimate generation times from short slim runs
+import sys,subprocess as sp,re,os,numpy as np,pyslim
+from tqdm import tqdm
+import argparse
+import spatial_slim_ts as sps
+
+parser=argparse.ArgumentParser()
+parser.add_argument("--infile",dest="infile",type=str)
+args=parser.parse_args()
+
+args=argparse.Namespace(infile="/Users/cj/spaceness/sims/slimout/spatial/W50_run3/sigma_0.25_.trees_1661344")
+
+sigma=float(re.split("_",os.path.basename(args.infile))[1])
+
+id=int(np.random.uniform(1e7,1e8-1,1))
+sp.run("cd ~/spaceness/; slim -d sigma="+str(sigma)+" -d id="+str(id)+" slim_recipes/gen_timer_spatial.slim",
+       shell=True)
+#ts=pyslim.load("/Users/cj/spaceness/tmp"+str(id)+".trees")
+ts = sps.SpatialSlimTreeSequence(pyslim.load("/Users/cj/spaceness/tmp"+str(id)+".trees"), dim=2)
+
+############ average generation time across all individuals
+node_times = ts.tables.nodes.time
+parent_nodes = ts.tables.edges.parent
+child_nodes = ts.tables.edges.child
+max_T = 200
+use_these = (node_times[parent_nodes] < max_T)
+parent_ages = node_times[parent_nodes] - node_times[child_nodes]
+genealogical_gen_time=np.mean(parent_ages[use_these])
+
+############ variance in reproductive output
+mid_node_ids=np.array([x.id for x in ts.nodes() if x.time in range(100,101)])
+
+
+np.var(n_offspring)
+np.mean(n_offspring)
+
+
+############# 'effective' generation time sampling inds proportional to final-generation ancestry
+alive_nodes=[x.id for x in ts.nodes() if x.time==0] #final-generation nodes
+#alive = ts.individuals_alive(0)
+#alive_nodes = ts.individual_nodes(np.where(alive)[0])
+node_ancestry = ts.proportion_ancestry_nodes(alive_nodes)[0]
+ind_ancestry = np.zeros(ts.num_individuals)
+for k, ind in enumerate(ts.individuals()):
+    ind_ancestry[k] = sum(node_ancestry[ind.nodes])
+
+# now find average parent age for everyone
+parents = ts.individual_parents_dict()
+mean_parent_ages = np.zeros(ts.num_individuals) - 1.0
[…]

### 2019-04-17 scripts/generation_times.py

@@ -1,0 +1,83 @@
+import os, numpy as np, msprime, pyslim, re
+from tqdm import tqdm
+import altair
+import pandas
+from matplotlib import pyplot as plt
+import scikit-misk
+os.chdir("/Users/cj/spaceness")
+
+#read through all-individual ts to get mean parent ages along a  sparsetree ...
+sims=[x for x in os.listdir("/Users/cj/Desktop/spaceviz/all_ind_trees/") if not x.startswith(".")]
+gen_times=np.zeros((len(sims),2))
+for i in range(len(sims)):
+    ts=pyslim.load(os.path.join("/Users/cj/Desktop/spaceviz/all_ind_trees/",sims[i]))
+    sigma=float(re.split("_|s",sims[i])[1])
+    node_times = ts.tables.nodes.time
+    parent_nodes = ts.tables.edges.parent
+    child_nodes = ts.tables.edges.child
+    max_T = 100 # number of time steps you recorded everyone for
+    # which edges in the EdgeTable correspond to reproduction events we have parent and offspring Remembered
+    use_these = (node_times[parent_nodes] < max_T)
+    parent_ages = node_times[parent_nodes] - node_times[child_nodes]
+    gen_times[i,:]= [np.mean(parent_ages[use_these]),sigma]
+   
+gen_times=pandas.DataFrame(gen_times,columns=["gen_time","sigma"])
+plt.scatter(gen_times["sigma"],gen_times["gen_time"],
+            edgecolors="black",facecolors="none")
+plt.show()
+plt.close()
+
+#fit loess curve to generation times, use for predicting gen time of no-mutation slim sims
+model=sm.nonparametric.lowess(gen_times["gen_time"],gen_times["sigma"])
+plt.plot()
+
+
+
+
+#estimate generation time by comparing n mutations generated by msprime.mutate() (muts/timestep) v slim (muts/generation)
+def guess_gen_time(ts):
+    t=pyslim.load(ts)
+    slim_muts=t.num_mutations
+    g=4
+    msp_t=msprime.mutate(t,1e-8/g,keep=False) #slim mutation rate was set to 1e-8
+    msp_muts=msp_t.num_mutations
+    while abs(msp_muts-slim_muts) > 0.01*slim_muts:
+        new_g=g+np.random.uniform(-1,1,1)
[…]

### 2019-04-17 scripts/gwas_viz.R

@@ -1,43 +1,376 @@
 #visualizing GWAS on SLiM simulations
 library(ggplot2);library(magrittr);library(plyr);library(data.table)
 library(qvalue);library(cowplot);library(wesanderson)
-setwd("~/spaceness/gwas/")
-source("~/R/ggthemes.R")
+setwd("~/projects/spaceness/gwas/")
+pal <- grDevices::colorRampPalette(color=c("steelblue4","skyblue","lightgoldenrod1","orangered","red4"),
+                                   bias=1,space="rgb",interpolate="linear")
+theme_set(theme_classic()+theme(
+          axis.text=element_text(size=6),
+          axis.title=element_text(size=6),
+          strip.text=element_text(size=6),
+          legend.text=element_text(size=6),
+          legend.title = element_text(size=6),
+          strip.background = element_blank(),
+          title=element_text(size=6)))
+
+#####################################################
+###################  qq plots  ######################
+#####################################################
+#helper function for qqplot confidence interval given binomial sampling for alpha=0.05
+panel.qqconf<-function(n, conf.points=1000, conf.col="gray", conf.alpha=.05, ...) { 
+  require(grid)
+  conf.points = min(conf.points, n-1);
+  mpts<-matrix(nrow=conf.points*2, ncol=2)
+  for(i in seq(from=1, to=conf.points)) {
+    mpts[i,1]<- -log10((i-.5)/n)
+    mpts[i,2]<- -log10(qbeta(1-conf.alpha/2, i, n-i))
+    mpts[conf.points*2+1-i,1]<- -log10((i-.5)/n)
+    mpts[conf.points*2+1-i,2]<- -log10(qbeta(conf.alpha/2, i, n-i))
+  }
+  #grid.polygon(x=mpts[,1],y=mpts[,2], gp=gpar(fill=conf.col, lty=0), default.units="native")
+  return(mpts)
+}
+setwd("random_sampling")
+files <- list.files("out_corner",full.names = T) %>% grep(".assoc.linear",.,value = T)
+qq_corner <- data.table(P=NA,sigma=NA,observed=NA,expected=NA,neighbors=NA)[-1]
+for(f in files){
+  sigma <- strsplit(f,"_") %>% unlist() %>% .[3] %>% as.numeric()
+  if(sigma < 4){
+    gwas <- fread(f)
[…]

### 2019-04-17 scripts/ibs_viz.R

@@ -1,138 +1,188 @@
 setwd("~/spaceness/")
-library(plyr);library(ggplot2);library(magrittr);library(data.table);library(wesanderson)
-
-theme_set(theme_classic()+theme(axis.text=element_text(size=7),
-                                axis.title=element_text(size=7),
-                                strip.text = element_text(size=7),
-                                title = element_text(size=8,hjust=0.5),
-                                legend.title = element_text(size=7),
-                                legend.text = element_text(size=7),
+library(plyr);library(ggplot2);library(magrittr);library(data.table);library(car);library(wesanderson)
+library(cowplot)
+source("~/R/colordict.R")
+pal <- grDevices::colorRampPalette(color=c("steelblue4","skyblue","lightgoldenrod1","orangered","red4"),
+                                   bias=1,space="rgb",interpolate="linear")
+theme_set(theme_classic()+theme(axis.text=element_text(size=8),
+                                axis.title=element_text(size=8),
+                                strip.text = element_text(size=8),
+                                title = element_text(size=9,hjust=0.5),
+                                legend.title = element_text(size=8),
+                                legend.text = element_text(size=8),
                                 strip.background = element_blank()))
 
-ss_m <- fread("sumstats/ss_spatial_midpoint_W35.txt",data.table=F)
-rm_m <- fread("sumstats/ss_randmates_midpoint_W35.txt",data.table=F)
-ss_r <- fread("sumstats/ss_spatial_W35.txt",data.table=F)
-rm_r <- fread("sumstats/ss_randmates_W35.txt",data.table=F)
-ss_p <- fread("sumstats/ss_spatial_point_W35.txt",data.table=F)
-rm_p <- fread("sumstats/ss_randmates_point_W35.txt",data.table=F)
+ss_m <- fread("sumstats/ss_spatial_midpoint_W50.txt",data.table=F)
+ss_r <- fread("sumstats/ss_spatial_random_W50.txt",data.table=F)
+ss_p <- fread("sumstats/ss_spatial_point_W50.txt",data.table=F)
[…]

### 2019-04-17 scripts/msp_grid_sim.py

@@ -1,10 +1,14 @@
 #msprime island model grid simulation
-import numpy as np, msprime as msp, os, allel, itertools,scipy
+import numpy as np, msprime as msp, os, allel, itertools, scipy, math
 from scipy.spatial.distance import pdist, squareform
 import seaborn
 from matplotlib import pyplot as plt
+import matplotlib
+from sklearn.linear_model import LinearRegression
+from slimtools import *
 
-#utility functions
+#######################utility functions####################################
+##migration matrix for square landscapes with or without migration along diagonals
 def get_grid_migration_matrix(width,rate,diagonal):
     mig_matrix=np.zeros((width*width,width*width))
     lefts=np.arange(0,width**2,width)
@@ -52,6 +56,7 @@ def get_grid_migration_matrix(width,rate,diagonal):
                     mig_matrix[i][i+(width+1)]=np.sqrt(2)*rate
     return mig_matrix
 
+##migration matrix for toroidal landscapes (no edges)
 def get_torus_migration_matrix(npops,rate,diagonal):
     npops=26
     if not np.sqrt(npops)%1 == 0:
@@ -70,6 +75,35 @@ def get_torus_migration_matrix(npops,rate,diagonal):
         mig_matrix[i][i+(width+1)]=np.sqrt(2)*rate
     return mig_matrix
 
+##get parameters in grid format (xloc,yloc,migration_right,migration_left,migration_up,migration_down)
+def get_grid_params(width,rate):
+    popy=np.repeat(np.arange(0,width,1),width)
+    popx=[np.arange(0,width,1) for x in range(width)] #man... fuck python
+    popx=[x for y in popx for x in y]
+    mr=np.repeat(rate,width**2)
+    ml=np.repeat(rate,width**2)
+    md=np.repeat(rate,width**2)
+    mu=np.repeat(rate,width**2)
+    return(np.transpose([popx,popy,ml,mr,md,mu]))
+#grid_params=get_grid_params(10,.1/(1e5/100))
+
+##convert grid parameters to migration matrix.
+def grid_params_to_migration_matrix(grid_params):
+    npops=np.shape(grid_params)[0]
+    width=math.sqrt(npops)
[…]

### 2019-04-17 scripts/msp_to_smcpp.py

@@ -1,0 +1,40 @@
+#run an msprime simulation and analyze the genotypes in smcpp
+import msprime, sys, os, numpy as np, subprocess as sp
+import argparse
+
+parser=argparse.ArgumentParser()
+parser.add_argument("i")
+args=parser.parse_args()
+
+#args=argparse.Namespace(i=1)
+os.chdir("/home/cbattey2/spaceness/demography/smcpp/out_msp_growth")
+#os.chdir("/users/cj/spaceness/demography/smcpp/out_msp")
+
+
+sp.run(["mkdir",str(args.i)])
+pop_config=msprime.PopulationConfiguration(sample_size=40,initial_size=6.1e5,growth_rate=1e-3)
+tree=msprime.simulate(population_configurations=[pop_config],length=1e8,recombination_rate=1e-9,mutation_rate=1e-8)
+
+tree.write_vcf(open(str(args.i)+"/"+str(args.i)+".vcf","w"),2)
+sp.check_output(["bgzip",str(args.i)+"/"+str(args.i)+".vcf"])
+sp.check_output(["tabix",str(args.i)+"/"+str(args.i)+".vcf"+".gz"])
+inds=",".join(["msp_"+str(x) for x in range(20)])
+for j in range(1):
+    command=["smc++",
+             "vcf2smc",
+             #"-d","msp_"+str(int(np.random.uniform(0,19,1))),"msp_"+str(int(np.random.uniform(0,19,1))),
+             "--length","100000000",
+             "--cores","1",
+             str(args.i)+"/"+str(args.i)+".vcf"+".gz",
+             str(args.i)+"/"+str(args.i)+"_set"+str(j)+".smcpp",
+             "1",
+             "pop1:"+inds]
+    sp.run(command)
+command=("smc++ estimate --cores 1 -o "+
+         str(args.i)+" -p 0 -r 1e-9 --unfold 1e-8 "+
+         str(args.i)+"/"+str(args.i)+"_set"+"*")
+sp.run(command,shell=True)
+
+sp.run(["smc++","plot","--csv",
+        str(args.i)+".pdf",
+        str(args.i)+"/"+"model.final.json"])

### 2019-04-17 scripts/msp_to_stairwayplot.py

@@ -1,0 +1,156 @@
+import msprime as msp, numpy as np, allel, argparse, subprocess
+from os.path import basename, splitext, join
+
+parser = argparse.ArgumentParser(
+  description='Run stairwayplot on a tree sequence output from msprime.\
+               Requires gnu-parallel and Stairwayplot including \
+               the command-line version of Stairway_plot_output_summary,\
+               which can be downloaded from: \
+               https://sites.google.com/site/jpopgen/stairway-plot.')
+parser.add_argument('infile',
+                    help='path to tree sequence')
+parser.add_argument('outdir',
+                    help='full path to directory for writing output files')
+parser.add_argument('nboots',type=int,
+                    help='number of bootstrap replicates to run.')
+parser.add_argument('stairwayplot_dir',
+                    help="path to stairwayplot directory")
+parser.add_argument('cores',
+                    help="number of cores to use when fitting demographic \
+                    models.")
+parser.add_argument('seed',type=int,
+                    help="random number seed for bootstrap resampling.")
+parser.add_argument('mutation_rate',type=float,
+                    help="mutation rate (per base per generation)")
+parser.add_argument('generation_time',type=float,
+                    help="generation time")
+parser.add_argument('plot',type=bool,
+                    help='plot Ne ~ t ? T/F')
+args=parser.parse_args()
+
+np.random.seed(args.seed)
+
+#read in tree sequence
+ts=msp.load(args.infile)
+
+#count alleles, bootstrap over sites, return the SFS minus the 0% bin
+haps=np.array(ts.genotype_matrix())
+genotypes=allel.HaplotypeArray(haps).to_genotypes(ploidy=2)
+allele_counts=genotypes.count_alleles()
+sfs=allel.sfs(allele_counts[:,1])
+sfs=sfs[1:len(sfs)]
+
+#tmp directory for input files
+command=("cd "+args.outdir+";"+
+         "mkdir infiles")
+subprocess.run(command,shell=True)
+
+#write stairwayplot input
+out=open((join(args.outdir,
+               "infiles",
+               splitext(basename(args.infile))[0])+
+        "_strwyplt.txt"),
+        "w")
[…]

### 2019-04-17 scripts/msp_ts_to_sumstats.py

@@ -1,0 +1,126 @@
+from slimtools import *
+import argparse
+parser=argparse.ArgumentParser()
+parser.add_argument("--infile")
+parser.add_argument("--outfile")
+parser.add_argument("--pop_config")
+args=parser.parse_args()
+
+ts=msp.load(args.infile)
+ts=msp.mutate(ts,1e-8)
+
+
+haps=np.array(ts.genotype_matrix())
+positions=np.array([s.position for s in ts.sites()])
+
+if args.pop_config=="grid":
+    label=float(re.sub(".trees","",os.path.basename(args.infile)))
+    pops=[ts.node(x).population for x in ts.samples()]
+    locs=[[(np.floor(x/5)+.5)*10,((x%5)+.5)*10] for x in pops]
+    locs=np.array(locs)[np.arange(0,len(locs),2)]
+if args.pop_config=="onepop":
+    locs=np.array([[np.random.uniform(1,50,1)[0],np.random.uniform(1,50,1)[0]] for x in range(50)]) #random locations for nonspatial sim(?)
+    label=float(re.split("_",args.infile)[1])
+
+getSLiMSumStats(haps=haps,
+                positions=positions,
+                label=label,
+                locs=locs,
+                outfile=args.outfile,
+                maxlen=1e8,
+                ibs_tracts=True,
+                verbose=True,
+                min_len_to_keep=2)
+
+###########~~~~~~~debugging zone~~~~~~~~~###########
+ts=msp.load("/Users/cj/spaceness/sims/msp_grid/0.0010278777692654174.trees")
+ts=msp.mutate(ts,1e-8)
+label=0.06
+pops=[ts.node(x).population for x in ts.samples()]
+locs=[[(np.floor(x/5)+.5)*10,((x%5)+.5)*10] for x in pops]
+locs=np.array(locs)[np.arange(0,len(locs),2)]
+haps=np.array(ts.genotype_matrix())
+positions=np.array([s.position for s in ts.sites()])
+
+#load in genotypes etc into scikit-allel
+genotypes=allel.HaplotypeArray(haps).to_genotypes(ploidy=2)
+allele_counts=genotypes.count_alleles()
+genotype_allele_counts=genotypes.to_allele_counts()
+
+#nonspatial population-wide summary statistics
[…]

### 2019-04-17 scripts/msprime_gwas_pipeline.py

@@ -1,0 +1,28 @@
+#msprime gwas pipeline
+import msprime, numpy as np, os, re
+
+#neutral coalescent simulation
+ts=msprime.simulate(Ne=1e5,sample_size=100,length=1e7,recombination_rate=1e-8,mutation_rate=1e-8)
+haps=ts.genotype_matrix()
+positions=np.array([s.position for s in ts.sites()])
+
+#choose 100 random SNPs with allele freqs bw 0.1 and 0.9
+snp_indices=[]
+for i in range(100):
+  phen_snp_af=0
+  while phen_snp_af<0.1 or phen_snp_af>0.9:
+    rsnp=np.random.choice(np.arange(0,np.shape(haps)[0],1))
+    phen_snp_af=sum(haps[rsnp,:])/len(haps[rsnp,:])
+  snp_indices.append(rsnp)
+  
+#generate phenotypes by drawing from a normal distribution with sd = 1 and mean proportional to n "phenotype" snps
+phenotypes=[]
+i=1
+for i in range(100):
+  snp_count=sum(haps[snp_indices,i])
+  phenotypes.append(np.random.normal(snp_count*0.02,1,1)[0]) 
+
+#outputs to use for GWAS-style inference
+out=[haps,positions,phenotypes]
+
+

### 2019-04-17 scripts/offspring_variance.R

@@ -1,0 +1,80 @@
+#estimating and plotting genealogical parameters (var(nOff),census_n,gen_time)
+library(data.table);library(magrittr);library(plyr);library(progress)
+library(foreach);library(doMC)
+setwd("~/spaceness/")
+source("~/R/colordict.R")
+registerDoMC(cores=10)
+
+colornames <- list('red_orange','dark_medici_blue')
+theme_set(theme_classic()+theme(axis.text=element_text(size=8),
+                                axis.title=element_text(size=8),
+                                strip.text = element_text(size=8),
+                                title = element_text(size=9,hjust=0.5),
+                                legend.title = element_text(size=8),
+                                legend.text = element_text(size=8),
+                                strip.background = element_blank(),
+                                axis.title.y=element_blank()))
+
+#get variance in offspring by model and sigma
+files <- list.files("~/Desktop/spaceviz/10kpeds/",full.names = T)
+getvaroff <- function(f) {
+  ped <- fread(f)
+  sigma <- strsplit(basename(f),"_") %>% unlist() %>% .[3] %>% as.numeric()
+  model <- strsplit(basename(f),"_") %>% unlist() %>% .[1]
+  ped <- ddply(ped,.(ind_ID),function(e) e[e$timestep==max(e$timestep),]) #trim to last generation entry for each individual
+  ped <- subset(ped,timestep<150)
+  data.frame(sigma=sigma,model=model,varoff=var(ped$n_offspring))
+}
+varoff <- foreach(f=files,.combine = rbind) %dopar% getvaroff(f)
+varoff$model <- mapvalues(varoff$model,from=c("rm","sp"),to=c("random mating","spatial"))
+varoff$stat <- "var(n offspring)"
+varoff <- varoff[,c(1,3,2,4)]
+
+#estimate number of individuals born per timestep
+ped <- fread(files[1])
+nbirths <- c()
+for(i in 130:170){
+  t <- subset(ped,timestep==i)
+  nbirths <- append(nbirths,sum(!t$ind_ID %in% ped$ind_ID[ped$timestep<i]))
+}
+print(mean(nbirths))
+
+
+#census population size
+popsizes <- fread("~/spaceness/sumstats/ss_spatial_random_W50.txt.popsizes")
[…]

### 2019-04-17 scripts/pain_in_the_torus_example.R

@@ -1,0 +1,82 @@
+#toy example of Felsenstein's pain in the torus problem
+library(ggplot2)
+theme_set(theme_classic()+theme(axis.line=element_blank(),
+                                axis.title = element_blank(),
+                                axis.ticks=element_blank(),
+                                axis.text.y=element_blank()))
+
+#infinite landscape
+positions <- runif(100,1,100)
+
+ggplot(data=data.frame(positions),aes(x=positions))+
+  coord_polar()+
+  xlim(1,100)+
+  geom_point(aes(y=0))
+
+ #ngen <- 100
+
+#for(gen in 1:ngen){
+  offlocs <- c()
+  for(i in 1:length(positions)){
+    nOff <- rpois(1,1)
+    if(nOff){
+      for(j in 1:nOff){
+        newpos <- positions[i]+rnorm(1,0,1)
+        if(newpos > 100) newpos-100
+        if(newpos < 1) newpos+100
+        offlocs <- append(offlocs,newpos)
+      }
+    }
+  }
+  positions <- offlocs
+  
+  ggplot(data=data.frame(positions),aes(x=positions))+
+    #ggtitle(gen)+
+    xlim(1,100)+
+    coord_polar()+
+    geom_point(aes(y=0))
+#  
+#  png(paste0("~/Desktop/spaceviz/torus_pain_ex/",gen,".png"),width=4,height=4,unit="in",res=400)
+#  print(p)
+#  dev.off()
+#}
+system("cd ~/Desktop/spaceviz/;
+        convert -delay 20 torus_pain_ex/{1..99}.png -delay 200 torus_pain_ex/100.png torus_pain_ex.gif")
+
+
+#finite landscape with reflected offspring positions
+positions <- runif(100,1,100)
+
+ggplot(data=data.frame(positions),aes(x=positions))+
+  xlim(1,100)+
+  geom_point(aes(y=0))+
+  geom_density()
+
+ngen <- 100
+for(gen in 1:ngen){
+  offlocs <- c()
+  for(i in 1:length(positions)){
+    nOff <- rpois(1,1)
+    if(nOff){
+      for(j in 1:nOff){
+        newpos <- positions[i]+rnorm(1,0,1)
+        if(newpos > 100) newpos <- 100-(newpos-100)
+        if(newpos < 1) newpos <- abs(newpos)
+        offlocs <- append(offlocs,newpos)
+      }
+    }
+  }
[…]

### 2019-04-17 scripts/plot_msp_sumstats.R

@@ -1,0 +1,15 @@
+#msp grid sims summary stat viz 
+setwd("~/spaceness/")
+library(ggplot2);library(plyr);library(magrittr);library(data.table)
+
+ss <- fread("sumstats/ss_msp_grid.txt",data.table = F)
+ss <- ss[,!grepl("sfs",colnames(ss))]
+ss$sigma2 <- ss$sigma*4
+mss <- melt(ss,id.vars=c("sigma","sigma2"))
+ggplot(data=mss,aes(x=sigma2*10,y=value))+
+  theme_classic()+
+  xlab("sigma")+
+  facet_wrap(~variable,scales="free")+
+  geom_point(shape=21,stroke=0.3,size=1)+geom_smooth(span=.2,fill=NA,col="orange",lwd=0.5)
+
+

### 2019-04-17 scripts/plot_pop_config.R

@@ -1,0 +1,88 @@
+#msprime population configuration plotting script
+library(ggplot2);library(argparse)
+parser <- ArgumentParser(description='plot population configuration',python_cmd="python")
+parser$add_argument('--migration_matrix',metavar='migmat',type="character",
+                    help='path to model params file (comma-delimited text with columns x,y,Ne,mright,mleft,mup,mdown)')
+parser$add_argument('--outpath',metavar="outpath",type="character",
+                    help='output file path')
+args <- parser$parse_args()
+
+#debug params
+args <- list(migmat="~/Desktop/mig_matrix.txt",outpath="~/Desktop/test_pop_config.pdf")
+
+##### convert migration matrix to param table for plotting #####
+mig_matrix_to_plot_params <- function(mig_matrix_path){
+  mig_matrix <- read.table(mig_matrix_path)
+  npops <- ncol(mig_matrix)
+  width <- sqrt(npops)
+  params <- data.frame(matrix(ncol=6,nrow=npops))
+  for(i in 1:npops){
+    x <- i-width*(ceiling(i/width)-1)
+    y <- ceiling(i/width)
+    mr <- mig_matrix[i,i+1]
+    ml <- mig_matrix[i,i-1]
+    mu <- mig_matrix[i,i+width]
+    md <- mig_matrix[i,i-width]
+    row <- sapply(list(x,y,mr,ml,mu,md),function(e) if(class(e)!="numeric") 0 else e)
+    params[i,] <- row
+  }
+  colnames(params) <- c("x","y","mr","ml","mu","md")
+  return(params)
+}
+
+plot_params_to_mig_matrix <- function(plot_params,to_file=T,outpath){
+  npops <- nrow(plot_params)
+  width <- sqrt(npops)
+  migmat <- matrix(data=0,nrow=npops,ncol=npops)
+  for(i in 1:nrow(plot_params)){
+    row <- plot_params[i,]
+    if(i+1<=npops) migmat[i,i+1] <- row$mr
+    if(i-1>0) migmat[i,i-1] <- row$ml
+    if(i+width<=npops) migmat[i,i+width] <- row$mu
+    if(i-width>0) migmat[i,i-width] <- row$md
+  }
+  if(to_file) write.table(migmat,outpath,col.names = F,row.names = F)
[…]

### 2019-04-17 scripts/plot_sumstats.R

@@ -1,0 +1,190 @@
+setwd("~/spaceness/")
+library(data.table);library(ggplot2);library(magrittr);library(cowplot);library(plyr);library(wesanderson)
+theme_set(theme_classic()+theme(axis.text=element_text(size=7),
+                                axis.title=element_text(size=7),
+                                strip.text = element_text(size=7),
+                                title = element_text(size=8,hjust=0.5),
+                                legend.title = element_text(size=7),
+                                legend.text = element_text(size=7),
+                                strip.background = element_blank()))
+discrete_palette <- c("steelblue1","forestgreen","chartreuse3","slateblue3")
+
+
+spat <- fread("sumstats/ss_spatial.txt")
+spat$type <- "spatial\nmate choice\n"
+spat$popsize <- fread("sumstats/popsize_spatial.txt",header = F)$V1
+rm <- fread("sumstats/ss_random_mating.txt")
+rm$type <- "random\nmating\n"
+rm$popsize <- fread("sumstats/popsize_random_mating.txt",header = F)$V1
+
+pd <- rbind(spat,rm)
+names(pd) <- c("dispersal","segsites","pi","thetaW","tajD","het_o","Fis",paste0(0:100),"type","Census N")
+sfs <- pd[,8:(ncol(pd)-1)]
+sfs$dispersal <- pd$dispersal
+sfs$segsites <- pd$segsites
+sfs$type <- pd$type
+pd$mean_neighbors <- ((pi*pd$dispersal^2)/16^2)*1000
+pd$max_neighbors <- (3*((pi*pd$dispersal^2)/16^2))*1000
+pd <- melt(pd[,c("segsites","pi","tajD","het_o","Fis","dispersal","type","Census N","mean_neighbors","max_neighbors")],
+           id.vars=c("dispersal","type","mean_neighbors","max_neighbors"))
+
+#summary stat plot
+pd$variable <- factor(pd$variable,levels=c("Census N","segsites","pi","tajD","het_o","Fis"))
+ss_by_dispersal <- ggplot(data=pd,aes(x=dispersal/16*100,y=value,fill=type,col=type))+
+  theme_classic()+
+  theme(axis.title.y=element_blank(),
[…]

### 2019-04-17 scripts/regress_sigma.py

@@ -1,0 +1,52 @@
+#extra trees regression on mate selection radius sd
+import pandas, numpy as np, sklearn, os
+from sklearn import model_selection
+from sklearn import ensemble
+import pickle
+import matplotlib.pyplot as plt
+#os.chdir("/projects/kernlab/cbattey2/pytheas/nwf/")
+os.chdir("/Users/cj/spaceness")
+
+############################ train extra trees classifier ###########################
+#load summary stats
+ss=np.loadtxt("sumstats/ss_msp_grid.txt",skiprows=1)
+ss=pandas.DataFrame(ss)
+ss=ss.drop(6,1)
+ss=np.array(ss)
+statnames=open("sumstats/ss_msp_grid.txt","r").readline().split(" ")
+
+rand_order=np.random.choice(np.arange(0,len(ss),1),len(ss),replace=False)
+ss=ss[rand_order,:]
+train_stats=ss[0:int(len(ss)*.9),1:]
+test_stats=ss[int(len(ss)*.9)+1:len(ss),1:]
+train_lab=ss[0:int(len(ss)*.9),0]
+test_lab=ss[int(len(ss)*.9)+1:len(ss),0]
+
+#train extra trees regressor
+param_grid_forest = {"max_depth": [3, 5, 10,50],
+                     "min_samples_split": [2, 3, 10, 50],
+                     "min_samples_leaf": [1, 3, 10, 50]}
+clf = sklearn.ensemble.ExtraTreesRegressor(n_estimators=100,bootstrap=False)
+clf_grid = sklearn.model_selection.GridSearchCV(clf,
+                                            param_grid=param_grid_forest,
+                                            cv=3,n_jobs=10)
+clf_grid.fit(train_stats, train_lab)
+clf_grid.best_params_
+pytheas=clf_grid.best_estimator_
+
+r2=pytheas.score(test_stats,test_lab)
+pred=pytheas.predict(test_stats)
+
+fig = plt.figure(figsize=(2,4),dpi=200)
+plt.rcParams.update({'font.size': 7})
+ax1=fig.add_axes([0,0.55,1,.45])
+ax1.plot(test_lab,test_lab,"-",color="k")
+ax1.plot(test_lab,pred,"o",color="green")
+ax1.set_ylabel("Predicted Sigma")
+ax1.set_xlabel("Simulated Sigma")
[…]

### 2019-04-17 scripts/run_gwas.py

@@ -4,6 +4,8 @@ import argparse
 import allel
 import scipy
 import random
+from matplotlib import pyplot as plt
+import pandas
 
 parser = argparse.ArgumentParser(description='Run a GWAS in plink on individuals \
                                               sampled from a SLiM simulation.\
@@ -21,9 +23,8 @@ parser.add_argument('--nSamples', dest='nSamples', type=int,
 parser.add_argument('--sampling',dest='sampling',
                    help='Type of sampling to use. Options: \
                    random (nSamples random individuals). \
-                   point (nSamples individuals closest to sampling_location). \
-                   pair (nSamples individuals closest to sampling_location \
-                   and its reflection over the x axis)')
+                   point (nSamples individuals from 4 points on the map). \
+                   midpoint (nSamples individuals closest to the midpoint)')
 parser.add_argument('--mu', dest='mu',type=float,
                    help='mutation rate per base per unit time (msprime generations or SLiM time steps)')
 parser.add_argument('--phenotype', dest='phenotype',
@@ -36,6 +37,8 @@ parser.add_argument('--phenotype_mean', dest='phenotype_mean',type=float,
                    help='mean for gaussian phenotypes')
 parser.add_argument('--phenotype_sd', dest='phenotype_sd',type=float,
                    help='standard deviation for gaussian phenotypes')
+parser.add_argument('--gentimes',dest='gentimes',
+                    help='path to a file with sigmas in the first column and generation times in the second.')
 parser.add_argument('--seed', dest='seed',type=int,
                    help='Random seed for sampling individuals and adding mutations to tree sequences.')
 
@@ -45,12 +48,14 @@ args=parser.parse_args()
 # args = argparse.Namespace(infile='/Users/cj/spaceness/sims/slimout/spatial/W50_run3/sigma_0.5196493297511975_.trees_3918732',
 #                           outdir="/Users/cj/Desktop/",
 #                           plink_path="plink",
+#                           sampling="point",
 #                           vcftools_path="vcftools",
-#                           nSamples=10000,
[…]

### 2019-04-17 scripts/run_gwas_local.py

@@ -1,0 +1,119 @@
+from slimtools import *
+import msprime as msp,pyslim,numpy as np,os,subprocess as sp
+import argparse
+import allel
+
+
+files=os.listdir("/projects/kernlab/cbattey2/spaceness/gwas/sims/")
+
+for i in files:
+    print("processing file"+str(i))
+    args = argparse.Namespace(treeseq=os.path.join("/projects/kernlab/cbattey2/spaceness/gwas/sims/",i),
+                              outdir="/projects/kernlab/cbattey2/spaceness/gwas/out/",
+                              plink_path="/projects/kernlab/cbattey2/plink-1.07-x86_64/plink",
+                              vcftools_path="/projects/kernlab/cbattey2/vcftools_0.1.13/bin/vcftools",
+                              nSamples=1000,
+                              mu=0.25e-8,
+                              phenotype="transform_coord")
+
+    #sample individuals and add mutations
+    simname=os.path.basename(args.treeseq)
+    ts=pyslim.load(args.treeseq)
+    sample_inds=np.unique([ts.node(j).individual for j in ts.samples()]) #final-generation individuals
+    subsample=np.random.choice(sample_inds,args.nSamples,replace=False) #get nSamples random sample inds
+    subsample_nodes=[ts.individual(x).nodes for x in subsample] #node numbers for sampled individuals
+    subsample_nodes=[a for b in subsample_nodes for a in b] #flatten the list
+    subsample_nodes=np.sort(np.array(subsample_nodes))
+    ts=ts.simplify(subsample_nodes)
+    ts=msp.mutate(ts,args.mu)
+
+    #get haplotypes and locations
+    haps=ts.genotype_matrix()
+    sample_inds=np.unique([ts.node(j).individual for j in ts.samples()])
+    locs=[[ts.individual(x).location[0],ts.individual(x).location[1]] for x in sample_inds]
+
+    #run a PCA
+    genotype_counts=allel.HaplotypeArray(haps).to_genotypes(ploidy=2).to_allele_counts()
+    pca=allel.pca(genotype_counts[:,:,0])
+    pcfile=open(os.path.join(args.outdir,simname)+".pca","w")
+    for i in range(args.nSamples):
+        pcfile.write("msp_"+str(i)+" "+"msp_"+str(i)+" ")
+        for j in range(10):
+            pcfile.write(str(pca[0][i][j])+" ")
+        pcfile.write("\n")
[…]

### 2019-04-17 scripts/run_msp_grid.py

@@ -1,0 +1,139 @@
+from slimtools import *
+# import argparse
+# parser=argparse.ArgumentParser()
+# parser.add_argument("--infile")
+# parser.add_argument("--outfile")
+# args=parser.parse_args()
+
+
+#######################utility functions####################################
+##migration matrix for square landscapes with or without migration along diagonals
+def get_grid_migration_matrix(width,rate,diagonal):
+    mig_matrix=np.zeros((width*width,width*width))
+    lefts=np.arange(0,width**2,width)
+    rights=np.arange(width-1,width**2,width)
+    for i in np.arange(0,width**2,1):
+        if i in lefts:
+            mig_matrix[i][i+1]=rate
+            if i+width <= width**2-1:
+                mig_matrix[i][i+width]=rate
+            if i-width >= 0:
+                mig_matrix[i][i-width]=rate
+            if diagonal:
+                if i-(width-1) >= 0:
+                    mig_matrix[i][i-(width-1)]=np.sqrt(2)*rate
+                if i+(width+1) <= width**2-1:
+                    mig_matrix[i][i+(width+1)]=np.sqrt(2)*rate
+        if i in rights:
+            mig_matrix[i][i-1]=rate
+            if i+width <= width**2-1:
+                mig_matrix[i][i+width]=rate
+            if i-width >= 0:
+                mig_matrix[i][i-width]=rate
+            if diagonal:
+                if i-(width+1) >= 0:
+                    mig_matrix[i][i-(width+1)]=np.sqrt(2)*rate
+                if i+(width-1) <= width**2-1:
+                    mig_matrix[i][i+(width-1)]=np.sqrt(2)*rate
+        if i not in lefts and i not in rights:
+            if i+1 <= width**2-1:
+                mig_matrix[i][i+1]=rate
+            if i-1 >= 0:
+                mig_matrix[i][i-1]=rate
+            if i+width <= width**2-1:
+                mig_matrix[i][i+width]=rate
+            if i-width >= 0:
+                mig_matrix[i][i-width]=rate
+            if diagonal:
+                if i-(width+1) >= 0:
+                    mig_matrix[i][i-(width+1)]=np.sqrt(2)*rate
+                if i+(width-1) <= width**2-1:
+                    mig_matrix[i][i+(width-1)]=np.sqrt(2)*rate
[…]

### 2019-04-17 scripts/run_msp_sim.py

@@ -1,0 +1,17 @@
+#run msprime simulations with Ne equal to census sizes from spatial SLiM simulations
+import pandas, numpy as np, msprime, os
+os.chdir("/Users/cj/spaceness/sumstats/")
+popsizes=np.loadtxt("ss_spatial_random_W50.txt.popsizes")
+popsizes[400,0]
+
+for i in range(len(popsizes)):
+    trees=msprime.simulate(sample_size=100,
+                           Ne=popsizes[i,1]*2,
+                           recombination_rate=1e-9,
+                           mutation_rate=0,
+                           length=1e8)
+    outfile=("/Users/cj/spaceness/sims/msp/sigma_"+
+             str(popsizes[i,0])+
+             "_.trees_"+
+             str(int(np.random.uniform(1e7,1e8-1,1))))
+    trees.dump(outfile)

### 2019-04-17 scripts/run_msp_sims.py

@@ -1,0 +1,16 @@
+#run msprime simulations with Ne equal to census sizes from spatial SLiM simulations
+import pandas, numpy as np, msprime, os
+os.chdir("/Users/cj/spaceness/sumstats/")
+popsizes=np.loadtxt("ss_spatial_random_W50.txt.popsizes")
+
+for i in range(len(popsizes)):
+    trees=msprime.simulate(sample_size=120,
+                           Ne=popsizes[i,1],
+                           recombination_rate=1e-9,
+                           mutation_rate=0,
+                           length=1e8)
+    outfile=("/Users/cj/spaceness/sims/msp/sigma_"+
+             str(popsizes[i,0])+
+             "_.trees_"+
+             str(int(np.random.uniform(1e7,1e8-1,1))))
+    trees.dump(outfile)

### 2019-04-17 scripts/run_smcpp.py

@@ -1,0 +1,76 @@
+#run smcpp analysis on an input tree sequence
+import msprime,pyslim,os,itertools,re,glob
+import numpy as np
+import subprocess as sp
+from slimtools import *
+import argparse
+
+parser=argparse.ArgumentParser()
+parser.add_argument("--infile",dest="infile")
+parser.add_argument("--gentimes",dest="gentimes")
+parser.add_argument("--outdir",dest="outdir")
+args=parser.parse_args()
+
+#debug
+# args=argparse.Namespace(infile = "/Users/cj/spaceness/sims/slimout/spatial/W50_run3/sigma_0.3332292532035142_.trees_2054475",
+#                         outdir = "/Users/cj/Desktop/")
+
+#temporary folder for intermediate output
+simname=os.path.basename(args.infile)
+label=float(re.split("_",simname)[1])
+id=re.split("_",simname)[3]
+tmpout=os.path.join(args.outdir,id)
+sp.run(["mkdir",tmpout])
+
+#load treeseq, sample, mutate, export VCF, compress, index
+ts=pyslim.load(args.infile)
+ts=sample_treeseq(infile=ts,
+                  outfile="",
+                  nSamples=20,
+                  sampling="random",
+                  recapitate=False,
+                  recombination_rate=1e-9,
+                  write_to_file=False,
+                  sampling_locs="",
+                  plot=False,
+                  seed=12345)
+gentimes=np.loadtxt(args.gentimes)
+gentime=[x[0] for x in gentimes if np.round(x[1],5)==np.round(label,5)]
+
+ts=msp.mutate(ts,args.mu/gentime[0],random_seed=98675)
+
+ts.write_vcf(open(os.path.join(tmpout,simname)+".vcf","w"),2)
+sp.check_output(["bgzip",os.path.join(tmpout,simname)+".vcf"])
+sp.check_output(["tabix",os.path.join(tmpout,simname)+".vcf"+".gz"])
+
+#convert VCF to smcpp allele counts format, looping over randomly selected "distinguished individuals"
+inds=",".join(["msp_"+str(x) for x in range(20)])
+#for i in range(5):
+command=["smc++",
+         "vcf2smc",
+         #"-d","msp_"+str(int(np.random.uniform(0,19,1))),"msp_"+str(int(np.random.uniform(0,19,1))),
+         "--length","100000000",
[…]

### 2019-04-17 scripts/sample_and_mutate_treeseqs.py

@@ -1,0 +1,28 @@
+from slimtools import *
+
+#check treeseq directory for coalescence
+treeseqs,out=check_treeseq_coalescence("/Users/cj/spaceness/sims/slimout/wf/")
+np.mean(out[0]) #mean proportion uncoalesced trees
+np.mean(out[1]) #mean roots per tree
+
+sample_treeseq_directory(indir="/Users/cj/spaceness/sims/slimout/wf/",
+                         outdir="/Users/cj/spaceness/sims/sampled/wf/",
+                         nSamples=50,
+                         recapitate=False,
+                         recombination_rate=1e-8)
+
+mutate_treeseqs("/Users/cj/spaceness/sims/sampled/wf",
+                "/Users/cj/spaceness/sims/mutated/wf",
+                4e-8)
+
+haps,positions,labels,locs=get_ms_outs("/Users/cj/spaceness/sims/mutated/wf",False)
+
+positions=discretize_snp_positions(positions)
+
+popsizes=get_pop_sizes(indir="/Users/cj/spaceness/sims/slimout/wf/",
+              outpath="/Users/cj/spaceness/sumstats/popsize_wfN1000.txt")
+
+ss=getHaplotypeSumStats(haps,positions,labels,locs,
+                        outfile="/Users/cj/spaceness/sumstats/sumstats_wf_N1000.txt",
+                        ibs_tracts=True,
+                        verbose=True,maxlen=1e8)

### 2019-04-17 scripts/sample_treeseqs.py

@@ -1,0 +1,40 @@
+def sample_treeseq_directory(indir,
+                             type="random",
+                             outdir,
+                             nSamples,
+                             recapitate,
+                             recombination_rate,
+                             n_localities,
+                             width):
+    '''
+    loop sample_treeseq over a directory.
+    '''
+    trees=os.listdir(indir)
+    for i in range(len(trees)):
+        ts=pyslim.load(os.path.join(indir,trees[i]))
+        if(type=="random"):
+            sample_inds=np.unique([ts.node(j).individual for j in ts.samples()]) #final-generation individuals
+            subsample=np.random.choice(sample_inds,nSamples,replace=False) #get nSamples random sample inds
+            subsample_nodes=[ts.individual(x).nodes for x in subsample] #node numbers for sampled individuals
+            subsample_nodes=[a for b in subsample_nodes for a in b] #flatten the list
+            subsample_nodes=np.sort(np.array(subsample_nodes))
+            o=ts.simplify(subsample_nodes)
+        else if(type=="grid"):
+            sampling_locs=width/n_localities
+
+        if(recapitate):
+            ts=ts.recapitate(recombination_rate=recombination_rate)
+        o.dump(os.path.join(outdir,trees[i]))
+
+    return None
+
+
+indir="/Users/cj/spaceness/sims/slimout/k5/"
+type="grid"
+nSamples=50
+recapitate=True
+recombination_rate=1e-9
+n_localities=4
+width=35
+
+sample_treeseq_directory()

### 2019-04-17 scripts/slim_ts_to_stairwayplot.py

@@ -1,0 +1,151 @@
+import msprime as msp, numpy as np, allel, argparse, subprocess,os,re
+from os.path import basename, splitext, join
+from slimtools import *
+
+parser = argparse.ArgumentParser(
+  description='Run stairwayplot on a tree sequence output from msprime.\
+               Requires gnu-parallel and Stairwayplot including \
+               the command-line version of Stairway_plot_output_summary,\
+               which can be downloaded from: \
+               https://sites.google.com/site/jpopgen/stairway-plot.')
+parser.add_argument('--infile',dest='infile',
+                    help='path to tree sequence')
+parser.add_argument('--sampling',dest='sampling',
+                    help='sampling scheme (random, midpoint, or point).')
+parser.add_argument('--outdir',dest='outdir',
+                    help='full path to directory for writing output files')
+parser.add_argument('--nboots',dest='nboots',type=int,
+                    help='number of bootstrap replicates to run.')
+parser.add_argument('--stairwayplot_dir',dest='stairwayplot_dir',
+                    help="path to stairwayplot directory")
+parser.add_argument('--cores',dest='cores',
+                    help="number of cores to use when fitting demographic \
+                    models.")
+parser.add_argument('--seed',dest='seed',type=int,
+                    help="random number seed for bootstrap resampling.")
+parser.add_argument('--mutation_rate',dest='mutation_rate',type=float,
+                    help="mutation rate (per base per generation)")
+parser.add_argument('--gentimes_file',dest='gentimes_file',
+                    help='path to file listing sigmas and generation times.')
+parser.add_argument('--generation_time',dest='generation_time',type=float,
+                    help="generation time for plotting")
+parser.add_argument('--plot',dest='plot',type=bool,
+                    help='plot Ne ~ t ? T/F')
+args=parser.parse_args()
+
+np.random.seed(args.seed)
+
+#read in and sample treeseq
+simname=os.path.basename(args.infile)
+label=float(re.split("_",simname)[1])
[…]

### 2019-04-17 scripts/slim_v_msp_mutation_check.R

@@ -1,0 +1,62 @@
+#plots to compare msprime and slim mutation effects on summary stats
+setwd("~/spaceness/sumstats")
+
+slim_spat <- fread("ss_spatial_random_W16muts.txt")
+slim_spat$model <- "spatial"
+slim_spat$muts <- "slim"
+msp_spat <- fread("ss_spatial_random_W16_msp_muts.txt")
+msp_spat$model <- "spatial"
+msp_spat$muts <- "msp"
+slim_rand <- fread("ss_random_random_W16muts.txt")
+slim_rand$model <- "random"
+slim_rand$muts <- "slim"
+msp_rand <- fread("ss_random_random_W16_msp_muts.txt")
+msp_rand$model <- "random"
+msp_rand$muts <- "msp"
+ss <- rbind(slim_spat,msp_spat,slim_rand,msp_rand)
+ss <- data.frame(ss)[,!grepl("sfs",colnames(ss))]
+
+popsize_spat <- fread("ss_spatial_random_W16muts.txt.popsizes")
+popsize_rand <- fread("ss_random_random_W16muts.txt.popsizes")
+popsize <- rbind(popsize_spat,popsize_rand)
+colnames(popsize) <- c("sigma","census_n")
+ss <- merge(ss,popsize,by="sigma")
+ss$theta <- 4*(ss$census_n/2)*1e-8
+ss$census_n2 <- ss$census_n
+
+mss <- melt(ss,id.vars=c("model","muts","sigma","census_n2"))
+pd <- subset(mss,variable %in% c("pi","theta"))
+p <- ggplot(data=pd,aes(x=sigma,y=value,col=paste(variable,model)))+
+  ggtitle("Ne estimates for constant mu applied in msprime v SLiM")+
+  facet_wrap(~muts)+ylab("")+
+  scale_color_brewer(palette = "Paired")+
+  geom_point(shape=21,size=0.8,stroke=0.2)+
+  geom_smooth(fill=NA,lwd=0.5,span=0.6)
+
+pdf("~/spaceness/figures/msp_v_slim_mutations.pdf",width=6,height=2.5,useDingbats = F)
+print(p)
+dev.off()
+
+
+
+mss$variable <- factor(mss$variable,levels=c("segsites","pi","thetaW","theta","tajD",
+                                             "het_o","fis","gen_dist_var","gen_dist_skew",
+                                             "ibs_mean","ibs_var","ibs_skew","ibs_95p",
[…]

### 2019-04-17 scripts/slimtools.py

@@ -140,7 +140,7 @@ def sample_treeseq(infile,
     recombination_rate - float. Recombination rate to use for recapitation, in recombination events per base per generation.
     write_to_file - boolean.
     outfile - string. File path for output.
-    sampling_locs - list. locations to draw samples from when using sampling='point', as [[x1,y1],[x2,y2]]
+    sampling_locs - list. locations to draw samples from when using sampling='point', as [[x1,y1],[x2,y2],...]
     plot - boolean. plot location of samples?
     seed - random seed for all numpy operations. 0 uses the system default.
     '''
@@ -176,7 +176,7 @@ def sample_treeseq(infile,
         inds=np.array([x.id for x in ts.individuals()])
         middle=[np.mean(locs[:,0]),np.mean(locs[:,1])]
         dists_to_middle=[scipy.spatial.distance.euclidean(locs[x],middle) for x in range(len(locs))]
-        weights=[1/x**4 for x in dists_to_middle]
+        weights=[1/x**2 for x in dists_to_middle]
         weights=weights/np.sum(weights)
         subsample=np.random.choice(inds,nSamples,p=weights,replace=False)
         #closest_inds=sorted(range(len(dists_to_middle)), key=lambda e: dists_to_middle[e],reverse=True)[-nSamples*2:] #this works for sampling half the closest nSamples individuals
@@ -338,7 +338,7 @@ def getSLiMSumStats(haps,positions,label,locs,outfile,maxlen,ibs_tracts=True,ver
     thetaW=allel.watterson_theta(pos=positions,ac=allele_counts,start=1,stop=maxlen)
     het_o=np.mean(allel.heterozygosity_observed(genotypes))
     fis=np.mean(allel.inbreeding_coefficient(genotypes))
-    sfs=allel.sfs(allele_counts[:,1]) #last entry seems to be the highest *non-zero* SFS entry (wtf?) so adding zeros
+    sfs=allel.sfs(allele_counts[:,1])
     sfs=np.append(sfs,[np.repeat(0,np.shape(haps)[1]-len(sfs)+1)])
     #Isolation by distance
     if(verbose):
@@ -348,8 +348,8 @@ def getSLiMSumStats(haps,positions,label,locs,outfile,maxlen,ibs_tracts=True,ver
                                 start=0,
                                 stop=maxlen)
[…]

### 2019-04-17 scripts/smcpp_msp_tests.py

@@ -1,0 +1,63 @@
+"""
+Utilities for working with scm++
+"""
+import logging
+import subprocess
+import tskit
+import msprime
+import os
+def write_smcpp_file(path):
+    """
+    Writes a smcpp input file given a treesequence
+    """
+    ts = tskit.load(path)
+    # write a vcf intermediate input
+    with open(path+".vcf", "w") as vcf_file:
+        ts.write_vcf(vcf_file, 2)
+    # index the vcf
+    cmd = f"bgzip {path}.vcf"
+    logging.info("Running:" + cmd)
+    subprocess.run(cmd, shell=True, check=True)
+    vz_file = f"{path}.vcf.gz"
+    cmd = f"tabix {vz_file}"
+    logging.info("Running:" + cmd)
+    subprocess.run(cmd, shell=True, check=True)
+    # run smc++ for vcf conversion
+    smc_file = f"{path}.smc.gz"
+    cmd = f"smc++ vcf2smc {vz_file} {smc_file} 1 pop1:"
+    for n in range(ts.num_samples // 2):
+        cmd = cmd + f"msp_{n},"
+    cmd = cmd[0:-1]
+    logging.info("Running:" + cmd)
+    subprocess.run(cmd, shell=True, check=True)
+
+
+def run_smcpp_estimate(input_file, mutation_rate, ncores):
+    """
+    Runs smc++ estimate on the specified file, resulting in the output being written
+    to the file input_file.final.jason".
+    """
+    cmd = (
+        f"smc++ estimate "
+        f"{mutation_rate} {input_file} --base {input_file} --cores {ncores}")
+    logging.info("Running:" + cmd)
+    subprocess.run(cmd, shell=True, check=True)
+
+
+def run_smcpp_plot(input_file, generation_time):
+    """
+    Runs smc++ plot on the specified file, resulting in the output being written
+    to the file input_file.png".
+    """
+    cmd = (
+        f"smc++ plot {input_file}.png {input_file} -g {generation_time} -c")
+    logging.info("Running:" + cmd)
+    subprocess.run(cmd, shell=True, check=True)
+
+os.chdir("/home/cbattey2/spaceness/demography/msp_sims/")
+for i in range(10):
+    t=msprime.simulate(sample_size=40,Ne=6.1e3,mutation_rate=1e-8,recombination_rate=1e-9,length=1e8)
+    t.dump(str(i)+".trees")
[…]

### 2019-04-17 scripts/smcpp_slim_pipeline.py

@@ -1,0 +1,85 @@
+"""
+Utilities for working with scm++
+"""
+import logging
+import subprocess
+import tskit
+import msprime
+import os
+from slimtools import *
+def write_smcpp_file(path):
+    """
+    Writes a smcpp input file given a treesequence
+    """
+    ts = tskit.load(path)
+    # write a vcf intermediate input
+    with open(path+".vcf", "w") as vcf_file:
+        ts.write_vcf(vcf_file, 2)
+    # index the vcf
+    cmd = f"bgzip {path}.vcf"
+    logging.info("Running:" + cmd)
+    subprocess.run(cmd, shell=True, check=True)
+    vz_file = f"{path}.vcf.gz"
+    cmd = f"tabix {vz_file}"
+    logging.info("Running:" + cmd)
+    subprocess.run(cmd, shell=True, check=True)
+    # run smc++ for vcf conversion
+    smc_file = f"{path}.smc.gz"
+    cmd = f"smc++ vcf2smc {vz_file} {smc_file} 1 pop1:"
+    for n in range(ts.num_samples // 2):
+        cmd = cmd + f"msp_{n},"
+    cmd = cmd[0:-1]
+    logging.info("Running:" + cmd)
+    subprocess.run(cmd, shell=True, check=True)
+
+
+def run_smcpp_estimate(input_file, mutation_rate, ncores):
+    """
+    Runs smc++ estimate on the specified file, resulting in the output being written
+    to the file input_file.final.jason".
+    """
+    cmd = (
+        f"smc++ estimate "
+        f"{mutation_rate} {input_file} --base {input_file} --cores {ncores}")
+    logging.info("Running:" + cmd)
+    subprocess.run(cmd, shell=True, check=True)
+
+
+def run_smcpp_plot(input_file, generation_time):
+    """
+    Runs smc++ plot on the specified file, resulting in the output being written
+    to the file input_file.png".
+    """
+    cmd = (
+        f"smc++ plot {input_file}.png {input_file} -g {generation_time} -c")
+    logging.info("Running:" + cmd)
+    subprocess.run(cmd, shell=True, check=True)
+
+simdir="/home/cbattey2/spaceness/sims/slimout/spatial/W50_run3"
+outdir="/home/cbattey2/spaceness/demography/smcpp/random_sampling/"
+
+os.chdir(outdir)
+sims=os.listdir(simdir)
+#f=sims[1]
+for f in sims:
[…]

### 2019-04-17 scripts/spaceness_runtime.sh

@@ -1,0 +1,7 @@
+#!/bin/bash
+for sigma in `seq 0.2 .25 3`
+do
+echo $sigma
+slim -t -d sigma=$sigma -d outpath="'~/Desktop/test'" slim_recipes/spaceness.slim | grep "CPU time used"
+done
+

### 2019-04-17 scripts/spaceness_viz.py

@@ -1,6 +1,5 @@
 from slimtools import *
 from numpy.polynomial.polynomial import polyfit
-ts=pyslim.load("/Users/cj/spaceness/sims/slimout/spatial/W35/sigma_0.3003441596629088_.trees")
 maxlen=1e8
 ts=sample_treeseq(infile="/Users/cj/spaceness/sims/slimout/spatial/W35/sigma_1.065973462406566_.trees",
                outfile="",

### 2019-04-17 scripts/spatial_slim_ts.py

@@ -1,0 +1,399 @@
+import pyslim, msprime
+import numpy as np
+import tqdm
+import scipy.sparse as sparse
+
+class SpatialSlimTreeSequence(pyslim.SlimTreeSequence):
+
+    def __init__(self, ts, dim=3):
+        super(SpatialSlimTreeSequence, self).__init__(ts)
+        self.dim = dim
+
+    def individual_distance_to_point(self, point):
+        """
+        Returns the array of distances of each individual's location to the point
+        whose coordinates are given by `point`.
+        :param float point: The coordinates of the point. If this has length less than
+            three, remaining coordinates will be set to 0.0.
+        """
+        point = np.concatenate((np.array(point), np.zeros(self.dim - len(point))))
+        if point.shape != (self.dim,):
+            raise ValueError("point not of the correct shape: must be coercible to"
+                             + "a vector of length self.dim or less.")
+
+        return np.sqrt(np.sum((self.individual_locations() - point) ** 2, axis=1))
+
+    def individuals_in_circle(self, center, radius, time=None):
+        """
+        Returns the IDs of individuals at distance less than or equal to `radius` of `center.
+        :param float center: The coordinates of the center.
+        :param float radius: The radius of the circle.
+        """
+        dists = self.individual_distance_to_point(center)
+        these = np.logical_and(dists <= radius, self.individuals_alive(time))
+        return np.where(these)[0]
+
+    def individuals_alive(self, time):
+        """
+        Returns a logical array of length equal to the number of individuals that
+        indicates whether the given individual was alive at the given time.
+        TODO: Use the pyslim.INDIVIDUAL_ALIVE flag?
+        :param float time: The time ago. If `time` is None, then everyone will be
+            counted as being alive.
+        """
+        if time is None:
+            out = np.repeat(True, self.num_individuals)
+        else:
+            births = self.individual_times()
+            ages = self.individual_ages()
+            out = np.logical_and(births >= time, births - ages <= time)
+        return out
+
+    def individuals_age(self, time):
+        """
+        Returns the *ages* of each individual at the corresponding time ago,
+        which will be `nan` if the individual is either not born yet or dead.
+        This is computed as the time ago the indivdiual was born (found by the
+        `time` associated with the the indivdiual's nodes) minus the `time`
+        argument; while "death" is inferred from the individual's `age`,
+        recorded in metadata.
+        :param float time: The reference time ago.
+        """
[…]

### 2019-04-17 scripts/sumstats_from_treeseq.py

@@ -17,18 +17,21 @@ parser.add_argument('--sampling',dest='sampling',
                    midpoint (nSamples individuals proportional to distance from the midpoint). \
                    point (nSamples/len(sampling_locs) individuals proportional to distance from each sampling location. \
                    and its reflection over the x axis)')
-parser.add_argument('--mu', dest='mu',
+parser.add_argument('--mu', dest='mu',type=float,
                    help='mutation rate per base per unit time in SLiM simulations.')
 parser.add_argument('--sampling_locs', dest='sampling_locs',
                    help='Currently not active - see hard coding in sample_treeseq call. \
                    Sampling site location given as "x1,y2 x2,y2 x3,y3..."')
+parser.add_argument('--gentimes',dest='gentimes',
+                    help='path to a file with sigmas in the first column and generation times in the second.')
 args=parser.parse_args()
 
 #params for debugging
-# args=argparse.Namespace(infile='/Users/cj/spaceness/sims/slimout/spatial/W35/sigma_0.2130878654199541_.trees',
+# args=argparse.Namespace(infile='/Users/cj/spaceness/sims/slimout/spatial/W50_run3/sigma_0.70554722351671_.trees_3600681',
 #                         outfile='/Users/cj/Desktop/test_sumstats.txt',
 #                         sampling='random',
-#                         mu=0.25e-8,
+#                         mu=1e-8,
+#                         gentimes="/Users/cj/spaceness/W50sp_gentimes.txt",
 #                         seed=12345)
 
 print("sampling individuals and adding mutations")
@@ -40,19 +43,23 @@ out=open(args.outfile+".popsizes",'a')
 out.write(str(label)+" "+str(nind)+"\n")
 out.close()
 
+#get generation time estimated from short all-individual simulations (see ~/spaceness/scripts/estimate_gen_times.py)
+gentimes=np.loadtxt(args.gentimes)
+gentime=[x[0] for x in gentimes if np.round(x[1],5)==np.round(label,5)]
+
 #sample and mutate treeseq
 ts=sample_treeseq(infile=args.infile,
                outfile="",
-               nSamples=100,
+               nSamples=60,
                recapitate=False,
                recombination_rate=1e-8,
                write_to_file=False,
                sampling=args.sampling,
-               sampling_locs=[[8,8],[27,27],[27,8],[8,27]],
[…]

### 2019-04-17 scripts/testmuts.py

@@ -1,0 +1,32 @@
+from slimtools import *
+
+muts=pyslim.load("/Users/cj/Desktop/sigma_0.793843124378123_.trees_7796460_822000")
+muts.num_mutations
+muts=sample_treeseq(infile=muts,
+               outfile="",
+               nSamples=60,
+               recapitate=False,
+               recombination_rate=1e-8,
+               write_to_file=False,
+               sampling="random",
+               sampling_locs=[[12.5,12.5],[12.5,37.5],[37.5,37.5],[37.5,12.5]],
+               plot=False,
+               seed=12345)
+muts.num_mutations
+
+nomuts=pyslim.load("/Users/cj/spaceness/sims/slimout/random_mating/W50_run3/sigma_0.7857423531825174_.trees_9690340")
+nomuts=msp.mutate(nomuts,1e-8/4.8)
+nomuts.num_mutations
+nomuts=sample_treeseq(infile=nomuts,
+               outfile="",
+               nSamples=60,
+               recapitate=False,
+               recombination_rate=1e-8,
+               write_to_file=False,
+               sampling="random",
+               sampling_locs=[[12.5,12.5],[12.5,37.5],[37.5,37.5],[37.5,12.5]],
+               plot=False,
+               seed=12345)
+nomuts.num_mutations
+
+#seems fine so wtf is up with pi

### 2019-04-17 scripts/ts_to_stairwayplot.py

@@ -1,0 +1,78 @@
+from slimtools import *
+import subprocess
+import pandas
+import argparse
+#os.chdir("/Users/cj/spaceness")
+
+parser = argparse.ArgumentParser(
+            description='Convert slim tree sequences to stairwayplot input')
+parser.add_argument('--infile', dest='infile',
+                    help='full path to tree sequence')
+parser.add_argument('--outfile', dest='outfile',
+                   help='full path for output summary statistics.')
+parser.add_argument('--seed', dest='seed',type=int,
+                   help='Random seed for sampling individuals and adding mutations to tree sequences.')
+parser.add_argument('--sampling',dest='sampling')
+parser.add_argument('--mu', dest='mu',type=float,
+                   help='mutation rate per base per unit time in SLiM simulations.')
+args=parser.parse_args()
+
+#params for debugging
+# ts=msp.simulate(sample_size=120,Ne=1e4,mutation_rate=1e-8,recombination_rate=1e-8,length=1e8)
+# args=argparse.Namespace(infile='/Users/cj/spaceness/sims/slimout/spatial/W50_run3/sigma_0.2357863131214634_.trees_3600449',
+#                         outfile='/Users/cj/Desktop/test.txt',
+#                         sampling='random',
+#                         mu=1e-8,
+#                         seed=12345)
+
+#folder for input files
+subprocess.run("mkdir /Users/cj/spaceness/demography/stairwayplot/input/"+re.split("\\_",os.path.basename(args.infile))[1],
+               shell=True)
+
+#load tree sequence
+label=float(re.sub("sigma_|_.trees*|.trees","",os.path.basename(args.infile)))
+t=pyslim.load(args.infile)
+
+#get generation time estimated from short all-individual simulations
+gentimes=np.loadtxt("/Users/cj/spaceness/W50sp_gentimes.txt")
+gentime=[x[0] for x in gentimes if np.round(x[1],5)==np.round(label,5)]
+
+print("sigma="+str(label)+"  gentime="+str(gentime[0]))
+
+#sample and mutate treeseq
+ts=sample_treeseq(infile=args.infile,
+                  outfile="",
+                  nSamples=60,
+                  recapitate=False,
+                  recombination_rate=1e-8,
[…]

### 2019-04-24 ms/spaceness.tex
pre-PR edits

@@
 %Evolution on the Final Frontier: How Dispersal and Sampling Shape Inference From Genetic Data in Continuous Space
-\title{Effects of Continuous Spatial Structure on Analysis of Population Genetic Data}
+\title{Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data}
 
@@
 Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations.
-As a result many methods of analyzing genetic data assume that samples are a random draw from a random-mating population, but are applied to real populations that are structured over space by limited dispersal.
-Here we implement a forward-time simulation of evolution in continuous space and use it to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and association studies.
-We find that estimates of population size trajectories and many common summary statistics vary significantly with dispersal, with most bias occurring when Wright's neighborhood size is less than 100 and sampling is spatially clustered.
-In association studies the combination of spatial variation in the environment and limited dispersal creates spurious signals of genetic association with purely environmental phenotypes across much of the genome, and this bias is only partially corrected by regressing out principal components positions in the analysis.
-Last we describe the genealogical process leading to spatial effects we observed on simulated gentoypes, summarize estimates of neighborhood size from empirical studies across a range of organisms, and discuss the relevance of our simulation results for inference from genetic variation in real organisms.
+As a result many methods of analyzing genetic data assume that samples are a random draw from a randomly mating population, but are applied to real populations that are structured over space by limited dispersal.
+Here we implement a forward-time simulation of evolution in continuous space and use it to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and genome-wide association studies.
+We find that continuous space skews the distribution of several common summary statistics, with the most bias occurring when Wright's neighborhood size is less than 100 and sampling is spatially clustered.
+In turn estimates of population size history based on the site frequency spectrum become biased in the face of limited dispersal.
+We show that in the context of genome-wide association studies the combination of spatial variation in the environment and limited dispersal creates spurious signals of genetic association with purely environmental phenotypes across much of the genome and that this bias is only partially corrected by regressing out principal components of ancestry.
+Last we discuss the relevance of our simulation results for inference from genetic variation in real organisms.
 \end{abstract}
@@
 This pattern of "isolation by distance" \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, but classical mathematical models describing it are flawed approximations of the underlying process \citep{Felsenstein1975}.
-As a result many studies focus on describing geographic structure as a set of random-mating populations connected by migration \citep{Malecot1948,Wright1931}, and analyze variation within clusters of genetic variation inferred by programs like \textit{structure} \citep{Pritchard2000} as random-mating units.
+As a result many studies focus on describing geographic structure as a set of populations connected by migration \citep{Malecot1948,Wright1931}, and analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000} assuming these are randomly mating units.
 
 Though discrete models of between-population differentiation account for much of the genetic variation in species, the assumption that populations are randomly mating at some level has important implications for downstream inference of selection and demography.
-Methods based on the coalescent \citep{Kingman1982} assume that the sampled individuals are a random draw from a random mating population.
+Methods based on the coalescent \citep{Kingman1982} assume that the sampled individuals are a random draw from a randomly mating population.
 If dispersal or mate selection is limited by space these assumptions are violated -- nearby individuals will be more closely related than an average random pair and drawing multiple samples from the same point on the landscape will represent a biased sample of the genetic variation present in the whole population.
@@
 However this statistic is essentially a summary of the site frequency spectrum, which itself reflects variation in branch lengths of the underlying genealogy of sampled individuals.
-Because the spatial extent of dispersal and mate selection determines the rate of drift in a continuously distributed population \citep{Maruyama1972}, these genealogical summaries may also reflect variation in the scale of spatial interactions both over time and across populations.
-Similarly, the distribution of haplotype block lengths (i.e.
-stretches of identical sequence) among individuals is sometimes interpreted as reflecting historic demography \citep{Harris2013} or evidence of selection \citep{Garud2015}.
-However as this distribution is also an indirect estimate of the branch lengths of the sample genealogy it is also sensitive to variation in dispersal and mate selection \citep{Ringbauer2017,Baharian2016}.
+Because the spatial extent of dispersal and mate selection determines the rate of drift in a continuously distributed population \citep{Maruyama1972}, these population genetic summaries should also reflect variation in the scale of spatial interactions both over time and across populations.
+Similarly, the distribution of tract lengths of identity by state among individuals is often interpreted either through the lens of historic demography \citep{Harris2013} or selection \citep{Garud2015}.
+While that is so this distribution also contains abundant information about dispersal and mate selection \citep{Ringbauer2017,Baharian2016}.
+We are particularly keen on examining how such summaries will be affected by models that incorporate continuous space.
 
@@
 
-The issue of fine-scale structure for putatively "within-population" analyses is particularly important for studies of polygenic selection.
+The issue of fine-scale structure for putatively "within-population" analyses may have important implications for genome-wide association studies as well.
 This is because the genome-wide allele frequency variation caused by isolation by distance is similar to that expected when many sites in the genome are associated with a given phenotype \citep{Bulik-Sullivan2015}.
 When population structure and the environment covary across space genome-wide association studies (GWAS) may incorrectly infer phenotypic association at many SNPs that have differentiated by drift \citep{Mathieson2012}.
-Indeed two recent studies found that previous evidence of polygenic selection on human height in Europe was confounded by population structure \citep{Sohail2018,Berg2018}, suggesting that existing methods to correct for population structure in GWAS are insufficient.
-However we have little quantitative idea of the population and environmental parameters that can be expected to lead to bias in GWAS.
+This may be a particularly important effect when using GWAS to study selection on polygenic traits.
+Indeed two recent studies found that previous evidence of polygenic selection on human height in Europe was confounded by subtle population structure \citep{Sohail2018,Berg2018}, suggesting that existing methods to correct for population structure in GWAS are insufficient.
+However we have little quantitative idea of the population and environmental parameters that can be expected to lead to biases in GWAS.
 As the scale of sequence data now available for many species allows inference of increasingly fine-scale patterns of selection and demography, understanding how and when subtle spatial structure is likely to bias results is an important task for population genetics.
@@
 The model scales to chromosome-scale alignments across tens of thousands of individuals and outputs the full genealogy and recombination history of all final-generation individuals.
-We use this simulation to test how sampling strategy interacts with isolation by distance to cause systematic variation in population genetic summary statistics typically analyzed under discrete population models.
+We use this simulation to test how sampling strategy interacts with isolation by distance to cause systematic variation in population genetic summary statistics typically analyzed assuming discrete population models.
 We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum.
+Lastly we examine the impacts of evolution in continuous space on genome-wide association studies (GWAS) and identify regions of parameter space under which the results from GWAS may be misleading.
 
-Last we use our model to study the impacts of isolation by distance in continuous space on genome-wide association studies (GWAS).
+\ak{I think we can cut the following paragraph if we just add the last sentence above} 
+\cjb{agree - commenting out}
+%Last we use our model to study the impacts of isolation by distance in continuous space on genome-wide association studies (GWAS).
 Variation in ancestry proportions between case and control cohorts is expected to cause inflation of test statistics at SNPs with different allele frequencies among populations \citep{Price2006}.
@@
 For all simulations the full genealogy and recombination history of final-generation individuals were stored as tree sequences \citep{Kelleher2018}.
-\ak{add a link to the github repo with code}
+Scripts for figures and analyses are available at https://github.com/petrelharp/spaceness.
 
@@
 Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among samples.
-The full distribution of lengths of IBS tracts for each pair of individuals was first calculated with a custom python function \ak{point to code on github}.
+The full distribution of lengths of IBS tracts for each pair of individuals was first calculated with a custom python function (https://github.com/petrelharp/spaceness/blob/master/scripts/slimtools.py).
 We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $1e6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.
@@
 Phenotype values were set to vary by two standard deviations across the landscape in a rough approximation of the variation seen in height across Europe, which has recently been found to be confounded with population structure in large scale GWAS \citep{Berg2018,Sohail2018}.
-Conceptually our approach is similar to that taken in \citep{Mathieson2012}, though here we model fully continuous spatial variation and focus on genome-wide effects rather than low-frequency alleles specifically.
+Conceptually our approach is similar to that taken in \citep{Mathieson2012}, though here we model fully continuous spatial variation and compare GWAS output across a range of dispersal distances.
 
-In the first simulation phenotypes for all individuals were drawn from a normal distribution with mean 110 and standard deviation 10.
+In the first simulation phenotypes for all individuals were drawn from a normal distribution with mean 100 and standard deviation 10.
 \begin{equation}
-    P \sim \mathcal{N}(110,10)
+    P \sim \mathcal{N}(100,10)
 \end{equation}
@@
 \begin{equation}
-    P \sim \mathcal{N}(110+\frac{2x}{5},10)
+    P \sim \mathcal{N}(100+\frac{2x}{5},10)
 \end{equation}
@@
 \begin{cases}
-    \mathcal{N}(130,10),& \text{if } x< 20 \text{ and } y<20\\
-    \mathcal{N}(110,10),& \text{otherwise}\\
+    \mathcal{N}(120,10),& \text{if } x< 20 \text{ and } y<20\\
+    \mathcal{N}(100,10),& \text{otherwise}\\
 \end{cases}
@@
 \begin{cases}
-    \mathcal{N}(130,10),& \text{if } d < 3\\
-    \mathcal{N}(110,10),& \text{otherwise}\\
+    \mathcal{N}(120,10),& \text{if } d < 3\\
+    \mathcal{N}(100,10),& \text{otherwise}\\
 \end{cases}
@@
 inflation in Tajima's $D$ and observed heterozygosity under midpoint sampling).
-This general pattern is similar to that predicted by the original analytic model of \cite{Wright1943}, which predicts that populations with neighborhood sizes under 100 will differ substantially from random mating, while those above 10,000 will be nearly indistinguishable from panmixia.
+Under random sampling the general pattern is similar to expectations of the original analytic model of \cite{Wright1943}, which predicts that populations with neighborhood sizes under 100 will differ substantially from random mating, while those above 10,000 will be nearly indistinguishable from panmixia.
 
@@
 For example, in ornithology a sample of 10 individuals per species per locality is a common target when collecting for natural history museums.
-In classical studies of Drosophila variation the situation is considerably worse, in which a single orchard might be sampled with bated traps for instance.
+In classical studies of \textit{Drosophila} variation the situation is considerably worse, in which a single orchard might be sampled with bated traps for instance.
 
@@
 Here we found that one method of inference of historic $N_{e}$ based on genome-wide estimates of the site frequency spectrum, Stairwayplot \citep{Liu2015}, is relatively robust to variation in dispersal distance when sampling is random and neighborhood size is over 20.
-However, non-random sampling, and particularly midpoint sampling, causes the method to infer inflated estimates of past population sizes and a series of recent bottlenecks (Figure \ref{fig:demography}.
+However, non-random sampling, and particularly midpoint sampling, causes the method to infer inflated estimates of past population sizes and a series of recent bottlenecks (Figure \ref{fig:demography}).
 All sampling strategies lead to inflated ancient and deflated recent $N_{e}$ when neighborhood sizes were less than 20.
@@
 In summary, spatial covariation in population structure and the environment confound the interpretation of GWAS $p$-values, and correction using principal components is insufficient to fully separate these signals for polygenic traits under a variety of environmental and population parameter regimes.
-How more sophisticated mixed-model methods would perform under our simulations is an interesting question that we plan to pursue in a future study, but statistical methods can only take us so far in the absence of controlled experiments.
+How more sophisticated mixed-model methods would perform under our simulations is an interesting question that we plan to pursue in a future study, but statistical methods can only take us so far in the absence of controlled environments.
 One currently popular approach to estimating the degree of bias in GWAS caused by population structure is LD score regression \citep{Bulik-Sullivan2015}.
@@
   \makecell[l]{\textit{Ipomopsis}\\\textit{aggregata}} & flowering plant & 12.60 - 37.80 & Genetic & \citep{Campbell1992}\\
-  \makecell[l]{\textit{Linanthus}\\\textit{parryae}} & flowering plant & 14 - 27 & Genetic & \citep{Wright1943}\\ 
+  %\makecell[l]{\textit{Linanthus}\\\textit{parryae}} & flowering plant & 14 - 27 & Genetic & \citep{Wright1943}\\ %this one got debunked by a series of studies in the 90's that found that spatial differentiation in flower color is driven by environmental variability across years.
+See https://doi.org/10.1111/j.1558-5646.2007.00219.x
   \makecell[l]{\textit{Borrichia}\\\textit{frutescens}} & salt marsh plant & 20 - 30 & Genetic+Survey & \citep{Antlfinger1982} \\ 
@@
 
+One complication in the inference of any spatial demographic parameter is the balance between local and global process.
+Many species are structured locally by limited dispersal, but also contain deeply divergent lineages in different regions that reflect signals of ancient episodes of geographic isolation or strong barriers to dispersal.
+Gene flow upon secondary contact of two previously isolated lineages should create clinal patterns similar to isolation by distance, and it will be difficult to determine when inferred dispersal parameters are reflecting recent demographic process versus the historic patterns of geographic isolation.
+In addition, spatially varying selection will create allele frequency variation over space that may mimic isolation by distance.
+Indeed, a series of field studies described in \cite{Schemske2007} found that in Wright's original empirical example of isolation by distance, the flowering plant \textit{Linanthus parryae}, patterns of flower color differentiation over space primarily reflect temporal and spatial variation in selection rather than limited dispersal.
+Studies simulating selection and dispersal interacting in space (e.g.
+\cite{Ralph2010}) and testing for identifiability of inferred dispersal or selective parameters may offer new insight into the extent of our ability to accurately infer evolutionary processes in real systems.
+
+%Recent methods that aim to estimate geographic barriers to dispersal \citep{Ringbauer2018,Petkova2015} partially address this issue by attempting to infer the location of regions resistant to gene flow, but here again the balance between historical and current demographic processes represented in our inference is unclear.
+Limiting analyses to low-frequency alleles \citep{Novembre2009} or between long haplotype blocks \citep{Ringbauer2017} is one approach to limiting inference to recent demography (because these genomic features should arise primarily from patterns in the most recent branches of the genealogy), but as we showed spatial processes also cause shifts in branch lengths throughout the genealogy.
+Further studies on the ways in which the sample genealogy responds to variation in dispersal both over time and across space are a good direction for future research.
+
 Though our continuous space simulation allows incorporation of realistic demographic and spatial processes and is much faster than previous individual-based models, it is inevitably limited by the computational burden of tracking tens of thousands of individuals in every generation.
 In particular the calculations required for our mate selection and competition steps involve summarizing distances across all pairs of individuals and so scale very poorly \ak{how exactly does this scale?
-$O(N^2)$?} as the number of individuals within three-$\sigma$ increases.
+$O(N^2)$?} as the number of individuals within a three-$\sigma$ radius increases.
 In part the issue of runtime scaling is a function of the spatial process itself -- under very limited dispersal we observed that coalescence requires over 30N generations, so forward-time methods must be run for a very long time to create a complete genealogy as underlies all real genome sequence data.
 The reverse-time model of continuous space evolution described in \cite{Barton2010} and implemented in \cite{Kelleher2014} may allow exploration of parameter regimes with population and landscape sizes more directly comparable to empirical cases like humans.
-However, incorporating selection or other processes into such models will likely be difficult.
+However, incorporating selection or other processes into such models will be difficult.
 
@@
 
-Finally we believe that the difficulties in correcting for population structure in continuous populations using principal components analysis or similar decompositions is a difficult issue, well worth considering on its own-- what does it mean to correct for population structure in a world without discrete demes?
+Finally we believe that the difficulties in correcting for population structure in continuous populations using principal components analysis or similar decompositions is a difficult issue, well worth considering on its own -- what does it mean to correct for population structure in a world without discrete demes?
 What are the boundaries between proper corrections between ancestry and underpowering the search for local genetic variation?
-We posit that there is progress yet to be made in deriving either process driven descriptions of ancestry or more generalized unsupervised methods that can better account for shared relatedness, say in the context of GWAS, for populations that exist in space.
-\ak{this paragraph is a bit out there.
-rein me in}
+We posit that there is progress yet to be made in deriving either process driven descriptions of ancestry or more generalized unsupervised methods that can better account for shared relatedness, say in the context of GWAS, for populations that are structured over space.
+\ak{this paragraph is a bit out there. rein me in}
 
 \section{Data Availability}
-\ak{github repo}
+Scripts used for all analyses and figures are available at https://github.com/petrelharp/spaceness.
 
@@
 \begin{table}
-\label{table:sumstats}
 \small
@@
 \end{tabular}
+\label{table:sumstats}
 \end{table}

### 2019-04-25 ms/spaceness.tex
peter mostly edited until Results

@@
 \usepackage{changepage}
+\usepackage{afterpage}
 \setstretch{1.5}
@@
 \begin{abstract}
-Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations.
-As a result many methods of analyzing genetic data assume that samples are a random draw from a randomly mating population, but are applied to real populations that are structured over space by limited dispersal.
-Here we implement a forward-time simulation of evolution in continuous space and use it to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and genome-wide association studies.
-We find that continuous space skews the distribution of several common summary statistics, with the most bias occurring when Wright's neighborhood size is less than 100 and sampling is spatially clustered.
-In turn estimates of population size history based on the site frequency spectrum become biased in the face of limited dispersal.
-We show that in the context of genome-wide association studies the combination of spatial variation in the environment and limited dispersal creates spurious signals of genetic association with purely environmental phenotypes across much of the genome and that this bias is only partially corrected by regressing out principal components of ancestry.
-Last we discuss the relevance of our simulation results for inference from genetic variation in real organisms.
+    Real geography is continuous,
+    but standard models in population genetics are based on discrete, well-mixed populations. 
+    As a result many methods of analyzing genetic data assume that samples are a random draw from a well-mixed population, 
+    but are applied to real populations that have significant geographic structure.
+    Here we use simulations of populations living in continuous geography 
+    to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and genome-wide association studies. 
+    We find that several common summary statistics have distributions that differ substantially from that seen in well-mixed populations, 
+    especially when Wright's neighborhood size is less than 100 and sampling is spatially clustered. 
+    Standard estimates of population size history based on the site frequency spectrum become biased with lower dispersal. 
+    We also show that the combination of spatially autocorrelated environments and limited dispersal
+    can cause genome-wide association studies to identify numerous spurious signals of genetic association 
+    with purely environmentally determined phenotypes,
+    and that this bias is only partially corrected by regressing out principal components of ancestry. 
+    Last, we discuss the relevance of our simulation results for inference from genetic variation in real organisms. 
 \end{abstract}
@@
 
+\plr{note: address confounding of $\sigma$ and neighborhood size in disco}
+
+\plr{note: conclusions are that Ne(t) is not very biased}
+
+%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
-The inescapable reality that biological organisms live, move, and reproduce in continuous spatial landscapes has been all but ignored in population genetic models.
+
+The inescapable reality that biological organisms live, move, and reproduce in continuous geography is usually omitted from population genetic models. 
 Indeed, a near universal rule of reproduction is that individuals mate with other nearby individuals, leading to a positive correlation between genetic and geographic distances. 
-This pattern of "isolation by distance" \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, but classical mathematical models describing it are flawed approximations of the underlying process \citep{Felsenstein1975}.
-As a result many studies focus on describing geographic structure as a set of populations connected by migration \citep{Malecot1948,Wright1931}, and analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000} assuming these are randomly mating units.
+This pattern of ``isolation by distance'' \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, 
+but classical mathematical models describing it \citep{Malecot1948} are flawed approximations of the underlying process \citep{Felsenstein1975,barton2002neutral}.
+Mathematical difficulties have made further development of population models in continuous space difficult,
+so most models describe geographic structure as a set of populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical}, 
+and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000},
+effectively assuming these are randomly mating units.
 
-Though discrete models of between-population differentiation account for much of the genetic variation in species, the assumption that populations are randomly mating at some level has important implications for downstream inference of selection and demography.
-Methods based on the coalescent \citep{Kingman1982} assume that the sampled individuals are a random draw from a randomly mating population.
-If dispersal or mate selection is limited by space these assumptions are violated -- nearby individuals will be more closely related than an average random pair and drawing multiple samples from the same point on the landscape will represent a biased sample of the genetic variation present in the whole population.
-All downstream inferences drawn from the patterns of relationship inferred among sampled individuals will be subject to bias, but the extent depends on the degree of variation in ancestry across the landscape.
+The assumption that populations are ``well-mixed'' has important implications for downstream inference of selection and demography. 
+Methods based on the coalescent \citep{Kingman1982,wakeley2005coalescent} assume that the sampled individuals are a random draw from a well-mixed population
+that is much larger than the sample \citep{wakeley2003gene}.
+The key assumption is actually that the individuals of each generation are \emph{exchangeable},
+so that there is no correlation between the fate or fecundity of a parent and that of their offspring \citep{Huillet2011}.
+If dispersal or mate selection is limited by geographic proximity, this assumption can be violated in many ways.
+For instance, if mean viability or fecundity is spatially autocorrelated,
+then limited geographic dispersal will lead to parent--offspring correlations.
+Futhermore, nearby individuals will be more closely related than an average random pair,
+so drawing multiple samples from the same area of on the landscape will represent a biased sample of the genetic variation present in the whole population. 
+It has long been appreciated that this model misspecification subjects downstream inferences to bias, 
+but the extent and nature of these effects remain largely uninvestigated.
 
-Our understanding of how many common summary statistics respond to variation in dispersal distance is limited.
-Tajima's $D$, for example, is often interpreted as reflecting evidence of selection or past population size changes (CITE \cjb{CJ: what's a good citation here?}).
-However this statistic is essentially a summary of the site frequency spectrum, which itself reflects variation in branch lengths of the underlying genealogy of sampled individuals.
-Because the spatial extent of dispersal and mate selection determines the rate of drift in a continuously distributed population \citep{Maruyama1972}, these population genetic summaries should also reflect variation in the scale of spatial interactions both over time and across populations.
-Similarly, the distribution of tract lengths of identity by state among individuals is often interpreted either through the lens of historic demography \citep{Harris2013} or selection \citep{Garud2015}.
-While that is so this distribution also contains abundant information about dispersal and mate selection \citep{Ringbauer2017,Baharian2016}.
-We are particularly keen on examining how such summaries will be affected by models that incorporate continuous space.
+For instance, nonzero values of Tajima's $D$ are often interpreted as reflecting evidence of selection or past population size changes \citep{tajima1989statistical}.
+This statistic is a summary of the site frequency spectrum, which itself reflects variation in branch lengths and tree structure of the underlying genealogies of sampled individuals.
+Geographically limited mate choice distorts the distribution of these genealogies \citep{Maruyama1972},
+which can affect the value of Tajima's $D$.
+% Because the spatial extent of dispersal and mate selection determines the rate of drift in a continuously distributed population 
+% these population genetic summaries should also reflect variation in the scale of spatial interactions both over time and across populations. 
+Similarly, the distribution of tract lengths of identity by state among individuals contains information about not only
+historical demography \citep{Harris2013,ralph2013geography} and selection \citep{Garud2015}, but also dispersal and mate choice \citep{Ringbauer2017,Baharian2016}. 
+We are particularly keen to examine how such summaries will be affected by models that incorporate continuous space,
+both to evaluate the assumptions underlying existing methods, but also to identify where the most promising signals of geography lie.
 
-A related issue is the spatial distribution of sampling effort.
-Though individuals in a random-mating population are exchangeable, those in a continuously structured population are not.
-If sampling is clustered at a series of points on the landscape, as is common in empirical studies because of the practical difficulties of sample collection, samples drawn from a spatially structured population may partially reflect spatial biases in the sampling process rather than in the underlying population.
-In addition, range edges have been observed to create complex patterns of heterozygosity in stepping-stone simulations \citep{Neel2013,Shirk2014}, but the effects of this process on many downstream inference procedures is unknown.
+A related issue is the spatial distribution of sampling effort --
+since nearby individuals are more likely to be closely related than distant ones,
+observed patterns of relatedness are expected to depend on the geographic sampling scheme,
+perhaps strongly.
+In addition, range edges have been observed to create complex patterns of heterozygosity in stepping-stone simulations \citep{Neel2013,Shirk2014}, 
+but the effects of this process on many downstream inference procedures is unknown.
 \cjb{must be more out there on sampling -- any recs for papers to read+cite?}
 
-The issue of fine-scale structure for putatively "within-population" analyses may have important implications for genome-wide association studies as well.
-This is because the genome-wide allele frequency variation caused by isolation by distance is similar to that expected when many sites in the genome are associated with a given phenotype \citep{Bulik-Sullivan2015}.
-When population structure and the environment covary across space genome-wide association studies (GWAS) may incorrectly infer phenotypic association at many SNPs that have differentiated by drift \citep{Mathieson2012}.
-This may be a particularly important effect when using GWAS to study selection on polygenic traits.
-Indeed two recent studies found that previous evidence of polygenic selection on human height in Europe was confounded by subtle population structure \citep{Sohail2018,Berg2018}, suggesting that existing methods to correct for population structure in GWAS are insufficient.
+The issue of fine-scale geographic structure may have particularly important implications for genome-wide association studies (GWAS). 
+This is because many phenotypes of interest have strong geographic differences due to the (nongenetic) influence of environmental or socioeconomic factors,
+which can therefore show spurious correlations with spatially patterned allele frequencies \citep{Bulik-Sullivan2015,Mathieson2012}. 
+This may be particularly important for the study of selection on polygenic traits,
+whose heritable genetic components are determined by many loci of weak effect,
+those hardest to disentangle from spurious correlations. 
+Indeed, two recent studies found that previous evidence of polygenic selection on human height in Europe was confounded by subtle population structure \citep{Sohail2018,Berg2018}, suggesting that existing methods to correct for population structure in GWAS are insufficient.
 However we have little quantitative idea of the population and environmental parameters that can be expected to lead to biases in GWAS.
@@
 
-Here we describe an implementation of an individual-based model in continuous space that incorporates overlapping generations, a Gaussian dispersal kernel, and density-dependent fitness.
-The model scales to chromosome-scale alignments across tens of thousands of individuals and outputs the full genealogy and recombination history of all final-generation individuals.
-We use this simulation to test how sampling strategy interacts with isolation by distance to cause systematic variation in population genetic summary statistics typically analyzed assuming discrete population models.
+To study this, we have implemented an individual-based model in continuous geography 
+that incorporates overlapping generations, local dispersal of offspring, and density-dependent survival. 
+We simluate chromosome-scale genomic data in tens of thousands of individuals, and output the full genealogy and recombination history of all final-generation individuals. 
+We use these simulations to test how sampling strategy interacts with geographic population structure to cause systematic variation in population genetic summary statistics 
+typically analyzed assuming discrete population models. 
 We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum. 
-Lastly we examine the impacts of evolution in continuous space on genome-wide association studies (GWAS) and identify regions of parameter space under which the results from GWAS may be misleading.
+Lastly, we examine the impacts of evolution in continuous space on genome-wide association studies (GWAS) and identify regions of parameter space under which the results from GWAS may be misleading.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Materials and Methods}
@@
 
+%%%%%%%%%%%%%
 \subsection{Modeling Evolution in Continuous Space}
+
+\plr{PETER TO REVISIT THIS SECTION}
+
 The best-studied approaches to population genetics in continuous space were developed by \cite{Wright1942,Wright1943} and \cite{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals.
-A key finding of Wright's model is that many important aspects of continuous populations can be described in terms of "neighborhood size" -- the number of potential mates for an individual in a given generation, defined as $4\pi\sigma^2d$, where $\sigma$ is the average dispersal distance and $d$ is population density.
+A key finding of Wright's model is that many important aspects of continuous populations can be described in terms of ``neighborhood size'' -- the number of potential mates for an individual in a given generation, defined as $4\pi\sigma^2d$, where $\sigma$ is the average dispersal distance and $d$ is population density.
 \cite{Maruyama1972} found that the rate of decline in genetic diversity in a 2-dimensional continuous population approaches the random mating expectation when $d\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.
@@
 One method for modeling continuous populations is then to assume the existence of a grid of discrete randomly-mating populations connected by migration, which prevents clustering by forcing all regions to be occupied in every generation.
-Among many other important results drawn from this class of "lattice" or "stepping stone" models, \cite{Rousset1997} showed that the slope of the a linear regression of genetic differentiation ($F_{st}$) against the logarithm of spatial distance is an estimate of neighborhood size.
+Among many other important results drawn from this class of ``lattice'' or ``stepping stone'' models, \cite{Rousset1997} showed that the slope of the a linear regression of genetic differentiation ($F_{ST}$) against the logarithm of spatial distance is an estimate of neighborhood size.
 Though good approximations of continuous structure given high dispersal, these models are not truly continuous, force a uniform realized population density across landscapes, and limit investigation of spatial structure below the level of the deme.
@@
 By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.
+
+%%%%%%%%%%%%
 \subsection{A Forward-Time Model of Evolution in Continuous Space}
 
-We implemented our model using the non-Wright-Fisher module in the program SLiM v3.0 \citep{Haller2019}.
-Each time step consists of three stages: reproduction, dispersal, and competition.
-To reduce the parameter space we use the same parameter, denoted $\sigma$, to modulate the spatial scale of interactions at all three stages by adjusting the standard deviation of the corresponding Gaussian functions.
-As in previous work \citep{Wright1943,Ringbauer2017}, $\sigma$ as applied in our dispersal step is approximately equal to the mean parent-offspring distance.
+We simulated populations using the non-Wright-Fisher module in the program SLiM v3.0 \citep{Haller2019}. 
+Each time step consists of three stages: reproduction, dispersal, and mortality. 
+To reduce the parameter space we use the same parameter, denoted $\sigma$, 
+to modulate the spatial scale of interactions at all three stages by adjusting the standard deviation of the corresponding Gaussian functions. 
+As in previous work \citep{Wright1943,Ringbauer2017}, $\sigma$ as applied in our dispersal step is equal to the mean parent-offspring distance.  
+A key parameter we report below is
+\citeauthor{Wright1943} citeyearpar{Wright1943} ``neighborhood size'', 
+defined to be $N_W = 4\pi\sigma^2 \rho$ where $\rho$ is the population density.
+This the approximate number of individuals available for mating in our simulation.
 
-At the beginning of the simulation individuals are distributed randomly on a continuous landscape.
-Mates are selected proportional to distances among individuals weighted by a Gaussian function with mean $1/(2\pi\sigma^2)$ and standard deviation $\sigma$.
-Wright's \cite{Wright1943} "Neighborhood Size", defined as $4\pi\sigma^2 K$ where $K$ is the population density, is then the approximate number of individuals available for mating in our simulation.
-The number of offspring is drawn from a Poisson distribution with mean $1/L$, where $L$ is the average lifespan.
-If new offspring are produced, we simulate dispersal by taking two draws from a normal distribution with mean 0 and standard deviation $\sigma$ and adding these to the x and y coordinates of the first parent.
+At the beginning of the simulation individuals are distributed uniformly at random on a continuous, square landscape. 
+Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$ 
+who disperse a random, Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
+in both the $x$ and $y$ coordinates, reflected to stay within the species range.
+Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
+with probability of choosing a neighbor at distance $x$ proportional to $\exp(-x^2 / 2 \sigma^2)$.
 
-Density-dependent competition is simulated by adjusting the probability of survival of each individual proportional to the local density of individuals up to a distance $3\sigma$ away, scaled according to the same Gaussian function used for mate selection.
-Given a per-unit carrying capacity $K$ and average fecundity $F$, the probability of survival $d$ for individual $i$ after adjustment for density is:
+To maintain a stable population, mortality increases with local population density.
+To do this we say that individuals at distance $d$ have a competitive interaction with strength $g(d)$,
+where $g$ is the Gaussian density with mean zero and standard deviation $\sigma$.
+Then, the sum of all competitive interactions with individual $i$ is $n_i = \sum_j g(d_{ij})$,
+where $d_{ij}$ is the distance between individuals $i$ and $j$ and the sum is over all neighbors within distance $3 \sigma$.
+Since $g$ is a probability density, $n_i$ is an estimate of the number of nearby individuals per unit area.
+Then, given a per-unit carrying capacity $K$, the probability of survival until the next time step for individual $i$ is
+\begin{equation}
+    p_i \min(0.95, \frac{1}{1 + n_i / (K (1+L)} .
+\end{equation}
+We chose this functional form so that the equilibrium population density per unit area is around $K$,
+and the mean lifetime is around $L$.
 
+An important step in creating any spatial model is dealing with range edges. 
+Because local population density is used to model competition, edge or corner populations can be assigned artificially high fitness values 
+because they lack neighbors within their interaction radius but outside the bounds of the simulation.
+We approximate a decline in habitat suitability near edges by decreasing the probability of survival proportional to the square root of distance to edges in units of $\sigma$. 
+The final probability of survival for individual $i$ is then
 \begin{equation}
-    d_{i}=min(0.95,\frac{1}{1+\rho*\sum_{i}{G(x)}})
+    s_{i} = p_{i} 
+        \min(1,\sqrt{x_{i}/\sigma})
+        \min(1,\sqrt{y_{i}/\sigma})\\
+        \min(1,\sqrt{(W - x_{i})/\sigma})
+        \min(1,\sqrt{(W - y_{i})/\sigma)}
 \end{equation}
-where 
-\begin{equation}
-    \rho = F/((1+F)*K)
-\end{equation} 
-and $G(x)$ is the Gaussian function used to weight distances to other individuals.
+where $x_i$ and $y_i$ are the spatial coordinates of individual $i$, and $W$ is the width (and height) of the square habitat.
+This buffer roughly counteracts the increase in fitness individuals close to the edge would otherwise have.
 
-A major challenge in all spatial models is dealing with range edges.
-When local population density is used to model competition, edge or corner populations can be assigned artificially high fitness values because they lack neighbors within their interaction radius but outside the bounds of the simulation.
-We approximate a decline in habitat suitability near edges by decreasing the probability of survival proportional to the square root of distance to edges in units of $\sigma$.
-The final probability of survival for individual \textit{i} is then
+To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, and density-dependent fitness, 
+we also implemented simulations identical to those above except that mates are selected uniformly random from the population,
+and offspring disperse to a uniform random location on the landscape. 
+We refer to this model as the ``random mating'' model, in contrast to the first, ``spatial'' model.
 
-\begin{equation}
-    s_{i}=d_{i} min(1,\sqrt{x_{i}/\sigma})
-    min(1,\sqrt{y_{i}/\sigma})\\
-    min(1,\sqrt{(x_{max}-x_{i})/\sigma})
-    min(1,\sqrt{(y_{max}-y_{i})/\sigma)}
-\end{equation}
+We stored the full genealogy and recombination history of final-generation individuals as tree sequences \citep{Kelleher2018},
+as implemented in SLiM \citep{haller2019treesequence}. 
+Scripts for figures and analyses are available at \url{https://github.com/petrelharp/spaceness}.
 
-where $x$ and $y$ are spatial coordinates. In practice this creates a one $\sigma$ buffer around the edge of the map in which fitness is depressed.
-
-To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, and density-dependent fitness, we  also implemented a second version of the SLiM simulation in which mates are selected at random and offspring disperse to a random location on the landscape.
-We refer to these two models as the "spatial" and "random mating" models.
-For all simulations the full genealogy and recombination history of final-generation individuals were stored as tree sequences \citep{Kelleher2018}.
-Scripts for figures and analyses are available at https://github.com/petrelharp/spaceness.
-
-We ran 400 simulations for the spatial and random-mating SLiM models on a square landscape of size 50 with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), average lifetime $L=4$, genome size $1\times10^{8}$, recombination rate $1\times10^{-9}$, and drawing $/sigma$ values from a uniform distribution bounded by 0.2 and 4.
-To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM and later applied mutations to the coalesced tree sequence with msprime's 'mutate( )' function.
-Because msprime assumes that timesteps are in units of generations, we divided the per-generation rate of $1\times10^{-8} mut/site/gen$ by the generation time estimated for each value of $\sigma$ (see 'Genealogical Parameters' below) to convert the rate to units of mutations/site/timestep.
-We then verified that this procedure produced the correct number of mutations by comparing a subset of simulations with SLiM-generated mutations (which are applied only at mating events) with mutations added by msprime.
-Simulations were run for 1.6 million timesteps (approximately 30N generations), or until all extant individuals shared a common ancestor within the simulation (i.e.
-the tree sequence had coalesced).
+We ran 400 simulations for the spatial and random-mating models on a square landscape of width $W=50$ 
+with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), 
+average lifetime $L=4$, 
+genome size $10^{8}$, 
+recombination rate $10^{-9}$, 
+and drawing $\sigma$ values from a uniform distribution between 0.2 and 4. 
+To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM 
+and later applied mutations to the tree sequence with msprime's \texttt{mutate} function \citep{Kelleher2016}. 
+Because msprime applies mutations proportionally to elapsed time,
+we divided the mutation rate of $10^{-8}$ mutations per site per generation
+by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time.
+(We verified that this procedure produced the correct number of mutations by comparing to a subset of simulations with SLiM-generated mutations, which are applied only at meiosis.)
+Simulations were run for 1.6 million timesteps (approximately $30N$ generations), or until all extant individuals shared a common ancestor within the simulation (i.e., the tree sequence had coalesced).
 \ak{maybe worth including a table with some basic runtime results in the supplement?}
 
-\subsection{Genealogical Parameters}
-Because basic elements of our simulations like census population size and generation time are not directly parameterized in our model (instead they emerge from the interaction of parameters including average population density, average lifespan, and the spatial scale of interaction), we first monitored a range of genealogical parameters in all simulations.
-We recorded the census population size and estimated the variance in number of offspring and generation time for all simulations.
-To estimate generation times we stored the age of all parents for 200 timesteps after a 100 generation burn-in and took the mean.
-To estimate variance in offspring we tracked the number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period, subset the resulting table to include only the last timestep recorded for each individual, and calculated the variance in number of offspring across all individuals in timesteps 50-100.
+%%%%%%%%%%%%%
+\subsection{Demographic Parameters}
 
+Our demographic model includes paramters for population density ($K$),
+mean life span ($L$), and dispersal distance ($\sigma$).
+However, nonlinearity of local demographic stochasticity causes
+actual realized averages of these demographic quantitites to deviate from the specified values
+in a way that depends on the neighborhood size.
+Therefore, to properly compare to theoretical expectations,
+we empirically calculated these demographic quantities in simulations. 
+We recorded the census population size in all simulations. 
+To estimate generation times, we stored ages of the parents of every new individual born across 200 timesteps, 
+after a 100 generation burn-in, and took the mean. 
+To estimate variance in offspring number, 
+we tracked the number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period, subset the resulting table to include only the last timestep recorded for each individual, and calculated the variance in number of offspring across all individuals in timesteps 50-100.
+All calculations were performed with information recorded in the tree sequence,
+using pyslim \citetext{\url{https://github.com/tskit-dev/pyslim}}.
+
+%%%%%%%%%%%%%
 \subsection{Sampling}
-Our model records the genealogy and sequence variation of the complete population, but in real data genotypes are only observed from a relatively small number of sampled individuals.
+Our model records the genealogy and sequence variation of the complete population, but in real data, genotypes are only observed from a relatively small number of sampled individuals.
 We modeled three sampling strategies similar to common data collection methods in empirical genetic studies (Figure \ref{fig:samplemap}).
-"Random" sampling selects individuals at random from across the full landscape, "point" sampling selects individuals proportional to their distance from four equally spaced points on the landscape, and "midpoint" sampling selects individuals in proportion to their distance from the middle of the landscape.
+``Random'' sampling selects individuals at random from across the full landscape, ``point'' sampling selects individuals proportional to their distance from four equally spaced points on the landscape, and ``midpoint'' sampling selects individuals in proportion to their distance from the middle of the landscape.
 Downstream analyses were repeated across all sampling strategies.
@@
 \includegraphics{figures/sampling_maps.pdf}
-\caption{Example sampling maps for 60 individuals on a 50x50 landscape under midpoint, point, and random sampling strategies.}
+\caption{Example sampling maps for 60 individuals on a 50$\times$50 landscape for midpoint, point, and random sampling strategies, respectively.}
 \label{fig:samplemap}
@@
 
+%%%%%%%%%%%%%
 \subsection{Summary Statistics}
 We calculated the site frequency spectrum and a set of 18 summary statistics (Table \ref{table:sumstats}) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}.
-Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{is}$), and Tajima's D, as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distances \citep{Rousset1997}, which we summarized as the correlation coefficient between $log10(spatial distance)$ and the proportion of identical base pairs among individuals.
+Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{IS}$), and Tajima's $D$, 
+as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distances \citep{Rousset1997}, 
+which we summarized as the correlation coefficient between the logarithm of the spatial distance and the proportion of identical base pairs across pairs of individuals.
 
 Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among samples.
-The full distribution of lengths of IBS tracts for each pair of individuals was first calculated with a custom python function (https://github.com/petrelharp/spaceness/blob/master/scripts/slimtools.py).
+The full distribution of lengths of IBS tracts for each pair of chromosomes was first calculated with a custom python function.
+%\url{https://github.com/petrelharp/spaceness/blob/master/scripts/slimtools.py}. 
 We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $1e6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.
@@
 most negative) when dispersal is low.
-The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks so should reflect recent demographic events \citep{Ringbauer2017}.
+The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks so should reflect recent demographic events \citep{chapman2002effect}.
 
@@
 
+%%%%%%%%%%%
 \subsection{Demographic Modeling}
+
 We fit single-population demographic models to the site frequency spectra of 20 individuals from each spatial SLiM simulation with the program Stairwayplot \citep{Liu2015}.
 This analysis was replicated across random, point, and midpoint sampling strategies.
-Site frequency spectra used for input data were calculated in scikit-allel \citep{Miles2017}, and 100 bootstrap replicates were generated for each simulation by resampling over sites.
-Models were fit across all bootstrap replicates using default settings in Stairwayplot and the median estimate of $Ne$ per generation was used to represent the output of each simulation.
+Site frequency spectra used for input data were calculated in scikit-allel \citep{Miles2017},
+and 100 bootstrap replicates were generated for each simulation by resampling over sites. 
+\plr{what were bootstrap replicates used for? if anything, need to say more precisely what you mean here}
+Models were fit across all bootstrap replicates using default settings in Stairwayplot and the median estimate of $N_e$ per generation was used to represent the output of each simulation.
 
-In preliminary runs we found that inferred population histories were relatively variable even when simulating under a coalescent model, suggesting that some of the differences in demographic estimates for spatial models are caused by the behavior of the optimization algorithm rather than bias in the SFS caused by spatial mate choice and dispersal.
+In preliminary runs we found that inferred population histories were highly variable even when simulating under a coalescent model, suggesting that some of the differences in demographic estimates for spatial models are caused by the behavior of the optimization algorithm rather than bias in the SFS caused by spatial mate choice and dispersal.
 To separate these effects we ran 100 coalescent simulations with constant population size $6.1\times 10^{-3}$ (the mean $N_{e}$ of random-mating SLiM models estimated from $\Theta_{\pi}$) and fit stairwayplot models using the same script as for our spatial models.
@@
 
+%%%%%%%%%%%
 \subsection{Association Studies}
-To assess the degree to which spatial structure confounds GWAS we simulated four types of nongenetic phenotype variation for 1000 randomly sampled individuals in each spatial SLiM simulation and conducted a linear regression GWAS with PC covariates in PLINK \citep{PURCELL2007}.
+
+To assess the degree to which spatial structure confounds GWAS we simulated four types of nongenetic phenotype variation for 1000 randomly sampled individuals in each spatial SLiM simulation and conducted a linear regression GWAS with principal components as covariates in PLINK \citep{PURCELL2007}.
 SNPs with a minor allele frequency less than 0.5\% were excluded from this analysis.
@@
 
-In the first simulation phenotypes for all individuals were drawn from a normal distribution with mean 100 and standard deviation 10.
-\begin{equation}
-    P \sim \mathcal{N}(100,10)
-\end{equation}
-Next we simulated clinal environmental influences on phenotype by drawing the phenotypes from independent normal distributions in which the mean was scaled by an individual's x position such that it varied by 2 standard deviations across the map.
-If $x$ is an individual's x coordinate on a 50x50 landscape, the phenotype is then:
-\begin{equation}
-    P \sim \mathcal{N}(100+\frac{2x}{5},10)
-\end{equation}
-Third, we approximate concentrated environmental effects by drawing phenotypes for individuals with x and y coordinates below 20 from a normal distribution with mean 2 standard deviations above the rest of the map.
-\begin{equation}
-    P\sim
-\begin{cases}
-    \mathcal{N}(120,10),& \text{if } x< 20 \text{ and } y<20\\
-    \mathcal{N}(100,10),& \text{otherwise}\\
-\end{cases}
-\end{equation}
+In all simulations, the phenotype of each individual
+is determined by adding independent Gaussian noise with mean zero and standard deviation 10
+to a mean that may depend on spatial position.
+We adjust the geographic pattern of mean phenotype to create spatially autocorrelated environmental influences on phenotype.
+In the first simulation of \emph{nonspatial} phenotypes, the mean did not change,
+so that all individuals' phenotypes were drawn independently from a Gaussian distribution with mean 110 and standard deviation 10.
+Next, to simulate \emph{clinal} environmental influences on phenotype, 
+we increased the mean phenotype form 90 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
+Concretely, an individual at position $(x,y)$ in a $50 \times 50$ landscape has mean phenotype $110 + 2x / 5$.
+Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype 
+for individuals with both $x$ and $y$ coordinates below 20 to 130 (two standard deviations above the rest of the map). 
+Finally, in ``\emph{patchy}'' simulations we selected 10 random points on the map
+and set the mean phenotype of all individuals within three map units of each of these points
+to 130.
 
-Last, we simulated a "patchy" environmental influence on phenotypes by selecting 10 random points on the map and drawing phenotypes for all individuals within two map units of any selected point from a normal distribution with mean 2 standard deviations above the rest of the map.
-
-\begin{equation}
-    P\sim
-\begin{cases}
-    \mathcal{N}(120,10),& \text{if } d < 3\\
-    \mathcal{N}(100,10),& \text{otherwise}\\
-\end{cases}
-\end{equation}
-where $d$ is the distance to the nearest patch.
-
-Principal components analysis (PCA) was conducted in scikit allel on the matrix of derived allele counts by individual for each simulation.
+We performed principal components analysis (PCA) using scikit-allel \citep{Miles2017} on the matrix of derived allele counts by individual for each simulation. 
 SNPs were first filtered to remove strongly linked sites by calculating LD between all pairs of SNPs in a 200-SNP moving window and dropping one of each pair of sites with an $R^2$ over 0.1.
@@
 
-We ran linear-model GWAS both with and without the first 10 principal components as covariates in PLINK and summarized results across simulations by counting the number of significant SNPs with an expected false positive rate of less than 5\% after adjusting p values with the R function p.adjust(...,method="fdr").
+We ran linear-model GWAS both with and without the first 10 principal components as covariates in PLINK and summarized results across simulations by counting the number of SNPs 
+with $p$-value below $0.05$ after adjusting for an expected false positive rate of less than 5\% \citep{benjamini2001control}. 
+% using the R function p.adjust(...,method=``fdr''). 
 We also examined $p$ values for systemic inflation by estimating the expected values from a uniform distribution (because no SNPs were used when generating phenotypes), plotting observed against expected values for all simulations, and summarizing across simulations by finding the mean $\sigma$ value in each region of quantile-quantile space.
-Results from all analyses were summarized and plotted with the 'ggplot2' \citep{Wickham2016} and "cowplot" \citep{Wilke2019} packages in R \citep{Rcore2018}.
+Results from all analyses were summarized and plotted with the 'ggplot2' \citep{Wickham2016} and ``cowplot'' \citep{Wilke2019} packages in R \citep{Rcore2018}.
 
+%%%%%%%%%%%%%%%%%%
 \section{Results}
@@
 
-Patterns from observed heterozygosity and its derivative $F_{is}$ also depend heavily on neighborhood size under spatial models as well as the sampling scheme.
-$F_{is}$ is inflated above the expectation across most of the parameter space examined and across all sampling strategies.
+Patterns from observed heterozygosity and its derivative $F_{IS}$ also depend heavily on neighborhood size under spatial models as well as the sampling scheme.
+$F_{IS}$ is inflated above the expectation across most of the parameter space examined and across all sampling strategies.
 This effect is caused by a deficit of heterozygous individuals in low-dispersal simulations -- a continuous-space version of the Wahlund effect \citep{Wahlund1928}.
-Indeed for random sampling under the spatial model $F_{is}$ does not approach the random mating equivalent until neighborhood sizes of nearly 1000.
+Indeed for random sampling under the spatial model $F_{IS}$ does not approach the random mating equivalent until neighborhood sizes of nearly 1000.
 The patterns for raw observed heterozygosity are more complex.
@@
 \subsubsection{Correlations of summary statistics with geographic distance}
-Correlating population genetic summaries such as $F_{st}$ against geographic distance has shown great utility in empirical population genetics \citep{Rousset1997}.
+Correlating population genetic summaries such as $F_{ST}$ against geographic distance has shown great utility in empirical population genetics \citep{Rousset1997}.
 As we know the exact locations of individuals that are sampled in our simulated populations we have examined the relationship of geographic distance among samples with a number of our summary statistics (Figure \ref{fig:sumstats} and Figure \ref{fig:allsumstats}).
@@
 This is because identical-by-state tract lengths reflect the impacts of two processes acting along the branches of the underlying genealogy -- both mutation and recombination -- rather than just mutation as is the case when looking at the site frequency spectrum or related summaries.
-This means that the pairwise distribution of haplotype block lengths carries with it important information about genealogical variation in the population, and correlation coefficients between moments of the this distribution and geographic location contain signal similar to the correlations between $F_{st}$ or $D_{xy}$ and space \citep{Rousset1997}.
+This means that the pairwise distribution of haplotype block lengths carries with it important information about genealogical variation in the population, and correlation coefficients between moments of the this distribution and geographic location contain signal similar to the correlations between $F_{ST}$ or $D_{xy}$ and space \citep{Rousset1997}.
 Indeed this basic logic underlies two recent studies explicitly estimating dispersal from the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}.
@@
 However this strategy is often impractical for empirical studies.
-In reality the difficulty of traveling through all parts of a species range and the inefficiency of collecting single individuals at each sampling site means that most studies follow something closer to the "point" sampling strategy we simulated, in which multiple individuals are sampled from nearby points on the landscape.
+In reality the difficulty of traveling through all parts of a species range and the inefficiency of collecting single individuals at each sampling site means that most studies follow something closer to the ``point'' sampling strategy we simulated, in which multiple individuals are sampled from nearby points on the landscape.
 For example, in ornithology a sample of 10 individuals per species per locality is a common target when collecting for natural history museums.
@@
 For some data summaries, such as Tajima's $D$, Watterson's $\Theta$, or the correlation coefficient between spatial distance and the count of long haplotype blocks, this can result in large differences in estimates between random and point sampling (Figure \ref{fig:sumstats}).
-Inferring underlying demographic parameters from these summary statistics -- for example, estimating dispersal distance as the slope of a regression of $F_{st}$ against the logarithm of spatial distance \citep{Rousset1997} -- may then be subject to bias if sampling is not random across the landscape.
+Inferring underlying demographic parameters from these summary statistics -- for example, estimating dispersal distance as the slope of a regression of $F_{ST}$ against the logarithm of spatial distance \citep{Rousset1997} -- may then be subject to bias if sampling is not random across the landscape.
 
-However, the largest sampling effects we observed occurred in our "midpoint" sampling strategy.
+However, the largest sampling effects we observed occurred in our ``midpoint'' sampling strategy.
 This model is meant to reflect a bias in sampling effort towards the middle of a species' range.
@@
 
-Incorporating PC positions as covariates in the analysis \citep{Price2006} is designed to address this difficulty by regressing out a baseline level of "average" differentiation.
+Incorporating PC positions as covariates in the analysis \citep{Price2006} is designed to address this difficulty by regressing out a baseline level of ``average'' differentiation.
 However while this approach is quite useful, it does not truly separate the confounding signals of environment and spatially varying genotypes.
-In essence with a PC-corrected GWAS we are asking "what regions of the genome are more associated with this phenotype than the average genome-wide association observed across populations?" In our simulations we observed that this procedure can fail under a variety of circumstances.
+In essence with a PC-corrected GWAS we are asking ``what regions of the genome are more associated with this phenotype than the average genome-wide association observed across populations?'' In our simulations we observed that this procedure can fail under a variety of circumstances.
 If dispersal is limited and environmental variation is clustered in space (i.e.
@@
 If significant associations with sample locality are observed after correcting for population structure through PC regression or a kinship matrix, the structure corrections are insufficient.
-This is essentially the approach taken in our "clinal" model (though we also include normally distributed variation in our phenotypes).
+This is essentially the approach taken in our ``clinal'' model (though we also include normally distributed variation in our phenotypes).
 Of course it is possible that genotypes indirectly affect individual locations by adjusting organismal fitness and thus habitat selection across spatially varying environments, but we believe that this hypothesis should be tested against a null of stratification bias inflation rather than accepted as true based on GWAS results.
@@
 A mixed approach may be possible by combining forward- and reverse-time models, as was recently done for a continuous-space Wright-Fisher model in \cite{Lotterhos2019} and for a simulation with linked selection in \cite{Buffalo2019}.
-This would allow us to generate short runs of complex, realistic simulations in forward time in SLiM \citep{Haller2019} and then "finish" the simulations as a coalescent simulation in msprime \citep{Kelleher2016}.
-However a significant difficulty in this "recapitation" approach is scaling the variance in reproductive output and generation time across forward- and reverse-time methods.
+This would allow us to generate short runs of complex, realistic simulations in forward time in SLiM \citep{Haller2019} and then ``finish'' the simulations as a coalescent simulation in msprime \citep{Kelleher2016}.
+However a significant difficulty in this ``recapitation'' approach is scaling the variance in reproductive output and generation time across forward- and reverse-time methods.
 Further development of our understanding of how to merge forward- and reverse-time models is a promising avenue for future research that will be necessary for scaling continuous-space simulations to millions or billions of individuals.
@@
 Observed Heterzygosity & Proportion of heterozygous individuals in the sample \\
-$F_{is}$ & Wright's inbreeding coefficient $1-H_{e}/H_{o}$ \\
+$F_{IS}$ & Wright's inbreeding coefficient $1-H_{e}/H_{o}$ \\
 $var(D_{xy})$ & Variance in the distribution of pairwise genetic distances \\
@@
 \centering
-\caption{\bf Anova and Levene's test $p$ values for differences by sampling strategy. Bolded values are rejected at \alpha=0.05}
+\caption{\bf Anova and Levene's test $p$ values for differences by sampling strategy. Bolded values are rejected at $\alpha=0.05$}
 \begin{tableminipage}{\textwidth}

### 2019-04-27 ms/spaceness.tex
edited until discussion

@@
 Because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within continuous populations can also deflate $p$ values in GWAS of traits influenced by the environment \citep{Mathieson2012}.
-Here we simulate both the underlying genealogy and phenotypes of a continuously distributed population and seek to identify regions of parameter space -- i.e.
-the strength of isolation by distance and the spatial distribution of environmental effects on phenotypes -- in which common methods employed for GWAS are subject to confounding of environmental and genetic effects on phenotypes.
+Here we simulate both the underlying genealogy and phenotypes of a continuously distributed population and seek to identify regions of parameter space -- i.e., the strength of isolation by distance and the spatial distribution of environmental effects on phenotypes -- in which common methods employed for GWAS are subject to confounding of environmental and genetic effects on phenotypes.
 
@@
 \begin{equation}
-    p_i \min(0.95, \frac{1}{1 + n_i / (K (1+L)} .
+    p_i 
+    =
+    \min\left( 0.95, \frac{1}{1 + n_i / (K (1+L))} \right) .
 \end{equation}
@@
 We then estimated correlation coefficients between spatial distance and each moment of the pairwise IBS tract distribution.
-Because more closely related individuals on average share longer haplotype blocks we expect that spatial distance will be negatively correlated with mean haplotype block length, and that this correlation will be strongest (i.e.
-most negative) when dispersal is low.
+Because more closely related individuals on average share longer haplotype blocks we expect that spatial distance will be negatively correlated with mean haplotype block length, and that this correlation will be strongest (i.e., most negative) when dispersal is low.
 The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks so should reflect recent demographic events \citep{chapman2002effect}.
@@
 
-\subsection{Genealogical Parameters}
-Having implemented our continuous space forward time population genetic simulation, the first thing we were interested in examining was the effect of spatial parameters on population parameters such as the generation time, the census population size, and the variance in offspring number.
-In contrast to intuition based on standard population genetic models, in the context of our non-Wright-Fisher model basic quantities like the generation time depend on life history parameters such as the dispersal distance and carrying capacity of the landscape.
-We first examined how the population size, generation time, and  variance in number of offspring vary with neighborhood size, which itself is a function of $\sigma$ (Figure \ref{fig:genparams}).
-There are a few things to note.
-First all three quantities are non-linear with respect to neighborhood size.
+%%%%%%%%%%%
+\subsection{Demographic Parameters}
+
+Adjusting the spatial dispersal and interaction distance, $\sigma$,
+has a surprisingly large effect on demographic quantities that are usually fixed in Wright-Fisher models --
+the generation time, census population size, and variance in offspring number.
+These are shown in Figure~\ref{fig:genparams}.
+This occurs because,
+even through the ``population density'' ($K$) and ``mean lifetime'' ($L$) parameters
+were the same in all simulations,
+the strength of stochastic effects depends strongly on $\sigma$.
+For instance, 
+the population density near to individual $i$ (denoted $n_i$ above)
+is computed by averaging over roughly $N_W = 4 \pi K \sigma^2$ individuals,
+and so has standard deviation proportional to $1/\sqrt{N_W}$ --
+it is more variable at lower densities.
+(Recall that $N_W$ is Wright's neighborhood size.)
+Since the probability of survival is a nonlinear function of $n_i$,
+actual equilibrium densities and lifetimes differ from $K$ and $L$.
+This is the reason that we included \emph{random mating} simulations --
+where mate choice and offspring dispersal are both nonspatial --
+since this should preserve the random fluctuations in local population density
+while destroying any spatial genetic structure.
+We verified that random mating models retained no geographic signal
+by showing that summary statistics did not differ significantly between sampling regimes
+(Table~\ref{table:sampling}), unlike in spatial models (discussed below).
+
+There are a few additional things to note about Figure~\ref{fig:genparams}. 
+First, all three quantities are non-monotone with neighborhood size. 
 Census size largely declines as neighborhood size increases for both the spatial and random mating models. 
-However for spatial models this decline only begins for neighborhood size $\geq 10$.
-By a neighborhood size $\geq 100$ the spatial and random mating models are indistinguishable from one another, a sign that our simulations are performing as expected.
+However, for spatial models this decline only begins for neighborhood size $\geq 10$. 
+By a neighborhood sizes larger than 100, the spatial and random mating models are indistinguishable from one another, 
+a sign that our simulations are performing as expected. 
 Census sizes range from $\approx 14,000$ at low $\sigma$ in the random mating model to $\approx 10,000$ for both models when neighborhood sizes approach 1,000.
@@
 
-Generation times similarly show complex behavior with respect to neighborhood sizes, and vary across the parameter range explored between 5.2 and 4.9 timesteps per generation.
-To stress this point further, the generation time here varies because under our non-Wright-Fisher dynamics individuals are free to reproduce more than once over their lifespan and the lifespan of an individual is only indirectly determined by the parameterization of density-dependent competition the model.
-Under both the spatial and random mating regimes generation time reaches a nadir at a neighborhood size of $\approx 50$.
-Interestingly under the range of neighborhood sizes that we examined, generation times between the random mating and spatial models are never quite equivalent -- presumably this would cease to be the case at neighborhood sizes higher than we simulated here.
+Generation time similarly shows complex behavior with respect to neighborhood sizes, and varies between 5.2 and 4.9 timesteps per genration across the parameter range explored. 
+% This occurs because individuals may reproduce more than once over their lifespan, and the lifespan of an individual is determined by density-dependent competition the model. 
+Under both the spatial and random mating models, 
+generation time reaches a minimum at a neighborhood size of around 50. 
+Interestingly, under the range of neighborhood sizes that we examined, generation times between the random mating and spatial models are never quite equivalent -- presumably this would cease to be the case at neighborhood sizes higher than we simulated here.
 
-Last we looked at the effect of neighborhood size on the variance in number of offspring in the population -- a key parameter determining the effective population size.
-Surprisingly the spatial and random mating model behave quite differently with respect to the neighborhood size: while the variance in offspring number increases monotonically, or nearly so, under the spatial model, the random mating model actually shows a decline in the variance in offspring number until a neighborhood size $\approx 10$ before it increases and eventually equals what we observe in the spatial case.
+Last, we looked at the variance in number of offspring -- a key parameter determining the effective population size. 
+Surprisingly, the spatial and random mating models behave quite differently: 
+while the variance in offspring number increases nearly monotonically under the spatial model, the random mating model actually shows a decline in the variance in offspring number until a neighborhood size $\approx 10$ before it increases and eventually equals what we observe in the spatial case.
 
@@
 
+%%%%%%%%%%%%
 \subsection{Impacts of Continuous Space on Population Genetic Summary Statistics}
-One of the main goals of this study is to examine the effect of continuous space on canonical summaries of population genetic variation.
-Indeed we have little knowledge to date about how some of our most beloved population genetic summary statistics might be affected by such realistic models.
-Moreover, as we will show, sampling strategies of individuals with respect to space can affect summaries of variation at least as strongly as the underlying population dynamics.
+
+Even though certain aspects of population demography depend on the scale of spatial interactions,
+it still could be that population genetic variation is well-described by a well-mixed population model.
+Indeed, mathematical results suggest that genetic variation in some spatial models should be well-approximated by a Wright-Fisher population
+if neighborhood size is large and all samples are geographically widely separated \citep{wilkins2004separation,zahle2005stepping}.
+% One of the main goals of this study is to examine the effect of continuous space on canonical summaries of population genetic variation. 
+However, the behavior of most common population genetic summary statistics has not yet been described in realistic geographic models.
+Moreover, as we will show, spatial sampling strategies can affect summaries of variation at least as strongly as the underlying population dynamics.
 
@@
 
+%%%%%%%%%%%%%%%%
 \subsubsection{Site Frequency Spectra and Summaries of Diversity}
-In Figure \ref{fig:sumstats} we examine the effect of varying neighborhood size and sampling strategy on the site frequency spectrum (Figure \ref{fig:sumstats}A) and several canonical population genetic summary statistics (Figure \ref{fig:sumstats}B).
-For populations in continuous space we observe a significant skew in the SFS for smaller neighborhood sizes ($\leq 100$) that is exacerbated by mid point and point sampling of individuals (see Figure \ref{fig:samplemap}).
-As expected these effects are strongest for the smallest neighborhood sizes examined and are consistently in the direction of an enrichment of intermediate frequency variants in comparison to the standard neutral expectation.
-This can be seen succinctly in the response of Tajima's \textit{D} to variation in neighborhood sizes and sampling regime shown in Figure \ref{fig:sumstats}B, whereby we see \textit{D} turning quite positive at small neighborhood sizes.
-This effect is particularly strong for the midpoint and point sampling regimes, but also occurs under random sampling.
-Notably, the point at which Tajima's $D$ approaches 0 differs strongly across sampling strategies -- varying from a neighborhood size of $\approx50$ for random sampling to $\geq1000$ for midpoint sampling.
 
-One of the most commonly used summaries of variation is Tajima's estimator of the neutral mutation rate $\widehat{\theta}_{\pi}$.
-As can be seen in Figure \ref{fig:sumstats}B $\widehat{\theta}_{\pi}$ can vary as much as nearly three-fold for the smallest neighborhood sizes between the random mating and spatial models depending on the sampling strategies.
-Each sampling strategy approaches the random mating expectation at its own rate, but by neighborhood size of $\approx 100$ all models are equivalent.
-These results are to be expected as at small neighborhood sizes the time to the most recent common ancestor for the population should be pushed back-- i.e.
-it takes longer for lineages to coalesce if dispersal is limited.
-Interestingly the rate of approach to random mating expectations across sampling strategies is reversed relative to that observed in Tajima's D -- midpoint sampling reaches random mating expectations at neighborhood sizes $\approx50$ while random sampling is inflated until neighborhood size $/approx$ 100.
-In effect this means that any estimates of population size based on expected heterozygosity from populations with limited dispersal should be treated with caution.
+Figure \ref{fig:sumstats} shows the effect of varying neighborhood size and sampling strategy on the site frequency spectrum (Figure~\ref{fig:sumstats}A) 
+and several standard population genetic summary statistics (Figure~\ref{fig:sumstats}B). 
+These show a significant enrichment of intermediate frequency variants in comparison to the nonspatial expectation,
+for smaller neighborhood sizes ($\leq 100$) that is exacerbated by midpoint and point sampling of individuals (depicted in Figure~\ref{fig:samplemap}). 
+Reflecting this, Tajima's $D$ is quite positive in the same situations (Figure~\ref{fig:sumstats}B).
+Notably, the point at which Tajima's $D$ approaches 0 differs strongly across sampling strategies 
+-- varying from a neighborhood size of roughly 50 for random sampling to at least 1000 for midpoint sampling.
 
-Patterns from observed heterozygosity and its derivative $F_{IS}$ also depend heavily on neighborhood size under spatial models as well as the sampling scheme.
+One of the most commonly used summaries of variation is Tajima's summary of nucleotide divergence, $\widehat{\theta}_{\pi}$,
+\plr{In the figure this has no hat on; should we remove these in the text? The notation is not standard.}
+calculated as the mean density of nucleotide differences, averaged across pairs of samples.
+As can be seen in Figure \ref{fig:sumstats}B, $\widehat{\theta}_{\pi}$ differs as much as nearly three-fold 
+between the random mating and spatial models.
+$\widehat{\theta}_{\pi}$ using each sampling strategy approaches the random mating expectation at its own rate, 
+but by a neighborhood size of around 100 all models are equivalent. 
+\plr{To interpret this we need to compare to census size over variance in offspring number.}
+The differences between spatial and random mating simulations are much greater than expected 
+from differences in census size divided by variance in offspring number.
+This likely occurs because $\widehat{\theta}_{\pi}$ is a measure of mean time to most recent common ancestor between two samples,
+and at small values of $\sigma$, the time for dispersal to mix ancestry across the range exceeds the mean nonspatial coalescent time.
+(For instance, at the smallest value of $\sigma=0.2$, the range is 250 dispersal distances wide,
+and since the location of a diffusively moving lineage after $k$ generations has variance $k \sigma^2$, 
+it takes around $250^2 = 62500$ generations to mix across the range,
+which is roughly ten times larger than the random mating effective population size.)
+Interestingly, the effect of sampling strategy is reversed relative to that observed in Tajima's D 
+-- midpoint sampling reaches random mating expectations around neighborhood size 50, 
+while random sampling is inflated until around neighborhood size 100. 
+% In effect, this means that any estimates of population size based on expected heterozygosity from populations with limited dispersal should be treated with caution.
+
+Values of observed heterozygosity and its derivative $F_{IS}$ also depend heavily on neighborhood size under spatial models as well as the sampling scheme. 
 $F_{IS}$ is inflated above the expectation across most of the parameter space examined and across all sampling strategies. 
-This effect is caused by a deficit of heterozygous individuals in low-dispersal simulations -- a continuous-space version of the Wahlund effect \citep{Wahlund1928}.
-Indeed for random sampling under the spatial model $F_{IS}$ does not approach the random mating equivalent until neighborhood sizes of nearly 1000.
-The patterns for raw observed heterozygosity are more complex.
-Under midpoint sampling observed heterozygosity is inflated even over the random mating expectation, as a result of the a higher proportion of heterozygotesn occurring in the middle of the landscape (Figure S\ref{fig:hetmap}).
+This effect is caused by a deficit of heterozygous individuals in low-dispersal simulations 
+-- a continuous-space version of the Wahlund effect \citep{Wahlund1928}. 
+Indeed, for random sampling under the spatial model, $F_{IS}$ does not approach the random mating equivalent until neighborhood sizes of nearly 1000. 
+On the other hand, the dependency of raw observed heterozygosity on neighborhood size is not monotone.
+Under midpoint sampling observed heterozygosity is inflated even over the random mating expectation, 
+as a result of the a higher proportion of heterozygotes occurring in the middle of the landscape (Figure~\ref{fig:hetmap}). 
 This squares with a report from \cite{Shirk2014} who observed a similar excess of heterozygosity in the middle of the landscape when simulating under a lattice model.
 
+\plr{peter to revisit this paragraph}
 Trends in pairwise haplotype block sharing parallel those in allele-frequency-based diversity estimates (Figure \ref{fig:sumstats}, Supplementary Figure \ref{fig:allsumstats}).
-At low dispersal the distribution of IBS block lengths in a set of samples is shifted towards smaller values with respect to the random mating expectation-- resulting in lower means and fewer long IBS blocks.
+At low dispersal the distribution of IBS block lengths in a set of samples is shifted towards smaller values with respect to the random mating expectation -- resulting in lower means and fewer long IBS blocks.
 The variance and skew of the distribution of haplotype block lengths are only minorly affected by neighborhood size in our simulations when calculated across all pairs of individuals; however, they are strongly dependent on sampling regimes.
@@
 Correlating population genetic summaries such as $F_{ST}$ against geographic distance has shown great utility in empirical population genetics \citep{Rousset1997}.
-As we know the exact locations of individuals that are sampled in our simulated populations we have examined the relationship of geographic distance among samples with a number of our summary statistics (Figure \ref{fig:sumstats} and Figure \ref{fig:allsumstats}).
-Starting with a measure of population differentiation,$D_{xy}$, we observe a positive correlation with distance that declines as dispersal increases, as expected under the theory developed by \citep{Rousset1997} and others.
-This relationship varies across sampling strategies of course, with the weakest correlations observed for midpoint sampling.
-Moreover there is clearly strong signal for  neighborhood size in the strength of the correlation between $D_{xy}$ and geographic distance.
+As we know the exact locations of individuals that are sampled in our simulated populations we have examined the relationship of geographic distance between samples with a number of summary statistics (Figures \ref{fig:sumstats} and \ref{fig:allsumstats}).
+Mean density of nucleotide differences between individuals, $D_{xy}$, is positively correlated with the geographic distance between the individuals,
+and the strength of this correlation declines as dispersal increases, as expected under theory \citep{Wright1943,Rousset1997}. 
+This relationship varies across sampling strategies, with the weakest correlations observed for midpoint sampling,
+perhaps due to a dearth of long-distance comparisons.
+% Moreover there is clearly strong signal for neighborhood size in the strength of the correlation between $D_{xy}$ and geographic distance.
 
+\plr{peter to revisit this paragraph}
 We next turn our attention to the effect of geographic distance on haplotype block length sharing. 
-As in \citep{Ringbauer2017} and \citep{Baharian2016} we found that the pairwise distribution of haplotype block lengths is more strongly left-skewed under limited dispersal.
+As in \citet{Ringbauer2017} and \citet{Baharian2016} we found that the pairwise distribution of haplotype block lengths is more strongly left-skewed under limited dispersal.
 This is reflected in negative correlation coefficients between spatial distance and the mean, variance, skew, and count of long blocks from the pairwise distribution of identical-by-state block lengths (Figure \ref{fig:sumstats} and Figure \ref{fig:allsumstats}).
@@
 
-\subsubsection{Effects of Sampling on Random Mating and Spatial Models}
-To summarize the effect of sampling strategy on genetic diversity in random mating and spatial models we also asked if summary statistics varied significantly across sampling regimes.
-In table \ref{table:sampling} we show that for spatial models the mean and variance of nearly all summary statistics was significantly different across sampling strategies (Table \ref{table:sampling}) for spatial models, but not for random mating models.
-In summary, sampling strategy can be safely ignored when the population is randomly mating but will shape estimates of genetic diversity in any population with limited dispersal.
+%%%%%%%%%%%
+\subsubsection{Spatial distribution of allele copies}
+\plr{insert bit describing that plot here}
 
@@
 
-In Figure \ref{fig:demography} we show inferred population size histories from our simulations as a function of neighborhood size binned in to one of four categories.
-In general we observe that demographic models from Stairwayplot tended to infer patterns of ancient population increases and recent declines when neighborhood sizes were below 20 under all sampling strategies (Figure~\ref{fig:demography}).
+\plr{revsit with smc++}
+Figure \ref{fig:demography} shows inferred population size histories, grouped by neighborhood size and sampling strategy.
+In general, Stairwayplot tends to infer ancient population increases and recent declines when neighborhood sizes were below 20 under all sampling strategies (Figure~\ref{fig:demography}). 
 This is consistent with our observations of the SFS from which Stairwayplot is doing its inference.
@@
 While that is so individual model fits were highly variable and often inferred five-fold or greater population fluctuations even in high-dispersal simulations.
-To test whether the variation we observed in inferred demographic histories from Stairwayplot was the result of spatial effects in our simulations rather than the behavior of the optimization routine we compared standard deviations of inferred population sizes in each sampling/neighborhood size bin with those returned by an equilibrium coalescent simulation.
+We compared this range of variation to Stairwayplot results run on coalescent simulations with constant population size,
+and found that the noisiest results ($N_W < 20$, or midpoint sampling and $N_W < 100$)
+were noiser than expected, but remaining simulation results were consistent with Stairwayplot's behavior on data from random mating models (Table~\ref{table:demography}).
+% To test whether the variation we observed in inferred demographic histories from Stairwayplot was the result of spatial effects in our simulations rather than the behavior of the optimization routine we compared standard deviations of inferred population sizes in each sampling/neighborhood size bin with those returned by an equilibrium coalescent simulation.
 Were these to be equal we could assume that the variation were purely as to be expected under the noise inherent in the genealogical process.
@@
 
+%%%%%%%%%%%%%%%%%
 \subsection{GWAS}
-To ask what confounding effects spatial genetic variation might have on genome-wide association studies we performed GWAS on our simulations using phenotypes that were determined solely by the environment.
-In general we found that for simulations with limited dispersal, i.e.
-neighborhood size $< 100$, spatial variation in the environment causes GWAS to infer significant associations with purely environmental phenotypes at over 25\% of sampled SNPs if no correction for genetic relatedness among samples is performed (Figure \ref{fig:gwas}).
-This effect is particularly strong for clinal and corner environments, for which the lowest dispersal levels cause over 60\% of SNPs in the sample to return significant associations.
-Patchy environmental distributions, which result in fewer extreme phenotype values in the sample (Figure \ref{fig:gwas}A), cause fewer false-positives overall but still produce spurious associations at roughly 10\% of sites at the lowest neighborhood sizes.
-Notably no simulations with nonspatial environments returned more than one significant association (Figure \ref{fig:gwas}C), demonstrating that this effect is caused specifically by the interaction of population structure and spatial variation in the environment rather than by population structure itself.
 
-When PCA positions are included as covariates to control for population structure in GWAS the vast majority of SNPs no longer surpass a 5\% FDR significance threshold, but up to 1.5\% of SNPs are significantly associated at low dispersal distances under corner and patchy environmental distributions (Figure \ref{fig:gwas}C).
-At neighborhood sizes $> 500$ up to 0.31\% of SNPs were significant for corner and clinal environments.
+To ask what confounding effects spatial genetic variation might have on genome-wide association studies we performed GWAS on our simulations using phenotypes that were determined solely by the environment
+-- so, any SNP showing statistically significant correlation with phenotype is a false positive. 
+As expected, spatial autocorrelation in the environment causes spurious associations 
+across much of the genome if no correction for genetic relatedness among samples is performed (Figure \ref{fig:gwas}). 
+This effect is particularly strong for clinal and corner environments, 
+for which the lowest dispersal levels cause over 60\% of SNPs in the sample to return significant associations. 
+Patchy environmental distributions, which is less strongly spatially correlated (Figure \ref{fig:gwas}A), 
+cause fewer false positives overall but still produce spurious associations at roughly 10\% of sites at the lowest neighborhood sizes. 
+Notably, no simulations with nonspatial environments returned more than one significant association (Figure \ref{fig:gwas}C), 
+demonstrating that this effect is caused specifically by the interaction of population structure and spatial variation in the environment rather than by population structure itself.
+
+The confounding effects of geographic structure are well-known, 
+and it is common practice to control for this by including principal components (PCs) as covariates to control for these effects.
+This mostly works in our simulations -- after doing this, the vast majority of SNPs no longer surpass a 5\% FDR significance threshold. 
+However, a substantial number of SNPs -- up to 1.5\% of SNPs -- still surpass this threshold (and thus would be false positives in a GWAS),
+especially under ``corner'' and ``patchy'' environmental distributions (Figure~\ref{fig:gwas}C). 
+At neighborhood sizes larger than 500, up to 0.31\% of SNPs were significant for corner and clinal environments. 
 Given an average of 132,000 SNPs across simulations after MAF filtering, this translates to up to 382 false-positive associations. 
@@
 
-Clinal environments cause an interesting pattern in false positives after PC correction: at low neighborhood sizes the correction removes nearly all significant associations, but at neighborhood sizes above $\approx250$ the proportion of significant SNPs increases to up to 0.4\% (Figure \ref{fig:gwas}.
-This appears to reflect a loss of descriptive power in the PCA -- as neighborhood size increases, the total proportion of variance explained by the first 10 PC axes declines from roughly 0.1 to 0.04 (Figure \ref{fig:gwas}B).
+Clinal environments cause an interesting pattern in false positives after PC correction: at low neighborhood sizes the correction removes nearly all significant associations, 
+but at neighborhood sizes above $\approx250$ the proportion of significant SNPs increases to up to 0.4\% (Figure~\ref{fig:gwas}). 
+This may be due to a loss of descriptive power of the PCs
+-- as neighborhood size increases, the total proportion of variance explained by the first 10 PC axes declines from roughly 10\% to 4\% (Figure \ref{fig:gwas}B). 
 Essentially, PCA seems unable to effectively summarize the weak population structure present in large-neighborhood simulations, but these populations continue to have enough spatial structure to create significant correlations between genotypes and the environment.
@@
 
-In Figure \ref{fig:gwas}D we present quantile-quantile plots that show the degree of genome-wide inflation of test statistics in PC-corrected GWAS across all simulations and environmental distributions.
-For clinal environments $-log_{10}(p)$ values are most inflated when neighborhood sizes are large, consistent with the pattern observed in the count of significant associations after PC regression.
-In contrast corner and patchy environments cause the greatest inflation in $-log_{10}(p)$ at neighborhood sizes $<100$, which likely reflects the inability of PCA to account for fine-scale structure caused by very limited dispersal.
-Finally, we observed that PC regression appears to cause some degree of overcorrection for all phenotype distributions, visible in Figure \ref{fig:gwas}D as points falling below the 1:1 line.
+Figure \ref{fig:gwas}D shows quantile-quantile plots that show the degree of genome-wide inflation of test statistics in PC-corrected GWAS across all simulations and environmental distributions. 
+For clinal environments, $-\log_{10}(p)$ values are most inflated when neighborhood sizes are large, consistent with the pattern observed in the count of significant associations after PC regression.
+In contrast corner and patchy environments cause the greatest inflation in $-\log_{10}(p)$ at neighborhood sizes $<100$, which likely reflects the inability of PCA to account for fine-scale structure caused by very limited dispersal.
+Finally, we observed that PC regression appears to overfit to some degree for all phenotype distributions, visible in Figure \ref{fig:gwas}D as points falling below the 1:1 line.
+
+\plr{Add a few manhattan plots to the supplement and refer to them somewhere.}
 
@@
 Points above the 1:1 line reflect deflated $p$ values.
+    \plr{suggestion: (a) log-scale the y axes in (C) and add another y-axis on the right showing (rough number of human genome SNPs above 5\% times this percentage) so we can see absolute numbers}
 }
@@
 Patterns of genetic variation are influenced by the fact that organisms are more likely, on average, to reproduce with others of their species that are geographically proximate, in particular when dispersal distances are short.
-While historically some analytical progress has been made in describing the influence of continuous space on population genetics (e.g.
-\cite{Wright1943,Rousset1997,Ringbauer2017,Barton2010}), the theoretical work is challenging and often untenable for realistic biological models.
+While historically some analytical progress has been made in describing the influence of continuous space on population genetics (e.g., \cite{Wright1943,Rousset1997,Ringbauer2017,Barton2010}), the theoretical work is challenging and often untenable for realistic biological models.
 Instead we here use efficient forward time population genetic simulations to describe the myriad influence of space on genetic variation.
@@
 At the extreme low end of dispersal distance this can result in an up to three-fold increase in genetic diversity relative to random-mating expectations.
-These effects are strongest when neighborhood sizes are below 100, but in combination with the effects of nonrandom sampling they can persist up to neighborhood sizes of at least 1000 (e.g.
-inflation in Tajima's $D$ and observed heterozygosity under midpoint sampling).
+These effects are strongest when neighborhood sizes are below 100, but in combination with the effects of nonrandom sampling they can persist up to neighborhood sizes of at least 1000 (e.g., inflation in Tajima's $D$ and observed heterozygosity under midpoint sampling).
 Under random sampling the general pattern is similar to expectations of the original analytic model of \cite{Wright1943}, which predicts that populations with neighborhood sizes under 100 will differ substantially from random mating, while those above 10,000 will be nearly indistinguishable from panmixia.
@@
 In our simulations we observed that groups of nearby individuals tend to coalesce rapidly, while coalescence among groups at opposite ends of the landscape takes much longer than under random mating.
-Tip branches (i.e.
-branches subtending only one individual) are then relatively short, and branches in the middle of the genealogy connecting local groups of individuals relatively long.
+Tip branches (i.e., branches subtending only one individual) are then relatively short, and branches in the middle of the genealogy connecting local groups of individuals relatively long.
 These patterns then create the biases we observed in the SFS -- the lowest frequency bins are deflated by the short branch lengths connecting nearby individuals, while mid-frequency bins are inflated by the long branches connecting local groups.
@@
 In summary, empirical researchers should collect individuals in as random a manner as practical and should not bias their sample towards the middle of a species' range.
-When sampling is clustered, summary statistics based on segregating sites (e.g.
-Watterson's $\Theta$ and Tajima's $D$), heterozygosity, or the distribution of long haplotype blocks, can be expected to depart significantly from what would be observed under random sampling.
+When sampling is clustered, summary statistics based on segregating sites (e.g., Watterson's $\Theta$ and Tajima's $D$), heterozygosity, or the distribution of long haplotype blocks, can be expected to depart significantly from what would be observed under random sampling.
 Comparing the results of analyses conducted on all individuals versus those limited to single individuals per locality may help to reveal any effect of sampling bias if sufficient samples are available.
@@
 
-Stepping back from the mechanics of GWAS specifically, the approach of quantitative genetics is to decompose the variance in phenotypes into environmental and genetic effects, e.g. 
+Stepping back from the mechanics of GWAS specifically, the approach of quantitative genetics is to decompose the variance in phenotypes into environmental and genetic effects, e.g., 
 \begin{equation}
@@
 In essence with a PC-corrected GWAS we are asking ``what regions of the genome are more associated with this phenotype than the average genome-wide association observed across populations?'' In our simulations we observed that this procedure can fail under a variety of circumstances.
-If dispersal is limited and environmental variation is clustered in space (i.e.
-corner or patchy distributions in our simulations), PCA positions fail to capture the fine-scale spatial structure required to remove all signals of association.
+If dispersal is limited and environmental variation is clustered in space (i.e., corner or patchy distributions in our simulations), PCA positions fail to capture the fine-scale spatial structure required to remove all signals of association.
 Conversely when dispersal is high we found that PCA loses power to describe population structure before the spatial scale of dispersal breaks down the relationship between genotype and the environment.
@@
 This result is consistent with a recent empirical study of heritability in human height and body mass index, which found that increasing the number of PC axes used as covariates caused the total proportion of variance explained by SNPs to decline from $\approx0.8$ to $\approx0.75$ \citep{Wainschtein2019}.
-Indeed SNPs with minor alleles frequencies of 0.001 - 0.01, which are expected to reflect fine scale population structure \citep{Mathieson2012,Novembre2009}, are estimated to explain \textit{negative} proportions of the total phenotypic variance in \citep{Wainschtein2019}\ak{i don't understand what the authors mean here-- how can we have a negative variance?}.
+Indeed SNPs with minor alleles frequencies of 0.001 - 0.01, which are expected to reflect fine scale population structure \citep{Mathieson2012,Novembre2009}, are estimated to explain \textit{negative} proportions of the total phenotypic variance in \citep{Wainschtein2019}\ak{i don't understand what the authors mean here -- how can we have a negative variance?}.
 Searching for genetic associations of polygenic traits that vary systematically across but not within populations through existing GWAS approaches is then unlikely to be successful: the signals are fully confounded, and new analytic methods or experiments controlling for variation in the environment will be necessary to rigorously identify causal variants.
@@
 As we have shown, a large number of population genetic summary statistics contain information about spatial population processes.
-We imagine that combinations of such summaries might be sufficient for the construction of supervised machine learning regressors (e.g.
-\cite{Schrider2018}) for the accurate estimation of dispersal from genetic data.
+We imagine that combinations of such summaries might be sufficient for the construction of supervised machine learning regressors (e.g., \cite{Schrider2018}) for the accurate estimation of dispersal from genetic data.
 Indeed \cite{Ashander2018} found that inverse interpolation on a vector of summary statistics provided a powerful method of estimating dispersal distances.
@@
 Indeed, a series of field studies described in \cite{Schemske2007} found that in Wright's original empirical example of isolation by distance, the flowering plant \textit{Linanthus parryae}, patterns of flower color differentiation over space primarily reflect temporal and spatial variation in selection rather than limited dispersal.
-Studies simulating selection and dispersal interacting in space (e.g.
-\cite{Ralph2010}) and testing for identifiability of inferred dispersal or selective parameters may offer new insight into the extent of our ability to accurately infer evolutionary processes in real systems.
+Studies simulating selection and dispersal interacting in space (e.g., \cite{Ralph2010}) and testing for identifiability of inferred dispersal or selective parameters may offer new insight into the extent of our ability to accurately infer evolutionary processes in real systems.
 
@@
 $mean(IBS>1e6)$ & \makecell[l]{Mean number of IBS tracts over $1\times10^6$bp per pair \\across all pairs in the sample.} \\ 
-$corr(D_{xy},dist)$ & Pearson correlation between genetic distance and $log_{10}(spatial\text{ }distance)$ \\
-$corr(mean(IBS),dist)$ & \makecell[l]{Pearson correlation between the mean of the IBS tract distribution \\for each pair of samples and $log_{10}(spatial\text{ }distance)$} \\
-$corr(nIBS,dist)$ & \makecell[l]{Pearson correlation between the number of IBS tracts \\for each pair of samples and $log_{10}(spatial\text{ }distance)$} \\
-$corr(IBS>1e6,dist)$ & \makecell[l]{Pearson correlation between the number of IBS tracts $> 1\times10^6$bp \\for each pair of samples and $log_{10}(spatial\text{ }distance)$} \\
-$corr(skew(IBS),dist)$ & \makecell[l]{Pearson correlation between the skew of the distribution of pairwise haplotype\\ block lengths for each pair of samples and $log_{10}(spatial\text{ }distance)$} \\
+$corr(D_{xy},dist)$ & Pearson correlation between genetic distance and $\log_{10}(spatial\text{ }distance)$ \\
+$corr(mean(IBS),dist)$ & \makecell[l]{Pearson correlation between the mean of the IBS tract distribution \\for each pair of samples and $\log_{10}(spatial\text{ }distance)$} \\
+$corr(nIBS,dist)$ & \makecell[l]{Pearson correlation between the number of IBS tracts \\for each pair of samples and $\log_{10}(spatial\text{ }distance)$} \\
+$corr(IBS>1e6,dist)$ & \makecell[l]{Pearson correlation between the number of IBS tracts $> 1\times10^6$bp \\for each pair of samples and $\log_{10}(spatial\text{ }distance)$} \\
+$corr(skew(IBS),dist)$ & \makecell[l]{Pearson correlation between the skew of the distribution of pairwise haplotype\\ block lengths for each pair of samples and $\log_{10}(spatial\text{ }distance)$} \\
 \end{tabular}

### 2019-04-28 ms/spaceness.tex
peter edited mostly through

@@
 Here we simulate both the underlying genealogy and phenotypes of a continuously distributed population and seek to identify regions of parameter space -- i.e., the strength of isolation by distance and the spatial distribution of environmental effects on phenotypes -- in which common methods employed for GWAS are subject to confounding of environmental and genetic effects on phenotypes.
+
+\plr{Moved from disco: integrate citations.}
+Patterns of genetic variation are influenced by the fact that organisms are more likely, on average,
+to live nearby to both their parents, in particular when dispersal and mate choice distances are short. 
+Analytical work describing the genetics of populations distributed across continuous geography
+(e.g., \cite{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation})
+the theoretical work is challenging and often untenable for realistic biological models. 
+Also discuss lots of work on grids, e.g., \citet{Shirk2014,rousset2011likelihoodbased}.
 
@@
 Given an average of 132,000 SNPs across simulations after MAF filtering, this translates to up to 382 false-positive associations. 
-In most cases the $p$ values for these associations were significant after FDR correction but would not pass the threshold for significance under the more conservative Bonferroni correction, suggesting that the most strongly associated variants from many GWAS of mono- or oligogenic traits are robust to this variety of stratification bias.
+In most cases the $p$ values for these associations were significant after FDR correction but would not pass the threshold for significance under the more conservative Bonferroni correction.
 
@@
 Spatially correlated environmental factors cause false positives at a large proportion of SNPs, which is partially but not entirely corrected by adding PC positions as covariates (\textbf{C}).
-Quantile-quantile plots in \textbf{D} show inflation of $-log10(p)$ after PC correction across all simulations and environmental distributions, with colors scaled by the median neighborhood size in each region of q-q space.
+Quantile-quantile plots in \textbf{D} show inflation of $-\log_{10}(p)$ after PC correction across all simulations and environmental distributions, with colors scaled by the median neighborhood size in each region of q-q space.
 %A: example environmental distributions and sampling maps, with colors scaled to phenotype values.
@@
 
+%%%%%%%%%%%%%%%%%%%%
 \section{Discussion}
-Patterns of genetic variation are influenced by the fact that organisms are more likely, on average, to reproduce with others of their species that are geographically proximate, in particular when dispersal distances are short.
-While historically some analytical progress has been made in describing the influence of continuous space on population genetics (e.g., \cite{Wright1943,Rousset1997,Ringbauer2017,Barton2010}), the theoretical work is challenging and often untenable for realistic biological models.
-Instead we here use efficient forward time population genetic simulations to describe the myriad influence of space on genetic variation.
-In particular we examine three axes of variation across three different sampling strategies -- 1) population genetic summary statistics, 2) inference of population size history, and 3) the consequences on genome-wide association studies (GWAS).
-We are in particular interested in asking how our empirical inferences from data might be affected by spatial processes.
-As we show below the answers seems to be - often space matters, both because of how populations are sampled with respect to space, and because of the inherent dispersal properties of those populations.
 
+In this study,
+we have used efficient forward time population genetic simulations 
+to describe the myriad influence of continuous geography on genetic variation. 
+In particular, we examine how three main types of downstream empirical inference are affected by unmodeled spatial population structure
+-- 1) population genetic summary statistics, 2) inference of population size history, and 3) genome-wide association studies (GWAS). 
+As discussed above, 
+space often matters (and sometimes dramatically),
+both because of how samples are arranged in space, and because of the inherent patterns of relatedness established by geography.
+
+%%%%%%%%%%
 \subsection{Effects of Dispersal}
-Limited dispersal inflates effective population size, creates correlations between genetic and spatial distances, and introduces subtle biases in the site frequency spectrum that are reflected in a positive Tajima's $D$ (Figure \ref{fig:sumstats}).
-At the extreme low end of dispersal distance this can result in an up to three-fold increase in genetic diversity relative to random-mating expectations.
+
+Limited dispersal inflates effective population size, creates correlations between genetic and spatial distances, 
+and introduces strong distortions in the site frequency spectrum that are reflected in a positive Tajima's $D$ (Figure \ref{fig:sumstats}). 
+At the lowest dispersal distances, this can increase genetic diversity threefold relative to random-mating expectations. 
 These effects are strongest when neighborhood sizes are below 100, but in combination with the effects of nonrandom sampling they can persist up to neighborhood sizes of at least 1000 (e.g., inflation in Tajima's $D$ and observed heterozygosity under midpoint sampling).
@@
 Nearby individuals coalesce rapidly under limited dispersal and so are connected by short branch lengths, while distant individuals take much longer to coalesce than they would under random mating.
-Mutation and recombination events in our simulation both occur at a constant rate along branches of the genealogy, so the genetic distance and number of recombination events separating two individuals is simply a noisy estimate of the branch lengths connecting them.
-These genealogical patterns also relate directly to the site frequency spectrum.
-In our simulations we observed that groups of nearby individuals tend to coalesce rapidly, while coalescence among groups at opposite ends of the landscape takes much longer than under random mating.
+Mutation and recombination events in our simulation both occur at a constant rate along branches of the genealogy, 
+so the genetic distance and number of recombination events separating sampled individuals simply gives a noisy picture of the genealogies connecting them. 
+% These genealogical patterns also relate directly to the site frequency spectrum. 
+% In our simulations we observed that groups of nearby individuals tend to coalesce rapidly, while coalescence among groups at opposite ends of the landscape takes much longer than under random mating.
 Tip branches (i.e., branches subtending only one individual) are then relatively short, and branches in the middle of the genealogy connecting local groups of individuals relatively long.
@@
 This is because identical-by-state tract lengths reflect the impacts of two processes acting along the branches of the underlying genealogy -- both mutation and recombination -- rather than just mutation as is the case when looking at the site frequency spectrum or related summaries.
-This means that the pairwise distribution of haplotype block lengths carries with it important information about genealogical variation in the population, and correlation coefficients between moments of the this distribution and geographic location contain signal similar to the correlations between $F_{ST}$ or $D_{xy}$ and space \citep{Rousset1997}.
+This means that the pairwise distribution of haplotype block lengths carries with it important information about genealogical variation in the population, and correlation coefficients between moments of the this distribution and geographic location contain signal similar to the correlations between $F_{ST}$ or $D_{xy}$ and geographic distance \citep{Rousset1997}.
 Indeed this basic logic underlies two recent studies explicitly estimating dispersal from the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}.
@@
 \subsection{Effects of Sampling}
-One of the most important differences between random mating and spatial models is the effect of sampling: in a randomly mating population the spatial distribution of sampling effort has no effect on estimates of genetic variation, but when dispersal is limited sampling strategy can compound spatial patterns in the underlying genealogy and create pervasive impacts on all downstream genetic analyses.
-As expected we found that random sampling provides the most accurate summary of genetic diversity across the landscape.
-However this strategy is often impractical for empirical studies.
-In reality the difficulty of traveling through all parts of a species range and the inefficiency of collecting single individuals at each sampling site means that most studies follow something closer to the ``point'' sampling strategy we simulated, in which multiple individuals are sampled from nearby points on the landscape.
+One of the most important differences between random mating and spatial models is the effect of sampling: 
+in a randomly mating population the spatial distribution of sampling effort has no effect on estimates of genetic variation, 
+but when dispersal is limited sampling strategy can compound spatial patterns in the underlying genealogy and create pervasive impacts on all downstream genetic analyses. 
+In most species,
+the difficulty of traveling through all parts of a species range and the inefficiency of collecting single individuals at each sampling site 
+means that most studies follow something closest to the ``point'' sampling strategy we simulated, 
+in which multiple individuals are sampled from nearby points on the landscape. 
 For example, in ornithology a sample of 10 individuals per species per locality is a common target when collecting for natural history museums. 
-In classical studies of \textit{Drosophila} variation the situation is considerably worse, in which a single orchard might be sampled with bated traps for instance.
+In classical studies of \textit{Drosophila} variation the situation is considerably worse, in which a single orchard might be extensively sampled.
 
@@
 For some data summaries, such as Tajima's $D$, Watterson's $\Theta$, or the correlation coefficient between spatial distance and the count of long haplotype blocks, this can result in large differences in estimates between random and point sampling (Figure \ref{fig:sumstats}).
-Inferring underlying demographic parameters from these summary statistics -- for example, estimating dispersal distance as the slope of a regression of $F_{ST}$ against the logarithm of spatial distance \citep{Rousset1997} -- may then be subject to bias if sampling is not random across the landscape.
+Inferring underlying demographic parameters from these summary statistics 
+-- unless the nature of the sampling is somehow taken into account 
+-- will be subject to bias if sampling is not random across the landscape.
 
@@
 This model is meant to reflect a bias in sampling effort towards the middle of a species' range. 
-In empirical studies this sampling strategy could arise if, for example, researchers choose to sample the center of the range and avoid range edges to maximize probability of locating individuals during a short field season.
+In empirical studies this sampling strategy could arise if, for example, 
+researchers choose to sample the center of the range and avoid range edges to maximize probability of locating individuals during a short field season. 
 Because midpoint sampling provides limited spatial resolution it dramatically reduces the magnitude of observed correlations between spatial and genetic distances. 
 More surprisingly, midpoint sampling also leads to strongly positive Tajima's $D$ and an inflation in the proportion of heterozygous individuals in the sample. 
-This increase in observed heterozygosity appears to reflect the effects of range edges, which are a fundamental facet of spatial genetic variation that have often been ignored by analytic approaches focusing on infinite toroidal landscapes \citep{Felsenstein1975}.
+This increase in observed heterozygosity appears to reflect the effects of range edges, which are a fundamental facet of spatial genetic variation. 
 If individuals move randomly in a finite two-dimensional landscape then regions in the middle of the landscape receive migrants from all directions while those on the edge receive no migrants from at least one direction.
@@
 Though here we used only a single parameterization of fitness decline at range edges we believe this is a general property of non-infinite landscapes as it has also been observed in previous studies simulating under lattice models \citep{Neel2013,Shirk2014}.
-For empirical studies this suggests that sampling only the middle of the landscape will produce a biased view of many aspects of the genetic variation in a population.
 
-In summary, empirical researchers should collect individuals in as random a manner as practical and should not bias their sample towards the middle of a species' range.
-When sampling is clustered, summary statistics based on segregating sites (e.g., Watterson's $\Theta$ and Tajima's $D$), heterozygosity, or the distribution of long haplotype blocks, can be expected to depart significantly from what would be observed under random sampling.
-Comparing the results of analyses conducted on all individuals versus those limited to single individuals per locality may help to reveal any effect of sampling bias if sufficient samples are available.
-Alternatively, sampling strategy could be incorporated directly into a simulation-based inferential framework by, for example, setting the  distribution of samples in a simulation to mimic that used for the empirical target case.
-This could be readily achieved in a supervised machine learning or approximate Bayesian setting (CITES).
+In summary, empirical researchers should collect individuals from across as much of the species' range as practical,
+choosing samples at a variety of spatial scales.
+Many summary statistics are designed for well-mixed populations,
+and so provide different insights into genetic variation
+when applied to different subsets of the population.
+Applied to a cluster of samples, summary statistics based on segregating sites 
+(e.g., Watterson's $\Theta$ and Tajima's $D$), 
+heterozygosity, or the distribution of long haplotype blocks, 
+can be expected to depart significantly from what would be obtained from a wider distribution of samples.
+Comparing the results of analyses conducted on all individuals versus those limited to single individuals per locality 
+can provide an informative contrast.
 
+%%%%%%%%%%%
 \subsection{Demography}
-Classical population genetic models collapse many elements of life history variation into a single parameter, $N_{e}$, which is then taken to reflect the degree of variation present in the population when modeling the effects of selection or migration.
-Inferring $N_{e}$ in the past is now a common goal of population genomic analyses and an important step in establishing baseline expectations of genetic variation when searching for signals of selection.
+
+Classical results in population genetics collapse many aspects of life history into a single compound parameter, effective population size ($N_{e}$).
+Inferring effective population size back through time is now a common goal of population genomic analyses 
+and an important step in establishing baseline expectations of genetic variation, e.g., when searching for signals of selection. 
 Here we found that one method of inference of historic $N_{e}$ based on genome-wide estimates of the site frequency spectrum, Stairwayplot \citep{Liu2015}, is relatively robust to variation in dispersal distance when sampling is random and neighborhood size is over 20.
@@
 
+%%%%%%%%%%%
 \subsection{GWAS}
-Over the last twenty years genome-wide association studies (GWAS) have identified tens of thousands of correlations between genetic variation and phentoypes, both in humans and other species.
-This technique is increasingly applied to questions of human health through methods like polygenic risk scores that sum the effect sizes estimated from GWAS to predict an individual's phenotype or disease risk \citep{Khera2018}.
-The most common approach to GWAS, which we followed in our analyses of simulated data here, is to regress the count of derived alleles at a site against individual phenotypes, taking the slope of this regression as an estimate of the effect of the allele on the phenotype.
+
+Over the last twenty years, genome-wide association studies (GWAS) have identified tens of thousands of correlations between genetic variation and phentoypes, 
+both in humans and other species. 
+This technique is increasingly applied to questions of human health through methods like polygenic risk scores 
+that sum the effect sizes estimated from GWAS to predict an individual's phenotype or disease risk \citep{Khera2018}. 
+The most common method, which we followed in our analyses of simulated data here, is to regress the count of derived alleles at a site against individual phenotypes, 
+taking the slope of this regression as an estimate of the effect of the allele on the phenotype. 
 As recently reviewed by \cite{Visscher2019}, this is exactly the approach outlined by \cite{Fisher1918} at the dawn of quantitative genetics.
 
-Stepping back from the mechanics of GWAS specifically, the approach of quantitative genetics is to decompose the variance in phenotypes into environmental and genetic effects, e.g., 
-\begin{equation}
-P=G+E
-\end{equation}
-\begin{equation}
-var(P)=var(G)+var(E)+cov(G,E)
-\end{equation}
-For GWAS in structured populations, the bias identified in many previous studies \citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015} in which test statistics reflect population structure rather than phenotype association arises because of the presence of a positive covariance between genotypic and environmental variation (the third term above).
-When the environment and genotype covary across space their effects are confounded.
-Note this does not require interaction between genotype and environment (so-called GxE effects), which will introduce an additional source of bias.
-Here we refer simply to the fact that the allele frequency at a site will often covary with the environment when both breeding structure and the environment vary over space.
-To some degree the success of quantitative genetics in fields like agriculture is likely due to the absence of this covariation -- crop and animal breeding operations can be conducted so that the environment is identical (or nearly so) or randomized across populations.
-However in natural populations this situation is extremely unlikely.
-Most populations are structured by a combination of limited dispersal and geographic barriers, and nearly all environments vary over space.
-GWAS in natural populations is forced to confront the confounding effects of population structure and the environment directly.
+Stepping back from specific methods,
+the general goal is to identify alleles that are more correlated with the phenotype of interest
+than one would expect due to chance.
+Both phenotype and genotype are often spatially autocorrelated (for different reasons),
+which makes spurious correlations a major obstacle,
+both by creating false positives and reducing power to identify truly causal alleles 
+\citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015}.
+It may be that the success of quantitative genetics in agriculture is in part because 
+crop and animal breeding operations can be conducted so that the environment is identical (or nearly so) or randomized across populations. 
+However, most natural populations are structured by a combination of limited dispersal and geographic barriers, 
+and nearly all environments vary over space.
 
 Incorporating PC positions as covariates in the analysis \citep{Price2006} is designed to address this difficulty by regressing out a baseline level of ``average'' differentiation. 
-However while this approach is quite useful, it does not truly separate the confounding signals of environment and spatially varying genotypes.
-In essence with a PC-corrected GWAS we are asking ``what regions of the genome are more associated with this phenotype than the average genome-wide association observed across populations?'' In our simulations we observed that this procedure can fail under a variety of circumstances.
-If dispersal is limited and environmental variation is clustered in space (i.e., corner or patchy distributions in our simulations), PCA positions fail to capture the fine-scale spatial structure required to remove all signals of association.
-Conversely when dispersal is high we found that PCA loses power to describe population structure before the spatial scale of dispersal breaks down the relationship between genotype and the environment.
-These effects were observed in all spatially varying environmental distributions, but were particularly pronounced for concentrated environmental effects in one region, as was also found in \cite{Mathieson2012}.
+% This approach is extremely useful, but does not truly separate the confounding signals of environment and spatially varying genotypes. 
+In essence, a PC-corrected GWAS asks ``what regions of the genome are more associated with this phenotype than the average genome-wide association observed across populations?'' 
+In our simulations, we observed that this procedure can fail under a variety of circumstances. 
+If dispersal is limited and environmental variation is clustered in space 
+(i.e., corner or patchy distributions in our simulations), 
+PCA positions fail to capture the fine-scale spatial structure required to remove all signals of association. 
+Conversely, as dispersal is increased, PCA loses power to describe population structure 
+before spatial mixing breaks down the relationship between genotype and the environment. 
+These effects were observed with all spatially correlated environmental patterns,
+but were particularly pronounced if environmental effects are concentrated in one region, as was also found by \citet{Mathieson2012}. 
+\plr{need to highlight a statement like this more}
 As a result we can expect to see several thousand weak false-positive associations in a PC-corrected GWAS conducted on a human-sized genome in species with neighborhood sizes up to at least 1000.
 
-This does not mean that GWAS is not useful, but does put some limits on the extent of valid interpretation.
-Very few of the associations we identified would be significant at a conservative Bonferroni-adjusted $p$-value cutoff, suggesting that most of the very strong signals of association signals observed in studies of mono- or oligogenic traits are robust to stratification bias.
-Further, the most dramatic effects of stratification inflation we observed occurred at neighborhood sizes below 100 -- smaller than the vast majority of modern human populations (but see below for further discussion of empirical cases).
-However, as recently identified in studies of gentoype associations for human height in Europe \citep{Berg2018,Sohail2018}, PC regression GWAS in modern human populations does leave residual signal of population structure in large-scale GWAS of polygenic traits.
+This suggests some caveats for interpretation of GWAS results.
+Very few of the spurious associations we identified would be significant at a conservative Bonferroni-adjusted $p$-value cutoff, 
+suggesting that most of the very strong signals of association observed in studies of mono- or oligogenic traits are robust to these concerns.
+Further, the most dramatic $p$-value inflations occurred at neighborhood sizes below 100 
+-- smaller than most recent human populations (but see below for further discussion of empirical cases). 
+However, as recently identified in studies of gentoype associations for human height in Europe \citep{Berg2018,Sohail2018}, 
+PC regression GWAS in modern human populations does leave residual signal of population structure in large-scale GWAS of polygenic traits. 
+\plr{Is there any indication that mixed models do better about this? (Where's the ``for precisely this reason'' come from?)}
 Indeed, studies in strongly structured species like \textit{Arabadopsis} have long relied on more sophisticated mixed model approaches to correcting for population structure for precisely this reason \citep{Aranzana2005,Sasaki2015}.
 
-A second point that has received less attention in the literature is the issue of overcorrection in GWAS.
-If a truly causal allele segregates at different frequencies in different populations, then correcting for population structure in a regression analysis will result in an underestimate of effect sizes.
+A second point that has received less attention in the literature is the issue of overfitting in GWAS. 
+If a truly causal allele segregates at different frequencies in different populations, then correcting for population structure in a regression analysis will result in an underestimate of its effect size.
 Though our simulations had no causal alleles, we observed some evidence of this effect in the distribution of $p$-values across the genome (Figure \ref{fig:gwas}D): after PC regression many analyses resulted in $p$-values falling below their expected values from a uniform distribution.
-This result is consistent with a recent empirical study of heritability in human height and body mass index, which found that increasing the number of PC axes used as covariates caused the total proportion of variance explained by SNPs to decline from $\approx0.8$ to $\approx0.75$ \citep{Wainschtein2019}.
-Indeed SNPs with minor alleles frequencies of 0.001 - 0.01, which are expected to reflect fine scale population structure \citep{Mathieson2012,Novembre2009}, are estimated to explain \textit{negative} proportions of the total phenotypic variance in \citep{Wainschtein2019}\ak{i don't understand what the authors mean here -- how can we have a negative variance?}.
-Searching for genetic associations of polygenic traits that vary systematically across but not within populations through existing GWAS approaches is then unlikely to be successful: the signals are fully confounded, and new analytic methods or experiments controlling for variation in the environment will be necessary to rigorously identify causal variants.
+This is consistent with a recent empirical study of heritability in human height and body mass index, which found that increasing the number of PC axes used as covariates caused the total proportion of variance explained by SNPs to decline from $\approx0.8$ to $\approx0.75$ \citep{Wainschtein2019}.
+Indeed SNPs with minor alleles frequencies of 0.001 - 0.01, which are expected to reflect fine scale population structure \citep{Mathieson2012,Novembre2009}, are estimated to explain \textit{negative} proportions of the total phenotypic variance in \citep{Wainschtein2019}.
+\ak{i don't understand what the authors mean here -- how can we have a negative variance?}
+\plr{peter revisit this}
+% Searching for genetic associations of polygenic traits that vary systematically across but not within populations through existing GWAS approaches is then unlikely to be successful: the signals are fully confounded, and new analytic methods or experiments controlling for variation in the environment will be necessary to rigorously identify causal variants.
 
-In summary, spatial covariation in population structure and the environment confound the interpretation of GWAS $p$-values, and correction using principal components is insufficient to fully separate these signals for polygenic traits under a variety of environmental and population parameter regimes.
-How more sophisticated mixed-model methods would perform under our simulations is an interesting question that we plan to pursue in a future study, but statistical methods can only take us so far in the absence of controlled environments.
-One currently popular approach to estimating the degree of bias in GWAS caused by population structure is LD score regression \citep{Bulik-Sullivan2015}.
+In summary, spatial covariation in population structure and the environment confound the interpretation of GWAS $p$-values, 
+and correction using principal components is insufficient to fully separate these signals for polygenic traits under a variety of environmental and population parameter regimes. 
+Other GWAS methods may be less sensitive to this confounding, but there is no obvious reason that this should be so.
+One approach to estimating the degree of bias in GWAS caused by population structure is LD score regression \citep{Bulik-Sullivan2015}. 
 Though this approach appears to work well in practice, its interpretation is not always straightforward and it is likely biased by the presence of linked selection \citep{Berg2018}. 
-We suggest a straightforward alternative for species in which the primary axes of population differentiation is space (note this is likely not the case for many modern human populations): run a GWAS with spatial coordinates as phenotypes and check for $p$-value inflation or significant associations.
-If significant associations with sample locality are observed after correcting for population structure through PC regression or a kinship matrix, the structure corrections are insufficient.
-This is essentially the approach taken in our ``clinal'' model (though we also include normally distributed variation in our phenotypes).
-Of course it is possible that genotypes indirectly affect individual locations by adjusting organismal fitness and thus habitat selection across spatially varying environments, but we believe that this hypothesis should be tested against a null of stratification bias inflation rather than accepted as true based on GWAS results.
+We suggest a straightforward alternative for species in which the primary axes of population differentiation is space 
+(note this is likely not the case for many modern human populations): 
+run a GWAS with spatial coordinates as phenotypes and check for $p$-value inflation or significant associations. 
+If significant associations with sample locality are observed after correcting for population structure through PC regression or a kinship matrix, 
+the method is sensitive to false positives induced by spatial structure. 
+This is essentially the approach taken in our ``clinal'' model (though we add normally distributed noise to our phenotypes). 
+Of course, it is possible that genotypes indirectly affect individual locations by adjusting organismal fitness and thus habitat selection across spatially varying environments, but we believe that this hypothesis should be tested against a null of stratification bias inflation rather than accepted as true based on GWAS results.
 \cjb{what do you think of the second half here?}
+\plr{love it}
 
@@
 
+%%%%%%%%%%%%
 \subsection{Where are natural populations on this spectrum?}
-For how much of the tree of life do spatial patterns circumscribe genomic variation?
-In Table \ref{table:NStable} we gathered estimates of neighborhood size from a range of organisms to get an idea of how likely dispersal is to play an important role in patterns of variation.
-Though this sample is almost certainly biased towards small-neighborhood species (because few studies have quantified neighborhood size in species with very high dispersal or population density), we find that neighborhood sizes in the range we simulated are fairly common across a range of taxa.
+
+How strong is the spatial patterning of genetic variation in extant species?
+% For how much of the tree of life do spatial patterns circumscribe genomic variation? 
+To get an idea of this, we gathered estimates of neighborhood size from a range of organisms (Table~\ref{table:NStable}).
+Although this sample is almost certainly biased towards small-neighborhood species 
+(because few studies have quantified neighborhood size in species with very high dispersal or population density), 
+we find that neighborhood sizes in the range we simulated are fairly common across a range of taxa. 
 At the extreme low end of empirical neighborhood size estimates we see some flowering plants, large mammals, and colonial insects like ants. 
 Species such as this have neighborhood size estimates small enough that spatial processes are likely to strongly influence inference. 
-These include some human populations such as the Gainj- and Kalam-speaking people of Papua New Guinea, in which the estimated neighborhood sizes in \citep{Rousset1997} range from 40 to 410 depending on the method of estimation.
-Many more species occur in a middle range of neighborhood sizes between 100 and 1000 -- a range in which spatial processes play a minor role in our analyses under random spatial sampling but are important when sampling of individuals in space is clustered.
-Last, many species likely have neighborhood sizes much larger than we simulated, including modern humans in NE Europe \citep{Ringbauer2017}.
-For these species demographic inference and summary statistics are likely to reflect minimal bias from spatial effects as long as dispersal is truly continuous across the landscape.
-While that is so we caution that association studies in which the effects of population structure are confounded with spatial variation in the environment are still sensitive to dispersal even at these large neighborhood sizes.
+These include some human populations such as the Gainj- and Kalam-speaking people of Papua New Guinea, 
+for whom the estimated neighborhood sizes in \citet{Rousset1997} range from 40 to 410 depending on the method of estimation. 
+Many more species occur in a middle range of neighborhood sizes between 100 and 1000 
+-- a range in which spatial processes play a minor role in our analyses under random spatial sampling but are important when sampling of individuals in space is clustered. 
+Last, many species likely have neighborhood sizes much larger than we simulated, including modern humans in northeastern Europe \citep{Ringbauer2017}. 
+If these species' ranges and population densities have been constant for a long period of time (unlike humans),
+demographic inference methods that assume well-mixed populations may show minimal bias from spatial effects.
+% While that is so, we caution that association studies in which the effects of population structure are confounded with spatial variation in the environment are still sensitive to dispersal even at these large neighborhood sizes.
 
+%%%%%%%%%%%
 \subsection{Future Directions and Limitations}
+
 As we have shown, a large number of population genetic summary statistics contain information about spatial population processes.
 We imagine that combinations of such summaries might be sufficient for the construction of supervised machine learning regressors (e.g., \cite{Schrider2018}) for the accurate estimation of dispersal from genetic data.
-Indeed \cite{Ashander2018} found that inverse interpolation on a vector of summary statistics provided a powerful method of estimating dispersal distances.
+Indeed, \cite{Ashander2018} found that inverse interpolation on a vector of summary statistics provided a powerful method of estimating dispersal distances.
 Expanding this approach to include the haplotype-based summary statistics studied here and applying machine learning regressors built for general inference of nonlinear relationships from high-dimensional data may allow precise estimation of spatial parameters under a range of complex models.
@@
 
-Though our continuous space simulation allows incorporation of realistic demographic and spatial processes and is much faster than previous individual-based models, it is inevitably limited by the computational burden of tracking tens of thousands of individuals in every generation.
-In particular the calculations required for our mate selection and competition steps involve summarizing distances across all pairs of individuals and so scale very poorly \ak{how exactly does this scale?
-$O(N^2)$?} as the number of individuals within a three-$\sigma$ radius increases.
-In part the issue of runtime scaling is a function of the spatial process itself -- under very limited dispersal we observed that coalescence requires over 30N generations, so forward-time methods must be run for a very long time to create a complete genealogy as underlies all real genome sequence data.
-The reverse-time model of continuous space evolution described in \cite{Barton2010} and implemented in \cite{Kelleher2014} may allow exploration of parameter regimes with population and landscape sizes more directly comparable to empirical cases like humans.
-However, incorporating selection or other processes into such models will be difficult.
+Though our continuous space simulation allows incorporation of realistic demographic and spatial processes and is much faster than previous individual-based models, 
+it is inevitably limited by the computational burden of tracking tens or hundreds of thousands of individuals in every generation. 
+In particular, computations required for mate selection and spatial competition caled as the number of individuals within a three-$\sigma$ radius increases. 
+% In part the issue of runtime scaling is a function of the spatial process itself -- under very limited dispersal we observed that coalescence requires over 30N generations, so forward-time methods must be run for a very long time to create a complete genealogy as underlies all real genome sequence data.
+The reverse-time model of continuous space evolution described by \citet{Barton2010} and implemented by \citet{Kelleher2014} 
+allows exploration of parameter regimes with population and landscape sizes more directly comparable to empirical cases like humans. 
+Alternatively, implementation of parallelized simulations may allow progress with forwards-time simulations.
 
-A mixed approach may be possible by combining forward- and reverse-time models, as was recently done for a continuous-space Wright-Fisher model in \cite{Lotterhos2019} and for a simulation with linked selection in \cite{Buffalo2019}.
-This would allow us to generate short runs of complex, realistic simulations in forward time in SLiM \citep{Haller2019} and then ``finish'' the simulations as a coalescent simulation in msprime \citep{Kelleher2016}.
-However a significant difficulty in this ``recapitation'' approach is scaling the variance in reproductive output and generation time across forward- and reverse-time methods.
-Further development of our understanding of how to merge forward- and reverse-time models is a promising avenue for future research that will be necessary for scaling continuous-space simulations to millions or billions of individuals.
+A mixed approach can combine forward- and reverse-time models, 
+as recently done for a continuous-space Wright-Fisher model by \citet{Lotterhos2019} 
+and a simulation with linked selection by \citet{Buffalo2019}. 
+This allows us to generate short runs of large, realistic simulations in forward time in SLiM \citep{Haller2019} 
+and then ``finish'' the simulations as a coalescent simulation in msprime \citep{Kelleher2016},
+combining the two using an operation called ``recapitation'' \citep{haller2019treesequence}
+As we have seen in this work, a nontrivial step in this
+approach is scaling the variance in reproductive output and generation time across forward- and reverse-time methods. 
+Further development of our understanding of how to merge forward- and reverse-time models 
+is a promising avenue for future research that will be necessary for scaling continuous-space simulations to millions or billions of individuals.
 
-Finally we believe that the difficulties in correcting for population structure in continuous populations using principal components analysis or similar decompositions is a difficult issue, well worth considering on its own -- what does it mean to correct for population structure in a world without discrete demes?
-What are the boundaries between proper corrections between ancestry and underpowering the search for local genetic variation?
-We posit that there is progress yet to be made in deriving either process driven descriptions of ancestry or more generalized unsupervised methods that can better account for shared relatedness, say in the context of GWAS, for populations that are structured over space.
-\ak{this paragraph is a bit out there. rein me in}
+Finally, we believe that the difficulties in correcting for population structure in continuous populations 
+using principal components analysis or similar decompositions is a difficult issue, well worth considering on its own.
+%-- what does it mean to correct for population structure in a world without discrete demes? 
+How can we best avoid spurious correlations while correlating genetic and phenotypic variation 
+without underpowering the methods?
+Perhaps optimistically,
+% What are the boundaries between proper corrections between ancestry and underpowering the search for local genetic variation? 
+we posit that process-driven descriptions of ancestry 
+and/or more generalized unsupervised methods may be able to better account for carry out this task.
 
 \section{Data Availability}
-Scripts used for all analyses and figures are available at https://github.com/petrelharp/spaceness.
+Scripts used for all analyses and figures are available at \url{https://github.com/petrelharp/spaceness}.
 

### 2019-04-29 ms/spaceness.tex
peter edited through

@@
 
-\plr{note: conclusions are that Ne(t) is not very biased}
+\plr{note: what are conclusions about bias or not of Ne(t)?}
 
@@
 The inescapable reality that biological organisms live, move, and reproduce in continuous geography is usually omitted from population genetic models. 
-Indeed, a near universal rule of reproduction is that individuals mate with other nearby individuals, leading to a positive correlation between genetic and geographic distances. 
-This pattern of ``isolation by distance'' \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, 
-but classical mathematical models describing it \citep{Malecot1948} are flawed approximations of the underlying process \citep{Felsenstein1975,barton2002neutral}.
-Mathematical difficulties have made further development of population models in continuous space difficult,
-so most models describe geographic structure as a set of populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical}, 
+However, mates tend to live near to one another and to their offspring,
+leading to a positive correlation between genetic and geographic distances. 
+% Patterns of genetic variation are influenced by the fact that organisms are more likely, on average, to live nearby to both their parents, in particular when dispersal and mate choice distances are short.
+This pattern of ``isolation by distance'' \citep{Wright1943}, describing mean relatedness as a function of distance in flat landscapes,
+is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}.
+Despite a long history of analytical work describing the genetics of populations distributed across continuous geography
+(e.g., \cite{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation}),
+much modern work still describes geographic structure as a set of populations connected by migration 
+\citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations}.
+For this reason,
+most population genetics statistics are interpreted with reference to discrete, well-mixed populations,
 and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000},
@@
 
-\ak{I think we can cut the following paragraph if we just add the last sentence above} 
-\cjb{agree - commenting out}
-%Last we use our model to study the impacts of isolation by distance in continuous space on genome-wide association studies (GWAS).
-Variation in ancestry proportions between case and control cohorts is expected to cause inflation of test statistics at SNPs with different allele frequencies among populations \citep{Price2006}.
-Because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within continuous populations can also deflate $p$ values in GWAS of traits influenced by the environment \citep{Mathieson2012}.
-Here we simulate both the underlying genealogy and phenotypes of a continuously distributed population and seek to identify regions of parameter space -- i.e., the strength of isolation by distance and the spatial distribution of environmental effects on phenotypes -- in which common methods employed for GWAS are subject to confounding of environmental and genetic effects on phenotypes.
-
-\plr{Moved from disco: integrate citations.}
-Patterns of genetic variation are influenced by the fact that organisms are more likely, on average,
-to live nearby to both their parents, in particular when dispersal and mate choice distances are short. 
-Analytical work describing the genetics of populations distributed across continuous geography
-(e.g., \cite{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation})
-the theoretical work is challenging and often untenable for realistic biological models. 
-Also discuss lots of work on grids, e.g., \citet{Shirk2014,rousset2011likelihoodbased}.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\plr{PETER TO REVISIT THIS SECTION}
-
-The best-studied approaches to population genetics in continuous space were developed by \cite{Wright1942,Wright1943} and \cite{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals.
-A key finding of Wright's model is that many important aspects of continuous populations can be described in terms of ``neighborhood size'' -- the number of potential mates for an individual in a given generation, defined as $4\pi\sigma^2d$, where $\sigma$ is the average dispersal distance and $d$ is population density.
-\cite{Maruyama1972} found that the rate of decline in genetic diversity in a 2-dimensional continuous population approaches the random mating expectation when $d\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.
-
-Though some aspects of continuous populations are well described by the Wright and Malécot models, \cite{Felsenstein1975} showed that the assumptions of independent dispersal and Poisson distributed offspring that are the basis of these models are incompatible.
-Over time, a population meeting them will clump into a small number of geographic clusters occupying only a part of the available range.
-Although real populations are often clumped on landscapes due to factors like varying habitat quality and competition among species, the Wright and Malécot models produce much more extreme clumping than is observed in practice and fail to account for the density-dependent declines in population growth rate that are widely observed in real populations (CITE).
-
-One method for modeling continuous populations is then to assume the existence of a grid of discrete randomly-mating populations connected by migration, which prevents clustering by forcing all regions to be occupied in every generation.
-Among many other important results drawn from this class of ``lattice'' or ``stepping stone'' models, \cite{Rousset1997} showed that the slope of the a linear regression of genetic differentiation ($F_{ST}$) against the logarithm of spatial distance is an estimate of neighborhood size.
-Though good approximations of continuous structure given high dispersal, these models are not truly continuous, force a uniform realized population density across landscapes, and limit investigation of spatial structure below the level of the deme.
-An alternative method is to model the geographic spread of ancestry backwards in time through a diffusion approximation -- an approach that has recently made significant progress in modeling both dispersal and demographic parameters \citep{Barton2010,Kelleher2014,Ringbauer2017,Ringbauer2018}.
-
-We took a direct approach to the clustering problem of classical forward-time models by incorporating density dependence into an individual-based model similar to the analytic models developed by Wright and Malécot.
-By scaling the probability of survival in each timestep to local population density we shift reproductive output towards regions of low-density and prevent populations from clustering.
-A similar approach was taken previously by \citep{Doebeli2003} who used an individual based model with continuous space and density dependent fitness to study the probability of speciation along continuous environmental gradients.
-However to our knowledge previous implementations of continuous space models have focused on a small number of genetic loci as the unit of analysis, which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms.
-By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.
+% The best-studied approaches to population genetics in continuous space were developed by \cite{Wright1942,Wright1943} and \cite{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals.
+The degree to which genetic relationships are geographically correlated
+depends on the chance that two geographically nearby individuals are close relatives
+-- in modern terms, by the tension between migration (the chance that one is descended from a distant location)
+and coalescence (the chance that they share a parent).
+A key early observation Wright model is that this balance is nicely summarized by the ``neighborhood size'',
+defined to be $N_W = 4 \pi \rho \sigma^2$, where $\sigma$ is the mean parent--offspring distance and $\rho$ is population density. 
+This can be thought of as proportional to the average number of potential mates for an individual (those within distance $2 \sigma)$,
+or the number of potential parents of a randomly chosen individual.
+\cite{Maruyama1972} found that the rate of decline in genetic diversity in a two-dimensional continuous population 
+approaches the random mating expectation when $\rho\sigma^2 > 1$, 
+and proposed that this had the important implication that most population genetic expectations for randomly mating populations 
+could be applied to continuously distributed populations with relatively little error.
+
+The first approach to modeling continuously-distributed populations
+was to endow individuals in a Wright-Fisher model with locations in continuous space.
+However, since the total size of the population is constrained, this introduces interactions between arbitrarily distant individuals,
+which (aside from being implausible) \cite{Felsenstein1975} showed would eventually lead to unrealistic population clumping
+if the species range is sufficiently large.
+After global population regulation,
+the next method for modeling spatial populations might to assume the existence of a grid of discrete randomly-mating populations connected by migration, 
+thus enforcing regular population density by edict.
+Among many other important results drawn from this class of ``lattice'' or ``stepping stone'' models, 
+\cite{Rousset1997} showed that the slope of the a linear regression of genetic differentiation ($F_{ST}$) against the logarithm of spatial distance is an estimate of neighborhood size. 
+Although these grid models are likely good approximations of continuous geography in many situations,
+these do not model demographic fluctuations, and limit investigation of spatial structure below the level of the deme,
+assumptions whose impacts are unknown. 
+An alternative method for dealing with continuous geography is a new class of coalescent models,
+the Spatial Lambda Fleming-Viot models \citep{Barton2010,Kelleher2014}.
+
+To avoid questionable assumptions,
+we worked with forwards-time, individual-based simulations.
+By scaling the probability of survival in each timestep to local population density, 
+we shift reproductive output towards regions of low-density, which prevents populations from clustering. 
+% LOTS of people have done models like this: classic ecology?
+% A similar approach was taken previously by \citet{Doebeli2003} who used an individual based model with continuous space and density dependent fitness to study the probability of speciation along continuous environmental gradients.
+Such models have been used extensively in ecological modeling but rarely in population genetics,
+where to our knowledge previous implementations of continuous space models have focused on a small number of genetic loci, 
+which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. 
+By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations 
+and replicate the sampling designs and analyses commonly conducted on real genomic data.
 
@@
 
-\plr{peter to revisit this paragraph}
+\plr{revisit this paragraph}
 Trends in pairwise haplotype block sharing parallel those in allele-frequency-based diversity estimates (Figure \ref{fig:sumstats}, Supplementary Figure \ref{fig:allsumstats}).
@@
 
-\plr{peter to revisit this paragraph}
+\plr{revisit this paragraph}
 We next turn our attention to the effect of geographic distance on haplotype block length sharing. 
@@
 If a truly causal allele segregates at different frequencies in different populations, then correcting for population structure in a regression analysis will result in an underestimate of its effect size.
-Though our simulations had no causal alleles, we observed some evidence of this effect in the distribution of $p$-values across the genome (Figure \ref{fig:gwas}D): after PC regression many analyses resulted in $p$-values falling below their expected values from a uniform distribution.
-This is consistent with a recent empirical study of heritability in human height and body mass index, which found that increasing the number of PC axes used as covariates caused the total proportion of variance explained by SNPs to decline from $\approx0.8$ to $\approx0.75$ \citep{Wainschtein2019}.
-Indeed SNPs with minor alleles frequencies of 0.001 - 0.01, which are expected to reflect fine scale population structure \citep{Mathieson2012,Novembre2009}, are estimated to explain \textit{negative} proportions of the total phenotypic variance in \citep{Wainschtein2019}.
-\ak{i don't understand what the authors mean here -- how can we have a negative variance?}
-\plr{peter revisit this}
+Though our simulations had no causal alleles, we observed some evidence of this effect in the distribution of $p$-values across the genome (Figure \ref{fig:gwas}D): 
+after PC regression many analyses resulted in $p$-values falling below their expected values from a uniform distribution. 
+This is consistent with a recent empirical study of heritability in human height and body mass index, 
+which found that increasing the number of PC axes used as covariates caused the total proportion of variance explained by SNPs to decline from $\approx0.8$ to $\approx0.75$ \citep{Wainschtein2019}. 
+% Indeed SNPs with minor alleles frequencies of 0.001 - 0.01, which are expected to reflect fine scale population structure \citep{Mathieson2012,Novembre2009}, 
+% are estimated to explain \textit{negative} proportions of the total phenotypic variance in \citep{Wainschtein2019}.
+% \ak{i don't understand what the authors mean here -- how can we have a negative variance?}
 % Searching for genetic associations of polygenic traits that vary systematically across but not within populations through existing GWAS approaches is then unlikely to be successful: the signals are fully confounded, and new analytic methods or experiments controlling for variation in the environment will be necessary to rigorously identify causal variants.

### 2019-05-09 ms/spaceness.tex
andy & cj overleaf edits

@@
 \author[$\ast$,1]{C.J. Battey}
-\author[$\ast$]{Peter L. Ralph}
-\author[$\ast$]{Andrew D. Kern}
+\author[$\ast, \dagger$]{Peter L. Ralph}
+\author[$\ast, \dagger$]{Andrew D. Kern}
 
@@
     but standard models in population genetics are based on discrete, well-mixed populations. 
-    As a result many methods of analyzing genetic data assume that samples are a random draw from a well-mixed population, 
-    but are applied to real populations that have significant geographic structure.
+    As a result many methods of analyzing genetic data assume that samples are a random draw from a well-mixed population, but are applied to clustered samples from populations that are structured clinally over space.
     Here we use simulations of populations living in continuous geography 
@@
     especially when Wright's neighborhood size is less than 100 and sampling is spatially clustered.
-    Standard estimates of population size history based on the site frequency spectrum become biased with lower dispersal. 
+    The combination of low dispersal and clustered sampling causes demographic inference from the site frequency spectrum to infer more turbulent demographic histories, but averaged results across multiple simulations were surprisingly robust to isolation by distance.
     We also show that the combination of spatially autocorrelated environments and limited dispersal
@@
 \correspondingauthoraffiliation{1}{301 Pacific Hall, University of Oregon Dept. Biology, Institute for Ecology and Evolution. cbattey2@uoregon.edu.}
+\correspondingauthoraffiliation{$\dagger$}{these authors co-supervised this project}
 \vspace{-35pt}% Only used for adjusting extra space in the left column of the first page
-
-\plr{note: address confounding of $\sigma$ and neighborhood size in disco}
-
-\plr{note: what are conclusions about bias or not of Ne(t)?}
 
@@
 The inescapable reality that biological organisms live, move, and reproduce in continuous geography is usually omitted from population genetic models.
-However, mates tend to live near to one another and to their offspring,
-leading to a positive correlation between genetic and geographic distances. 
-% Patterns of genetic variation are influenced by the fact that organisms are more likely, on average, to live nearby to both their parents, in particular when dispersal and mate choice distances are short.
-This pattern of ``isolation by distance'' \citep{Wright1943}, describing mean relatedness as a function of distance in flat landscapes,
-is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}.
-Despite a long history of analytical work describing the genetics of populations distributed across continuous geography
-(e.g., \cite{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation}),
-much modern work still describes geographic structure as a set of populations connected by migration 
-\citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations}.
-For this reason,
-most population genetics statistics are interpreted with reference to discrete, well-mixed populations,
-and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000},
-effectively assuming these are randomly mating units.
+However, mates tend to live near to one another and to their offspring, leading to a positive correlation between genetic differentiation and geographic distance.
+This pattern of ``isolation by distance'' \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}.
+Despite a long history of analytical work describing the genetics of populations distributed across continuous geography (e.g., \cite{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation}), much modern work still describes geographic structure as a set of populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations}.
+For this reason, most population genetics statistics are interpreted with reference to discrete, well-mixed populations, and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000} with methods that assume these are randomly mating units.
 
@@
 that is much larger than the sample \citep{wakeley2003gene}.
-The key assumption is actually that the individuals of each generation are \emph{exchangeable},
-so that there is no correlation between the fate or fecundity of a parent and that of their offspring \citep{Huillet2011}.
+The key assumption is that the individuals of each generation are \emph{exchangeable}, so that there is no correlation between the fate or fecundity of a parent and that of their offspring \citep{Huillet2011}.
 If dispersal or mate selection is limited by geographic proximity, this assumption can be violated in many ways.
-For instance, if mean viability or fecundity is spatially autocorrelated,
-then limited geographic dispersal will lead to parent--offspring correlations.
-Futhermore, nearby individuals will be more closely related than an average random pair,
-so drawing multiple samples from the same area of on the landscape will represent a biased sample of the genetic variation present in the whole population. 
-It has long been appreciated that this model misspecification subjects downstream inferences to bias, 
-but the extent and nature of these effects remain largely uninvestigated.
-
-For instance, nonzero values of Tajima's $D$ are often interpreted as reflecting evidence of selection or past population size changes \citep{tajima1989statistical}.
-This statistic is a summary of the site frequency spectrum, which itself reflects variation in branch lengths and tree structure of the underlying genealogies of sampled individuals.
-Geographically limited mate choice distorts the distribution of these genealogies \citep{Maruyama1972},
-which can affect the value of Tajima's $D$.
-% Because the spatial extent of dispersal and mate selection determines the rate of drift in a continuously distributed population 
-% these population genetic summaries should also reflect variation in the scale of spatial interactions both over time and across populations. 
-Similarly, the distribution of tract lengths of identity by state among individuals contains information about not only
-historical demography \citep{Harris2013,ralph2013geography} and selection \citep{Garud2015}, but also dispersal and mate choice \citep{Ringbauer2017,Baharian2016}. 
-We are particularly keen to examine how such summaries will be affected by models that incorporate continuous space,
-both to evaluate the assumptions underlying existing methods, but also to identify where the most promising signals of geography lie.
-
-A related issue is the spatial distribution of sampling effort --
-since nearby individuals are more likely to be closely related than distant ones,
-observed patterns of relatedness are expected to depend on the geographic sampling scheme,
-perhaps strongly.
-In addition, range edges have been observed to create complex patterns of heterozygosity in stepping-stone simulations \citep{Neel2013,Shirk2014}, 
-but the effects of this process on many downstream inference procedures is unknown.
-\cjb{must be more out there on sampling -- any recs for papers to read+cite?}
-
-The issue of fine-scale geographic structure may have particularly important implications for genome-wide association studies (GWAS). 
+For instance, if mean viability or fecundity is spatially autocorrelated, then limited geographic dispersal will lead to parent--offspring correlations.
+Furthermore, nearby individuals will be more closely related than an average random pair, so drawing multiple samples from the same area of on the landscape will represent a biased sample of the genetic variation present in the whole population.
+It has long been appreciated that this model misspecification subjects downstream inferences to bias, but the extent and nature of these effects remain largely uninvestigated.
+
+Two areas in which spatial structure may be particularly important are demographic inference and genome-wide association studies (GWAS).
+Effective population size is a fundamental parameter of population genetics and shapes expectations of the efficacy of selection, the rate of drift, among other evolutionary forces (\textit{c.f.} \cite{gillespie2001population}).
+Inferring a trajectory of population sizes through time has therefore become a major goal in population genetics, with many recent methods proposed to accomplish this task \citep{Liu2015,Terhorst2016,Schiffels2014,Sheehan2013}.
+All of these methods assume that the populations being modeled are approximately randomly mating (though this also depends on how researchers interpret the inferred parameter $\Theta=4N\mu$).
+Because isolation by distance is expected to change the rate of drift in a population \citep{Maruyama1972}, it may cause biases in inferred population sizes.
+However the effects of continuous spatial structure on demographic inference has not to our knowledge been previously investigated.
+
+Spatial structure is also a major challenge for interpreting the results of genome-wide association studies (GWAS).
 This is because many phenotypes of interest have strong geographic differences due to the (nongenetic) influence of environmental or socioeconomic factors,
 which can therefore show spurious correlations with spatially patterned allele frequencies \citep{Bulik-Sullivan2015,Mathieson2012}.
-This may be particularly important for the study of selection on polygenic traits,
-whose heritable genetic components are determined by many loci of weak effect,
-those hardest to disentangle from spurious correlations. 
 Indeed, two recent studies found that previous evidence of polygenic selection on human height in Europe was confounded by subtle population structure \citep{Sohail2018,Berg2018}, suggesting that existing methods to correct for population structure in GWAS are insufficient.
 However we have little quantitative idea of the population and environmental parameters that can be expected to lead to biases in GWAS.
-As the scale of sequence data now available for many species allows inference of increasingly fine-scale patterns of selection and demography, understanding how and when subtle spatial structure is likely to bias results is an important task for population genetics.
+
+Last, some of the most basic tools of population genetics are summary statistics like $F_{is}$ and Tajima's $D$, which are often interpreted as reflecting the influence of selection or demography on sampled populations \citep{tajima1989statistical}.
+Statistics like Tajima's $D$ are essentially summaries of the site frequency spectrum, which itself reflects variation in branch lengths and tree structure of the underlying genealogies of sampled individuals.
+Geographically limited mate choice distorts the distribution of these genealogies \citep{Maruyama1972}, which can affect the value of Tajima's $D$.
+Similarly, the distribution of tract lengths of identity by state among individuals contains information about not only historical demography \citep{Harris2013,ralph2013geography} and selection \citep{Garud2015}, but also dispersal and mate choice \citep{Ringbauer2017,Baharian2016}.
+We are particularly keen to examine how such summaries will be affected by models that incorporate continuous space, both to evaluate the assumptions underlying existing methods and to identify where the most promising signals of geography lie.
 
@@
 that incorporates overlapping generations, local dispersal of offspring, and density-dependent survival. 
-We simluate chromosome-scale genomic data in tens of thousands of individuals, and output the full genealogy and recombination history of all final-generation individuals. 
+We simulate chromosome-scale genomic data in tens of thousands of individuals from parameter regimes relevant to common subjects of population genetic investigation such as humans and Drosophila, and output the full genealogy and recombination history of all final-generation individuals.
 We use these simulations to test how sampling strategy interacts with geographic population structure to cause systematic variation in population genetic summary statistics 
@@
 We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum. 
-Lastly, we examine the impacts of evolution in continuous space on genome-wide association studies (GWAS) and identify regions of parameter space under which the results from GWAS may be misleading.
+Last, we examine the impacts of evolution in continuous space on genome-wide association studies (GWAS) and identify regions of parameter space under which the results from GWAS may be misleading.
 
@@
 The degree to which genetic relationships are geographically correlated
-depends on the chance that two geographically nearby individuals are close relatives
--- in modern terms, by the tension between migration (the chance that one is descended from a distant location)
+is determined by the interaction between rates of migration (the chance that one is descended from a distant location)
 and coalescence (the chance that they share a parent).
@@
 or the number of potential parents of a randomly chosen individual.
+Empirical estimates of neighborhood size vary hugely across species (see Discussion), but for reference estimates in human populations range from 40 to over 5,000 depending on the population and method of estimation (\ref{table:NStable}), and \cite{Wright1946} estimated values of 500-1000 for \textit{Drosophila pseudoobscura}.
 \cite{Maruyama1972} found that the rate of decline in genetic diversity in a two-dimensional continuous population 
@@
 which (aside from being implausible) \cite{Felsenstein1975} showed would eventually lead to unrealistic population clumping
-if the species range is sufficiently large.
-After global population regulation,
-the next method for modeling spatial populations might to assume the existence of a grid of discrete randomly-mating populations connected by migration, 
+if the range is sufficiently large.
+Another method for modeling spatial populations is to assume the existence of a grid of discrete randomly-mating populations connected by migration, 
 thus enforcing regular population density by edict.
@@
 Although these grid models are likely good approximations of continuous geography in many situations,
-these do not model demographic fluctuations, and limit investigation of spatial structure below the level of the deme,
+they do not model demographic fluctuations, and limit investigation of spatial structure below the level of the deme,
 assumptions whose impacts are unknown. 
@@
 
-To avoid questionable assumptions,
-we worked with forwards-time, individual-based simulations.
-By scaling the probability of survival in each timestep to local population density, 
-we shift reproductive output towards regions of low-density, which prevents populations from clustering. 
-% LOTS of people have done models like this: classic ecology?
-% A similar approach was taken previously by \citet{Doebeli2003} who used an individual based model with continuous space and density dependent fitness to study the probability of speciation along continuous environmental gradients.
-Such models have been used extensively in ecological modeling but rarely in population genetics,
-where to our knowledge previous implementations of continuous space models have focused on a small number of genetic loci, 
-which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. 
-By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations 
-and replicate the sampling designs and analyses commonly conducted on real genomic data.
+To avoid questionable assumptions, we here used forwards-time, individual-based simulations.
+By scaling the probability of survival in each timestep to local population density we shift reproductive output towards low-density regions, which prevents populations from clustering.
+Such models have been used extensively in ecological modeling but rarely in population genetics, where to our knowledge previous implementations of continuous space models have focused on a small number of genetic loci, which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms.
+By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.
 
@@
 As in previous work \citep{Wright1943,Ringbauer2017}, $\sigma$ as applied in our dispersal step is equal to the mean parent-offspring distance.
-A key parameter we report below is
-\citeauthor{Wright1943} citeyearpar{Wright1943} ``neighborhood size'', 
-defined to be $N_W = 4\pi\sigma^2 \rho$ where $\rho$ is the population density.
-This the approximate number of individuals available for mating in our simulation.
 
 At the beginning of the simulation individuals are distributed uniformly at random on a continuous, square landscape. 
-Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$ 
-who disperse a random, Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
+Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$ where $L$ is the expected lifespan.
+who disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
 in both the $x$ and $y$ coordinates, reflected to stay within the species range.
@@
 average lifetime $L=4$, 
-genome size $10^{8}$, 
-recombination rate $10^{-9}$, 
+genome size = $10^{8}$, 
+recombination rate = $10^{-9}$, 
 and drawing $\sigma$ values from a uniform distribution between 0.2 and 4. 
@@
 (We verified that this procedure produced the correct number of mutations by comparing to a subset of simulations with SLiM-generated mutations, which are applied only at meiosis.)
-Simulations were run for 1.6 million timesteps (approximately $30N$ generations), or until all extant individuals shared a common ancestor within the simulation (i.e., the tree sequence had coalesced).
-\ak{maybe worth including a table with some basic runtime results in the supplement?}
+Simulations were run for 1.6 million timesteps (approximately $30N$ generations).
 
@@
 
-Our demographic model includes paramters for population density ($K$),
+Our demographic model includes parameters for population density ($K$),
 mean life span ($L$), and dispersal distance ($\sigma$).
@@
 %\url{https://github.com/petrelharp/spaceness/blob/master/scripts/slimtools.py}. 
-We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $1e6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.
+We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $10^6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.
 
@@
 Because more closely related individuals on average share longer haplotype blocks we expect that spatial distance will be negatively correlated with mean haplotype block length, and that this correlation will be strongest (i.e., most negative) when dispersal is low.
-The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks so should reflect recent demographic events \citep{chapman2002effect}.
+The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks, and so should reflect recent demographic events \citep{chapman2002effect}.
+For a subset of simulations, we also calculated cumulative distributions for IBS tract lengths across pairs of distant (>48 map units) and nearby (<2 map units) individuals.
+Last, we examined the relationship between allele frequency and the spatial dispersion of an allele by calculating the average distance among individuals carrying each derived allele in a set of simulations representing a range of neighborhood sizes.
 
@@
 \subsection{Demographic Modeling}
-
-We fit single-population demographic models to the site frequency spectra of 20 individuals from each spatial SLiM simulation with the program Stairwayplot \citep{Liu2015}.
-This analysis was replicated across random, point, and midpoint sampling strategies.
-Site frequency spectra used for input data were calculated in scikit-allel \citep{Miles2017},
-and 100 bootstrap replicates were generated for each simulation by resampling over sites. 
-\plr{what were bootstrap replicates used for? if anything, need to say more precisely what you mean here}
-Models were fit across all bootstrap replicates using default settings in Stairwayplot and the median estimate of $N_e$ per generation was used to represent the output of each simulation.
-
-In preliminary runs we found that inferred population histories were highly variable even when simulating under a coalescent model, suggesting that some of the differences in demographic estimates for spatial models are caused by the behavior of the optimization algorithm rather than bias in the SFS caused by spatial mate choice and dispersal.
+To assess the impacts of continuous spatial structure on demographic inference we inferred population size histories for all simulations using two approaches: Stairwayplot \citep{Liu2015} and SMC++ \citep{Terhorst2016}.
+Stairwayplot fits its model to a genome-wide estimate of the SFS, while SMC++ also incorporates information about linkage disequilibrium across the genome in a sequentially markovian coalescent framework.
+For both methods we sampled 20 individuals from all spatial simulations using random, midpoint, and point sampling strategies.
+
+Stairwayplot documentation suggests fitting models across multiple bootstrap replicates drawn from empirical genomic data and taking the median inferred $N_{e}$ per unit time as the best estimate.
+We calculated site frequency spectra in scikit-allel \citep{Miles2017}, generated 100 bootstrap replicates per simulation by resampling over sites, and fit models for all bootstrap samples using default settings in stairwayplot.
+
+For SMC++ we first converted simulated genotype matrices to VCFs with msprime and then used SMC++'s standard pipeline for preparing input files assuming no polarization error in the SFS.
+We used the first individual in the VCF as the "designated individual" when fitting models, and allowed the program to estimate the recombination rate during optimiziation.
+We fit models using the 'estimate' command rather than the now recommended cross-validation approach because our simulations had only a single contig.
+
+To evaluate the performance of these methods we binned simulations by neighborhood size, took a rolling median of inferred $N_{e}$ trajectories across all model fits in a bin for each method and sampling strategy, and compared these values to the mean $N_{e}$ of random-mating simulations estimated from $\Theta_{\pi}$.
+We also examined how varying levels of isolation by distance impacted the variance of $N_{e}$ estimates by calculating the standard deviation of $N_{e}$ from each best-fit model and plotting these against neighborhood size.
+
+%In preliminary runs we found that inferred population histories were highly variable even when simulating under a coalescent model, suggesting that some of the differences in demographic estimates for spatial models are caused by the behavior of the optimization algorithm rather than bias in the SFS caused by spatial mate choice and dispersal.
 To separate these effects we ran 100 coalescent simulations with constant population size $6.1\times 10^{-3}$ (the mean $N_{e}$ of random-mating SLiM models estimated from $\Theta_{\pi}$) and fit stairwayplot models using the same script as for our spatial models.
@@
 SNPs with a minor allele frequency less than 0.5\% were excluded from this analysis.
-Phenotype values were set to vary by two standard deviations across the landscape in a rough approximation of the variation seen in height across Europe, which has recently been found to be confounded with population structure in large scale GWAS \citep{Berg2018,Sohail2018}.
-Conceptually our approach is similar to that taken in \citep{Mathieson2012}, though here we model fully continuous spatial variation and compare GWAS output across a range of dispersal distances.
+Phenotype values were set to vary by two standard deviations across the landscape in a rough approximation of the variation seen in height across Europe \ak{need citation}.
+Conceptually our approach is similar to that taken in \cite{Mathieson2012}, though here we model fully continuous spatial variation and compare GWAS output across a range of dispersal distances.
 
@@
 to a mean that may depend on spatial position.
-We adjust the geographic pattern of mean phenotype to create spatially autocorrelated environmental influences on phenotype.
-In the first simulation of \emph{nonspatial} phenotypes, the mean did not change,
+We adjust the geographic pattern of mean phenotype to reflect spatially autocorrelated environmental influences on phenotype.
+In the first simulation of \emph{nonspatial} environments, the mean did not change,
 so that all individuals' phenotypes were drawn independently from a Gaussian distribution with mean 110 and standard deviation 10.
 Next, to simulate \emph{clinal} environmental influences on phenotype, 
-we increased the mean phenotype form 90 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
-Concretely, an individual at position $(x,y)$ in a $50 \times 50$ landscape has mean phenotype $110 + 2x / 5$.
+we increased the mean phenotype from 100 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
+Concretely, an individual at position $(x,y)$ in a $50 \times 50$ landscape has mean phenotype $100 + 2x / 5$.
 Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype 
-for individuals with both $x$ and $y$ coordinates below 20 to 130 (two standard deviations above the rest of the map). 
+for individuals with both $x$ and $y$ coordinates below 20 to 120 (two standard deviations above the rest of the map). 
 Finally, in ``\emph{patchy}'' simulations we selected 10 random points on the map
 and set the mean phenotype of all individuals within three map units of each of these points
-to 130.
+to 120.
 
@@
 SNPs were first filtered to remove strongly linked sites by calculating LD between all pairs of SNPs in a 200-SNP moving window and dropping one of each pair of sites with an $R^2$ over 0.1.
-The LD-pruned allele count matrix was then centered and all sites scaled to unit variance when conducting the PCA, following recommendations in \citep{Patterson2006}.
+The LD-pruned allele count matrix was then centered and all sites scaled to unit variance when conducting the PCA, following recommendations in \cite{Patterson2006}.
 
@@
 with $p$-value below $0.05$ after adjusting for an expected false positive rate of less than 5\% \citep{benjamini2001control}. 
-% using the R function p.adjust(...,method=``fdr''). 
 We also examined $p$ values for systemic inflation by estimating the expected values from a uniform distribution (because no SNPs were used when generating phenotypes), plotting observed against expected values for all simulations, and summarizing across simulations by finding the mean $\sigma$ value in each region of quantile-quantile space.
@@
 
-Even though certain aspects of population demography depend on the scale of spatial interactions,
-it still could be that population genetic variation is well-described by a well-mixed population model.
-Indeed, mathematical results suggest that genetic variation in some spatial models should be well-approximated by a Wright-Fisher population
-if neighborhood size is large and all samples are geographically widely separated \citep{wilkins2004separation,zahle2005stepping}.
-% One of the main goals of this study is to examine the effect of continuous space on canonical summaries of population genetic variation. 
+Even though certain aspects of population demography depend on the scale of spatial interactions, it still could be that population genetic variation is well-described by a well-mixed population model.
+Indeed, mathematical results suggest that genetic variation in some spatial models should be well-approximated by a Wright-Fisher population if neighborhood size is large and all samples are geographically widely separated \citep{wilkins2004separation,zahle2005stepping}.
 However, the behavior of most common population genetic summary statistics has not yet been described in realistic geographic models.
@@
 
-Figure \ref{fig:sumstats} shows the effect of varying neighborhood size and sampling strategy on the site frequency spectrum (Figure~\ref{fig:sumstats}A) 
-and several standard population genetic summary statistics (Figure~\ref{fig:sumstats}B). 
-These show a significant enrichment of intermediate frequency variants in comparison to the nonspatial expectation,
-for smaller neighborhood sizes ($\leq 100$) that is exacerbated by midpoint and point sampling of individuals (depicted in Figure~\ref{fig:samplemap}). 
+Figure \ref{fig:sumstats} shows the effect of varying neighborhood size and sampling strategy on the site frequency spectrum (Figure~\ref{fig:sumstats}A) and several standard population genetic summary statistics (Figure~\ref{fig:sumstats}B).
+SFS comparisons show a significant enrichment of intermediate frequency variants in comparison to the nonspatial expectation, for smaller neighborhood sizes ($\leq 100$) that is exacerbated by midpoint and point sampling of individuals (depicted in Figure~\ref{fig:samplemap}).
 Reflecting this, Tajima's $D$ is quite positive in the same situations (Figure~\ref{fig:sumstats}B).
-Notably, the point at which Tajima's $D$ approaches 0 differs strongly across sampling strategies 
--- varying from a neighborhood size of roughly 50 for random sampling to at least 1000 for midpoint sampling.
-
-One of the most commonly used summaries of variation is Tajima's summary of nucleotide divergence, $\widehat{\theta}_{\pi}$,
-\plr{In the figure this has no hat on; should we remove these in the text? The notation is not standard.}
-calculated as the mean density of nucleotide differences, averaged across pairs of samples.
-As can be seen in Figure \ref{fig:sumstats}B, $\widehat{\theta}_{\pi}$ differs as much as nearly three-fold 
-between the random mating and spatial models.
-$\widehat{\theta}_{\pi}$ using each sampling strategy approaches the random mating expectation at its own rate, 
-but by a neighborhood size of around 100 all models are equivalent. 
-\plr{To interpret this we need to compare to census size over variance in offspring number.}
-The differences between spatial and random mating simulations are much greater than expected 
-from differences in census size divided by variance in offspring number.
-This likely occurs because $\widehat{\theta}_{\pi}$ is a measure of mean time to most recent common ancestor between two samples,
-and at small values of $\sigma$, the time for dispersal to mix ancestry across the range exceeds the mean nonspatial coalescent time.
-(For instance, at the smallest value of $\sigma=0.2$, the range is 250 dispersal distances wide,
-and since the location of a diffusively moving lineage after $k$ generations has variance $k \sigma^2$, 
-it takes around $250^2 = 62500$ generations to mix across the range,
-which is roughly ten times larger than the random mating effective population size.)
-Interestingly, the effect of sampling strategy is reversed relative to that observed in Tajima's D 
--- midpoint sampling reaches random mating expectations around neighborhood size 50, 
-while random sampling is inflated until around neighborhood size 100. 
-% In effect, this means that any estimates of population size based on expected heterozygosity from populations with limited dispersal should be treated with caution.
+Notably, the point at which Tajima's $D$ approaches 0 differs strongly across sampling strategies -- varying from a neighborhood size of roughly 50 for random sampling to at least 1000 for midpoint sampling.
+
+One of the most commonly used summaries of variation is Tajima's summary of nucleotide divergence, $\theta_{\pi}$, calculated as the mean density of nucleotide differences, averaged across pairs of samples.
+As can be seen in Figure \ref{fig:sumstats}B, $\theta_{\pi}$ differs as much as nearly three-fold between the random mating and spatial models.
+$\theta_{\pi}$ using each sampling strategy approaches the random mating expectation at its own rate, but by a neighborhood size of around 100 all models are equivalent.
+The differences between spatial and random mating simulations are much greater than expected from differences in census size divided by variance in offspring number.
+This likely occurs because $\theta_{\pi}$ is a measure of mean time to most recent common ancestor between two samples, and at small values of $\sigma$, the time for dispersal to mix ancestry across the range exceeds the mean nonspatial coalescent time.
+(For instance, at the smallest value of $\sigma=0.2$, the range is 250 dispersal distances wide, and since the location of a diffusively moving lineage after $k$ generations has variance $k \sigma^2$, it takes around $250^2 = 62500$ generations to mix across the range, which is roughly ten times larger than the random mating effective population size.) Interestingly, the effect of sampling strategy is reversed relative to that observed in Tajima's D -- midpoint sampling reaches random mating expectations around neighborhood size 50, while random sampling is inflated until around neighborhood size 100.
 
@@
 $F_{IS}$ is inflated above the expectation across most of the parameter space examined and across all sampling strategies.
-This effect is caused by a deficit of heterozygous individuals in low-dispersal simulations 
--- a continuous-space version of the Wahlund effect \citep{Wahlund1928}. 
+This effect is caused by a deficit of heterozygous individuals in low-dispersal simulations -- a continuous-space version of the Wahlund effect \citep{Wahlund1928}.
 Indeed, for random sampling under the spatial model, $F_{IS}$ does not approach the random mating equivalent until neighborhood sizes of nearly 1000.
 On the other hand, the dependency of raw observed heterozygosity on neighborhood size is not monotone.
-Under midpoint sampling observed heterozygosity is inflated even over the random mating expectation, 
-as a result of the a higher proportion of heterozygotes occurring in the middle of the landscape (Figure~\ref{fig:hetmap}). 
+Under midpoint sampling observed heterozygosity is inflated even over the random mating expectation, as a result of the a higher proportion of heterozygotes occurring in the middle of the landscape (Figure~\ref{fig:hetmap}).
 This squares with a report from \cite{Shirk2014} who observed a similar excess of heterozygosity in the middle of the landscape when simulating under a lattice model.
 
-\plr{revisit this paragraph}
-Trends in pairwise haplotype block sharing parallel those in allele-frequency-based diversity estimates (Figure \ref{fig:sumstats}, Supplementary Figure \ref{fig:allsumstats}).
-At low dispersal the distribution of IBS block lengths in a set of samples is shifted towards smaller values with respect to the random mating expectation -- resulting in lower means and fewer long IBS blocks.
-The variance and skew of the distribution of haplotype block lengths are only minorly affected by neighborhood size in our simulations when calculated across all pairs of individuals; however, they are strongly dependent on sampling regimes.
-For example, the number of long haplotype blocks declines as neighborhood size increases under midpoint sampling but changes very little across neighborhood sizes under point or random sampling.
-Thus sampling strategies with respect to geography will affect conclusions drawn from haplotype length distributions quite dramatically.
-
-\subsubsection{Correlations of summary statistics with geographic distance}
-Correlating population genetic summaries such as $F_{ST}$ against geographic distance has shown great utility in empirical population genetics \citep{Rousset1997}.
-As we know the exact locations of individuals that are sampled in our simulated populations we have examined the relationship of geographic distance between samples with a number of summary statistics (Figures \ref{fig:sumstats} and \ref{fig:allsumstats}).
-Mean density of nucleotide differences between individuals, $D_{xy}$, is positively correlated with the geographic distance between the individuals,
-and the strength of this correlation declines as dispersal increases, as expected under theory \citep{Wright1943,Rousset1997}. 
-This relationship varies across sampling strategies, with the weakest correlations observed for midpoint sampling,
-perhaps due to a dearth of long-distance comparisons.
-% Moreover there is clearly strong signal for neighborhood size in the strength of the correlation between $D_{xy}$ and geographic distance.
-
-\plr{revisit this paragraph}
-We next turn our attention to the effect of geographic distance on haplotype block length sharing. 
-As in \citet{Ringbauer2017} and \citet{Baharian2016} we found that the pairwise distribution of haplotype block lengths is more strongly left-skewed under limited dispersal.
-This is reflected in negative correlation coefficients between spatial distance and the mean, variance, skew, and count of long blocks from the pairwise distribution of identical-by-state block lengths (Figure \ref{fig:sumstats} and Figure \ref{fig:allsumstats}).
-Of these summaries the mean of the IBS tract length distribution is only weakly affected by neighborhood size, likely because it is heavily influenced by the small number of very long IBS tracts.
-In contrast the count of long IBS blocks and the skew of the pairwise IBS block distribution are strongly dependent on distance among individuals, and the magnitude of this correlation declines predictably with neighborhood size.
-In all spatial correlations random and point sampling are similarly correlated with space across neighborhood sizes, but midpoint sampling causes weaker correlations because it incorporates less genetic and geographic distance than the full sample.
+\begin{figure*}[p]
+\centering
+\includegraphics[width=.5\textwidth]{figures/ibs_dists2.pdf}
+\caption{Cumulative distributions for IBS tract lengths per pair of individuals at different geographic distances, across three neighborhood sizes ($N_{W}$).}
+\label{fig:ibs}
+\end{figure*}
+
+\subsubsection{IBS tracts and correlations with geographic distance}
+We next turn our attention to the effect of geographic distance on haplotype block length sharing, summarized for sets of nearby and distant individuals in Figure \ref{fig:ibs}.
+There are two main patterns to note.
+First, nearby individuals share more long IBS tracts than distant individuals (as expected because they are on average more closely related).
+Second, the difference in the number of long IBS tracts between nearby and distant individuals decreases as neighborhood size increases.
+This reflects the faster spatial mixing of populations with higher dispersal, which breaks down the correlation between the IBS tract length distribution and geographic distance.
+This can also be seen in the bottom row of Figure \ref{fig:sumstats}B, where the correlation coefficients between the summaries of the IBS tract length distribution (the mean, skew, and count of tracts over $1\times10^6bp$) and geographic distance approaches 0 as neighborhood size increases.
+
+%Last, in figure \ref{fig:ibs} we can see that most of the difference both between nearby and distant individuals and across different neighborhood sizes is found in the upper tail of the IBS tract length distribution.
+This is consistent with the methods for inferring dispersal distances described in \cite{Ringbauer2017} and \cite{Baharian2016}, which focus on long IBS blocks rather than the full distribution.
+
+The patterns observed for correlations of IBS tract lengths with geographic distance are similar to those observed in the more familiar regression of allele frequency measures such as $D_{xy}$ (i.e.
+"genetic distance") or $F_{st}$ against geographic distance \citep{Rousset1997}.
+$D_{xy}$ is positively correlated with the geographic distance between the individuals, and the strength of this correlation declines as dispersal increases (Figure \ref{fig:sumstats}B), as expected under theory \citep{Wright1943,Rousset1997}.
+This relationship is very similar across random and point sampling strategies, but is weaker for midpoint sampling, perhaps due to a dearth of long-distance comparisons.
+In much of empirical population genetics a regression of genetic differentiation against spatial distance is a de-facto metric of the significance of isolation by distance.
+The similar behavior of moments of the pairwise distribution of IBS tract lengths shows why haplotype block sharing has recently emerged as a promising source of information on spatial demography through methods described in \cite{Ringbauer2017} and \cite{Baharian2016}.
 
 %%%%%%%%%%%
+\begin{figure*}[p]
+\centering
+\includegraphics[width=\textwidth]{figures/distance_by_DAF.pdf}
+\caption{Trends in the distance among allele copies at varying derived allele frequencies and neighborhood sizes.}
+\label{fig:dafdists}
+\end{figure*}
+
 \subsubsection{Spatial distribution of allele copies}
-\plr{insert bit describing that plot here}
+Mutations occur in individuals and spread geographically over time.
+Because low frequency alleles generally represent recent mutations \citep{sawyer1977past, griffiths1999ages}, the geographic dispersion of an allele may covary along with its frequency in the population.
+To visualize this relationship we calculated the average distance among individuals carrying a focal derived allele across simulations with varying neighborhood sizes, shown in Figure \ref{fig:dafdists}.
+On average we find that low frequency alleles are the most geographically restricted, and that the extent to which geography and allele frequency are related depends on the amount of dispersal in the population.
+For populations with large neighborhood sizes we found that even very low frequency alleles can be found across the full landscape, whereas in populations with low neighborhood sizes the relationship between distance among allele copies and their frequency is quite strong.
+This is the basic process underlying \cite{Novembre2009}'s method for estimating dispersal distances based on the distribution of low frequency alleles, and also generates the greater degree of bias in GWAS effect sizes for low frequency alleles identified in \cite{Mathieson2012}.
 
@@
 One of the most important uses for population genetic data is inferring demographic history of populations.
-As demonstrated above, the site frequency spectrum varies across neighborhood sizes and sampling strategies.
+As demonstrated above, the site frequency spectrum and the distribution of IBS tracts varies across neighborhood sizes and sampling strategies.
 Does this variation lead to different inferences of past population sizes?
-To ask this we inferred population size histories from samples drawn from our simulated populations using a popular software package that uses the SFS as its information, Stairwayplot \citep{Liu2015}.
-
-\plr{revsit with smc++}
-Figure \ref{fig:demography} shows inferred population size histories, grouped by neighborhood size and sampling strategy.
-In general, Stairwayplot tends to infer ancient population increases and recent declines when neighborhood sizes were below 20 under all sampling strategies (Figure~\ref{fig:demography}). 
-This is consistent with our observations of the SFS from which Stairwayplot is doing its inference.
-Inflated past population sizes were seen in both point and random sampling, demonstrating that the relatively minor shift in the site frequency spectrum observed among sampling regimes is enough to alter demographic estimates.
-More alarmingly, inference of severe population bottlenecks was  common at neighborhood sizes under 100 for midpoint and point sampling strategies.
-Above neighborhood sizes of 100 the average inferred demography across all simulations was relatively accurate, with minor fluctuations slightly above the expected variance $Ne$.
-While that is so individual model fits were highly variable and often inferred five-fold or greater population fluctuations even in high-dispersal simulations.
-We compared this range of variation to Stairwayplot results run on coalescent simulations with constant population size,
-and found that the noisiest results ($N_W < 20$, or midpoint sampling and $N_W < 100$)
-were noiser than expected, but remaining simulation results were consistent with Stairwayplot's behavior on data from random mating models (Table~\ref{table:demography}).
-% To test whether the variation we observed in inferred demographic histories from Stairwayplot was the result of spatial effects in our simulations rather than the behavior of the optimization routine we compared standard deviations of inferred population sizes in each sampling/neighborhood size bin with those returned by an equilibrium coalescent simulation.
-Were these to be equal we could assume that the variation were purely as to be expected under the noise inherent in the genealogical process.
-Instead we found that standard deviations were significantly greater in all sampling strategies for neighborhood sizes under 20 and for midpoint sampling with neighborhood sizes 20-100, but all other sets performed similarly to the coalescent model (Table \ref{table:demography}).
-In summary, spatial mate choice and dispersal causes strong bias in SFS-based demographic estimates for neighborhood sizes below 20 or when sampling is clustered, but otherwise any biases are within the range of variability regularly inferred by Stairwayplot.
-This underscores the fact that some \textit{a priori} knowledge about the population dynamics at play will be important to interpreting results of demographic estimation routines.
+To ask this we inferred population size histories from samples drawn from our simulated populations with two approaches: Stairwayplot \citep{Liu2015}, which uses a genome-wide estimate of the site frequency spectrum (SFS), and SMC++ \citep{Terhorst2016}, which incorporates information on both the SFS and linkage disequilibrium across the genome.
+
+Figure \ref{fig:demography}A shows the median inferred population size histories from each method, grouped by neighborhood size and sampling strategy.
+In general these methods tend to slightly overestimate ancient population sizes and infer recent population declines when neighborhood sizes are below 20 and sampling is spatially clustered (Figure \ref{fig:demography}A, Figure \ref{fig:demography_supplement}).
+The overestimation of ancient population sizes however is relatively minor, averaging around a two-fold inflation at 10,000 generations before present in the worst-affected bins.
+For stairwayplot we found that many runs infer dramatic population bottlenecks in the last 1,000 generations when sampling is spatially concentrated, resulting in ten-fold or greater underestimates of recent population sizes.
+However SMC++ appeared more robust to this error, with runs on point- and midpoint-sampled simulations at the lowest neighborhood sizes underestimating recent population sizes by roughly half and those on randomly-sampled simulations showing little error.
+Above neighborhood sizes of around 100, both methods performed relatively well when averaging across results from multiple simulations.
+
+However, individual model fits from both methods frequently reflected turbulent demographic histories (Figure \ref{fig:demography_supplement}), with the standard deviation of inferred $N_{e}$ across time points often exceeding the expected $N_{e}$ for both methods (Figure \ref{fig:demography}B).
+That is, despite the constant population sizes in our simulations, both methods tended to infer large fluctuations in population size over time, which could potentially result in incorrect biological interpretations.
+On average the variance of inferred population sizes was elevated at the lowest neighborhood sizes and declines as dispersal increases, with the strongest effects seen in stairwayplot model fits with for clustered sampling and neighborhood sizes less than 20 (Figure \ref{fig:demography}B).
+This suggests that very limited dispersal, particularly when combined with clustered sampling, may bias these methods towards inferring more variable demographic histories.
 
@@
 \centering
-\includegraphics[width=\textwidth]{figures/stairwayplot_facet_rollmean.pdf}
-\caption{Inferred demographic histories for spatial SLiM simulations from Stairwayplot, by sampling scheme and neighborhood size (NS) range.
-The thick line is a rolling mean and thin lines are individual model fits.
-Dashed horizontal lines are the average $N_{e}$ across random-mating SLiM models estimated from $\theta_{\pi}$.}
+\includegraphics[width=\textwidth]{figures/demography_rollmedian_w_variance.pdf}
+\caption{A: Rolling median inferred $N_{e}$ trajectories for stairwayplot and smc++ across sampling strategies and neighborhood size bins.
+The dotted line shows the mean $N_{e}$ of random-mating simulations.
+B: Standard deviation of individual inferred $N_{e}$ trajectories, by neighborhood size and sampling strategy.
+Black lines are loess curves.
+Plots including individual model fits are shown in Figure \ref{fig:demography_supplement}.}
 \label{fig:demography}
@@
 As expected, spatial autocorrelation in the environment causes spurious associations 
-across much of the genome if no correction for genetic relatedness among samples is performed (Figure \ref{fig:gwas}). 
+across much of the genome if no correction for genetic relatedness among samples is performed (Figure \ref{fig:gwas}, figure \ref{fig:manhattan_plots}). 
 This effect is particularly strong for clinal and corner environments, 
 for which the lowest dispersal levels cause over 60\% of SNPs in the sample to return significant associations. 
-Patchy environmental distributions, which is less strongly spatially correlated (Figure \ref{fig:gwas}A), 
+Patchy environmental distributions, which are less strongly spatially correlated (Figure \ref{fig:gwas}A), 
 cause fewer false positives overall but still produce spurious associations at roughly 10\% of sites at the lowest neighborhood sizes. 
-Notably, no simulations with nonspatial environments returned more than one significant association (Figure \ref{fig:gwas}C), 
-demonstrating that this effect is caused specifically by the interaction of population structure and spatial variation in the environment rather than by population structure itself.
-
-The confounding effects of geographic structure are well-known, 
+Interestingly we also observed a small number of false-positive associations in roughly 3\% of analyses on simulations with nonspatial environments, both with and without PC covariates included in the regression.
+
+The confounding effects of geographic structure are well known, 
 and it is common practice to control for this by including principal components (PCs) as covariates to control for these effects.
 This mostly works in our simulations -- after doing this, the vast majority of SNPs no longer surpass a 5\% FDR significance threshold. 
-However, a substantial number of SNPs -- up to 1.5\% of SNPs -- still surpass this threshold (and thus would be false positives in a GWAS),
+However, a substantial number of SNPs -- up to 1.5\% at the lowest dispersal distances -- still surpass this threshold (and thus would be false positives in a GWAS),
 especially under ``corner'' and ``patchy'' environmental distributions (Figure~\ref{fig:gwas}C). 
@@
 Given an average of 132,000 SNPs across simulations after MAF filtering, this translates to up to 382 false-positive associations. 
-In most cases the $p$ values for these associations were significant after FDR correction but would not pass the threshold for significance under the more conservative Bonferroni correction.
+In most cases the $p$ values for these associations were significant after FDR correction but would not pass the threshold for significance under the more conservative Bonferroni correction (see example manhattan plots in figure \ref{fig:manhattan_plots}).
 
@@
 
-\plr{Add a few manhattan plots to the supplement and refer to them somewhere.}
-
 \begin{figure*}[p]
 \centering
-\includegraphics[width=\textwidth]{figures/gwas_summary.pdf}
+\includegraphics[width=\textwidth]{figures/gwas_summary_loglog.pdf}
     \caption{Impacts of spatially varying environments and isolation by distance on linear regression GWAS.
@@
 Quantile-quantile plots in \textbf{D} show inflation of $-\log_{10}(p)$ after PC correction across all simulations and environmental distributions, with colors scaled by the median neighborhood size in each region of q-q space.
-%A: example environmental distributions and sampling maps, with colors scaled to phenotype values.
-B: proportion of total variance explained by the first 10 PC axes, by neighborhood size.
-C: Proportion of significant SNPs after FDR correction for GWAS conducted with (top) or without (bottom) PC covariates.
-D: Quantile-quantile plots of observed $p$ values relative to those expected from a uniform distribution.
-The dotted lines show the 95\% confidence region assuming binomial sampling and an alpha level of 0.05.
-Points above the 1:1 line reflect deflated $p$ values.
-    \plr{suggestion: (a) log-scale the y axes in (C) and add another y-axis on the right showing (rough number of human genome SNPs above 5\% times this percentage) so we can see absolute numbers}
 }
@@
 so the genetic distance and number of recombination events separating sampled individuals simply gives a noisy picture of the genealogies connecting them. 
-% These genealogical patterns also relate directly to the site frequency spectrum. 
-% In our simulations we observed that groups of nearby individuals tend to coalesce rapidly, while coalescence among groups at opposite ends of the landscape takes much longer than under random mating.
-Tip branches (i.e., branches subtending only one individual) are then relatively short, and branches in the middle of the genealogy connecting local groups of individuals relatively long.
-These patterns then create the biases we observed in the SFS -- the lowest frequency bins are deflated by the short branch lengths connecting nearby individuals, while mid-frequency bins are inflated by the long branches connecting local groups.
+Tip branches (i.e., branches subtending only one individual) are then relatively short, and branches in the middle of the genealogy connecting local groups of individuals relatively long, leading to the biases in the site frequency spectrum shown in Figure \ref{fig:sumstats}.
 
@@
 One of the most important differences between random mating and spatial models is the effect of sampling: 
-in a randomly mating population the spatial distribution of sampling effort has no effect on estimates of genetic variation, 
+in a randomly mating population the spatial distribution of sampling effort has no effect on estimates of genetic variation (Table \ref{table:sumstats}) , 
 but when dispersal is limited sampling strategy can compound spatial patterns in the underlying genealogy and create pervasive impacts on all downstream genetic analyses. 
@@
 can provide an informative contrast.
+Finally we wish to point out that the bias towards intermediate allele frequencies that we observe may mean that the importance of linked selection, at least as is gleaned from the site frequency spectrum, may be systematically underestimated currently.
 
@@
 Classical results in population genetics collapse many aspects of life history into a single compound parameter, effective population size ($N_{e}$).
-Inferring effective population size back through time is now a common goal of population genomic analyses 
-and an important step in establishing baseline expectations of genetic variation, e.g., when searching for signals of selection. 
-Here we found that one method of inference of historic $N_{e}$ based on genome-wide estimates of the site frequency spectrum, Stairwayplot \citep{Liu2015}, is relatively robust to variation in dispersal distance when sampling is random and neighborhood size is over 20.
-However, non-random sampling, and particularly midpoint sampling, causes the method to infer inflated estimates of past population sizes and a series of recent bottlenecks (Figure \ref{fig:demography}).
-All sampling strategies lead to inflated ancient and deflated recent $N_{e}$ when neighborhood sizes were less than 20.
-
-These predictions match the biases visible in the raw site frequency spectrum (Figure \ref{fig:sumstats}) -- the deficit of low frequency alleles corresponds to the recent bottlenecks while the inflation of mid-frequency alleles corresponds to the high ancestral $N_{e}$.
-Though we found that Stairwayplot is a noisy estimator of equilibrium demography in general, there was no significant bias in demographic estimates for any sampling strategy for neighborhood sizes over 100.
-Thus many existing analyses are likely robust to biases in inferred $N_{e}$ caused by limited dispersal in continuous landscapes.
-However  barriers to dispersal will likely lead to higher levels of differentiation than we simulated here, and may mimic those seen at the low end of continuous dispersal we simulated.
-
-\cjb{could be another paragraph here discussing the relationship between Ne, the distribution of coalescence times, and dispersal.}
-\ak{wonder if it is worth revisting MSMC or similar now that we know more about how to run the method?}
+Inferring effective population size back through time is now a common goal of population genomic analyses and an important step in establishing baseline expectations of genetic variation, e.g., when searching for signals of selection.
+Here we found that these methods are subject to bias when neighborhood sizes are less than 20 and sampling is spatially clustered, with Stairwayplot in particular inferring dramatic recent population bottlenecks for many simulations during the last 1000 generations.
+However we were surprised to observe that, at least when averaged across multiple simulations, the two methods we tested were relatively robust to violations of the assumption of random mating for most continuously structured populations with neighborhood sizes over 100.
+Indeed for SMC++ we found that the errors in the worst-affected bins had a roughly a two-fold underestimate of recent populations and a two-fold overestimate of ancient populations.
+Though this degree of variation in population size is certainly meaningful in an ecological context, it is relatively minor in population genetic terms.
+
+A more worrying pattern was the high level of variance in inferred $N_{e}$ trajectories for individual model fits using these methods, which was highest in simulations with the smallest neighborhood size.
+Often this reflected population crashes or booms in the recent past, but in practice it is rarely clear how far in the past we have to go before interpretting the $N_{e}$ values returned by these methods.
+This suggests that, at a minimum, researchers working with empirical data should replicate analyses multiple times and take a rolling average if model fits are inconsistent across runs.
+Splitting samples and running replicates on separate subsets -- the closest an empirical study can come to our design of averaging the results from multiple simulations -- may also alleviate this issue.
+
+In general however, our analysis suggests that many empirical analyses of population size history using methods like SMC++ are somewhat robust to error caused by continuous spatial structure.
+Inferences drawn from static SFS-based methods like Stairwayplot should be treated with caution when there are signs of isolation by distance in the underlying data (for example, if a regression of $F_{st}$ against the logarithm of geographic distance has a significantly positive slope), and in particular an inference of population bottlenecks in the last 1000 years should be discounted if sampling is clustered, but estimates of deeper time patterns are likely to be fairly accurate.
+The biases in the SFS and haplotype structure we identified above are apparently small enough that they fall within the range of variability regularly inferred by these approaches, at least on datasets of the size we simulated.
 
@@
 
-Over the last twenty years, genome-wide association studies (GWAS) have identified tens of thousands of correlations between genetic variation and phentoypes, 
-both in humans and other species. 
-This technique is increasingly applied to questions of human health through methods like polygenic risk scores 
-that sum the effect sizes estimated from GWAS to predict an individual's phenotype or disease risk \citep{Khera2018}. 
-The most common method, which we followed in our analyses of simulated data here, is to regress the count of derived alleles at a site against individual phenotypes, 
-taking the slope of this regression as an estimate of the effect of the allele on the phenotype. 
+Over the last twenty years, genome-wide association studies (GWAS) have identified tens of thousands of correlations between genetic variation and phentoypes, both in humans and other species.
+Though originally intended to locate causal alleles for phenotypes controlled by a limited number of sites in the genome, GWAS is now routinely used as the first step in a phenotype-prediction pipeline that sums effect sizes across thousands or millions of variants to generate a "polygenic risk score" (PRS; \citep{Khera2018}).
+The most common method for estimating these effect sizes, which we followed in our analyses of simulated data here, is to regress the count of derived alleles at a site against individual phenotypes, taking the slope of this regression as an estimate of the effect of the allele on the phenotype.
 As recently reviewed by \cite{Visscher2019}, this is exactly the approach outlined by \cite{Fisher1918} at the dawn of quantitative genetics.
 
-Stepping back from specific methods,
-the general goal is to identify alleles that are more correlated with the phenotype of interest
-than one would expect due to chance.
-Both phenotype and genotype are often spatially autocorrelated (for different reasons),
-which makes spurious correlations a major obstacle,
-both by creating false positives and reducing power to identify truly causal alleles 
-\citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015}.
-It may be that the success of quantitative genetics in agriculture is in part because 
-crop and animal breeding operations can be conducted so that the environment is identical (or nearly so) or randomized across populations. 
-However, most natural populations are structured by a combination of limited dispersal and geographic barriers, 
-and nearly all environments vary over space.
+%Stepping back from the mechanics of GWAS specifically, Fisher's approach was to decompose the variance in phenotypes into environmental and genetic effects, e.g. 
+%\begin{equation}
+%P=G+E
+%\end{equation}
+%\begin{equation}
+%var(P)=var(G)+var(E)+cov(G,E)
+%\end{equation}
+For GWAS in structured populations, much of the bias identified in previous studies \citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015}, in which test statistics reflect population structure rather than phenotype association, arises because of the presence of a positive covariance between genotypic and environmental variation.
+When the environment and genotype covary across space their effects are confounded.
+Note this does not require interaction between genotype and environment (so-called GxE effects), which will introduce an additional source of bias.
+Here we refer simply to the fact that the allele frequency at a site will often covary with the environment when both breeding structure and the environment vary over space.
+To some degree the success of quantitative genetics in fields like agriculture may be due to the absence of this covariation -- crop and animal breeding operations can be conducted so that the environment is either controlled or measured across populations \citep{Wray2019}.
+However in natural populations this situation is extremely unlikely.
+Most populations are structured by a combination of limited dispersal and geographic barriers, and nearly all environments vary over space.
+GWAS in natural populations is then forced to confront the confounding effects of population structure and the environment directly.
 
 Incorporating PC positions as covariates in the analysis \citep{Price2006} is designed to address this difficulty by regressing out a baseline level of ``average'' differentiation. 
-% This approach is extremely useful, but does not truly separate the confounding signals of environment and spatially varying genotypes. 
 In essence, a PC-corrected GWAS asks ``what regions of the genome are more associated with this phenotype than the average genome-wide association observed across populations?'' 
 In our simulations, we observed that this procedure can fail under a variety of circumstances. 
-If dispersal is limited and environmental variation is clustered in space 
-(i.e., corner or patchy distributions in our simulations), 
-PCA positions fail to capture the fine-scale spatial structure required to remove all signals of association. 
-Conversely, as dispersal is increased, PCA loses power to describe population structure 
-before spatial mixing breaks down the relationship between genotype and the environment. 
-These effects were observed with all spatially correlated environmental patterns,
-but were particularly pronounced if environmental effects are concentrated in one region, as was also found by \citet{Mathieson2012}. 
-\plr{need to highlight a statement like this more}
-As a result we can expect to see several thousand weak false-positive associations in a PC-corrected GWAS conducted on a human-sized genome in species with neighborhood sizes up to at least 1000.
-
-This suggests some caveats for interpretation of GWAS results.
-Very few of the spurious associations we identified would be significant at a conservative Bonferroni-adjusted $p$-value cutoff, 
-suggesting that most of the very strong signals of association observed in studies of mono- or oligogenic traits are robust to these concerns.
-Further, the most dramatic $p$-value inflations occurred at neighborhood sizes below 100 
--- smaller than most recent human populations (but see below for further discussion of empirical cases). 
-However, as recently identified in studies of gentoype associations for human height in Europe \citep{Berg2018,Sohail2018}, 
-PC regression GWAS in modern human populations does leave residual signal of population structure in large-scale GWAS of polygenic traits. 
-\plr{Is there any indication that mixed models do better about this? (Where's the ``for precisely this reason'' come from?)}
-Indeed, studies in strongly structured species like \textit{Arabadopsis} have long relied on more sophisticated mixed model approaches to correcting for population structure for precisely this reason \citep{Aranzana2005,Sasaki2015}.
-
-A second point that has received less attention in the literature is the issue of overfitting in GWAS. 
+If dispersal is limited and environmental variation is clustered in space (i.e., corner or patchy distributions in our simulations), PCA positions fail to capture the fine-scale spatial structure required to remove all signals of association.
+Conversely, as dispersal increases, PCA loses power to describe population structure before spatial mixing breaks down the relationship between genotype and the environment.
+These effects were observed with all spatially correlated environmental patterns, but were particularly pronounced if environmental effects are concentrated in one region, as was also found by \citet{Mathieson2012}.
+
+Here we simulated a single chromosome with size roughly comparable to one human chromosome.
+If we scale the number of false-positive associations identified in our analyses to a GWAS conducted on whole-genome data from humans, we would expect to see several thousand weak false-positive associations after PC corrections in a population with neighborhood sizes up to at least 1000 (which should include values appropriate for many human populations).
+Notably, very few of the spurious associations we identified would be significant at a conservative Bonferroni-adjusted $p$-value cutoff (see Figure \ref{fig:manhattan_plots}).
+This suggests that GWAS focused on finding strongly associated alleles for traits controlled by a limited number of variants in the genome are likely robust to the impacts of continuous spatial structure.
+However, methods that analyze the combined effects of thousands or millions of weakly associated variants such as polygenic risk scores \citep{Khera2018} are likely to be affected by subtle population structure.
+Indeed as recently identified in studies of gentoype associations for human height in Europe \citep{Berg2018,Sohail2018}, PC regression GWAS in modern human populations do include residual signal of population structure in large-scale GWAS of polygenic traits.
+When attempting to make predictions across populations with different environmental exposures, PRS affected by population structure -- often termed "stratification bias" in the GWAS literature -- can be expected to offer low predictive power, as was shown in a recent study finding lower performance of PRS outside European populations \citep{Martin2019}.
+
+A second point that has received less attention in the literature (but see \cite{Lawson2019}) is the issue of overfitting in GWAS. 
 If a truly causal allele segregates at different frequencies in different populations, then correcting for population structure in a regression analysis will result in an underestimate of its effect size.
@@
 which found that increasing the number of PC axes used as covariates caused the total proportion of variance explained by SNPs to decline from $\approx0.8$ to $\approx0.75$ \citep{Wainschtein2019}.
-% Indeed SNPs with minor alleles frequencies of 0.001 - 0.01, which are expected to reflect fine scale population structure \citep{Mathieson2012,Novembre2009}, 
-% are estimated to explain \textit{negative} proportions of the total phenotypic variance in \citep{Wainschtein2019}.
-% \ak{i don't understand what the authors mean here -- how can we have a negative variance?}
-% Searching for genetic associations of polygenic traits that vary systematically across but not within populations through existing GWAS approaches is then unlikely to be successful: the signals are fully confounded, and new analytic methods or experiments controlling for variation in the environment will be necessary to rigorously identify causal variants.
-
-In summary, spatial covariation in population structure and the environment confound the interpretation of GWAS $p$-values, 
-and correction using principal components is insufficient to fully separate these signals for polygenic traits under a variety of environmental and population parameter regimes. 
+
+In summary, spatial covariation in population structure and the environment confound the interpretation of GWAS $p$-values, and correction using principal components is insufficient to fully separate these signals for polygenic traits under a variety of environmental and population parameter regimes.
 Other GWAS methods may be less sensitive to this confounding, but there is no obvious reason that this should be so.
@@
 Though this approach appears to work well in practice, its interpretation is not always straightforward and it is likely biased by the presence of linked selection \citep{Berg2018}.
+In addition, we observed that in many cases the false-positive SNPs we identified appeared to be concentrated in LD peaks similar to those expected from truly causal sites (Figure \ref{fig:manhattan_plots}).
 We suggest a straightforward alternative for species in which the primary axes of population differentiation is space 
@@
 Of course, it is possible that genotypes indirectly affect individual locations by adjusting organismal fitness and thus habitat selection across spatially varying environments, but we believe that this hypothesis should be tested against a null of stratification bias inflation rather than accepted as true based on GWAS results.
-\cjb{what do you think of the second half here?}
-\plr{love it}
 
@@
 
-%%%%%%%%%%%%
 \subsection{Where are natural populations on this spectrum?}
-
-How strong is the spatial patterning of genetic variation in extant species?
-% For how much of the tree of life do spatial patterns circumscribe genomic variation? 
-To get an idea of this, we gathered estimates of neighborhood size from a range of organisms (Table~\ref{table:NStable}).
-Although this sample is almost certainly biased towards small-neighborhood species 
-(because few studies have quantified neighborhood size in species with very high dispersal or population density), 
-we find that neighborhood sizes in the range we simulated are fairly common across a range of taxa. 
+For how much of the tree of life do spatial patterns circumscribe genomic variation?
+In Table \ref{table:NStable} we gathered estimates of neighborhood size from a range of organisms to get an idea of how likely dispersal is to play an important role in patterns of variation.
+Though this sample is almost certainly biased towards small-neighborhood species (because few studies have quantified neighborhood size in species with very high dispersal or population density), we find that neighborhood sizes in the range we simulated are fairly common across a range of taxa.
 At the extreme low end of empirical neighborhood size estimates we see some flowering plants, large mammals, and colonial insects like ants.
 Species such as this have neighborhood size estimates small enough that spatial processes are likely to strongly influence inference.
-These include some human populations such as the Gainj- and Kalam-speaking people of Papua New Guinea, 
-for whom the estimated neighborhood sizes in \citet{Rousset1997} range from 40 to 410 depending on the method of estimation. 
-Many more species occur in a middle range of neighborhood sizes between 100 and 1000 
--- a range in which spatial processes play a minor role in our analyses under random spatial sampling but are important when sampling of individuals in space is clustered. 
-Last, many species likely have neighborhood sizes much larger than we simulated, including modern humans in northeastern Europe \citep{Ringbauer2017}. 
-If these species' ranges and population densities have been constant for a long period of time (unlike humans),
-demographic inference methods that assume well-mixed populations may show minimal bias from spatial effects.
-% While that is so, we caution that association studies in which the effects of population structure are confounded with spatial variation in the environment are still sensitive to dispersal even at these large neighborhood sizes.
+These include some human populations such as the Gainj- and Kalam-speaking people of Papua New Guinea, in which the estimated neighborhood sizes in \citep{Rousset1997} range from 40 to 410 depending on the method of estimation.
+Many more species occur in a middle range of neighborhood sizes between 100 and 1000 -- a range in which spatial processes play a minor role in our analyses under random spatial sampling but are important when sampling of individuals in space is clustered.
+Last, many species likely have neighborhood sizes much larger than we simulated, including modern humans in NE Europe \citep{Ringbauer2017}.
+For these species demographic inference and summary statistics are likely to reflect minimal bias from spatial effects as long as dispersal is truly continuous across the landscape.
+While that is so we caution that association studies in which the effects of population structure are confounded with spatial variation in the environment are still sensitive to dispersal even at these large neighborhood sizes.
 
@@
 
-One complication in the inference of any spatial demographic parameter is the balance between local and global process.
+%One complication in the inference of any spatial demographic parameter is the balance between local and global process.
 Many species are structured locally by limited dispersal, but also contain deeply divergent lineages in different regions that reflect signals of ancient episodes of geographic isolation or strong barriers to dispersal.
@@
 
-%Recent methods that aim to estimate geographic barriers to dispersal \citep{Ringbauer2018,Petkova2015} partially address this issue by attempting to infer the location of regions resistant to gene flow, but here again the balance between historical and current demographic processes represented in our inference is unclear.
-Limiting analyses to low-frequency alleles \citep{Novembre2009} or between long haplotype blocks \citep{Ringbauer2017} is one approach to limiting inference to recent demography (because these genomic features should arise primarily from patterns in the most recent branches of the genealogy), but as we showed spatial processes also cause shifts in branch lengths throughout the genealogy.
-Further studies on the ways in which the sample genealogy responds to variation in dispersal both over time and across space are a good direction for future research.
-
-Though our continuous space simulation allows incorporation of realistic demographic and spatial processes and is much faster than previous individual-based models, 
+One facet of spatial variation that we did not address in this study is the confounding of dispersal and population density implicit in the definition of Wright's Neighborhood Size.
+Our simulations were run under constant densities, but \cite{Ringbauer2017}'s approach to demographic inference in space suggests that density and dispersal can in some cases be estimated separately from genetic data.
+Much additional work remains to be done to better understand how these parameters interact to shape genetic variation in continuous space, which we leave to future studies.
+
+Though our simulation allows incorporation of realistic demographic and spatial processes, 
 it is inevitably limited by the computational burden of tracking tens or hundreds of thousands of individuals in every generation. 
-In particular, computations required for mate selection and spatial competition caled as the number of individuals within a three-$\sigma$ radius increases. 
-% In part the issue of runtime scaling is a function of the spatial process itself -- under very limited dispersal we observed that coalescence requires over 30N generations, so forward-time methods must be run for a very long time to create a complete genealogy as underlies all real genome sequence data.
+In particular, computations required for mate selection and spatial competition scale \cjb{Peter -- what's the correct scaling here?} as the number of individuals within a three-$\sigma$ radius increases.
 The reverse-time model of continuous space evolution described by \citet{Barton2010} and implemented by \citet{Kelleher2014} 
@@
 using principal components analysis or similar decompositions is a difficult issue, well worth considering on its own.
-%-- what does it mean to correct for population structure in a world without discrete demes? 
 How can we best avoid spurious correlations while correlating genetic and phenotypic variation 
@@
 Perhaps optimistically,
-% What are the boundaries between proper corrections between ancestry and underpowering the search for local genetic variation? 
 we posit that process-driven descriptions of ancestry 
@@
 \section{Acknowledgements}
-We thank folks for reading and thinking about this manuscript. 
-We thank the Hearth for having such good nearby coffee. 
+We thank Brandon Cooper, and others for reading and thinking about this manuscript. 
+We thank the Hearth for having such good, nearby coffee. 
 Falling Sky Brewing was key to the success of this research endeavor. 
@@
 \label{fig:hetmap}
+\end{figure*}
+
+\begin{figure}[p]
+\centering
+\includegraphics[width=\textwidth]{figures/demography_combined.pdf}
+\caption{Inferred demographic histories for spatial SLiM simulations, by sampling scheme and neighborhood size (NS) range.
+Thick lines are rolling medians across all simulations in a bin and thin lines are best fit models for each simulation.
+Dashed horizontal lines are the average $N_{e}$ across random-mating SLiM models estimated from $\theta_{\pi}$.}
+\label{fig:demography_supplement}
+\end{figure}
+
+\afterpage{\clearpage}
+\begin{figure*}[p]
+\centering
+\includegraphics[width=\textwidth]{figures/manhattan_plots_combo-01.png}
+\caption{Manhattan plots for a sample of simulations at varying neighborhood sizes.
+Labels on the right of each plot describes the spatial distribution of environmental factors (described in the methods section of the main text).
+Points in red are significantly associated with a nongenetic phenotype using a 5\% FDR threshold (dashed line).
+For runs with significant associations the dotted line is a Bonferroni-adjusted cutoff for p=0.05.}
+\label{fig:manhattan_plots}
 \end{figure*}
@@
 $skew(IBS)$ & \makecell[l]{Skew of the distribution of pairwise identical-by-state (IBS) \\tract lengths taken over all pairs.} \\
-$nIBS$ & Number of IBS tracts with length > 2bp across all pairs of individuals. \\
-$mean(IBS>1e6)$ & \makecell[l]{Mean number of IBS tracts over $1\times10^6$bp per pair \\across all pairs in the sample.} \\ 
+$nIBS$ & Mean number of IBS tracts with length > 2bp across all pairs in the sample. \\
+$nIBS>1e6$ & \makecell[l]{Mean number of IBS tracts over $1\times10^6$bp per pair \\across all pairs in the sample.} \\ 
 $corr(D_{xy},dist)$ & Pearson correlation between genetic distance and $\log_{10}(spatial\text{ }distance)$ \\
@@
 segsites & random mating & 0.998190 & 0.980730 \\ 
-pi & random mating & 0.997750 & 0.996450 \\ 
-thetaW & random mating & 0.998190 & 0.980730 \\ 
-tajD & random mating & 0.879690 & 0.188770 \\ 
-het\_o & random mating & 0.531540 & 0.433230 \\ 
-fis & random mating & 0.474790 & 0.785730 \\ 
-gen\_dist\_mean & random mating & 0.997770 & 0.996510 \\ 
-gen\_dist\_var & random mating & 0.283630 & 0.647240 \\ 
-gen\_dist\_skew & random mating & 0.958320 & 0.260750 \\ 
-gen\_sp\_corr & random mating & 0.601980 &\textbf{0.000000} \\ 
-ibs\_mean & random mating & 0.997960 & 0.997730 \\ 
-ibs\_var & random mating & 0.486450 & 0.399490 \\ 
-ibs\_skew & random mating & 0.117980 & 0.069770 \\ 
-ibs\_blocks\_per\_pair & random mating & 0.997680 & 0.996570 \\ 
-ibs\_blocks\_over\_1e6\_per\_pair & random mating & 0.834870 & 0.888730 \\ 
-ibs\_mean\_spat\_corr & random mating & 0.073270 & 0.308420 \\ 
-ibs\_1e6blocks\_spat\_corr & random mating & 0.268440 & \textbf{0.002100} \\ 
-ibs\_skew\_spat\_corr & random mating & 0.396920 & \textbf{0.000620} \\ 
-ibs\_blocks\_spat\_corr & random mating & 0.581090 & \textbf{0.000000} \\ 
+$\Theta_{\pi}$ & random mating & 0.997750 & 0.996450 \\ 
+$\Theta_{W}$ & random mating & 0.998190 & 0.980730 \\ 
+Tajima's $D$ & random mating & 0.879690 & 0.188770 \\ 
+observed heterozygosity & random mating & 0.531540 & 0.433230 \\ 
+$F_{is}$ & random mating & 0.474790 & 0.785730 \\ 
+$mean(D_{xy})$ & random mating & 0.997770 & 0.996510 \\ 
+$var(D_{xy})$ & random mating & 0.283630 & 0.647240 \\ 
+$skew(D_{xy})$ & random mating & 0.958320 & 0.260750 \\ 
+$corr(Dxy,dist)$ & random mating & 0.601980 &\textbf{0.000000} \\ 
+$mean(IBS)$ & random mating & 0.997960 & 0.997730 \\ 
+$var(IBS)$ & random mating & 0.486450 & 0.399490 \\ 
+$skew(IBS)$ & random mating & 0.117980 & 0.069770 \\ 
+$nIBS$ & random mating & 0.997680 & 0.996570 \\ 
+$nIBS>1e6$ & random mating & 0.834870 & 0.888730 \\ 
+$corr(mean(IBS),dist)$ & random mating & 0.073270 & 0.308420 \\ 
+$corr(IBS>1e6,dist)$ & random mating & 0.268440 & \textbf{0.002100} \\ 
+$corr(skew(IBS),dist)$ & random mating & 0.396920 & \textbf{0.000620} \\ 
+$corr(nIBS,dist)$ & random mating & 0.581090 & \textbf{0.000000} \\ 
 segsites & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
-pi & spatial & \textbf{0.026510} & \textbf{0.013440} \\ 
-thetaW & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
-tajD & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
-het\_o & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
-fis & spatial & \textbf{0.000000} & \textbf{0.000120} \\ 
-gen\_dist\_mean & spatial & \textbf{0.025390} & \textbf{0.012910} \\ 
-gen\_dist\_var & spatial & \textbf{0.004970} & \textbf{0.006230} \\ 
-gen\_dist\_skew & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
-gen\_sp\_corr & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
-ibs\_mean & spatial & 0.272400 & 0.114250 \\ 
-ibs\_var & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
-ibs\_skew & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
-ibs\_blocks\_per\_pair & spatial & \textbf{0.033920} & \textbf{0.016640} \\ 
-ibs\_blocks\_over\_1e6\_per\_pair & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
-ibs\_mean\_spat\_corr & spatial & \textbf{0.000000} & 0.590540 \\ 
-ibs\_1e6blocks\_spat\_corr & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
-ibs\_skew\_spat\_corr & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
-ibs\_blocks\_spat\_corr & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+$\Theta_{\pi}$ & spatial & \textbf{0.026510} & \textbf{0.013440} \\ 
+$\Theta_{W}$ & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+Tajima's $D$ & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+observed heterozygosity & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+$F_{is}$ & spatial & \textbf{0.000000} & \textbf{0.000120} \\ 
+$mean(D_{xy})$ & spatial & \textbf{0.025390} & \textbf{0.012910} \\ 
+$var(D_{xy})$ & spatial & \textbf{0.004970} & \textbf{0.006230} \\ 
+$skew(D_{xy})$ & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+$corr(Dxy,dist)$ & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+$mean(IBS)$ & spatial & 0.272400 & 0.114250 \\ 
+$var(IBS)$ & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+$skew(IBS)$ & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+$nIBS$ & spatial & \textbf{0.033920} & \textbf{0.016640} \\ 
+$nIBS>1e6$ & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+$corr(mean(IBS),dist)$ & spatial & \textbf{0.000000} & 0.590540 \\ 
+$corr(IBS>1e6,dist)$ & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+$corr(skew(IBS),dist)$ & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
+$corr(nIBS,dist)$ & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
 \hline
@@
 % Fri Mar 22 13:50:47 2019
-\begin{table}[ht]
-\small
-\centering
-\caption{\bf T-test results comparing standard deviations of inferred $N_{e}$ between spatial and coalescent models, by neighborhood size (NS) and sampling strategy.
-$p$ is the probability that spatial models have higher standard deviations.}
-\begin{tabular}{rllrrrrr}
-  \hline
-sampling & NS range & t & df & $p$ \\ 
-  \hline
-random & 2-20 & 4.2572 & 41.6166 & 0.0001 \\ 
-random & 20-100 & -1.8473 & 171.9905 & 0.9668 \\ 
-random & 100-500 & -2.1297 & 164.3864  & 0.9827 \\ 
-random & 500-1000 & -3.9681 & 147.0497 & 0.9999 \\ 
-point & 2-20 & 7.0802 & 44.3615  & 0.0000 \\ 
-point & 20-100 & -0.2038 & 169.3799 & 0.5806 \\ 
-point & 100-500 & -2.4945 & 152.5000 & 0.9932 \\ 
-point & 500-1000 & -3.8329 & 162.6443& 0.9999 \\ 
-midpoint & 2-20 & 5.9253 & 59.5462 & 0.0000 \\ 
-midpoint & 20-100 & 3.8940 & 171.7005  & 0.0001 \\ 
-midpoint & 100-500 & -2.2764 & 139.5221 & 0.9878 \\ 
-midpoint & 500-1000 & -3.2223 & 165.0792 & 0.9992 \\ 
-   \hline
-\end{tabular}
-\label{table:demography}
-\end{table}
+%\begin{table}[ht]
+%\small
+%\centering
+%\caption{\bf T-test results comparing standard deviations of inferred $N_{e}$ between spatial %and coalescent models, by neighborhood size (NS) and sampling strategy.
+$p$ is the probability %that spatial models have higher standard deviations.}
+%\begin{tabular}{rllrrrrr}
+%  \hline
+%sampling & NS range & t & df & $p$ \\ 
+%  \hline
+%random & 2-20 & 4.2572 & 41.6166 & 0.0001 \\ 
+%random & 20-100 & -1.8473 & 171.9905 & 0.9668 \\ 
+%random & 100-500 & -2.1297 & 164.3864  & 0.9827 \\ 
+%random & 500-1000 & -3.9681 & 147.0497 & 0.9999 \\ 
+%point & 2-20 & 7.0802 & 44.3615  & 0.0000 \\ 
+%point & 20-100 & -0.2038 & 169.3799 & 0.5806 \\ 
+%point & 100-500 & -2.4945 & 152.5000 & 0.9932 \\ 
+%point & 500-1000 & -3.8329 & 162.6443& 0.9999 \\ 
+%midpoint & 2-20 & 5.9253 & 59.5462 & 0.0000 \\ 
+%midpoint & 20-100 & 3.8940 & 171.7005  & 0.0001 \\ 
+%midpoint & 100-500 & -2.2764 & 139.5221 & 0.9878 \\ 
+%midpoint & 500-1000 & -3.2223 & 165.0792 & 0.9992 \\ 
+%   \hline
+%\end{tabular}
+%\label{table:demography}
+%\end{table}
 

### 2019-05-09 ms/spaceness.tex
andy & cj overleaf edits w typo fixes

@@
 Two areas in which spatial structure may be particularly important are demographic inference and genome-wide association studies (GWAS).
-Effective population size is a fundamental parameter of population genetics and shapes expectations of the efficacy of selection, the rate of drift, among other evolutionary forces (\textit{c.f.} \cite{gillespie2001population}).
+Effective population size is a fundamental parameter of population genetics and shapes expectations of the efficacy of selection and the rate of drift, among other evolutionary forces (\textit{c.f.} \cite{gillespie2001population}).
 Inferring a trajectory of population sizes through time has therefore become a major goal in population genetics, with many recent methods proposed to accomplish this task \citep{Liu2015,Terhorst2016,Schiffels2014,Sheehan2013}.
@@
 Because isolation by distance is expected to change the rate of drift in a population \citep{Maruyama1972}, it may cause biases in inferred population sizes.
-However the effects of continuous spatial structure on demographic inference has not to our knowledge been previously investigated.
+However the effect of continuous spatial structure on demographic inference has not to our knowledge been previously investigated.
 
@@
 or the number of potential parents of a randomly chosen individual.
-Empirical estimates of neighborhood size vary hugely across species (see Discussion), but for reference estimates in human populations range from 40 to over 5,000 depending on the population and method of estimation (\ref{table:NStable}), and \cite{Wright1946} estimated values of 500-1000 for \textit{Drosophila pseudoobscura}.
-\cite{Maruyama1972} found that the rate of decline in genetic diversity in a two-dimensional continuous population 
-approaches the random mating expectation when $\rho\sigma^2 > 1$, 
-and proposed that this had the important implication that most population genetic expectations for randomly mating populations 
-could be applied to continuously distributed populations with relatively little error.
+Empirical estimates of neighborhood size vary hugely across species (see Discussion), but for reference estimates in human populations range from 40 to over 5,000 depending on the population and method of estimation (Table \ref{table:NStable}).
+%\cite{Maruyama1972} found that the rate of decline in genetic diversity in a two-dimensional continuous population approaches the random mating expectation when $\rho\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.
 
@@
 Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$ where $L$ is the expected lifespan.
-who disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
+Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
 in both the $x$ and $y$ coordinates, reflected to stay within the species range.

### 2019-05-14 ms/spaceness.tex
demography lit updates

@@
     to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and genome-wide association studies. 
-    We find that several common summary statistics have distributions that differ substantially from that seen in well-mixed populations, 
+    We find that most common summary statistics have distributions that differ substantially from that seen in well-mixed populations, 
     especially when Wright's neighborhood size is less than 100 and sampling is spatially clustered.
@@
     We also show that the combination of spatially autocorrelated environments and limited dispersal
-    can cause genome-wide association studies to identify numerous spurious signals of genetic association 
+    causes genome-wide association studies to identify spurious signals of genetic association 
     with purely environmentally determined phenotypes,
@@
 For instance, if mean viability or fecundity is spatially autocorrelated, then limited geographic dispersal will lead to parent--offspring correlations.
-Furthermore, nearby individuals will be more closely related than an average random pair, so drawing multiple samples from the same area of on the landscape will represent a biased sample of the genetic variation present in the whole population.
-It has long been appreciated that this model misspecification subjects downstream inferences to bias, but the extent and nature of these effects remain largely uninvestigated.
+Furthermore, nearby individuals will be more closely related than an average random pair, so drawing multiple samples from the same area of on the landscape will represent a biased sample of the genetic variation present in the whole population \citep{Stadler2009}.% It has long been appreciated that this model misspecification subjects downstream inferences to bias, but the extent and nature of these effects remain largely uninvestigated.
 
-Two areas in which spatial structure may be particularly important are demographic inference and genome-wide association studies (GWAS).
-Effective population size is a fundamental parameter of population genetics and shapes expectations of the efficacy of selection and the rate of drift, among other evolutionary forces (\textit{c.f.} \cite{gillespie2001population}).
-Inferring a trajectory of population sizes through time has therefore become a major goal in population genetics, with many recent methods proposed to accomplish this task \citep{Liu2015,Terhorst2016,Schiffels2014,Sheehan2013}.
-All of these methods assume that the populations being modeled are approximately randomly mating (though this also depends on how researchers interpret the inferred parameter $\Theta=4N\mu$).
-Because isolation by distance is expected to change the rate of drift in a population \citep{Maruyama1972}, it may cause biases in inferred population sizes.
-However the effect of continuous spatial structure on demographic inference has not to our knowledge been previously investigated.
+\textcolor{red}{Two areas in which spatial structure may be particularly important are demographic inference and genome-wide association studies (GWAS).
+Previous work has found that discrete population structure can create false signatures of population bottlenecks when attempting to infer demographic histories from microsatellite variation \citep{Chikhi2010}, statistics summarizing the site frequency spectrum (SFS) \citep{Ptak2002,Stadler2009,Onge2012}, or runs of homozygosity in a single individual \citep{Mazet2015}.
+The increasing availability of whole-genome data has led to the development of many methods that attempt to infer detailed trajectories of population sizes through time based on a variety of summaries of genetic data \citep{Liu2015,Schiffels2014,Sheehan2013,Terhorst2016}.
+Because all of these methods assume that the populations being modeled are approximately randomly mating, they are likely affected by spatial biases in the genealogy of sampled individuals \citep{Wakeley1999}, which may lead to incorrect inference of population changes over time \citep{Mazet2015}.
+However, previous investigations of these effects have focused on discrete rather than continuous space models, and the level of isolation by distance at which inference of population size trajectories become biased by structure is not well known.
+Here we test how two methods suitable for use with large samples of individuals -- Stairwayplot \citep{Liu2015} and SMC++ \citep{Terhorst2016} -- perform when applied to populations evolving in continuous space with varying sampling strategies and levels of dispersal.
+}
 
@@
 Statistics like Tajima's $D$ are essentially summaries of the site frequency spectrum, which itself reflects variation in branch lengths and tree structure of the underlying genealogies of sampled individuals.
-Geographically limited mate choice distorts the distribution of these genealogies \citep{Maruyama1972}, which can affect the value of Tajima's $D$.
+Geographically limited mate choice distorts the distribution of these genealogies \citep{Maruyama1972,Wakeley1999}, which can affect the value of Tajima's $D$ \citep{Stadler2009}.
 Similarly, the distribution of tract lengths of identity by state among individuals contains information about not only historical demography \citep{Harris2013,ralph2013geography} and selection \citep{Garud2015}, but also dispersal and mate choice \citep{Ringbauer2017,Baharian2016}.
@@
 Indeed, mathematical results suggest that genetic variation in some spatial models should be well-approximated by a Wright-Fisher population if neighborhood size is large and all samples are geographically widely separated \citep{wilkins2004separation,zahle2005stepping}.
-However, the behavior of most common population genetic summary statistics has not yet been described in realistic geographic models.
+However, the behavior of most common population genetic summary statistics \textcolor{red}{other than Tajima's $D$ \citep{Stadler2009}} has not yet been described in realistic geographic models.
 Moreover, as we will show, spatial sampling strategies can affect summaries of variation at least as strongly as the underlying population dynamics.
@@
 Figure \ref{fig:sumstats} shows the effect of varying neighborhood size and sampling strategy on the site frequency spectrum (Figure~\ref{fig:sumstats}A) and several standard population genetic summary statistics (Figure~\ref{fig:sumstats}B).
-SFS comparisons show a significant enrichment of intermediate frequency variants in comparison to the nonspatial expectation, for smaller neighborhood sizes ($\leq 100$) that is exacerbated by midpoint and point sampling of individuals (depicted in Figure~\ref{fig:samplemap}).
+\textcolor{red}{Consistent with findings in n-island and stepping stone simulations \citep{Stadler2009}, the SFS shows a significant enrichment of intermediate frequency variants in comparison to the nonspatial expectation.
+This bias is most pronounced below neighborhood sizes $\leq 100$ and is exacerbated by midpoint and point sampling of individuals (depicted in Figure~\ref{fig:samplemap}).}
 Reflecting this, Tajima's $D$ is quite positive in the same situations (Figure~\ref{fig:sumstats}B).
@@
 
-One of the most commonly used summaries of variation is Tajima's summary of nucleotide divergence, $\theta_{\pi}$, calculated as the mean density of nucleotide differences, averaged across pairs of samples.
+One of the most commonly used summaries of variation is Tajima's summary of nucleotide divergence, $\theta_{\pi}$, calculated as the mean density of nucleotide differences averaged across pairs of samples.
 As can be seen in Figure \ref{fig:sumstats}B, $\theta_{\pi}$ differs as much as nearly three-fold between the random mating and spatial models.
@@
 Does this variation lead to different inferences of past population sizes?
-To ask this we inferred population size histories from samples drawn from our simulated populations with two approaches: Stairwayplot \citep{Liu2015}, which uses a genome-wide estimate of the site frequency spectrum (SFS), and SMC++ \citep{Terhorst2016}, which incorporates information on both the SFS and linkage disequilibrium across the genome.
+To ask this we inferred population size histories from samples drawn from our simulated populations with two approaches: Stairwayplot \citep{Liu2015}, which uses a genome-wide estimate of the SFS, and SMC++ \citep{Terhorst2016}, which incorporates information on both the SFS and linkage disequilibrium across the genome.
 
@@
 On average the variance of inferred population sizes was elevated at the lowest neighborhood sizes and declines as dispersal increases, with the strongest effects seen in stairwayplot model fits with for clustered sampling and neighborhood sizes less than 20 (Figure \ref{fig:demography}B).
-This suggests that very limited dispersal, particularly when combined with clustered sampling, may bias these methods towards inferring more variable demographic histories.
 
@@
 \subsection{Effects of Sampling}
+
 One of the most important differences between random mating and spatial models is the effect of sampling: 
 in a randomly mating population the spatial distribution of sampling effort has no effect on estimates of genetic variation (Table \ref{table:sumstats}) , 
-but when dispersal is limited sampling strategy can compound spatial patterns in the underlying genealogy and create pervasive impacts on all downstream genetic analyses. 
+but when dispersal is limited sampling strategy can compound spatial patterns in the underlying genealogy and create pervasive impacts on all downstream genetic analyses \textcolor{red}{(see also \cite{Stadler2009})}.
 In most species,
@@
 Because midpoint sampling provides limited spatial resolution it dramatically reduces the magnitude of observed correlations between spatial and genetic distances. 
-More surprisingly, midpoint sampling also leads to strongly positive Tajima's $D$ and an inflation in the proportion of heterozygous individuals in the sample. 
+More surprisingly, midpoint sampling also leads to strongly positive Tajima's $D$ and an inflation in the proportion of heterozygous individuals in the sample -- similar to the effect of sampling a single deme in an n-island model as reported in \citep{Stadler2009}.
 This increase in observed heterozygosity appears to reflect the effects of range edges, which are a fundamental facet of spatial genetic variation. 
@@
 
-Classical results in population genetics collapse many aspects of life history into a single compound parameter, effective population size ($N_{e}$).
-Inferring effective population size back through time is now a common goal of population genomic analyses and an important step in establishing baseline expectations of genetic variation, e.g., when searching for signals of selection.
-Here we found that these methods are subject to bias when neighborhood sizes are less than 20 and sampling is spatially clustered, with Stairwayplot in particular inferring dramatic recent population bottlenecks for many simulations during the last 1000 generations.
-However we were surprised to observe that, at least when averaged across multiple simulations, the two methods we tested were relatively robust to violations of the assumption of random mating for most continuously structured populations with neighborhood sizes over 100.
-Indeed for SMC++ we found that the errors in the worst-affected bins had a roughly a two-fold underestimate of recent populations and a two-fold overestimate of ancient populations.
-Though this degree of variation in population size is certainly meaningful in an ecological context, it is relatively minor in population genetic terms.
+\textcolor{red}{Previous studies have found that population structure and nonrandom sampling can create spurious signals of population bottlenecks when attempting to infer demographic history with microsatellite variation, summary statistics, or runs of homozygosity \citep{Chikhi2010,Stadler2009,Ptak2002,Mazet2015}.
+Here we found that methods that infer detailed population trajectories through time based on the SFS and patterns of LD across the genome are also subject to this bias, with some combinations of dispersal and sampling strategy inferring deep recent population bottlenecks and overestimating ancient $N_{e}$ by a around a factor of 2.
+However we were surprised to see that both Stairwayplot and SMC++ can tolerate relatively strong isolation by distance -- i.e.
+neighborhood sizes of 20 -- and still perform well when averaging results across multiple simulations.
+Inference in populations with neighborhood sizes over 20 was relatively accurate unless samples were concentrated in the middle of the range (Figure \ref{fig:demography}).
+As we will discuss below, most empirical estimates of neighborhood size, including all estimates for human populations, are large enough that population size trajectories inferred by these approaches should not be strongly affected by spatial biases created by dispersal in continuous landscapes.
+In contrast, \cite{Mazet2015} found that varying migration rates through time could create strong biases in inferred population trajectories from an $n$-island model with parameters relevant for human history, suggesting that changes in migration rates through time are more likely to drive variation in inferred $N_{e}$ than isolation by distance.}
 
-A more worrying pattern was the high level of variance in inferred $N_{e}$ trajectories for individual model fits using these methods, which was highest in simulations with the smallest neighborhood size.
-Often this reflected population crashes or booms in the recent past, but in practice it is rarely clear how far in the past we have to go before interpretting the $N_{e}$ values returned by these methods.
+\textcolor{red}{Between the two methods we tested, we found that SMC++ was more robust to spatial biases than Stairwayplot, with the worst-affected bins underestimating recent populations by roughly half rather than nearly 10-fold as with Stairwayplot.} Though this degree of variation in population size is certainly meaningful in an ecological context, it is relatively minor in population genetic terms.
+A more worrying pattern was the high level of variance in inferred $N_{e}$ trajectories for individual model fits using these methods, which was highest in simulations with the smallest neighborhood size (Figure \ref{fig:demography}, Figure \ref{fig:demography_supplement}).
 This suggests that, at a minimum, researchers working with empirical data should replicate analyses multiple times and take a rolling average if model fits are inconsistent across runs.
@@
 
-In general however, our analysis suggests that many empirical analyses of population size history using methods like SMC++ are somewhat robust to error caused by continuous spatial structure.
+In general however, our analysis suggests that many empirical analyses of population size history using methods like SMC++ are robust to error caused by spatial structure within continuous landscapes.
 Inferences drawn from static SFS-based methods like Stairwayplot should be treated with caution when there are signs of isolation by distance in the underlying data (for example, if a regression of $F_{st}$ against the logarithm of geographic distance has a significantly positive slope), and in particular an inference of population bottlenecks in the last 1000 years should be discounted if sampling is clustered, but estimates of deeper time patterns are likely to be fairly accurate.
-The biases in the SFS and haplotype structure we identified above are apparently small enough that they fall within the range of variability regularly inferred by these approaches, at least on datasets of the size we simulated.
+The biases in the SFS and haplotype structure identified above (see also \citep{Wakeley1999,Chikhi2010,Stadler2009}) are apparently small enough that they fall within the range of variability regularly inferred by these approaches, at least on datasets of the size we simulated.
 
@@
 
-Over the last twenty years, genome-wide association studies (GWAS) have identified tens of thousands of correlations between genetic variation and phentoypes, both in humans and other species.
+\textcolor{red}{Spatial structure is particularly challenging for genome-wide association studies, because the effects of dispersal on genetic variation are compounded by spatial variation in the environment \citep{Mathieson2012}.
+Spatially restricted mate choice and dispersal causes variation in allele frequencies across the range of a species.
+If environmental factors affecting the phenotype of interest also vary over space, then groups of individuals in different regions will both experience different environments and have different allele frequencies at many sites in the genome.
+In this scenario an uncorrected GWAS will infer genetic associations with a purely environmental phenotype at any site in the genome that is differentiated over space, and the relative degree of bias will be a function of the degree of covariation in allele frequencies and the environment (i.e.
+Figure \ref{fig:gwas}C, bottom panel).
+This pattern is now well known and has been demonstrated in a variety of simulation and empirical contexts \citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015,Berg2018,Sohail2018}.}
+
+Incorporating PC positions as covariates in a linear-regression GWAS \citep{Price2006} is designed to address this challenge by regressing out a baseline level of ``average'' differentiation. 
+In essence, a PC-corrected GWAS asks ``what regions of the genome are more associated with this phenotype than the average genome-wide association observed across populations?'' 
+In our simulations, we observed that this procedure can fail under a variety of circumstances. 
+If dispersal is limited and environmental variation is clustered in space (i.e., corner or patchy distributions in our simulations), PCA positions fail to capture the fine-scale spatial structure required to remove all signals of association.
+Conversely, as dispersal increases, PCA loses power to describe population structure before spatial mixing breaks down the relationship between genotype and the environment.
+These effects were observed with all spatially correlated environmental patterns, but were particularly pronounced if environmental effects are concentrated in one region, as was also found by \citet{Mathieson2012}.
+Though increasing the number of PC axes used in the analysis may reduce the false-positive rate, this may also decrease the power of the test to detect truly causal alleles \citep{Lawson2019}.
+
+Here we simulated a single chromosome with size roughly comparable to one human chromosome.
+If we scale the number of false-positive associations identified in our analyses to a GWAS conducted on whole-genome data from humans, we would expect to see several thousand weak false-positive associations after PC corrections in a population with neighborhood sizes up to at least 1000 (which should include values appropriate for many human populations).
+Notably, very few of the spurious associations we identified would be significant at a conservative Bonferroni-adjusted $p$-value cutoff (see Figure \ref{fig:manhattan_plots}).
+This suggests that GWAS focused on finding strongly associated alleles for traits controlled by a limited number of variants in the genome are likely robust to the impacts of continuous spatial structure.
+However, methods that analyze the combined effects of thousands or millions of weakly associated variants such as polygenic risk scores \citep{Khera2018} are likely to be affected by subtle population structure.
+Indeed as recently identified in studies of gentoype associations for human height in Europe \citep{Berg2018,Sohail2018}, PC regression GWAS in modern human populations do include residual signal of population structure in large-scale analyses of polygenic traits.
+When attempting to make predictions across populations with different environmental exposures, PRS affected by population structure can be expected to offer low predictive power, as was shown in a recent study finding lower performance of PRS outside European populations \citep{Martin2019}.
+
+In summary, spatial covariation in population structure and the environment confound the interpretation of GWAS $p$-values, and correction using principal components is insufficient to fully separate these signals for polygenic traits under a variety of environmental and population parameter regimes.
+Other GWAS methods may be less sensitive to this confounding, but there is no obvious reason that this should be so.
+One approach to estimating the degree of bias in GWAS caused by population structure is LD score regression \citep{Bulik-Sullivan2015}. 
+Though this approach appears to work well in practice, its interpretation is not always straightforward and it is likely biased by the presence of linked selection \citep{Berg2018}.
+\textcolor{red}{In addition, we observed that in many cases the false-positive SNPs we identified appeared to be concentrated in LD peaks similar to those expected from truly causal sites (Figure \ref{fig:manhattan_plots}), which may confound LD score regression.}
+We suggest a straightforward alternative for species in which the primary axes of population differentiation is space 
+(note this is likely not the case for some modern human populations): 
+run a GWAS with spatial coordinates as phenotypes and check for $p$-value inflation or significant associations. 
+If significant associations with sample locality are observed after correcting for population structure, 
+the method is sensitive to false positives induced by spatial structure. 
+This is essentially the approach taken in our ``clinal'' model (though we add normally distributed noise to our phenotypes). 
+Of course, it is possible that genotypes indirectly affect individual locations by adjusting organismal fitness and thus habitat selection across spatially varying environments, but we believe that this hypothesis should be tested against a null of stratification bias inflation rather than accepted as true based on GWAS results.
+
+%%%%%% old text from gwas section
+%In a classic quantitative genetics framework phenotypes are the sum of genetic and environmental effects, $P=G+E$, with variance $var(P)=var(G)+var(E)+cov(G,E)$.
+Much of the bias identified in previous studies \citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015}, in which GWAS test statistics reflect population structure rather than phenotype association, arises because of a positive covariance between genotypic and environmental variation (the third term above).
+Because the environment is typically not measured in GWAS of natural populations, this covariation confounds genetic and environmental effects on the phenotype of interest and leads to overestimated effect sizes and deflated $p$ values.
+
+%Over the last twenty years, genome-wide association studies (GWAS) have identified tens of thousands of correlations between genetic variation and phentoypes, both in humans and other species.
 Though originally intended to locate causal alleles for phenotypes controlled by a limited number of sites in the genome, GWAS is now routinely used as the first step in a phenotype-prediction pipeline that sums effect sizes across thousands or millions of variants to generate a "polygenic risk score" (PRS; \citep{Khera2018}).
@@
 %\end{equation}
-For GWAS in structured populations, much of the bias identified in previous studies \citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015}, in which test statistics reflect population structure rather than phenotype association, arises because of the presence of a positive covariance between genotypic and environmental variation.
+%For GWAS in structured populations, much of the bias identified in previous studies \citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015}, in which test statistics reflect population structure rather than phenotype association, arises because of the presence of a positive covariance between genotypic and environmental variation.
 When the environment and genotype covary across space their effects are confounded.
@@
 
-Incorporating PC positions as covariates in the analysis \citep{Price2006} is designed to address this difficulty by regressing out a baseline level of ``average'' differentiation. 
-In essence, a PC-corrected GWAS asks ``what regions of the genome are more associated with this phenotype than the average genome-wide association observed across populations?'' 
-In our simulations, we observed that this procedure can fail under a variety of circumstances. 
-If dispersal is limited and environmental variation is clustered in space (i.e., corner or patchy distributions in our simulations), PCA positions fail to capture the fine-scale spatial structure required to remove all signals of association.
-Conversely, as dispersal increases, PCA loses power to describe population structure before spatial mixing breaks down the relationship between genotype and the environment.
-These effects were observed with all spatially correlated environmental patterns, but were particularly pronounced if environmental effects are concentrated in one region, as was also found by \citet{Mathieson2012}.
-
-Here we simulated a single chromosome with size roughly comparable to one human chromosome.
-If we scale the number of false-positive associations identified in our analyses to a GWAS conducted on whole-genome data from humans, we would expect to see several thousand weak false-positive associations after PC corrections in a population with neighborhood sizes up to at least 1000 (which should include values appropriate for many human populations).
-Notably, very few of the spurious associations we identified would be significant at a conservative Bonferroni-adjusted $p$-value cutoff (see Figure \ref{fig:manhattan_plots}).
-This suggests that GWAS focused on finding strongly associated alleles for traits controlled by a limited number of variants in the genome are likely robust to the impacts of continuous spatial structure.
-However, methods that analyze the combined effects of thousands or millions of weakly associated variants such as polygenic risk scores \citep{Khera2018} are likely to be affected by subtle population structure.
-Indeed as recently identified in studies of gentoype associations for human height in Europe \citep{Berg2018,Sohail2018}, PC regression GWAS in modern human populations do include residual signal of population structure in large-scale GWAS of polygenic traits.
-When attempting to make predictions across populations with different environmental exposures, PRS affected by population structure -- often termed "stratification bias" in the GWAS literature -- can be expected to offer low predictive power, as was shown in a recent study finding lower performance of PRS outside European populations \citep{Martin2019}.
-
-A second point that has received less attention in the literature (but see \cite{Lawson2019}) is the issue of overfitting in GWAS. 
+%A second point that has received less attention in the literature (but see \cite{Lawson2019}) is the issue of overfitting in GWAS.
 If a truly causal allele segregates at different frequencies in different populations, then correcting for population structure in a regression analysis will result in an underestimate of its effect size.
-Though our simulations had no causal alleles, we observed some evidence of this effect in the distribution of $p$-values across the genome (Figure \ref{fig:gwas}D): 
-after PC regression many analyses resulted in $p$-values falling below their expected values from a uniform distribution. 
-This is consistent with a recent empirical study of heritability in human height and body mass index, 
-which found that increasing the number of PC axes used as covariates caused the total proportion of variance explained by SNPs to decline from $\approx0.8$ to $\approx0.75$ \citep{Wainschtein2019}.
-
-In summary, spatial covariation in population structure and the environment confound the interpretation of GWAS $p$-values, and correction using principal components is insufficient to fully separate these signals for polygenic traits under a variety of environmental and population parameter regimes.
-Other GWAS methods may be less sensitive to this confounding, but there is no obvious reason that this should be so.
-One approach to estimating the degree of bias in GWAS caused by population structure is LD score regression \citep{Bulik-Sullivan2015}. 
-Though this approach appears to work well in practice, its interpretation is not always straightforward and it is likely biased by the presence of linked selection \citep{Berg2018}.
-In addition, we observed that in many cases the false-positive SNPs we identified appeared to be concentrated in LD peaks similar to those expected from truly causal sites (Figure \ref{fig:manhattan_plots}).
-We suggest a straightforward alternative for species in which the primary axes of population differentiation is space 
-(note this is likely not the case for many modern human populations): 
-run a GWAS with spatial coordinates as phenotypes and check for $p$-value inflation or significant associations. 
-If significant associations with sample locality are observed after correcting for population structure through PC regression or a kinship matrix, 
-the method is sensitive to false positives induced by spatial structure. 
-This is essentially the approach taken in our ``clinal'' model (though we add normally distributed noise to our phenotypes). 
-Of course, it is possible that genotypes indirectly affect individual locations by adjusting organismal fitness and thus habitat selection across spatially varying environments, but we believe that this hypothesis should be tested against a null of stratification bias inflation rather than accepted as true based on GWAS results.
+Though our simulations had no causal alleles, we observed some evidence of this effect in the distribution of $p$-values across the genome (Figure \ref{fig:gwas}D): after PC regression many analyses resulted in $p$-values falling below their expected values from a uniform distribution, suggesting that the PC regression procedure was leading to overfitting in the model.
+This is consistent with a recent empirical study of heritability in human height and body mass index, which found that increasing the number of PC axes used as covariates caused the total proportion of variance explained by SNPs to decline from $\approx0.8$ to $\approx0.75$ \citep{Wainschtein2019}.
 
@@
 it is inevitably limited by the computational burden of tracking tens or hundreds of thousands of individuals in every generation. 
-In particular, computations required for mate selection and spatial competition scale \cjb{Peter -- what's the correct scaling here?} as the number of individuals within a three-$\sigma$ radius increases.
+In particular, computations required for mate selection and spatial competition scale approximately with the product of the total census size and the neighborhood size and so increases rapidly for large populations and dispersal distances.
 The reverse-time model of continuous space evolution described by \citet{Barton2010} and implemented by \citet{Kelleher2014} 
 allows exploration of parameter regimes with population and landscape sizes more directly comparable to empirical cases like humans. 
-Alternatively, implementation of parallelized simulations may allow progress with forwards-time simulations.
+Alternatively, implementation of parallelized calculations may allow progress with forwards-time simulations.
 
-A mixed approach can combine forward- and reverse-time models, 
-as recently done for a continuous-space Wright-Fisher model by \citet{Lotterhos2019} 
-and a simulation with linked selection by \citet{Buffalo2019}. 
-This allows us to generate short runs of large, realistic simulations in forward time in SLiM \citep{Haller2019} 
-and then ``finish'' the simulations as a coalescent simulation in msprime \citep{Kelleher2016},
-combining the two using an operation called ``recapitation'' \citep{haller2019treesequence}
-As we have seen in this work, a nontrivial step in this
-approach is scaling the variance in reproductive output and generation time across forward- and reverse-time methods. 
-Further development of our understanding of how to merge forward- and reverse-time models 
-is a promising avenue for future research that will be necessary for scaling continuous-space simulations to millions or billions of individuals.
+\cjb{removed section on recapitation here}
+%A mixed approach can combine forward- and reverse-time models, as recently done for a continuous-space Wright-Fisher model by \citet{Lotterhos2019} and a simulation with linked selection by \citet{Buffalo2019}.
+This allows us to generate short runs of large, realistic simulations in forward time in SLiM \citep{Haller2019} and then ``finish'' the simulations as a coalescent simulation in msprime \citep{Kelleher2016}, combining the two using an operation called ``recapitation'' \citep{haller2019treesequence}.
+As we have seen in this work, a nontrivial step in this approach is scaling the variance in reproductive output and generation time across forward- and reverse-time methods.
+Further development of our understanding of how to merge forward- and reverse-time models is a promising avenue for future research that will be necessary for scaling continuous-space simulations to millions or billions of individuals.
 

### 2019-05-16 ms/spaceness.tex
peter edits

@@
 This pattern of ``isolation by distance'' \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}.
-Despite a long history of analytical work describing the genetics of populations distributed across continuous geography (e.g., \cite{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation}), much modern work still describes geographic structure as a set of populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations}.
+Despite a long history of analytical work describing the genetics of populations distributed across continuous geography (e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation}), much modern work still describes geographic structure as a set of discrete populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations}.
 For this reason, most population genetics statistics are interpreted with reference to discrete, well-mixed populations, and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000} with methods that assume these are randomly mating units.
@@
 For instance, if mean viability or fecundity is spatially autocorrelated, then limited geographic dispersal will lead to parent--offspring correlations.
-Furthermore, nearby individuals will be more closely related than an average random pair, so drawing multiple samples from the same area of on the landscape will represent a biased sample of the genetic variation present in the whole population \citep{Stadler2009}.% It has long been appreciated that this model misspecification subjects downstream inferences to bias, but the extent and nature of these effects remain largely uninvestigated.
+Furthermore, nearby individuals will be more closely related than an average random pair, so drawing multiple samples from the same area of the landscape will represent a biased sample of the genetic variation present in the whole population \citep{Stadler2009}.% It has long been appreciated that this model misspecification subjects downstream inferences to bias, but the extent and nature of these effects remain largely uninvestigated.
 
@@
 However, previous investigations of these effects have focused on discrete rather than continuous space models, and the level of isolation by distance at which inference of population size trajectories become biased by structure is not well known.
-Here we test how two methods suitable for use with large samples of individuals -- Stairwayplot \citep{Liu2015} and SMC++ \citep{Terhorst2016} -- perform when applied to populations evolving in continuous space with varying sampling strategies and levels of dispersal.
+Here we test how two methods suitable for use with large samples of individuals -- stairwayplot \citep{Liu2015} and SMC++ \citep{Terhorst2016} -- perform when applied to populations evolving in continuous space with varying sampling strategies and levels of dispersal.
 }
@@
 
-Last, some of the most basic tools of population genetics are summary statistics like $F_{is}$ and Tajima's $D$, which are often interpreted as reflecting the influence of selection or demography on sampled populations \citep{tajima1989statistical}.
+Last, some of the most basic tools of population genetics are summary statistics like $F_{IS}$ and Tajima's $D$, which are often interpreted as reflecting the influence of selection or demography on sampled populations \citep{tajima1989statistical}.
 Statistics like Tajima's $D$ are essentially summaries of the site frequency spectrum, which itself reflects variation in branch lengths and tree structure of the underlying genealogies of sampled individuals.
@@
 We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum. 
-Last, we examine the impacts of evolution in continuous space on genome-wide association studies (GWAS) and identify regions of parameter space under which the results from GWAS may be misleading.
+Last, we examine the impacts of continuous geography on genome-wide association studies (GWAS) and identify regions of parameter space under which the results from GWAS may be misleading.
 
@@
 
-% The best-studied approaches to population genetics in continuous space were developed by \cite{Wright1942,Wright1943} and \cite{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals.
+% The best-studied approaches to population genetics in continuous space were developed by \citet{Wright1942,Wright1943} and \citet{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals.
 The degree to which genetic relationships are geographically correlated
-is determined by the interaction between rates of migration (the chance that one is descended from a distant location)
+depends on the chance that two geographically nearby individuals are close relatives
+-- in modern terms, by the tension between
+migration (the chance that one is descended from a distant location)
 and coalescence (the chance that they share a parent).
-A key early observation Wright model is that this balance is nicely summarized by the ``neighborhood size'',
+A key early observation \citep[by][]{Wright1946} is that this balance is often nicely summarized by the ``neighborhood size'',
 defined to be $N_W = 4 \pi \rho \sigma^2$, where $\sigma$ is the mean parent--offspring distance and $\rho$ is population density. 
 This can be thought of as proportional to the average number of potential mates for an individual (those within distance $2 \sigma)$,
-or the number of potential parents of a randomly chosen individual.
-Empirical estimates of neighborhood size vary hugely across species (see Discussion), but for reference estimates in human populations range from 40 to over 5,000 depending on the population and method of estimation (Table \ref{table:NStable}).
-%\cite{Maruyama1972} found that the rate of decline in genetic diversity in a two-dimensional continuous population approaches the random mating expectation when $\rho\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.
+or the number of potential parents of a randomly chosen individual. Empirical estimates of neighborhood size vary hugely across species
+-- even in human populations,
+estimates range from 40 to over 5,000 depending on the population and method of estimation (Table~\ref{table:NStable}).
+%\citet{Maruyama1972} found that the rate of decline in genetic diversity in a two-dimensional continuous population approaches the random mating expectation when $\rho\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.
 
@@
 However, since the total size of the population is constrained, this introduces interactions between arbitrarily distant individuals,
-which (aside from being implausible) \cite{Felsenstein1975} showed would eventually lead to unrealistic population clumping
+which (aside from being implausible) 
+was shown by \citet{Felsenstein1975} to eventually lead to unrealistic population clumping
 if the range is sufficiently large.
@@
 Among many other important results drawn from this class of ``lattice'' or ``stepping stone'' models, 
-\cite{Rousset1997} showed that the slope of the a linear regression of genetic differentiation ($F_{ST}$) against the logarithm of spatial distance is an estimate of neighborhood size. 
-Although these grid models are likely good approximations of continuous geography in many situations,
+\citet{Rousset1997} showed that the slope of the linear regression of genetic differentiation ($F_{ST}$) against the logarithm of spatial distance is an estimate of neighborhood size. 
+Although these grid models may be good approximations of continuous geography in many situations,
 they do not model demographic fluctuations, and limit investigation of spatial structure below the level of the deme,
@@
 
-We simulated populations using the non-Wright-Fisher module in the program SLiM v3.0 \citep{Haller2019}. 
+We simulated populations using the non-Wright-Fisher module in the program SLiM v3.1 \citep{Haller2019}. 
 Each time step consists of three stages: reproduction, dispersal, and mortality. 
@@
 to modulate the spatial scale of interactions at all three stages by adjusting the standard deviation of the corresponding Gaussian functions. 
-As in previous work \citep{Wright1943,Ringbauer2017}, $\sigma$ as applied in our dispersal step is equal to the mean parent-offspring distance.
+As in previous work \citep{Wright1943,Ringbauer2017}, $\sigma$ is equal to the mean parent-offspring distance.
 
@@
 To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, and density-dependent fitness, 
-we also implemented simulations identical to those above except that mates are selected uniformly random from the population,
+we also implemented simulations identical to those above except that mates are selected uniformly at random from the population,
 and offspring disperse to a uniform random location on the landscape. 
@@
 Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{IS}$), and Tajima's $D$, 
-as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distances \citep{Rousset1997}, 
+as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distance \citep{Rousset1997}, 
 which we summarized as the correlation coefficient between the logarithm of the spatial distance and the proportion of identical base pairs across pairs of individuals.
 
-Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among samples.
+Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among sampled chromosomes.
 The full distribution of lengths of IBS tracts for each pair of chromosomes was first calculated with a custom python function.
@@
 The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks, and so should reflect recent demographic events \citep{chapman2002effect}.
-For a subset of simulations, we also calculated cumulative distributions for IBS tract lengths across pairs of distant (>48 map units) and nearby (<2 map units) individuals.
+For a subset of simulations, we also calculated cumulative distributions for IBS tract lengths across pairs of distant (${}>48$ map units) and nearby (${}<2$ map units) individuals.
 Last, we examined the relationship between allele frequency and the spatial dispersion of an allele by calculating the average distance among individuals carrying each derived allele in a set of simulations representing a range of neighborhood sizes.
@@
 \subsection{Demographic Modeling}
-To assess the impacts of continuous spatial structure on demographic inference we inferred population size histories for all simulations using two approaches: Stairwayplot \citep{Liu2015} and SMC++ \citep{Terhorst2016}.
-Stairwayplot fits its model to a genome-wide estimate of the SFS, while SMC++ also incorporates information about linkage disequilibrium across the genome in a sequentially markovian coalescent framework.
+
+To assess the impacts of continuous spatial structure on demographic inference we inferred population size histories for all simulations using two approaches: stairwayplot \citep{Liu2015} and SMC++ \citep{Terhorst2016}.
+Stairwayplot fits its model to a genome-wide estimate of the SFS, while SMC++ also incorporates linkage information.
+% in a sequentially Markovian coalescent framework.
 For both methods we sampled 20 individuals from all spatial simulations using random, midpoint, and point sampling strategies.
 
-Stairwayplot documentation suggests fitting models across multiple bootstrap replicates drawn from empirical genomic data and taking the median inferred $N_{e}$ per unit time as the best estimate.
-We calculated site frequency spectra in scikit-allel \citep{Miles2017}, generated 100 bootstrap replicates per simulation by resampling over sites, and fit models for all bootstrap samples using default settings in stairwayplot.
+As recommended by its documentation, we used stairwayplot to fit models with multiple bootstrap replicates drawn from empirical genomic data, and took the median inferred $N_{e}$ per unit time as the best estimate.
+We calculated site frequency spectra with scikit-allel \citep{Miles2017}, generated 100 bootstrap replicates per simulation by resampling over sites, and fit models for all bootstrap samples using default settings.
 
-For SMC++ we first converted simulated genotype matrices to VCFs with msprime and then used SMC++'s standard pipeline for preparing input files assuming no polarization error in the SFS.
-We used the first individual in the VCF as the "designated individual" when fitting models, and allowed the program to estimate the recombination rate during optimiziation.
-We fit models using the 'estimate' command rather than the now recommended cross-validation approach because our simulations had only a single contig.
+For SMC++, we first output genotypes as VCF with msprime and then used SMC++'s standard pipeline for preparing input files assuming no polarization error in the SFS.
+We used the first individual in the VCF as the ``designated individual'' when fitting models, and allowed the program to estimate the recombination rate during optimization.
+We fit models using the `estimate' command rather than the now recommended cross-validation approach because our simulations had only a single contig.
 
-To evaluate the performance of these methods we binned simulations by neighborhood size, took a rolling median of inferred $N_{e}$ trajectories across all model fits in a bin for each method and sampling strategy, and compared these values to the mean $N_{e}$ of random-mating simulations estimated from $\Theta_{\pi}$.
+To evaluate the performance of these methods we binned simulations by neighborhood size, took a rolling median of inferred $N_{e}$ trajectories across all model fits in a bin for each method and sampling strategy.
 We also examined how varying levels of isolation by distance impacted the variance of $N_{e}$ estimates by calculating the standard deviation of $N_{e}$ from each best-fit model and plotting these against neighborhood size.
@@
 Phenotype values were set to vary by two standard deviations across the landscape in a rough approximation of the variation seen in height across Europe \ak{need citation}.
-Conceptually our approach is similar to that taken in \cite{Mathieson2012}, though here we model fully continuous spatial variation and compare GWAS output across a range of dispersal distances.
+Conceptually our approach is similar to that taken by \citet{Mathieson2012}, though here we model fully continuous spatial variation and compare GWAS output across a range of dispersal distances.
 
@@
 to a mean that may depend on spatial position.
-We adjust the geographic pattern of mean phenotype to reflect spatially autocorrelated environmental influences on phenotype.
+We adjust the geographic pattern of mean phenotype to create spatially autocorrelated environmental influences on phenotype.
 In the first simulation of \emph{nonspatial} environments, the mean did not change,
@@
 SNPs were first filtered to remove strongly linked sites by calculating LD between all pairs of SNPs in a 200-SNP moving window and dropping one of each pair of sites with an $R^2$ over 0.1.
-The LD-pruned allele count matrix was then centered and all sites scaled to unit variance when conducting the PCA, following recommendations in \cite{Patterson2006}.
+The LD-pruned allele count matrix was then centered and all sites scaled to unit variance when conducting the PCA, following recommendations in \citet{Patterson2006}.
 
@@
 We also examined $p$ values for systemic inflation by estimating the expected values from a uniform distribution (because no SNPs were used when generating phenotypes), plotting observed against expected values for all simulations, and summarizing across simulations by finding the mean $\sigma$ value in each region of quantile-quantile space.
-Results from all analyses were summarized and plotted with the 'ggplot2' \citep{Wickham2016} and ``cowplot'' \citep{Wilke2019} packages in R \citep{Rcore2018}.
+Results from all analyses were summarized and plotted with the ``ggplot2'' \citep{Wickham2016} and ``cowplot'' \citep{Wilke2019} packages in R \citep{Rcore2018}.
 
@@
 Figure \ref{fig:sumstats} shows the effect of varying neighborhood size and sampling strategy on the site frequency spectrum (Figure~\ref{fig:sumstats}A) and several standard population genetic summary statistics (Figure~\ref{fig:sumstats}B).
-\textcolor{red}{Consistent with findings in n-island and stepping stone simulations \citep{Stadler2009}, the SFS shows a significant enrichment of intermediate frequency variants in comparison to the nonspatial expectation.
+\textcolor{red}{Consistent with findings in island and stepping stone simulations \citep{Stadler2009}, the SFS shows a significant enrichment of intermediate frequency variants in comparison to the nonspatial expectation.
 This bias is most pronounced below neighborhood sizes $\leq 100$ and is exacerbated by midpoint and point sampling of individuals (depicted in Figure~\ref{fig:samplemap}).}
@@
 Under midpoint sampling observed heterozygosity is inflated even over the random mating expectation, as a result of the a higher proportion of heterozygotes occurring in the middle of the landscape (Figure~\ref{fig:hetmap}).
-This squares with a report from \cite{Shirk2014} who observed a similar excess of heterozygosity in the middle of the landscape when simulating under a lattice model.
+This squares with a report from \citet{Shirk2014} who observed a similar excess of heterozygosity in the middle of the landscape when simulating under a lattice model.
 
@@
 
+%%%%%%%%%%%%%%%
 \subsubsection{IBS tracts and correlations with geographic distance}
+
 We next turn our attention to the effect of geographic distance on haplotype block length sharing, summarized for sets of nearby and distant individuals in Figure \ref{fig:ibs}.
@@
 This reflects the faster spatial mixing of populations with higher dispersal, which breaks down the correlation between the IBS tract length distribution and geographic distance.
-This can also be seen in the bottom row of Figure \ref{fig:sumstats}B, where the correlation coefficients between the summaries of the IBS tract length distribution (the mean, skew, and count of tracts over $1\times10^6bp$) and geographic distance approaches 0 as neighborhood size increases.
+This can also be seen in the bottom row of Figure \ref{fig:sumstats}B, where the correlation coefficients between the summaries of the IBS tract length distribution (the mean, skew, and count of tracts over $10^6$bp) and geographic distance approaches 0 as neighborhood size increases.
 
 %Last, in figure \ref{fig:ibs} we can see that most of the difference both between nearby and distant individuals and across different neighborhood sizes is found in the upper tail of the IBS tract length distribution.
-This is consistent with the methods for inferring dispersal distances described in \cite{Ringbauer2017} and \cite{Baharian2016}, which focus on long IBS blocks rather than the full distribution.
+This is consistent with the methods for inferring dispersal distances described in \citet{Ringbauer2017} and \citet{Baharian2016}, which focus on long IBS blocks rather than the full distribution.
 
-The patterns observed for correlations of IBS tract lengths with geographic distance are similar to those observed in the more familiar regression of allele frequency measures such as $D_{xy}$ (i.e.
-"genetic distance") or $F_{st}$ against geographic distance \citep{Rousset1997}.
-$D_{xy}$ is positively correlated with the geographic distance between the individuals, and the strength of this correlation declines as dispersal increases (Figure \ref{fig:sumstats}B), as expected under theory \citep{Wright1943,Rousset1997}.
+The patterns observed for correlations of IBS tract lengths with geographic distance are similar to those observed in the more familiar regression of allele frequency measures such as $D_{xy}$ (i.e., ``genetic distance'') or $F_{ST}$ against geographic distance \citep{Rousset1997}.
+$D_{xy}$ is positively correlated with the geographic distance between the individuals, and the strength of this correlation declines as dispersal increases (Figure \ref{fig:sumstats}B), as expected \citep{Wright1943,Rousset1997}.
 This relationship is very similar across random and point sampling strategies, but is weaker for midpoint sampling, perhaps due to a dearth of long-distance comparisons.
 In much of empirical population genetics a regression of genetic differentiation against spatial distance is a de-facto metric of the significance of isolation by distance.
-The similar behavior of moments of the pairwise distribution of IBS tract lengths shows why haplotype block sharing has recently emerged as a promising source of information on spatial demography through methods described in \cite{Ringbauer2017} and \cite{Baharian2016}.
+The similar behavior of moments of the pairwise distribution of IBS tract lengths shows why haplotype block sharing has recently emerged as a promising source of information on spatial demography through methods described in \citet{Ringbauer2017} and \citet{Baharian2016}.
 
@@
 
+%%%%%%%%%%%%%%%%%%%
 \subsubsection{Spatial distribution of allele copies}
+
 Mutations occur in individuals and spread geographically over time.
@@
 For populations with large neighborhood sizes we found that even very low frequency alleles can be found across the full landscape, whereas in populations with low neighborhood sizes the relationship between distance among allele copies and their frequency is quite strong.
-This is the basic process underlying \cite{Novembre2009}'s method for estimating dispersal distances based on the distribution of low frequency alleles, and also generates the greater degree of bias in GWAS effect sizes for low frequency alleles identified in \cite{Mathieson2012}.
+This is the basic process underlying \citeauthor{Novembre2009}'s \citeyearpar{Novembre2009} method for estimating dispersal distances based on the distribution of low frequency alleles, and also generates the greater degree of bias in GWAS effect sizes for low frequency alleles identified in \citet{Mathieson2012}.
 
+%%%%%%%%%%%%%%%%%
 \subsection{Effects of Space on Demographic Inference}
+
 One of the most important uses for population genetic data is inferring demographic history of populations.
@@
 Does this variation lead to different inferences of past population sizes?
-To ask this we inferred population size histories from samples drawn from our simulated populations with two approaches: Stairwayplot \citep{Liu2015}, which uses a genome-wide estimate of the SFS, and SMC++ \citep{Terhorst2016}, which incorporates information on both the SFS and linkage disequilibrium across the genome.
+To ask this we inferred population size histories from samples drawn from our simulated populations with two approaches: stairwayplot \citep{Liu2015}, which uses a genome-wide estimate of the SFS, and SMC++ \citep{Terhorst2016}, which incorporates information on both the SFS and linkage disequilibrium across the genome.
 
+\plr{median of how many? should be fig S4 here instead?}
 Figure \ref{fig:demography}A shows the median inferred population size histories from each method, grouped by neighborhood size and sampling strategy.
@@
 As expected, spatial autocorrelation in the environment causes spurious associations 
-across much of the genome if no correction for genetic relatedness among samples is performed (Figure \ref{fig:gwas}, figure \ref{fig:manhattan_plots}). 
+across much of the genome if no correction for genetic relatedness among samples is performed (Figures \ref{fig:gwas} and \ref{fig:manhattan_plots}). 
 This effect is particularly strong for clinal and corner environments, 
@@
 cause fewer false positives overall but still produce spurious associations at roughly 10\% of sites at the lowest neighborhood sizes. 
-Interestingly we also observed a small number of false-positive associations in roughly 3\% of analyses on simulations with nonspatial environments, both with and without PC covariates included in the regression.
+Interestingly we also observed a small number of false positives in roughly 3\% of analyses on simulations with nonspatial environments, both with and without PC covariates included in the regression.
 
@@
 and it is common practice to control for this by including principal components (PCs) as covariates to control for these effects.
-This mostly works in our simulations -- after doing this, the vast majority of SNPs no longer surpass a 5\% FDR significance threshold. 
+This mostly works in our simulations -- after doing this, the vast majority of SNPs no longer surpass a significance threshold chosen to have a 5\% false discovery rate (FDR).
 However, a substantial number of SNPs -- up to 1.5\% at the lowest dispersal distances -- still surpass this threshold (and thus would be false positives in a GWAS),
@@
 At neighborhood sizes larger than 500, up to 0.31\% of SNPs were significant for corner and clinal environments. 
-Given an average of 132,000 SNPs across simulations after MAF filtering, this translates to up to 382 false-positive associations. 
-In most cases the $p$ values for these associations were significant after FDR correction but would not pass the threshold for significance under the more conservative Bonferroni correction (see example manhattan plots in figure \ref{fig:manhattan_plots}).
+Given an average of 132,000 SNPs across simulations after MAF filtering, this translates to up to 382 false-positive associations; for human-sized genomes, this number would be much larger.
+In most cases the $p$ values for these associations were significant after FDR correction but would not pass the threshold for significance under the more conservative Bonferroni correction (see example Manhattan plots in figure \ref{fig:manhattan_plots}).
 
 Clinal environments cause an interesting pattern in false positives after PC correction: at low neighborhood sizes the correction removes nearly all significant associations, 
-but at neighborhood sizes above $\approx250$ the proportion of significant SNPs increases to up to 0.4\% (Figure~\ref{fig:gwas}). 
+but at neighborhood sizes above roughly 250 the proportion of significant SNPs increases to up to 0.4\% (Figure~\ref{fig:gwas}). 
 This may be due to a loss of descriptive power of the PCs
@@
 For clinal environments, $-\log_{10}(p)$ values are most inflated when neighborhood sizes are large, consistent with the pattern observed in the count of significant associations after PC regression.
-In contrast corner and patchy environments cause the greatest inflation in $-\log_{10}(p)$ at neighborhood sizes $<100$, which likely reflects the inability of PCA to account for fine-scale structure caused by very limited dispersal.
+In contrast corner and patchy environments cause the greatest inflation in $-\log_{10}(p)$ at neighborhood sizes less than 100, which likely reflects the inability of PCA to account for fine-scale structure caused by very limited dispersal.
 Finally, we observed that PC regression appears to overfit to some degree for all phenotype distributions, visible in Figure \ref{fig:gwas}D as points falling below the 1:1 line.
@@
 These effects are strongest when neighborhood sizes are below 100, but in combination with the effects of nonrandom sampling they can persist up to neighborhood sizes of at least 1000 (e.g., inflation in Tajima's $D$ and observed heterozygosity under midpoint sampling).
-Under random sampling the general pattern is similar to expectations of the original analytic model of \cite{Wright1943}, which predicts that populations with neighborhood sizes under 100 will differ substantially from random mating, while those above 10,000 will be nearly indistinguishable from panmixia.
+If samples are chosen uniformly from across space, 
+the general pattern is similar to expectations of the original analytic model of \citet{Wright1943}, which predicts that populations with neighborhood sizes under 100 will differ substantially from random mating,
+while those above 10,000 will be nearly indistinguishable from panmixia.
 
@@
 in a randomly mating population the spatial distribution of sampling effort has no effect on estimates of genetic variation (Table \ref{table:sumstats}) , 
-but when dispersal is limited sampling strategy can compound spatial patterns in the underlying genealogy and create pervasive impacts on all downstream genetic analyses \textcolor{red}{(see also \cite{Stadler2009})}.
+but when dispersal is limited sampling strategy can compound spatial patterns in the underlying genealogy and create pervasive impacts on all downstream genetic analyses \textcolor{red}{(see also \citet{Stadler2009})}.
 In most species,
@@
 
-However, the largest sampling effects we observed occurred in our ``midpoint'' sampling strategy. 
+However, we observed the largest sampling effects using ``midpoint'' sampling.
 This model is meant to reflect a bias in sampling effort towards the middle of a species' range. 
@@
 Because midpoint sampling provides limited spatial resolution it dramatically reduces the magnitude of observed correlations between spatial and genetic distances. 
-More surprisingly, midpoint sampling also leads to strongly positive Tajima's $D$ and an inflation in the proportion of heterozygous individuals in the sample -- similar to the effect of sampling a single deme in an n-island model as reported in \citep{Stadler2009}.
+More surprisingly, midpoint sampling also leads to strongly positive Tajima's $D$ and an inflation in the proportion of heterozygous individuals in the sample -- similar to the effect of sampling a single deme in an island model as reported in \citep{Stadler2009}.
 This increase in observed heterozygosity appears to reflect the effects of range edges, which are a fundamental facet of spatial genetic variation. 
@@
 
-In summary, empirical researchers should collect individuals from across as much of the species' range as practical,
-choosing samples at a variety of spatial scales.
+In summary, 
+we recommend that empirical researchers collect individuals from across as much of the species' range as practical,
+choosing samples separated by a range of spatial scales.
 Many summary statistics are designed for well-mixed populations,
@@
 \textcolor{red}{Previous studies have found that population structure and nonrandom sampling can create spurious signals of population bottlenecks when attempting to infer demographic history with microsatellite variation, summary statistics, or runs of homozygosity \citep{Chikhi2010,Stadler2009,Ptak2002,Mazet2015}.
-Here we found that methods that infer detailed population trajectories through time based on the SFS and patterns of LD across the genome are also subject to this bias, with some combinations of dispersal and sampling strategy inferring deep recent population bottlenecks and overestimating ancient $N_{e}$ by a around a factor of 2.
-However we were surprised to see that both Stairwayplot and SMC++ can tolerate relatively strong isolation by distance -- i.e.
-neighborhood sizes of 20 -- and still perform well when averaging results across multiple simulations.
-Inference in populations with neighborhood sizes over 20 was relatively accurate unless samples were concentrated in the middle of the range (Figure \ref{fig:demography}).
+Here we found that methods that infer detailed population trajectories through time based on the SFS and patterns of LD across the genome are also subject to this bias, with some combinations of dispersal and sampling strategy systematically inferring deep recent population bottlenecks and overestimating ancient $N_{e}$ by a around a factor of 2.
+We were surprised to see that both stairwayplot and SMC++ can tolerate relatively strong isolation by distance -- i.e., neighborhood sizes of 20 -- and still perform well when averaging results across multiple simulations.
+Inference in populations with neighborhood sizes over 20 was relatively unbiased unless samples were concentrated in the middle of the range (Figure \ref{fig:demography}).
+Although median demography estimates across many independent simulations were fairly accurate,
+empirical work has only a single estimate to work with,
+and individual model fits (Figure \ref{fig:demography_supplement})
+suggest that spuriously inferred population size changes and bottlenecks are common, especially at small neighborhood sizes.
 As we will discuss below, most empirical estimates of neighborhood size, including all estimates for human populations, are large enough that population size trajectories inferred by these approaches should not be strongly affected by spatial biases created by dispersal in continuous landscapes.
-In contrast, \cite{Mazet2015} found that varying migration rates through time could create strong biases in inferred population trajectories from an $n$-island model with parameters relevant for human history, suggesting that changes in migration rates through time are more likely to drive variation in inferred $N_{e}$ than isolation by distance.}
+In contrast, \citet{Mazet2015} found that varying migration rates through time could create strong biases in inferred population trajectories from an $n$-island model with parameters relevant for human history, suggesting that changes in migration rates through time are more likely to drive variation in inferred $N_{e}$ than isolation by distance.}
 
-\textcolor{red}{Between the two methods we tested, we found that SMC++ was more robust to spatial biases than Stairwayplot, with the worst-affected bins underestimating recent populations by roughly half rather than nearly 10-fold as with Stairwayplot.} Though this degree of variation in population size is certainly meaningful in an ecological context, it is relatively minor in population genetic terms.
+\textcolor{red}{We found that SMC++ was more robust to the effects of space than stairwayplot, underestimating recent populations by roughly half in the worst time periods rather than nearly 10-fold as with stairwayplot.} Though this degree of variation in population size is certainly meaningful in an ecological context, it is relatively minor in population genetic terms.
 A more worrying pattern was the high level of variance in inferred $N_{e}$ trajectories for individual model fits using these methods, which was highest in simulations with the smallest neighborhood size (Figure \ref{fig:demography}, Figure \ref{fig:demography_supplement}).
@@
 
-In general however, our analysis suggests that many empirical analyses of population size history using methods like SMC++ are robust to error caused by spatial structure within continuous landscapes.
-Inferences drawn from static SFS-based methods like Stairwayplot should be treated with caution when there are signs of isolation by distance in the underlying data (for example, if a regression of $F_{st}$ against the logarithm of geographic distance has a significantly positive slope), and in particular an inference of population bottlenecks in the last 1000 years should be discounted if sampling is clustered, but estimates of deeper time patterns are likely to be fairly accurate.
-The biases in the SFS and haplotype structure identified above (see also \citep{Wakeley1999,Chikhi2010,Stadler2009}) are apparently small enough that they fall within the range of variability regularly inferred by these approaches, at least on datasets of the size we simulated.
+Our analysis suggests that many empirical analyses of population size history using methods like SMC++ are robust to error caused by spatial structure within continuous landscapes.
+Inferences drawn from static SFS-based methods like stairwayplot should be treated with caution when there are signs of isolation by distance in the underlying data (for example, if a regression of $F_{ST}$ against the logarithm of geographic distance has a significantly positive slope), and in particular an inference of population bottlenecks in the last 1000 years should be discounted if sampling is clustered, but estimates of deeper time patterns are likely to be fairly accurate.
+The biases in the SFS and haplotype structure identified above \citep[see also][]{Wakeley1999,Chikhi2010,Stadler2009} are apparently small enough that they fall within the range of variability regularly inferred by these approaches, at least on datasets of the size we simulated.
 
@@
 If environmental factors affecting the phenotype of interest also vary over space, then groups of individuals in different regions will both experience different environments and have different allele frequencies at many sites in the genome.
-In this scenario an uncorrected GWAS will infer genetic associations with a purely environmental phenotype at any site in the genome that is differentiated over space, and the relative degree of bias will be a function of the degree of covariation in allele frequencies and the environment (i.e.
-Figure \ref{fig:gwas}C, bottom panel).
+In this scenario an uncorrected GWAS will infer genetic associations with a purely environmental phenotype at any site in the genome that is differentiated over space, and the relative degree of bias will be a function of the degree of covariation in allele frequencies and the environment (i.e., Figure \ref{fig:gwas}C, bottom panel).
 This pattern is now well known and has been demonstrated in a variety of simulation and empirical contexts \citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015,Berg2018,Sohail2018}.}
@@
 
-Here we simulated a single chromosome with size roughly comparable to one human chromosome.
+In this work we simulated a single chromosome with size roughly comparable to one human chromosome.
 If we scale the number of false-positive associations identified in our analyses to a GWAS conducted on whole-genome data from humans, we would expect to see several thousand weak false-positive associations after PC corrections in a population with neighborhood sizes up to at least 1000 (which should include values appropriate for many human populations).
@@
 However, methods that analyze the combined effects of thousands or millions of weakly associated variants such as polygenic risk scores \citep{Khera2018} are likely to be affected by subtle population structure.
-Indeed as recently identified in studies of gentoype associations for human height in Europe \citep{Berg2018,Sohail2018}, PC regression GWAS in modern human populations do include residual signal of population structure in large-scale analyses of polygenic traits.
-When attempting to make predictions across populations with different environmental exposures, PRS affected by population structure can be expected to offer low predictive power, as was shown in a recent study finding lower performance of PRS outside European populations \citep{Martin2019}.
+Indeed as recently identified in studies of genotype associations for human height in Europe \citep{Berg2018,Sohail2018}, PC regression GWAS in modern human populations do include residual signal of population structure in large-scale analyses of polygenic traits.
+When attempting to make predictions across populations with different environmental exposures, polygenic risk scores affected by population structure can be expected to offer low predictive power, as was shown in a recent study finding lower performance outside European populations \citep{Martin2019}.
 
-In summary, spatial covariation in population structure and the environment confound the interpretation of GWAS $p$-values, and correction using principal components is insufficient to fully separate these signals for polygenic traits under a variety of environmental and population parameter regimes.
+In summary, spatial covariation in population structure and the environment confounds the interpretation of GWAS $p$-values, and correction using principal components is insufficient to fully separate these signals for polygenic traits under a variety of environmental and population parameter regimes.
 Other GWAS methods may be less sensitive to this confounding, but there is no obvious reason that this should be so.
@@
 \textcolor{red}{In addition, we observed that in many cases the false-positive SNPs we identified appeared to be concentrated in LD peaks similar to those expected from truly causal sites (Figure \ref{fig:manhattan_plots}), which may confound LD score regression.}
+
 We suggest a straightforward alternative for species in which the primary axes of population differentiation is space 
@@
 %Over the last twenty years, genome-wide association studies (GWAS) have identified tens of thousands of correlations between genetic variation and phentoypes, both in humans and other species.
-Though originally intended to locate causal alleles for phenotypes controlled by a limited number of sites in the genome, GWAS is now routinely used as the first step in a phenotype-prediction pipeline that sums effect sizes across thousands or millions of variants to generate a "polygenic risk score" (PRS; \citep{Khera2018}).
+Though originally intended to locate causal alleles for phenotypes controlled by a limited number of sites in the genome, GWAS is now routinely used as the first step in a phenotype-prediction pipeline that sums effect sizes across thousands or millions of variants to generate a ``polygenic risk score'' (PRS; \citep{Khera2018}).
 The most common method for estimating these effect sizes, which we followed in our analyses of simulated data here, is to regress the count of derived alleles at a site against individual phenotypes, taking the slope of this regression as an estimate of the effect of the allele on the phenotype.
-As recently reviewed by \cite{Visscher2019}, this is exactly the approach outlined by \cite{Fisher1918} at the dawn of quantitative genetics.
+As recently reviewed by \citet{Visscher2019}, this is exactly the approach outlined by \citet{Fisher1918} at the dawn of quantitative genetics.
 
@@
 
-%A second point that has received less attention in the literature (but see \cite{Lawson2019}) is the issue of overfitting in GWAS.
+%A second point that has received less attention in the literature (but see \citet{Lawson2019}) is the issue of overfitting in GWAS.
 If a truly causal allele segregates at different frequencies in different populations, then correcting for population structure in a regression analysis will result in an underestimate of its effect size.
@@
 Many more species occur in a middle range of neighborhood sizes between 100 and 1000 -- a range in which spatial processes play a minor role in our analyses under random spatial sampling but are important when sampling of individuals in space is clustered.
-Last, many species likely have neighborhood sizes much larger than we simulated, including modern humans in NE Europe \citep{Ringbauer2017}.
+Last, many species likely have neighborhood sizes much larger than we simulated, including modern humans in northeastern Europe \citep{Ringbauer2017}.
 For these species demographic inference and summary statistics are likely to reflect minimal bias from spatial effects as long as dispersal is truly continuous across the landscape.
@@
 As we have shown, a large number of population genetic summary statistics contain information about spatial population processes.
-We imagine that combinations of such summaries might be sufficient for the construction of supervised machine learning regressors (e.g., \cite{Schrider2018}) for the accurate estimation of dispersal from genetic data.
-Indeed, \cite{Ashander2018} found that inverse interpolation on a vector of summary statistics provided a powerful method of estimating dispersal distances.
+We imagine that combinations of such summaries might be sufficient for the construction of supervised machine learning regressors \citep[e.g.,][]{Schrider2018} for the accurate estimation of dispersal from genetic data.
+Indeed, \citet{Ashander2018} found that inverse interpolation on a vector of summary statistics provided a powerful method of estimating dispersal distances.
 Expanding this approach to include the haplotype-based summary statistics studied here and applying machine learning regressors built for general inference of nonlinear relationships from high-dimensional data may allow precise estimation of spatial parameters under a range of complex models.
@@
 In addition, spatially varying selection will create allele frequency variation over space that may mimic isolation by distance.
-Indeed, a series of field studies described in \cite{Schemske2007} found that in Wright's original empirical example of isolation by distance, the flowering plant \textit{Linanthus parryae}, patterns of flower color differentiation over space primarily reflect temporal and spatial variation in selection rather than limited dispersal.
-Studies simulating selection and dispersal interacting in space (e.g., \cite{Ralph2010}) and testing for identifiability of inferred dispersal or selective parameters may offer new insight into the extent of our ability to accurately infer evolutionary processes in real systems.
+Indeed, a series of field studies described in \citet{Schemske2007} found that in Wright's original empirical example of isolation by distance, the flowering plant \textit{Linanthus parryae}, patterns of flower color differentiation over space primarily reflect temporal and spatial variation in selection rather than limited dispersal.
+Studies simulating selection and dispersal interacting in space (e.g., \citet{Ralph2010}) and testing for identifiability of inferred dispersal or selective parameters may offer new insight into the extent of our ability to accurately infer evolutionary processes in real systems.
 
-One facet of spatial variation that we did not address in this study is the confounding of dispersal and population density implicit in the definition of Wright's Neighborhood Size.
-Our simulations were run under constant densities, but \cite{Ringbauer2017}'s approach to demographic inference in space suggests that density and dispersal can in some cases be estimated separately from genetic data.
+One facet of spatial variation that we did not address in this study is the confounding of dispersal and population density implicit in the definition of Wright's neighborhood size.
+Our simulations were run under constant densities, but \citet{Ringbauer2017}'s approach to demographic inference in space suggests that density and dispersal can in some cases be estimated separately from genetic data.
 Much additional work remains to be done to better understand how these parameters interact to shape genetic variation in continuous space, which we leave to future studies.
@@
 \section{Acknowledgements}
-We thank Brandon Cooper, and others for reading and thinking about this manuscript. 
+We thank Brandon Cooper, Matt Hahn and others for reading and thinking about this manuscript. 
 We thank the Hearth for having such good, nearby coffee. 
@@
 observed heterozygosity & random mating & 0.531540 & 0.433230 \\ 
-$F_{is}$ & random mating & 0.474790 & 0.785730 \\ 
+$F_{IS}$ & random mating & 0.474790 & 0.785730 \\ 
 $mean(D_{xy})$ & random mating & 0.997770 & 0.996510 \\ 
@@
 observed heterozygosity & spatial & \textbf{0.000000} & \textbf{0.000000} \\ 
-$F_{is}$ & spatial & \textbf{0.000000} & \textbf{0.000120} \\ 
+$F_{IS}$ & spatial & \textbf{0.000000} & \textbf{0.000120} \\ 
 $mean(D_{xy})$ & spatial & \textbf{0.025390} & \textbf{0.012910} \\ 

### 2019-06-03 ms/spaceness.tex
biorxiv ms

@@
 
-\documentclass[11pt,twoside,lineno]{preprint}
+\documentclass[11pt,twoside,lineno,hidelinks]{preprint}
 \usepackage{amsmath}
@@
 
-\textcolor{red}{Two areas in which spatial structure may be particularly important are demographic inference and genome-wide association studies (GWAS).
+Two areas in which spatial structure may be particularly important are demographic inference and genome-wide association studies (GWAS).
 Previous work has found that discrete population structure can create false signatures of population bottlenecks when attempting to infer demographic histories from microsatellite variation \citep{Chikhi2010}, statistics summarizing the site frequency spectrum (SFS) \citep{Ptak2002,Stadler2009,Onge2012}, or runs of homozygosity in a single individual \citep{Mazet2015}.
@@
 Here we test how two methods suitable for use with large samples of individuals -- stairwayplot \citep{Liu2015} and SMC++ \citep{Terhorst2016} -- perform when applied to populations evolving in continuous space with varying sampling strategies and levels of dispersal.
-}
 
@@
 that incorporates overlapping generations, local dispersal of offspring, and density-dependent survival. 
-We simulate chromosome-scale genomic data in tens of thousands of individuals from parameter regimes relevant to common subjects of population genetic investigation such as humans and Drosophila, and output the full genealogy and recombination history of all final-generation individuals.
+We simulate chromosome-scale genomic data in tens of thousands of individuals from parameter regimes relevant to common subjects of population genetic investigation such as humans and \textit{Drosophila}, and output the full genealogy and recombination history of all final-generation individuals.
 We use these simulations to test how sampling strategy interacts with geographic population structure to cause systematic variation in population genetic summary statistics 
@@
 and coalescence (the chance that they share a parent).
-A key early observation \citep[by][]{Wright1946} is that this balance is often nicely summarized by the ``neighborhood size'',
+A key early observation by Wright \citep{Wright1946} is that this balance is often nicely summarized by the ``neighborhood size'',
 defined to be $N_W = 4 \pi \rho \sigma^2$, where $\sigma$ is the mean parent--offspring distance and $\rho$ is population density. 
@@
 
-The first approach to modeling continuously-distributed populations
+The first approach to modeling continuously distributed populations
 was to endow individuals in a Wright-Fisher model with locations in continuous space.
@@
 if the range is sufficiently large.
-Another method for modeling spatial populations is to assume the existence of a grid of discrete randomly-mating populations connected by migration, 
+Another method for modeling spatial populations is to assume the existence of a grid of discrete randomly mating populations connected by migration, 
 thus enforcing regular population density by edict.
@@
 
-To avoid questionable assumptions, we here used forwards-time, individual-based simulations.
+To avoid questionable assumptions, we here used forward-time, individual-based simulations.
 By scaling the probability of survival in each timestep to local population density we shift reproductive output towards low-density regions, which prevents populations from clustering.
@@
 Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
-in both the $x$ and $y$ coordinates, reflected to stay within the species range.
+in both the $x$ and $y$ coordinates.
 Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
@@
 Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{IS}$), and Tajima's $D$, 
-as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distance \citep{Rousset1997}, 
+as well as an isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distance analogous to \cite{Rousset1997}'s approach, 
 which we summarized as the correlation coefficient between the logarithm of the spatial distance and the proportion of identical base pairs across pairs of individuals.
@@
 SNPs with a minor allele frequency less than 0.5\% were excluded from this analysis.
-Phenotype values were set to vary by two standard deviations across the landscape in a rough approximation of the variation seen in height across Europe \ak{need citation}.
+Phenotype values were set to vary by two standard deviations across the landscape in a rough approximation of the variation seen in height across Europe \citep{Turchin2012,garcia2006,garcia2007}.
 Conceptually our approach is similar to that taken by \citet{Mathieson2012}, though here we model fully continuous spatial variation and compare GWAS output across a range of dispersal distances.
@@
 
-Generation time similarly shows complex behavior with respect to neighborhood sizes, and varies between 5.2 and 4.9 timesteps per genration across the parameter range explored. 
+Generation time similarly shows complex behavior with respect to neighborhood sizes, and varies between 5.2 and 4.9 timesteps per generation across the parameter range explored. 
 % This occurs because individuals may reproduce more than once over their lifespan, and the lifespan of an individual is determined by density-dependent competition the model. 
@@
 Indeed, mathematical results suggest that genetic variation in some spatial models should be well-approximated by a Wright-Fisher population if neighborhood size is large and all samples are geographically widely separated \citep{wilkins2004separation,zahle2005stepping}.
-However, the behavior of most common population genetic summary statistics \textcolor{red}{other than Tajima's $D$ \citep{Stadler2009}} has not yet been described in realistic geographic models.
+However, the behavior of most common population genetic summary statistics other than Tajima's $D$ \citep{Stadler2009} has not yet been described in realistic geographic models.
 Moreover, as we will show, spatial sampling strategies can affect summaries of variation at least as strongly as the underlying population dynamics.
@@
 Figure \ref{fig:sumstats} shows the effect of varying neighborhood size and sampling strategy on the site frequency spectrum (Figure~\ref{fig:sumstats}A) and several standard population genetic summary statistics (Figure~\ref{fig:sumstats}B).
-\textcolor{red}{Consistent with findings in island and stepping stone simulations \citep{Stadler2009}, the SFS shows a significant enrichment of intermediate frequency variants in comparison to the nonspatial expectation.
-This bias is most pronounced below neighborhood sizes $\leq 100$ and is exacerbated by midpoint and point sampling of individuals (depicted in Figure~\ref{fig:samplemap}).}
+Consistent with findings in island and stepping stone simulations \citep{Stadler2009}, the SFS shows a significant enrichment of intermediate frequency variants in comparison to the nonspatial expectation.
+This bias is most pronounced below neighborhood sizes $\leq 100$ and is exacerbated by midpoint and point sampling of individuals (depicted in Figure~\ref{fig:samplemap}).
 Reflecting this, Tajima's $D$ is quite positive in the same situations (Figure~\ref{fig:sumstats}B).
@@
 One of the most commonly used summaries of variation is Tajima's summary of nucleotide divergence, $\theta_{\pi}$, calculated as the mean density of nucleotide differences averaged across pairs of samples.
-As can be seen in Figure \ref{fig:sumstats}B, $\theta_{\pi}$ differs as much as nearly three-fold between the random mating and spatial models.
-$\theta_{\pi}$ using each sampling strategy approaches the random mating expectation at its own rate, but by a neighborhood size of around 100 all models are equivalent.
-The differences between spatial and random mating simulations are much greater than expected from differences in census size divided by variance in offspring number.
-This likely occurs because $\theta_{\pi}$ is a measure of mean time to most recent common ancestor between two samples, and at small values of $\sigma$, the time for dispersal to mix ancestry across the range exceeds the mean nonspatial coalescent time.
-(For instance, at the smallest value of $\sigma=0.2$, the range is 250 dispersal distances wide, and since the location of a diffusively moving lineage after $k$ generations has variance $k \sigma^2$, it takes around $250^2 = 62500$ generations to mix across the range, which is roughly ten times larger than the random mating effective population size.) Interestingly, the effect of sampling strategy is reversed relative to that observed in Tajima's D -- midpoint sampling reaches random mating expectations around neighborhood size 50, while random sampling is inflated until around neighborhood size 100.
+As can be seen in Figure \ref{fig:sumstats}B, $\theta_{\pi}$ in the spatial model is inflated by up to three-fold relative to the random mating model.
+This pattern is opposite the expectation from census population size (Figure \ref{fig:genparams}), because the spatial model has \emph{lower} census size than the random mating model at neighborhood sizes less than 100.
+Differences between these models likely occur because $\theta_{\pi}$ is a measure of mean time to most recent common ancestor between two samples, and at small values of $\sigma$, the time for dispersal to mix ancestry across the range exceeds the mean coalescent time under random mating.
+(For instance, at the smallest value of $\sigma=0.2$, the range is 250 dispersal distances wide, and since the location of a diffusively moving lineage after $k$ generations has variance $k \sigma^2$, it takes around $250^2 = 62500$ generations to mix across the range, which is roughly ten times larger than the random mating effective population size).
+$\theta_{\pi}$ using each sampling strategy approaches the random mating expectation at its own rate, but by a neighborhood size of around 100 all models are roughly equivalent.
+Interestingly, the effect of sampling strategy is reversed relative to that observed in Tajima's D -- midpoint sampling reaches random mating expectations around neighborhood size 50, while random sampling is inflated until around neighborhood size 100.
 
@@
 Under midpoint sampling observed heterozygosity is inflated even over the random mating expectation, as a result of the a higher proportion of heterozygotes occurring in the middle of the landscape (Figure~\ref{fig:hetmap}).
-This squares with a report from \citet{Shirk2014} who observed a similar excess of heterozygosity in the middle of the landscape when simulating under a lattice model.
+This echoes a report from \citet{Shirk2014} who observed a similar excess of heterozygosity in the middle of the landscape when simulating under a lattice model.
 
@@
 
-\plr{median of how many? should be fig S4 here instead?}
-Figure \ref{fig:demography}A shows the median inferred population size histories from each method, grouped by neighborhood size and sampling strategy.
+Figure \ref{fig:demography}A shows the median inferred population size histories from each method across all simulations, grouped by neighborhood size and sampling strategy.
 In general these methods tend to slightly overestimate ancient population sizes and infer recent population declines when neighborhood sizes are below 20 and sampling is spatially clustered (Figure \ref{fig:demography}A, Figure \ref{fig:demography_supplement}).
@@
 For stairwayplot we found that many runs infer dramatic population bottlenecks in the last 1,000 generations when sampling is spatially concentrated, resulting in ten-fold or greater underestimates of recent population sizes.
-However SMC++ appeared more robust to this error, with runs on point- and midpoint-sampled simulations at the lowest neighborhood sizes underestimating recent population sizes by roughly half and those on randomly-sampled simulations showing little error.
+However SMC++ appeared more robust to this error, with runs on point- and midpoint-sampled simulations at the lowest neighborhood sizes underestimating recent population sizes by roughly half and those on randomly sampled simulations showing little error.
 Above neighborhood sizes of around 100, both methods performed relatively well when averaging across results from multiple simulations.
@@
 and it is common practice to control for this by including principal components (PCs) as covariates to control for these effects.
-This mostly works in our simulations -- after doing this, the vast majority of SNPs no longer surpass a significance threshold chosen to have a 5\% false discovery rate (FDR).
+This mostly works in our simulations -- after incorporating the first ten PC axes as covariates, the vast majority of SNPs no longer surpass a significance threshold chosen to have a 5\% false discovery rate (FDR).
 However, a substantial number of SNPs -- up to 1.5\% at the lowest dispersal distances -- still surpass this threshold (and thus would be false positives in a GWAS),
@@
 As neighborhood size increases a PCA explains less of the total variation in the data (\textbf{B}).
-Spatially correlated environmental factors cause false positives at a large proportion of SNPs, which is partially but not entirely corrected by adding PC positions as covariates (\textbf{C}).
+Spatially correlated environmental factors cause false positives at a large proportion of SNPs, which is partially but not entirely corrected by adding the first 10 PC coordinates as covariates (\textbf{C}).
 Quantile-quantile plots in \textbf{D} show inflation of $-\log_{10}(p)$ after PC correction across all simulations and environmental distributions, with colors scaled by the median neighborhood size in each region of q-q space.
@@
 in a randomly mating population the spatial distribution of sampling effort has no effect on estimates of genetic variation (Table \ref{table:sumstats}) , 
-but when dispersal is limited sampling strategy can compound spatial patterns in the underlying genealogy and create pervasive impacts on all downstream genetic analyses \textcolor{red}{(see also \citet{Stadler2009})}.
+but when dispersal is limited sampling strategy can compound spatial patterns in the underlying genealogy and create pervasive impacts on all downstream genetic analyses (see also \citet{Stadler2009}).
 In most species,
@@
 
-\textcolor{red}{Previous studies have found that population structure and nonrandom sampling can create spurious signals of population bottlenecks when attempting to infer demographic history with microsatellite variation, summary statistics, or runs of homozygosity \citep{Chikhi2010,Stadler2009,Ptak2002,Mazet2015}.
+Previous studies have found that population structure and nonrandom sampling can create spurious signals of population bottlenecks when attempting to infer demographic history with microsatellite variation, summary statistics, or runs of homozygosity \citep{Chikhi2010,Stadler2009,Ptak2002,Mazet2015}.
 Here we found that methods that infer detailed population trajectories through time based on the SFS and patterns of LD across the genome are also subject to this bias, with some combinations of dispersal and sampling strategy systematically inferring deep recent population bottlenecks and overestimating ancient $N_{e}$ by a around a factor of 2.
@@
 As we will discuss below, most empirical estimates of neighborhood size, including all estimates for human populations, are large enough that population size trajectories inferred by these approaches should not be strongly affected by spatial biases created by dispersal in continuous landscapes.
-In contrast, \citet{Mazet2015} found that varying migration rates through time could create strong biases in inferred population trajectories from an $n$-island model with parameters relevant for human history, suggesting that changes in migration rates through time are more likely to drive variation in inferred $N_{e}$ than isolation by distance.}
+In contrast, \citet{Mazet2015} found that varying migration rates through time could create strong biases in inferred population trajectories from an $n$-island model with parameters relevant for human history, suggesting that changes in migration rates through time are more likely to drive variation in inferred $N_{e}$ than isolation by distance.
 
-\textcolor{red}{We found that SMC++ was more robust to the effects of space than stairwayplot, underestimating recent populations by roughly half in the worst time periods rather than nearly 10-fold as with stairwayplot.} Though this degree of variation in population size is certainly meaningful in an ecological context, it is relatively minor in population genetic terms.
+We found that SMC++ was more robust to the effects of space than stairwayplot, underestimating recent populations by roughly half in the worst time periods rather than nearly 10-fold as with stairwayplot.
+Though this degree of variation in population size is certainly meaningful in an ecological context, it is relatively minor in population genetic terms.
 A more worrying pattern was the high level of variance in inferred $N_{e}$ trajectories for individual model fits using these methods, which was highest in simulations with the smallest neighborhood size (Figure \ref{fig:demography}, Figure \ref{fig:demography_supplement}).
@@
 
-\textcolor{red}{Spatial structure is particularly challenging for genome-wide association studies, because the effects of dispersal on genetic variation are compounded by spatial variation in the environment \citep{Mathieson2012}.
+Spatial structure is particularly challenging for genome-wide association studies, because the effects of dispersal on genetic variation are compounded by spatial variation in the environment \citep{Mathieson2012}.
 Spatially restricted mate choice and dispersal causes variation in allele frequencies across the range of a species.
-If environmental factors affecting the phenotype of interest also vary over space, then groups of individuals in different regions will both experience different environments and have different allele frequencies at many sites in the genome.
+If environmental factors affecting the phenotype of interest also vary over space, then groups of individuals in different regions will allele frequencies and environmental exposures will covary over space.
 In this scenario an uncorrected GWAS will infer genetic associations with a purely environmental phenotype at any site in the genome that is differentiated over space, and the relative degree of bias will be a function of the degree of covariation in allele frequencies and the environment (i.e., Figure \ref{fig:gwas}C, bottom panel).
-This pattern is now well known and has been demonstrated in a variety of simulation and empirical contexts \citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015,Berg2018,Sohail2018}.}
+This pattern has been demonstrated in a variety of simulation and empirical contexts \citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015,Berg2018,Sohail2018}.
 
@@
 Though this approach appears to work well in practice, its interpretation is not always straightforward and it is likely biased by the presence of linked selection \citep{Berg2018}.
-\textcolor{red}{In addition, we observed that in many cases the false-positive SNPs we identified appeared to be concentrated in LD peaks similar to those expected from truly causal sites (Figure \ref{fig:manhattan_plots}), which may confound LD score regression.}
+In addition, we observed that in many cases the false-positive SNPs we identified appeared to be concentrated in LD peaks similar to those expected from truly causal sites (Figure \ref{fig:manhattan_plots}), which may confound LD score regression.
 
@@
   \makecell[l]{\textit{Dieffenbachia}\\\textit{longispatha}} & \makecell[l]{beetle-pollinated\\colonial herb} & 227 - 611 & Survey & \citep{Young1988} \\ 
+  \makecell[l]{\textit{Aedes aegypti}} & \makecell[l]{Yellow-fever mosquito} & 268 & Genetic & \citep{Jasper2019} \\ 
   \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{Gainj- and Kalam- \\speaking people,\\Papua New Guinea} & 410 & Survey & \citep{Rousset1997} \\ 
@@
 Many more species occur in a middle range of neighborhood sizes between 100 and 1000 -- a range in which spatial processes play a minor role in our analyses under random spatial sampling but are important when sampling of individuals in space is clustered.
+Surprisingly, even some flying insects with huge census population sizes fall in this group, including fruit flies (\textit{D.
+melanogaster}) and mosquitoes (\textit{A.
+aegypti}).
 Last, many species likely have neighborhood sizes much larger than we simulated, including modern humans in northeastern Europe \citep{Ringbauer2017}.
@@
 it is inevitably limited by the computational burden of tracking tens or hundreds of thousands of individuals in every generation. 
-In particular, computations required for mate selection and spatial competition scale approximately with the product of the total census size and the neighborhood size and so increases rapidly for large populations and dispersal distances.
+In particular, computations required for mate selection and spatial competition scale approximately with the product of the total census size and the neighborhood size and so increase rapidly for large populations and dispersal distances.
 The reverse-time model of continuous space evolution described by \citet{Barton2010} and implemented by \citet{Kelleher2014} 
 allows exploration of parameter regimes with population and landscape sizes more directly comparable to empirical cases like humans. 
-Alternatively, implementation of parallelized calculations may allow progress with forwards-time simulations.
+Alternatively, implementation of parallelized calculations may allow progress with forward-time simulations.
 
-\cjb{removed section on recapitation here}
 %A mixed approach can combine forward- and reverse-time models, as recently done for a continuous-space Wright-Fisher model by \citet{Lotterhos2019} and a simulation with linked selection by \citet{Buffalo2019}.
@@
 \section{Acknowledgements}
-We thank Brandon Cooper, Matt Hahn and others for reading and thinking about this manuscript. 
-We thank the Hearth for having such good, nearby coffee. 
-Falling Sky Brewing was key to the success of this research endeavor. 
+We thank Brandon Cooper, Matt Hahn, Doc Edge, and others for reading and thinking about this manuscript. 
+We thank the Hearth for having such good, nearby coffee, and
+Falling Sky Brewing for creative support during manuscript drafting.
 CJB and ADK were supported by NIH award R01GM117241.
@@
 \includegraphics[]{figures/het_z_by_ind.pdf}
-\caption{Normalized mean observed heterozygosity by location across 200 randomly-sampled individuals}
+\caption{Normalized mean observed heterozygosity by location across 200 randomly sampled individuals}
 \label{fig:hetmap}
@@
 \centering
-\includegraphics[width=\textwidth]{figures/manhattan_plots_combo-01.png}
+\includegraphics[height=0.9\textheight]{figures/manhattan_plots_combo-01.png}
 \caption{Manhattan plots for a sample of simulations at varying neighborhood sizes.
@@
 Points in red are significantly associated with a nongenetic phenotype using a 5\% FDR threshold (dashed line).
-For runs with significant associations the dotted line is a Bonferroni-adjusted cutoff for p=0.05.}
+For runs with significant associations the dotted line is a Bonferroni-adjusted cutoff for $p=0.05$.}
 \label{fig:manhattan_plots}

### 2019-06-04 scripts/run_gwas.py
preprint updates

@@ -44,21 +44,20 @@ parser.add_argument('--seed', dest='seed',type=int,
 
 args=parser.parse_args()
 
-# ## debug
+## debug
 # args = argparse.Namespace(infile='/Users/cj/spaceness/sims/slimout/spatial/W50_run3/sigma_0.5196493297511975_.trees_3918732',
 #                           outdir="/Users/cj/Desktop/",
 #                           plink_path="plink",
 #                           sampling="point",
 #                           vcftools_path="vcftools",
 #                           nSamples=1000,
-#                           mu=0.25e-8,
-#                           phenotype="random_snps",
+#                           mu=1e-8,
+#                           phenotype="corner_bimodal",
 #                           phenotype_mean=100,
 #                           phenotype_sd=10,
 #                           gentimes="/Users/cj/spaceness/W50sp_gentimes.txt",
 #                           seed=123)
-#ts=pyslim.load('/Users/cj/Desktop/sigma_0.3130183601576863_.trees2000000.trees')
-#ts=ts.recapitate(recombination_rate=1e-9)
+
 
 #sample individuals and add mutations
 np.random.seed(args.seed)
@@ -83,6 +82,7 @@ ts=msp.mutate(ts,args.mu/gentime[0],random_seed=args.seed)
 
 #get haplotypes and locations
 haps=ts.genotype_matrix()
+positions=np.array([s.position for s in ts.sites()])
 sample_inds=np.unique([ts.node(j).individual for j in ts.samples()]) #add check that nodes corresponding to individuals are sequential for future versions?
 locs=[[ts.individual(x).location[0],ts.individual(x).location[1]] for x in sample_inds]
 np.savetxt(os.path.join(args.outdir,simname)+"_locs.txt",locs)
@@ -227,6 +227,16 @@ sp.check_output([args.plink_path,
                  "--out",
                  os.path.join(args.outdir,simname),
                  "--assoc"])
+
+#get per-window LD
+# LD_windows=allel.windowed_r_squared(pos=positions,
+#                                     gn=genotype_counts[:,:,1],
+#                                     size=100000,start=1,stop=100000000)
+# out=np.array([LD_windows[0],LD_windows[1][:,0],LD_windows[1][:,1]])
+# np.savetxt(X=np.transpose(out),fname=os.path.join(args.outdir,simname)+".LD")
+
+#TODO: LD scores + regression against summary stats with LDSC
+
 
 #clean up big files
 sp.check_output(["rm",

### 2019-06-04 scripts/slimtools.py

@@ -197,21 +197,17 @@ def sample_treeseq(infile,
         extant_nodes=[ts.individual(x).nodes for x in extant_inds] #node numbers for final-generation individuals
         extant_nodes=[a for b in extant_nodes for a in b]
         ts=ts.simplify(extant_nodes)
-        #sample nSamples individuals proportional to their distance from the midpoint where p(sample)=(1/dist**4)/sum(dists_to_point)
         locs=np.array([[x.location[0],x.location[1]] for x in ts.individuals()])
         inds=np.array([x.id for x in ts.individuals()])
-        if nSamples%len(sampling_locs) == 0:
-            inds_per_pt= int(nSamples/len(sampling_locs))
-        else:
-            print("Error: nSamples must be evenly divisible by len(sampling_locs).")
-            return
-        subsample=[]
-        for i in sampling_locs:
-            dists=[scipy.spatial.distance.euclidean(locs[x],i) for x in range(len(locs))]
-            weights=[1/x**4 for x in dists]
-            weights=weights/np.sum(weights)
-            subsample.append(np.random.choice(inds,inds_per_pt,p=weights,replace=False))
-        subsample=[a for b in subsample for a in b]
+        mindists=[]
+        for i in range(len(extant_inds)):
+            row=np.array([])
+            for j in range(len(sampling_locs)):
+                row=np.append(row,spatial.distance.euclidean(u=locs[i],v=sampling_locs[j]))
+            mindists=np.append(mindists,np.min(row))
+        mindists=1/mindists**3
+        mindists=mindists/np.sum(mindists)
+        subsample=np.random.choice(extant_inds,nSamples,p=mindists,replace=False)
         subsample_nodes=[ts.individual(x).nodes for x in subsample] #node numbers for sampled individuals
         subsample_nodes=[a for b in subsample_nodes for a in b] #flatten the list
         subsample_nodes=np.sort(np.array(subsample_nodes))
@@ -349,7 +345,7 @@ def getSLiMSumStats(haps,positions,label,locs,outfile,maxlen,ibs_tracts=True,ver
                                 stop=maxlen)
     sp_dist=np.array(scipy.spatial.distance.pdist(locs))
     gen_sp_corr=np.corrcoef(gen_dist[sp_dist>0],np.log(sp_dist[sp_dist>0]))[0,1]
-    print(gen_sp_corr)
+    #print(gen_sp_corr)
[…]

### 2019-06-04 scripts/smcpp_slim_pipeline.py

@@ -55,16 +55,17 @@ def run_smcpp_plot(input_file, generation_time):
     logging.info("Running:" + cmd)
     subprocess.run(cmd, shell=True, check=True)
 
+#the business
 simdir="/home/cbattey2/spaceness/sims/slimout/spatial/W50_run3"
 outdir="/home/cbattey2/spaceness/demography/smcpp/random_sampling/"
-
 os.chdir(outdir)
 sims=os.listdir(simdir)
+existingfits=os.listdir(outdir)
 #f=sims[1]
 for f in sims:
     simname=os.path.basename(f)
     label=float(re.split("_",simname)[1])
-    if label < 4:
+    if label < 4 and str(label)+".trees" not in existingfits:
         ts=sample_treeseq(infile=os.path.join(simdir,f),
                           outfile="",
                           nSamples=20,
@@ -81,5 +82,8 @@ for f in sims:
         ts=msp.mutate(ts,1e-8/gentime[0],random_seed=12345)
         ts.dump(os.path.join(outdir,str(label)+".trees"))
         write_smcpp_file(str(label)+".trees")
-        run_smcpp_estimate(str(label)+".trees"+".smc.gz",1e-8,50)
-        run_smcpp_plot(str(label)+".trees"+".smc.gz"+".final.json",1)
+        try:
+            run_smcpp_estimate(str(label)+".trees"+".smc.gz",1e-8,50)
+            run_smcpp_plot(str(label)+".trees"+".smc.gz"+".final.json",1)
+        except:
+            print("estimate failed for sigma="+str(label))

### 2019-08-14 ms/review-responses.tex
review response stuff

%%%%%%
%%
%%  Don't reorder the reviewer points; that'll mess up the automatic referencing!
%%
%%%%%

\begin{minipage}[b]{2.5in}
  Resubmission Cover Letter \\
    {\it Genetics}
\end{minipage}
\hfill
\begin{minipage}[b]{2.5in}
    C.~J.~Battey, \\
    Peter Ralph, \\
    \emph{and} Andrew Kern \\
  \today
\end{minipage}

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are writing to submit a revised version of our manuscript, 
``Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data''.

\vspace{5em}

\noindent \hspace{4em}
\begin{minipage}{3in}
\noindent
{\bf Sincerely,}

\vskip 2em

{\bf 
C.~J.~Battey, Peter Ralph, and Andrew Kern
}\\
\end{minipage}

\vskip 4em

\pagebreak
\setcounter{page}{1}

%%%%%%%%%%%%%%
\reviewersection{AE}

\begin{quote}
    The manuscript admirably explores a lot of consequences of
    isolation-by-distance in the context of a novel model that is easily amenable
    to forward simulation; however, given that this model may be used in a lot of
    future studies based on the precedent set here, there is some concern about the
    model and its support. Reviewers 2 and 3 highlight this in particular (it
    underlies the main 2 points of reviewer 2's review, and the core of Reviewer
    3's comment), and I agree. Whatever can be done to strengthen the standing of
    this model, and/or connect it to more thoroughly studied models, will be
    helpful for the manuscript. The concern would be that there are peculiarities
    of this model that do not generalize well. A new supplemental section or opener
    to the results section establishing the model more thoroughly would make the
    strongest response.
\end{quote}

\plr{I would generally cut down the quoted bits like the above to only what's essential,
     but haven't done that yet.}

 \plr{IMPORTANT: don't reorder or delete "points" below - it messes up the automatic numbering!}

\begin{point}{Line 35:}
    Also cite Wilkins and Wakeley, Genetics 2002; Wilkins 2004
\end{point}

\reply{
}

\begin{point}{\revref}
    "Such models have been used extensively in ecological modeling but rarely in population genetics " Detailing these previous uses via citations and elaboration may help alleviate the major concern about the provenenance of this model and its unique behaviors (see general comments above and R2 and R3 comments).
\end{point}

\reply{
}

\begin{point}{\revref}
    Please describe computation time needed per replicate
\end{point}

\reply{
}

\begin{point}{\revref}
    I read the acknowledgement to the Hearth and Creative Sky Brewing with a sense of familiarity in feeling of gratitude to my own favorite cafes and breweries, but I it's not a great precedent for Acknowledgements to be filled this way. Please cut.
\end{point}

\reply{
    Good point; we have done this.
}

\begin{point}{Figure 4:}
    Show random-mating expectation
\end{point}

\reply{
}

\begin{point}{Figure 3A, S2:}
    Perhaps more revealing to show on log-log scale?
\end{point}

\reply{
}

\begin{point}{Figure S3:}
    Caption seems to be missing detail 
\end{point}

\reply{
}

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
    This study explores biases arising in population-based inference when 1)
    real population samples are coming from spatial habitat with various degree
    of structuring while inference is made assuming random mating population;
    2) imperfect sampling in practice that fails to represent full diversity
    across entire population habitat; 3) phenotypes that vary across geography
    and create spurious associations with genotypes. While earlier studies
    explored the effect of strong structure on population genetic inference and
    GWAS, this work focuses on less extreme scenarios of structuring that
    arises in populations evolving in continuous habitat. By using
    non-Wright-Fisher model, authors simulated chromosome-scale samples from
    populations that evolved in continuous space, and that can model
    environmental factors to create phenotypes varying over space. As a result,
    this study identified spatial structuring scenarios (small neighbourhood
    size ~ 10-100) that coupled with imperfect sampling strategies lead to a
    biased inference of widely used population genetic statistics (altogether
    18 statistics) such as pi (average pairwise sequence differences),
    heterozygosity (and inbreeding coefficient), and IBS tract sharing.
    Accordingly, inference of the effective population size history was also
    strongly affected under these parameter ranges. Finally, the authors use
    their spatial modelling to demonstrate that typical GWAS with PC-based
    correction cannot entirely remove spurious signals of genotype-phenotype
    associations arising from purely environmental factors.  Overall, the
    authors explore an important but often neglected source of bias that can
    affect inference in many population-based studies (in medical genetics,
    evolutionary biology and ecology). This study can be of interest to a
    broader audience of readership, and I have only minor comments to improve
    clarity and increase accessibility for readers:
\end{quote}

\begin{point}{}
    When neighbourhood size is small (10-100), the mean number of IBS tracts > 2bp (nIBS as in Table S1) is elevated similar to Wright's inbreeding coefficient, but mean of the distribution of pairwise IBS (mean(IBS)) is decreased. What could be the source of this discrepancy? How exactly mean(IBS) was calculated?
\end{point}

\reply{
}

\begin{point}{}
    The authors use K to denote both carrying capacity \llname{ll:K_defn}
    and population density \llname{ll:K_again}. 
    It might be better to use a different notation for these quantities since carrying capacity is fixed while density is an emergent quantity in the non-Wright-Fisher model. Use of K to denote carrying capacity and density is a bit confusing. For example, on \llname{ll:K_more} it is said that ' the "population density" (K) and "mean lifetime" (L) parameters were the same in all simulations'. Here K seems to indicate carrying capacity rather than density? The latter is an emergent quantity and varies across simulation runs?
\end{point}

\reply{
    We agree that this distinction is worth emphasizing!
    We've adjusted our language to hopefully remind the reader that
    $K$ is a parameter that controls population density, rather than being equal to it,
    at \llname{ll:K_again} and \revref{} and \llname{ll:K_more}.
}

\begin{point}{}
    Concerning the non-Wright-Fisher model used, it would be helpful to emphasize that some of the parameters are emergent in contrast to Wright-Fisher model. For example, on Page 11, lines 306-308, the author's goal was to look at census size variation and variation in other quantities. This would be better understood if to emphasize that these parameters are emergent properties in the non-Wright-Fisher model used.
\end{point}

\reply{
}

\begin{point}{}
    Page 9, line 242, Perhaps 'Demographic Inference' might better reflect the content of this section.
\end{point}

\reply{
}

\begin{point}{\revref}
    This sentence with 'Gaussian noise with mean zero and standard deviation 10' is confusing since it was mentioned earlier that the modelled phenotype must vary as human height across Europe, and human height varies 2 standard deviations. Only after reading the whole paragraph it becomes clear that 'standard deviation 10' here referes to unit of height. Please consider rephrasing this sentence.
\end{point}

\reply{
}

\begin{point}{\revref}
    In the sentence, 'We also examined p values for systemic inflation' I think the authors meant 'systematic inflation'.
\end{point}

\reply{
    Whoops; thanks. Fixed.
}

\begin{point}{}
    Page 11, Please correct the legend in Figure 2: must be 'spatial model' and 'random mating' model.
\end{point}

\reply{
}

\begin{point}{}
    Optional: a dashed line in Figure 2 that shows the total carrying capacity of 50*50*5=12500 would be helpful.
\end{point}

\reply{
}

\begin{point}{}
    Page 13, line 349, The phrase 'affect summaries of variation' is better to replace with 'summaries of genetic variation'.
\end{point}

\reply{
}

\begin{point}{}
    Please add or correct references to supplementary figures: For example, Figure S2 was probably meant to accompany Figure 3A, while Figure S1 Figure 3B, but references in the text are absent. In fact, the first reference is made to Figure S3 on page 15.
\end{point}

\reply{
}

\begin{point}{}
    There are also several typos and errors in the text. For example, on Page 12, lines 309; Page 27, line 655.
\end{point}

\reply{
}

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
    Battey et al. use spatially explicit population genetic simulations to
    analyze the effects of spatial structure on (i) the estimation of key
    population genetic parameters, in turn used to (ii) make inferences about
    population history, and on (iii) confounding in genome-wide association
    studies (GWAS). I Liked the paper a lot. It's interesting, well-written and
    addresses an important question - the effect of spatial population
    structure on population genetic statistics and inference-and I enjoyed
    reading it. The most positive aspects were:

    \begin{enumerate}
        \item It nice to actually see spatially explicit simulations and I'm happy that forward simulation is now fast enough that you can do this sort of thing.
        \item The paper is very clear and well-written, easy to understand the motivation and most of the details. That's not always the case for this sort of paper.
        \item I felt that the section about the effect on GWAS was the most interesting and novel part of the paper and gave me some intuition that I hadn't had before.
    \end{enumerate}

    I don't have any major criticisms. There were a few aspects that I thought might warrant some additional discussion, and a few specific questions below. The general questions I had after reading it were:
\end{quote}

\begin{point}{}
    To what extent are any of the results dependent on the exact method of simulation. There are a number of choices about the exact details of the simulations (e.g. the way the overlapping generations are handled, the edge effects and, particularly, the form of Equation 1 - see below). It's not so much that these are non-standard (since I don't think there is a standard) and they all sort of make sense heuristically, and I was left wondering whether these sorts of choices actually make a difference. Do the authors have some thoughts/intuition/results about that? Given that the results in Fig. 3 seem quite consistent with expectations, I suspect that on some level it doesn't make much difference but then there are intermediate results like Fig. 2 which seem a bit counter-intuitive and I wonder if those aspects depend on the simulation scheme.
\end{point}

\reply{
}

\begin{point}{}
    Related to the first point, to what extent are the results qualitatively different to those that would be obtained in a stepping-stone model? My interpretation is that they are actually very similar, but I didn't see whether that was explicitly discussed. In some sense, it's still easier to do large simulations in a stepping-stone model so it would be nice to be reassured that that's still ok.
\end{point}

\reply{
}

\begin{point}{}
    The source of equation (1) is not obvious to me. I sort of see how it makes sense, but a little but more intuition or a brief derivation or an illuminating either in the main text or the supplement, would be helpful.
\end{point}

\reply{
}

\begin{point}{}
    The authors use a scaling factor in equation (2) to counteract the increase in fitness of individuals at the edges. Can they provide a figure showing that this is the case. What does "roughly" mean on line 164. Perhaps a heatmap of the fitness of individuals across the grid with and without the scaling factor?
\end{point}

\reply{
}

\begin{point}{}
    It would be helpful provide the figure showing that generating mutations during the forward simulations in SLIM is equivalent to applying mutations using msprime on pre-generated trees (line 185)? It sounds like this procedure would underestimate the variance in the number of mutations, since you remove the effect of random generation time. Is this effect small?
\end{point}

\reply{
}

\begin{point}{}
    Can the authors provide a bit more intuition behind the patterns of variation seen in generation time, census population size, and variance in the number of offspring with respect to neighborhood size seen in Figure 2? For example, it is not obvious to me why the census population size, for example, should decline systematically with respect to neighborhood size. Presumably this isn't just due to the local demographic stochasticity. Could the authors briefly interpret the observed patterns or cite appropriate literature?
\end{point}

\reply{
}

\begin{point}{}
    Fig. 7D: I am surprised by the extent to which the observed values of -log10(p) fall below the y=x line. Particularly in the lower right panel for large neighbourhood sizes. I would expect that to be close to panmictic - why are the P-values underdispersed? That seems like a potential bug, or else something weird is going on.
\end{point}

\reply{
}

\begin{point}{}
    Lines 706-716, It might be worth citing Haworth et al Nature Communications 2019 (https://doi.org/10.1038/s41467-018-08219-1) who do the proposed test (GWAS for birth location) in UK Biobank to illustrate the population structure.
\end{point}

\reply{
}

\begin{point}{}
    The analysis and discussion around the effect of GWAS is focused on PCA correction. Do mixed models help at all?
\end{point}

\reply{
}

\begin{point}{}
    The github link to the code didn't work for me. I assume it will be made public later, but at this point I can't tell whether the code is available/useable.
\end{point}

\reply{
}

%%%%%%%%%%%%%%%%%%%
\reviewersection{3}

\begin{quote}
    The present study deals with a "hot topic" in spatial population genetics. Most
    inferential and descriptive methods in statistical spatial population
    genetic rely on a discrete approximation of space and it is not clear what
    impact this approximation may have when individuals migrate along a
    continuum instead. Spatial patterns in sampling is also another major issue
    which is often simply dismissed, mainly because of the paucity of
    statistical methods to deal with it. This work touches on these important
    issues in a timely manner.
\end{quote}

\begin{quote}
    Although I was enthusiastic about the topic, I was quite disappointed with
    the core of the study, i.e., the forward-in-time simulation of populations in
    continuous space. The field has been struggling with this issue for decades --
    examples of spectacular failures like the Wright-Malecot model (see
    Felsenstein's "pain in the torus' article, 1975) or, more recently, the
    ``mugration' or ``discrete trait analysis' model in phylodynamics (see De Maio
    et al. 2015) have probably mostly harmed our research field -- that one cannot
    make the economy of using a sound probabilistic model for generating
    geo-referenced genetic data. It does not seem to be the case here
    unfortunately.  
\end{quote}

\begin{point}{}
    First, the simulation starts with individuals distributed uniformly at random in space. Is there any indication that the three-step algorithm used here maintains this distribution during the course of evolution? If it does not, then is there any stationary regime and how many generations does one need to wait before reaching it? I do appreciate that the competitive interaction term was introduced in order to avoid seeing the ``clumping' of individuals that hampers the Wright-Malecot model. Yet, just because there are no such clusters does not mean that the spatial distribution of individuals reaches a stable regime and that the distribution reached, if any, is reasonable from a biological perspective.
\end{point}

\reply{
}

Second, the demographic process used here involves birth and death of individuals. Does the population survive asymptotically or, like any birth-death process, eventually dies with probability one? In fact, one needs to know a little about the dynamics of the population size to decide whether the corresponding process is reasonable from a biological standpoint.
\begin{point}{}

\end{point}

\reply{
}

\begin{point}{}
    Third, it is not clear what the relationship between the expected lifespan and the probability of survival is. The expected lifespan, L, is first defined as the inverse of the expected number of offspring produced by a parent. The authors also define the probability of survival of a given individual at a given point in space, $p_i$. Hence, the expected lifespan at a point in space (and time) is the mean of a geometric distribution with parameter $p_i$, i.e., $1/p_i$. Now, it is far from being obvious what the relationship between these two approaches for defining the expected lifespan actually is.
\end{point}

\reply{
}

\begin{point}{}
    Also, the web page https://github.com/petrelharp/spaceness does not seem to exist so that I was not able to experiment with the forward-in-time generator used here unfortunately.
\end{point}

\reply{
}

\begin{point}{}
    All in all, more efforts need to be made here in my opinion to show that the forward-in-time simulations generate sensible outcomes. Sensible in terms of the behavior of the population demography at equilibrium (provided such equilibrium indeed exists) along with that of the spatial distribution of individuals. The authors could provide some guarantee of the good behavior of their model as evidenced from simulations using a broad range of parameter values for generating data. Alternatively, they could elect to use the spatial-Lambda-Fleming-Viot model for their simulations, which, in my opinion would seem the most sensible option given that (1) it is possible to run backward-in-time simulations under this model, thereby saving a lot of computation time and (2) it is a well-studied model with good mathematical and biological properties and (3) it is implemented in a publicly available software program (https://github.com/jeromekelleher/discsim)
\end{point}

\reply{
}

\begin{point}{}
    Figure 2: I do not understand why the neighborhood size varies to the same extent in the random mating model as it does for the spatial model. For the random mating model, I would have expected the neighborhood size to be equal to the census size since all individuals have the same probability of being a parent of any given offspring. From lines 166->171, it is clear that the spatial model would converge to the random mating model when the mean parent-offspring distance tends to infinity only if we were to ignore the impact of range edges. I am thus wondering whether the variation of neighborhood size one observes in Fig 2 for the random mating model is just a consequence of border effects. If that is the case, then the authors should state it clearly and try to justify it from a biological perspective.
\end{point}

\reply{
}

\begin{point}{}
    Line 729-731: "Many more species occur in a middle range of neighborhood sizes between 100 and 1000 - a range in which spatial processes play a minor role in our analyses [...]". Do the authors think that the spatial processes would still play a minor role when neighborhood sizes exceed 100-1000 if the habitat was larger than that taken in the present simulations? It would also probably be useful to mention that neighborhood sizes given in Table 1 should be compared with extreme caution since the size of the corresponding habitats vary across species. More generally, I suspect that the size of the habitat has a substantial impact on the vast majority of statistics examined in this study. Indeed, the mean parent-offspring distance, which is at the core of the definition of Wright's neighborhood size, is only small or large relative to the size of the habitat.
\end{point}

\reply{
}

\begin{point}{}
    Line 753-757: please add a reference to Guindon, Guo and Welch (2016). This study clearly shows that population density and dispersal parameters are identifiable and can indeed be estimated in practice under the spatial Lambda-Fleming-Viot model.
\end{point}

\reply{
}

%%%%%%%%%%%%%%%%%%%
\reviewersection{4}

\begin{quote}
    The manuscript by Battey et al explores the consequence of a well-known
    violation to population genetic models: the fact that populations are
    spatially structured and mate along a geographical cline, rather than
    randomly. This topic is important, particularly in light of recent working
    describing how spatially correlated genetic and environmental impacts can
    confound some population genetic insights, such as positive selection for
    height in Europe. The analyses and investigations presented here are
    thorough and sensible, and my comments are primarily intended to broaden
    accessibility for this interesting topic.
\end{quote}

\begin{point}{}
    Introduction. The discussion is very clear, articulating the three primary goals of the project: the impact of failing to model spatial population structure on 1) population genetic summary statistics, 2) inference on demographic history from population genetic data, and 3) impacts on GWAS summary statistics. I found the discussion a bit easier to follow than the introduction and would suggest streamlining and introducing the topic a bit more. Since the paper follows the flow described in the discussion, it might help orient readers by introducing these topics in the same order.
\end{point}

\reply{
}

\begin{point}{}
    I agree that most modern work describes structure as discrete populations connected by migration. However, some methods/studies have explicitly modeled spatial structure, e.g. especially in ecology or using methods like dadi (diffusion approximations). Highlighting some examples of previously identified structure not possible to infer without modeling geography would be helpful to contextualize this work.
\end{point}

\reply{
}

\begin{point}{}
    There is some reference to spatial models using grids (e.g. Rousset 1997). Some additional discussion contextualizing more recent methods like EEMS that also construct demes and model migration through divergence between neighboring demes would be helpful and interesting.
\end{point}

\reply{
}

\begin{point}{}
    Demographic modeling. Both approaches tested, stairwayplot and SMC++, are most sensitive to older demographic events, and consequently are very noisy and underestimate effect population sizes, especially in smaller neighborhood sizes. Models that consider haplotype structure are much better suited to this time period. It would be helpful to either 1) discuss the varying time sensitivities of different classes of demographic inference methods and how spatial patterns of genetic variation would influence these inferences, or 2) apply a method of this class (many options, e.g. DoRIS, IBDNe, Tracts, Globetrotter, etc) and show how it performs.
\end{point}

\reply{
}

\begin{point}{}
    GWAS mixed models. To what extent can spatial signals (e.g. corner, patchy) be corrected with mixed models, e.g. with PCs and PC-adjusted GRM as in Conomos et al, 2016 using PC-AiR and PC-Relate)? Is patchiness related to dispersal? I'm curious how this relates to the predictive ability of GWAS phenotypes with some spatial association that may or may not be associated with environmental effects.
\end{point}

\reply{
}

\begin{point}{}
    Code availability. This github link doesn't work, but is important to be able to evaluate for review: https://github.com/petrelharp/spaceness
\end{point}

\reply{
}

\begin{point}{}
    Definitions and interpretations. There are quite a large number of metrics discussed in Figure 3B, and it's a lot to take in. It might be helpful to have a table with a reminder of what the metric is, its interpretation, and how it is computed.
\end{point}

\reply{
}

\begin{point}{}
    Notation: "Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$ in both the x and y coordinates. Each offspring is produced with a mate selected randomly from those within distance $3\sigma$, with probability of choosing a neighbor at distance $x$ proportional to $\exp(-x^2/2\sigma^2)$." I think x may be overloaded here, or I'm confused. Clarify?
\end{point}

\reply{
}

\begin{point}{}
    When introducing the "spatial model" as opposed to this "random model," the more concrete illustration in Figure 1 hasn't yet been referenced, which makes it harder to follow. It would be helpful to introduce this figure with the model. Additionally, when Figure 1 is introduced, the order is from right to left (random, then point, then midpoint). It would be helpful to rearrange the figure to mirror what's in the text.
\end{point}

\reply{
}

\begin{point}{}
    Not sure I follow this example: "Concretely, an individual at position (x, y) in a 50 × 50 landscape has mean phenotype 100 + 2x/5."
\end{point}

\reply{
}

\begin{point}{}
    Minor typo (through vs though): "This occurs because, even through the "population density" (K) and "mean lifetime" (L) parameters..."
\end{point}

\reply{
}

\begin{point}{}
    Define NS abbreviation in Figure 5.
\end{point}

\reply{
}

### 2019-08-14 ms/spaceness.tex
review response stuff

@@
 \newcommand{\ak}[1]{\ifcomments{{\color{red} \it (#1)}}\else{}\fi}
+
+%%%%%%%%%%%
+% for responses to reviewers
+% set this to show line numbers and include responses to reviews or not
+\newif\ifreviewresponses
+\reviewresponsestrue  % include them
+% \reviewresponsesfalse  % don't include them
+\input{review-response-commands}
+\newcommand{\responsefile}{review-responses.tex}  % name of the review reponses file
 
@@
 Such models have been used extensively in ecological modeling but rarely in population genetics, where to our knowledge previous implementations of continuous space models have focused on a small number of genetic loci, which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms.
+\revpoint{AE}{2}
 By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.
@@
 Since $g$ is a probability density, $n_i$ is an estimate of the number of nearby individuals per unit area.
-Then, given a per-unit carrying capacity $K$, the probability of survival until the next time step for individual $i$ is
+Then, given a per-unit carrying capacity $K$, \llabel{ll:K_defn}
+the probability of survival until the next time step for individual $i$ is
 \begin{equation}
@@
 \end{equation}
-We chose this functional form so that the equilibrium population density per unit area is around $K$,
+We chose this functional form so that the equilibrium population density per unit area is close to $K$, \llabel{ll:K_again}
 and the mean lifetime is around $L$.
@@
 
-We ran 400 simulations for the spatial and random-mating models on a square landscape of width $W=50$ 
+We ran 400 simulations for the spatial and random-mating models on a square landscape of width $W=50$  \revpoint{AE}{3}
 with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), 
@@
 
-Our demographic model includes parameters for population density ($K$),
+Our demographic model includes parameters that control population density ($K$), \revpoint{1}{2}
 mean life span ($L$), and dispersal distance ($\sigma$).
@@
 In all simulations, the phenotype of each individual
-is determined by adding independent Gaussian noise with mean zero and standard deviation 10
+is determined by adding independent Gaussian noise with mean zero and standard deviation 10 \revpoint{1}{5}
 to a mean that may depend on spatial position.
@@
 with $p$-value below $0.05$ after adjusting for an expected false positive rate of less than 5\% \citep{benjamini2001control}. 
-We also examined $p$ values for systemic inflation by estimating the expected values from a uniform distribution (because no SNPs were used when generating phenotypes), plotting observed against expected values for all simulations, and summarizing across simulations by finding the mean $\sigma$ value in each region of quantile-quantile space.
+We also examined $p$ values for systematic inflation \revpoint{1}{6}
+by estimating the expected values from a uniform distribution (because no SNPs were used when generating phenotypes), plotting observed against expected values for all simulations, and summarizing across simulations by finding the mean $\sigma$ value in each region of quantile-quantile space.
 Results from all analyses were summarized and plotted with the ``ggplot2'' \citep{Wickham2016} and ``cowplot'' \citep{Wilke2019} packages in R \citep{Rcore2018}.
@@
 This occurs because,
-even through the ``population density'' ($K$) and ``mean lifetime'' ($L$) parameters
+even through the parameters $K$ and $L$ that control population density and mean lifetime respectively \llabel{ll:K_more}
 were the same in all simulations,
@@
 We thank Brandon Cooper, Matt Hahn, Doc Edge, and others for reading and thinking about this manuscript. 
-We thank the Hearth for having such good, nearby coffee, and
-Falling Sky Brewing for creative support during manuscript drafting.
+% We thank the Hearth for having such good, nearby coffee, and
+% Falling Sky Brewing for creative support during manuscript drafting.
 CJB and ADK were supported by NIH award R01GM117241. 
+\revpoint{AE}{4}
 
@@
 
+\newpage
+
+\includereviews
+
 \end{document}

### 2019-09-12 ms/spaceness.tex
msp grid updates

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Supplement %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Appendix 1}
+\subsection{Comparisons with Alternate Spatial Models}
+Stepping-stone models have been extremely useful in population genetics and, when simulated in reverse time, remain much faster than our implementation of continuous-space evolution in forward time.
+Similarly, the Wright-Malécot models have lead to analytic results that have proved reasonable approximations for estimating neighborhood sizes from empirical data \citep{Rousset1997}.
+When are these models "good enough"?
+
+First we address a two-dimensional Wright-Malécot model.
+SLiM includes an implementation of a similar model in recipe 14.1.
+This model simulates a Wright-Fisher population with discrete generations and fixed total population size.
+Individuals are initially distributed randomly across the landscape, and each generation
+
 \section{Supplementary Figures and Tables}

### 2019-09-12 batch_scripts/1_run_slim_simulations.sh

@@ -1,5 +1,5 @@
 #!/bin/bash
-for i in {1..500} 
+for i in {1..100} 
 do
 	sbatch /projects/kernlab/cbattey2/spaceness/batch_scripts/runslim_spatial.srun  
 	sbatch /projects/kernlab/cbattey2/spaceness/batch_scripts/runslim_random_mating.srun

### 2019-09-12 batch_scripts/2_get_sumstats.sh

@@ -3,16 +3,17 @@
 cd ~/projects/spaceness
 
 #spatial mate choice sims
-files=~/projects/spaceness/sims/slimout/spatial/W35/*
+files=~/projects/spaceness/sims/slimout/spatial/W50_run3/*
 for f in $files
 do
  
  command="python scripts/sumstats_from_treeseq.py \
  --infile $f \
- --outfile ~/projects/spaceness/sumstats/ss_spatial_random_W35.txt \
- --sampling random \
- --mu 0.25e-8 \
- --sampling_locs 17,17 \
+ --outfile ~/projects/spaceness/sumstats/ss_spatial_point_W50.txt \
+ --sampling point \
+ --mu 1e-8 \
+ --sampling_locs 12.5,12.5 \
+ --gentimes ~/projects/spaceness/W50sp_gentimes.txt \
  --seed 12345"
 
 cp ~/projects/spaceness/batch_scripts/slurm_sumstats_header.srun ~/projects/spaceness/batch_scripts/tmp.srun
@@ -26,16 +27,17 @@ sleep 0.05
 done
 
 #random mating sims
-files=~/projects/spaceness/sims/slimout/random_mating/W35/*
+files=~/projects/spaceness/sims/slimout/random_mating/W50_run3/*
 for f in $files
 do
  
  command="python scripts/sumstats_from_treeseq.py \
  --infile $f \
- --outfile ~/projects/spaceness/sumstats/ss_randmates_random_W35.txt \
- --sampling random \
- --mu 0.25e-8 \
- --sampling_locs 17,17 \
+ --outfile ~/projects/spaceness/sumstats/ss_randmates_point_W50.txt \
+ --sampling point \
+ --mu 1e-8 \
+ --sampling_locs 12.5,12.5 \
+ --gentimes ~/projects/spaceness/W50rm_gentimes.txt \
  --seed 12345"
 
 cp ~/projects/spaceness/batch_scripts/slurm_sumstats_header.srun ~/projects/spaceness/batch_scripts/tmp.srun

### 2019-09-12 batch_scripts/3_run_gwas.sh

@@ -3,22 +3,23 @@
 cd ~/projects/spaceness
 
 #spatial mate choice sims
-files=~/projects/spaceness/sims/slimout/spatial/W50/coalesced/*
+files=~/projects/spaceness/sims/slimout/spatial/W50_run3/*
 for f in $files
 do
 
  command="python scripts/run_gwas.py \
  --infile $f \
- --outdir ~/projects/spaceness/gwas/out_normal \
+ --outdir ~/projects/spaceness/gwas/point_sampling/out_normal \
  --plink_path ~/projects/plink-1.07-x86_64/plink \
  --vcftools_path ~/projects/vcftools_0.1.13/bin/vcftools \
  --nSamples 1000 \
- --sampling random \
- --mu 0.25e-8 \
+ --sampling point \
+ --mu 1e-8 \
  --phenotype gaussian \
  --phenotype_mean 100 \
  --phenotype_sd 10 \
- --seed 85904"
+ --gentimes ~/projects/spaceness/W50sp_gentimes.txt \
+ --seed 12345"
  
 cp ~/projects/spaceness/batch_scripts/slurm_gwas_header.srun ~/projects/spaceness/batch_scripts/tmp.srun
 
@@ -28,15 +29,16 @@ sbatch ~/projects/spaceness/batch_scripts/tmp.srun
 
 command="python scripts/run_gwas.py \
 --infile $f \
---outdir ~/projects/spaceness/gwas/out_clinal \
+--outdir ~/projects/spaceness/gwas/point_sampling/out_clinal \
 --plink_path ~/projects/plink-1.07-x86_64/plink \
 --vcftools_path ~/projects/vcftools_0.1.13/bin/vcftools \
 --nSamples 1000 \
---sampling random \
---mu 0.25e-8 \
+--sampling point \
+--mu 1e-8 \
 --phenotype transform_coord \
 --phenotype_mean 100 \
 --phenotype_sd 10 \
+--gentimes ~/projects/spaceness/W50sp_gentimes.txt \
 --seed 12345"
 
 cp ~/projects/spaceness/batch_scripts/slurm_gwas_header.srun ~/projects/spaceness/batch_scripts/tmp.srun
@@ -47,16 +49,17 @@ sbatch ~/projects/spaceness/batch_scripts/tmp.srun
 
 command="python scripts/run_gwas.py \
  --infile $f \
- --outdir ~/projects/spaceness/gwas/out_corner \
[…]

### 2019-09-12 batch_scripts/batch_slim_gwas.sh

@@ -1,0 +1,23 @@
+#!/bin/bash
+cd ~/projects/spaceness/gwas/sims/
+files=./*
+for f in $files
+do
+ 
+command="Rscript ~/projects/spaceness/demography/smcpp/run_smcpp.R \
+--infile $f \
+--outdir ~/projects/spaceness/demography/smcpp/output \
+--cores 28 \
+--recomb 1e-8 \
+--mu 1e-8"
+
+cp ~/projects/spaceness/demography/smcpp/smcpp_slurm_header.srun ~/projects/spaceness/demography/smcpp/tmp.srun
+
+echo $command >> ~/projects/spaceness/demography/smcpp/tmp.srun
+
+sbatch ~/projects/spaceness/demography/smcpp/tmp.srun
+	
+sleep 2
+
+done
+

### 2019-09-12 batch_scripts/get_sumstats_msprime.sh

@@ -1,0 +1,22 @@
+#!/bin/bash
+# get summary statistics for all tree sequences in a directory
+cd ~/projects/spaceness
+
+#spatial mate choice sims
+files=~/projects/spaceness/sims/msp/*
+for f in $files
+do
+ 
+ command="python scripts/msp_ts_to_sumstats.py \
+ --infile $f \
+ --outfile ~/projects/spaceness/sumstats/ss_msp.txt"
+
+cp ~/projects/spaceness/batch_scripts/slurm_sumstats_header.srun ~/projects/spaceness/batch_scripts/tmp.srun
+
+echo $command >> ~/projects/spaceness/batch_scripts/tmp.srun
+
+sbatch ~/projects/spaceness/batch_scripts/tmp.srun
+	
+sleep 0.05
+
+done

### 2019-09-12 batch_scripts/get_sumstats_sesame.sh

@@ -24,8 +24,7 @@ for f in $files
 	getsumstats "$f" & 
 done
 )
-i=1
-wait
+wait $!
 
 #midpoint sampling
 getsumstats(){
@@ -46,8 +45,7 @@ for f in $files
 	getsumstats "$f" & 
 done
 )
-i=1
-wait
+wait $!
 
 #point sampling
 getsumstats(){
@@ -68,8 +66,7 @@ for f in $files
 	getsumstats "$f" & 
 done
 )
-i=1
-wait
+wait $!
 
 ############### random mating sims ###############
 files=~/spaceness/sims/slimout/random_mating/W35/*
@@ -92,8 +89,7 @@ for f in $files
 	getsumstats "$f" & 
 done
 )
-i=1
-wait 
+wait $!
 
 #midpoint sampling
 getsumstats(){
@@ -114,8 +110,7 @@ for f in $files
 	getsumstats "$f" & 
 done
 )
-i=1
-wait
+wait $!
 
 #point sampling
 getsumstats(){
@@ -136,7 +131,5 @@ for f in $files
 	getsumstats "$f" & 
 done
 )
-i=1
-wait
-
+wait $!
 

### 2019-09-12 batch_scripts/run_slim_batch_10kouts.sh

@@ -1,0 +1,6 @@
+#!/bin/bash
+for i in {1..100} 
+do
+	sbatch /projects/kernlab/cbattey2/spaceness/scripts/runslim_10kouts.srun
+done
+

### 2019-09-12 batch_scripts/run_slim_batch_random_mating.sh

@@ -1,0 +1,6 @@
+#!/bin/bash
+for i in {1..200} 
+do
+	sbatch /projects/kernlab/cbattey2/spaceness/batch_scripts/runslim_random_mating.srun
+done
+

### 2019-09-12 batch_scripts/run_slim_batch_small_long.sh

@@ -1,0 +1,6 @@
+#!/bin/bash
+for i in {1..100} 
+do
+	sbatch /projects/kernlab/cbattey2/spaceness/scripts/runslim_small_long.srun
+done
+

### 2019-09-12 batch_scripts/run_slim_batch_spatial.sh

@@ -1,0 +1,6 @@
+#!/bin/bash
+for i in {1..50} 
+do
+	sbatch /projects/kernlab/cbattey2/spaceness/batch_scripts/runslim_spatial.srun
+done
+

### 2019-09-12 batch_scripts/run_slim_batch_vcfout.sh

@@ -1,0 +1,11 @@
+#!/bin/bash
+
+for i in {1..10}
+do 
+	simnum=$i
+	command="slim -d sigma=0.2 -d simnum=$simnum /projects/kernlab/cbattey2/spaceness/slim_recipes/flat_map_vcfout.slim"
+	cp ~/projects/spaceness/batch_scripts/runslim_header.srun ~/projects/spaceness/batch_scripts/tmp.srun
+	echo $command >> ~/projects/spaceness/batch_scripts/tmp.srun
+	#tail ~/projects/spaceness/batch_scripts/tmp.srun
+	sbatch ~/projects/spaceness/batch_scripts/tmp.srun
+done

### 2019-09-12 batch_scripts/run_slim_batch_wf.sh

@@ -1,0 +1,6 @@
+#!/bin/bash
+for i in {1..200} 
+do
+	sbatch /projects/kernlab/cbattey2/spaceness/scripts/runslim_wf.srun
+done
+

### 2019-09-12 batch_scripts/runslim_vcfout.sh

@@ -1,0 +1,12 @@
+#!/bin/bash
+#SBATCH --partition=kern,preempt       ### queue to submit to
+#SBATCH --job-name=slimstats    ### job name
+#SBATCH --output=slimstats.out   ### file in which to store job stdout
+#SBATCH --error=slimstats.err    ### file in which to store job stderr
+#SBATCH --time=240                ### wall-clock time limit, in minutes
+#SBATCH --mem=4G              ### memory limit
+#SBATCH --nodes=1               ### number of nodes to use
+#SBATCH --ntasks-per-node=1     ### number of tasks to launch per node
+#SBATCH --cpus-per-task=1       ### number of cores for each task
+
+slim -d sigma=0.2 -d simnum=1 ~/spaceness/slim_recipes/flat_map_vcfout.slim

### 2019-09-12 scripts/IBS_plots.py

@@ -1,0 +1,157 @@
+from slimtools import *
+os.chdir("/Users/cj/spaceness")
+
+#sample and mutate treeseq
+def get_ibs_dist(infile,sampling,nSamples,min_len_to_keep):
+    np.random.seed(12345)
+    #infile="sims/slimout/spatial/W50_run3/sigma_3.05686209988018_.trees_7807882"
+    simname=os.path.basename(infile)
+    label=float(re.split("_",simname)[1])
+    ts=sample_treeseq(infile=infile,
+                      outfile="",
+                      nSamples=nSamples,
+                      sampling=sampling,
+                      recapitate=False,
+                      recombination_rate=1e-8,
+                      write_to_file=False,
+                      sampling_locs=[[12.5,12.5],[12.5,37.5],[37.5,37.5],[37.5,12.5]],
+                      plot=False,
+                      seed=12345)
+
+    #get generation times estimated from short all-individual simulations
+    gentimes=np.loadtxt("/Users/cj/spaceness/W50sp_gentimes.txt")
+    gentime=[x[0] for x in gentimes if np.round(x[1],5)==np.round(label,5)]
+    ts=msp.mutate(ts,1e-8/gentime[0],random_seed=12345)
+    haps,positions,locs=get_ms_outs(ts)
+    positions=discretize_snp_positions(positions)
+    maxlen=1e8
+    verbose=True
+
+    #load in genotypes etc into scikit-allel
+    genotypes=allel.HaplotypeArray(haps).to_genotypes(ploidy=2)
+    allele_counts=genotypes.count_alleles()
+    genotype_allele_counts=genotypes.to_allele_counts()
+
+    #nonspatial population-wide summary statistics
+    segsites=np.shape(genotypes)[0]
+    #mpd=np.mean(allel.mean_pairwise_difference(allele_counts))
+    #pi=(mpd*segsites)/n_accessible_sites                               #alternate pi if not all sites are accessible
+    pi=allel.sequence_diversity(positions,allele_counts,start=1,stop=1e8)
+    tajD=allel.tajima_d(ac=allele_counts,start=1,stop=maxlen) #NOTE check for 0 v 1 positions
[…]

### 2019-09-12 scripts/estimate_gen_times.py

@@ -8,69 +8,70 @@ parser=argparse.ArgumentParser()
 parser.add_argument("--infile",dest="infile",type=str)
 args=parser.parse_args()
 
-args=argparse.Namespace(infile="/Users/cj/spaceness/sims/slimout/spatial/W50_run3/sigma_0.25_.trees_1661344")
+#args=argparse.Namespace(infile="/Users/cj/spaceness/sims/slimout/spatial/W50_muts/sigma_0.4494235649796564_.trees_5319950_1272000")
 
 sigma=float(re.split("_",os.path.basename(args.infile))[1])
 
 id=int(np.random.uniform(1e7,1e8-1,1))
 sp.run("cd ~/spaceness/; slim -d sigma="+str(sigma)+" -d id="+str(id)+" slim_recipes/gen_timer_spatial.slim",
        shell=True)
-#ts=pyslim.load("/Users/cj/spaceness/tmp"+str(id)+".trees")
-ts = sps.SpatialSlimTreeSequence(pyslim.load("/Users/cj/spaceness/tmp"+str(id)+".trees"), dim=2)
+filename=re.search(os.listdir("/Users/cj/spaceness/tmp")[0],str(id))
+ts = sps.SpatialSlimTreeSequence(pyslim.load("/Users/cj/spaceness/tmp/"+str(sigma)[0:6]+"_"+str(id)+".trees"), dim=2)
 
 ############ average generation time across all individuals
 node_times = ts.tables.nodes.time
 parent_nodes = ts.tables.edges.parent
 child_nodes = ts.tables.edges.child
-max_T = 200
+max_T = 150
 use_these = (node_times[parent_nodes] < max_T)
 parent_ages = node_times[parent_nodes] - node_times[child_nodes]
-genealogical_gen_time=np.mean(parent_ages[use_these])
-
-############ variance in reproductive output
-mid_node_ids=np.array([x.id for x in ts.nodes() if x.time in range(100,101)])
-
-
-np.var(n_offspring)
-np.mean(n_offspring)
-
-
-############# 'effective' generation time sampling inds proportional to final-generation ancestry
-alive_nodes=[x.id for x in ts.nodes() if x.time==0] #final-generation nodes
-#alive = ts.individuals_alive(0)
-#alive_nodes = ts.individual_nodes(np.where(alive)[0])
-node_ancestry = ts.proportion_ancestry_nodes(alive_nodes)[0]
-ind_ancestry = np.zeros(ts.num_individuals)
-for k, ind in enumerate(ts.individuals()):
[…]

### 2019-09-12 scripts/gwas_viz.R

@@ -2,6 +2,7 @@
 library(ggplot2);library(magrittr);library(plyr);library(data.table)
 library(qvalue);library(cowplot);library(wesanderson)
 setwd("~/projects/spaceness/gwas/")
+setwd("~/spaceness/gwas")
 pal <- grDevices::colorRampPalette(color=c("steelblue4","skyblue","lightgoldenrod1","orangered","red4"),
                                    bias=1,space="rgb",interpolate="linear")
 theme_set(theme_classic()+theme(
@@ -30,71 +31,71 @@ panel.qqconf<-function(n, conf.points=1000, conf.col="gray", conf.alpha=.05, ...
   #grid.polygon(x=mpts[,1],y=mpts[,2], gp=gpar(fill=conf.col, lty=0), default.units="native")
   return(mpts)
 }
-setwd("random_sampling")
-files <- list.files("out_corner",full.names = T) %>% grep(".assoc.linear",.,value = T)
-qq_corner <- data.table(P=NA,sigma=NA,observed=NA,expected=NA,neighbors=NA)[-1]
-for(f in files){
-  sigma <- strsplit(f,"_") %>% unlist() %>% .[3] %>% as.numeric()
-  if(sigma < 4){
-    gwas <- fread(f)
-    gwas <- subset(gwas,TEST=="ADD")
-    expected <- -log10(1:length(gwas$P)/length(gwas$P))
-    observed <- -log10(sort(gwas$P))
-    qq_corner <- rbind(qq_corner,data.table(P=gwas$P,sigma=sigma,observed=observed,expected=expected,neighbors=4*pi*sigma^2*5))
-    print(f)
-  }
-}
-qq_corner$phenotype <- "corner"
-
-files <- list.files("out_clinal",full.names = T) %>% grep(".assoc.linear",.,value = T)
-qq_clinal <- data.table(P=NA,sigma=NA,observed=NA,expected=NA,neighbors=NA)[-1]
-for(f in files){
-  sigma <- strsplit(f,"_") %>% unlist() %>% .[3] %>% as.numeric()
-  if(sigma < 4){
-    gwas <- fread(f)
-    gwas <- subset(gwas,TEST=="ADD")
-    expected <- -log10(1:length(gwas$P)/length(gwas$P))
-    observed <- -log10(sort(gwas$P))
[…]

### 2019-09-12 scripts/ibs_viz.R

@@ -41,8 +41,14 @@ ss <- ss[,!grepl("sfs",colnames(ss))]
 ss$neighborhood_size <- 4*pi*ss$sigma^2*5
 ss$col <- paste(ss$model,ss$sampling)
 
-#test for differences by sampling in spatial and random mating summary stat distributions
+
+#check Ne scaling (??)
+ss$theta <- 4*ss$census_n/2.5*1e-8 #variance in offspring for 
 mss <- melt(ss,id.vars=c("sigma","model","sampling","neighborhood_size","col"))
+ggplot(data=subset(mss,variable %in% c("pi","theta")),aes(x=2*pi*sigma^2*5,y=value,col=variable))+
+  geom_point(size=0.3,shape=1)+facet_wrap(~model)+scale_x_log10()
+
+#test for differences by sampling in spatial and random mating summary stat distributions
 mss <- subset(mss,variable!="census_n")
 sampling_diffs <- ddply(mss,.(variable,model),function(e){
   meantest <- summary(aov(value~sampling,data=e))[[1]][,5][1]
@@ -51,7 +57,7 @@ sampling_diffs <- ddply(mss,.(variable,model),function(e){
 }) #40 warnings for forcing sampling to factor, which is fine
 sampling_diffs <- arrange(sampling_diffs,model,variable)
 colnames(sampling_diffs) <- c("variable","model","p(different\nmeans)","p(different\nvariance)")
-xtable(sampling_diffs,digits=6)
+#xtable(sampling_diffs,digits=6)
 
 #drop alternate sampling schemes for random mating simulations
 ss <- rbind(ss_m,ss_r,ss_p,rm_r)
@@ -84,7 +90,7 @@ mss$variable <- mapvalues(mss$variable,from=c("segsites","pi","thetaW","tajD",
                                               "census_n"),
                           to=c("segregating\nsites","pi","Watterson's\ntheta","Tajima's\nD",
                              "observed\nheterozygosity","Fis","var(dxy)","skew(dxy)",
-                             "mean(IBS)","var(IBS)","skew(IBS)","nIBS","mean(IBS>1e6)",
+                             "mean(IBS)","var(IBS)","skew(IBS)","nIBS","nIBS>1e6",
                              "corr(dxy,dist)","corr(mean(IBS),dist)","corr(nIBS,dist)",
[…]

### 2019-09-12 scripts/msp_grid_sim.py

@@ -117,7 +117,7 @@ def get_coords_per_sample(width,haploids_per_pop):
 
 #########################running a simulation#################################
 #simulation params equivalent to W50 slim runs w/250-ind demes
-width=10
+width=zz
 haploids_per_pop=20
 joint_N=100000
 rate=10/(joint_N/width**2) #numerator is n migrants per generation

### 2019-09-12 scripts/msp_ts_to_sumstats.py

@@ -1,22 +1,33 @@
+#import os;os.chdir("/Users/cj/spaceness/scripts/")
 from slimtools import *
 import argparse
 parser=argparse.ArgumentParser()
 parser.add_argument("--infile")
 parser.add_argument("--outfile")
+parser.add_argument("--width",type=int)
 parser.add_argument("--pop_config")
 args=parser.parse_args()
 
 ts=msp.load(args.infile)
-ts=msp.mutate(ts,1e-8)
+#ts=msp.load("/Users/cj/spaceness/sims/sigma_0.21313291689796326_sim_911380669.trees")
+#ts=msp.mutate(ts,1e-8)
 
-
+#sample 60 individuals
+nodes_to_sample=np.random.choice(np.arange(0,499,2),60,replace=False)
+node_partners=nodes_to_sample+1
+nodes_to_sample=np.concatenate((nodes_to_sample,node_partners))
+nodes_to_sample=np.sort(nodes_to_sample)
+nodes_to_sample=np.array(nodes_to_sample,dtype="int32")
+ts=ts.simplify(nodes_to_sample)
 haps=np.array(ts.genotype_matrix())
 positions=np.array([s.position for s in ts.sites()])
 
+#get locations etc for all individuals
 if args.pop_config=="grid":
-    label=float(re.sub(".trees","",os.path.basename(args.infile)))
+    #label=float(re.sub(".trees|sigma_|","",os.path.basename(args.infile)))
+    label=float(re.split("_",os.path.basename(args.infile))[1])
     pops=[ts.node(x).population for x in ts.samples()]
-    locs=[[(np.floor(x/5)+.5)*10,((x%5)+.5)*10] for x in pops]
+    locs=[[(np.floor(x/args.width)+.5)*(50/args.width),((x%args.width)+.5)*(50/args.width)] for x in pops]
     locs=np.array(locs)[np.arange(0,len(locs),2)]
 if args.pop_config=="onepop":
     locs=np.array([[np.random.uniform(1,50,1)[0],np.random.uniform(1,50,1)[0]] for x in range(50)]) #random locations for nonspatial sim(?)
@@ -28,99 +39,99 @@ getSLiMSumStats(haps=haps,
                 locs=locs,
                 outfile=args.outfile,
                 maxlen=1e8,
[…]

### 2019-09-12 scripts/plot_msp_sumstats.R

@@ -2,14 +2,93 @@
 setwd("~/spaceness/")
 library(ggplot2);library(plyr);library(magrittr);library(data.table)
 
+theme_set(theme_classic()+theme(axis.text=element_text(size=8),
+                                axis.title=element_text(size=8),
+                                strip.text = element_text(size=8),
+                                title = element_text(size=9,hjust=0.5),
+                                legend.title = element_text(size=8),
+                                legend.text = element_text(size=8),
+                                strip.background = element_blank()))
+
 ss <- fread("sumstats/ss_msp_grid.txt",data.table = F)
+
 ss <- ss[,!grepl("sfs",colnames(ss))]
-ss$sigma2 <- ss$sigma*4
-mss <- melt(ss,id.vars=c("sigma","sigma2"))
-ggplot(data=mss,aes(x=sigma2*10,y=value))+
-  theme_classic()+
+mss <- melt(ss,id.vars=c("sigma"))
+
+ggplot(data=mss,aes(x=sigma,y=value))+
   xlab("sigma")+
+  scale_x_log10(labels = function(x) format(x, scientific = FALSE))+
   facet_wrap(~variable,scales="free")+
   geom_point(shape=21,stroke=0.3,size=1)+geom_smooth(span=.2,fill=NA,col="orange",lwd=0.5)
+
+spaceness <- fread("sumstats/ss_spatial_random_W50.txt")
+spaceness <- spaceness[,c("sigma","segsites","pi","thetaW","tajD","het_o","gen_sp_corr")]
+spaceness$width <- Inf
+
+ss5 <- fread("sumstats/ss_msp_grid_Ne6100_w5.txt",data.table=F)
+ss5 <- ss5[,!grepl("sfs",colnames(ss5))]
+ss5$width <- 5
+ss10 <- fread("sumstats/ss_msp_grid_Ne6100_w10.txt",data.table=F)
+ss10 <- ss10[,!grepl("sfs",colnames(ss10))]
+ss10$width <- 10
+ss20 <- fread("sumstats/ss_msp_grid_Ne6100_w20.txt",data.table=F)
+ss20 <- ss20[,!grepl("sfs",colnames(ss20))]
+ss20$width <- 20
+#ss50 <- fread("sumstats/ss_msp_grid_Ne6100_w50.txt",data.table=F)
[…]

### 2019-09-12 scripts/plot_speciation_times.R

@@ -1,0 +1,12 @@
+setwd("~/Dropbox/speciation_cyclical_migration/simple_model_output/")
+library(tidyverse);library(data.table)
+as0 <- fread("allopatry_s0.txt")
+
+ggplot(data=as0,aes(x=V1))+
+  theme_classic()+
+  geom_histogram(fill=NA,col="black",bins=100)+
+  geom_vline(xintercept=mean(as0$V1),col="red")+
+  annotate(geom="text",x=mean(as0$V1)+1.2*sd(as0$V1),y=50,label=paste("mean =",round(mean(as0$V1))))
+
+hist(as0$V1,breaks=50)
+mean(as0$V1)

### 2019-09-12 scripts/run_msp_grid.py

@@ -1,3 +1,5 @@
+#import os
+#os.chdir("/Users/cj/spaceness/scripts/")
 from slimtools import *
 # import argparse
 # parser=argparse.ArgumentParser()
@@ -55,49 +57,16 @@ def get_grid_migration_matrix(width,rate,diagonal):
                     mig_matrix[i][i+(width+1)]=np.sqrt(2)*rate
     return mig_matrix
 
-##migration matrix for toroidal landscapes (no edges)
-def get_torus_migration_matrix(npops,rate,diagonal):
-    npops=26
-    if not np.sqrt(npops)%1 == 0:
-        print("error - npops should make a square")
-        return
-    width=round(np.sqrt(npops))
-    mig_matrix=np.zeros((width*width,width*width))
-    mig_matrix[i][i+1]=rate
-    mig_matrix[i][i-1]=rate
-    mig_matrix[i][i+width]=rate
-    mig_matrix[i][i-width]=rate
-    if diagonal:
-        mig_matrix[i][i-(width+1)]=np.sqrt(2)*rate
-        mig_matrix[i][i+(width-1)]=np.sqrt(2)*rate
-        mig_matrix[i][i-(width-1)]=np.sqrt(2)*rate
-        mig_matrix[i][i+(width+1)]=np.sqrt(2)*rate
-    return mig_matrix
-
-#returns coordinates per node (haploid chromosome)
-def get_coords_per_sample(width,haploids_per_pop):
-    samples=np.arange(0,haploids_per_pop*width**2,1)
-    locs=[]
-    x=0
-    y=0
-    for x in range(width):
-        for y in range(width):
-            for z in range(int(haploids_per_pop)):
-                locs.append([x,y])
-    return np.array(locs)
-
 #########################set params#################################
-#simulation params equivalent to W50 slim runs w/250-ind demes
 width=50
-haploids_per_pop=10
-joint_N=2.5e6
-#nm=np.random.uniform(0,100)
-nm=100
-rate=nm/(joint_N/width**2) #numerator is n migrants per generation
+haploids_per_pop=2
+joint_N=6.1e3
+#NS=np.random.uniform(2,1000,400)
+#sigma=np.sqrt(NS/(4*3.14159*(joint_N/(50*2))))
+sigma=np.random.uniform(0.2,4)
+m=sigma/((50/width)*4)
 
-mig_matrix=get_grid_migration_matrix(width,rate,diagonal=True)
-
[…]

### 2019-09-12 scripts/slimtools.py

@@ -329,11 +329,11 @@ def getSLiMSumStats(haps,positions,label,locs,outfile,maxlen,ibs_tracts=True,ver
     segsites=np.shape(genotypes)[0]
     #mpd=np.mean(allel.mean_pairwise_difference(allele_counts))
     #pi=(mpd*segsites)/n_accessible_sites                               #alternate pi if not all sites are accessible
-    pi=allel.sequence_diversity(positions,allele_counts,start=1,stop=1e8)
+    pi=allel.sequence_diversity(positions,allele_counts,start=1,stop=maxlen)
     tajD=allel.tajima_d(ac=allele_counts,start=1,stop=maxlen) #NOTE check for 0 v 1 positions
     thetaW=allel.watterson_theta(pos=positions,ac=allele_counts,start=1,stop=maxlen)
     het_o=np.mean(allel.heterozygosity_observed(genotypes))
-    fis=np.mean(allel.inbreeding_coefficient(genotypes))
+    fis=np.nanmean(allel.inbreeding_coefficient(genotypes))
     sfs=allel.sfs(allele_counts[:,1])
     sfs=np.append(sfs,[np.repeat(0,np.shape(haps)[1]-len(sfs)+1)])
     #Isolation by distance
@@ -409,7 +409,7 @@ def getSLiMSumStats(haps,positions,label,locs,outfile,maxlen,ibs_tracts=True,ver
             sfsnames=["sfs_"+str(x)+" " for x in range(np.shape(haps)[1])]
             sfsnames.append("sfs_"+str(np.shape(haps)[1]))
             out=open(outfile,"w")
-            out.write("sigma segsites pi thetaW tajD het_o fis gen_dist_mean gen_dist_var gen_dist_skew gen_sp_corr"+"".join(sfsnames)+"\n")
+            out.write("sigma segsites pi thetaW tajD het_o fis gen_sp_corr "+"".join(sfsnames)+"\n")
         row=[label,segsites,pi,thetaW,tajD,het_o,fis,gen_sp_corr]
         row=np.append(row,sfs)
         row=" ".join(map(str, row))

### 2019-09-12 scripts/testmuts.py

@@ -1,32 +1,64 @@
+import os
+os.chdir("/Users/cj/spaceness/scripts/")
 from slimtools import *
+os.chdir("/Users/cj/spaceness/")
+import pandas as pd
+#get generation time predictor from all-individual simulations
 
-muts=pyslim.load("/Users/cj/Desktop/sigma_0.793843124378123_.trees_7796460_822000")
-muts.num_mutations
-muts=sample_treeseq(infile=muts,
-               outfile="",
-               nSamples=60,
-               recapitate=False,
-               recombination_rate=1e-8,
-               write_to_file=False,
-               sampling="random",
-               sampling_locs=[[12.5,12.5],[12.5,37.5],[37.5,37.5],[37.5,12.5]],
-               plot=False,
-               seed=12345)
-muts.num_mutations
+gentimes=pd.read_csv("/Users/cj/spaceness/W50sp_slimMuts_gentimes.txt",sep=" ",header=None,float_precision='round_trip')
 
-nomuts=pyslim.load("/Users/cj/spaceness/sims/slimout/random_mating/W50_run3/sigma_0.7857423531825174_.trees_9690340")
-nomuts=msp.mutate(nomuts,1e-8/4.8)
-nomuts.num_mutations
-nomuts=sample_treeseq(infile=nomuts,
-               outfile="",
-               nSamples=60,
-               recapitate=False,
-               recombination_rate=1e-8,
-               write_to_file=False,
-               sampling="random",
-               sampling_locs=[[12.5,12.5],[12.5,37.5],[37.5,37.5],[37.5,12.5]],
-               plot=False,
-               seed=12345)
-nomuts.num_mutations
+slim_muts=[];msp_muts=[];sigmas=[];msp_muts_unscaled=[]
+files=[x for x in os.listdir("/Users/cj/spaceness/sims/slimout/spatial/W50_muts/") if not x.startswith(".")]
+for sim in tqdm(files):
+  sigma=float(re.split("_",os.path.basename(sim))[1])
+  gt=gentimes[np.round(gentimes[0],5)==np.round(sigma,5)][1]
+  a=pyslim.load(os.path.join("/Users/cj/spaceness/sims/slimout/spatial/W50_muts/",sim))
+  slim_muts.append(a.num_mutations)
+  msp_a=msp.mutate(a,1e-8/(gt),keep=False)
+  msp_muts.append(msp_a.num_mutations)
[…]

### 2019-09-19 ms/spaceness.tex

--- 2019-08-14 ms/spaceness.tex
@@
      }
-
-\documentclass[11pt,twoside,lineno,hidelinks]{preprint}
+\newcommand{\beginappendix}{%
+        \setcounter{table}{0}
+        \renewcommand{\thetable}{A\arabic{table}}%
+        \setcounter{figure}{0}
+        \renewcommand{\thefigure}{A\arabic{figure}}%
+     }
+\newcommand{\stopappendix}{%
+        \setcounter{table}{0}
+        \renewcommand{\thetable}{\arabic{table}}%
+        \setcounter{figure}{0}
+        \renewcommand{\thefigure}{\arabic{figure}}%
+     }
+
+\documentclass[10pt,twoside,lineno,hidelinks]{preprint}
 \usepackage{amsmath}
@@
 \usepackage{afterpage}
-\setstretch{1.5}
+\setstretch{1}
 \NeedsTeXFormat{LaTeX2e}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Supplement %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\beginappendix
+\section{Appendix 1}
+\subsection{Comparisons with Alternate Spatial Models}
+Stepping-stone models have been extremely useful in population genetics and, when simulated in reverse time, remain much faster than our implementation of continuous-space evolution in forward time.
+Similarly, Wright-Malécot models have lead to analytic results that have proved reasonable approximations for estimating neighborhood sizes from empirical data \citep{Rousset1997}.
+Here we address how these classes of models compare with our implementation of density-dependant continuous space evolution, and try to assess when they are "good enough".
+
+\subsubsection{Stepping Stone Models}
+In this class of models we imagine an $n \times n$ grid of populations exchanging migrants with their vertical and horizontal neighbors at a rate $m$.
+This model can be implemented in standard coalescent simulators like msprime \citep{Kelleher2016} by setting the migration matrix to allow migration only between adjacent populations.
+To compare with our continuous-space simulation we then need to estimate neighborhood sizes for the stepping-stone setting.
+Recall that $NS\approx4\pi\sigma^2d$.
+To set $d$ (population density) equal among models, we first estimated the effective population size of a random-mating non-Wright-Fisher SLiM model with a census size of 10,000 diploid individuals.
+We used the $\theta_{\pi}$ estimator and took the mean across all random-mating SLiM simulations (roughly 6100).
+By distributing this $N_{e}$ evenly among demes in a stepping-stone model we achieve roughly equivalent population densities.
+
+Next we need to relate migration rates to dispersal distances such that the total number of individuals within an expected dispersal radius is equivalent between stepping-stone and continuous models.
+Recall that $\sigma$ is the expected dispersal distance for an offspring from its parent in our continuous model.
+In our stepping stone model the probability of an individual within a deme being a migrant is $4m$ for all interior populations (i.e.
+two vertical neighbors and two horizontal neighbors, each with probability $m$ of being the source of a migrant), and all migrants move one deme across the grid.
+To translate this distance to the area containing an equivalent number of individuals in the continuous model we need to scale it by the ratio of landscape widths.
+That is, for a $5 \times 5$ stepping-stone landscape moving one population is equivalent to moving 20\% of the total landscape width, or 10 map units on the $50 x 50$ continuous landscape in our spatial SLiM model.
+If $W_{continuous}$ is the width of a continuous landscape and $W_{grid}$ is the number of demes per side in a stepping-stone model with equivalent total $N_{e}$, $m$ can then be related to $\sigma$ as $\sigma\approx4m\frac{W_{continuous}}{W_{grid}}$.
+By rearranging we can then sample $\sigma$ values from the same uniform distribution used for our continuous-space simulations and convert to the corresponding migration rate with $m=\sigma/(4\frac{W_{continuous}}{W_{grid}})$.
+
+Using this approach we ran stepping-stone models in msprime by evenly distributing a total Ne of 6100 among $n^2$ populations while drawing $\sigma$ from a $U(0.2,4)$ as in our continuous space model.
+We ran 500 simulations each for $5 \times 5$, $10 \times 10$, and $20 \times 20$ stepping-stone landscapes.
+We then calculated the same set of summary statistics as in our continuous model (assigning all individuals in a deme the location of the center of that deme), and plotted them on the same axes for comparison (Figure X).
+
+%the two-dimensional Wright-Malécot model.
+SLiM v3 includes an implementation of a similar model in recipe 14.1.
+This model simulates a Wright-Fisher population with discrete generations and fixed total population size.
+Individuals are initially distributed randomly across the landscape.
+In each generation, mates are selected proportional to distance and offspring disperse a normally-distributed distance from the first parent.
+The total number of offspring per pair is Poisson distributed
+\stopappendix
+\clearpage
+
 \section{Supplementary Figures and Tables}

### 2019-09-19 ms/spaceness.tex

--- 2019-08-14 ms/spaceness.tex
@@
 by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time.
-(We verified that this procedure produced the correct number of mutations by comparing to a subset of simulations with SLiM-generated mutations, which are applied only at meiosis.)
+We show that this procedure produces the same site-frequency spectrum as applying mutations directly in SLiM in Figure XXXXX.
 Simulations were run for 1.6 million timesteps (approximately $30N$ generations).
@@
 The effects of sampling on summary statistic estimates were summarized by testing for differences in mean (ANOVA, \citep{Rcore2018}) and variance (Levene's test, \citep{Fox2011}) across sampling strategies for each summary statistic.
+
+Last, we used a subset of summary statistics to compare how our continuous model performs relative to a reverse-time stepping stone model (Appendix 1).
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Supplement %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\beginappendix
+\section{Appendix 1}
+
+%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Comparisons with Stepping-Stone Models}
+We checked that our model produces reasonable results by comparing it to a reverse-time stepping-stone model implemented in msprime \citep{Kelleher2016}.
+In this class of models we imagine an $n \times n$ grid of populations exchanging migrants with neighboring populations at rate $m$.
+If these models are good approximations of the continuous case we expect that results will converge as $n \to \infty$, so we ran simulations while varying $n$ from 5 to 50 (Table \ref{table:grid_params}).
+To compare with continuous models we first distributed the the same "effective" number of individuals across the landscape as in our continuous-space simulations ($\approx 6100$, estimated from $\theta_{\pi}$ of random-mating continuous-space simulations).
+We then approximate the mean per-generation dispersal distance $\sigma$ given a total landscape width $W$ as the product of the probability of an individual being a migrant and the distance traveled by migrants: $\sigma \approx 4m(W/n)$.
+We ran 500 simulations per $n$ while sampling $\sigma$ from $U(0.2,4)$.
+We then randomly selected 60 diploid individuals from each simulation (approximating diploidy by combining pairs of chromosomes with contiguous indices within demes) and calculated a set of six summary statistics using the scripts described in the summary statistics portion of the main text.
+
+\begin{table}[h]
+\centering
+\begin{tabular}{ccc}
+     demes per side ($n$)&$N_e$ per deme  & samples per deme  \\
+     \hline
+     5 & 244 & 20 \\
+     10 & 61 & 10 \\
+     20 & 15.25 & 2 \\
+     50 & 2.44 & 1 \\
+\end{tabular}
+\captionsetup{justification=centering}
+\caption{stepping-stone simulation parameters}
+\label{table:grid_params}
+\end{table}
+
+\begin{figure*}[h]
+\centering
+\includegraphics[width=\textwidth]{ms/figures/grid_ss_curves.pdf}
+\caption{Summary statistics for 2-dimensional coalescent stepping-stone models with fixed total $N_{e}$ and varying numbers of demes per side.
+The black "infinite" points are from our forward-time continuous space model.
+Inter-deme migration rates are related to $\sigma$ as described above.}
+\label{fig:grid_stats}
+\end{figure*}
+
+In general we find many of the qualitative trends are similar among continuous and stepping-stone models and that, in most cases, statistics from stepping-stone models approach the continuous model as the resolution of the grid increases.
+For example, $\theta_\pi$ is inflated at low neighborhood sizes (i.e.
+low $m$), and the extent of the inflation increases to approach the continuous case as the resolution of the landscape increases.
+Similar pattterns are observed for $F_{is}$ and observed heterozygosity.
+However, $\theta_{W}$ behaves differently, with increased grid resolution leading to lower values.
+This in turn drives an even more positive Tajima's $D$ in grid simulations at small neighborhood sizes.
+
+\begin{figure*}[h]
+\centering
+\includegraphics[]{ms/figures/grid_thetaW_tajD_zoom.pdf}
+\caption{Zoomed-in view of $\theta_{W}$ and Tajima's $D$ for stepping-stone models under varying neighborhood sizes.}
+\label{fig:grid_stats_zoom}
+\end{figure*}
+
+These differences relative to our continuous model mainly reflect two shortcomings of the reverse-time stepping stone model.
+If we simulate a coarse grid with relatively large populations in each deme, we cannot accurately capture the dynamics of small neighborhood sizes because mating within each deme remains random regardless of the migration rate connecting demes.
+This likely explains the trends in $\pi$, observed heterozygosity, and $F_{is}$.
+However increasing the number of demes while holding the total number of individuals constant results in small within-deme populations for which even the minimum sample size of 1 approaches the local $N_e$ (Table \ref{table:grid_params}).
+This results in an excess of short terminal branches in the coalescent tree, which decreases the total branch length and leads to fewer segregating sites, deflated $\theta_W$, and inflated Tajima's $D$.
+Overall then our continuous model reproduces important features of spatial structure approximated by reverse-time stepping-stone models at moderate neighborhood sizes while more accurately reflecting the dynamics of populations with small neighborhood sizes.
+
+\stopappendix
+\clearpage
+
 \section{Supplementary Figures and Tables}

### 2019-10-10 ms/review-responses.tex
edits to andy and peter

@@
 \reply{
+Done.
 }
@@
 
+\cjb{Peter - did you have any in mind here? I can come up with a couple but don't know the ecology lit on this very well}
 \reply{
+
 }
@@
 \reply{
+We have added a figure (Figure \ref{fig:runtimes}) and short discussion \llname{runtimesxn} of run times.
 }
@@
 \reply{
+    \cjb{working on this}
 }
@@
 \reply{
+Good suggestion -- the SFS in Fig 3A is now on a log-log scale, which shows the slight decrease in low frequency SNPs a little better. 
 }
@@
 \reply{
+    Thanks for catching this - we have revised this caption to add details including the simulation parameters. \llname{ll:figs3caption}
 }
@@
 \reply{
+The mean of the IBS tract distribution is calculated by building a (very large) vector of lengths of all IBS tracts between all pairs of individuals and taking the mean over this entire vector.
+This distribution is quite strongly skewed (i.e.
+see figure 4 and the skew(IBs) panel of fig S1) such that the highest density is in low values.
+We explain the depressed mean value by noting that at low neighborhood sizes the coalescence time of distant individuals is inflated, which allows more time for recombination and mutation to break up IBS tracts.
+The flip side of this is that there are then more IBS tracts in total across a random sample of individuals, explaining the corresponding inflation in the number of tracts.
+Perhaps confusing we also see an increase in the number of very long IBS tracts at low neighborhood sizes, which corresponds to an excess of very recent coalescent events for groups of nearby individuals.
+\cjb{need to think of something to say we did about this in the text}
+
 }
@@
 \reply{
+    We have added to the text at the beginning of the results to emphasize that this analysis is necessary because these parameters are emergent rather than fixed \llname{emergentparams}. 
 }
@@
 \reply{
+Good suggestion -- we have changed the section heading to "demographic inference" 
 }
@@
 \reply{
+We have revised this sentence to clarify that we aim to produce a variation in mean phenotype of two standard deviations across the landscape \llname{phenotypescalingsd}.
 }
@@
 \reply{
+ah, a confusing legend title placement! we moved "model" to after "spatial" as suggested.
 }
@@
 \reply{
+    This is a good suggestion, but we decided to not include this as we don't have straightforward expectations for the other parameters shown. 
 }
@@
 \reply{
+    done. \llname{spatialgenetic}
 }
@@
 \reply{
+\cjb{check on final pass}
 }
@@
 \reply{
+    Thank you for noting these -- they have been corrected. 
 }
@@
 \reply{
+This is an excellent point also raised by other reviewers.
+We have added an appendix section comparing our model to a reverse-time stepping stone analysis on several relevant summary statistics \llname{appendix1}.
+Briefly, we find that many of the qualitative trends we see also occur in a stepping-stone setting, suggesting that they are driven purely by the particularities of our parameterization.
+We also see several interesting departures from the stepping-stone model reflecting shortcomings in the gridded approach to approximating continuous landscapes, which we think our model helps to avoid.
+We are also curious how all the other analyses we test would be affected using other simulation schemes, but hope to explore that aspect in future work.
 }
@@
 \reply{
+    See above.
+It is certainly easier and faster, but comes with significant issues either when neighborhood size is lower than the population of a deme, or when demes are small enough that the sample size approaches local $N_e$ (at least in reverse-time simulations).
 }
@@
 \reply{
+    \cjb{peter - any suggestions here?}
 }
@@
 \reply{
+Good suggestion -- we have added a supplemental figure to show the distribution of individual fitness across the landscape with and without our edge-scaling approach.
+MAKE THIS FIG AND POINT TO IT HERE.
 }
@@
 \reply{
+    We have added a figure showing sample site-frequency-spectra generated from a subset of simulations run with SLiM mutations, and then using msprime to apply mutations to the same tree sequences with our generation-time scaling approach.
+As you can see in \ref{fig:slim_v_msp_sfs}, these approaches yield extremely similar spectra so we believe any error in the approximation is quite small.
+We also now show variation in simulation times with different mutation and tree sequence recording schemes in Figure \ref{fig:runtimes}, and included a brief discussion of this in \llname{runtimesxn}
 }
@@
 \reply{
+
 }
@@
 \reply{
+    We have checked the code to the best of our abilities and did not find a bug causing the underdispersion.
+It seems to reflect overcorrection in the regression when using PC coordinates as covariates -- the PCA is capturing some information about the spatial genetic variation which itself covaries only weakly with the phenotype, and as a result we see anomalously low $-log10(p)$ when regressing genotype against phenotype.
+\cjb{Peter or Andy -- any advice here?
+}
 }
@@
 \reply{
+Done - thank you for pointing us to this study. \llname{haworthcite}
 }
@@
 \reply{
+    We are also very interested to know how mixed models perform here, but think that adding a second GWAS method would make this section too large for the current paper.
+We have added a note to \llname{mixedmodelcallout} specifically citing mixed models as alternate methods that may perform better.
 }
@@
 \reply{
+Apologies, I had forgotten to make the repo public. The link should work for everyone now. 
 }
@@
 \reply{
+    This is a good point.
+We evaluated this while developing the simulation using the built-in visualization tools of the SLiM GUI, and have now added a supplementary figure (Figure \ref{fig:wm_v_dd}) showing the distribution of individuals in our density-dependant spatial model and a continuous-space Wright-Fisher simulation without density dependance.
 }
@@
 \reply{
+    Similarly here we monitored the population size over time while running SLiM.
+All simulations maintained asymptotic populations, and none of the runs we started ended because populations crashed.
+\cjb{is there something else we should show here?}
 }
@@
 \reply{
+    \cjb{Peter -- ideas here?}
 }
@@
 \reply{
+I apologize -- I had forgotten to set the repo to public. It should work now. 
 }
@@
 \reply{
+    This is a good point also raised by other reviewers.
+We have added an Appendix comparing our model to a stepping-stone, rather than the spatial-Labmda-Fleming-Viot, because we think these are the most familiar and widely used class of spatial models.
+See Appendix 1.
+We find that many features of our model are well approximated by stepping-stone models, and that for statistics like $\theta_{\pi}$ the stepping stone model results approach our continuous space model as the number of demes used to describe the landscape increases.
 }
@@
 \reply{
+    Census population in the random mating model scales with "neighborhood size" because of two effects -- the edge scaling and the competition
+
+    \cjb{need a better answer here...}
 }
@@
 \reply{
+This is a good point.
+Wright's work \citep{Wright1943} suggests some aspects of genetic variation such as variance in allele frequencies and inbreeding coefficients can be estimated by looking only at what he would later \citep{Wright1946} call "neighborhood size", but certainly other aspects like the number of segregating sits will also depend on total landscape size.
+We now note on \llname{popsizecaveat} that we have evaluated only one landscape size, and have added a sentence to the discussion noting that exploration of these patterns in varying landscape sizes is an important avenue for further research \llname{landscapesizecaveat}.
+
 }
@@
 \reply{
+Done. Thank you for pointing us to this study. 
 }
@@
 \reply{
+    \cjb{working on this}
+    Thank you for this suggestion.
+We have slightly revised the introduction and hope it is now clearer; however since we want to cover a little history and motivation for our continuous model vs stepping-stone approaches in the intro it does have a different flow from the discussion.
+
 }
@@
 \reply{
+    Good point on EEMS. We have added the most recent EEMS paper to the citations on \llname{eemscitation}
 }
@@
 \reply{
+    We now discuss haplotype methods in the relevant discussion section \llname{ibdnesxn}.
+However though these methods should be more accurate for recent events it is not clear that this will improve performance per se.
+The dips in recent inferred Ne from stairwayplot are not just prediction noise, but actually reflect an underlying genealogy in which terminal branches are shorter than expected from a constant-size random-mating population (see e.g.
+figure \ref{fig:sumstats}A and \ref{fig:ibs}).
+The interpretation error is that these short branches are generated by spatial structure rather than changes in population size over time -- a point also made in the \citep{Mazet2015} paper we discuss in the introduction and discussion.
 }
@@
 \reply{
+\cjb{is this an ok response?
+two reviewers asked about mixed models but I feel like including them would require expanding the gwas section quite a bit and I'd rather now.
+But if either of you feel like it's critical I could probably get it done next week}
+    Good question -- we are also interested to know how mixed models perform here, but think that to properly test that we would want to change our design to generate phenotypes from simulated genotypes.
+This would allow us to evaluate false-negatives in addition to false-positives.
+This is important because, if mixed models do provide stronger control for stratification they are also likely to remove true signals of causal SNPs if those SNPs covary with spatial structure.
+We now point to these methods explicitly in the discussion \llname{mixedmodelcallout}, but think that incorporating that study here would make this paper too long.
+We also think the PC results are still quite relevant as the method is still seen in many studies.
 }
@@
 \reply{
+Apologies, it was accidentally set to private. The link should work now. 
 }
@@
 \reply{
+    We have included a table describing the summary statistics in Figure \ref{table:sumstats}.
 }
@@
 \reply{
+\cjb{Peter and Andy -- ideas here?}
 }
@@
 \reply{
+We have rearranged the figure as suggested.
+
 }
@@
 \reply{
+Thanks, this sentence has been revised and fixed. 
 }
@@
 \reply{
+Done. 
 }

### 2019-10-10 ms/spaceness.tex
edits to andy and peter

@@
 \setstretch{1}
+\usepackage{placeins}
 \NeedsTeXFormat{LaTeX2e}
@@
 This pattern of ``isolation by distance'' \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}.
-Despite a long history of analytical work describing the genetics of populations distributed across continuous geography (e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation}), much modern work still describes geographic structure as a set of discrete populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations}.
+Despite a long history of analytical work describing the genetics of populations distributed across continuous geography (e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation,Wilkins2002linearcoalescent,Wilkins2004septimescales}), much modern work still describes geographic structure as a set of discrete populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations,al2019estimating}\llabel{eemscitation}.
 For this reason, most population genetics statistics are interpreted with reference to discrete, well-mixed populations, and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000} with methods that assume these are randomly mating units.
@@
 by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time.
-We show that this procedure produces the same site-frequency spectrum as applying mutations directly in SLiM in Figure XXXXX.
+We show that this procedure produces the same site-frequency spectrum as applying mutations directly in SLiM in Figure \ref{fig:slim_v_msp_sfs}.
 Simulations were run for 1.6 million timesteps (approximately $30N$ generations).
+
+To check that our model produces reasonable results, we compared its output to that of a stepping-stone model implemented in msprime \citep{Kelleher2016}.
+These results are discussed in detail in Appendix 1, but in general we find that our model produces many of patterns generated by stepping stone models while avoiding some artifacts associated with discretization of the landscape.
 
@@
 \centering
-\includegraphics{figures/sampling_maps.pdf}
+\includegraphics{ms/figures/sampling_maps.pdf}
 \caption{Example sampling maps for 60 individuals on a 50$\times$50 landscape for midpoint, point, and random sampling strategies, respectively.}
@@
 %%%%%%%%%%%
-\subsection{Demographic Modeling}
+\subsection{Demographic Inference}
 
@@
 In all simulations, the phenotype of each individual
-is determined by adding independent Gaussian noise with mean zero and standard deviation 10 \revpoint{1}{5}
-to a mean that may depend on spatial position.
-We adjust the geographic pattern of mean phenotype to create spatially autocorrelated environmental influences on phenotype.
+is determined by adding independent Gaussian noise with mean zero \revpoint{1}{5}
+to a mean that may depend on spatial position such that the mean of spatially-correlated phenotypes varies by two standard deviations across the landscape.\llabel{phenotypescalingsd}
+We then adjust the geographic pattern of mean phenotype to create four spatially autocorrelated environmental influences on phenotype.
 In the first simulation of \emph{nonspatial} environments, the mean did not change,
@@
 %%%%%%%%%%%
-\subsection{Demographic Parameters}
+\subsection{Demographic Parameters and Run Times}
 
@@
 has a surprisingly large effect on demographic quantities that are usually fixed in Wright-Fisher models --
-the generation time, census population size, and variance in offspring number.
-These are shown in Figure~\ref{fig:genparams}.
-This occurs because,
-even through the parameters $K$ and $L$ that control population density and mean lifetime respectively \llabel{ll:K_more}
+the generation time, census population size, and variance in offspring number, shown in Figure~\ref{fig:genparams}.
+Because our simulation is paramaterized on an individual level, these population parameters emerge as a property of the interactions among individuals rather than being directly set \llabel{emergentparams}.
+Variation across runs occurs because,
+even though the parameters $K$ and $L$ that control population density and mean lifetime respectively \llabel{ll:K_more}
 were the same in all simulations,
-the strength of stochastic effects depends strongly on $\sigma$.
+the strength of stochastic effects depends strongly on the spatial interaction distance $\sigma$.
 For instance, 
@@
 
+\begin{figure}[htbp]
+\centering
+\includegraphics{figures/runtimes.pdf}
+\caption{Run times of continuous space simulations with landscape width 50 and expected density 5 under varying neighborhood size.
+Times are shown for simulations run with mutations applied directly in SLiM (dashed lines) or later applied to tree sequences with msprime (solid lines).
+Times for simulations run with tree sequence recording disabled are shown in grey.}
+\label{fig:runtimes}
+\end{figure}
+
 Generation time similarly shows complex behavior with respect to neighborhood sizes, and varies between 5.2 and 4.9 timesteps per generation across the parameter range explored. 
@@
 
+Run times for our model scale approximately linearly with neighborhood size (Figure \ref{fig:runtimes}), with the lowest neighborhood sizes reaching 30N generations in around a day and those with neighborhood size approaching 1,000 requiring up to two weeks of computation.
+At small neighborhood sizes the overhead caused by processing mutations and simplifying tree sequences can add $\approx30\%$ to the total run time, but above neighborhood sizes of $\approx$ 100 the effects of neighborhood size dominate because our simulation requires pairwise calculations for sets of individuals within a $3\sigma$ radius.
+As currently implemented running simulations at neighborhood sizes $>1000$ to coalescence is likely impractical, though running these models for more limited timescales is possible.
+\llabel{runtimesxn}
+
 %The fitness scaling procedure we use to avoid high fitness at range edges shrinks the suitable area by one $\sigma$ on all sides and likely explains part of the variation in census size, but scaling in offspring variance and the U-shaped distribution of census size in spatial models suggests additional processes at work.
@@
 However, the behavior of most common population genetic summary statistics other than Tajima's $D$ \citep{Stadler2009} has not yet been described in realistic geographic models.
-Moreover, as we will show, spatial sampling strategies can affect summaries of variation at least as strongly as the underlying population dynamics.
+Moreover, as we will show, spatial sampling strategies can affect summaries of genetic variation at least as strongly as the underlying population dynamics \llabel{spatialgenetic}.
 
@@
 \centering
-\includegraphics[width=\textwidth]{figures/sfs_w_sumstats.pdf}
-\caption{Site frequency spectrum (A) and summary statistic distributions (B) by sampling strategy and neighborhood size.}
+\includegraphics[width=\textwidth]{ms/figures/sfs_w_sumstats_loglog.pdf}
+\caption{Log-scaled site frequency spectrum (A) and summary statistic distributions (B) by sampling strategy and neighborhood size.}
 \label{fig:sumstats}
@@
 \centering
-\includegraphics[width=.5\textwidth]{figures/ibs_dists2.pdf}
-\caption{Cumulative distributions for IBS tract lengths per pair of individuals at different geographic distances, across three neighborhood sizes ($N_{W}$).}
+\includegraphics[width=.5\textwidth]{ms/figures/ibs_cumdists_w_rm.pdf}
+\caption{Cumulative distributions for IBS tract lengths per pair of individuals at different geographic distances, across three neighborhood sizes (NS).}
 \label{fig:ibs}
@@
 \includegraphics[width=\textwidth]{figures/distance_by_DAF.pdf}
-\caption{Trends in the distance among allele copies at varying derived allele frequencies and neighborhood sizes.}
+\caption{Trends in the distance among allele copies at varying derived allele frequencies and neighborhood sizes (NS).}
 \label{fig:dafdists}
@@
 Though this degree of variation in population size is certainly meaningful in an ecological context, it is relatively minor in population genetic terms.
+\llabel{ibdnesxn} In general methods directly assessing haplotype structure in phased data (for example, \citet{browning2015accurate}) are thought to provide increased resolution for recent demographic events, but in this case the error we observed was essentially an accurate reflection of underlying genealogies in which terminal branches are anomalously short.
+Combined with our analysis of IBS tract length variation (Figure \ref{fig:ibs}) this suggests that haplotype-based methods are likely to be affected by similar biases.
+
 A more worrying pattern was the high level of variance in inferred $N_{e}$ trajectories for individual model fits using these methods, which was highest in simulations with the smallest neighborhood size (Figure \ref{fig:demography}, Figure \ref{fig:demography_supplement}).
@@
 Spatially restricted mate choice and dispersal causes variation in allele frequencies across the range of a species.
-If environmental factors affecting the phenotype of interest also vary over space, then groups of individuals in different regions will allele frequencies and environmental exposures will covary over space.
+If environmental factors affecting the phenotype of interest also vary over space, then allele frequencies and environmental exposures will covary over space.
 In this scenario an uncorrected GWAS will infer genetic associations with a purely environmental phenotype at any site in the genome that is differentiated over space, and the relative degree of bias will be a function of the degree of covariation in allele frequencies and the environment (i.e., Figure \ref{fig:gwas}C, bottom panel).
@@
 In summary, spatial covariation in population structure and the environment confounds the interpretation of GWAS $p$-values, and correction using principal components is insufficient to fully separate these signals for polygenic traits under a variety of environmental and population parameter regimes.
-Other GWAS methods may be less sensitive to this confounding, but there is no obvious reason that this should be so.
+Other GWAS methods such as mixed models \citep{Kang2008} \llabel{mixedmodelcallout} may be less sensitive to this confounding, but there is no obvious reason that this should be so. 
 One approach to estimating the degree of bias in GWAS caused by population structure is LD score regression \citep{Bulik-Sullivan2015}. 
@@
 This is essentially the approach taken in our ``clinal'' model (though we add normally distributed noise to our phenotypes).
+This approach has recently been taken with polygenic scores for UK Biobank samples in \cite{haworth2019apparent}, finding that scores are correlated with birth location even in this relatively homogenous sample \llabel{haworthcite}.
 Of course, it is possible that genotypes indirectly affect individual locations by adjusting organismal fitness and thus habitat selection across spatially varying environments, but we believe that this hypothesis should be tested against a null of stratification bias inflation rather than accepted as true based on GWAS results.
@@
 In Table \ref{table:NStable} we gathered estimates of neighborhood size from a range of organisms to get an idea of how likely dispersal is to play an important role in patterns of variation.
-Though this sample is almost certainly biased towards small-neighborhood species (because few studies have quantified neighborhood size in species with very high dispersal or population density), we find that neighborhood sizes in the range we simulated are fairly common across a range of taxa.
+These values should be compared to our simulation results with some caution, as though we expect neighborhood size to have similar effects across species of varying global $N_e$ \citep{Wright1946}, in our study we evaluated only a relatively small population of $\approx 10,000$ \llabel{popsizecaveat}.
+In addition, these empirical examples are likely biased towards small-neighborhood species (because few studies have quantified neighborhood size in species with very high dispersal or population density).
+However, from the available data we find that neighborhood sizes in the range we simulated are fairly common across a range of taxa.
 At the extreme low end of empirical neighborhood size estimates we see some flowering plants, large mammals, and colonial insects like ants.
@@
 One facet of spatial variation that we did not address in this study is the confounding of dispersal and population density implicit in the definition of Wright's neighborhood size.
-Our simulations were run under constant densities, but \citet{Ringbauer2017}'s approach to demographic inference in space suggests that density and dispersal can in some cases be estimated separately from genetic data.
+Our simulations were run under constant densities, but \cite{guindon2016demographic} and \citet{Ringbauer2017} have shown that these parameters are identifiable under some continuous models.
+Similarly, though the scaling effects of dispersal we show in Figure \ref{fig:sumstats} should occur in populations of any total size, other aspects such as the number of segregating sites are also likely affected by the total landscape size (and so total census $N$)\llabel{landscapesizecaveat}.
 Much additional work remains to be done to better understand how these parameters interact to shape genetic variation in continuous space, which we leave to future studies.
@@
 \section{Appendix 1}
-
+\llabel{appendix1}
 %%%%%%%%%%%%%%%%%%%%%%%
@@
 \centering
-\includegraphics[width=\textwidth]{ms/figures/grid_ss_curves.pdf}
+\includegraphics[width=\textwidth]{ms/figures/grid_ss_curves_edit.pdf}
 \caption{Summary statistics for 2-dimensional coalescent stepping-stone models with fixed total $N_{e}$ and varying numbers of demes per side.
@@
 
-\begin{figure*}[h]
-\centering
-\includegraphics[]{ms/figures/grid_thetaW_tajD_zoom.pdf}
-\caption{Zoomed-in view of $\theta_{W}$ and Tajima's $D$ for stepping-stone models under varying neighborhood sizes.}
-\label{fig:grid_stats_zoom}
-\end{figure*}
-
 These differences relative to our continuous model mainly reflect two shortcomings of the reverse-time stepping stone model.
@@
 This results in an excess of short terminal branches in the coalescent tree, which decreases the total branch length and leads to fewer segregating sites, deflated $\theta_W$, and inflated Tajima's $D$.
-Overall then our continuous model reproduces important features of spatial structure approximated by reverse-time stepping-stone models at moderate neighborhood sizes while more accurately reflecting the dynamics of populations with small neighborhood sizes.
+Overall then our continuous model reproduces important features of spatial structure approximated by reverse-time stepping-stone models at moderate neighborhood sizes while avoiding some artifacts caused binning the landscape into discrete demes.
 
@@
 
+\beginsupplement
 \section{Supplementary Figures and Tables}
-\beginsupplement
+\FloatBarrier
+
+\begin{figure*}[p]
+\centering
+\includegraphics[width=\textwidth]{ms/figures/wm_v_dd_maps.pdf}
+\caption{Maps of individual locations in a continuous-space Wright-Malécot model with independent dispersal of all individuals (top) and under our continuous space model incorporating density-dependant fitness (bottom).
+The clustering seen in the top row is the "Pain in the Torus" described by \citet{Felsenstein1975}.}
+\label{fig:slim_v_msp_sfs} 
+\end{figure*}
+\FloatBarrier
+
+\begin{figure*}[p]
+\centering
+\includegraphics[]{figures/slim_v_msp_sfs.pdf}
+\caption{Site frequency spectra from a simulation with neighborhood size = 12.5 when mutations are recorded directly in SLiM (blue line) or applied later in msprime (red line).}
+\label{fig:slim_v_msp_sfs} 
+\end{figure*}
+\FloatBarrier
 
@@
 \end{figure*}
+\FloatBarrier
 
@@
 \end{figure*}
-
-\afterpage{\clearpage}
+\FloatBarrier
+
+\llabel{ll:figs3caption}
 \begin{figure*}[p]
@@
 \includegraphics[]{figures/het_z_by_ind.pdf}
-\caption{Normalized mean observed heterozygosity by location across 200 randomly sampled individuals}
+\caption{Variation in observed heterozygosity (i.e.
+proportion of heterozygous individuals) in hexagonal bins across the landscape, estimated from a random sample of 200 individuals from the final generation of a simulation with neighborhood size $\approx$25.
+Values were Z-normalized for plotting.}
 \label{fig:hetmap}
 \end{figure*}
+\FloatBarrier
 
@@
 \end{figure}
+\FloatBarrier
 
@@
 \end{table}
+\FloatBarrier
 
@@
 \end{table*}
+\FloatBarrier
 

### 2019-10-11 ms/review-responses.tex
andy edits

@@
 \reviewersection{1}
+We thank the reviewer for their very constructive comments. Responses follow below:
 
@@
 see figure 4 and the skew(IBs) panel of fig S1) such that the highest density is in low values.
-We explain the depressed mean value by noting that at low neighborhood sizes the coalescence time of distant individuals is inflated, which allows more time for recombination and mutation to break up IBS tracts.
+We explain the depressed mean value by noting that at low neighborhood sizes the coalescence time of distant (in space) individuals is inflated, which allows more time for recombination and mutation to break up IBS tracts.
 The flip side of this is that there are then more IBS tracts in total across a random sample of individuals, explaining the corresponding inflation in the number of tracts.
-Perhaps confusing we also see an increase in the number of very long IBS tracts at low neighborhood sizes, which corresponds to an excess of very recent coalescent events for groups of nearby individuals.
+We also see an increase in the number of very long IBS tracts at low neighborhood sizes, which corresponds to an excess of very recent coalescent events for groups of nearby individuals.
 \cjb{need to think of something to say we did about this in the text}
@@
 \reply{
-ah, a confusing legend title placement! we moved "model" to after "spatial" as suggested.
+Thanks for catching our confusing legend title placement! We have moved "model" to after "spatial" as suggested.
 }
@@
 \reply{
-This is an excellent point also raised by other reviewers.
-We have added an appendix section comparing our model to a reverse-time stepping stone analysis on several relevant summary statistics \llname{appendix1}.
-Briefly, we find that many of the qualitative trends we see also occur in a stepping-stone setting, suggesting that they are driven purely by the particularities of our parameterization.
-We also see several interesting departures from the stepping-stone model reflecting shortcomings in the gridded approach to approximating continuous landscapes, which we think our model helps to avoid.
+This is a good point also raised by other reviewers.
+We have added an appendix section comparing our model to a reverse-time stepping stone simulation on several relevant summary statistics \llname{appendix1}.
+We show that for $\theta_\pi$ and several other statistics the stepping stone model approaches the continuous model as the resolution of the landscape increases.
+We also see some interesting differences that seem to reflect artifacts from discretization.
 We are also curious how all the other analyses we test would be affected using other simulation schemes, but hope to explore that aspect in future work.
@@
     See above.
-It is certainly easier and faster, but comes with significant issues either when neighborhood size is lower than the population of a deme, or when demes are small enough that the sample size approaches local $N_e$ (at least in reverse-time simulations).
+It is certainly easier and faster, but comes with significant issues either when neighborhood size is lower than the population of a deme, or when demes are small enough that the sample size approaches local $N_e$, at least in reverse-time simulations.
 }
@@
 \reply{
-    We have added a figure showing sample site-frequency-spectra generated from a subset of simulations run with SLiM mutations, and then using msprime to apply mutations to the same tree sequences with our generation-time scaling approach.
-As you can see in \ref{fig:slim_v_msp_sfs}, these approaches yield extremely similar spectra so we believe any error in the approximation is quite small.
+    We have added a figure showing sample site-frequency-spectra generated from a subset of simulations run with SLiM mutations, and then using msprime to apply mutations to the same tree sequences with our generation-time scaling approach \ref{fig:slim_v_msp_sfs}.
+These approaches yield extremely similar spectra so we believe any error in the approximation is quite small.
 We also now show variation in simulation times with different mutation and tree sequence recording schemes in Figure \ref{fig:runtimes}, and included a brief discussion of this in \llname{runtimesxn}

### 2019-11-09 ms/spaceness.tex
removed redundant figure versions (replaced with symbolic links)

@@
 \centering
-\includegraphics{ms/figures/sampling_maps.pdf}
+\includegraphics{figures/sampling_maps.pdf}
 \caption{Example sampling maps for 60 individuals on a 50$\times$50 landscape for midpoint, point, and random sampling strategies, respectively.}
@@
 \centering
-\includegraphics[width=\textwidth]{ms/figures/sfs_w_sumstats_loglog.pdf}
+\includegraphics[width=\textwidth]{figures/sfs_w_sumstats_loglog.pdf}
 \caption{Log-scaled site frequency spectrum (A) and summary statistic distributions (B) by sampling strategy and neighborhood size.}
@@
 \centering
-\includegraphics[width=.5\textwidth]{ms/figures/ibs_cumdists_w_rm.pdf}
+\includegraphics[width=.5\textwidth]{figures/ibs_cumdists_w_rm.pdf}
 \caption{Cumulative distributions for IBS tract lengths per pair of individuals at different geographic distances, across three neighborhood sizes (NS).}
@@
 \centering
-\includegraphics[width=\textwidth]{ms/figures/grid_ss_curves_edit.pdf}
+\includegraphics[width=\textwidth]{figures/grid_ss_curves_edit.pdf}
 \caption{Summary statistics for 2-dimensional coalescent stepping-stone models with fixed total $N_{e}$ and varying numbers of demes per side.
@@
 \centering
-\includegraphics[width=\textwidth]{ms/figures/wm_v_dd_maps.pdf}
+\includegraphics[width=\textwidth]{figures/wm_v_dd_maps.pdf}
 \caption{Maps of individual locations in a continuous-space Wright-Malécot model with independent dispersal of all individuals (top) and under our continuous space model incorporating density-dependant fitness (bottom).

### 2019-11-09 ms/review-responses.tex
some cites

@@
 Done.
+\revref
 }
@@
 \begin{point}{\revref}
-    "Such models have been used extensively in ecological modeling but rarely in population genetics " Detailing these previous uses via citations and elaboration may help alleviate the major concern about the provenenance of this model and its unique behaviors (see general comments above and R2 and R3 comments).
+    ``Such models have been used extensively in ecological modeling but rarely in population genetics'': Detailing these previous uses via citations and elaboration may help alleviate the major concern about the provenenance of this model and its unique behaviors (see general comments above and R2 and R3 comments).
 \end{point}
 
-\cjb{Peter - did you have any in mind here? I can come up with a couple but don't know the ecology lit on this very well}
 \reply{
-
+    Good idea --
+    we have added some historical discussion and a few more citations to this section.
+    \revref
 }

### 2019-11-09 ms/spaceness.tex
some cites

@@
 This pattern of ``isolation by distance'' \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}.
-Despite a long history of analytical work describing the genetics of populations distributed across continuous geography (e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation,Wilkins2002linearcoalescent,Wilkins2004septimescales}), much modern work still describes geographic structure as a set of discrete populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations,al2019estimating}\llabel{eemscitation}.
+Despite a long history of analytical work describing the genetics of populations distributed across continuous geography
+(e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation,Wilkins2002linearcoalescent,Wilkins2004septimescales}), \revpoint{AE}{1}
+much modern work still describes geographic structure as a set of discrete populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations,al2019estimating}\llabel{eemscitation}.
 For this reason, most population genetics statistics are interpreted with reference to discrete, well-mixed populations, and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000} with methods that assume these are randomly mating units.
@@
 thus enforcing regular population density by edict.
-Among many other important results drawn from this class of ``lattice'' or ``stepping stone'' models, 
+Among many other results drawn from this class of ``lattice'' or ``stepping stone'' models
+\citep{epperson2003geographical}, 
 \citet{Rousset1997} showed that the slope of the linear regression of genetic differentiation ($F_{ST}$) against the logarithm of spatial distance is an estimate of neighborhood size. 
@@
 To avoid questionable assumptions, we here used forward-time, individual-based simulations.
-By scaling the probability of survival in each timestep to local population density we shift reproductive output towards low-density regions, which prevents populations from clustering.
-Such models have been used extensively in ecological modeling but rarely in population genetics, where to our knowledge previous implementations of continuous space models have focused on a small number of genetic loci, which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms.
-\revpoint{AE}{2}
+The question of what regulates real populations has a long history and many answers
+\citep{lloyd1967crowding,antonovics1980ecological},
+but it is clear that populations must have some density-dependent feedback on population size,
+or else they would face eventual extinction or explosion.
+In the absence of unrealistic global population regulation,
+this regulation must be local,
+and there are many ways to arrange this \citep{bolker2003spatial}.
+We decrease each individual's probability of survival with increasing local population density,
+which shifts reproductive output towards low-density regions 
+(and prevents the unrealistic population clustering seen by \citet{Felsenstein1975}). 
+Such models have been used extensively in ecological modeling
+\citep{durrett1994importance,bolker1997using,law2003population,fournier2004microscopic,champer2019suppression,}
+but rarely in population genetics, \revpoint{AE}{2}
+where to our knowledge implementations of continuous space models
+before their availability through SLiM \citep{Haller2019}
+have focused on a small number of genetic loci
+\citep{slatkin1989comparison,barton2002neutral,robledoarnuncio2010isolation},
+which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. 
 By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.
@@
 \bibliography{spaceness}
+\bibliography{more_refs}
 

### 2019-11-10 ms/review-responses.tex
a bit more

@@
 We have added a figure (Figure \ref{fig:runtimes}) and short discussion \llname{runtimesxn} of run times.
+\plr{what's up with those run times?}
 }
@@
 
-\begin{point}{Figure 4:}
+\begin{point}{Figure \ref{fig:ibs}:}
     Show random-mating expectation
@@
 \reply{
-    \cjb{working on this}
+    This is done.
 }
 
-\begin{point}{Figure 3A, S2:}
+\begin{point}{Figure \ref{fig:sumstats}A, \ref{fig:allsfs}:}
     Perhaps more revealing to show on log-log scale?
@@
 \reply{
-Good suggestion -- the SFS in Fig 3A is now on a log-log scale, which shows the slight decrease in low frequency SNPs a little better. 
+    Good suggestion -- the SFS in Fig \ref{fig:sumstats}A is now on a log-log scale,
+    which shows the slight decrease in low frequency SNPs a little better. 
+    We've left Fig \ref{fig:allsfs} as-is.
 }
@@
 \begin{point}{}
-    There are also several typos and errors in the text. For example, on Page 12, lines 309; Page 27, line 655.
+    There are also several typos and errors in the text. For example, \revref; \llname{ll:1:11:2}.
 \end{point}
@@
     population history, and on (iii) confounding in genome-wide association
-    studies (GWAS). I Liked the paper a lot. It's interesting, well-written and
+    studies (GWAS). I liked the paper a lot. It's interesting, well-written and
     addresses an important question - the effect of spatial population
@@
 \begin{point}{}
-    The source of equation (1) is not obvious to me.
+    The source of equation \eqref{eqn:p_survival} is not obvious to me.
 I sort of see how it makes sense, but a little but more intuition or a brief derivation or an illuminating either in the main text or the supplement, would be helpful.

### 2019-11-10 ms/spaceness.tex
a bit more

@@
 
-To avoid questionable assumptions, we here used forward-time, individual-based simulations.
+To avoid hard-to-evaluate approximations, we here used forward-time, individual-based simulations.
 The question of what regulates real populations has a long history and many answers
-\citep{lloyd1967crowding,antonovics1980ecological},
-but it is clear that populations must have some density-dependent feedback on population size,
+\citep[e.g.,][]{lloyd1967crowding,antonovics1980ecological,crawley1990population},
+but it is clear that populations must at some point have density-dependent feedback on population size,
 or else they would face eventual extinction or explosion.
@@
 and there are many ways to arrange this \citep{bolker2003spatial}.
-We decrease each individual's probability of survival with increasing local population density,
+In our simulations,
+we decrease each individual's probability of survival with increasing local population density,
 which shifts reproductive output towards low-density regions 
-(and prevents the unrealistic population clustering seen by \citet{Felsenstein1975}). 
+(and prevents the unrealistic population clustering seen by \citet{Felsenstein1975},
+see Supplemental \autoref{fig:slim_has_no_pain}).
 Such models have been used extensively in ecological modeling
-\citep{durrett1994importance,bolker1997using,law2003population,fournier2004microscopic,champer2019suppression,}
+\citep{durrett1994importance,bolker1997using,law2003population,fournier2004microscopic,champer2019suppression}
 but rarely in population genetics, \revpoint{AE}{2}
@@
 have focused on a small number of genetic loci
-\citep{slatkin1989comparison,barton2002neutral,robledoarnuncio2010isolation},
+\citep[e.g.,][]{slatkin1989comparison,barton2002neutral,robledoarnuncio2010isolation,rossine2014espao},
 which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. 
@@
 the probability of survival until the next time step for individual $i$ is
-\begin{equation}
+\begin{equation} \label{eqn:p_survival}
     p_i 
@@
 We chose this functional form so that the equilibrium population density per unit area is close to $K$, \llabel{ll:K_again}
-and the mean lifetime is around $L$.
+and the mean lifetime is around $L$;
+for more description see Appendix \ref{apx:demographic_model}.
 
@@
 Variation across runs occurs because,
-even though the parameters $K$ and $L$ that control population density and mean lifetime respectively \llabel{ll:K_more}
+even though the parameters $K$ and $L$ that control population density and mean lifetime respectively \llabel{ll:K_more} \revpoint{1}{11}
 were the same in all simulations,
@@
 If environmental factors affecting the phenotype of interest also vary over space, then allele frequencies and environmental exposures will covary over space.
+\llabel{ll:1:11:2}
 In this scenario an uncorrected GWAS will infer genetic associations with a purely environmental phenotype at any site in the genome that is differentiated over space, and the relative degree of bias will be a function of the degree of covariation in allele frequencies and the environment (i.e., Figure \ref{fig:gwas}C, bottom panel).
@@
 
-\bibliography{spaceness}
-\bibliography{more_refs}
+\bibliography{spaceness,more_refs}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Demographic model}
+\label{apx:demographic_model}
+
+We chose our demographic model so that every individual has on average $1/L$ offspring
+each time step,
+and if the local population density of an individual is $n$,
+then their probability of survival until the next time step is
+(equation \eqref{eqn:p_survival}):
+\begin{equation}
+    p = \min\left( 0.95, \frac{1}{1 + n / (K (1+L))} \right) .
+\end{equation}
+We capped survival at 0.95 so that we would not have exceptionally long-lived individuals
+in sparsely populated areas --
+otherwise, an isolated individual might live for a very long time.
+Since $1 - p \approx n / (K (1 + L))$,
+mortalitity goes up roughly linearly with the number of neighbors (on a scale given by $K$),
+as would be obtained if, for instance, mortality is due to agonistic interactions.
+Ignoring in/outmigration,
+a region is at demographic equilibrium if the per-capita probability of death
+is equal to the birth rate,
+i.e., if $1 - p = 1/L$.
+(Note that there is no effect of age in the model,
+which would make the analysis more complicated.)
+Solving this for $n$,
+we get that in a well-mixed population, the equilibrium density should be around
+\begin{equation}
+    n &= K \frac{L+1}{L-1}
+\end{equation}
+individuals per unit area.
+At this density, the per-capita death rate is $1/L$, so the mean lifetime is $L$.
+This equilibrium density is \emph{not} $K$, but (since $L=4$) is two-thirds larger.
+However, in practice this model leads to a total population size which is around $K$
+multiplied by total geographic area
+(but which depends on $\sigma$, as discussed above).
+The main reason for this is that
+since offspring tend to be near their parents, individuals tend to be ``clumped'',
+and so experience a higher average density than the ``density'' one would compute
+by dividing census size by geographic area \citep{lloyd1967crowding}.
+To maintain a constant expected total population size
+would require making (say) $K$ depend on $\sigma$;
+however, typical local population densities might then be more dissimilar.
+
 \stopappendix
@@
 The clustering seen in the top row is the "Pain in the Torus" described by \citet{Felsenstein1975}.}
-\label{fig:slim_v_msp_sfs} 
+\label{fig:slim_has_no_pain} 
 \end{figure*}
@@
 \caption{Site frequency spectra from a simulation with neighborhood size = 12.5 when mutations are recorded directly in SLiM (blue line) or applied later in msprime (red line).}
-\label{fig:slim_v_msp_sfs} 
+\label{fig:slim_v_msp_sfs2} 
 \end{figure*}

### 2019-11-18 ms/review-responses.tex

@@
 We have added a figure (Figure \ref{fig:runtimes}) and short discussion \llname{runtimesxn} of run times.
-\plr{what's up with those run times?}
+\plr{What's up with those run times?}
 }
@@
 Use of K to denote carrying capacity and density is a bit confusing.
-For example, on \llname{ll:K_more} it is said that ' the "population density" (K) and "mean lifetime" (L) parameters were the same in all simulations'.
+For example, on \llname{ll:K_more} it is said that ' the ``population density'' (K) and ``mean lifetime'' (L) parameters were the same in all simulations'.
 Here K seems to indicate carrying capacity rather than density?
@@
 
-\begin{point}{}
-    Page 9, line 242, Perhaps 'Demographic Inference' might better reflect the content of this section.
+\begin{point}{\revref}
+    Perhaps 'Demographic Inference' might better reflect the content of this section.
 \end{point}
@@
 \reply{
-Good suggestion -- we have changed the section heading to "demographic inference" 
+Good suggestion -- we have changed the section heading to ``demographic inference'' \revref.
 }
@@
 \begin{point}{}
-    Page 11, Please correct the legend in Figure 2: must be 'spatial model' and 'random mating' model.
+    Please correct the legend in \autoref{fig:genparams}:
+    must be `spatial model' and `random mating' model.
 \end{point}
@@
 \reply{
-Thanks for catching our confusing legend title placement! We have moved "model" to after "spatial" as suggested.
+Thanks for catching our confusing legend title placement!
+We have moved ``model'' to after ``spatial'' as suggested.
 }
@@
 \begin{point}{}
-    Optional: a dashed line in Figure 2 that shows the total carrying capacity of 50*50*5=12500 would be helpful.
+    Optional: a dashed line in \autoref{fig:genparams}
+    that shows the total carrying capacity of $50 \times 50 \times 5=12500$ would be helpful.
 \end{point}
@@
 
-\begin{point}{}
-    Page 13, line 349, The phrase 'affect summaries of variation' is better to replace with 'summaries of genetic variation'.
+\begin{point}{\revref}
+    The phrase `affect summaries of variation' is better to replace with `summaries of genetic variation'.
 \end{point}
@@
 \reply{
-    done. \llname{spatialgenetic}
+    Done. \revref
 }
@@
 Can they provide a figure showing that this is the case.
-What does "roughly" mean on line 164.
+What does ``roughly'' mean on line 164.
 Perhaps a heatmap of the fitness of individuals across the grid with and without the scaling factor?
@@
 \begin{quote}
-    The present study deals with a "hot topic" in spatial population genetics. Most
+    The present study deals with a ``hot topic'' in spatial population genetics. Most
     inferential and descriptive methods in statistical spatial population
@@
     examples of spectacular failures like the Wright-Malecot model (see
-    Felsenstein's "pain in the torus' article, 1975) or, more recently, the
-    ``mugration' or ``discrete trait analysis' model in phylodynamics (see De Maio
+    Felsenstein's ``pain in the torus'' article, 1975) or, more recently, the
+    ``mugration'' or ``discrete trait analysis'' model in phylodynamics (see De Maio
     et al. 2015) have probably mostly harmed our research field -- that one cannot
@@
 \reply{
-    Census population in the random mating model scales with "neighborhood size" because of two effects -- the edge scaling and the competition
+    Census population in the random mating model scales with ``neighborhood size'' because of two effects -- the edge scaling and the competition
 
@@
 \begin{point}{}
-    Line 729-731: "Many more species occur in a middle range of neighborhood sizes between 100 and 1000 - a range in which spatial processes play a minor role in our analyses [...]".
+    Line 729-731: ``Many more species occur in a middle range of neighborhood sizes between 100 and 1000 - a range in which spatial processes play a minor role in our analyses [...]''.
 Do the authors think that the spatial processes would still play a minor role when neighborhood sizes exceed 100-1000 if the habitat was larger than that taken in the present simulations?
@@
 This is a good point.
-Wright's work \citep{Wright1943} suggests some aspects of genetic variation such as variance in allele frequencies and inbreeding coefficients can be estimated by looking only at what he would later \citep{Wright1946} call "neighborhood size", but certainly other aspects like the number of segregating sits will also depend on total landscape size.
+Wright's work \citep{Wright1943} suggests some aspects of genetic variation such as variance in allele frequencies and inbreeding coefficients can be estimated by looking only at what he would later \citep{Wright1946} call ``neighborhood size'', but certainly other aspects like the number of segregating sits will also depend on total landscape size.
 We now note on \llname{popsizecaveat} that we have evaluated only one landscape size, and have added a sentence to the discussion noting that exploration of these patterns in varying landscape sizes is an important avenue for further research \llname{landscapesizecaveat}.
@@
 \begin{point}{}
-    Notation: "Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$ in both the x and y coordinates.
-Each offspring is produced with a mate selected randomly from those within distance $3\sigma$, with probability of choosing a neighbor at distance $x$ proportional to $\exp(-x^2/2\sigma^2)$." I think x may be overloaded here, or I'm confused.
+    Notation: ``Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$ in both the x and y coordinates.
+Each offspring is produced with a mate selected randomly from those within distance $3\sigma$, with probability of choosing a neighbor at distance $x$ proportional to $\exp(-x^2/2\sigma^2)$.'' I think x may be overloaded here, or I'm confused.
 Clarify?
@@
 \begin{point}{}
-    When introducing the "spatial model" as opposed to this "random model," the more concrete illustration in Figure 1 hasn't yet been referenced, which makes it harder to follow.
+    When introducing the ``spatial model'' as opposed to this ``random model,'' the more concrete illustration in Figure 1 hasn't yet been referenced, which makes it harder to follow.
 It would be helpful to introduce this figure with the model.
@@
 \begin{point}{}
-    Not sure I follow this example: "Concretely, an individual at position (x, y) in a 50 × 50 landscape has mean phenotype 100 + 2x/5."
+    Not sure I follow this example: ``Concretely, an individual at position (x, y) in a 50 × 50 landscape has mean phenotype 100 + 2x/5.''
 \end{point}
@@
 
-\begin{point}{}
-    Minor typo (through vs though): "This occurs because, even through the "population density" (K) and "mean lifetime" (L) parameters..."
+\begin{point}{\revreffull{1}{11}}
+    Minor typo (through vs though): \emph{This occurs because, even through the ``population density'' (K) and ``mean lifetime'' (L) parameters...}
 \end{point}
@@
 \begin{point}{}
-    Define NS abbreviation in Figure 5.
+    Define NS abbreviation in Figure \ref{fig:dafdists}.
 \end{point}

### 2019-11-18 ms/spaceness.tex

@@
 \subsection{Demographic Inference}
+\revpoint{1}{4}
 
@@
 However, the behavior of most common population genetic summary statistics other than Tajima's $D$ \citep{Stadler2009} has not yet been described in realistic geographic models.
-Moreover, as we will show, spatial sampling strategies can affect summaries of genetic variation at least as strongly as the underlying population dynamics \llabel{spatialgenetic}.
+Moreover, as we will show, spatial sampling strategies can affect summaries of genetic variation at least as strongly as the underlying population dynamics.
+\revpoint{1}{9}
 
@@
 \begin{equation}
-    n &= K \frac{L+1}{L-1}
+    n = K \frac{L+1}{L-1}
 \end{equation}

### 2019-11-18 ms/review-responses.tex
pass through PR edits, cleaned up response comments, added edge scaling fig

@@
 We have added a figure (Figure \ref{fig:runtimes}) and short discussion \llname{runtimesxn} of run times.
-\plr{What's up with those run times?}
 }
@@
 We also see an increase in the number of very long IBS tracts at low neighborhood sizes, which corresponds to an excess of very recent coalescent events for groups of nearby individuals.
-\cjb{need to think of something to say we did about this in the text}
 
@@
 \reply{
-    \cjb{peter - any suggestions here?}
+    Additional discussion of the model is now provided in the appendix (see "Demographic model") \llname{dem_model_app}
 }
@@
 Good suggestion -- we have added a supplemental figure to show the distribution of individual fitness across the landscape with and without our edge-scaling approach.
-MAKE THIS FIG AND POINT TO IT HERE.
+(Note though that we found the effect subtle and hard to visualize well).
 }
@@
 \reply{
-
+    We have added a brief discussion of the causes of the census size scaling to the relevant results section \llname{censusscaling}.
 }
@@
 It seems to reflect overcorrection in the regression when using PC coordinates as covariates -- the PCA is capturing some information about the spatial genetic variation which itself covaries only weakly with the phenotype, and as a result we see anomalously low $-log10(p)$ when regressing genotype against phenotype.
-\cjb{Peter or Andy -- any advice here?
-}
 }
@@
 \reply{
-    Similarly here we monitored the population size over time while running SLiM.
-All simulations maintained asymptotic populations, and none of the runs we started ended because populations crashed.
-\cjb{is there something else we should show here?}
+    Similarly here we monitored the population size over time while running SLiM. All simulations maintained asymptotic populations, and none of the runs we started ended because populations crashed.
 }
@@
 \reply{
-    \cjb{Peter -- ideas here?}
+   We have added additional discussion of this point to appendix A (see "Demographic Model") \llname{dem_model_app}
 }
@@
 \reply{
-    Census population in the random mating model scales with ``neighborhood size'' because of two effects -- the edge scaling and the competition
-
-    \cjb{need a better answer here...}
+    We have a added a brief discussion of the census size scaling to \llname{censusscaling}
 }
@@
 \reply{
-    \cjb{working on this}
     Thank you for this suggestion.
@@
 \reply{
-    \cjb{working on this}
+   We have expanded our citations of some of the relevant ecology literature \llname{ecolcites}, which we hope helps to contextualize the study better.
 }
@@
 \reply{
-    Good point on EEMS. We have added the most recent EEMS paper to the citations on \llname{eemscitation}
+    Good point on EEMS. We have added the most recent EEMS paper to the citations on \llname{eemscitation}, and have added an Appendix giving a more thorough comparison with stepping-stone models. 
 }
@@
 \reply{
-\cjb{is this an ok response?
-two reviewers asked about mixed models but I feel like including them would require expanding the gwas section quite a bit and I'd rather now.
-But if either of you feel like it's critical I could probably get it done next week}
     Good question -- we are also interested to know how mixed models perform here, but think that to properly test that we would want to change our design to generate phenotypes from simulated genotypes.

### 2019-11-18 ms/spaceness.tex
pass through PR edits, cleaned up response comments, added edge scaling fig

@@
 The question of what regulates real populations has a long history and many answers
-\citep[e.g.,][]{lloyd1967crowding,antonovics1980ecological,crawley1990population},
+\citep[e.g.,][]{lloyd1967crowding,antonovics1980ecological,crawley1990population}, \llabel{ecolcites}
 but it is clear that populations must at some point have density-dependent feedback on population size,
@@
 where $x_i$ and $y_i$ are the spatial coordinates of individual $i$, and $W$ is the width (and height) of the square habitat.
-This buffer roughly counteracts the increase in fitness individuals close to the edge would otherwise have.
+This buffer roughly counteracts the increase in fitness individuals close to the edge would otherwise have, though the effect is relatively subtle \ref{fig:edge_scaling}.
 
@@
 Census sizes range from $\approx 14,000$ at low $\sigma$ in the random mating model to $\approx 10,000$ for both models when neighborhood sizes approach 1,000.
+The scaling of census sizes in both random-mating and spatial models appears to be related to two consequences of the spatial competition function: the decline of fitness at range edges, which effectively reduces the habitable area by one $\sigma$ around the edge of the map and so results in a smaller habitable area at high $\sigma$ values; and variation in the equilibrium population density given varying competition radii.
+For spatial models we see an additional effect in which census size initially rises as neighborhood sizes scale from 2 - 10.
+This may reflect an Allee effect \citep{Allee1949} in which some individuals are unable to find mates when the mate selection radius is very small.
+\llabel{censusscaling}
 
@@
 
-Run times for our model scale approximately linearly with neighborhood size (Figure \ref{fig:runtimes}), with the lowest neighborhood sizes reaching 30N generations in around a day and those with neighborhood size approaching 1,000 requiring up to two weeks of computation.
-At small neighborhood sizes the overhead caused by processing mutations and simplifying tree sequences can add $\approx30\%$ to the total run time, but above neighborhood sizes of $\approx$ 100 the effects of neighborhood size dominate because our simulation requires pairwise calculations for sets of individuals within a $3\sigma$ radius.
-As currently implemented running simulations at neighborhood sizes $>1000$ to coalescence is likely impractical, though running these models for more limited timescales is possible.
+Run times for our model scale approximately linearly with neighborhood size (Figure \ref{fig:runtimes}), with the lowest neighborhood sizes reaching 30N generations in around a day and those with neighborhood size approaching 1,000 requiring up to three weeks of computation.
+As currently implemented running simulations at neighborhood sizes $>1000$ to coalescence is likely impractical, though running these models for more limited timescales and then "completing" the simulation by starting a reverse-time simulation from the resulting tree sequence in msprime is possible.
 \llabel{runtimesxn}
-
-%The fitness scaling procedure we use to avoid high fitness at range edges shrinks the suitable area by one $\sigma$ on all sides and likely explains part of the variation in census size, but scaling in offspring variance and the U-shaped distribution of census size in spatial models suggests additional processes at work.
-For spatial models census size initially rises as neighborhood sizes scale from 2 - 10, which seems to reflect an Allee effect \citep{Allee1949} in which some individuals are unable to find mates when the mate selection radius is very small.
 
@@
 \label{apx:demographic_model}
+\llabel{dem_model_app}
 
@@
 \label{fig:slim_has_no_pain} 
+\end{figure*}
+\FloatBarrier
+
+\begin{figure*}[p]
+\centering
+\includegraphics[width=0.6\textwidth]{figures/edge_scaling_comp.pdf}
+\caption{Comparison of individual fitness across the landscape in simulations with (right) and without (left) a decline in fitness approaching range edges.
+Note the slight excess of high-fitness individuals at edges on the left, which is (partially) counteracted by the scaling procedure.}
+\label{fig:edge_scaling} 
 \end{figure*}

### 2019-11-18 ms/review-responses.tex
more disco on demography and moar edits

@@
 
-\plr{I would generally cut down the quoted bits like the above to only what's essential,
-     but haven't done that yet.}
-
- \plr{IMPORTANT: don't reorder or delete "points" below - it messes up the automatic numbering!}
+This is a good point. There are many ways to go in the quest for demographic realism,
+and the fundamental question -- what regulates population density in real populations --
+is still very much an open question (or rather, a question with a great many answers).
+As highlighted below,
+we have included further discussion of our particular choice
+of local, density-mediated control on mortality,
+along with pointers into the literature, at several places in the manuscript;
+see in particular in the Methods when we introduce the model \revreffull{AE}{2},
+and \llname{demog_disco}.
 
@@
     Good idea --
-    we have added some historical discussion and a few more citations to this section.
-    \revref
+    we have added some historical discussion and a few more citations to this section,
+    \revref. Also see the additional discussion in the Discussion \llname{demog_disco}.
 }
@@
 \begin{point}{\revref}
-    I read the acknowledgement to the Hearth and Creative Sky Brewing with a sense of familiarity in feeling of gratitude to my own favorite cafes and breweries, but I it's not a great precedent for Acknowledgements to be filled this way.
+    I read the acknowledgement to the Hearth and Falling Sky Brewing with a sense of familiarity in feeling of gratitude to my own favorite cafes and breweries, but I it's not a great precedent for Acknowledgements to be filled this way.
 Please cut.
@@
 \reply{
-    Thanks for catching this - we have revised this caption to add details including the simulation parameters. \llname{ll:figs3caption}
+    Thanks for catching this - we have revised this caption to add details including the simulation parameters. (Supplemental Figure \ref{fig:hetmap}).
 }
@@
 \reviewersection{1}
-We thank the reviewer for their very constructive comments. Responses follow below:
 
@@
     this study identified spatial structuring scenarios (small neighbourhood
-    size ~ 10-100) that coupled with imperfect sampling strategies lead to a
+    size $\sim$ 10--100) that coupled with imperfect sampling strategies lead to a
     biased inference of widely used population genetic statistics (altogether
@@
 
+We thank the reviewer for their very constructive comments!
+Responses follow below:
+
 \begin{point}{}
-    When neighbourhood size is small (10-100), the mean number of IBS tracts > 2bp (nIBS as in Table S1) is elevated similar to Wright's inbreeding coefficient, but mean of the distribution of pairwise IBS (mean(IBS)) is decreased.
+    When neighbourhood size is small (10-100), the mean number of IBS tracts ${}>2$bp (nIBS as in Table S1) is elevated similar to Wright's inbreeding coefficient, but mean of the distribution of pairwise IBS (mean(IBS)) is decreased.
 What could be the source of this discrepancy?
@@
 \reply{
-The mean of the IBS tract distribution is calculated by building a (very large) vector of lengths of all IBS tracts between all pairs of individuals and taking the mean over this entire vector.
-This distribution is quite strongly skewed (i.e.
-see figure 4 and the skew(IBs) panel of fig S1) such that the highest density is in low values.
-We explain the depressed mean value by noting that at low neighborhood sizes the coalescence time of distant (in space) individuals is inflated, which allows more time for recombination and mutation to break up IBS tracts.
-The flip side of this is that there are then more IBS tracts in total across a random sample of individuals, explaining the corresponding inflation in the number of tracts.
-We also see an increase in the number of very long IBS tracts at low neighborhood sizes, which corresponds to an excess of very recent coalescent events for groups of nearby individuals.
-
+    The mean IBS is simply the mean of the vector of lengths of IBS tracts.
+    \plr{make sure this is stated in the text somewhere}
+    The short answer is that if one splits a chromosome up into more pieces (increasing nIBS),
+    the mean length of those pieces must necessarily be smaller.
+    At low neighborhood sizes, spatially distant individuals tend to have longer coalescence times,
+    leading to more (and hence smaller) IBS tracts.
+    This can be seen in Supplemental Figure \ref{fig:allsumstats}.
+    Note that skew(IBS) and number of long ($>$1e6) IBS tracts reflect other aspects of the distribution.
 }
@@
 \reply{
-\cjb{check on final pass}
+	Thanks for catching this, should be fixed now. 
 }
@@
 
+Thanks very much for the encouraging words!
+
 \begin{point}{}
@@
 Do the authors have some thoughts/intuition/results about that?
-Given that the results in Fig.
-3 seem quite consistent with expectations, I suspect that on some level it doesn't make much difference but then there are intermediate results like Fig.
-2 which seem a bit counter-intuitive and I wonder if those aspects depend on the simulation scheme.
+Given that the results in \autoref{fig:sumstats} seem quite consistent with expectations, I suspect that on some level it doesn't make much difference but then there are intermediate results like \autoref{fig:genparams} which seem a bit counter-intuitive and I wonder if those aspects depend on the simulation scheme.
 \end{point}
@@
     This is a good point also raised by other reviewers.
-We have added an appendix section comparing our model to a reverse-time stepping stone simulation on several relevant summary statistics \llname{appendix1}.
-We show that for $\theta_\pi$ and several other statistics the stepping stone model approaches the continuous model as the resolution of the landscape increases.
-We also see some interesting differences that seem to reflect artifacts from discretization.
-We are also curious how all the other analyses we test would be affected using other simulation schemes, but hope to explore that aspect in future work.
+    We have added some more discussion of the choice of demographic model
+    (\revreffull{AE}{2} and \llname{demog_disco}).
 }
@@
 \reply{
-    See above.
-It is certainly easier and faster, but comes with significant issues either when neighborhood size is lower than the population of a deme, or when demes are small enough that the sample size approaches local $N_e$, at least in reverse-time simulations.
+    We have added an appendix comparing our model to a reverse-time stepping stone simulation on several relevant summary statistics \llname{apx:stepping_stone}.
+    For some of these statistics (e.g., $\theta_\pi$),
+    the behavior of the stepping stone model becomes more similar (although not identical)
+    to the continuous model as the resolution of the landscape increases,
+    but other statistics seems to be more strongly affected by discretization artifacts.
+    We are also curious how all the other analyses we test would be affected using other simulation schemes, but hope to explore that aspect in future work. 
+    Furthermore, discrete-deme coalescent simulations are certainly faster,
+    but come with significant issues either when neighborhood size is lower than the population of a deme,
+    or when demes are small enough that the sample size approaches local $N_e$.
 }
@@
 \reply{
-    Additional discussion of the model is now provided in the appendix (see "Demographic model") \llname{dem_model_app}
+    Additional discussion of the model is now provided in the appendix (see "Demographic model") \llname{apx:demographic_model}
 }
@@
 \reply{
-    We have added a figure showing sample site-frequency-spectra generated from a subset of simulations run with SLiM mutations, and then using msprime to apply mutations to the same tree sequences with our generation-time scaling approach \ref{fig:slim_v_msp_sfs}.
+    We have added a figure showing sample site-frequency-spectra generated from a subset of simulations run with SLiM mutations, and then using msprime to apply mutations to the same tree sequences with our generation-time scaling approach (Figure \ref{fig:slim_v_msp_sfs}).
 These approaches yield extremely similar spectra so we believe any error in the approximation is quite small.
-We also now show variation in simulation times with different mutation and tree sequence recording schemes in Figure \ref{fig:runtimes}, and included a brief discussion of this in \llname{runtimesxn}
 }
@@
 \reply{
-   We have added additional discussion of this point to appendix A (see "Demographic Model") \llname{dem_model_app}
+   We have added additional discussion of this point to appendix A (see "Demographic Model") \llname{apx:demographic_model}
 }
@@
 The authors could provide some guarantee of the good behavior of their model as evidenced from simulations using a broad range of parameter values for generating data.
-Alternatively, they could elect to use the spatial-Lambda-Fleming-Viot model for their simulations, which, in my opinion would seem the most sensible option given that (1) it is possible to run backward-in-time simulations under this model, thereby saving a lot of computation time and (2) it is a well-studied model with good mathematical and biological properties and (3) it is implemented in a publicly available software program (https://github.com/jeromekelleher/discsim)
+Alternatively, they could elect to use the spatial-Lambda-Fleming-Viot model for their simulations, which, in my opinion would seem the most sensible option given that (1) it is possible to run backward-in-time simulations under this model, thereby saving a lot of computation time and (2) it is a well-studied model with good mathematical and biological properties and (3) it is implemented in a publicly available software program (\url{https://github.com/jeromekelleher/discsim}).
 \end{point}
@@
     This is a good point also raised by other reviewers.
-We have added an Appendix comparing our model to a stepping-stone, rather than the spatial-Labmda-Fleming-Viot, because we think these are the most familiar and widely used class of spatial models.
-See Appendix 1.
+We have added an Appendix comparing our model to a stepping-stone \llname{apx:stepping_stone}, rather than the spatial-Labmda-Fleming-Viot, because we think these are the most familiar and widely used class of spatial models.
 We find that many features of our model are well approximated by stepping-stone models, and that for statistics like $\theta_{\pi}$ the stepping stone model results approach our continuous space model as the number of demes used to describe the landscape increases.
@@
 \reply{
-\cjb{Peter and Andy -- ideas here?}
+	We switched the second instance (referring to the distance among individuals) to $d$. 
 }
@@
 \reply{
+	We have clarified by switching to "the phenotype $p$ for an individual at location (x,y) is then $p=100+2x/5$
 }

### 2019-11-19 ms/review-responses.tex
last(?) pass for reviewer 4

--- 2019-11-18 ms/review-responses.tex
@@
 
-\plr{I would generally cut down the quoted bits like the above to only what's essential,
-     but haven't done that yet.}
+%\plr{I would generally cut down the quoted bits like the above to only what's essential, but haven't done that yet.}
 
- \plr{IMPORTANT: don't reorder or delete "points" below - it messes up the automatic numbering!}
+% \plr{IMPORTANT: don't reorder or delete "points" below - it messes up the automatic numbering!}
 
@@
 \reply{
-    Thanks for catching this - we have revised this caption to add details including the simulation parameters. \llname{ll:figs3caption}
+    Thanks for catching this - we have revised this caption to add details including the simulation parameters.
 }

### 2019-11-19 ms/spaceness.tex
last(?) pass for reviewer 4

@@
 Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
-with probability of choosing a neighbor at distance $x$ proportional to $\exp(-x^2 / 2 \sigma^2)$.
+with probability of choosing a neighbor at distance $d$ proportional to $\exp(-d^2 / 2 \sigma^2)$.
 
@@
 we increased the mean phenotype from 100 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
-Concretely, an individual at position $(x,y)$ in a $50 \times 50$ landscape has mean phenotype $100 + 2x / 5$.
+Concretely, the phenotype $p$ for an individual at position $(x,y)$ is $p=100 + 2x / 5$.
 Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype 
@@
 
-\afterpage{\clearpage}
 \begin{figure*}[p]
@@
 
-\llabel{ll:figs3caption}
 \begin{figure*}[p]

### 2019-11-19 ms/spaceness.tex
correcting supp fig references

@@
 where $x_i$ and $y_i$ are the spatial coordinates of individual $i$, and $W$ is the width (and height) of the square habitat.
-This buffer roughly counteracts the increase in fitness individuals close to the edge would otherwise have, though the effect is relatively subtle \ref{fig:edge_scaling}.
+This buffer roughly counteracts the increase in fitness individuals close to the edge would otherwise have, though the effect is relatively subtle (Figure \ref{fig:edge_scaling}).
 
@@
 
-Figure \ref{fig:sumstats} shows the effect of varying neighborhood size and sampling strategy on the site frequency spectrum (Figure~\ref{fig:sumstats}A) and several standard population genetic summary statistics (Figure~\ref{fig:sumstats}B).
+Figure \ref{fig:sumstats} shows the effect of varying neighborhood size and sampling strategy on the site frequency spectrum (Figure~\ref{fig:sumstats}, Figure \ref{fig:allsfs}) and several standard population genetic summary statistics (Figure~\ref{fig:sumstats}B; additional statistics are shown in Figure \ref{fig:allsumstats}).
 Consistent with findings in island and stepping stone simulations \citep{Stadler2009}, the SFS shows a significant enrichment of intermediate frequency variants in comparison to the nonspatial expectation.
@@
 \caption{Site frequency spectra from a simulation with neighborhood size = 12.5 when mutations are recorded directly in SLiM (blue line) or applied later in msprime (red line).}
-\label{fig:slim_v_msp_sfs2} 
+\label{fig:slim_v_msp_sfs} 
 \end{figure*}

### 2019-11-20 ms/review-responses.tex
switched to Peter's workding on response 1.1

@@
 \reply{
-The mean of the IBS tract distribution is calculated by building a (very large) vector of lengths of all IBS tracts between all pairs of individuals and taking the mean over this entire vector.
-This distribution is quite strongly skewed (i.e.
-see figure 4 and the skew(IBs) panel of fig S1) such that the highest density is in low values.
-We explain the depressed mean value by noting that at low neighborhood sizes the coalescence time of distant (in space) individuals is inflated, which allows more time for recombination and mutation to break up IBS tracts.
-The flip side of this is that there are then more IBS tracts in total across a random sample of individuals, explaining the corresponding inflation in the number of tracts.
-We also see an increase in the number of very long IBS tracts at low neighborhood sizes, which corresponds to an excess of very recent coalescent events for groups of nearby individuals.
+The mean IBS is simply the mean of the vector of lengths of IBS tracts. 
+The short answer is that if one splits a chromosome up into more pieces (increasing nIBS),
+    the mean length of those pieces must necessarily be smaller.
+    At low neighborhood sizes, spatially distant individuals tend to have longer coalescence times,
+    leading to more (and hence smaller) IBS tracts.
+    This can be seen in Supplemental Figure \ref{fig:allsumstats}.
+    Note that skew(IBS) and number of long ($>$1e6) IBS tracts reflect other aspects of the distribution.
 

### 2019-11-20 ms/spaceness.tex

@@
 %\documentclass{article}
-\newcommand{\beginsupplement}{%
-        \setcounter{table}{0}
-        \renewcommand{\thetable}{S\arabic{table}}%
-        \setcounter{figure}{0}
-        \renewcommand{\thefigure}{S\arabic{figure}}%
-     }
-\newcommand{\stopsupplement}{%
-        \setcounter{table}{0}
-        \renewcommand{\thetable}{\arabic{table}}%
-        \setcounter{figure}{0}
-        \renewcommand{\thefigure}{\arabic{figure}}%
-     }
-\newcommand{\beginappendix}{%
-        \setcounter{table}{0}
-        \renewcommand{\thetable}{A\arabic{table}}%
-        \setcounter{figure}{0}
-        \renewcommand{\thefigure}{A\arabic{figure}}%
-     }
-\newcommand{\stopappendix}{%
-        \setcounter{table}{0}
-        \renewcommand{\thetable}{\arabic{table}}%
-        \setcounter{figure}{0}
-        \renewcommand{\thefigure}{\arabic{figure}}%
-     }
-
 \documentclass[10pt,twoside,lineno,hidelinks]{preprint}
@@
 In Table \ref{table:NStable} we gathered estimates of neighborhood size from a range of organisms to get an idea of how likely dispersal is to play an important role in patterns of variation.
-These values should be compared to our simulation results with some caution, as though we expect neighborhood size to have similar effects across species of varying global $N_e$ \citep{Wright1946}, in our study we evaluated only a relatively small population of $\approx 10,000$ \llabel{popsizecaveat}.
+These values should be compared to our simulation results with some caution, as though we expect neighborhood size to have similar effects across species of varying global $N_e$ \citep{Wright1946}, in our study we evaluated only a relatively small population of $\approx 10,000$.
+\llabel{popsizecaveat}
 In addition, these empirical examples are likely biased towards small-neighborhood species (because few studies have quantified neighborhood size in species with very high dispersal or population density).
@@
 Many more species occur in a middle range of neighborhood sizes between 100 and 1000 -- a range in which spatial processes play a minor role in our analyses under random spatial sampling but are important when sampling of individuals in space is clustered.
-Surprisingly, even some flying insects with huge census population sizes fall in this group, including fruit flies (\textit{D.
-melanogaster}) and mosquitoes (\textit{A.
+Surprisingly, even some flying insects with huge census population sizes fall in this group, including fruit flies (\textit{D.~melanogaster}) and mosquitoes (\textit{A.
 aegypti}).
@@
 While that is so we caution that association studies in which the effects of population structure are confounded with spatial variation in the environment are still sensitive to dispersal even at these large neighborhood sizes.
+
+%%%%%%%%%%%
+\subsection{Other demographic models}
+
+Any simulation of a population of reproducing organisms for a reasonably long amount of time
+requires some kind of control on population sizes,
+or else the population will either die out or grow very large. \llabel{demog_disco}
+The usual choice of population regulation for population genetics --
+a constant size, as in the Wright--Fisher model --
+implies biologically unrealistic interactions between geographically distant parts of the species range.
+Our choice to regulate population size by including a local density-dependent control on mortality
+is only one of many possible ways to do this.
+We could have instead regulated fecundity, or recruitment, or both;
+this general class of models
+is sometimes referred to as the ``Bolker--Pacala model'' \citep{bolker1997using}.
+It is not currently clear how much different choices of vital rates,
+or of functional forms for the regulation,
+might quantitatively affect our results,
+although the general predictions should be robust to similar forms of regulation.
+Since populations are still entirely \emph{intrinsically} regulated,
+our model still has a very strong ``population genetics'' flavor.
+Alternatively, population size could be regulated by interactions with other species
+(e.g., a Lotka-Volterra model),
+or extrinsically specified by local resource availability
+(e.g., by food or nest site availability).
+Indeed, our model could be interpreted as a caricature of such a model:
+as local density increases, good habitat is increasingly occupied,
+pushing individuals into more marginal habitat and increasing their mortality.
+Many such models should be behave similarly to ours,
+but others (especially those with local population cycling), may differ dramatically.
+There is a clear need for greater exploration of 
+the consequences for population genetics of ecologically realistic population models.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Supplement %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\beginappendix
-\section{Appendix 1}
-\llabel{appendix1}
+% \beginappendix
+\appendix
+\setcounter{table}{0}
+\renewcommand{\thetable}{A\arabic{table}}%
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{A\arabic{figure}}%
+
 %%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Comparisons with Stepping-Stone Models}
+\section{Comparisons with Stepping-Stone Models}
+\llabel{apx:stepping_stone}
+
 We checked that our model produces reasonable results by comparing it to a reverse-time stepping-stone model implemented in msprime \citep{Kelleher2016}.
@@
 low $m$), and the extent of the inflation increases to approach the continuous case as the resolution of the landscape increases.
-Similar pattterns are observed for $F_{is}$ and observed heterozygosity.
+Similar patterns are observed for $F_{is}$ and observed heterozygosity.
 However, $\theta_{W}$ behaves differently, with increased grid resolution leading to lower values.
@@
 \section{Demographic model}
-\label{apx:demographic_model}
-\llabel{dem_model_app}
+\llabel{apx:demographic_model}
 
@@
 
-\stopappendix
 \clearpage
 
-\beginsupplement
+% \beginsupplement
+\setcounter{table}{0}
+\renewcommand{\thetable}{S\arabic{table}}%
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{S\arabic{figure}}%
+
 \section{Supplementary Figures and Tables}
@@
 Tajima's $D$ & Difference in $\Theta_{/pi}$ and $\Theta_{/W}$ over its standard deviation\\
-Observed Heterzygosity & Proportion of heterozygous individuals in the sample \\
+Observed Heterozygosity & Proportion of heterozygous individuals in the sample \\
 $F_{IS}$ & Wright's inbreeding coefficient $1-H_{e}/H_{o}$ \\
@@
 
-\stopsupplement
-
 \newpage

### 2019-11-20 ms/review-responses.tex
grrr r3

@@
 
-%\plr{I would generally cut down the quoted bits like the above to only what's essential, but haven't done that yet.}
-
-% \plr{IMPORTANT: don't reorder or delete "points" below - it messes up the automatic numbering!}
+This is a good point. There are many ways to go in the quest for demographic realism,
+and the fundamental question -- what regulates population density in real populations --
+is still very much an open question (or rather, a question with a great many answers).
+As highlighted below,
+we have included further discussion of our particular choice
+of local, density-mediated control on mortality,
+along with pointers into the literature, at several places in the manuscript;
+see in particular in the Methods when we introduce the model \revreffull{AE}{2},
+and \llname{demog_disco}.
 
@@
 \reply{
-    Thanks for catching this - we have revised this caption to add details including the simulation parameters.
+    Thanks for catching this - we have revised this caption to add details including the simulation parameters. (Supplemental Figure \ref{fig:hetmap}).
 }
@@
     The mean IBS is simply the mean of the vector of lengths of IBS tracts.
+    \plr{make sure this is stated in the text somewhere}
     The short answer is that if one splits a chromosome up into more pieces (increasing nIBS),
@@
     Note that skew(IBS) and number of long ($>$1e6) IBS tracts reflect other aspects of the distribution.
-
 }
@@
 \reply{
-Good suggestion -- we have added a supplemental figure to show the distribution of individual fitness across the landscape with and without our edge-scaling approach.
-(Note though that we found the effect subtle and hard to visualize well).
+    Good suggestion -- we have added a supplemental figure (Figure \ref{fig:edge_scaling})
+    to show the distribution of individual fitness across the landscape
+    with and without our edge-scaling approach.
 }
@@
 \begin{point}{}
-    It would be helpful provide the figure showing that generating mutations during the forward simulations in SLIM is equivalent to applying mutations using msprime on pre-generated trees (line 185)?
+    It would be helpful provide the figure showing that generating mutations during the forward simulations in SLIM is equivalent to applying mutations using msprime on pre-generated trees \revref?
 It sounds like this procedure would underestimate the variance in the number of mutations, since you remove the effect of random generation time.
@@
 \reply{
-    We have added a figure showing sample site-frequency-spectra generated from a subset of simulations run with SLiM mutations, and then using msprime to apply mutations to the same tree sequences with our generation-time scaling approach (Figure \ref{fig:slim_v_msp_sfs}).
-These approaches yield extremely similar spectra so we believe any error in the approximation is quite small.
+    Theory says that any difference here would be extremely small
+    and would affect only the variance, not the mean.
+    Nonetheless, we have added a figure showing sample site frequency spectra
+    generated from a subset of simulations run with SLiM mutations,
+    and then using msprime to apply mutations to the same tree sequences with our generation-time scaling approach (Figure \ref{fig:slim_v_msp_sfs}).
+    These approaches yield extremely similar spectra.
 }
@@
 \reply{
-    We have added a brief discussion of the causes of the census size scaling to the relevant results section \llname{censusscaling}.
+    We agree that this is hard to intuit: there are a lot of factors at work.
+    We have added some more discussion of these phenomena \llname{censusscaling}.
 }
@@
 \begin{point}{}
-    Fig.
-7D: I am surprised by the extent to which the observed values of -log10(p) fall below the y=x line.
+    \autoref{fig:gwas}D: I am surprised by the extent to which the observed values of $-\log_{10}(p)$ fall below the $y=x$ line.
 Particularly in the lower right panel for large neighbourhood sizes.
@@
 
-\begin{point}{}
-    Lines 706-716, It might be worth citing Haworth et al Nature Communications 2019 (https://doi.org/10.1038/s41467-018-08219-1) who do the proposed test (GWAS for birth location) in UK Biobank to illustrate the population structure.
+\begin{point}{\revref}
+    It might be worth citing \citep{haworth2019apparent,},
+    who do the proposed test (GWAS for birth location) in UK Biobank to illustrate the population structure.
 \end{point}
@@
 \reply{
-Done - thank you for pointing us to this study. \llname{haworthcite}
+Done - thank you for pointing us to this study. \revref
 }
@@
 \reply{
-Apologies, I had forgotten to make the repo public. The link should work for everyone now. 
+We apologize -- we had forgotten to make the repository public. Now it is.
 }
@@
 
+We are glad you agree this is an important topic to explore!
+
 \begin{quote}
@@
 
+We were surprised that the reviewer seems to be saying that local, density-mediated control of mortality
+doesn't constitute a ``sound probabilisitic model'',
+but hope that our additional discussion of the choice,
+and citations to the history of this sort of model,
+help to better motivate our choice.
+
 \begin{point}{}
@@
 \reply{
-    This is a good point.
-We evaluated this while developing the simulation using the built-in visualization tools of the SLiM GUI, and have now added a supplementary figure (Figure \ref{fig:wm_v_dd}) showing the distribution of individuals in our density-dependant spatial model and a continuous-space Wright-Fisher simulation without density dependance.
+    This is a natural question that we asked ourselves while developing the simulation using the built-in visualization tools of the SLiM GUI.
+We have now added a supplementary figure (Figure \ref{fig:wm_v_dd}) showing the distribution of individuals in our density-dependant spatial model and a continuous-space Wright-Fisher simulation without density dependance.
+    We encourage the reviewer to run our SLiM recipe using the GUI at different parameter values
+    to see for themselves how quickly equilibrium is attained --
+    we are tempted to include movies of this as supplementary material,
+    but that seems excessive.
+    Also note that we did \emph{not} include competitive interaction to avoid clumping --
+    we included it to make the model biologically realistic
+    (so that, for instance, the population does not grow without bound),
+    and lack of clumping is a \emph{consequence} of this choice.
+    We have clarified this in the text. \revref
 }
 
+\begin{point}{}
 Second, the demographic process used here involves birth and death of individuals.
@@
 In fact, one needs to know a little about the dynamics of the population size to decide whether the corresponding process is reasonable from a biological standpoint.
-\begin{point}{}
-
 \end{point}
@@
 \reply{
-    Similarly here we monitored the population size over time while running SLiM. All simulations maintained asymptotic populations, and none of the runs we started ended because populations crashed.
+    As in the last point, as part of sanity checking our models
+    we monitored the population size over time while with SLiM.
+    All simulations maintained asymptotic populations, and none of the runs we started ended because populations crashed.
+    Although the population \emph{is} a Markov process with a single accessible absorbing state (death),
+    and (since it is not a branching process, so won't grow without bound)
+    will eventually die out with probability 1,
+    it will take an extremely long time to hit that state.
+    We have added a clarifying note about this. \revref
 }
@@
 \reply{
-   We have added additional discussion of this point to appendix A (see "Demographic Model") \llname{dem_model_app}
+   We explained the relationship here,
+   in a new Appendix (`Demographic Model'; \llname{apx:demographic_model}).
 }
@@
 \begin{point}{}
-    Also, the web page https://github.com/petrelharp/spaceness does not seem to exist so that I was not able to experiment with the forward-in-time generator used here unfortunately.
+    Also, the web page \url{https://github.com/petrelharp/spaceness} does not seem to exist so that I was not able to experiment with the forward-in-time generator used here unfortunately.
 \end{point}
@@
 \reply{
-I apologize -- I had forgotten to set the repo to public. It should work now. 
+We apologize -- we had forgotten to make the repository public. Now it is.
 }
@@
 \reply{
-    This is a good point also raised by other reviewers.
-We have added an Appendix comparing our model to a stepping-stone, rather than the spatial-Labmda-Fleming-Viot, because we think these are the most familiar and widely used class of spatial models.
-See Appendix 1.
+    Hopefully, the additional plots 
+    and discussion around demographic modeling (notably, the discussion at \llname{demog_disco})
+    helps to increase the transparency in this model.
+    We agree that demographic realism is a very important consideration,
+    and due to the often hard-to-intuit nature of spatial demography,
+    simulations need to be carefully sanity checked.
+    (Furthermore, what is ``realistic'' for one species will not be for another!)
+    We have added an Appendix comparing our model to a stepping-stone \llname{apx:stepping_stone}, 
+    because we think these are the most familiar and widely used class of spatial models.
     We find that many features of our model are well approximated by stepping-stone models, and that for statistics like $\theta_{\pi}$ the stepping stone model results approach our continuous space model as the number of demes used to describe the landscape increases.
@@
 
-\begin{point}{}
-    Figure 2: I do not understand why the neighborhood size varies to the same extent in the random mating model as it does for the spatial model.
+    We definately think that a comparison to the spatial Lambda-Fleming-Viot
+    would be an extremely useful thing to do, for the reasons the reviewer mentions,
+    but this would constitute an entirely different study.
+    The SLFV model is only known to be an approximation of one particular biologically explicit model
+    (of patchy extinction-recolonization)
+    and is only conjectured to be a good approximation
+    for spatial models more generally.
+    To reiterate, the definition of the SLFV was explicitly motivated by a desire to approximate
+    spatial models, especially Bolker-Pacala models like ours.
+    Therefore, our study is an important prerequisite
+    for simulation study of the spatial Lambda-Fleming-Viot,
+    as it works through some of the issues in simulating biologically realistic spatial models,
+    and describes some of the patterns.
+
+\begin{point}{\autoref{fig:genparams}:}
+    I do not understand why the neighborhood size varies to the same extent in the random mating model as it does for the spatial model.
 For the random mating model, I would have expected the neighborhood size to be equal to the census size since all individuals have the same probability of being a parent of any given offspring.
-From lines 166->171, it is clear that the spatial model would converge to the random mating model when the mean parent-offspring distance tends to infinity only if we were to ignore the impact of range edges.
+From [the paragraph at \revref], it is clear that the spatial model would converge to the random mating model when the mean parent-offspring distance tends to infinity only if we were to ignore the impact of range edges.
 I am thus wondering whether the variation of neighborhood size one observes in Fig 2 for the random mating model is just a consequence of border effects.

### 2019-11-20 ms/spaceness.tex
grrr r3

@@
 we decrease each individual's probability of survival with increasing local population density,
-which shifts reproductive output towards low-density regions 
-(and prevents the unrealistic population clustering seen by \citet{Felsenstein1975},
-see Supplemental \autoref{fig:slim_has_no_pain}).
+which shifts reproductive output towards low-density regions,
+and produces total census sizes that fluctuate around an equilibrium. \revpoint{3}{2}
+This also prevents the population clumping seen by \citet{Felsenstein1975}
+(Supplemental \autoref{fig:wm_v_dd})). \revpoint{3}{1}
 Such models have been used extensively in ecological modeling
@@
 we divided the mutation rate of $10^{-8}$ mutations per site per generation
-by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time.
-We show that this procedure produces the same site-frequency spectrum as applying mutations directly in SLiM in Figure \ref{fig:slim_v_msp_sfs}.
+by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time. \revpoint{2}{5}
+We verify that this procedure produces the same site frequency spectrum as applying mutations directly in SLiM in Figure \ref{fig:slim_v_msp_sfs},
+as predicted by theory \citep{ralph2019efficiently}.
 Simulations were run for 1.6 million timesteps (approximately $30N$ generations).
@@
 However, for spatial models this decline only begins for neighborhood size $\geq 10$. 
-By a neighborhood sizes larger than 100, the spatial and random mating models are indistinguishable from one another, 
-a sign that our simulations are performing as expected. 
+Spatial and random mating models are indistinguishable from one another
+for neighborhood sizes larger than 100.
 Census sizes range from $\approx 14,000$ at low $\sigma$ in the random mating model to $\approx 10,000$ for both models when neighborhood sizes approach 1,000.
 The scaling of census sizes in both random-mating and spatial models appears to be related to two consequences of the spatial competition function: the decline of fitness at range edges, which effectively reduces the habitable area by one $\sigma$ around the edge of the map and so results in a smaller habitable area at high $\sigma$ values; and variation in the equilibrium population density given varying competition radii.
-For spatial models we see an additional effect in which census size initially rises as neighborhood sizes scale from 2 - 10.
-This may reflect an Allee effect \citep{Allee1949} in which some individuals are unable to find mates when the mate selection radius is very small.
-\llabel{censusscaling}
+Furthermore, census size increases in spatial models as neighborhood size increases from 2 to 10.
+This may reflect an Allee effect \citep{Allee1949} in which some individuals are unable to find mates when the mate selection radius is very small.  \llabel{censusscaling}
 
@@
 Spatially correlated environmental factors cause false positives at a large proportion of SNPs, which is partially but not entirely corrected by adding the first 10 PC coordinates as covariates (\textbf{C}).
-Quantile-quantile plots in \textbf{D} show inflation of $-\log_{10}(p)$ after PC correction across all simulations and environmental distributions, with colors scaled by the median neighborhood size in each region of q-q space.
+Quantile-quantile plots in (\textbf{D}) show inflation of $-\log_{10}(p)$ after PC correction across all simulations and environmental distributions, with colors scaled by the median neighborhood size in each region of q-q space.
 }
@@
 This is essentially the approach taken in our ``clinal'' model (though we add normally distributed noise to our phenotypes).
-This approach has recently been taken with polygenic scores for UK Biobank samples in \cite{haworth2019apparent}, finding that scores are correlated with birth location even in this relatively homogenous sample \llabel{haworthcite}.
+This approach has recently been taken with polygenic scores for UK Biobank samples in \cite{haworth2019apparent}, finding that scores are correlated with birth location even in this relatively homogenous sample.
+\revpoint{2}{8}
 Of course, it is possible that genotypes indirectly affect individual locations by adjusting organismal fitness and thus habitat selection across spatially varying environments, but we believe that this hypothesis should be tested against a null of stratification bias inflation rather than accepted as true based on GWAS results.
@@
 \section{Supplementary Figures and Tables}
-\FloatBarrier
 
@@
 The clustering seen in the top row is the "Pain in the Torus" described by \citet{Felsenstein1975}.}
-\label{fig:slim_has_no_pain} 
+\label{fig:wm_v_dd} 
 \end{figure*}

### 2019-11-20 ms/review-responses.tex
.

@@
 \reply{
-    We have a added a brief discussion of the census size scaling to \llname{censusscaling}
+    This is indeed part of what's going on.
+    We have a added some discussion of the census size scaling \llname{censusscaling}.
 }
 
-\begin{point}{}
-    Line 729-731: ``Many more species occur in a middle range of neighborhood sizes between 100 and 1000 - a range in which spatial processes play a minor role in our analyses [...]''.
+\begin{point}{\revref}
+    ``Many more species occur in a middle range of neighborhood sizes between 100 and 1000 - a range in which spatial processes play a minor role in our analyses [...]''.
 Do the authors think that the spatial processes would still play a minor role when neighborhood sizes exceed 100-1000 if the habitat was larger than that taken in the present simulations?

### 2019-11-20 ms/spaceness.tex
.

@@
 
-To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, and density-dependent fitness, 
+To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, and density-dependent fitness,  \revpoint{3}{6}
 we also implemented simulations identical to those above except that mates are selected uniformly at random from the population,
@@
 These include some human populations such as the Gainj- and Kalam-speaking people of Papua New Guinea, in which the estimated neighborhood sizes in \citep{Rousset1997} range from 40 to 410 depending on the method of estimation.
+\revpoint{3}{7}
 Many more species occur in a middle range of neighborhood sizes between 100 and 1000 -- a range in which spatial processes play a minor role in our analyses under random spatial sampling but are important when sampling of individuals in space is clustered.

### 2019-11-20 ms/spaceness.tex
.

@@
 One facet of spatial variation that we did not address in this study is the confounding of dispersal and population density implicit in the definition of Wright's neighborhood size.
-Our simulations were run under constant densities, but \cite{guindon2016demographic} and \citet{Ringbauer2017} have shown that these parameters are identifiable under some continuous models.
+Our simulations were run under constant densities, but \cite{guindon2016demographic} and \citet{Ringbauer2017} \revpoint{3}{8}
+have shown that these parameters are identifiable under some continuous models.
 Similarly, though the scaling effects of dispersal we show in Figure \ref{fig:sumstats} should occur in populations of any total size, other aspects such as the number of segregating sites are also likely affected by the total landscape size (and so total census $N$)\llabel{landscapesizecaveat}.

### 2019-11-21 ms/review-responses.tex
.
line qqplots with associated text edits

@@
 \reply{
-    We have checked the code to the best of our abilities and did not find a bug causing the underdispersion.
-It seems to reflect overcorrection in the regression when using PC coordinates as covariates -- the PCA is capturing some information about the spatial genetic variation which itself covaries only weakly with the phenotype, and as a result we see anomalously low $-log10(p)$ when regressing genotype against phenotype.
+    We have checked the code to the best of our abilities and did not find a bug causing the underdispersion, but we have switched our visualization in Figure 8D to now show a subset of qqplots as lines, rather than the hexagonal binning over all simulations.
+This lets us see the density in addition to the range of values, and shows that relatively few points fall below the 1:1 lines.
+Those that do seem to reflect overcorrection in the regression when using PC coordinates as covariates -- the PCA is capturing some information about the spatial genetic variation which itself covaries only weakly with the phenotype, and as a result we see anomalously low $-log10(p)$ when regressing genotype against phenotype.
 }
@@
     This is indeed part of what's going on.
-    We have a added some discussion of the census size scaling \llname{censusscaling}.
+    We have added some discussion of the census size scaling \llname{censusscaling}.
 }
@@
 
-\begin{point}{}
-    Line 753-757: please add a reference to Guindon, Guo and Welch (2016).
+\begin{point}{\revref}
+    Please add a reference to Guindon, Guo and Welch (2016).
 This study clearly shows that population density and dispersal parameters are identifiable and can indeed be estimated in practice under the spatial Lambda-Fleming-Viot model.
@@
 \reply{
-Done. Thank you for pointing us to this study. 
+Done. Thank you for pointing us to this study.  \revref
 }

### 2019-11-21 ms/spaceness.tex
line qqplots with associated text edits

@@
 
-Figure \ref{fig:gwas}D shows quantile-quantile plots that show the degree of genome-wide inflation of test statistics in PC-corrected GWAS across all simulations and environmental distributions. 
+Figure \ref{fig:gwas}D shows quantile-quantile plots for a subset of simulations that show the degree of genome-wide inflation of test statistics in PC-corrected GWAS across all simulations and environmental distributions.
 For clinal environments, $-\log_{10}(p)$ values are most inflated when neighborhood sizes are large, consistent with the pattern observed in the count of significant associations after PC regression.
@@
 \centering
-\includegraphics[width=\textwidth]{figures/gwas_summary_loglog.pdf}
+\includegraphics[width=\textwidth]{figures/gwas_summary_loglog_qqlines.pdf}
     \caption{Impacts of spatially varying environments and isolation by distance on linear regression GWAS.
@@
 Spatially correlated environmental factors cause false positives at a large proportion of SNPs, which is partially but not entirely corrected by adding the first 10 PC coordinates as covariates (\textbf{C}).
-Quantile-quantile plots in (\textbf{D}) show inflation of $-\log_{10}(p)$ after PC correction across all simulations and environmental distributions, with colors scaled by the median neighborhood size in each region of q-q space.
+Quantile-quantile plots in (\textbf{D}) show inflation of $-\log_{10}(p)$ after PC correction for simulations with spatially structure environments, with line colors scaled to the neighborhood size of each simulation.
 }

### 2019-11-21 ms/spaceness.tex
adding hexbin qqplots as supp fig

@@
 Figure \ref{fig:gwas}D shows quantile-quantile plots for a subset of simulations that show the degree of genome-wide inflation of test statistics in PC-corrected GWAS across all simulations and environmental distributions.
+An alternate visualization is also included in figure \ref{fig:qqplot_hexbins}.
 For clinal environments, $-\log_{10}(p)$ values are most inflated when neighborhood sizes are large, consistent with the pattern observed in the count of significant associations after PC regression.
@@
 \label{fig:manhattan_plots}
+\end{figure*}
+\FloatBarrier
+
+\begin{figure*}[p]
+\centering
+\includegraphics[width=0.8\textwidth]{figures/qqplots_hexbin.pdf}
+\caption{Quantile-quantile plots showing observed and expected $-log10(p)$ for PC-corrected GWAS run on simulations with varying neighborhood sizes and environmental distributions.
+Hexagonal bins are colored by the average neighborhood size of simulations with points falling in a given region of quantile-quantile space.
+Qqplots for a subset of these simulations are shown as lines in Figure \ref{fig:gwas}D.}
+\label{fig:qqplot_hexbins}
 \end{figure*}

### 2019-11-22 ms/review-responses.tex
my pass on responses

@@
 \begin{quote}
-    This study explores biases arising in population-based inference when 1)
-    real population samples are coming from spatial habitat with various degree
-    of structuring while inference is made assuming random mating population;
-    2) imperfect sampling in practice that fails to represent full diversity
-    across entire population habitat; 3) phenotypes that vary across geography
-    and create spurious associations with genotypes. While earlier studies
-    explored the effect of strong structure on population genetic inference and
-    GWAS, this work focuses on less extreme scenarios of structuring that
-    arises in populations evolving in continuous habitat. By using
-    non-Wright-Fisher model, authors simulated chromosome-scale samples from
-    populations that evolved in continuous space, and that can model
-    environmental factors to create phenotypes varying over space. As a result,
-    this study identified spatial structuring scenarios (small neighbourhood
-    size $\sim$ 10--100) that coupled with imperfect sampling strategies lead to a
-    biased inference of widely used population genetic statistics (altogether
-    18 statistics) such as pi (average pairwise sequence differences),
-    heterozygosity (and inbreeding coefficient), and IBS tract sharing.
-    Accordingly, inference of the effective population size history was also
-    strongly affected under these parameter ranges. Finally, the authors use
-    their spatial modelling to demonstrate that typical GWAS with PC-based
-    correction cannot entirely remove spurious signals of genotype-phenotype
-    associations arising from purely environmental factors.  Overall, the
+    % This study explores biases arising in population-based inference when 1)
+    % real population samples are coming from spatial habitat with various degree
+    % of structuring while inference is made assuming random mating population;
+    % 2) imperfect sampling in practice that fails to represent full diversity
+    % across entire population habitat; 3) phenotypes that vary across geography
+    % and create spurious associations with genotypes. While earlier studies
+    % explored the effect of strong structure on population genetic inference and
+    % GWAS, this work focuses on less extreme scenarios of structuring that
+    % arises in populations evolving in continuous habitat. By using
+    % non-Wright-Fisher model, authors simulated chromosome-scale samples from
+    % populations that evolved in continuous space, and that can model
+    % environmental factors to create phenotypes varying over space. As a result,
+    % this study identified spatial structuring scenarios (small neighbourhood
+    % size $\sim$ 10--100) that coupled with imperfect sampling strategies lead to a
+    % biased inference of widely used population genetic statistics (altogether
+    % 18 statistics) such as pi (average pairwise sequence differences),
+    % heterozygosity (and inbreeding coefficient), and IBS tract sharing.
+    % Accordingly, inference of the effective population size history was also
+    % strongly affected under these parameter ranges. Finally, the authors use
+    % their spatial modelling to demonstrate that typical GWAS with PC-based
+    % correction cannot entirely remove spurious signals of genotype-phenotype
+    % associations arising from purely environmental factors.
+    Overall, the
     authors explore an important but often neglected source of bias that can
@@
 \reply{
-    We agree that this distinction is worth emphasizing!
-    We've adjusted our language to hopefully remind the reader that
+    We agree and have
+    adjusted our language to underscore the point that
     $K$ is a parameter that controls population density, rather than being equal to it,
@@
     to the continuous model as the resolution of the landscape increases,
-    but other statistics seems to be more strongly affected by discretization artifacts.
-    We are also curious how all the other analyses we test would be affected using other simulation schemes, but hope to explore that aspect in future work. 
+    but other statistics are more strongly affected by discretization.
+    We are also curious how all the other analyses we perform would be affected using other simulation schemes, but hope to explore that aspect in future work. 
     Furthermore, discrete-deme coalescent simulations are certainly faster,
@@
 \reply{
-We apologize -- we had forgotten to make the repository public. Now it is.
+We apologize -- we had forgotten to make the repository public. This has been corrected.
 }
@@
 If it does not, then is there any stationary regime and how many generations does one need to wait before reaching it?
-I do appreciate that the competitive interaction term was introduced in order to avoid seeing the ``clumping' of individuals that hampers the Wright-Malecot model.
+I do appreciate that the competitive interaction term was introduced in order to avoid seeing the ``clumping'' of individuals that hampers the Wright-Malecot model.
 Yet, just because there are no such clusters does not mean that the spatial distribution of individuals reaches a stable regime and that the distribution reached, if any, is reasonable from a biological perspective.
@@
     would be an extremely useful thing to do, for the reasons the reviewer mentions,
-    but this would constitute an entirely different study.
+    but this is beyond the scope of the current study.
     The SLFV model is only known to be an approximation of one particular biologically explicit model
@@
 \reply{
-    Good point on EEMS. We have added the most recent EEMS paper to the citations on \llname{eemscitation}, and have added an Appendix giving a more thorough comparison with stepping-stone models. 
+    We agree that not noting EEMS was an oversight.
+We have added the most recent EEMS paper to the citations on \llname{eemscitation}, and have added an Appendix giving a more thorough comparison with stepping-stone models.
 }
@@
 This is important because, if mixed models do provide stronger control for stratification they are also likely to remove true signals of causal SNPs if those SNPs covary with spatial structure.
-We now point to these methods explicitly in the discussion \llname{mixedmodelcallout}, but think that incorporating that study here would make this paper too long.
-We also think the PC results are still quite relevant as the method is still seen in many studies.
+We now point to these methods explicitly in the discussion \llname{mixedmodelcallout}, but think that incorporating that study here would make this paper too long and we plan to explore this timely issue in the future.
+We also think the PC results are still quite relevant as the method is quite population in current studies.
 }
@@
 \reply{
-	We switched the second instance (referring to the distance among individuals) to $d$. 
+	We have clarified the notation for the second instance (referring to the distance among individuals) to $d$. 
 }

### 2019-11-22 ms/spaceness.tex
took a pass on the main text

@@
 this regulation must be local,
-and there are many ways to arrange this \citep{bolker2003spatial}.
+and there are many ways to achieve this \citep{bolker2003spatial}.
 In our simulations,
@@
 Since $1 - p \approx n / (K (1 + L))$,
-mortalitity goes up roughly linearly with the number of neighbors (on a scale given by $K$),
+mortality goes up roughly linearly with the number of neighbors (on a scale given by $K$),
 as would be obtained if, for instance, mortality is due to agonistic interactions.
-Ignoring in/outmigration,
+Ignoring migration,
 a region is at demographic equilibrium if the per-capita probability of death

### 2019-11-22 ms/review-responses.tex
editz

@@
     All simulations maintained asymptotic populations, and none of the runs we started ended because populations crashed.
-    Although the population \emph{is} a Markov process with a single accessible absorbing state (death),
-    and (since it is not a branching process, so won't grow without bound)
-    will eventually die out with probability 1,
-    it will take an extremely long time to hit that state.
+    We agree that the population model will die out eventually with probability 1
+    (it is a Markov process with a single accessible absorbing state (death),
+    and since it isn't a branching process, doesn't grow without bound).
+    However, it will take an extremely long time to hit that state,
+    so that extinction is not a practical consideration.
     We have added a clarifying note about this. \revref
@@
     spatial models, especially Bolker-Pacala models like ours.
-    Therefore, our study is an important prerequisite
+    Therefore, our study does important preliminary work
     for simulation study of the spatial Lambda-Fleming-Viot,
-    as it works through some of the issues in simulating biologically realistic spatial models,
+    since it works through some of the issues in simulating biologically realistic spatial models,
     and describes some of the patterns.
@@
 \reply{
+    It's a good point that neighborhood size, as defined to be ``number of possible parents''
+    should indeed be the census size.
+    However, we are using $N_W$ defined in the same way for both spatial and random mating simulations
+    as a way of comparing spatial to ``nonspatial'' models
+    keeping other aspects of demography (almost) the same.
+    We've added a note clarifying this, at \revref and \llname{randomNW}.
     This is indeed part of what's going on.
-    We have added some discussion of the census size scaling \llname{censusscaling}.
+    We have added some discussion of the census size scaling \llname{censusscaling},
+    but haven't added a detailed discussion of whether or not most species experience
+    decreased habitat quality at range edges,
+    since that seems like it is getting too far into the weeds.
 }

### 2019-11-22 ms/spaceness.tex
added a sentence to the demog model section for clarity
editz

@@
 we empirically calculated these demographic quantities in simulations. 
-We recorded the census population size in all simulations. 
+We recorded the census population size in all simulations,
+and used mean population density ($\rho$, census size divided by total area)
+to compute neighborhood size as $N_W = 4 \pi \rho \sigma^2$. \revpoint{3}{6}
 To estimate generation times, we stored ages of the parents of every new individual born across 200 timesteps, 
@@
 for neighborhood sizes larger than 100.
-Census sizes range from $\approx 14,000$ at low $\sigma$ in the random mating model to $\approx 10,000$ for both models when neighborhood sizes approach 1,000.
+Census sizes range from around 14,000 at low $\sigma$ in the random mating model to 10,000 for both models when neighborhood sizes approach 1,000.
 The scaling of census sizes in both random-mating and spatial models appears to be related to two consequences of the spatial competition function: the decline of fitness at range edges, which effectively reduces the habitable area by one $\sigma$ around the edge of the map and so results in a smaller habitable area at high $\sigma$ values; and variation in the equilibrium population density given varying competition radii.
+Furthermore, lack of available mates contributes to lower census sizes at small $N_W$
+for the spatial model
+(We calculate $N_W$ for both in the same way,
+even though the number of ``available mates'' is much larger with random mating). \llabel{randomNW}
 Furthermore, census size increases in spatial models as neighborhood size increases from 2 to 10.
@@
     Spatially correlated environmental factors cause false positives at a large proportion of SNPs, which is partially but not entirely corrected by adding the first 10 PC coordinates as covariates (\textbf{C}).
-Quantile-quantile plots in (\textbf{D}) show inflation of $-\log_{10}(p)$ after PC correction for simulations with spatially structure environments, with line colors scaled to the neighborhood size of each simulation.
+    Quantile-quantile plots in (\textbf{D}) show inflation of $-\log_{10}(p)$ after PC correction for simulations with spatially structured environments, with line colors showing the neighborhood size of each simulation.
 }
@@
 For how much of the tree of life do spatial patterns circumscribe genomic variation?
-In Table \ref{table:NStable} we gathered estimates of neighborhood size from a range of organisms to get an idea of how likely dispersal is to play an important role in patterns of variation.
-These values should be compared to our simulation results with some caution, as though we expect neighborhood size to have similar effects across species of varying global $N_e$ \citep{Wright1946}, in our study we evaluated only a relatively small population of $\approx 10,000$.
+In Table \ref{table:NStable} we gathered estimates of neighborhood size from a range of organisms
+to get an idea of how strongly local geographic dispersal affects patterns of variation.
+This is an imperfect measure:
+some aspects of genetic variation are most strongly determined by neighborhood size \citep{Wright1946},
+others (e.g., number of segregating sites)
+are more strongly determined by global $N_e$ or by the ratio of the two.
+% These values should be compared to our simulation results with some caution, as though we expect neighborhood size to have similar effects across species of varying global $N_e$ \citep{Wright1946}, in our study we evaluated only a relatively small population of $\approx 10,000$.
 \llabel{popsizecaveat}
@@
 \llabel{apx:demographic_model}
-
+Local population regulation is controlled by two parameters, $L$, and $K$.
+Here, we show that these should be close to the average lifespan of an individual
+and the average number of individuals per unit area, respectively.
 We chose our demographic model so that every individual has on average $1/L$ offspring

### 2019-12-04 ms/spaceness.tex
minor edit to appendix
all the way through

@@
 (e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation,Wilkins2002linearcoalescent,Wilkins2004septimescales}), \revpoint{AE}{1}
-much modern work still describes geographic structure as a set of discrete populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations,al2019estimating}\llabel{eemscitation}.
+much modern work still describes geographic structure as a set of discrete populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations} or as an average over such discrete models \citep{Petkova2015,al2019estimating}.
+\revpoint{4}{3}
 For this reason, most population genetics statistics are interpreted with reference to discrete, well-mixed populations, and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000} with methods that assume these are randomly mating units.
@@
 Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
-in both the $x$ and $y$ coordinates.
+in both the $x$ and $y$ coordinates. \revpoint{4}{8}
 Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
@@
 
-To check that our model produces reasonable results, we compared its output to that of a stepping-stone model implemented in msprime \citep{Kelleher2016}.
-These results are discussed in detail in Appendix 1, but in general we find that our model produces many of patterns generated by stepping stone models while avoiding some artifacts associated with discretization of the landscape.
+We also compared our model's output to a commonly-used approximation of continuous space,
+the stepping-stone model, which we simulated with \texttt{msprime} \citep{Kelleher2016}.
+These results are discussed in detail in the Appendix,
+but in general we find that the demographic structure of a stepping-stone model
+can depend strongly on the chosen discretization,
+and some artifacts of discretization do not seem to dissappear in the limit of a fine grid.
+For many summary statistics, finer discretizations (we used a $50 \times 50$ grid)
+produced similar results to the continuous model,
+but this was not true for others (e.g., $F_{IS}$ and Tajima's $D$),
+which differed from the continuous model \emph{more} at finer discretizations.
 
@@
 
-Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among sampled chromosomes.
+Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among sampled chromosomes,
+defined to be the set of distances between adjacent sites that differ between the two chromosomes
+of each diploid. \revpoint{1}{1}
 The full distribution of lengths of IBS tracts for each pair of chromosomes was first calculated with a custom python function. 
-%\url{https://github.com/petrelharp/spaceness/blob/master/scripts/slimtools.py}. 
+%\url{https://github.com/cjbattey/spaceness/blob/master/scripts/slimtools.py}. 
 We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $10^6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.
@@
 The effects of sampling on summary statistic estimates were summarized by testing for differences in mean (ANOVA, \citep{Rcore2018}) and variance (Levene's test, \citep{Fox2011}) across sampling strategies for each summary statistic.
-
-Last, we used a subset of summary statistics to compare how our continuous model performs relative to a reverse-time stepping stone model (Appendix 1).
 
@@
 we increased the mean phenotype from 100 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
-Concretely, the phenotype $p$ for an individual at position $(x,y)$ is $p=100 + 2x / 5$.
+Concretely, the phenotype $p$ for an individual at position $(x,y)$ is $p=100 + 2x / 5$. \revpoint{4}{10}
 Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype 
@@
 Run times for our model scale approximately linearly with neighborhood size (Figure \ref{fig:runtimes}), with the lowest neighborhood sizes reaching 30N generations in around a day and those with neighborhood size approaching 1,000 requiring up to three weeks of computation.
-As currently implemented running simulations at neighborhood sizes $>1000$ to coalescence is likely impractical, though running these models for more limited timescales and then "completing" the simulation by starting a reverse-time simulation from the resulting tree sequence in msprime is possible.
+As currently implemented running simulations at neighborhood sizes $>1000$ to coalescence is likely impractical, though running these models for more limited timescales and then ``completing'' the simulation by starting a reverse-time simulation from the resulting tree sequence in msprime is possible.
 \llabel{runtimesxn}
@@
 \includegraphics[width=\textwidth]{figures/sfs_w_sumstats_loglog.pdf}
-\caption{Log-scaled site frequency spectrum (A) and summary statistic distributions (B) by sampling strategy and neighborhood size.}
+\caption{Site frequency spectrum (A; note axes are log-scaled) and summary statistic distributions (B) by sampling strategy and neighborhood size.}
 \label{fig:sumstats}
@@
 Though this degree of variation in population size is certainly meaningful in an ecological context, it is relatively minor in population genetic terms.
-\llabel{ibdnesxn} In general methods directly assessing haplotype structure in phased data (for example, \citet{browning2015accurate}) are thought to provide increased resolution for recent demographic events, but in this case the error we observed was essentially an accurate reflection of underlying genealogies in which terminal branches are anomalously short.
+\llabel{ibdnesxn}
+Methods directly assessing haplotype structure in phased data (for example, \citet{browning2015accurate}) are thought to provide increased resolution for recent demographic events,
+but in this case the error we observed was essentially an accurate reflection of underlying genealogies in which terminal branches are anomalously short.
 Combined with our analysis of IBS tract length variation (Figure \ref{fig:ibs}) this suggests that haplotype-based methods are likely to be affected by similar biases.
@@
 
-We checked that our model produces reasonable results by comparing it to a reverse-time stepping-stone model implemented in msprime \citep{Kelleher2016}.
+We also compared our model results to a regular grid of discrete populations,
+which is commonly used as an approximation of continuous geography.
+An important reason that this approximation is often made
+is that it allows more efficient, coalescent simulations;
+we implemented these using \texttt{msprime} \citep{Kelleher2016}.
 In this class of models we imagine an $n \times n$ grid of populations exchanging migrants with neighboring populations at rate $m$.
-If these models are good approximations of the continuous case we expect that results will converge as $n \to \infty$, so we ran simulations while varying $n$ from 5 to 50 (Table \ref{table:grid_params}).
-To compare with continuous models we first distributed the the same "effective" number of individuals across the landscape as in our continuous-space simulations ($\approx 6100$, estimated from $\theta_{\pi}$ of random-mating continuous-space simulations).
-We then approximate the mean per-generation dispersal distance $\sigma$ given a total landscape width $W$ as the product of the probability of an individual being a migrant and the distance traveled by migrants: $\sigma \approx 4m(W/n)$.
-We ran 500 simulations per $n$ while sampling $\sigma$ from $U(0.2,4)$.
+If these models are good approximations of the continuous case we expect that results
+will converge as $n \to \infty$
+(while scaling $m$ appropriately and keeping total population size fixed),
+so we ran simulations while varying $n$ from 5 to 50 (Table \ref{table:grid_params}).
+To compare with continuous models we first distributed the the same ``effective'' number of individuals across the landscape as in our continuous-space simulations ($\approx 6100$, estimated from $\theta_{\pi}$ of random-mating continuous-space simulations).
+We then approximate the mean per-generation dispersal distance $\sigma$ given a total landscape width $W$ as the product of the probability of an individual being a migrant and the distance traveled by migrants: $\sigma = 4m(W/n)$.
+This means that $m$ in different simulations with the same $\sigma$ scales with $\sqrt{n}$.
+We ran 500 simulations for each value of $n$ while sampling $\sigma$ from $U(0.2,4)$.
 We then randomly selected 60 diploid individuals from each simulation (approximating diploidy by combining pairs of chromosomes with contiguous indices within demes) and calculated a set of six summary statistics using the scripts described in the summary statistics portion of the main text.
@@
 \caption{Summary statistics for 2-dimensional coalescent stepping-stone models with fixed total $N_{e}$ and varying numbers of demes per side.
-The black "infinite" points are from our forward-time continuous space model.
+The black ``infinite'' points are from our forward-time continuous space model.
 Inter-deme migration rates are related to $\sigma$ as described above.}
@@
 \caption{Maps of individual locations in a continuous-space Wright-Malécot model with independent dispersal of all individuals (top) and under our continuous space model incorporating density-dependant fitness (bottom).
-The clustering seen in the top row is the "Pain in the Torus" described by \citet{Felsenstein1975}.}
+The clustering seen in the top row is the ``Pain in the Torus'' described by \citet{Felsenstein1975}.}
 \label{fig:wm_v_dd} 

### 2019-12-04 ms/review-responses.tex

@@
 ``Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data''.
+We thank the AE and revewiers for constructive comments,
+which we have addressed in detail, below.
+The main changes we have made have to do with our choice of model.
+We have added some more discussion
+of the biological motivation for the particular model of continuous space we use
+(with references to the ecological literature, where it appears) to the Methods (around \llname{ecolcites}),
+and in the Discussion (at \llname{demog_disco}).
+We have included more explanation of some of the more opaque consequences
+of a more realistic demographic model, around \llname{censusscaling}.
+Finally, we have added a substantial comparison to the discrete stepping-stone model,
+most of which is in the Appendix.
 
@@
 \reply{
-    The mean IBS is simply the mean of the vector of lengths of IBS tracts.
-    \plr{make sure this is stated in the text somewhere}
+    The mean IBS is simply the mean of the vector of lengths of IBS tracts \revref.
     The short answer is that if one splits a chromosome up into more pieces (increasing nIBS),
@@
 \reply{
-    Additional discussion of the model is now provided in the appendix (see "Demographic model") \llname{apx:demographic_model}
+    Additional discussion of the model is now provided in the appendix (see ``Demographic model'') \llname{apx:demographic_model}
 }
@@
 \begin{point}{\revref}
-    It might be worth citing \citep{haworth2019apparent,},
+    It might be worth citing \citep{haworth2019apparent},
     who do the proposed test (GWAS for birth location) in UK Biobank to illustrate the population structure.
@@
 \reply{
-    It's a good point that neighborhood size, as defined to be ``number of possible parents''
+    It's a good point that neighborhood size, as defined to be ``number of possible parents'',
     should indeed be the census size.
@@
     keeping other aspects of demography (almost) the same.
-    We've added a note clarifying this, at \revref and \llname{randomNW}.
+    We've added a note clarifying this, at \revref{} and \llname{randomNW}.
     This is indeed part of what's going on.
@@
    We have expanded our citations of some of the relevant ecology literature \llname{ecolcites}, which we hope helps to contextualize the study better.
+   We haven't cited dadi, which works with discrete populations
+   (the ``diffusion'' is in frequency space).
 }
@@
     We agree that not noting EEMS was an oversight.
-We have added the most recent EEMS paper to the citations on \llname{eemscitation}, and have added an Appendix giving a more thorough comparison with stepping-stone models.
+    We have added the most recent EEMS paper to the citations at \revref,
+    and have added an Appendix giving a more thorough comparison with stepping-stone models. 
 }
@@
 \begin{point}{}
-    Code availability. This github link doesn't work, but is important to be able to evaluate for review: https://github.com/petrelharp/spaceness
+    Code availability. This github link doesn't work, but is important to be able to evaluate for review: \url{https://github.com/petrelharp/spaceness}.
 \end{point}
@@
 \reply{
-Apologies, it was accidentally set to private. The link should work now. 
+Apologies, it was accidentally set to private. The link should work now
+(and has been replaced by the primary link, \url{https://github.com/cjbattey/spaceness}).
 }
@@
     Definitions and interpretations.
-There are quite a large number of metrics discussed in Figure 3B, and it's a lot to take in.
+There are quite a large number of metrics discussed in Figure \ref{fig:sumstats}, and it's a lot to take in.
 It might be helpful to have a table with a reminder of what the metric is, its interpretation, and how it is computed.
@@
 \reply{
-    We have included a table describing the summary statistics in Figure \ref{table:sumstats}.
+    We have included a table describing the summary statistics in \autoref{table:sumstats}.
 }
 
-\begin{point}{}
+\begin{point}{\revref}
     Notation: ``Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$ in both the x and y coordinates.
@@
 \reply{
-	We have clarified the notation for the second instance (referring to the distance among individuals) to $d$. 
+	We have clarified the notation for the second instance (referring to the distance among individuals) to $d$. \revref
 }
@@
 
-\begin{point}{}
+\begin{point}{\revref}
     Not sure I follow this example: ``Concretely, an individual at position (x, y) in a 50 × 50 landscape has mean phenotype 100 + 2x/5.''
@@
 \reply{
-	We have clarified by switching to "the phenotype $p$ for an individual at location (x,y) is then $p=100+2x/5$
+	We have clarified by switching to ``the phenotype $p$ for an individual at location (x,y) is then $p=100+2x/5$''. \revref
 }

### 2019-12-05 ms/review-responses.tex
fine-toothed comb pass

@@
 \reply{
-We have revised this sentence to clarify that we aim to produce a variation in mean phenotype of two standard deviations across the landscape \llname{phenotypescalingsd}.
+We have revised this sentence to clarify that we aim to produce a variation in mean phenotype of two standard deviations across the landscape \revref.
 }
@@
     keeping other aspects of demography (almost) the same.
-    We've added a note clarifying this, at \revref{} and \llname{randomNW}.
     This is indeed part of what's going on.
-    We have added some discussion of the census size scaling \llname{censusscaling},
+    We have added some discussion of the census size scaling at \revref{} and \llname{censusscaling},
     but haven't added a detailed discussion of whether or not most species experience
@@
 Apologies, it was accidentally set to private. The link should work now
-(and has been replaced by the primary link, \url{https://github.com/cjbattey/spaceness}).
+(and has been replaced by the primary link, \url{https://github.com/kern-lab/spaceness}).
 }

### 2019-12-05 ms/spaceness.tex
fine-toothed comb pass

@@
 Despite a long history of analytical work describing the genetics of populations distributed across continuous geography
-(e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation,Wilkins2002linearcoalescent,Wilkins2004septimescales}), \revpoint{AE}{1}
-much modern work still describes geographic structure as a set of discrete populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations} or as an average over such discrete models \citep{Petkova2015,al2019estimating}.
+(e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation,Wilkins2002linearcoalescent,Wilkins2004separation}), \revpoint{AE}{1}
+much modern work still describes geographic structure as a set of discrete populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2019populations} or as an average over such discrete models \citep{Petkova2015,al2019estimating}.
 \revpoint{4}{3}
@@
 and coalescence (the chance that they share a parent).
-A key early observation by Wright \citep{Wright1946} is that this balance is often nicely summarized by the ``neighborhood size'',
+A key early observation by \citet{Wright1946} is that this balance is often nicely summarized by the ``neighborhood size'',
 defined to be $N_W = 4 \pi \rho \sigma^2$, where $\sigma$ is the mean parent--offspring distance and $\rho$ is population density. 
@@
 Although these grid models may be good approximations of continuous geography in many situations,
-they do not model demographic fluctuations, and limit investigation of spatial structure below the level of the deme,
+they do not model demographic fluctuations,
+and limit investigation of spatial structure below the level of the deme,
 assumptions whose impacts are unknown. 
@@
 
-To avoid hard-to-evaluate approximations, we here used forward-time, individual-based simulations.
+To avoid hard-to-evaluate approximations, we here used forward-time, individual-based simulations
+across continuous geographical space.
 The question of what regulates real populations has a long history and many answers
@@
 In our simulations,
-we decrease each individual's probability of survival with increasing local population density,
+each individual's probability of survival is a decreasing function of local population density,
 which shifts reproductive output towards low-density regions,
@@
 
-We simulated populations using the non-Wright-Fisher module in the program SLiM v3.1 \citep{Haller2019}. 
+We simulated populations using the program SLiM v3.1 \citep{Haller2019}. 
 Each time step consists of three stages: reproduction, dispersal, and mortality. 
@@
 At the beginning of the simulation individuals are distributed uniformly at random on a continuous, square landscape. 
-Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$ where $L$ is the expected lifespan.
+Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$.
 Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
@@
 and the mean lifetime is around $L$;
-for more description see Appendix \ref{apx:demographic_model}.
+for more description see the Appendix. % \ref{apx:demographic_model}.
 
@@
 as implemented in SLiM \citep{haller2019treesequence}. 
-Scripts for figures and analyses are available at \url{https://github.com/petrelharp/spaceness}.
+Scripts for figures and analyses are available at \url{https://github.com/kern-lab/spaceness}.
 
@@
 average lifetime $L=4$, 
-genome size = $10^{8}$, 
-recombination rate = $10^{-9}$, 
+genome size $10^{8}$bp, 
+recombination rate $10^{-9}$ per bp per generation, 
 and drawing $\sigma$ values from a uniform distribution between 0.2 and 4. 
@@
 We verify that this procedure produces the same site frequency spectrum as applying mutations directly in SLiM in Figure \ref{fig:slim_v_msp_sfs},
-as predicted by theory \citep{ralph2019efficiently}.
+in agreement with theory \citep{ralph2019efficiently}.
 Simulations were run for 1.6 million timesteps (approximately $30N$ generations).
@@
 To estimate variance in offspring number, 
-we tracked the number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period, subset the resulting table to include only the last timestep recorded for each individual, and calculated the variance in number of offspring across all individuals in timesteps 50-100.
+we tracked the lifetime total number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period,
+% subset the resulting table to include only the last timestep recorded for each individual,
+and calculated the variance in number of offspring across all individuals in timesteps 50-100.
 All calculations were performed with information recorded in the tree sequence,
-using pyslim \citetext{\url{https://github.com/tskit-dev/pyslim}}.
+using pyslim \citep{pyslim}.
 
@@
 Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{IS}$), and Tajima's $D$, 
-as well as an isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distance analogous to \cite{Rousset1997}'s approach, 
-which we summarized as the correlation coefficient between the logarithm of the spatial distance and the proportion of identical base pairs across pairs of individuals.
-
-Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among sampled chromosomes,
-defined to be the set of distances between adjacent sites that differ between the two chromosomes
-of each diploid. \revpoint{1}{1}
+as well (motivated by \citet{Rousset1997}'s results)
+the correlation coefficient between the logarithm of the spatial distance and the proportion of identical base pairs across pairs of individuals. 
+% as an isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distance analogous to \citet{Rousset1997}'s approach, which we summarized as
+
+Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep[e.g.,][]{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among sampled chromosomes,
+defined to be the set of distances between adjacent sites that differ between the two chromosomes. \revpoint{1}{1}
 The full distribution of lengths of IBS tracts for each pair of chromosomes was first calculated with a custom python function. 
@@
 We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $10^6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.
-
-We then estimated correlation coefficients between spatial distance and each moment of the pairwise IBS tract distribution.
+We then calculated correlation coefficients between spatial distance and each moment of the pairwise IBS tract distribution.
 Because more closely related individuals on average share longer haplotype blocks we expect that spatial distance will be negatively correlated with mean haplotype block length, and that this correlation will be strongest (i.e., most negative) when dispersal is low.
 The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks, and so should reflect recent demographic events \citep{chapman2002effect}.
-For a subset of simulations, we also calculated cumulative distributions for IBS tract lengths across pairs of distant (${}>48$ map units) and nearby (${}<2$ map units) individuals.
-Last, we examined the relationship between allele frequency and the spatial dispersion of an allele by calculating the average distance among individuals carrying each derived allele in a set of simulations representing a range of neighborhood sizes.
+For a subset of simulations, we also calculated cumulative distributions for IBS tract lengths across pairs of distant (more than 48 map units) and nearby (less than 2 map units) individuals.
+Last, we examined the relationship between allele frequency and the spatial dispersion of an allele by calculating the average distance between individuals carrying each derived allele.
 
@@
 To evaluate the performance of these methods we binned simulations by neighborhood size, took a rolling median of inferred $N_{e}$ trajectories across all model fits in a bin for each method and sampling strategy.
-We also examined how varying levels of isolation by distance impacted the variance of $N_{e}$ estimates by calculating the standard deviation of $N_{e}$ from each best-fit model and plotting these against neighborhood size.
+We also examined how varying levels of isolation by distance impacted the variance of $N_{e}$ estimates by calculating the standard deviation of $N_{e}$ from each best-fit model.
+% and plotting these against neighborhood size.
 
@@
 In all simulations, the phenotype of each individual
-is determined by adding independent Gaussian noise with mean zero \revpoint{1}{5}
-to a mean that may depend on spatial position such that the mean of spatially-correlated phenotypes varies by two standard deviations across the landscape.\llabel{phenotypescalingsd}
-We then adjust the geographic pattern of mean phenotype to create four spatially autocorrelated environmental influences on phenotype.
+is determined by drawing from a Gaussian distribution with standard deviation 10 \revpoint{1}{5}
+and a mean that may depend on spatial position.
+In spatially varying models, the mean phenotype differs by two standard deviations across the landscape.
+We then adjust the geographic pattern of mean phenotype to create four types of spatially autocorrelated environmental influences on phenotype.
 In the first simulation of \emph{nonspatial} environments, the mean did not change,
@@
 we increased the mean phenotype from 100 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
-Concretely, the phenotype $p$ for an individual at position $(x,y)$ is $p=100 + 2x / 5$. \revpoint{4}{10}
-Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype 
-for individuals with both $x$ and $y$ coordinates below 20 to 120 (two standard deviations above the rest of the map). 
+Concretely, the mean phenotype $p$ for an individual at position $(x,y)$ is $p=100 + 2x / 5$. \revpoint{4}{10}
+Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype to 120
+for individuals with both $x$ and $y$ coordinates below 20 (two standard deviations above the rest of the map). 
 Finally, in ``\emph{patchy}'' simulations we selected 10 random points on the map
@@
 We also examined $p$ values for systematic inflation \revpoint{1}{6}
-by estimating the expected values from a uniform distribution (because no SNPs were used when generating phenotypes), plotting observed against expected values for all simulations, and summarizing across simulations by finding the mean $\sigma$ value in each region of quantile-quantile space.
+by comparing to the values expected from a uniform distribution
+(because no SNPs were used when generating phenotypes, well-calibrated $p$-values should be uniform).
+
 Results from all analyses were summarized and plotted with the ``ggplot2'' \citep{Wickham2016} and ``cowplot'' \citep{Wilke2019} packages in R \citep{Rcore2018}.
@@
 the generation time, census population size, and variance in offspring number, shown in Figure~\ref{fig:genparams}.
-Because our simulation is paramaterized on an individual level, these population parameters emerge as a property of the interactions among individuals rather than being directly set \llabel{emergentparams}.
+Because our simulation is paramaterized on an individual level, these population parameters emerge as a property of the interactions among individuals rather than being directly set.
+\llabel{emergentparams}
 Variation across runs occurs because,
@@
 The scaling of census sizes in both random-mating and spatial models appears to be related to two consequences of the spatial competition function: the decline of fitness at range edges, which effectively reduces the habitable area by one $\sigma$ around the edge of the map and so results in a smaller habitable area at high $\sigma$ values; and variation in the equilibrium population density given varying competition radii.
-Furthermore, lack of available mates contributes to lower census sizes at small $N_W$
-for the spatial model
-(We calculate $N_W$ for both in the same way,
-even though the number of ``available mates'' is much larger with random mating). \llabel{randomNW}
 Furthermore, census size increases in spatial models as neighborhood size increases from 2 to 10.
@@
 Surprisingly, the spatial and random mating models behave quite differently: 
-while the variance in offspring number increases nearly monotonically under the spatial model, the random mating model actually shows a decline in the variance in offspring number until a neighborhood size $\approx 10$ before it increases and eventually equals what we observe in the spatial case.
+while the variance in offspring number increases nearly monotonically under the spatial model, the random mating model actually shows a decline in the variance in offspring number until a neighborhood size of around 10 before it increases and eventually equals what we observe in the spatial case.
 
 Run times for our model scale approximately linearly with neighborhood size (Figure \ref{fig:runtimes}), with the lowest neighborhood sizes reaching 30N generations in around a day and those with neighborhood size approaching 1,000 requiring up to three weeks of computation.
-As currently implemented running simulations at neighborhood sizes $>1000$ to coalescence is likely impractical, though running these models for more limited timescales and then ``completing'' the simulation by starting a reverse-time simulation from the resulting tree sequence in msprime is possible.
+As currently implemented running simulations at neighborhood sizes more than 1,000 to coalescence is likely impractical, though running these models for more limited timescales and then ``recapitating'' the simulation using reverse-time simulation from the resulting tree sequence in msprime is possible \citep{haller2019treesequence}.
 \llabel{runtimesxn}
@@
 Consistent with findings in island and stepping stone simulations \citep{Stadler2009}, the SFS shows a significant enrichment of intermediate frequency variants in comparison to the nonspatial expectation.
-This bias is most pronounced below neighborhood sizes $\leq 100$ and is exacerbated by midpoint and point sampling of individuals (depicted in Figure~\ref{fig:samplemap}).
+This bias is most pronounced below a neighborhood size of 100 and is exacerbated by midpoint and point sampling of individuals (depicted in Figure~\ref{fig:samplemap}).
 Reflecting this, Tajima's $D$ is quite positive in the same situations (Figure~\ref{fig:sumstats}B).
@@
 
-One of the most commonly used summaries of variation is Tajima's summary of nucleotide divergence, $\theta_{\pi}$, calculated as the mean density of nucleotide differences averaged across pairs of samples.
+One of the most commonly used summaries of variation is Tajima's summary of nucleotide diversity, $\theta_{\pi}$, calculated as the mean density of nucleotide differences averaged across pairs of samples.
 As can be seen in Figure \ref{fig:sumstats}B, $\theta_{\pi}$ in the spatial model is inflated by up to three-fold relative to the random mating model.
@@
 \includegraphics[width=.5\textwidth]{figures/ibs_cumdists_w_rm.pdf}
-\caption{Cumulative distributions for IBS tract lengths per pair of individuals at different geographic distances, across three neighborhood sizes (NS).}
+\caption{
+    Cumulative distributions for IBS tract lengths per pair of individuals at different geographic distances, across three neighborhood sizes (NS).
+    Nearby pairs (red curves) share many more long IBS tracts than do distant pairs (blue curves),
+    except in the random mating model.
+    The distribution of long IBS tracts between nearby individuals are very similar across neighborhood sizes,
+    but distant individuals are much more likely to share long IBS tracts at high neighborhood size
+    than at low neighborhood size.
+}
 \label{fig:ibs}
@@
 
-The patterns observed for correlations of IBS tract lengths with geographic distance are similar to those observed in the more familiar regression of allele frequency measures such as $D_{xy}$ (i.e., ``genetic distance'') or $F_{ST}$ against geographic distance \citep{Rousset1997}.
+The patterns observed for correlations of IBS tract lengths with geographic distance are similar to those observed in the more familiar correlation of allele frequency measures such as $D_{xy}$ (i.e., ``genetic distance'') or $F_{ST}$ against geographic distance \citep{Rousset1997}.
 $D_{xy}$ is positively correlated with the geographic distance between the individuals, and the strength of this correlation declines as dispersal increases (Figure \ref{fig:sumstats}B), as expected \citep{Wright1943,Rousset1997}.
@@
 \includegraphics[width=\textwidth]{figures/distance_by_DAF.pdf}
-\caption{Trends in the distance among allele copies at varying derived allele frequencies and neighborhood sizes (NS).}
+\caption{
+    Spatial spread of rare alleles by neighborhood size (NS):
+    Each plot shows the distribution (across derived alleles and simulations)
+    of average pairwise distance between individuals carrying a focal derived allele
+    and derived allele frequency.
+}
 \label{fig:dafdists}
@@
 Mutations occur in individuals and spread geographically over time.
-Because low frequency alleles generally represent recent mutations \citep{sawyer1977past, griffiths1999ages}, the geographic dispersion of an allele may covary along with its frequency in the population.
+Because low frequency alleles generally represent recent mutations \citep{sawyer1977past, griffiths1999ages}, the geographic spread of an allele may covary along with its frequency in the population.
 To visualize this relationship we calculated the average distance among individuals carrying a focal derived allele across simulations with varying neighborhood sizes, shown in Figure \ref{fig:dafdists}.
@@
 
-Figure \ref{fig:demography}A shows the median inferred population size histories from each method across all simulations, grouped by neighborhood size and sampling strategy.
-In general these methods tend to slightly overestimate ancient population sizes and infer recent population declines when neighborhood sizes are below 20 and sampling is spatially clustered (Figure \ref{fig:demography}A, Figure \ref{fig:demography_supplement}).
+Figure \ref{fig:demography}A shows rolling medians of inferred population size histories from each method across all simulations, grouped by neighborhood size and sampling strategy.
+In general these methods tend to slightly overestimate ancient population sizes and infer recent population declines when neighborhood sizes are below 20 and sampling is spatially clustered.
 The overestimation of ancient population sizes however is relatively minor, averaging around a two-fold inflation at 10,000 generations before present in the worst-affected bins.
@@
 
-However, individual model fits from both methods frequently reflected turbulent demographic histories (Figure \ref{fig:demography_supplement}), with the standard deviation of inferred $N_{e}$ across time points often exceeding the expected $N_{e}$ for both methods (Figure \ref{fig:demography}B).
-That is, despite the constant population sizes in our simulations, both methods tended to infer large fluctuations in population size over time, which could potentially result in incorrect biological interpretations.
-On average the variance of inferred population sizes was elevated at the lowest neighborhood sizes and declines as dispersal increases, with the strongest effects seen in stairwayplot model fits with for clustered sampling and neighborhood sizes less than 20 (Figure \ref{fig:demography}B).
+However, individual simulations were often inferred to have turbulent demographic histories,
+as shown by the individually inferred histories (shown in Figure \ref{fig:demography_supplement}).
+Indeed, the standard deviation of inferred $N_{e}$ across time points (shown in Figure \ref{fig:demography}B) often exceeds the expected $N_{e}$ for both methods.
+That is, despite the nearly constant population sizes in our simulations, both methods tended to infer large fluctuations in population size over time, which could potentially result in incorrect biological interpretations.
+On average the variance of inferred population sizes was elevated at the lowest neighborhood sizes and declines as dispersal increases, with the strongest effects seen in stairwayplot results with clustered sampling and neighborhood sizes less than 20 (Figure \ref{fig:demography}B).
 
@@
 to describe the myriad influence of continuous geography on genetic variation. 
-In particular, we examine how three main types of downstream empirical inference are affected by unmodeled spatial population structure
--- 1) population genetic summary statistics, 2) inference of population size history, and 3) genome-wide association studies (GWAS). 
+In particular, we examine how three main types of downstream empirical inference are affected by unmodeled spatial population structure --
+population genetic summary statistics, inference of population size history, and genome-wide association studies (GWAS). 
 As discussed above, 
@@
 This leads to fewer mutations and recombination events occurring since their last common ancestor, causing a random set of individuals to share longer average IBS tracts and have fewer nucleotide differences.
-For some data summaries, such as Tajima's $D$, Watterson's $\Theta$, or the correlation coefficient between spatial distance and the count of long haplotype blocks, this can result in large differences in estimates between random and point sampling (Figure \ref{fig:sumstats}).
+For some data summaries, such as Tajima's $D$, Watterson's $\theta$, or the correlation coefficient between spatial distance and the count of long haplotype blocks, this can result in large differences in estimates between random and point sampling (Figure \ref{fig:sumstats}).
 Inferring underlying demographic parameters from these summary statistics 
--- unless the nature of the sampling is somehow taken into account 
--- will be subject to bias if sampling is not random across the landscape.
-
-However, we observed the largest sampling effects using ``midpoint'' sampling.
+-- unless the spatial locations of the sampled individuals are somehow taken into account 
+-- will likely be subject to bias.
+
+We observed the largest sampling effects using ``midpoint'' sampling.
 This model is meant to reflect a bias in sampling effort towards the middle of a species' range. 
@@
 Because midpoint sampling provides limited spatial resolution it dramatically reduces the magnitude of observed correlations between spatial and genetic distances. 
-More surprisingly, midpoint sampling also leads to strongly positive Tajima's $D$ and an inflation in the proportion of heterozygous individuals in the sample -- similar to the effect of sampling a single deme in an island model as reported in \citep{Stadler2009}.
+More surprisingly, midpoint sampling also leads to strongly positive Tajima's $D$ and an inflation in the proportion of heterozygous individuals in the sample -- similar to the effect of sampling a single deme in an island model as reported in \citet{Stadler2009}.
 This increase in observed heterozygosity appears to reflect the effects of range edges, which are a fundamental facet of spatial genetic variation. 
@@
 Applied to a cluster of samples, summary statistics based on segregating sites 
-(e.g., Watterson's $\Theta$ and Tajima's $D$), 
+(e.g., Watterson's $\theta$ and Tajima's $D$), 
 heterozygosity, or the distribution of long haplotype blocks, 
@@
 Previous studies have found that population structure and nonrandom sampling can create spurious signals of population bottlenecks when attempting to infer demographic history with microsatellite variation, summary statistics, or runs of homozygosity \citep{Chikhi2010,Stadler2009,Ptak2002,Mazet2015}.
-Here we found that methods that infer detailed population trajectories through time based on the SFS and patterns of LD across the genome are also subject to this bias, with some combinations of dispersal and sampling strategy systematically inferring deep recent population bottlenecks and overestimating ancient $N_{e}$ by a around a factor of 2.
+Here we found that methods that infer detailed population trajectories through time based on the SFS and patterns of LD across the genome are also subject to this bias, with some combinations of dispersal and sampling strategy systematically inferring deep recent population bottlenecks and overestimating ancient $N_{e}$ by around a factor of 2.
 We were surprised to see that both stairwayplot and SMC++ can tolerate relatively strong isolation by distance -- i.e., neighborhood sizes of 20 -- and still perform well when averaging results across multiple simulations.
@@
 \llabel{ibdnesxn}
-Methods directly assessing haplotype structure in phased data (for example, \citet{browning2015accurate}) are thought to provide increased resolution for recent demographic events,
+Methods directly assessing haplotype structure in phased data example, \citep[e.g.,][]{browning2015accurate} are thought to provide increased resolution for recent demographic events,
 but in this case the error we observed was essentially an accurate reflection of underlying genealogies in which terminal branches are anomalously short.
@@
 In our simulations, we observed that this procedure can fail under a variety of circumstances. 
-If dispersal is limited and environmental variation is clustered in space (i.e., corner or patchy distributions in our simulations), PCA positions fail to capture the fine-scale spatial structure required to remove all signals of association.
+If dispersal is limited and environmental variation is clustered in space (i.e., corner or patchy distributions in our simulations), PC positions fail to capture the fine-scale spatial structure required to remove all signals of association.
 Conversely, as dispersal increases, PCA loses power to describe population structure before spatial mixing breaks down the relationship between genotype and the environment.
@@
 This is essentially the approach taken in our ``clinal'' model (though we add normally distributed noise to our phenotypes).
-This approach has recently been taken with polygenic scores for UK Biobank samples in \cite{haworth2019apparent}, finding that scores are correlated with birth location even in this relatively homogenous sample.
+This approach has recently been taken with polygenic scores for UK Biobank samples in \citet{haworth2019apparent}, finding that scores are correlated with birth location even in this relatively homogenous sample.
 \revpoint{2}{8}
@@
 Species such as this have neighborhood size estimates small enough that spatial processes are likely to strongly influence inference.
-These include some human populations such as the Gainj- and Kalam-speaking people of Papua New Guinea, in which the estimated neighborhood sizes in \citep{Rousset1997} range from 40 to 410 depending on the method of estimation.
+These include some human populations such as the Gainj- and Kalam-speaking people of Papua New Guinea, in which the estimated neighborhood sizes in \citet{Rousset1997} range from 40 to 410 depending on the method of estimation.
 \revpoint{3}{7}
 Many more species occur in a middle range of neighborhood sizes between 100 and 1000 -- a range in which spatial processes play a minor role in our analyses under random spatial sampling but are important when sampling of individuals in space is clustered.
-Surprisingly, even some flying insects with huge census population sizes fall in this group, including fruit flies (\textit{D.~melanogaster}) and mosquitoes (\textit{A.
-aegypti}).
-Last, many species likely have neighborhood sizes much larger than we simulated, including modern humans in northeastern Europe \citep{Ringbauer2017}.
+Surprisingly, even some flying insects with huge census population sizes fall in this group, including fruit flies (\textit{D.~melanogaster}) and mosquitoes (\textit{A.~aegypti}).
+Last, many species likely have neighborhood sizes much larger than we simulated, including the recent ancestors of modern humans in northeastern Europe \citep{Ringbauer2017}.
 For these species demographic inference and summary statistics are likely to reflect minimal bias from spatial effects as long as dispersal is truly continuous across the landscape.
@@
 One facet of spatial variation that we did not address in this study is the confounding of dispersal and population density implicit in the definition of Wright's neighborhood size.
-Our simulations were run under constant densities, but \cite{guindon2016demographic} and \citet{Ringbauer2017} \revpoint{3}{8}
+Our simulations were run under constant densities, but \citet{guindon2016demographic} and \citet{Ringbauer2017} \revpoint{3}{8}
 have shown that these parameters are identifiable under some continuous models.
-Similarly, though the scaling effects of dispersal we show in Figure \ref{fig:sumstats} should occur in populations of any total size, other aspects such as the number of segregating sites are also likely affected by the total landscape size (and so total census $N$)\llabel{landscapesizecaveat}.
+Similarly, though the scaling effects of dispersal we show in Figure \ref{fig:sumstats} should occur in populations of any total size, other aspects such as the number of segregating sites are also likely affected by the total landscape size (and so total census size).
+\llabel{landscapesizecaveat}
 Much additional work remains to be done to better understand how these parameters interact to shape genetic variation in continuous space, which we leave to future studies.
@@
 In particular, computations required for mate selection and spatial competition scale approximately with the product of the total census size and the neighborhood size and so increase rapidly for large populations and dispersal distances.
-The reverse-time model of continuous space evolution described by \citet{Barton2010} and implemented by \citet{Kelleher2014} 
-allows exploration of parameter regimes with population and landscape sizes more directly comparable to empirical cases like humans. 
+The reverse-time spatial Lambda--Fleming--Viot model described by \citet{Barton2010} and implemented by \citet{Kelleher2014} 
+allows exploration of larger population and landscape sizes,
+but the precise connection of these models to forward-time demography is not yet clear.
 Alternatively, implementation of parallelized calculations may allow progress with forward-time simulations.
@@
 \section{Data Availability}
-Scripts used for all analyses and figures are available at \url{https://github.com/petrelharp/spaceness}.
+Scripts used for all analyses and figures are available at \url{https://github.com/kern-lab/spaceness}.
 

### 2019-12-05 ms/spaceness.tex

@@
 Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{IS}$), and Tajima's $D$, 
-as well (motivated by \citet{Rousset1997}'s results)
+as well as (motivated by \citet{Rousset1997}'s results)
 the correlation coefficient between the logarithm of the spatial distance and the proportion of identical base pairs across pairs of individuals. 

### 2019-12-05 ms/spaceness.tex
quick pass on the ms

@@
 is determined by drawing from a Gaussian distribution with standard deviation 10 \revpoint{1}{5}
-and a mean that may depend on spatial position.
+and a mean that depends on spatial position.
 In spatially varying models, the mean phenotype differs by two standard deviations across the landscape.
@@
 
-Any simulation of a population of reproducing organisms for a reasonably long amount of time
+Any simulation of a population of reproducing organisms
 requires some kind of control on population sizes,
@@
 is sometimes referred to as the ``Bolker--Pacala model'' \citep{bolker1997using}.
-It is not currently clear how much different choices of vital rates,
+It is not currently clear how much different choices of demographic parameters,
 or of functional forms for the regulation,
@@
 pushing individuals into more marginal habitat and increasing their mortality.
-Many such models should be behave similarly to ours,
+Many such models should behave similarly to ours,
 but others (especially those with local population cycling), may differ dramatically.

### 2019-12-05 ms/spaceness.tex
put the "may" back

@@
 is determined by drawing from a Gaussian distribution with standard deviation 10 \revpoint{1}{5}
-and a mean that depends on spatial position.
+and a mean that may depend on spatial position.
 In spatially varying models, the mean phenotype differs by two standard deviations across the landscape.

### 2019-12-05 ms/spaceness.tex
put caveat on run length back in

@@
 requires some kind of control on population sizes,
-or else the population will either die out or grow very large. \llabel{demog_disco}
+or else the population will either die out or grow very large after a sufficiently long period of time. \llabel{demog_disco}
 The usual choice of population regulation for population genetics --

### 2019-12-05 ms/spaceness.tex
revisions from meeting

@@
     especially when Wright's neighborhood size is less than 100 and sampling is spatially clustered. 
+    Stepping-stone models reproduce some of these effects, but discretizing the landscape introduces artifacts which in some cases are exacerbated at higher resolutions.
         The combination of low dispersal and clustered sampling causes demographic inference from the site frequency spectrum to infer more turbulent demographic histories, but averaged results across multiple simulations were surprisingly robust to isolation by distance.
@@
 can depend strongly on the chosen discretization,
-and some artifacts of discretization do not seem to dissappear in the limit of a fine grid.
+and some artifacts of discretization seem to become stronger in the limit of a fine grid.
 For many summary statistics, finer discretizations (we used a $50 \times 50$ grid)
@@
 
-In general we find many of the qualitative trends are similar among continuous and stepping-stone models and that, in most cases, statistics from stepping-stone models approach the continuous model as the resolution of the grid increases.
-For example, $\theta_\pi$ is inflated at low neighborhood sizes (i.e.
-low $m$), and the extent of the inflation increases to approach the continuous case as the resolution of the landscape increases.
-Similar patterns are observed for $F_{is}$ and observed heterozygosity.
-However, $\theta_{W}$ behaves differently, with increased grid resolution leading to lower values.
-This in turn drives an even more positive Tajima's $D$ in grid simulations at small neighborhood sizes.
+In general we find many of the qualitative trends are similar among continuous and stepping-stone models and that, at low neighborhood sizes, many  (but not all) statistics from stepping-stone models approach the continuous model as the resolution of the grid increases.
+For example, $\theta_\pi$ is inflated at low neighborhood sizes (i.e., low $m$), and the extent of the inflation increases to approach the continuous case as the resolution of the landscape increases.
+Similar patterns are observed for $F_{IS}$ and observed heterozygosity.
+However, $\theta_{W}$ behaves differently, showing a non-monotonic relationship with grid resolution.
+This results in an increasingly positive Tajima's $D$ in grid simulations at small neighborhood sizes, to a much greater extent than seen in a continuous model.
+In contrast to $\theta_\pi$, increasing the resolution of the grid causes Tajima's $D$ to deviate \emph{more} from what is seen in the continuous model.
 
@@
 If we simulate a coarse grid with relatively large populations in each deme, we cannot accurately capture the dynamics of small neighborhood sizes because mating within each deme remains random regardless of the migration rate connecting demes.
-This likely explains the trends in $\pi$, observed heterozygosity, and $F_{is}$.
+This likely explains the trends in $\theta_{\pi}$, observed heterozygosity, and $F_{IS}$.
 However increasing the number of demes while holding the total number of individuals constant results in small within-deme populations for which even the minimum sample size of 1 approaches the local $N_e$ (Table \ref{table:grid_params}).
 This results in an excess of short terminal branches in the coalescent tree, which decreases the total branch length and leads to fewer segregating sites, deflated $\theta_W$, and inflated Tajima's $D$.
-Overall then our continuous model reproduces important features of spatial structure approximated by reverse-time stepping-stone models at moderate neighborhood sizes while avoiding some artifacts caused binning the landscape into discrete demes.
+Overall, the stepping-stone model reproduces important features of spatial structure in our continuous space model, such as a decline in $\theta_\pi$ and correlations between spatial and genetic distance with increasing migration, but introduces artifacts caused by binning the landscape into discrete demes.
 

### 2019-12-05 ms/spaceness.tex
final edits(?)

@@
 have focused on a small number of genetic loci
-\citep[e.g.,][]{slatkin1989comparison,barton2002neutral,robledoarnuncio2010isolation,rossine2014espao},
+\citep[e.g.,][]{slatkin1989comparison,barton2002neutral,robledoarnuncio2010isolation,rossine2014espao,jackson2014landscape},
 which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. 
@@
 but others (especially those with local population cycling), may differ dramatically.
+
+Population genetic simulations often use grids of discrete demes, which are assumed to approximate continuous space. 
+However, there are theoretical reasons to expect that increasingly fine grids of discrete demes do not approach the continuous model 
+\citep{barton2002neutral}.
+If continuous space can be approximated by a limit of discrete models, this should be true regardless of the precise details of the discrete model. 
+Although we carefully chose parameters to match our continuous models, we found that some aspects of genetic variation
+diverged from the continuous case as the discretization got finer. 
+This suggests that these models do not converge in the limit. 
+However, many populations may indeed be well-modeled as a series of discrete, 
+randomly-mating demes if, for example, suitable habitats are 
+patchily distributed across the landscape. 
 There is a clear need for greater exploration of 

### 2019-12-05 ms/spaceness.tex
actual final edits

@@
 The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks, and so should reflect recent demographic events \citep{chapman2002effect}.
-For a subset of simulations, we also calculated cumulative distributions for IBS tract lengths across pairs of distant (more than 48 map units) and nearby (less than 2 map units) individuals.
+For a subset of simulations, we also calculated cumulative distributions for IBS tract lengths across pairs of distant (more than 40 map units) and nearby (less than 10 map units) individuals.
 Last, we examined the relationship between allele frequency and the spatial dispersion of an allele by calculating the average distance between individuals carrying each derived allele.

### 2019-12-06 ms/spaceness.tex
bibtex fixes

@@
 Despite a long history of analytical work describing the genetics of populations distributed across continuous geography
-(e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation,Wilkins2002linearcoalescent,Wilkins2004separation}), \revpoint{AE}{1}
+(e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation,Wilkins2002linearcoalescent,wilkins2004separation}), \revpoint{AE}{1}
 much modern work still describes geographic structure as a set of discrete populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2019populations} or as an average over such discrete models \citep{Petkova2015,al2019estimating}.

### 2020-01-31 ms/spaceness.tex
reviews

@@
 \input{review-response-commands}
-\newcommand{\responsefile}{review-responses.tex}  % name of the review reponses file
+\newcommand{\responsefile}{second-review-responses.tex}  % name of the review reponses file
 
@@
 Despite a long history of analytical work describing the genetics of populations distributed across continuous geography
-(e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation,Wilkins2002linearcoalescent,wilkins2004separation}), \revpoint{AE}{1}
+(e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation,Wilkins2002linearcoalescent,wilkins2004separation}), 
 much modern work still describes geographic structure as a set of discrete populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2019populations} or as an average over such discrete models \citep{Petkova2015,al2019estimating}.
-\revpoint{4}{3}
 For this reason, most population genetics statistics are interpreted with reference to discrete, well-mixed populations, and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000} with methods that assume these are randomly mating units.
@@
 which shifts reproductive output towards low-density regions,
-and produces total census sizes that fluctuate around an equilibrium. \revpoint{3}{2}
+and produces total census sizes that fluctuate around an equilibrium. 
 This also prevents the population clumping seen by \citet{Felsenstein1975}
-(Supplemental \autoref{fig:wm_v_dd})). \revpoint{3}{1}
+(Supplemental \autoref{fig:wm_v_dd})). 
 Such models have been used extensively in ecological modeling
 \citep{durrett1994importance,bolker1997using,law2003population,fournier2004microscopic,champer2019suppression}
-but rarely in population genetics, \revpoint{AE}{2}
+but rarely in population genetics, 
 where to our knowledge implementations of continuous space models
@@
 Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
-in both the $x$ and $y$ coordinates. \revpoint{4}{8}
+in both the $x$ and $y$ coordinates. 
 Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
@@
 
-To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, and density-dependent fitness,  \revpoint{3}{6}
+To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, and density-dependent fitness,  
 we also implemented simulations identical to those above except that mates are selected uniformly at random from the population,
@@
 
-We ran 400 simulations for the spatial and random-mating models on a square landscape of width $W=50$  \revpoint{AE}{3}
+We ran 400 simulations for the spatial and random-mating models on a square landscape of width $W=50$  
 with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), 
@@
 we divided the mutation rate of $10^{-8}$ mutations per site per generation
-by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time. \revpoint{2}{5}
+by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time. 
 We verify that this procedure produces the same site frequency spectrum as applying mutations directly in SLiM in Figure \ref{fig:slim_v_msp_sfs},
@@
 
-Our demographic model includes parameters that control population density ($K$), \revpoint{1}{2}
+Our demographic model includes parameters that control population density ($K$), 
 mean life span ($L$), and dispersal distance ($\sigma$).
@@
 and used mean population density ($\rho$, census size divided by total area)
-to compute neighborhood size as $N_W = 4 \pi \rho \sigma^2$. \revpoint{3}{6}
+to compute neighborhood size as $N_W = 4 \pi \rho \sigma^2$. 
 To estimate generation times, we stored ages of the parents of every new individual born across 200 timesteps, 
@@
 Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep[e.g.,][]{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among sampled chromosomes,
-defined to be the set of distances between adjacent sites that differ between the two chromosomes. \revpoint{1}{1}
+defined to be the set of distances between adjacent sites that differ between the two chromosomes. 
 The full distribution of lengths of IBS tracts for each pair of chromosomes was first calculated with a custom python function. 
@@
 \subsection{Demographic Inference}
-\revpoint{1}{4}
 
@@
 In all simulations, the phenotype of each individual
-is determined by drawing from a Gaussian distribution with standard deviation 10 \revpoint{1}{5}
+is determined by drawing from a Gaussian distribution with standard deviation 10 
 and a mean that may depend on spatial position.
@@
 we increased the mean phenotype from 100 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
-Concretely, the mean phenotype $p$ for an individual at position $(x,y)$ is $p=100 + 2x / 5$. \revpoint{4}{10}
+Concretely, the mean phenotype $p$ for an individual at position $(x,y)$ is $p=100 + 2x / 5$. 
 Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype to 120
@@
 with $p$-value below $0.05$ after adjusting for an expected false positive rate of less than 5\% \citep{benjamini2001control}. 
-We also examined $p$ values for systematic inflation \revpoint{1}{6}
+We also examined $p$ values for systematic inflation 
 by comparing to the values expected from a uniform distribution
@@
 Variation across runs occurs because,
-even though the parameters $K$ and $L$ that control population density and mean lifetime respectively \llabel{ll:K_more} \revpoint{1}{11}
+even though the parameters $K$ and $L$ that control population density and mean lifetime respectively \llabel{ll:K_more} 
 were the same in all simulations,
@@
 Moreover, as we will show, spatial sampling strategies can affect summaries of genetic variation at least as strongly as the underlying population dynamics.
-\revpoint{1}{9}
 
@@
 This approach has recently been taken with polygenic scores for UK Biobank samples in \citet{haworth2019apparent}, finding that scores are correlated with birth location even in this relatively homogenous sample.
-\revpoint{2}{8}
 Of course, it is possible that genotypes indirectly affect individual locations by adjusting organismal fitness and thus habitat selection across spatially varying environments, but we believe that this hypothesis should be tested against a null of stratification bias inflation rather than accepted as true based on GWAS results.
@@
 These include some human populations such as the Gainj- and Kalam-speaking people of Papua New Guinea, in which the estimated neighborhood sizes in \citet{Rousset1997} range from 40 to 410 depending on the method of estimation.
-\revpoint{3}{7}
 Many more species occur in a middle range of neighborhood sizes between 100 and 1000 -- a range in which spatial processes play a minor role in our analyses under random spatial sampling but are important when sampling of individuals in space is clustered.
@@
 One facet of spatial variation that we did not address in this study is the confounding of dispersal and population density implicit in the definition of Wright's neighborhood size.
-Our simulations were run under constant densities, but \citet{guindon2016demographic} and \citet{Ringbauer2017} \revpoint{3}{8}
+Our simulations were run under constant densities, but \citet{guindon2016demographic} and \citet{Ringbauer2017}
 have shown that these parameters are identifiable under some continuous models.
@@
 CJB and ADK were supported by NIH award R01GM117241.
-\revpoint{AE}{4}
 

### 2020-01-31 ms/spaceness.tex
added line ref

@@
 -- as neighborhood size increases, the total proportion of variance explained by the first 10 PC axes declines from roughly 10\% to 4\% (Figure \ref{fig:gwas}B). 
-Essentially, PCA seems unable to effectively summarize the weak population structure present in large-neighborhood simulations, but these populations continue to have enough spatial structure to create significant correlations between genotypes and the environment.
+Essentially, PCA seems unable to effectively summarize the weak population structure present in large-neighborhood simulations, \revpoint{2}{2}
+but these populations continue to have enough spatial structure to create significant correlations between genotypes and the environment.
 A similar process can also be seen in the corner phenotype distribution, in which the count of significant SNPs initially declines as neighborhood size increases and then increases at approximately the point at which the proportion of variance explained by PCA approaches its minimum.

### 2020-01-31 ms/second-review-responses.tex
actually add responses

%%%%%%
%%
%%  Don't reorder the reviewer points; that'll mess up the automatic referencing!
%%
%%%%%

\begin{minipage}[b]{2.5in}
  Resubmission Cover Letter \\
    {\it Genetics}
\end{minipage}
\hfill
\begin{minipage}[b]{2.5in}
    C.~J.~Battey, \\
    Peter Ralph, \\
    \emph{and} Andrew Kern \\
  \today
\end{minipage}

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are writing to submit another revised version of our manuscript, 
``Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data''.

\vspace{5em}

\noindent \hspace{4em}
\begin{minipage}{3in}
\noindent
{\bf Sincerely,}

\vskip 2em

{\bf 
C.~J.~Battey, Peter Ralph, and Andrew Kern
}\\
\end{minipage}

\vskip 4em

\pagebreak
\setcounter{page}{1}

%%%%%%%%%%%%%%
\reviewersection{AE}

The point of R3 on the bimodality in the realized dispersal distance in your model is interesting. You can make this more explicit in the paper and describe how there are realistic biological scenarios that motivate such a model (many/most plants) and demarcate how well you think you can generalize to others (the new appendix already helps in that regard). It strikes me your model has relevance for patrilocal/matrilocal systems where the paternal/maternal parent-offspring distances differ - though your model is hermaphroditic so it's an imperfect analogy. Perhaps, to explore more typical animal mating patterns, you can quickly check the impact for one or two outcomes of changing the simulation such that offspring are dispersed around the midpoint of the parental origins, as this would remove the bimodality. That said, please be sure we respect the work done is extensive and don't mean to invite/require a complete re-working. The core tension is between ensuring your results are generalizable if that is what you wish to claim; versus specific to a particular model that is biologically motivated. It's particularly key for the material on GWAS, where you are explicitly trying to address human GWAS results: those sections will be read and taken at face value by a large community even if this simulation scheme introduces some discrepancies from what might more realistically emerge. So at a minimum, some caveats directly in the GWAS discussion section would help.

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
Only minor changes were needed, and they were made. This study neatly summarises many biases arising from unmodelled continuous space (pertinent for real studies), and I would not extend it by further analyses. When polishing the final version, the authors can make some of their messages stronger. For example, Figure 7 and the accompanying text may give the impression that we are doing pretty well with smc++. It is only when panel B is seen one can appreciate the degree of uncertainty. So, in a way, Figure S7 is more telling as to how much we can trust our results in real analysis. After all, in reality, as authors mention, we only have one sample set, not replicates.
\end{quote}

Thanks!!!

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
The authors have done a commendable job of responding to the reviewers' comments. The extended discussion and additional supplemental figures helps with the clarity of the manuscript. The results for the comparison with the discrete stepping-stone model were useful - though see below. The authors have also provided sufficient detail justifying their choice of model parameters.
\end{quote}

\begin{quote}
I don't think I have any critical remaining issues or suggestions, but I did have a couple of thoughts. The authors might just want to think about them.
\end{quote}

\begin{point}{}
I appreciate the appendix comparing summary statistics in the continuous and stepping-stone models, although I must confess I am now even more uncertain than I was before about which models are more appropriate. I get that the coalescent models make assumptions that are not directly interpretable in terms of observable parameters (although are "interactions between geographically distant parts of the species range" really biologically unrealistic?). This forward model is more interpretable, in that sense, but does it actually produce more realistic data? I guess that's a question for future work, but that question was one of the main things I came away from this paper with.
\end{point}

\reply{
}

\begin{point}{\revref}
 "PCA is unable to capture". Is this a sample size effect? i.e. the precision of the estimates of the PCs depends on sample size, so it's possible that you just need a larger sample to estimate the PCs when you have weak structure. Or is it something else?
\end{point}

\reply{
}

\begin{point}{Line 685.}
 Is Khera et al (2018) really the right reference for the idea of polygenic scores? They have been around for a lot longer than that.
\end{point}

\reply{
}

\begin{point}{Line 686. }
 I don't know that the results of Berg and Sohail et al. are indicating a fundamental issue with PC correction, rather than a problem with meta-analysis.
\end{point}

\reply{
}

\begin{point}{Line 689-92. }
 Is the result of Martin et al. really due to population stratification? There seem to be a number of factors affecting transferability of PRS (LD differences, effect size heterogeneity, differences in causal variants etc..), but I don't know that anyone thinks residual structure is the major one, particularly for eg. UKB summary stats.
\end{point}

\reply{
}

\begin{point}{Line 752 "population genetics flavor"}
I don't understand what this means?
\end{point}

\reply{
}

\begin{point}{Line 800}
"process-driven descriptions of ancestry and/or more generalized unsupervised methods". Aren't these two things opposite? That's ok, but it sounds a bit like you have no idea what would help...
\end{point}

\reply{
}

\begin{point}{}
 Fix citation: Peter L. Ralph and Ashander ??? on pg. 5, line 216
\end{point}

\reply{
}

\begin{point}{}
'there' written twice on pg. 27, line 1033
\end{point}

\reply{
}

%%%%%%%%%%%%%%
\reviewersection{3}

\begin{quote}
I am fairly pleased overall with the revised version of this article. I commend the authors for the substantial amount of work conducted. This contribution now convincingly demonstrates how important it is to properly account for population structure when individuals are distributed along a spatial continuum.
\end{quote}

\begin{point}{}
A couple of minor points about the forward model are still puzzling me though. First of all, in one dimension, the variance of $L(o)$, the random variable corresponding to the offspring's spatial position, given $l(p)$, the parent's position, is equal to $E((L(o)-l(p))^2):=\sigma^2$. Hence, $\sigma$ is here the square root of the expected squared Euclidean distance between parents and offspring. In general, assuming isotropic migrations, $\sigma$ is the square root of the expected squared Euclidean distance taken along any dimension. It seems to me that this definition is distinct from the mean parent-offspring distance, as stated by the authors.
\end{point}

\reply{
}

\begin{point}{}
More importantly, I am still a bit confused about the reproduction scheme. My understanding is that, once mating takes place, the two parents involved (noted as P(1) and P(2)) each produce a Poisson number of offspring. The location of each offspring is determined by a Gaussian distribution of covariance matrix $\sigma^2 I$, centered on the location of the parent that ``produced' this offspring. Let $O(1,i)$ and $O(2,j)$ be the i-th and j-th offspring produced by $P(1)$ and $P(2)$ respectively. Now, if I am not mistaken, half of the genetic material of O(2,j) comes from P(1) (the other half coming from P(2)) while O(1,i) receives one chromosome from P(2) (and another one from its ``true' parent, i.e., P(1)). Hence, when it comes to the parent-offspring distance as measured from their genetic material, half of the chromosomes have a ``small' dispersal distance, equal to that of the parent-offspring distance, while the other half have a ``large' distance roughly equal to the parent-offspring distance plus the distance between mates.
\end{point}

\reply{
}

\begin{point}{}
It is important that the authors clarify this last point because most results presented in this study rely on Wright's neighborhood size, which derives from a model where dispersal of individuals and chromosomes follow well-defined spatial dynamics. A sensible comparison of neighborhood sizes requires that dispersal (along with effective population density) is quantified in a coherent way across different models.
\end{point}

\reply{
}

\begin{point}{}
Additionally, the neighborhood size as defined by the authors (see line 211) is a function of the census density rather than the effective density of individuals. I am therefore not fully convinced that the range of values for this parameter is comparable to that given in Table 1. Furthermore, the neighborhood size estimates obtained for Bebicium vittatum derive from the analysis of a one-dimensional habitat. The product $\sigma^2 \rho$ is thus expressed in this particular case as a number of individuals *per unit of space* (in two dimensions, this product is expressed as a number of individuals). Hence, the unit of space matters a lot here, prohibiting the direct comparison with other figures in the same table or with the simulations conducted in this study.
\end{point}

\reply{
}

\begin{point}{}
Finally, Leblois, Estoup and Streiff (2006) Molecular Ecology conducted a large simulation studies with a focus very close to that of the present study. In particular, they investigated the impact of sampling on the inference of summary statistics commonly used in population genetics. It would probably be relevant to compare some of the results presented in Leblois et al. to that put forward here.
\end{point}

\reply{
}

%%%%%%%%%%%%%%
\reviewersection{4}

\begin{quote}
The authors have thoroughly addressed nearly all of my comments in their timely manuscript, and I am satisfied with their responses.
\end{quote}

\begin{point}{}
Regarding Table S1, I would still suggest adding a column about interpretation for those variables and evaluated with respect to the simulated results.
\end{point}

\reply{
}

### 2020-02-10 ms/second-review-responses.tex
draft response 2

@@
 
+\reply{Thank you for your comments.
+We have responded to each reviewer below.
+In response to reviewer 2 we have added several caveats to the GWAS section noting other sources of error such as allele frequency and LD variation among populations.
+Reviewer 3 raised two good points on our definition of neighborhood size and the interpetation of $\sigma$ -- first, that $\sigma$ refers to one- rather than two-dimensional distances between parent and offspring (this is consistent with previous theoretical work, and we have corrected our definition in the text), and second, that the use of separate draws for mate selection and dispersal means the total distance traveled by a locus along a pedigree link is greater for "paternal" than "maternal" material (taking "maternal" to mean the parent from which dispersal is calculated).
+Because our simulated genome is autosomal and individuals hermaphroditic and non-selfing, a locus will experience a mix of "maternal" and "paternal" transmission events along its genealogy.
+So we see this as a scaling issue rather than one of a "bimodal" dispersal kernal per se -- the expected distance traveled by a locus along a pedigree link is $\sqrt{3/2} \sigma$, rather than $\sqrt{2} \sigma$.
+However we note that in terms of summary statistics the effects of scaling $\sigma$ by $sqrt{3/2}$ are weaker than those generated by factors like changing landscape resolution in lattice vs.
+continuous models (e.g.
+Appendix 1), so we don't think this significantly changes our interpretation of model outputs.
+We have instead expanded some text in the sections on empirical data and further directions, noting that the relationship between neighborhood size and any given summary statistic depends on the on assumed definitions of population density and details of the breeding system.
+}
+
 %%%%%%%%%%%%%%
@@
 
-Thanks!!!
+Thank you for your comments.
+While we agree that our analysis found problems with demographic inference under isolation by distance, since this manifests as higher variance in predictions rather than a straightforward bias as we expected (i.e.
+a reduction in recent inferred Ne), we have chosen to keep the somewhat more nuanced presentation of results in that section used in a previous version.
 
@@
 \reply{
+This is a good point.
+We have added a sentence to the discussion noting that while our continuous model has some advantages over discrete models, the best model for any empirical system will inevitably depend on the life history of the organism in question.
+\llname{bestmodelcaveat}
 }
@@
 \reply{
+We have added the caveat "given the sample sizes tested" to the above statement. \llname{samplesizenote}
 }
@@
 \reply{
+We think Berg et al.
+shows that population structure leaves signals in GWAS that resemble those of polygenic selection, and given that this was most apparent in GIANT data employing PCA corrections, indicates a problem with the GWAS that make up the GIANT study at least as much as any issues with meta-analysis.
 }
@@
 
-\reply{
+\reply{This is a good point -- differences in allele frequency and LD structure certainly have significant effects on PRS transferability (at least when the GWAS is based on SNP assays rather than whole-genome sequencing.
+We think the population structure issue is important to point to here because it will not be solved by simply generating more detailed genomic data from existing samples, but have added a caveat noting these other important potential sources of error at the head of that sentence.
+\llname{ldcaveats}
 }
@@
 \reply{
+Peter(???)
 }
@@
 \reply{
+They certainly entail research in somewhat different directions, but you are correct that we don't know which is the best way to go. 
 }
@@
 \reply{
+corrected - thanks. 
 }
@@
 \reply{
+thanks for catching this, corrected. 
 }
@@
 \reply{
+Good point here.
+We have corrected this to "$\sigma$ is the mean parent--offspring distance along each of the x and y axes" and now direct readers to \citep{Rousset1997} for discussion of how these parameters are defined in one- and two-dimensional habitats.
+The mean parent-offspring distance in two dimensions is $\sigma\sqrt{2}$, but we follow previous work in defining neighborhood size as $4 \pi \rho \sigma^2$ (i.e.
+without the factor of $\sqrt{2}$; see \citep{Rousset1997} for details).
+\llname{axialdist}
 }
@@
 \reply{
+Another important point here.
+In our simulation the average distance traveled by a locus along any link in the pedigree will be $ \sigma \sqrt{2} $ relative to the "maternal" parent, and $2 \sqrt{2} \sigma$ relative to the "paternal" parent, so the expected distance averaging over both parents is $\sqrt{3/2} \sigma$.
+We have ?????
+
 }
@@
 A sensible comparison of neighborhood sizes requires that dispersal (along with effective population density) is quantified in a coherent way across different models.
-\end{point}
 
-\reply{
-}
-
-\begin{point}{}
 Additionally, the neighborhood size as defined by the authors (see line 211) is a function of the census density rather than the effective density of individuals.
@@
 \reply{
+Good point, we have removed Bebicium vittatum from table 1, and added a sentence to the empirical data section noting that comparisons with our results depend on details of the reproduction model and population density and should thus be taken with a grain of salt \llname{rhocaveat}.
+Indeed we think our analysis of the stepping-stone model emphasizes this point quite well, showing how differences between classes of model (continuous- vs lattice) and even within lattice models (i.e.
+landscape resolution) can lead to qualitatively different results at a given neighborhood size.
+However our main conclusions are given in terms of orders of magnitude -- roughly, less than 100, 100-1000, and over 1000 -- and we believe the various models used to derive the values in table 1 are close enough for this interpretation to remain a useful yardstick.
 }
@@
 \reply{
+Thank you for bringing this study to our attention. We have added it to the citations in the demographic inference section. 
 }
@@
 \reply{
+Thank you for your comments.
+We chose to evaluate a few key statistics in detail in the main text and then provide an interpretation rooted in how the marginal genealogies of the tree sequence interact with space, and hope that this will assist readers when examining other statistics shown.
 }

### 2020-02-10 ms/spaceness.tex
draft response 2

--- 2020-01-31 ms/spaceness.tex
@@
 A key early observation by \citet{Wright1946} is that this balance is often nicely summarized by the ``neighborhood size'',
-defined to be $N_W = 4 \pi \rho \sigma^2$, where $\sigma$ is the mean parent--offspring distance and $\rho$ is population density. 
+defined to be $N_W = 4 \pi \rho \sigma^2$, where $\sigma$ is the mean parent--offspring distance along each of the x and y axes and $\rho$ is population density (see \citep{Rousset1997} for further discussion of parameter defiitions in one- and two-dimensional habitats) \llabel{axialdist}.
 This can be thought of as proportional to the average number of potential mates for an individual (those within distance $2 \sigma)$,
@@
 to modulate the spatial scale of interactions at all three stages by adjusting the standard deviation of the corresponding Gaussian functions. 
-As in previous work \citep{Wright1943,Ringbauer2017}, $\sigma$ is equal to the mean parent-offspring distance.
+As in previous work \citep{Wright1943,Ringbauer2017}, $\sigma$ is equal to the mean parent-offspring distance along each of the x and y axes.
 
@@
 -- as neighborhood size increases, the total proportion of variance explained by the first 10 PC axes declines from roughly 10\% to 4\% (Figure \ref{fig:gwas}B). 
-Essentially, PCA seems unable to effectively summarize the weak population structure present in large-neighborhood simulations, but these populations continue to have enough spatial structure to create significant correlations between genotypes and the environment.
+Essentially, PCA seems unable to effectively summarize the weak population structure present in large-neighborhood simulations given the sample sizes we tested, \revpoint{2}{2} \llabel{samplesizenote}
+but these populations continue to have enough spatial structure to create significant correlations between genotypes and the environment.
 A similar process can also be seen in the corner phenotype distribution, in which the count of significant SNPs initially declines as neighborhood size increases and then increases at approximately the point at which the proportion of variance explained by PCA approaches its minimum.
@@
 
-Previous studies have found that population structure and nonrandom sampling can create spurious signals of population bottlenecks when attempting to infer demographic history with microsatellite variation, summary statistics, or runs of homozygosity \citep{Chikhi2010,Stadler2009,Ptak2002,Mazet2015}.
+Previous studies have found that population structure and nonrandom sampling can create spurious signals of population bottlenecks when attempting to infer demographic history with microsatellite variation, summary statistics, or runs of homozygosity \citep{Chikhi2010,Stadler2009,Ptak2002,Mazet2015,leblois2006genetics}.
 Here we found that methods that infer detailed population trajectories through time based on the SFS and patterns of LD across the genome are also subject to this bias, with some combinations of dispersal and sampling strategy systematically inferring deep recent population bottlenecks and overestimating ancient $N_{e}$ by around a factor of 2.
@@
 Indeed as recently identified in studies of genotype associations for human height in Europe \citep{Berg2018,Sohail2018}, PC regression GWAS in modern human populations do include residual signal of population structure in large-scale analyses of polygenic traits.
-When attempting to make predictions across populations with different environmental exposures, polygenic risk scores affected by population structure can be expected to offer low predictive power, as was shown in a recent study finding lower performance outside European populations \citep{Martin2019}.
+In addition to error associated with varying patterns of linkage disequilibrium and allele frequency among populations, the confounding of environmental and genetic effects on phenotypes introduced by population structure is expected to lead to low predictive power when polygenic scores are generated for populations outside the original GWAS cohort, \llabel{ldcaveats}
+%When attempting to make predictions across populations with different environmental exposures, polygenic risk scores affected by population structure can be expected to offer low predictive power, 
+as was shown in a recent study finding lower polygenic score predictive power outside European populations \citep{Martin2019}.
 
@@
   \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{POPRES data\\NE Europe} & 1,342 - 5,425 & Genetic & \citep{Ringbauer2017} \\ 
-  \makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 240,000 & Survey & \citep{Rousset1997}\\ 
-  \makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 360,000 & Genetic & \citep{Rousset1997}\\ 
+  %\makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 240,000 & Survey & \citep{Rousset1997}\\ 
+  %\makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 360,000 & Genetic & \citep{Rousset1997}\\ 
    \hline
@@
 others (e.g., number of segregating sites)
-are more strongly determined by global $N_e$ or by the ratio of the two.
+are more strongly determined by global $N_e$ or by the ratio of the two, and definitions of "population density" in genetic versus ecological studies may lead to varying estimates of neighborhood size for a given species \llabel{rhocaveat}.
 % These values should be compared to our simulation results with some caution, as though we expect neighborhood size to have similar effects across species of varying global $N_e$ \citep{Wright1946}, in our study we evaluated only a relatively small population of $\approx 10,000$.
@@
 However, from the available data we find that neighborhood sizes in the range we simulated are fairly common across a range of taxa.
-At the extreme low end of empirical neighborhood size estimates we see some flowering plants, large mammals, and colonial insects like ants.
+At the extreme low end of empirical neighborhood size estimates we see some flowering plants, large mammals, and colonial insects like ants with neighborhood sizes less than roughly 100.
 Species such as this have neighborhood size estimates small enough that spatial processes are likely to strongly influence inference.
@@
 \llabel{landscapesizecaveat}
-Much additional work remains to be done to better understand how these parameters interact to shape genetic variation in continuous space, which we leave to future studies.
+Indeed, our finding that stepping-stone models reproduce only some of the effects on genetic variation seen in a continuous-space model (Figure \ref{fig:grid_stats}) suggests that qualitatively similar models can produce different results dependent on the specific parameterizations used.
+While we believe our continuous model is a more appropriate depiction of evolution in a continuous landscape than lattice models, it is likely that some populations and breeding systems do more closely resemble a series of interconnected random-mating populations.
+As with all population models, the best approximation for any empirical system will depend on the natural history of the species in question.
+\llabel{bestmodelcaveat}
+Much additional work remains to be done to better understand how life history and range size interact to shape genetic variation in continuous space, which we leave to future studies.
 

### 2020-02-10 ms/second-review-responses.tex
expanded rev 3 responses

@@
 Hence, when it comes to the parent-offspring distance as measured from their genetic material, half of the chromosomes have a ``small' dispersal distance, equal to that of the parent-offspring distance, while the other half have a ``large' distance roughly equal to the parent-offspring distance plus the distance between mates.
-\end{point}
 
-\reply{
-Another important point here.
-In our simulation the average distance traveled by a locus along any link in the pedigree will be $ \sigma \sqrt{2} $ relative to the "maternal" parent, and $2 \sqrt{2} \sigma$ relative to the "paternal" parent, so the expected distance averaging over both parents is $\sqrt{3/2} \sigma$.
-We have ?????
-
-}
-
-\begin{point}{}
 It is important that the authors clarify this last point because most results presented in this study rely on Wright's neighborhood size, which derives from a model where dispersal of individuals and chromosomes follow well-defined spatial dynamics.
@@
 I am therefore not fully convinced that the range of values for this parameter is comparable to that given in Table 1.
+\end{point}
+
+\reply{
+Another important point set of points on relating neighborhood sizes from different models here.
+In our simulation the average x axis displacement of a locus along any link in the pedigree is $\sigma$ for the "maternal" material (i.e.
+deriving from the parent from which dispersal is calculated), but for the "paternal" parent it is the sum of two independent normal draws each with variance $\sigma^2$.
+Averaging over a random series of maternal and paternal transmissions (because our simulated genome is autosomal), variance in axial position across pedigree links is then $(3/2)\sigma^2$ and the expected displacement $\sqrt{3/2}\sigma$.
+So from the perspective of genes sampled at the present time our neighborhood size calculations are lower than those in models like those described in \cite{Rousset1997} by a factor of (3/2) because of differences in the breeding model.
+
+This inflation in neighborhood size is then partially offset by our use of census rather than "effective" population size when calculating population density, because $N_e$ of our spatial model estimated from $\pi$ in random-mating spatial models is roughly (2/3) the census size.
+We chose this method for density and sigma calculations because we sought to parameterize our model with values that could in theory be observed in real populations.
+We also find the concept of "effective population size" in this context even more confusing than usual -- as we show, spatial processes in both evolution and sampling radically shift the distribution of coalescence times and the variance in offspring numbers in ways that mimic the effects of classic $N_{e}$ scaling in random-mating models, but are driven by fundamentally different processes.
+
+Thus relative to models like \cite{Rousset1997} our definition of \sigma is probably too low and \rho probably too high; while relative to neighborhood sizes calculated from mark-recapture studies measuring parent-offspring distances our figures should be consistent.
+We have expanded the caveats at the head of the discussion paragraph interpreting empirical studies noting that definitions of \rho in ecological vs genetic studies mean results should be interpreted with caution \llname{rhocaveat}; however, as we mainly interpret these figures in terms of order-of-magnitude differences (roughly less than 100, 100-1000, and over 1000) we think our results are general enough to be robust to these modeling differences while still giving a useful yardstick.
+Indeed for many of the empirical studies the range of NS values given is larger than the differences we expect from the parameter definitions described above.
+We also think that our study of lattice simulations in appendix 1 demonstrates nicely how qualitatively similar simulations can lead to quite different patterns of genetic diversity at a given neighborhood size as a result of differences in the underlying breeding model.
+
+\cjb{CJ: let's meet and talk again and talk about what edits to make here. I'm not sure what the best way to go is.}
+
+}
+
+\begin{point}{}
 Furthermore, the neighborhood size estimates obtained for Bebicium vittatum derive from the analysis of a one-dimensional habitat.
@@
 \reply{
-Good point, we have removed Bebicium vittatum from table 1, and added a sentence to the empirical data section noting that comparisons with our results depend on details of the reproduction model and population density and should thus be taken with a grain of salt \llname{rhocaveat}.
-Indeed we think our analysis of the stepping-stone model emphasizes this point quite well, showing how differences between classes of model (continuous- vs lattice) and even within lattice models (i.e.
-landscape resolution) can lead to qualitatively different results at a given neighborhood size.
-However our main conclusions are given in terms of orders of magnitude -- roughly, less than 100, 100-1000, and over 1000 -- and we believe the various models used to derive the values in table 1 are close enough for this interpretation to remain a useful yardstick.
+Good point, we have removed Bebicium vittatum from table 1.  
 }

### 2020-02-10 ms/spaceness.tex
expanded rev 3 responses

@@
 to modulate the spatial scale of interactions at all three stages by adjusting the standard deviation of the corresponding Gaussian functions. 
-As in previous work \citep{Wright1943,Ringbauer2017}, $\sigma$ is equal to the mean parent-offspring distance along each of the x and y axes.
+As in previous work \citep{Rousset1997}, $\sigma$ is equal to the mean parent-offspring distance along each of the x and y axes.
 
@@
 others (e.g., number of segregating sites)
-are more strongly determined by global $N_e$ or by the ratio of the two, and definitions of "population density" in genetic versus ecological studies may lead to varying estimates of neighborhood size for a given species \llabel{rhocaveat}.
-% These values should be compared to our simulation results with some caution, as though we expect neighborhood size to have similar effects across species of varying global $N_e$ \citep{Wright1946}, in our study we evaluated only a relatively small population of $\approx 10,000$.
+by global $N_e$, or the ratio of the two. 
 \llabel{popsizecaveat}
-In addition, these empirical examples are likely biased towards small-neighborhood species (because few studies have quantified neighborhood size in species with very high dispersal or population density).
+In addition, definitions of "population density" in genetic versus ecological studies may lead to varying estimates of neighborhood size for a given species \llabel{rhocaveat}, and these empirical examples may be biased towards small-neighborhood species because few studies have quantified neighborhood size in species with very high dispersal or population density.
+
 However, from the available data we find that neighborhood sizes in the range we simulated are fairly common across a range of taxa.

### 2020-02-10 ms/second-review-responses.tex
fixed missing math symbols in response to rev 3

@@
 
-Thus relative to models like \cite{Rousset1997} our definition of \sigma is probably too low and \rho probably too high; while relative to neighborhood sizes calculated from mark-recapture studies measuring parent-offspring distances our figures should be consistent.
-We have expanded the caveats at the head of the discussion paragraph interpreting empirical studies noting that definitions of \rho in ecological vs genetic studies mean results should be interpreted with caution \llname{rhocaveat}; however, as we mainly interpret these figures in terms of order-of-magnitude differences (roughly less than 100, 100-1000, and over 1000) we think our results are general enough to be robust to these modeling differences while still giving a useful yardstick.
+Thus relative to models like \cite{Rousset1997} our definition of $\sigma$ is probably too low and $\rho$ probably too high; while relative to neighborhood sizes calculated from mark-recapture studies measuring parent-offspring distances our figures should be consistent.
+We have expanded the caveats at the head of the discussion paragraph interpreting empirical studies noting that definitions of $\rho$ in ecological vs genetic studies mean results should be interpreted with caution \llname{rhocaveat}; however, as we mainly interpret these figures in terms of order-of-magnitude differences (roughly less than 100, 100-1000, and over 1000) we think our results are general enough to be robust to these modeling differences while still giving a useful yardstick.
 Indeed for many of the empirical studies the range of NS values given is larger than the differences we expect from the parameter definitions described above.

### 2020-02-11 ms/second-review-responses.tex
minor edits

@@
 Because our simulated genome is autosomal and individuals hermaphroditic and non-selfing, a locus will experience a mix of "maternal" and "paternal" transmission events along its genealogy.
-So we see this as a scaling issue rather than one of a "bimodal" dispersal kernal per se -- the expected distance traveled by a locus along a pedigree link is $\sqrt{3/2} \sigma$, rather than $\sqrt{2} \sigma$.
+So we see this as a scaling issue rather than one of a "bimodal" dispersal kernal per se -- the expected axial displacement of a locus along a pedigree link is $\sqrt{3/2} \sigma$, rather than $\sigma$.
 However we note that in terms of summary statistics the effects of scaling $\sigma$ by $sqrt{3/2}$ are weaker than those generated by factors like changing landscape resolution in lattice vs.
@@
 Appendix 1), so we don't think this significantly changes our interpretation of model outputs.
-We have instead expanded some text in the sections on empirical data and further directions, noting that the relationship between neighborhood size and any given summary statistic depends on the on assumed definitions of population density and details of the breeding system.
 }
@@
 
-\reply{This is a good point -- differences in allele frequency and LD structure certainly have significant effects on PRS transferability (at least when the GWAS is based on SNP assays rather than whole-genome sequencing.
+\reply{This is a good point -- differences in allele frequency and LD structure certainly have significant effects on PRS transferability (at least when the GWAS is based on SNP assays rather than whole-genome sequencing).
 We think the population structure issue is important to point to here because it will not be solved by simply generating more detailed genomic data from existing samples, but have added a caveat noting these other important potential sources of error at the head of that sentence.
@@
 
-This inflation in neighborhood size is then partially offset by our use of census rather than "effective" population size when calculating population density, because $N_e$ of our spatial model estimated from $\pi$ in random-mating spatial models is roughly (2/3) the census size.
-We chose this method for density and sigma calculations because we sought to parameterize our model with values that could in theory be observed in real populations.
+This inflation in neighborhood size is then partially offset by our use of census rather than "effective" population size when calculating population density, because $N_e$ of our spatial model estimated from $\pi$ in random-mating spatial models is roughly 2/3 the census size.
+We chose this method for density and sigma calculations because we sought to parameterize our model with values that could be observed in real populations.
 We also find the concept of "effective population size" in this context even more confusing than usual -- as we show, spatial processes in both evolution and sampling radically shift the distribution of coalescence times and the variance in offspring numbers in ways that mimic the effects of classic $N_{e}$ scaling in random-mating models, but are driven by fundamentally different processes.
 
-Thus relative to models like \cite{Rousset1997} our definition of $\sigma$ is probably too low and $\rho$ probably too high; while relative to neighborhood sizes calculated from mark-recapture studies measuring parent-offspring distances our figures should be consistent.
+Thus relative to models like \cite{Rousset1997} our definition of $\sigma$ is probably too low and $\rho$ probably too high such that the neighborhood sizes we show in plots are roughly consistent with previous theoretical work; while relative to neighborhood sizes calculated from mark-recapture studies measuring parent-offspring distances our figures should be consistent to the extent that the breeding and dispersal system matches our simulation.
 We have expanded the caveats at the head of the discussion paragraph interpreting empirical studies noting that definitions of $\rho$ in ecological vs genetic studies mean results should be interpreted with caution \llname{rhocaveat}; however, as we mainly interpret these figures in terms of order-of-magnitude differences (roughly less than 100, 100-1000, and over 1000) we think our results are general enough to be robust to these modeling differences while still giving a useful yardstick.

### 2020-02-11 ms/second-review-responses.tex

@@
 %%%%%%%%%%%%%%
-\reviewersection{AE}
+\reviewersection{Decision Letter}
 
+When you submit a revised manuscript please include a response to each of the reviewers' comments.
+It is most important that you address the following in a revised manuscript: 1) in response to reviewer 3, add text to help clarify how the reproduction scheme / dispersal model differs from models that do not induce the bimodality in the distance distribution the reviewer points out (see my comments below as well); 2) Respond to the concern about the definitions of sigma and neighborhood size; 3) Reviewer 2 has several minor comments that would be straightforward and helpful to address/add depth into your discussion (e.g., comments 2, 4, 5).
+4) The other minor comments from each reviewer.
+
+\reviewersection{AE Comments}
 The point of R3 on the bimodality in the realized dispersal distance in your model is interesting.
@@
 \reply{
-Another important point set of points on relating neighborhood sizes from different models here.
-In our simulation the average x axis displacement of a locus along any link in the pedigree is $\sigma$ for the "maternal" material (i.e.
-deriving from the parent from which dispersal is calculated), but for the "paternal" parent it is the sum of two independent normal draws each with variance $\sigma^2$.
+These are an important point set of points on relating neighborhood sizes from different models which we have aimed to clarify.
+In our simulation the x axis displacement of a locus along any link in the pedigree is normally distributed with standard deviation $\sigma$ for the "maternal" material (i.e.
+deriving from the parent from which dispersal is calculated), but for the "paternal" parent it is the sum of two independent normal draws each with with standard deviation $\sigma$.
 Averaging over a random series of maternal and paternal transmissions (because our simulated genome is autosomal), variance in axial position across pedigree links is then $(3/2)\sigma^2$ and the expected displacement $\sqrt{3/2}\sigma$.
@@
 }
+
+\cjb{ADK: I made some small edits but I'm pretty happy about how it reads}
 

### 2020-02-11 ms/spaceness.tex
minor touch ups

@@
 This suggests that GWAS focused on finding strongly associated alleles for traits controlled by a limited number of variants in the genome are likely robust to the impacts of continuous spatial structure.
-However, methods that analyze the combined effects of thousands or millions of weakly associated variants such as polygenic risk scores \citep{Khera2018} are likely to be affected by subtle population structure.
+However, methods that analyze the combined effects of thousands or millions of weakly associated variants such as polygenic risk scores \citep{wray2007prediction,international2009common} are likely to be affected by subtle population structure.
 Indeed as recently identified in studies of genotype associations for human height in Europe \citep{Berg2018,Sohail2018}, PC regression GWAS in modern human populations do include residual signal of population structure in large-scale analyses of polygenic traits.

### 2020-02-14 ms/second-review-responses.tex
minor touch ups
minor adjustments

@@
 In response to reviewer 2 we have added several caveats to the GWAS section noting other sources of error such as allele frequency and LD variation among populations.
-Reviewer 3 raised two good points on our definition of neighborhood size and the interpetation of $\sigma$ -- first, that $\sigma$ refers to one- rather than two-dimensional distances between parent and offspring (this is consistent with previous theoretical work, and we have corrected our definition in the text), and second, that the use of separate draws for mate selection and dispersal means the total distance traveled by a locus along a pedigree link is greater for "paternal" than "maternal" material (taking "maternal" to mean the parent from which dispersal is calculated).
-Because our simulated genome is autosomal and individuals hermaphroditic and non-selfing, a locus will experience a mix of "maternal" and "paternal" transmission events along its genealogy.
-So we see this as a scaling issue rather than one of a "bimodal" dispersal kernal per se -- the expected axial displacement of a locus along a pedigree link is $\sqrt{3/2} \sigma$, rather than $\sigma$.
-However we note that in terms of summary statistics the effects of scaling $\sigma$ by $sqrt{3/2}$ are weaker than those generated by factors like changing landscape resolution in lattice vs.
+Reviewer 3 raised two good points on our definition of neighborhood size and the interpetation of $\sigma$ -- first, that $\sigma$ refers to one- rather than two-dimensional distances between parent and offspring (this is consistent with previous theoretical work, and we have corrected our definition in the text), and second, that the use of separate draws for mate selection and dispersal means the total distance traveled by a locus along a pedigree link is greater for ``paternal'' than ``maternal'' material (taking ``maternal'' to mean the parent from which dispersal is calculated).
+Because our simulated genome is autosomal and individuals hermaphroditic and non-selfing, a locus will experience a mix of ``maternal'' and ``paternal'' transmission events along its genealogy.
+So we see this as a scaling issue rather than one of a ``bimodal'' dispersal kernal per se -- the expected axial displacement of a locus along a pedigree link is $\sqrt{3/2} \sigma$, rather than $\sigma$.
+However we note that in terms of summary statistics the effects of scaling $\sigma$ by $\sqrt{3/2}$ are weaker than those generated by factors like changing landscape resolution in lattice vs.
 continuous models (e.g.
@@
 I appreciate the appendix comparing summary statistics in the continuous and stepping-stone models, although I must confess I am now even more uncertain than I was before about which models are more appropriate.
-I get that the coalescent models make assumptions that are not directly interpretable in terms of observable parameters (although are "interactions between geographically distant parts of the species range" really biologically unrealistic?).
+I get that the coalescent models make assumptions that are not directly interpretable in terms of observable parameters (although are ``interactions between geographically distant parts of the species range'' really biologically unrealistic?).
 This forward model is more interpretable, in that sense, but does it actually produce more realistic data?
@@
 \begin{point}{\revref}
- "PCA is unable to capture".
+ ``PCA is unable to capture''.
 Is this a sample size effect?
@@
 \reply{
-We have added the caveat "given the sample sizes tested" to the above statement. \llname{samplesizenote}
+We have added the caveat ``given the sample sizes tested'' to the above statement. \revref
 }
 
-\begin{point}{Line 685.}
+\begin{point}{\revref}
  Is Khera et al (2018) really the right reference for the idea of polygenic scores? They have been around for a lot longer than that.
@@
 
-\reply{
+\reply{Yes this was an oversight. We now cite Wray et al. 2007 for individual disease risk prediction from GWAS estimates and the 
+first, to our knowledge, PGS application (International Schizophrenia Consortium, 2009).
 }
 
-\begin{point}{Line 686. }
+\begin{point}{\revref}
  I don't know that the results of Berg and Sohail et al. are indicating a fundamental issue with PC correction, rather than a problem with meta-analysis.
@@
 
-\begin{point}{Line 689-92. }
+\begin{point}{\llname{ldcaveats}}
  Is the result of Martin et al.
@@
 
-\begin{point}{Line 752 "population genetics flavor"}
+\begin{point}{``population genetics flavor'' \revref}
 I don't understand what this means?
@@
 \reply{
-Peter(???)
+We meant to contrast this sort of ``single-species'' model,
+common in population genetics,
+to a more ``ecological'' model involving other species as well.
+We've reworded this. \revref
 }
 
-\begin{point}{Line 800}
-"process-driven descriptions of ancestry and/or more generalized unsupervised methods". Aren't these two things opposite? That's ok, but it sounds a bit like you have no idea what would help...
+\begin{point}{\revref}
+``process-driven descriptions of ancestry and/or more generalized unsupervised methods''. Aren't these two things opposite? That's ok, but it sounds a bit like you have no idea what would help...
 \end{point}
@@
 Good point here.
-We have corrected this to "$\sigma$ is the mean parent--offspring distance along each of the x and y axes" and now direct readers to \citep{Rousset1997} for discussion of how these parameters are defined in one- and two-dimensional habitats.
+We have corrected this to ``$\sigma$ is the mean parent--offspring distance along each of the x and y axes'' and now direct readers to \citep{Rousset1997} for discussion of how these parameters are defined in one- and two-dimensional habitats.
 The mean parent-offspring distance in two dimensions is $\sigma\sqrt{2}$, but we follow previous work in defining neighborhood size as $4 \pi \rho \sigma^2$ (i.e.
@@
 
-Additionally, the neighborhood size as defined by the authors (see line 211) is a function of the census density rather than the effective density of individuals.
+    Additionally, the neighborhood size as defined by the authors (see \llname{ll:nw_defn}) is a function of the census density rather than the effective density of individuals.
 I am therefore not fully convinced that the range of values for this parameter is comparable to that given in Table 1.
@@
 These are an important point set of points on relating neighborhood sizes from different models which we have aimed to clarify.
-In our simulation the x axis displacement of a locus along any link in the pedigree is normally distributed with standard deviation $\sigma$ for the "maternal" material (i.e.
-deriving from the parent from which dispersal is calculated), but for the "paternal" parent it is the sum of two independent normal draws each with with standard deviation $\sigma$.
+In our simulation the x axis displacement of a locus along any link in the pedigree is normally distributed with standard deviation $\sigma$ for the ``maternal'' material (i.e.
+deriving from the parent from which dispersal is calculated), but for the ``paternal'' parent it is the sum of two independent normal draws each with with standard deviation $\sigma$.
 Averaging over a random series of maternal and paternal transmissions (because our simulated genome is autosomal), variance in axial position across pedigree links is then $(3/2)\sigma^2$ and the expected displacement $\sqrt{3/2}\sigma$.
@@
 
-This inflation in neighborhood size is then partially offset by our use of census rather than "effective" population size when calculating population density, because $N_e$ of our spatial model estimated from $\pi$ in random-mating spatial models is roughly 2/3 the census size.
+This inflation in neighborhood size is then partially offset by our use of census rather than ``effective'' population size when calculating population density, because $N_e$ of our spatial model estimated from $\pi$ in random-mating spatial models is roughly 2/3 the census size.
 We chose this method for density and sigma calculations because we sought to parameterize our model with values that could be observed in real populations.
-We also find the concept of "effective population size" in this context even more confusing than usual -- as we show, spatial processes in both evolution and sampling radically shift the distribution of coalescence times and the variance in offspring numbers in ways that mimic the effects of classic $N_{e}$ scaling in random-mating models, but are driven by fundamentally different processes.
+We also find the concept of ``effective population size'' in this context even more confusing than usual -- as we show, spatial processes in both evolution and sampling radically shift the distribution of coalescence times and the variance in offspring numbers in ways that mimic the effects of classic $N_{e}$ scaling in random-mating models, but are driven by fundamentally different processes.
 
@@
 
-\cjb{ADK: I made some small edits but I'm pretty happy about how it reads}
+\ak{ADK: I made some small edits but I'm pretty happy about how it reads}
 

### 2020-02-14 ms/spaceness.tex
minor adjustments

@@
 and used mean population density ($\rho$, census size divided by total area)
-to compute neighborhood size as $N_W = 4 \pi \rho \sigma^2$. 
+to compute neighborhood size as $N_W = 4 \pi \rho \sigma^2$.  \llabel{ll:nw_defn}
 To estimate generation times, we stored ages of the parents of every new individual born across 200 timesteps, 
@@
 -- as neighborhood size increases, the total proportion of variance explained by the first 10 PC axes declines from roughly 10\% to 4\% (Figure \ref{fig:gwas}B). 
-Essentially, PCA seems unable to effectively summarize the weak population structure present in large-neighborhood simulations given the sample sizes we tested, \revpoint{2}{2} \llabel{samplesizenote}
+Essentially, PCA seems unable to effectively summarize the weak population structure present in large-neighborhood simulations given the sample sizes we tested, \revpoint{2}{2}
 but these populations continue to have enough spatial structure to create significant correlations between genotypes and the environment.
@@
 However, methods that analyze the combined effects of thousands or millions of weakly associated variants such as polygenic risk scores \citep{wray2007prediction,international2009common} are likely to be affected by subtle population structure.
-Indeed as recently identified in studies of genotype associations for human height in Europe \citep{Berg2018,Sohail2018}, PC regression GWAS in modern human populations do include residual signal of population structure in large-scale analyses of polygenic traits.
+\revpoint{2}{3}
+Indeed as recently identified in studies of genotype associations for human height in Europe \citep{Berg2018,Sohail2018}, \revpoint{2}{4}
+PC regression GWAS in modern human populations do include residual signal of population structure in large-scale analyses of polygenic traits.
 In addition to error associated with varying patterns of linkage disequilibrium and allele frequency among populations, the confounding of environmental and genetic effects on phenotypes introduced by population structure is expected to lead to low predictive power when polygenic scores are generated for populations outside the original GWAS cohort, \llabel{ldcaveats}
@@
 although the general predictions should be robust to similar forms of regulation.
-Since populations are still entirely \emph{intrinsically} regulated,
-our model still has a very strong ``population genetics'' flavor.
+As is usual in population genetics, \revpoint{2}{6}
+the populations are entirely \emph{intrinsically} regulated.
 Alternatively, population size could be regulated by interactions with other species
@@
 Perhaps optimistically,
-we posit that process-driven descriptions of ancestry 
+we posit that process-driven descriptions of ancestry  \revpoint{2}{7}
 and/or more generalized unsupervised methods may be able to better account for carry out this task.

### 2020-02-14 ms/not_bimodal.R

@@ -1,0 +1,8 @@
+mat <- cbind(rnorm(1e6), rnorm(1e6))
+pat <- cbind(rnorm(1e6), rnorm(1e6)) * (runif(1e6) < 0.5)
+dist <- sqrt( rowSums((mat+pat)^2) )
+
+pdf(file="not_bimodal.pdf", width=4, height=2.5, pointsize=10)
+par(mar=c(4,4,1,1)+.1)
+    hist(dist, breaks=50, main=expression(abs(M + P * theta)), xlab='dispersal distance')
+dev.off()

### 2020-02-15 ms/second-review-responses.tex
plr edits

@@
 
-When you submit a revised manuscript please include a response to each of the reviewers' comments.
-It is most important that you address the following in a revised manuscript: 1) in response to reviewer 3, add text to help clarify how the reproduction scheme / dispersal model differs from models that do not induce the bimodality in the distance distribution the reviewer points out (see my comments below as well); 2) Respond to the concern about the definitions of sigma and neighborhood size; 3) Reviewer 2 has several minor comments that would be straightforward and helpful to address/add depth into your discussion (e.g., comments 2, 4, 5).
-4) The other minor comments from each reviewer.
+\begin{quote}
+    It is most important that you address the following in a revised manuscript:
+    \begin{enumerate}
+        \item in response to reviewer 3, add text to help clarify how the reproduction scheme / dispersal model differs from models that do not induce the bimodality in the distance distribution the reviewer points out (see my comments below as well);
+        \item Respond to the concern about the definitions of sigma and neighborhood size; 
+        \item Reviewer 2 has several minor comments that would be straightforward and helpful to address/add depth into your discussion (e.g., comments 2, 4, 5). 
+        \item The other minor comments from each reviewer. 
+    \end{enumerate}
+\end{quote}
+
+Thanks very much for the useful input.
+The inaccuracy in describing dispersal under the model was our mistake,
+and we have added more careful discussion of this point (see below).
+We have modified the manuscript to incorporate all remaining feedback,
+as enumerated below,
+and think this has improved the manuscript.
 
 \reviewersection{AE Comments}
+
+\begin{quote}
 The point of R3 on the bimodality in the realized dispersal distance in your model is interesting.
@@
 So at a minimum, some caveats directly in the GWAS discussion section would help.
+\end{quote}
 
-\reply{Thank you for your comments.
+Thank you for your comments.
 We have responded to each reviewer below.
@@
 So we see this as a scaling issue rather than one of a ``bimodal'' dispersal kernal per se -- the expected axial displacement of a locus along a pedigree link is $\sqrt{3/2} \sigma$, rather than $\sigma$.
-However we note that in terms of summary statistics the effects of scaling $\sigma$ by $\sqrt{3/2}$ are weaker than those generated by factors like changing landscape resolution in lattice vs.
+(Note that although the parent-offspring displacement is a mixture of mean-zero Gaussians
+with respect to a bimodal distribution of standard deviations,
+the norm of any such mixtures is never itself bimodal.)
+% However we note that in terms of summary statistics the effects of scaling $\sigma$ by $\sqrt{3/2}$ are weaker than those generated by factors like changing landscape resolution in lattice vs.
 continuous models (e.g.
 Appendix 1), so we don't think this significantly changes our interpretation of model outputs.
-}
+We have added a paragraph discussing these points (while trying to keep out of the weeds),
+to \llname{ll:sigma_effective}.
 
@@
 While we agree that our analysis found problems with demographic inference under isolation by distance, since this manifests as higher variance in predictions rather than a straightforward bias as we expected (i.e.
-a reduction in recent inferred Ne), we have chosen to keep the somewhat more nuanced presentation of results in that section used in a previous version.
+a reduction in recent inferred Ne), we have chosen to keep the somewhat more nuanced presentation of results in that section used in a previous version,
+although we added an additional note about the between-replicate variance to the discussion.
+\llname{ll:variance_note}
 
@@
 
+Thanks for the encouraging words and helpful points,
+as well as for your mindfulness of the burden of additional analyses.
+
 \begin{point}{}
@@
 This is a good point.
-We have added a sentence to the discussion noting that while our continuous model has some advantages over discrete models, the best model for any empirical system will inevitably depend on the life history of the organism in question.
+We have added additional discussion noting that while our continuous model has some advantages over discrete models, the best model for any empirical system will inevitably depend on the life history of the organism in question.
 \llname{bestmodelcaveat}
@@
 \reply{
-We think Berg et al.
-shows that population structure leaves signals in GWAS that resemble those of polygenic selection, and given that this was most apparent in GIANT data employing PCA corrections, indicates a problem with the GWAS that make up the GIANT study at least as much as any issues with meta-analysis.
+    We think \citet{Berg2018} shows that population structure leaves signals in GWAS that resemble those of polygenic selection, and given that this was most apparent in GIANT data employing PCA corrections, indicates a problem with the GWAS that make up the GIANT study at least as much as any issues with meta-analysis.
 }
@@
 common in population genetics,
-to a more ``ecological'' model involving other species as well.
+to a more ``ecological'' model involving explicit modeling or resources and/or other species.
 We've reworded this. \revref
@@
 
+Thanks. We are also fairly pleased with it, overall.
+\plr{this is probably too much =)}
+
 \begin{point}{}
@@
 Good point here.
-We have corrected this to ``$\sigma$ is the mean parent--offspring distance along each of the x and y axes'' and now direct readers to \citep{Rousset1997} for discussion of how these parameters are defined in one- and two-dimensional habitats.
-The mean parent-offspring distance in two dimensions is $\sigma\sqrt{2}$, but we follow previous work in defining neighborhood size as $4 \pi \rho \sigma^2$ (i.e.
-without the factor of $\sqrt{2}$; see \citep{Rousset1997} for details).
-\llname{axialdist}
+We did intend to define $\sigma$ as is usual in the literature,
+so have corrected the wording
+and now direct readers to \citep{Rousset1997} for discussion of how these parameters are defined in one- and two-dimensional habitats. \llname{axialdist}
 }
@@
 \reply{
+    You are entirely correct here; apologies for the confusion.
+    To clarify the definition of $\sigma$ and discuss these issues somewhat,
+    we have added a new paragraph at \llname{ll:sigma_effective}.
+    As we discuss there, we still calculate $N_W = 4 \sigma^2 \rho$,
+    where $\sigma$ is the mean mother-child dispersal distance
+    (not including the distance between mates)
+    and $\rho$ is the census population density.
+    This is not the quantity that is predicted
+    to best describe geographic patterns of genetic variation,
+    which would use \emph{effective} versions of both $\sigma$ and $\rho$.
+    This does not bother us for a few reasons.
+    First, we do not attempt to match statistics in our simulations
+    to theoretical results (e.g., the Wright--Malec\'ot formula).
+    Instead, we use $N_W$ as a proxy for ``strength of spatial structure'',
+    and in these simulations, the ``effective'' $N_W$ certainly has a monotonic
+    relationship with the $N_W$ we compute.
+    The remaining question is whether, as the reviewer asks,
+    the actual values of $N_W$ are comparable to those computed in the wild.
+    This depends, of course, on how those values were estimated in wild populations.
+    In our opinion, a straightforward estimate of $N_W$
+    using observed dispersal distances and census density or reproductive adults
+    (similar to how we compute $N_W$ for simulations)
+    is likely more robust and reflective of modern demography
+    than most methods to estimate ``effective'' vesions.
+
+    Furthermore, our imprecisions roughly cancel out.
+    As now noted in the text, an average of one and two dispersal distances
+    is a better model for the mean parent-offspring distance;
+    this would give us a value of $\sqrt{3/2} \sigma$,
+    so by using only $\sigma$ we have lowered $N_W$ by a factor of about 2/3.
+    However, effective population density is lower than census density
+    by a factor of also around 2/3,
+    which would cancel the first.
+    So, we are fairly confident that our values of $N_W$ match the theoretical versions
+    to within the precision we require (which is only order-of-magnitue, really).
+
 These are an important point set of points on relating neighborhood sizes from different models which we have aimed to clarify.
@@
 
+Thank you very much!
+
 \begin{point}{}

### 2020-02-18 ms/spaceness.tex

@@
     Stepping-stone models reproduce some of these effects, but discretizing the landscape introduces artifacts which in some cases are exacerbated at higher resolutions.
-        The combination of low dispersal and clustered sampling causes demographic inference from the site frequency spectrum to infer more turbulent demographic histories, but averaged results across multiple simulations were surprisingly robust to isolation by distance.
+    The combination of low dispersal and clustered sampling causes demographic inference from the site frequency spectrum to infer more turbulent demographic histories, but averaged results across multiple simulations
+    revealed surprisingly little systematic bias.
     We also show that the combination of spatially autocorrelated environments and limited dispersal
@@
 that incorporates overlapping generations, local dispersal of offspring, and density-dependent survival. 
-We simulate chromosome-scale genomic data in tens of thousands of individuals from parameter regimes relevant to common subjects of population genetic investigation such as humans and \textit{Drosophila}, and output the full genealogy and recombination history of all final-generation individuals.
+We simulate chromosome-scale genomic data in tens of thousands of individuals from parameter regimes relevant to common subjects of population genetic investigation, and output the full genealogy and recombination history of all final-generation individuals.
 We use these simulations to test how sampling strategy interacts with geographic population structure to cause systematic variation in population genetic summary statistics 
@@
 A key early observation by \citet{Wright1946} is that this balance is often nicely summarized by the ``neighborhood size'',
-defined to be $N_W = 4 \pi \rho \sigma^2$, where $\sigma$ is the mean parent--offspring distance along each of the x and y axes and $\rho$ is population density (see \citep{Rousset1997} for further discussion of parameter defiitions in one- and two-dimensional habitats) \llabel{axialdist}.
+defined in two dimensions to be $N_W = 4 \pi \rho \sigma^2$,
+where $\sigma^2$ is one half of the mean squared parent--offspring distance
+and $\rho$ is population density
+(see \citet{Rousset1997} for further discussion of parameter definitions in one- and two-dimensional habitats) \llabel{axialdist}. 
 This can be thought of as proportional to the average number of potential mates for an individual (those within distance $2 \sigma)$,
@@
 Each time step consists of three stages: reproduction, dispersal, and mortality. 
-To reduce the parameter space we use the same parameter, denoted $\sigma$, 
+To reduce the number of parameters we use the same parameter, denoted $\sigma$, 
 to modulate the spatial scale of interactions at all three stages by adjusting the standard deviation of the corresponding Gaussian functions. 
-As in previous work \citep{Rousset1997}, $\sigma$ is equal to the mean parent-offspring distance along each of the x and y axes.
+Informally, we think of $\sigma$ as the ``dispersal distance'',
+although only one of those stages is dispersal.
 
@@
 Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
-with probability of choosing a neighbor at distance $d$ proportional to $\exp(-d^2 / 2 \sigma^2)$.
+with probability of choosing a neighbor at distance $d$ proportional to the Gaussian density
+with mean zero and standard deviation $\sigma$, which is $g(d) = \exp(-d^2 / 2 \sigma^2)/ (2 \sigma^2)$.
 
 To maintain a stable population, mortality increases with local population density.
-To do this we say that individuals at distance $d$ have a competitive interaction with strength $g(d)$,
-where $g$ is the Gaussian density with mean zero and standard deviation $\sigma$.
+To do this we say that individuals at distance $d$ have a competitive interaction with strength $g(d)$.
 Then, the sum of all competitive interactions with individual $i$ is $n_i = \sum_j g(d_{ij})$,
@@
 However, nonlinearity of local demographic stochasticity causes
-actual realized averages of these demographic quantitites to deviate from the specified values
+actual realized averages of these demographic quantities to deviate from the specified values
 in a way that depends on the neighborhood size.
@@
 
+Also note that $\sigma$ controls the dispersal of offspring away from only \emph{one} parent \llabel{ll:sigma_effective}
+(e.g., the seed parent for plants),
+while the usual population genetics definition of dispersal distance
+involves the distance to a randomly chosen one of the two parents.
+(thus taking into account the distance from pollen parent as well).
+This second component -- the distance between mates --
+has in our simulations a distribution that is of order $\sigma$ but that depends on the population's patchiness.
+If both between-mate distance and dispersal distance has variance $\sigma^2$ along each axis,
+then the mean squared distance to a randomly chosen parent along that axis would be
+$(\sigma^2 + 2 \sigma^2)/2 = 3 \sigma^2 /2$.
+Also note that Wright's definition of neighborhood size should rightly use the \emph{effective}
+dispersal distance, i.e., the mean squared displacement along an axis
+between parent-child pairs found moving back along a lineage \citep{barton2002neutral}.
+Nonetheless, we use $\sigma$ and $\rho$ as defined here to compute $N_W$
+because these quantities are more easily observable in practice than their ``effective'' versions.
+(Also, note that effective population density is lower than the actual population density,
+while the $\sigma$ we use is an underestimate,
+so our numbers may be close to the ``effective'' values used in theory.
+
 %%%%%%%%%%%%%
@@
 
-The effects of sampling on summary statistic estimates were summarized by testing for differences in mean (ANOVA, \citep{Rcore2018}) and variance (Levene's test, \citep{Fox2011}) across sampling strategies for each summary statistic.
+The effects of sampling on summary statistic estimates were summarized by testing for differences in mean (ANOVA, \citet{Rcore2018}) and variance (Levene's test, \citet{Fox2011}) across sampling strategies for each summary statistic.
 
@@
 Run times for our model scale approximately linearly with neighborhood size (Figure \ref{fig:runtimes}), with the lowest neighborhood sizes reaching 30N generations in around a day and those with neighborhood size approaching 1,000 requiring up to three weeks of computation.
-As currently implemented running simulations at neighborhood sizes more than 1,000 to coalescence is likely impractical, though running these models for more limited timescales and then ``recapitating'' the simulation using reverse-time simulation from the resulting tree sequence in msprime is possible \citep{haller2019treesequence}.
+As currently implemented, running simulations at neighborhood sizes more than 1,000 to coalescence is likely impractical, though running these models for more limited timescales and then ``recapitating'' the simulation using reverse-time simulation from the resulting tree sequence in msprime is possible \citep{haller2019treesequence}.
 \llabel{runtimesxn}
@@
 We were surprised to see that both stairwayplot and SMC++ can tolerate relatively strong isolation by distance -- i.e., neighborhood sizes of 20 -- and still perform well when averaging results across multiple simulations.
+(However, note the high amount of between-simulation variance seen in \autoref{fig:demography_supplement}.) \llabel{ll:variance_note}
 Inference in populations with neighborhood sizes over 20 was relatively unbiased unless samples were concentrated in the middle of the range (Figure \ref{fig:demography}). 
@@
 
-We suggest a straightforward alternative for species in which the primary axes of population differentiation is space 
-(note this is likely not the case for some modern human populations): 
+We suggest a straightforward alternative for species in which the primary axes of population differentiation
+are associated with geography (note this is likely not the case for some modern human populations): 
 run a GWAS with spatial coordinates as phenotypes and check for $p$-value inflation or significant associations.
@@
 \llabel{landscapesizecaveat}
-Indeed, our finding that stepping-stone models reproduce only some of the effects on genetic variation seen in a continuous-space model (Figure \ref{fig:grid_stats}) suggests that qualitatively similar models can produce different results dependent on the specific parameterizations used.
-While we believe our continuous model is a more appropriate depiction of evolution in a continuous landscape than lattice models, it is likely that some populations and breeding systems do more closely resemble a series of interconnected random-mating populations.
+Indeed, our finding that stepping-stone and continuous-space models match
+in only certain aspects of genetic variation (Figure \ref{fig:grid_stats})
+shows that qualitatively similar models can produce different results dependent on the specific parameterizations used.
+While we believe our continuous model is a more appropriate depiction of 
+many species' demographies than lattice models,
+it is likely that some populations and breeding systems do more closely resemble a series of interconnected random-mating populations.
 As with all population models, the best approximation for any empirical system will depend on the natural history of the species in question.
 \llabel{bestmodelcaveat}
-Much additional work remains to be done to better understand how life history and range size interact to shape genetic variation in continuous space, which we leave to future studies.
+Much additional work remains to be done to better understand how life history, range size,
+and habitat geometry interact to shape genetic variation in continuous space, which we leave to future studies.
 
@@
 In general we find many of the qualitative trends are similar among continuous and stepping-stone models and that, at low neighborhood sizes, many  (but not all) statistics from stepping-stone models approach the continuous model as the resolution of the grid increases.
-For example, $\theta_\pi$ is inflated at low neighborhood sizes (i.e., low $m$), and the extent of the inflation increases to approach the continuous case as the resolution of the landscape increases.
-Similar patterns are observed for $F_{IS}$ and observed heterozygosity.
-However, $\theta_{W}$ behaves differently, showing a non-monotonic relationship with grid resolution.
+For example, $\theta_\pi$ is lower in stepping-stone models 
+at low neighborhood sizes (i.e., low $m$),
+but increases to approach the continuous case as the resolution of the landscape increases.
+Similar patterns are observed for observed heterozygosity. However, $\theta_{W}$ behaves differently, showing a non-monotonic relationship with grid resolution.
 This results in an increasingly positive Tajima's $D$ in grid simulations at small neighborhood sizes, to a much greater extent than seen in a continuous model.
 In contrast to $\theta_\pi$, increasing the resolution of the grid causes Tajima's $D$ to deviate \emph{more} from what is seen in the continuous model.
+Similarly, although $F_{IS}$ approaches the continuous case as the resolution of the grid increases at very small neighborhood sizes,
+at intermediate neighborhood sizes the continuous case best matches intermediate grid resolutions.
 

### 2020-02-19 ms/second-review-responses.tex
incorporating plr response and removing duplicates

@@
 
-Thanks. We are also fairly pleased with it, overall.
-\plr{this is probably too much =)}
+Thank you for your comments.
 
@@
     to within the precision we require (which is only order-of-magnitue, really).
-
-These are an important point set of points on relating neighborhood sizes from different models which we have aimed to clarify.
-In our simulation the x axis displacement of a locus along any link in the pedigree is normally distributed with standard deviation $\sigma$ for the ``maternal'' material (i.e.
-deriving from the parent from which dispersal is calculated), but for the ``paternal'' parent it is the sum of two independent normal draws each with with standard deviation $\sigma$.
-Averaging over a random series of maternal and paternal transmissions (because our simulated genome is autosomal), variance in axial position across pedigree links is then $(3/2)\sigma^2$ and the expected displacement $\sqrt{3/2}\sigma$.
-So from the perspective of genes sampled at the present time our neighborhood size calculations are lower than those in models like those described in \cite{Rousset1997} by a factor of (3/2) because of differences in the breeding model.
-
-This inflation in neighborhood size is then partially offset by our use of census rather than ``effective'' population size when calculating population density, because $N_e$ of our spatial model estimated from $\pi$ in random-mating spatial models is roughly 2/3 the census size.
-We chose this method for density and sigma calculations because we sought to parameterize our model with values that could be observed in real populations.
-We also find the concept of ``effective population size'' in this context even more confusing than usual -- as we show, spatial processes in both evolution and sampling radically shift the distribution of coalescence times and the variance in offspring numbers in ways that mimic the effects of classic $N_{e}$ scaling in random-mating models, but are driven by fundamentally different processes.
-
-Thus relative to models like \cite{Rousset1997} our definition of $\sigma$ is probably too low and $\rho$ probably too high such that the neighborhood sizes we show in plots are roughly consistent with previous theoretical work; while relative to neighborhood sizes calculated from mark-recapture studies measuring parent-offspring distances our figures should be consistent to the extent that the breeding and dispersal system matches our simulation.
-We have expanded the caveats at the head of the discussion paragraph interpreting empirical studies noting that definitions of $\rho$ in ecological vs genetic studies mean results should be interpreted with caution \llname{rhocaveat}; however, as we mainly interpret these figures in terms of order-of-magnitude differences (roughly less than 100, 100-1000, and over 1000) we think our results are general enough to be robust to these modeling differences while still giving a useful yardstick.
-Indeed for many of the empirical studies the range of NS values given is larger than the differences we expect from the parameter definitions described above.
-We also think that our study of lattice simulations in appendix 1 demonstrates nicely how qualitatively similar simulations can lead to quite different patterns of genetic diversity at a given neighborhood size as a result of differences in the underlying breeding model.
-
-\cjb{CJ: let's meet and talk again and talk about what edits to make here. I'm not sure what the best way to go is.}
-
 }
-
-\ak{ADK: I made some small edits but I'm pretty happy about how it reads}
 

### 2020-02-19 ms/spaceness.tex
added reviewer acknowledgements, misc typo fixes

@@
 because these quantities are more easily observable in practice than their ``effective'' versions.
-(Also, note that effective population density is lower than the actual population density,
+We also note that effective population density is lower than the actual population density,
 while the $\sigma$ we use is an underestimate,
@@
 We thank Brandon Cooper, Matt Hahn, Doc Edge, and others for reading and thinking about this manuscript. 
-% We thank the Hearth for having such good, nearby coffee, and
-% Falling Sky Brewing for creative support during manuscript drafting.
+Three anonymous reviewers and the suggestions from editors of GENETICS significantly improved the final manuscript. 
 CJB and ADK were supported by NIH award R01GM117241.

### 2020-02-19 ms/spaceness.tex
tv edits

@@
 
-Also note that $\sigma$ controls the dispersal of offspring away from only \emph{one} parent \llabel{ll:sigma_effective}
+ Note that $\sigma$ controls the dispersal of offspring away from only \emph{one} parent \llabel{ll:sigma_effective}
 (e.g., the seed parent for plants),
-while the usual population genetics definition of dispersal distance
-involves the distance to a randomly chosen one of the two parents.
+while population genetics usually defines "dispersal distance" to be the distance to a randomly chosen one of the two parents
 (thus taking into account the distance from pollen parent as well).
@@
 $(\sigma^2 + 2 \sigma^2)/2 = 3 \sigma^2 /2$.
-Also note that Wright's definition of neighborhood size should rightly use the \emph{effective}
+To match theory, neighborhood size should be defined in terms of \emph{effective}
 dispersal distance, i.e., the mean squared displacement along an axis
-between parent-child pairs found moving back along a lineage \citep{barton2002neutral}.
-Nonetheless, we use $\sigma$ and $\rho$ as defined here to compute $N_W$
+between parent-child pairs found moving back along a lineage \citep{barton2002neutral}, 
+and \emph{effective} population density \citep{Rousset1997}.
+However, we use $\sigma$ and $\rho$ as defined here to compute $N_W$
 because these quantities are more easily observable in practice than their ``effective'' versions.
-We also note that effective population density is lower than the actual population density,
-while the $\sigma$ we use is an underestimate,
-so our numbers may be close to the ``effective'' values used in theory.
 

### 2020-02-19 ms/second-review-responses.tex
final(?)

@@
 ``Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data''. 
+We have addressed reviewer comments below and made a number of changes to better explain our demographic parameters and alternate factors affecting GWAS transferability.
+Included here is a combined PDF including the revised text and response to reviewers, and a diff file comparing this revision and our previous submission (note that page and line numbers in the response to reviewers refers to the revised text, not the diff file).
+Thank you for your time and efforts.
 
@@
 Reviewer 3 raised two good points on our definition of neighborhood size and the interpetation of $\sigma$ -- first, that $\sigma$ refers to one- rather than two-dimensional distances between parent and offspring (this is consistent with previous theoretical work, and we have corrected our definition in the text), and second, that the use of separate draws for mate selection and dispersal means the total distance traveled by a locus along a pedigree link is greater for ``paternal'' than ``maternal'' material (taking ``maternal'' to mean the parent from which dispersal is calculated).
-Because our simulated genome is autosomal and individuals hermaphroditic and non-selfing, a locus will experience a mix of ``maternal'' and ``paternal'' transmission events along its genealogy.
-So we see this as a scaling issue rather than one of a ``bimodal'' dispersal kernal per se -- the expected axial displacement of a locus along a pedigree link is $\sqrt{3/2} \sigma$, rather than $\sigma$.
+However, our simulated genome is autosomal and individuals non-selfing, so a locus will experience a mix of ``maternal'' and ``paternal'' transmission events along its genealogy.
+We see this as a scaling issue rather than one of a ``bimodal'' dispersal kernal per se -- the expected axial displacement of a locus along a pedigree link is $\sqrt{3/2} \sigma$, rather than $\sigma$.
 (Note that although the parent-offspring displacement is a mixture of mean-zero Gaussians

### 2020-02-20 ms/spaceness_first_revision.tex
adding diff to first revision

--- 2020-02-18 ms/spaceness.tex
@@
 \input{review-response-commands}
-\newcommand{\responsefile}{second-review-responses.tex}  % name of the review reponses file
+\newcommand{\responsefile}{review-responses.tex}  % name of the review reponses file
 
@@
     Stepping-stone models reproduce some of these effects, but discretizing the landscape introduces artifacts which in some cases are exacerbated at higher resolutions.
-    The combination of low dispersal and clustered sampling causes demographic inference from the site frequency spectrum to infer more turbulent demographic histories, but averaged results across multiple simulations
-    revealed surprisingly little systematic bias.
+        The combination of low dispersal and clustered sampling causes demographic inference from the site frequency spectrum to infer more turbulent demographic histories, but averaged results across multiple simulations were surprisingly robust to isolation by distance.
     We also show that the combination of spatially autocorrelated environments and limited dispersal
@@
 Despite a long history of analytical work describing the genetics of populations distributed across continuous geography
-(e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation,Wilkins2002linearcoalescent,wilkins2004separation}), 
+(e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation,Wilkins2002linearcoalescent,wilkins2004separation}), \revpoint{AE}{1}
 much modern work still describes geographic structure as a set of discrete populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2019populations} or as an average over such discrete models \citep{Petkova2015,al2019estimating}.
+\revpoint{4}{3}
 For this reason, most population genetics statistics are interpreted with reference to discrete, well-mixed populations, and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000} with methods that assume these are randomly mating units.
@@
 that incorporates overlapping generations, local dispersal of offspring, and density-dependent survival. 
-We simulate chromosome-scale genomic data in tens of thousands of individuals from parameter regimes relevant to common subjects of population genetic investigation, and output the full genealogy and recombination history of all final-generation individuals.
+We simulate chromosome-scale genomic data in tens of thousands of individuals from parameter regimes relevant to common subjects of population genetic investigation such as humans and \textit{Drosophila}, and output the full genealogy and recombination history of all final-generation individuals.
 We use these simulations to test how sampling strategy interacts with geographic population structure to cause systematic variation in population genetic summary statistics 
@@
 A key early observation by \citet{Wright1946} is that this balance is often nicely summarized by the ``neighborhood size'',
-defined in two dimensions to be $N_W = 4 \pi \rho \sigma^2$,
-where $\sigma^2$ is one half of the mean squared parent--offspring distance
-and $\rho$ is population density
-(see \citet{Rousset1997} for further discussion of parameter definitions in one- and two-dimensional habitats) \llabel{axialdist}. 
+defined to be $N_W = 4 \pi \rho \sigma^2$, where $\sigma$ is the mean parent--offspring distance and $\rho$ is population density. 
 This can be thought of as proportional to the average number of potential mates for an individual (those within distance $2 \sigma)$,
@@
 which shifts reproductive output towards low-density regions,
-and produces total census sizes that fluctuate around an equilibrium. 
+and produces total census sizes that fluctuate around an equilibrium. \revpoint{3}{2}
 This also prevents the population clumping seen by \citet{Felsenstein1975}
-(Supplemental \autoref{fig:wm_v_dd})). 
+(Supplemental \autoref{fig:wm_v_dd})). \revpoint{3}{1}
 Such models have been used extensively in ecological modeling
 \citep{durrett1994importance,bolker1997using,law2003population,fournier2004microscopic,champer2019suppression}
-but rarely in population genetics, 
+but rarely in population genetics, \revpoint{AE}{2}
 where to our knowledge implementations of continuous space models
@@
 Each time step consists of three stages: reproduction, dispersal, and mortality. 
-To reduce the number of parameters we use the same parameter, denoted $\sigma$, 
+To reduce the parameter space we use the same parameter, denoted $\sigma$, 
 to modulate the spatial scale of interactions at all three stages by adjusting the standard deviation of the corresponding Gaussian functions. 
-Informally, we think of $\sigma$ as the ``dispersal distance'',
-although only one of those stages is dispersal.
+As in previous work \citep{Wright1943,Ringbauer2017}, $\sigma$ is equal to the mean parent-offspring distance.
 
@@
 Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
-in both the $x$ and $y$ coordinates. 
+in both the $x$ and $y$ coordinates. \revpoint{4}{8}
 Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
-with probability of choosing a neighbor at distance $d$ proportional to the Gaussian density
-with mean zero and standard deviation $\sigma$, which is $g(d) = \exp(-d^2 / 2 \sigma^2)/ (2 \sigma^2)$.
+with probability of choosing a neighbor at distance $d$ proportional to $\exp(-d^2 / 2 \sigma^2)$.
 
 To maintain a stable population, mortality increases with local population density.
-To do this we say that individuals at distance $d$ have a competitive interaction with strength $g(d)$.
+To do this we say that individuals at distance $d$ have a competitive interaction with strength $g(d)$,
+where $g$ is the Gaussian density with mean zero and standard deviation $\sigma$.
 Then, the sum of all competitive interactions with individual $i$ is $n_i = \sum_j g(d_{ij})$,
@@
 
-To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, and density-dependent fitness,  
+To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, and density-dependent fitness,  \revpoint{3}{6}
 we also implemented simulations identical to those above except that mates are selected uniformly at random from the population,
@@
 
-We ran 400 simulations for the spatial and random-mating models on a square landscape of width $W=50$  
+We ran 400 simulations for the spatial and random-mating models on a square landscape of width $W=50$  \revpoint{AE}{3}
 with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), 
@@
 we divided the mutation rate of $10^{-8}$ mutations per site per generation
-by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time. 
+by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time. \revpoint{2}{5}
 We verify that this procedure produces the same site frequency spectrum as applying mutations directly in SLiM in Figure \ref{fig:slim_v_msp_sfs},
@@
 
-Our demographic model includes parameters that control population density ($K$), 
+Our demographic model includes parameters that control population density ($K$), \revpoint{1}{2}
 mean life span ($L$), and dispersal distance ($\sigma$).
 However, nonlinearity of local demographic stochasticity causes
-actual realized averages of these demographic quantities to deviate from the specified values
+actual realized averages of these demographic quantitites to deviate from the specified values
 in a way that depends on the neighborhood size.
@@
 and used mean population density ($\rho$, census size divided by total area)
-to compute neighborhood size as $N_W = 4 \pi \rho \sigma^2$.  \llabel{ll:nw_defn}
+to compute neighborhood size as $N_W = 4 \pi \rho \sigma^2$. \revpoint{3}{6}
 To estimate generation times, we stored ages of the parents of every new individual born across 200 timesteps, 
@@
 
-Also note that $\sigma$ controls the dispersal of offspring away from only \emph{one} parent \llabel{ll:sigma_effective}
-(e.g., the seed parent for plants),
-while the usual population genetics definition of dispersal distance
-involves the distance to a randomly chosen one of the two parents.
-(thus taking into account the distance from pollen parent as well).
-This second component -- the distance between mates --
-has in our simulations a distribution that is of order $\sigma$ but that depends on the population's patchiness.
-If both between-mate distance and dispersal distance has variance $\sigma^2$ along each axis,
-then the mean squared distance to a randomly chosen parent along that axis would be
-$(\sigma^2 + 2 \sigma^2)/2 = 3 \sigma^2 /2$.
-Also note that Wright's definition of neighborhood size should rightly use the \emph{effective}
-dispersal distance, i.e., the mean squared displacement along an axis
-between parent-child pairs found moving back along a lineage \citep{barton2002neutral}.
-Nonetheless, we use $\sigma$ and $\rho$ as defined here to compute $N_W$
-because these quantities are more easily observable in practice than their ``effective'' versions.
-(Also, note that effective population density is lower than the actual population density,
-while the $\sigma$ we use is an underestimate,
-so our numbers may be close to the ``effective'' values used in theory.
-
 %%%%%%%%%%%%%
@@
 Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep[e.g.,][]{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among sampled chromosomes,
-defined to be the set of distances between adjacent sites that differ between the two chromosomes. 
+defined to be the set of distances between adjacent sites that differ between the two chromosomes. \revpoint{1}{1}
 The full distribution of lengths of IBS tracts for each pair of chromosomes was first calculated with a custom python function. 
@@
 
-The effects of sampling on summary statistic estimates were summarized by testing for differences in mean (ANOVA, \citet{Rcore2018}) and variance (Levene's test, \citet{Fox2011}) across sampling strategies for each summary statistic.
+The effects of sampling on summary statistic estimates were summarized by testing for differences in mean (ANOVA, \citep{Rcore2018}) and variance (Levene's test, \citep{Fox2011}) across sampling strategies for each summary statistic.
 
@@
 In all simulations, the phenotype of each individual
-is determined by drawing from a Gaussian distribution with standard deviation 10 
+is determined by drawing from a Gaussian distribution with standard deviation 10 \revpoint{1}{5}
 and a mean that may depend on spatial position.
@@
 we increased the mean phenotype from 100 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
-Concretely, the mean phenotype $p$ for an individual at position $(x,y)$ is $p=100 + 2x / 5$. 
+Concretely, the mean phenotype $p$ for an individual at position $(x,y)$ is $p=100 + 2x / 5$. \revpoint{4}{10}
 Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype to 120
@@
 with $p$-value below $0.05$ after adjusting for an expected false positive rate of less than 5\% \citep{benjamini2001control}. 
-We also examined $p$ values for systematic inflation 
+We also examined $p$ values for systematic inflation \revpoint{1}{6}
 by comparing to the values expected from a uniform distribution
@@
 Variation across runs occurs because,
-even though the parameters $K$ and $L$ that control population density and mean lifetime respectively \llabel{ll:K_more} 
+even though the parameters $K$ and $L$ that control population density and mean lifetime respectively \llabel{ll:K_more} \revpoint{1}{11}
 were the same in all simulations,
@@
 Run times for our model scale approximately linearly with neighborhood size (Figure \ref{fig:runtimes}), with the lowest neighborhood sizes reaching 30N generations in around a day and those with neighborhood size approaching 1,000 requiring up to three weeks of computation.
-As currently implemented, running simulations at neighborhood sizes more than 1,000 to coalescence is likely impractical, though running these models for more limited timescales and then ``recapitating'' the simulation using reverse-time simulation from the resulting tree sequence in msprime is possible \citep{haller2019treesequence}.
+As currently implemented running simulations at neighborhood sizes more than 1,000 to coalescence is likely impractical, though running these models for more limited timescales and then ``recapitating'' the simulation using reverse-time simulation from the resulting tree sequence in msprime is possible \citep{haller2019treesequence}.
 \llabel{runtimesxn}
@@
 Moreover, as we will show, spatial sampling strategies can affect summaries of genetic variation at least as strongly as the underlying population dynamics.
+\revpoint{1}{9}
 
@@
 -- as neighborhood size increases, the total proportion of variance explained by the first 10 PC axes declines from roughly 10\% to 4\% (Figure \ref{fig:gwas}B). 
-Essentially, PCA seems unable to effectively summarize the weak population structure present in large-neighborhood simulations given the sample sizes we tested, \revpoint{2}{2}
-but these populations continue to have enough spatial structure to create significant correlations between genotypes and the environment.
+Essentially, PCA seems unable to effectively summarize the weak population structure present in large-neighborhood simulations, but these populations continue to have enough spatial structure to create significant correlations between genotypes and the environment.
 A similar process can also be seen in the corner phenotype distribution, in which the count of significant SNPs initially declines as neighborhood size increases and then increases at approximately the point at which the proportion of variance explained by PCA approaches its minimum.
@@
 
-Previous studies have found that population structure and nonrandom sampling can create spurious signals of population bottlenecks when attempting to infer demographic history with microsatellite variation, summary statistics, or runs of homozygosity \citep{Chikhi2010,Stadler2009,Ptak2002,Mazet2015,leblois2006genetics}.
+Previous studies have found that population structure and nonrandom sampling can create spurious signals of population bottlenecks when attempting to infer demographic history with microsatellite variation, summary statistics, or runs of homozygosity \citep{Chikhi2010,Stadler2009,Ptak2002,Mazet2015}.
 Here we found that methods that infer detailed population trajectories through time based on the SFS and patterns of LD across the genome are also subject to this bias, with some combinations of dispersal and sampling strategy systematically inferring deep recent population bottlenecks and overestimating ancient $N_{e}$ by around a factor of 2.
 We were surprised to see that both stairwayplot and SMC++ can tolerate relatively strong isolation by distance -- i.e., neighborhood sizes of 20 -- and still perform well when averaging results across multiple simulations.
-(However, note the high amount of between-simulation variance seen in \autoref{fig:demography_supplement}.) \llabel{ll:variance_note}
 Inference in populations with neighborhood sizes over 20 was relatively unbiased unless samples were concentrated in the middle of the range (Figure \ref{fig:demography}).
@@
 This suggests that GWAS focused on finding strongly associated alleles for traits controlled by a limited number of variants in the genome are likely robust to the impacts of continuous spatial structure.
-However, methods that analyze the combined effects of thousands or millions of weakly associated variants such as polygenic risk scores \citep{wray2007prediction,international2009common} are likely to be affected by subtle population structure.
-\revpoint{2}{3}
-Indeed as recently identified in studies of genotype associations for human height in Europe \citep{Berg2018,Sohail2018}, \revpoint{2}{4}
-PC regression GWAS in modern human populations do include residual signal of population structure in large-scale analyses of polygenic traits.
-In addition to error associated with varying patterns of linkage disequilibrium and allele frequency among populations, the confounding of environmental and genetic effects on phenotypes introduced by population structure is expected to lead to low predictive power when polygenic scores are generated for populations outside the original GWAS cohort, \llabel{ldcaveats}
-%When attempting to make predictions across populations with different environmental exposures, polygenic risk scores affected by population structure can be expected to offer low predictive power, 
-as was shown in a recent study finding lower polygenic score predictive power outside European populations \citep{Martin2019}.
+However, methods that analyze the combined effects of thousands or millions of weakly associated variants such as polygenic risk scores \citep{Khera2018} are likely to be affected by subtle population structure.
+Indeed as recently identified in studies of genotype associations for human height in Europe \citep{Berg2018,Sohail2018}, PC regression GWAS in modern human populations do include residual signal of population structure in large-scale analyses of polygenic traits.
+When attempting to make predictions across populations with different environmental exposures, polygenic risk scores affected by population structure can be expected to offer low predictive power, as was shown in a recent study finding lower performance outside European populations \citep{Martin2019}.
 
@@
 
-We suggest a straightforward alternative for species in which the primary axes of population differentiation
-are associated with geography (note this is likely not the case for some modern human populations): 
+We suggest a straightforward alternative for species in which the primary axes of population differentiation is space 
+(note this is likely not the case for some modern human populations): 
 run a GWAS with spatial coordinates as phenotypes and check for $p$-value inflation or significant associations.
@@
 This approach has recently been taken with polygenic scores for UK Biobank samples in \citet{haworth2019apparent}, finding that scores are correlated with birth location even in this relatively homogenous sample.
+\revpoint{2}{8}
 Of course, it is possible that genotypes indirectly affect individual locations by adjusting organismal fitness and thus habitat selection across spatially varying environments, but we believe that this hypothesis should be tested against a null of stratification bias inflation rather than accepted as true based on GWAS results.
@@
   \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{POPRES data\\NE Europe} & 1,342 - 5,425 & Genetic & \citep{Ringbauer2017} \\ 
-  %\makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 240,000 & Survey & \citep{Rousset1997}\\ 
-  %\makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 360,000 & Genetic & \citep{Rousset1997}\\ 
+  \makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 240,000 & Survey & \citep{Rousset1997}\\ 
+  \makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 360,000 & Genetic & \citep{Rousset1997}\\ 
    \hline
@@
 others (e.g., number of segregating sites)
-by global $N_e$, or the ratio of the two. 
+are more strongly determined by global $N_e$ or by the ratio of the two.
+% These values should be compared to our simulation results with some caution, as though we expect neighborhood size to have similar effects across species of varying global $N_e$ \citep{Wright1946}, in our study we evaluated only a relatively small population of $\approx 10,000$.
 \llabel{popsizecaveat}
-In addition, definitions of "population density" in genetic versus ecological studies may lead to varying estimates of neighborhood size for a given species \llabel{rhocaveat}, and these empirical examples may be biased towards small-neighborhood species because few studies have quantified neighborhood size in species with very high dispersal or population density.
-
+In addition, these empirical examples are likely biased towards small-neighborhood species (because few studies have quantified neighborhood size in species with very high dispersal or population density).
 However, from the available data we find that neighborhood sizes in the range we simulated are fairly common across a range of taxa.
-At the extreme low end of empirical neighborhood size estimates we see some flowering plants, large mammals, and colonial insects like ants with neighborhood sizes less than roughly 100.
+At the extreme low end of empirical neighborhood size estimates we see some flowering plants, large mammals, and colonial insects like ants.
 Species such as this have neighborhood size estimates small enough that spatial processes are likely to strongly influence inference.
 These include some human populations such as the Gainj- and Kalam-speaking people of Papua New Guinea, in which the estimated neighborhood sizes in \citet{Rousset1997} range from 40 to 410 depending on the method of estimation.
+\revpoint{3}{7}
 Many more species occur in a middle range of neighborhood sizes between 100 and 1000 -- a range in which spatial processes play a minor role in our analyses under random spatial sampling but are important when sampling of individuals in space is clustered.
@@
 although the general predictions should be robust to similar forms of regulation.
-As is usual in population genetics, \revpoint{2}{6}
-the populations are entirely \emph{intrinsically} regulated.
+Since populations are still entirely \emph{intrinsically} regulated,
+our model still has a very strong ``population genetics'' flavor.
 Alternatively, population size could be regulated by interactions with other species
@@
 One facet of spatial variation that we did not address in this study is the confounding of dispersal and population density implicit in the definition of Wright's neighborhood size.
-Our simulations were run under constant densities, but \citet{guindon2016demographic} and \citet{Ringbauer2017}
+Our simulations were run under constant densities, but \citet{guindon2016demographic} and \citet{Ringbauer2017} \revpoint{3}{8}
 have shown that these parameters are identifiable under some continuous models.
@@
 \llabel{landscapesizecaveat}
-Indeed, our finding that stepping-stone and continuous-space models match
-in only certain aspects of genetic variation (Figure \ref{fig:grid_stats})
-shows that qualitatively similar models can produce different results dependent on the specific parameterizations used.
-While we believe our continuous model is a more appropriate depiction of 
-many species' demographies than lattice models,
-it is likely that some populations and breeding systems do more closely resemble a series of interconnected random-mating populations.
-As with all population models, the best approximation for any empirical system will depend on the natural history of the species in question.
-\llabel{bestmodelcaveat}
-Much additional work remains to be done to better understand how life history, range size,
-and habitat geometry interact to shape genetic variation in continuous space, which we leave to future studies.
+Much additional work remains to be done to better understand how these parameters interact to shape genetic variation in continuous space, which we leave to future studies.
 
@@
 Perhaps optimistically,
-we posit that process-driven descriptions of ancestry  \revpoint{2}{7}
+we posit that process-driven descriptions of ancestry 
 and/or more generalized unsupervised methods may be able to better account for carry out this task.
@@
 CJB and ADK were supported by NIH award R01GM117241. 
+\revpoint{AE}{4}
 
@@
 In general we find many of the qualitative trends are similar among continuous and stepping-stone models and that, at low neighborhood sizes, many  (but not all) statistics from stepping-stone models approach the continuous model as the resolution of the grid increases.
-For example, $\theta_\pi$ is lower in stepping-stone models 
-at low neighborhood sizes (i.e., low $m$),
-but increases to approach the continuous case as the resolution of the landscape increases.
-Similar patterns are observed for observed heterozygosity. However, $\theta_{W}$ behaves differently, showing a non-monotonic relationship with grid resolution.
+For example, $\theta_\pi$ is inflated at low neighborhood sizes (i.e., low $m$), and the extent of the inflation increases to approach the continuous case as the resolution of the landscape increases.
+Similar patterns are observed for $F_{IS}$ and observed heterozygosity.
+However, $\theta_{W}$ behaves differently, showing a non-monotonic relationship with grid resolution.
 This results in an increasingly positive Tajima's $D$ in grid simulations at small neighborhood sizes, to a much greater extent than seen in a continuous model.
 In contrast to $\theta_\pi$, increasing the resolution of the grid causes Tajima's $D$ to deviate \emph{more} from what is seen in the continuous model.
-Similarly, although $F_{IS}$ approaches the continuous case as the resolution of the grid increases at very small neighborhood sizes,
-at intermediate neighborhood sizes the continuous case best matches intermediate grid resolutions.
 

### 2020-04-10 ms/spaceness_final.tex
adding final ms files

--- 2020-02-19 ms/spaceness.tex
@@
 % Use the documentclass option 'lineno' to view line numbers
-
-% adding commenting commands
-\newif\ifcomments
-\commentstrue
-%\commentsfalse
-\newcommand{\cjb}[1]{\ifcomments{{\color{orange} \it (#1)}}\else{}\fi}
-\newcommand{\plr}[1]{\ifcomments{{\color{purple} \it (#1)}}\else{}\fi}
-\newcommand{\ak}[1]{\ifcomments{{\color{red} \it (#1)}}\else{}\fi}
 
@@
 For instance, if mean viability or fecundity is spatially autocorrelated, then limited geographic dispersal will lead to parent--offspring correlations.
-Furthermore, nearby individuals will be more closely related than an average random pair, so drawing multiple samples from the same area of the landscape will represent a biased sample of the genetic variation present in the whole population \citep{Stadler2009}.% It has long been appreciated that this model misspecification subjects downstream inferences to bias, but the extent and nature of these effects remain largely uninvestigated.
+Furthermore, nearby individuals will be more closely related than an average random pair, so drawing multiple samples from the same area of the landscape will represent a biased sample of the genetic variation present in the whole population \citep{Stadler2009}.
 
@@
 
-% The best-studied approaches to population genetics in continuous space were developed by \citet{Wright1942,Wright1943} and \citet{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals.
 The degree to which genetic relationships are geographically correlated
@@
 and $\rho$ is population density
-(see \citet{Rousset1997} for further discussion of parameter definitions in one- and two-dimensional habitats) \llabel{axialdist}. 
+(see \citet{Rousset1997} for further discussion of parameter definitions in one- and two-dimensional habitats). 
 This can be thought of as proportional to the average number of potential mates for an individual (those within distance $2 \sigma)$,
@@
 estimates range from 40 to over 5,000 depending on the population and method of estimation (Table~\ref{table:NStable}).
-%\citet{Maruyama1972} found that the rate of decline in genetic diversity in a two-dimensional continuous population approaches the random mating expectation when $\rho\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.
 
@@
 The question of what regulates real populations has a long history and many answers
-\citep[e.g.,][]{lloyd1967crowding,antonovics1980ecological,crawley1990population}, \llabel{ecolcites}
+\citep[e.g.,][]{lloyd1967crowding,antonovics1980ecological,crawley1990population}, 
 but it is clear that populations must at some point have density-dependent feedback on population size,
@@
 Since $g$ is a probability density, $n_i$ is an estimate of the number of nearby individuals per unit area.
-Then, given a per-unit carrying capacity $K$, \llabel{ll:K_defn}
+Then, given a per-unit carrying capacity $K$, 
 the probability of survival until the next time step for individual $i$ is
@@
 \end{equation}
-We chose this functional form so that the equilibrium population density per unit area is close to $K$, \llabel{ll:K_again}
+We chose this functional form so that the equilibrium population density per unit area is close to $K$, 
 and the mean lifetime is around $L$;
-for more description see the Appendix. % \ref{apx:demographic_model}.
+for more description see the Appendix.
 
@@
 and used mean population density ($\rho$, census size divided by total area)
-to compute neighborhood size as $N_W = 4 \pi \rho \sigma^2$.  \llabel{ll:nw_defn}
+to compute neighborhood size as $N_W = 4 \pi \rho \sigma^2$.  
 To estimate generation times, we stored ages of the parents of every new individual born across 200 timesteps, 
@@
 we tracked the lifetime total number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period,
-% subset the resulting table to include only the last timestep recorded for each individual,
 and calculated the variance in number of offspring across all individuals in timesteps 50-100.
@@
 
- Note that $\sigma$ controls the dispersal of offspring away from only \emph{one} parent \llabel{ll:sigma_effective}
-(e.g., the seed parent for plants),
+ Note that $\sigma$ controls the dispersal of offspring away from only \emph{one} parent (e.g., the seed parent for plants),
 while population genetics usually defines "dispersal distance" to be the distance to a randomly chosen one of the two parents
@@
 \centering
-\includegraphics{figures/sampling_maps.pdf}
+\includegraphics{final_figures/Fig1_sampling_maps.pdf}
 \caption{Example sampling maps for 60 individuals on a 50$\times$50 landscape for midpoint, point, and random sampling strategies, respectively.}
@@
 the correlation coefficient between the logarithm of the spatial distance and the proportion of identical base pairs across pairs of individuals.
-% as an isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distance analogous to \citet{Rousset1997}'s approach, which we summarized as
 
@@
 The full distribution of lengths of IBS tracts for each pair of chromosomes was first calculated with a custom python function. 
-%\url{https://github.com/cjbattey/spaceness/blob/master/scripts/slimtools.py}. 
 We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $10^6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.
@@
 We also examined how varying levels of isolation by distance impacted the variance of $N_{e}$ estimates by calculating the standard deviation of $N_{e}$ from each best-fit model.
-% and plotting these against neighborhood size.
-
-%In preliminary runs we found that inferred population histories were highly variable even when simulating under a coalescent model, suggesting that some of the differences in demographic estimates for spatial models are caused by the behavior of the optimization algorithm rather than bias in the SFS caused by spatial mate choice and dispersal.
-To separate these effects we ran 100 coalescent simulations with constant population size $6.1\times 10^{-3}$ (the mean $N_{e}$ of random-mating SLiM models estimated from $\Theta_{\pi}$) and fit stairwayplot models using the same script as for our spatial models.
-All coalescent simulations were performed using msprime \citep{Kelleher2016}.
-We then calculated the standard deviation of inferred $N_{e}$ in each stairwayplot model to summarize the degree of fluctuation around the simulated population size, and asked if standard deviations were higher in spatial relative to coalescent models with a one-tailed t-test.
 
@@
 
+\section{Data Availability}
+Scripts used for all analyses and figures are available at \url{https://github.com/kern-lab/spaceness}.
+
 %%%%%%%%%%%%%%%%%%
@@
 Because our simulation is paramaterized on an individual level, these population parameters emerge as a property of the interactions among individuals rather than being directly set.
-\llabel{emergentparams}
 Variation across runs occurs because,
-even though the parameters $K$ and $L$ that control population density and mean lifetime respectively \llabel{ll:K_more} 
+even though the parameters $K$ and $L$ that control population density and mean lifetime respectively  
 were the same in all simulations,
@@
 Furthermore, census size increases in spatial models as neighborhood size increases from 2 to 10.
-This may reflect an Allee effect \citep{Allee1949} in which some individuals are unable to find mates when the mate selection radius is very small.  \llabel{censusscaling}
+This may reflect an Allee effect \citep{Allee1949} in which some individuals are unable to find mates when the mate selection radius is very small.
 
@@
 \centering
-\includegraphics{figures/pop_params.pdf}
+\includegraphics{final_figures/Fig2_pop_params.pdf}
 \caption{Genealogical parameters from spatial and random mating SLiM simulations, by neighborhood size.}
@@
 \centering
-\includegraphics{figures/runtimes.pdf}
+\includegraphics{final_figures/Fig3_runtimes.pdf}
 \caption{Run times of continuous space simulations with landscape width 50 and expected density 5 under varying neighborhood size.
@@
 As currently implemented, running simulations at neighborhood sizes more than 1,000 to coalescence is likely impractical, though running these models for more limited timescales and then ``recapitating'' the simulation using reverse-time simulation from the resulting tree sequence in msprime is possible \citep{haller2019treesequence}.
-\llabel{runtimesxn}
 
@@
 \centering
-\includegraphics[width=\textwidth]{figures/sfs_w_sumstats_loglog.pdf}
+\includegraphics[width=\textwidth]{final_figures/Fig4_sfs_w_sumstats_loglog.pdf}
 \caption{Site frequency spectrum (A; note axes are log-scaled) and summary statistic distributions (B) by sampling strategy and neighborhood size.}
@@
 \centering
-\includegraphics[width=.5\textwidth]{figures/ibs_cumdists_w_rm.pdf}
+\includegraphics[width=.5\textwidth]{final_figures/Fig5_ibs_cumdists_w_rm.pdf}
 \caption{
@@
 \centering
-\includegraphics[width=\textwidth]{figures/distance_by_DAF.pdf}
+\includegraphics[width=\textwidth]{final_figures/Fig6_distance_by_DAF.pdf}
 \caption{
@@
 \centering
-\includegraphics[width=\textwidth]{figures/demography_rollmedian_w_variance.pdf}
+\includegraphics[width=\textwidth]{final_figures/Fig7_demography_rollmedian_w_variance.pdf}
 \caption{A: Rolling median inferred $N_{e}$ trajectories for stairwayplot and smc++ across sampling strategies and neighborhood size bins.
@@
 \centering
-\includegraphics[width=\textwidth]{figures/gwas_summary_loglog_qqlines.pdf}
+\includegraphics[width=\textwidth]{final_figures/Fig8_gwas_summary_loglog_qqlines.pdf}
     \caption{Impacts of spatially varying environments and isolation by distance on linear regression GWAS.
@@
 We were surprised to see that both stairwayplot and SMC++ can tolerate relatively strong isolation by distance -- i.e., neighborhood sizes of 20 -- and still perform well when averaging results across multiple simulations.
-(However, note the high amount of between-simulation variance seen in \autoref{fig:demography_supplement}.) \llabel{ll:variance_note}
+(However, note the high amount of between-simulation variance seen in \autoref{fig:demography_supplement}.) 
 Inference in populations with neighborhood sizes over 20 was relatively unbiased unless samples were concentrated in the middle of the range (Figure \ref{fig:demography}). 
@@
 Though this degree of variation in population size is certainly meaningful in an ecological context, it is relatively minor in population genetic terms.
-\llabel{ibdnesxn}
 Methods directly assessing haplotype structure in phased data example, \citep[e.g.,][]{browning2015accurate} are thought to provide increased resolution for recent demographic events,
@@
 If environmental factors affecting the phenotype of interest also vary over space, then allele frequencies and environmental exposures will covary over space.
-\llabel{ll:1:11:2}
 In this scenario an uncorrected GWAS will infer genetic associations with a purely environmental phenotype at any site in the genome that is differentiated over space, and the relative degree of bias will be a function of the degree of covariation in allele frequencies and the environment (i.e., Figure \ref{fig:gwas}C, bottom panel).
@@
 PC regression GWAS in modern human populations do include residual signal of population structure in large-scale analyses of polygenic traits.
-In addition to error associated with varying patterns of linkage disequilibrium and allele frequency among populations, the confounding of environmental and genetic effects on phenotypes introduced by population structure is expected to lead to low predictive power when polygenic scores are generated for populations outside the original GWAS cohort, \llabel{ldcaveats}
+In addition to error associated with varying patterns of linkage disequilibrium and allele frequency among populations, the confounding of environmental and genetic effects on phenotypes introduced by population structure is expected to lead to low predictive power when polygenic scores are generated for populations outside the original GWAS cohort,
 %When attempting to make predictions across populations with different environmental exposures, polygenic risk scores affected by population structure can be expected to offer low predictive power, 
@@
 In summary, spatial covariation in population structure and the environment confounds the interpretation of GWAS $p$-values, and correction using principal components is insufficient to fully separate these signals for polygenic traits under a variety of environmental and population parameter regimes.
-Other GWAS methods such as mixed models \citep{Kang2008} \llabel{mixedmodelcallout} may be less sensitive to this confounding, but there is no obvious reason that this should be so. 
+Other GWAS methods such as mixed models \citep{Kang2008} may be less sensitive to this confounding, but there is no obvious reason that this should be so. 
 One approach to estimating the degree of bias in GWAS caused by population structure is LD score regression \citep{Bulik-Sullivan2015}. 
@@
 by global $N_e$, or the ratio of the two. 
-\llabel{popsizecaveat}
-In addition, definitions of "population density" in genetic versus ecological studies may lead to varying estimates of neighborhood size for a given species \llabel{rhocaveat}, and these empirical examples may be biased towards small-neighborhood species because few studies have quantified neighborhood size in species with very high dispersal or population density.
+In addition, definitions of "population density" in genetic versus ecological studies may lead to varying estimates of neighborhood size for a given species, and these empirical examples may be biased towards small-neighborhood species because few studies have quantified neighborhood size in species with very high dispersal or population density.
 
@@
 requires some kind of control on population sizes,
-or else the population will either die out or grow very large after a sufficiently long period of time. \llabel{demog_disco}
+or else the population will either die out or grow very large after a sufficiently long period of time. 
 The usual choice of population regulation for population genetics --
@@
 Similarly, though the scaling effects of dispersal we show in Figure \ref{fig:sumstats} should occur in populations of any total size, other aspects such as the number of segregating sites are also likely affected by the total landscape size (and so total census size).
-\llabel{landscapesizecaveat}
 Indeed, our finding that stepping-stone and continuous-space models match
@@
 As with all population models, the best approximation for any empirical system will depend on the natural history of the species in question.
-\llabel{bestmodelcaveat}
 Much additional work remains to be done to better understand how life history, range size,
@@
 
-\section{Data Availability}
-Scripts used for all analyses and figures are available at \url{https://github.com/kern-lab/spaceness}.
-
 \section{Acknowledgements}
@@
 
-\bibliography{spaceness,more_refs}
+\bibliography{spaceness_final}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%
-\section{Comparisons with Stepping-Stone Models}
-\llabel{apx:stepping_stone}
+\clearpage
+\section{Appendix 1}
+
+\subsection{Comparisons with Stepping-Stone Models}
 
@@
 \centering
-\includegraphics[width=\textwidth]{figures/grid_ss_curves_edit.pdf}
+\includegraphics[width=\textwidth]{final_figures/FigA1_grid_ss_curves_edit.pdf}
 \caption{Summary statistics for 2-dimensional coalescent stepping-stone models with fixed total $N_{e}$ and varying numbers of demes per side.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Demographic model}
-\llabel{apx:demographic_model}
+\subsection{Demographic model}
 Local population regulation is controlled by two parameters, $L$, and $K$.
@@
 
-\includereviews
-
 \end{document}
